### 2021-08-15 README.Rmd
Initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# assessing.sealevel.dating

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/isakro/assessing.sealevel.dating/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your broswer or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2021-08-15 analysis/paper/paper.Rmd
Initial commit

---
title: "Title Goes Here"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

# Introduction

# Background

# Methods

# Results

```{r get-data, eval = FALSE}
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

```{r demo-inline-code}
x <- round(pi, 2)
```

# Discussion

# Conclusion

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2021-09-10 analysis/paper/supplementary_material.Rmd
Moved functions to script 3. Script 4 will be used to analyse the sites

---
title: "Supplementary material"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
institute:
  - UofO: University of One Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

```{r data_functions, eval = FALSE}
# Load in required functions and prepared data
library(here)
```

```{r data_functions, eval = TRUE, warning = FALSE}
# Load in required functions and prepared data
source(here("analysis/script/02script.R"))
source(here("analysis/script/03script.R"))
load(here("analysis/data/derived_data/01data.RData"))
load(here("analysis/data/derived_data/02data.RData"))
```

```{r , cache= = TRUE}

```

```{r demo-inline-code}

```

\newpage

# References

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

### 2021-09-10 analysis/script/02script.R

@@ -1,4 +1,5 @@
 library(tidyverse)
+library(here)
 library(sf)
 library(tmap)
 library(cowplot)
@@ -9,19 +10,19 @@ load(here::here("analysis/data/derived_data/01data.RData"))
 
 # Read in displacement curves
 arendal_raw <- read.csv(
-  here::here("analysis/data/raw_data/displacement_curves/bjornebu.csv"))
+  here("analysis/data/raw_data/displacement_curves/bjornebu.csv"))
 tvedestrand_raw <- read.csv(
-  here::here("analysis/data/raw_data/displacement_curves/hanto.csv"))
+  here("analysis/data/raw_data/displacement_curves/hanto.csv"))
 larvik_raw <- read.csv(
-  here::here("analysis/data/raw_data/displacement_curves/gunnarsrod.csv"))
+  here("analysis/data/raw_data/displacement_curves/gunnarsrod.csv"))
 skoppum_raw <- read.csv(
-  here::here("analysis/data/raw_data/displacement_curves/skoppum.csv"))
+  here("analysis/data/raw_data/displacement_curves/skoppum.csv"))
 
 # Read in coast polgyon
-coast <- read_sf(here::here("analysis/data/raw_data/coast.gpkg"))
+coast <- read_sf(here("analysis/data/raw_data/coast.gpkg"))
 
 # Read in isobase centre points
-centpts <- read_sf(here::here("analysis/data/raw_data/isobase_centrepts.gpkg"))
+centpts <- read_sf(here("analysis/data/raw_data/isobase_centrepts.gpkg"))
 
 #### Setting up displacement curves ####
 
@@ -149,49 +150,7 @@ isopolys$isobase2 <- c("Tvedestrand", "Arendal", "Tvedestrand", "Larvik",
                        "Larvik")
 isopolys$dir_rel_1 <- c("sw", "sw", "sw", "sw", "nw")
 
-# Define function to interpolate displacement curve for a given location,
-# based on distance to curves on two provided isobases on the
-# north-east to south-west axis perpendicular to the isobases.
-interpolate_curve <- function(years, isobase1, isobase2, target, dispdat,
-                              isodat, direction_rel_curve1){
-
-  # Distance between "northern" and "southern" isobase
-  dist <- st_distance(filter(isodat, name == isobase1),
-                      filter(isodat, name == isobase2))
-
-  curve1 <- filter(dispdat, name == isobase1)
-  curve2 <- filter(dispdat, name == isobase2)
-
[…]

### 2021-09-10 analysis/script/03script.R

@@ -1,19 +1,13 @@
-library(here)
-library(tidyverse)
-library(sf)
-library(ggridges)
-library(tmap)
-library(tmaptools)
-library(cowplot)
-library(terra)
-library(oxcAAR)
-library(vapour)
-library(topoDistance)
-quickSetupOxcal()
-
-source(here("analysis/script/02script.R"))
-load(here("analysis/data/derived_data/01data.RData"))
-load(here("analysis/data/derived_data/02data.RData"))
+# Function for creating a bounding box polygon around a feature,
+# the size of which can be adjusted. Used for plotting purposes
+# and ease of raster handling.
+bboxpoly <- function(feature, xy_adjustment) {
+  feature_bbox <- st_bbox(feature)
+  feature_bbox[1:2] <- feature_bbox[1:2] - xy_adjustment
+  feature_bbox[3:4] <-feature_bbox[3:4] + xy_adjustment
+  feature_bboxpoly <- st_as_sf(st_as_sfc(feature_bbox))
+  return(feature_bboxpoly)
+}
 
 # Function to identify and load the correct raster
 load_raster <- function(dtmfolder, sitelimit) {
@@ -48,6 +42,48 @@ load_raster <- function(dtmfolder, sitelimit) {
 
   # Return the correct raster
   return(rast(craster$path))
+}
+
+# Define function to interpolate displacement curve for a given location,
+# based on distance to curves on two provided isobases on the
+# north-east to south-west axis perpendicular to the isobases.
+interpolate_curve <- function(years, isobase1, isobase2, target, dispdat,
+                              isodat, direction_rel_curve1){
+
+  # Distance between "northern" and "southern" isobase
+  dist <- st_distance(filter(isodat, name == isobase1),
+                      filter(isodat, name == isobase2))
+
+  curve1 <- filter(dispdat, name == isobase1)
+  curve2 <- filter(dispdat, name == isobase2)
+
+  # Difference in displacement per meter between the the curves,
+  # upper confidence limit
+  prm_u <- (dplyr::select(curve1, upperelev) -
+              dplyr::select(curve2, upperelev))/ as.numeric(dist)
+  # Difference in difference per meter, lower confidence limit
+  prm_l <- (dplyr::select(curve1, lowerelev) -
+              dplyr::select(curve2, lowerelev))/ as.numeric(dist)
+
+  # Distance to target isobase from isobase of curve 1
[…]

### 2021-09-10 analysis/script/04script.R

@@ -1,0 +1,71 @@
+library(here)
+library(tidyverse)
+library(sf)
+library(ggridges)
+library(tmap)
+library(tmaptools)
+library(cowplot)
+library(terra)
+library(oxcAAR)
+library(vapour)
+library(topoDistance)
+quickSetupOxcal()
+
+# Load required functions and prepared data
+source(here("analysis/script/03script.R"))
+load(here("analysis/data/derived_data/01data.RData"))
+load(here("analysis/data/derived_data/02data.RData"))
+
+# Specify path to raster data (currently stored locally)
+dtmpath <- "/home/isak/phd/eaa_presentation/dtm1"
+
+# Assign closest isobases to site data (largest = TRUE makes sure that
+# sites overlapping isobase polygons are only ascribed the one they overlap the
+# most, so as not to create duplicate records).
+sites_sa <- st_join(st_make_valid(sites_sa), isopolys,
+                    join = st_intersects, largest = TRUE)
+
+# Example site
+sitename <- "Langangen Vestgård 1"
+
+sitel <- filter(sites_sa, name == sitename)
+siter <- filter(rcarb_sa, site_name == sitename)
+
+# Model dates
+datedat <- model_dates(siter)
+
+# Interpolate displacement curve to the site location
+sitecurve <- interpolate_curve(years = xvals,
+                               isobase1 = sitel$isobase1,
+                               isobase2 = sitel$isobase2,
+                               target = sitel,
+                               dispdat = displacement_curves,
+                               isodat = isobases,
+                               direction_rel_curve1 = sitel$dir_rel_1)
+# Add site name
+sitecurve$name <- sitename
+
+# Load correct regional raster
+dtm <- load_raster(dtmpath, sitel)
+
+# Create bounding box polygon
+location_bbox <- bboxpoly(sitel, 250)
+
+# Use this to clip the dtm to the site area
+sitearea <- terra::crop(dtm, location_bbox)
+
+# Simulate sea-level and retrieve distances (takes some time)
+output <- sample_shoreline(samps = 1000, sitel, sitecurve, sitearea,
+                           posteriorprobs)
+
+# Generate grid with dtm resolution holding number of overlaps for each cell
+# (also takes quite some time to execute)
+simsea <- sea_overlaps(sitearea, output$seapol)
+
+# Create plots
+datplot <- plot_dates(datedat, sitename)
+smap <- tmap_grob(shore_plot(simsea, sitel))
[…]

### 2021-09-10 README.Rmd
Established OSF project

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "assessing.sealevel.dating"
 ```
@@
 
-This repository contains the data and code for our paper:
+<!-- This repository contains the data and code for our paper: -->
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+<!-- > Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx> -->
 
-Our pre-print is online here:
+<!-- Our pre-print is online here: -->
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+<!-- > Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Roalkvam, I (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/10.17605/OSF.IO/7F9SU>
 

### 2021-09-10 analysis/script/02script.R

@@ -6,7 +6,7 @@ library(cowplot)
 library(era)
 
 # Load radiocarbon and site data from the first script
-load(here::here("analysis/data/derived_data/01data.RData"))
+load(here("analysis/data/derived_data/01data.RData"))
 
 # Read in displacement curves
 arendal_raw <- read.csv(
@@ -143,8 +143,8 @@ plot_grid(isocurves_plot, iso_map)
 outline <- st_as_sf(st_as_sfc(centpts_bb))
 # Then split this by the isobases
 isopolys <- st_collection_extract(lwgeom::st_split(outline, isobases))
-# Then assign isobasenames to the polygons and the direction (in case of ne of
-# both).
+# Then assign isobasenames to the polygons and the direction (in case both are
+# to the north-east).
 isopolys$isobase1 <- c("Arendal", "Tvedestrand", "Larvik", "Halden", "Halden")
 isopolys$isobase2 <- c("Tvedestrand", "Arendal", "Tvedestrand", "Larvik",
                        "Larvik")

### 2021-09-10 analysis/script/eaa.R

@@ -35,6 +35,65 @@ countries <- st_read(here::here('analysis/data/raw_data/naturalearth_countries.g
 sites_sa <- st_join(st_make_valid(sites_sa), isopolys, join = st_intersects, largest = TRUE)
 site_pts <- st_centroid(sites_sa)
 
+
+model_rcarb <- function(data){
+
+  # Format individual dates for R_Date in Oxcal
+  dates <- R_Date(data[["lab_code"]], data[["c14_bp"]], data[["error"]])
+
+  # Manually compile rest of OxCal code
+  oxcal_code <- paste0('
+  Plot()
+   {
+    Sequence()
+    {
+     Boundary("Start");
+     Phase()
+     {
+    ',
+                       dates,
+                       '
+     Sum("Sum");
+     };
+     Boundary("End");
+    };
+  };')
+
+  # Run OxCal code - wrapped this in some functions to  not print to
+  # rmarkdown
+  invisible(capture.output(oxcal_exec <- executeOxcalScript(oxcal_code)))
+  oxcal_read <- readOxcalOutput(oxcal_exec)
+  # Note that parseOxcalOutput() defaults to only returning individual dates,
+  # specified with only.R_Date
+  oxcal_data <- parseOxcalOutput(oxcal_read, only.R_Date = FALSE)
+
+  # Assemble data frame to be returned
+  posterior <- data.frame(
+    probabilities = oxcal_data$Sum$posterior_probabilities$probabilities,
+    date_grid = oxcal_data$Sum$posterior_probabilities$dates,
+    site = data[["site_name"]][1])
+
+  # Return data
+  return(posterior)
+}
+
+# Apply modelling function to the radiocarbon dates, grouped by site name
+posteriors <- rcarb_sa %>%
+  group_by(site_name) %>%
+  group_map(~ model_rcarb(.), .keep = TRUE)
+
+rcarb_data <- do.call(rbind, posteriors)
+
+ggplot(rcarb_data) +
+  geom_ridgeline(aes(x = date_grid, y = site, height = probabilities * 100,
+                     fill = site,
+                     colour = site)) +
+  xlab("Cal. BCE") +
+  theme_classic() +
+  theme(axis.title.y = element_blank()) +
+  guides(fill =  guide_legend(title.position = "top", title.hjust = 0.5),
+         colour = FALSE)
+
 # Create bounding box around the sites, both for extent indicator on the
 # overview map and to create basemap for the site overview.
 sitbbox2 <- st_bbox(site_pts)

### 2021-10-02 analysis/paper/paper.Rmd
Both Dybdalshei sites appear fine in terms of rasters and 250 m distance

@@
 ---
-title: "Title Goes Here"
+title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
 author:
@@
 # Introduction
+The association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C.
+Brøgger and his son, the archaeologist A.W.
+Brøgger, in their respective theses both published in 1905 (see also @maccurdy1906).
+Ever since, this finding has been fundamental to Stone Age research in Norway.
+One of its major implications is that relative sea-level change can be used for the dating of the sites.
+This is based on the fact that since most Stone Age sites were situated on or close to the contemporary shoreline during their original occupation, a reconstruction of the trajectory of shoreline displacement can be used to date the sites based on their elevation relative to the present day sea level.
+The method of shoreline dating is still fundamental to most Stone Age research in Norway, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
+Given the generally coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most accurate method by which one can hope to date the sites.
+Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age.
+This is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line.
+Consequently, adherence or deviation from this pattern has major implications for the socio-economic foundations of the societies in question.
+
+Despite the fundamental role of shoreline-dating in Norway, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement.
+The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway.
+The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age.
+This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement, it can potentially identify societal shifts if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the region.
 
 # Background
+After the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, meaning that the relative sea level has been dropping throughout prehistory.
+In other areas of Norway the situation is sometimes more complicated.
+In some areas relative sea-level change has transitioned between regression and transgression phases, and the archaeological remains can be destroyed.
+In other areas, the sea-level has been stable over longer periods of time, meaning coastal locations have often been reused multiple times over long time-spans, creating a palmpiset of settlement events that are impossible to disentangle.
+Comparatively, given the continuous and relatively rapid shoreline regression in south-eastern Norway, any one location in the region has only been shore-bound within a relatively limited time-span.
+As this general displacement trajectory is the same throughout the entirety of prehistory, this makes the region especially useful for testing of the assumption of a shore-bound settlement and to identify potential deviations from this pattern.
+
+The most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves.
+If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this.
+In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of of the overlap between shoreline displacement curves and the radiocarbon probability distributions.
+Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensisble if the site was *not* shore bound [e.g.
+@jaksland2014].
+This can for example pertain to accessibility.
+If the site is situated on a ledge in a steep and jagged area of the present day landscape, it would make intuitive sense that the location was used when the ocean reached the elevation of the ledge and the site would have been accessible by means of watercraft.
+This line of reasoning is often also used in survey projects which typically do not produce material suitable for radiocarbon dating.
+The evaluation of the degree to which sites were shorebound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations.
+Although it appears that the arguments for such site locations is sensible and for a large part appear to be correct, few attempts at a comprehensive evalation and quantification of this tendency has been done.
+
+One of the only larger scale evaulations of this relationship in Norwegian archaeology has been done by Solheim and collegues (@solheim20, see also @solheim2020), who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
+They found a overlap of the confidence bands of the radiocarbon dates with the shoreline displacement curve for 86.5 % of the sites.
+However, in several of the cases where there was a discrepancy, the
+
+Given that the aim of this study is to test whether or not there is a match between dates provided by means of shoreline displaceent curves and radiocarbon dates, it is appropriate to evaulate this relationship by means of statistical equivalence tests.
+This effectively shifts the common direction of the null-hypothesis, giving H0: The date given by means of radiocarbon dating and by shoreline displacement are different.
+Ha: The date given by means of radiocarbon dating and by shoreline displacement are the same.
+While this consitutes the formal dimension of the study, there must necessarily also be a qualitative and more subjectivly driven dimension as well.
+As was exemplified in Solheims study, and which has been convincingly argued in several individual studies, many of the sites where there is a mismatch between the radiocarbon dates and the shoreline dates, the assemblages are clearly related to the shoreline date rather than the radiocarbon dates.
+Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates or whether they are the result of contaminated samples or by pedoturbation can be difficult to evaluate and will here be entirely dependent on the discretion of the original excavation report.
+While such evaluations can be difficult, it is nonetheless a necessity that will be held strictly separate from the formal equivalence tests in the analysis to follow.
 
 # Methods
+Relative sea level is so named because the parameters of interest is the height of the sea surface relative to its present level.
+Sea level is here understood as the distance from the solid surface of the Earth to the mean sea level as averaged over time, also known as the geoid.
+Variation in this distance is determined by effects that can be divided into three main categories.
+The first of these is eustatic sea-level change.
+Eustatic sea level is understood as the sea level if the water has been evenly distributed across the Earth's surface and without accounting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself.
+The eustatic sea-level is most notably impacted by glaciation and de-glaciation which can either bind or release large amounts of water into the oceans, as well as volumetric changes in the Earth's crust following from .
+The second category is that of istostasy, which pertains to adjustments in the Earth's crust that results from processes to regain gravitational equilibrium relative to the underlying viscous mantle.
+This is typically the result of erosion of the crust which causes its weight to be redistributed and the surface to be pushed upwards, or glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, which either causes the lithosphere to be pushed into the mantle or to rebound and to lift upwards.
+Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.
 
-# Results
+This study necessaitated the compilation of multiple relative sea-level curves developed for different parts of the region.
+Most of the curves come provided with a confidence interval which is largely based on an evaluation by the geologists who developed the curve.
+The reason why these confidence intervals are not strictly mathematically derived is that the curves do not only contain uncertainy as related to radiometric dates, which are fairly well understood, but also hold potential error as related to the interpretation of sediment cores, placement of samples,  direction and adjustment to a common isobase based on the direction and tilt of the variation in rebound intensity, as well as, to name but a few.
+The error bands are consequently complex, and cannot necessarily be directly compared between curves as these have been developed at different times, using a wide range of different methods, and often using older and outdated calibration curves for both marine and terrestrial samples.
+However, any reinterpretation and standardisation of the curves is also hampered by a range of issues.
+The original data or necessary details about the sampling processes are not always available, and a reinterpretation based soley on radiometric dates, without taking into account the considerations and evaluations of the field geologists can be entirely misleading.
+In a rudimentary attempt to account for these differences when compiling the curves, the width of the confidence band for the sea level at a single date across the entire region is given as the largerst uncertainty of any one of the curves at that point in time.
+This averages to an uncertainty of +/- calendar years.
+This uncertainty is uniformly distributed within the confidence band.
+
+The goal is to test the synchroneity of two space-time processes, the occupation of a site and continuously shifting sea-levels.
+The synchroneity of these processes is tested by the comparison of singular events represented by the death of the dated organic materials.
+Additionally, defining what shore bound settlement represents can be problematic.
+In the literature this is often qualified with statements along the lines of "the sites were located on or close to the contemporenous shore-line".
+What constitutes close is of course highly variabl,e complicated further by the fact that vertical changes in sea level can results in widely different effects on the horisontal distance between a given point and the shore-line.
+Other elements of interest is the tidal range, and the exposure of the location in question, which can also have an impact on the distance of the site to the mean ocean surface.
 
@@
 ```
-
-# Discussion
-
-# Conclusion
 

### 2021-10-04 analysis/paper/paper.Rmd
A few more sites. Remembered to scale plot height when multiphase sites.

@@
 One of its major implications is that relative sea-level change can be used for the dating of the sites.
-This is based on the fact that since most Stone Age sites were situated on or close to the contemporary shoreline during their original occupation, a reconstruction of the trajectory of shoreline displacement can be used to date the sites based on their elevation relative to the present day sea level.
-The method of shoreline dating is still fundamental to most Stone Age research in Norway, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
+This is based on most Stone Age sites were situated on or close to the contemporary shoreline during their original occupation.
+A reconstruction of the trajectory of shoreline displacement can thus be used to date the original occupation of the sites based on their elevation relative to the present day sea level.
+The method of shoreline dating is still fundamental to Stone Age research in Norway, and is used both independently and to compliment the dating of sites where other sources of temporal data, such as typological indicators or radiometric dates, are either limited or not available.
 Given the generally coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most accurate method by which one can hope to date the sites.
@@
 This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement, it can potentially identify societal shifts if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the region.
+Furthermore, the method of Monte Carlo simulation to , can be extended to a wide range of use cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
 
 # Background
-After the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, meaning that the relative sea level has been dropping throughout prehistory.
+Relative sea level is so named because the parameters of interest is the elevation of the sea surface relative to land.
+Relative sea level is the distance from the solid surface of the Earth to the mean sea level, as averaged over time, and is also known as the geoid.
+Variation in this relative distance follows from effects that can be divided into three main categories.
+The first of these is eustatic sea-level change.
+Eustatic sea-level is understood as the sea level if the water has been evenly distributed across the Earth's surface and without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself.
+The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can either bind or release large amounts of water into the oceans, as well as by volumetric changes in the Earth's crust following from .
+The second category is that of istostasy, which pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle.
+This is typically the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, and from erosion of the crust which causes its weight to be redistributed.
+These effects thus causes the lithosphere to be either pushed into the mantle due to increased weight or to rebound and to lift upwards due to a lower weight.
+Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.
+
+Following the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, despite corresponding eustatic sea-level rise.
+In other words, the relative sea level has been dropping throughout prehistory, a process that also contiues to this day.
 In other areas of Norway the situation is sometimes more complicated.
-In some areas relative sea-level change has transitioned between regression and transgression phases, and the archaeological remains can be destroyed.
+On the outer coast, the relative sea-level change has transitioned between regression and transgression phases, and the archaeological remains can be destroyed.
 In other areas, the sea-level has been stable over longer periods of time, meaning coastal locations have often been reused multiple times over long time-spans, creating a palmpiset of settlement events that are impossible to disentangle.
@@
 
-# Methods
-Relative sea level is so named because the parameters of interest is the height of the sea surface relative to its present level.
-Sea level is here understood as the distance from the solid surface of the Earth to the mean sea level as averaged over time, also known as the geoid.
-Variation in this distance is determined by effects that can be divided into three main categories.
-The first of these is eustatic sea-level change.
-Eustatic sea level is understood as the sea level if the water has been evenly distributed across the Earth's surface and without accounting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself.
-The eustatic sea-level is most notably impacted by glaciation and de-glaciation which can either bind or release large amounts of water into the oceans, as well as volumetric changes in the Earth's crust following from .
-The second category is that of istostasy, which pertains to adjustments in the Earth's crust that results from processes to regain gravitational equilibrium relative to the underlying viscous mantle.
-This is typically the result of erosion of the crust which causes its weight to be redistributed and the surface to be pushed upwards, or glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, which either causes the lithosphere to be pushed into the mantle or to rebound and to lift upwards.
-Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.
-
 This study necessaitated the compilation of multiple relative sea-level curves developed for different parts of the region.
@@
 
-The goal is to test the synchroneity of two space-time processes, the occupation of a site and continuously shifting sea-levels.
+# Methods
+The goal is effectively to test the synchroneity of two space-time processes: The occupation of a site and continuously shifting sea-levels.
 The synchroneity of these processes is tested by the comparison of singular events represented by the death of the dated organic materials.

### 2021-10-04 analysis/script/04script.R

@@ -233,13 +233,11 @@ sitearea <- rast(file.path(siterpath,
                            paste0(str_replace(sitename, " ", "_"), ".tif")))
 
 plot_results(sitename, output$sitel, output$datedat, sitearea, bmap,
-             sites_sa, isobases, output, date_groups, scale_dist = 100,
-             s_tdist = 0.25, s_xpos = 110, s_ypos = 50, s_tsize = 2,  s_bheight = 0.05,
-             os_tdist = 0.03, os_bheight = 0.02, os_tsize = 2,
-             adjust_widths = c(1, 1, 1))
+             sites_sa, isobases, output, date_groups, scale_dist = 100)
 
 ggsave(file = here("analysis/figures/langangenv5.png"), width = 250,
-       height = 152, units = "mm")
+       height = 228, units = "mm")
+
 ######### Langangen Vestgård 6 #########
 sitename <- "Langangen Vestgård 6"
 date_groups <- rep(1, 9)
@@ -276,6 +274,41 @@ plot_results(sitename, output$sitel, output$datedat, sitearea, bmap,
 
 ggsave(file = here("analysis/figures/rognlien.png"), width = 250,
        height = 152, units = "mm")
+
+######### Dybdalshei 2 #########
+sitename <- "Dybdalshei 2"
+date_groups <- 1
+
+# output <- apply_functions(sitename, date_groups, dtmpath, displacement_curves,
+#                           isobases, nsamp = 1000, loc_bbox = 250, siterpath)
+load(here("analysis/data/derived_data/dybdalshei2.RData"))
+
+sitearea <- rast(file.path(siterpath,
+                           paste0(str_replace(sitename, " ", "_"), ".tif")))
+
+plot_results(sitename, output$sitel, output$datedat, sitearea, bmap,
+             sites_sa, isobases, output, date_groups, scale_dist = 100)
+
+ggsave(file = here("analysis/figures/dybdalshei2.png"), width = 250,
+       height = 152, units = "mm")
+
+######### Dybdalshei 1 #########
+sitename <- "Dybdalshei 1"
+date_groups <- c(1, 1, 1, 1, 2, 2, 3, 3)
+
+# output <- apply_functions(sitename, date_groups, dtmpath, displacement_curves,
+#                           isobases, nsamp = 1000, loc_bbox = 250, siterpath)
+load(here("analysis/data/derived_data/dybdalshei1.RData"))
+
+sitearea <- rast(file.path(siterpath,
[…]

### 2021-10-12 analysis/paper/paper.Rmd
Added Alveberget and Intercity-Vestfold

@@
 # Introduction
-The association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C.
+The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C.
 Brøgger and his son, the archaeologist A.W.
-Brøgger, in their respective theses both published in 1905 (see also @maccurdy1906).
-Ever since, this finding has been fundamental to Stone Age research in Norway.
-One of its major implications is that relative sea-level change can be used for the dating of the sites.
-This is based on most Stone Age sites were situated on or close to the contemporary shoreline during their original occupation.
-A reconstruction of the trajectory of shoreline displacement can thus be used to date the original occupation of the sites based on their elevation relative to the present day sea level.
-The method of shoreline dating is still fundamental to Stone Age research in Norway, and is used both independently and to compliment the dating of sites where other sources of temporal data, such as typological indicators or radiometric dates, are either limited or not available.
-Given the generally coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most accurate method by which one can hope to date the sites.
+Brøgger, in their respective theses of 1905 (see also @maccurdy1906).
+One of the major implications of this finding is that relative sea-level change can be used for the dating of the sites.
+If most Stone Age sites were situated on or close to the contemporary shoreline, a reconstruction of the trajectory of shoreline displacement can be used to date the original occupation of the sites based on their elevation relative to the present day sea level.
+The method of shoreline dating has consequently been fundamental to Stone Age research in Norway ever since, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
+Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites.
 Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age.
@@
 
-Despite the fundamental role of shoreline-dating in Norway, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement.
-The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway.
+Despite the fundamental role of shoreline-dating in Norway, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using what could be termed relatively coarse methods.
+The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach.
 The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age.
-This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement, it can potentially identify societal shifts if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the region.
-Furthermore, the method of Monte Carlo simulation to , can be extended to a wide range of use cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
+This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement and its precision.
+It can also help identify societal shifts if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the region.
+As presented in more detail below, this problem involves the combined evaluation of three major components.
+One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question.
+A Monte Carlo simulation-based approach was adopted to integrate all three dimensions in the analysis.
+While this approach is only used to inform the question of the distance between sites and shorelines here, this method can be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
 
@@
 Relative sea level is so named because the parameters of interest is the elevation of the sea surface relative to land.
-Relative sea level is the distance from the solid surface of the Earth to the mean sea level, as averaged over time, and is also known as the geoid.
 Variation in this relative distance follows from effects that can be divided into three main categories.
@@
 The second category is that of istostasy, which pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle.
-This is typically the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, and from erosion of the crust which causes its weight to be redistributed.
-These effects thus causes the lithosphere to be either pushed into the mantle due to increased weight or to rebound and to lift upwards due to a lower weight.
+This is often the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust which causes its weight to be redistributed.
+These effects thus causes the lithosphere to be either pushed into the mantle due to increased weight, or to rebound and to lift upwards due to lower weight.
 Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.
 
+sea level is the distance from the solid surface of the Earth to the mean sea level, as averaged over time, and is also known as the geoid.
+
 Following the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, despite corresponding eustatic sea-level rise.
-In other words, the relative sea level has been dropping throughout prehistory, a process that also contiues to this day.
+In other words, the relative sea level has been dropping throughout prehistory, a process that also continues to this day.
 In other areas of Norway the situation is sometimes more complicated.
@@
 The evaluation of the degree to which sites were shorebound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations.
-Although it appears that the arguments for such site locations is sensible and for a large part appear to be correct, few attempts at a comprehensive evalation and quantification of this tendency has been done.
+Although it appears that the arguments for such site locations is sensible and for a large part appear to be correct, few attempts at a comprehensive evaluation and quantification of this tendency has been done.
 
@@
 
+# Data
+All data that are mentioned in the text, and that are required to run the analyses are freely available in online repositories.
+
+The data compiled for the analysis consists of all excavated Stone Age sites from the region between Horten county in the north-east, to Arendal in the south-west.
+These number a total nrow() sites, of which nrow() have radiocarbon dates.
+Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age, and nrow() sites have radiocarbon dates to the Stone Age that were evaluated by the original excavators to relate to the lithic inventory on the sites (see also below).
+Spatial data in the form of features and site limits as defined by the excavation was retrieved from local databases at the MCH.
+Each radiocarbon date has been associated with site features or, where these weren't available, the spatial limit of the entire site.
+Due to somewhat variable practices between excavations, what spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports.
+This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hulls generated around the site features.
+A note on this is given for each site in the supplemental material.
+
 # Methods
-The goal is effectively to test the synchroneity of two space-time processes: The occupation of a site and continuously shifting sea-levels.
+The goal is effectively to evaluate the spatial relationship between two phenomena, each associated with their own range of temporal uncertainty.
+First, a single year was drawn from the posterior desnity estimate.
+This then corresponds to a possible elevation range in the shoreline displacement curve, adjusted to the site location.
+An elevation value was drawn uniformly from this range, using intervals of 5 cm.
+The sea-level was then raised to this elevation on the digital terrain model, and the vertical, horisontal and topographic distance between site and shore-line was measured.
+This process was repeated 1000 times for each phase for each site.
+
+The occupation of a site and shifting sea-levels.
 The synchroneity of these processes is tested by the comparison of singular events represented by the death of the dated organic materials.
@@
 In the literature this is often qualified with statements along the lines of "the sites were located on or close to the contemporenous shore-line".
-What constitutes close is of course highly variabl,e complicated further by the fact that vertical changes in sea level can results in widely different effects on the horisontal distance between a given point and the shore-line.
+What constitutes close is of course highly variable, complicated further by the fact that vertical changes in sea level can results in widely different effects on the horisontal distance between a given point and the shore-line.
 Other elements of interest is the tidal range, and the exposure of the location in question, which can also have an impact on the distance of the site to the mean ocean surface.

### 2021-10-19 analysis/paper/paper.Rmd
Reducing raster resolution to 10 m (i.e. using DTM10 from hoydedata.no)

@@
 One of the major implications of this finding is that relative sea-level change can be used for the dating of the sites.
-If most Stone Age sites were situated on or close to the contemporary shoreline, a reconstruction of the trajectory of shoreline displacement can be used to date the original occupation of the sites based on their elevation relative to the present day sea level.
+This follows from the fact that if most Stone Age sites were situated on or close to the contemporary shoreline, a reconstruction of the trajectory of shoreline displacement can be used to date the original occupation of the sites based on their elevation relative to the present day sea level.
 The method of shoreline dating has consequently been fundamental to Stone Age research in Norway ever since, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.

### 2021-10-19 analysis/script/03script.R

@@ -361,7 +361,8 @@ interpolate_curve <- function(years, isobase1, isobase2, target, dispdat,
   return(values)
 }
 
-# Function to identify and load the correct raster
+# Function to identify and load the correct raster if it is split up into
+#  multiple tiles (not in use).
 load_raster <- function(dtmfolder, sitelimit) {
 
   # Number of raster files in total
@@ -538,7 +539,7 @@ sea_overlaps <- function(sitearea, seapolygons){
 }
 
 # Function that combines all of the above
-apply_functions <- function(sitename, date_groups, dtmpath, displacement_curves,
+apply_functions <- function(sitename, date_groups, dtm, displacement_curves,
                             isobases, nsamp = 1000, loc_bbox, siterpath,
                             rcarbcor_true = FALSE){
 
@@ -568,9 +569,6 @@ apply_functions <- function(sitename, date_groups, dtmpath, displacement_curves,
                                  direction_rel_curve1 = sitel$dir_rel_1)
   # Add site name
   sitecurve$name <- sitename
-
-  # Load correct regional raster
-  dtm <- load_raster(dtmpath, sitel)
 
   # Create bounding box polygon
   location_bbox <- bboxpoly(sitel, loc_bbox)
@@ -725,9 +723,11 @@ shore_plot <- function(overlapgrid, sitelimit, dist, date_groups,
                                                        NA, overlaps))
 
   ggplot() +
-    geom_sf(data = overlapgrid, aes(colour = overlaps)) +
+    geom_sf(data = overlapgrid, aes(colour = overlaps, fill = overlaps)) +
     geom_sf(data = sitelimit, colour = "black", fill = NA) +
     scale_colour_gradient(low = "grey98", high = "grey40",
+                          na.value = "grey99") +
+    scale_fill_gradient(low = "grey98", high = "grey40",
                           na.value = "grey99") +
     ggsn::scalebar(data = sitelimit, dist = dist, dist_unit = "m",
                    transform = FALSE, st.size = s_tsize, height = s_bheight,

### 2021-10-24 analysis/paper/paper.Rmd
Moved to DTM 10. Started process of interpolating raster where there's railway/highway

--- 2021-10-12 analysis/paper/paper.Rmd
@@
 # Introduction
+The relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology, as archaeological phenomena that were originally shore-bound can be dated with reference to relative sea-level change .
+This follows from the fact that if the phenomena were situated on or close to the contemporary shoreline, a reconstruction of the trajectory of shoreline displacement can be used to date the original occupation of the sites based on their elevation relative to the present day sea level.
+
+Post-glacial isostatic rebound leading to shoreline fall
+
 The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C.
@@
 Brøgger, in their respective theses of 1905 (see also @maccurdy1906).
-One of the major implications of this finding is that relative sea-level change can be used for the dating of the sites.
-If most Stone Age sites were situated on or close to the contemporary shoreline, a reconstruction of the trajectory of shoreline displacement can be used to date the original occupation of the sites based on their elevation relative to the present day sea level.
-The method of shoreline dating has consequently been fundamental to Stone Age research in Norway ever since, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
+One of the major implications of this finding is thus that relative sea-level change can be used for the dating of the sites.
+The method of shoreline dating has consequently been fundamental to Stone Age research in Norway ever since.
+The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
 Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites.
 Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age.
-This is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line.
+However, this is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line.
 Consequently, adherence or deviation from this pattern has major implications for the socio-economic foundations of the societies in question.
@@
 The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age.
-This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement and its precision.
-It can also help identify societal shifts if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the region.
+This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement as well as its precision.
+The investigation can also help identify societal shifts, if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the area of analysis.
 As presented in more detail below, this problem involves the combined evaluation of three major components.
@@
 A Monte Carlo simulation-based approach was adopted to integrate all three dimensions in the analysis.
-While this approach is only used to inform the question of the distance between sites and shorelines here, this method can be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
+The approach is only used to inform the question of the distance between sites and shorelines here, but this method can also be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
 
@@
 If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this.
-In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of of the overlap between shoreline displacement curves and the radiocarbon probability distributions.
-Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensisble if the site was *not* shore bound [e.g.
+In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions.
+Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was *not* shore bound [e.g.
 @jaksland2014].

### 2021-10-28 analysis/supplementary-materials/supplementary.Rmd
Independent phases for radiocarbon dates. Set up ordering and red outline for non correlating dates

---
title: "Supplementary material"
author: "Isak Roalkvam"
date: "10/27/2021"
output: pdf_document
---

Simulation results for each individual site. The first plot on the first row for each site shows all radiocarbon dates from the site. The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase and each "sum" indicates that the dates for a phase have been modelled using the Boundary function and then summed. A red outline indicates that the date(s) were not seen as related to the lithic inventory in the original report (see also main text). The second plot on the first row displays the site location in the present day landscape on the edited DTM (i.e. highways and railways have been removed and elevation values are interpolated from the surroundings; see analysis/script/03script.R). The third plot on the first rows shows the site location within the study area relative to the isobases of the displacement curves in use (see main text for more on these).

```{r gunnarsrod5, include=FALSE}

```

Gunnarsrød 5 is today situated at the exit of a railway tunnel. The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated. While the interpolation appears to give reasonable results for the rest of the flatter marsh area where the railway runs today, the site itself would thus have been situated slightly higher, leading to the negative values on the distance measures. As the simulation results compared to the site map in   indicate that the site would not have been situated any significant distance from the shoreline, the values have therefore been set to zero for the final aggregative analysis.

### 2021-10-28 analysis/script/03script.R

@@ -40,8 +40,8 @@ execGRASS("r.in.gdal", flags = c("overwrite"),
 # Set computational region to the raster
 execGRASS("g.region", parameters = list(raster = "tmprast") )
 
-# Perform interpolation using regularized spline interpolation with tension
-# with default settings for the tension parameter.
+# Perform regularized spline interpolation with tension
+# using default settings for the tension parameter.
 execGRASS("r.fillnulls", flags = "overwrite",
           parameters = list(input = "tmprast",
                             output = "intrast",

### 2021-10-28 analysis/script/05script.R

@@ -311,22 +311,24 @@ ggsave(file = here("analysis/figures/langangenv5.png"), width = 250,
 
 ######### Langangen Vestgård 6 #########
 sitename <- "Langangen Vestgård 6"
-date_groups <- rep(1, 9)
+date_groups <- c(rep(1, 9), 2)
 
-# output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
-#                           isobases, nsamp = 1000, loc_bbox = 400, siterpath,
-#                           rcarbcor_true = TRUE)
+output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
+                           isobases, nsamp = 1000, loc_bbox = 400, siterpath)
 load(here("analysis/data/derived_data/langangenv6.RData"))
+
+save(output,
+     file = here::here("analysis/data/derived_data/langangenv6.RData.RData"))
 
 sitearea <- rast(file.path(siterpath,
                            paste0(str_replace(sitename, " ", "_"), ".tif")))
 
 plot_results(sitename, output$sitel, output$datedat, sitearea, bmap,
              sites_sa, isobases, output, date_groups, scale_dist = 150,
-             s_tdist = 0.5, s_xpos = 110, s_ypos = 50,  s_bheight = 0.3)
+             s_tdist = 1, s_xpos = 170, s_ypos = 80,  s_bheight = 0.5)
 
 ggsave(file = here("analysis/figures/langangenv6.png"), width = 250,
-       height = 152, units = "mm")
+       height = 228, units = "mm")
 
 ######### Rognlien #########
 sitename <- "Rognlien"
@@ -369,9 +371,11 @@ ggsave(file = here("analysis/figures/dybdalshei2.png"), width = 250,
 sitename <- "Dybdalshei 1"
 date_groups <- c(1, 1, 1, 1, 2, 2, 3, 3)
 
-# output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
-#                           isobases, nsamp = 1000, loc_bbox = 400, siterpath)
-load(here("analysis/data/derived_data/dybdalshei1.RData"))
+output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
+                           isobases, nsamp = 1000, loc_bbox = 400, siterpath)
+
+save(output,
+     file = here::here("analysis/data/derived_data/dybdalshei1.RData"))
 
 sitearea <- rast(file.path(siterpath,
                            paste0(str_replace(sitename, " ", "_"), ".tif")))
[…]

### 2021-10-29 analysis/script/04script.R
More or less set up to run through all analyses (missing Brunstad site limits, and need to enter two Kvastad survey dates, also check Kvastad duplication of dates).

@@ -843,8 +843,7 @@ plot_dates <- function(datedata, sitename, multigroup = TRUE, groupn = NA,
 
     plot <-
       datedata %>%
-      filter(!(name %in% c("Start",
-                           names(prior)[grep("End", names(prior))]))) %>%
+      filter(!grepl("Start|End", name)) %>%
       ggplot() +
       ggridges::geom_ridgeline(aes(x = dates,
       # 3 nested ordering levels: median age, group and sum/individual dates
@@ -865,7 +864,9 @@ plot_dates <- function(datedata, sitename, multigroup = TRUE, groupn = NA,
       labs(y = "", x = "") +
       theme(legend.position = "none")
   } else{
-    plot <- ggplot(datedata) +
+    plot <- datedata %>%
+      filter(!grepl("Start|End", name)) %>%
+      ggplot() +
       ggridges::geom_ridgeline(aes(x = dates, y = name,
                                    height = probabilities * 50,
                                    fill = as.factor(group),

### 2021-11-01 analysis/paper/paper.Rmd
Stepping through each site

@@
 # Introduction
-The relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology, as archaeological phenomena that were originally shore-bound can be dated with reference to relative sea-level change .
-This follows from the fact that if the phenomena were situated on or close to the contemporary shoreline, a reconstruction of the trajectory of shoreline displacement can be used to date the original occupation of the sites based on their elevation relative to the present day sea level.
+The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology.
+This follows not only from the dramatic changes to the landscape that this process represent, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
 
-Post-glacial isostatic rebound leading to shoreline fall
-
-The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C.
+The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was established by the geologist W.C.
 Brøgger and his son, the archaeologist A.W.
 Brøgger, in their respective theses of 1905 (see also @maccurdy1906).
-One of the major implications of this finding is thus that relative sea-level change can be used for the dating of the sites.
-The method of shoreline dating has consequently been fundamental to Stone Age research in Norway ever since.
-The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
+The method of shoreline dating has been fundamental to Stone Age research in Norway ever since, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
 Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites.
@@
 
-Despite the fundamental role of shoreline-dating in Norway, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using what could be termed relatively coarse methods.
+Post-glacial isostatic rebound leading to shoreline fall
+
+Despite the fundamental role of shoreline-dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods.
 The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach.
 The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age.
-This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement as well as its precision.
-The investigation can also help identify societal shifts, if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the area of analysis.
-As presented in more detail below, this problem involves the combined evaluation of three major components.
+This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement, as well as its precision.
+If there is chronological variation in the degree to which the sites were shore-bound, the investigation can also help identify societal shifts, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the area of analysis.
+As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions.
 One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question.
-A Monte Carlo simulation-based approach was adopted to integrate all three dimensions in the analysis.
+Taking inspiration from recent studies that have integrated multiple sources of spatio-temporal uncertainty, a Monte Carlo simulation-based approach was adopted for the analysis.
+
 The approach is only used to inform the question of the distance between sites and shorelines here, but this method can also be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
@@
 # Methods
-The goal is effectively to evaluate the spatial relationship between two phenomena, each associated with their own range of temporal uncertainty.
-First, a single year was drawn from the posterior desnity estimate.
-This then corresponds to a possible elevation range in the shoreline displacement curve, adjusted to the site location.
+
+The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shore-lines, each associated with their own range of temporal uncertainty.
+The first task was therefore to ascribe likely date ranges and a numerical expression of the associated uncertainty to the two phenomena.
+
+Due to the dsfg there is a gradient in the rate of isostatic rebound, with increasing rebound rates towards the north-east.
+To take account of this gradient, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase.
+This was done by first finding the difference between the two closest displacement curves.
+The elevation difference was found for each year along the entirety of the curves, with each value divided by the meter distance between the isobases.
+This was done for both ends of the confidence intervals.
+To find the adjusted displacement curve for a given site location the distance between the site polygon and closest isobase was multiplied with the values for difference per meter.
+If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east the values were instead added to the original curve.
+The result of this process is shown for an example site in Figure.
+
+For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve using OxCal v.
+.
+All dates not younger than 1700 BCE with 98 % certainty were then excluded from further analysis.
+Radiocarbon dates associated with each site were then inspected, and with reference to the evaluation done in the original reports, dates believed to belong to the same settlement phase were grouped.
+In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal.
+Multiple phases at a single site were treated as independent of each other (Figure x).
+
+As the excavation of archaeological sites typically follow from residential and commercial development and the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances.
+In addition to employing 10 m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster.
+This involved manually defining the extent of problem areas such as railways, highways, quarries and the like, using polygon features.
+The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain.
+This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls through GRASS GIS.
+In addition to being available in the , it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.
+
+Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern , the simulation was performed.
+Each simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase.
+This year then has a corresponding likely elevation range of the shoreline.
 An elevation value was drawn uniformly from this range, using intervals of 5 cm.
-The sea-level was then raised to this elevation on the digital terrain model, and the vertical, horisontal and topographic distance between site and shore-line was measured.
+The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing, and creating polygons of the resulting areas with missing values.
+The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the.
+The so-called topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM.
+This was done using the topoDist package.
+In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
+In the case that the sea-polygons completely contain the site, the horisontal and topographic distance measures were made negative.
+While its safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure and might also help identify discrepancies in displacement data or radiocarbon dates.
+
+horisontal and topographic distance between site and shore-line was measured.  
 This process was repeated 1000 times for each phase for each site.
+The choice of 1000 iterations follows an evaluation of convergence on median distances using the example site Hovland 3.
+At 1000 simulation runs the analysis of each site ranged from  hours, depending on the distance to the simulated shorelines and consequently the necessary size of the window of analysis.
+
+The distances were then measured.
 

### 2021-11-01 analysis/supplementary-materials/supplementary.Rmd
More or less set up to run through all analyses (missing Brunstad site limits, and need to enter two Kvastad survey dates, also check Kvastad duplication of dates).
Stepping through each site

@@
 
-Simulation results for each individual site.
-The first plot on the first row for each site shows all radiocarbon dates from the site.
-The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase and each "sum" indicates that the dates for a phase have been modelled using the Boundary function and then summed.
+This document includes some notes and shows the simulation results for each individual site analysed in the study.
+The first plot on the first row for each site shows all calibrated radiocarbon dates from the Stone Age associated with the site if there are more than a single date (see analysis/data/raw_data/radiocarbon.csv for all dates, including those falling outside of the Stone Age).
+The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase, and each sum indicates that the dates for a phase have been modelled using the Boundary function and then summed using OxCal.
+Each phase is assumed to be independent of other phases.
 A red outline indicates that the date(s) were not seen as related to the lithic inventory in the original report (see also main text).
 The second plot on the first row displays the site location in the present day landscape on the edited DTM (i.e.
-highways and railways have been removed and elevation values are interpolated from the surroundings; see analysis/script/03script.R).
-The third plot on the first rows shows the site location within the study area relative to the isobases of the displacement curves in use (see main text for more on these).
+highways and railways impacting the adjusted sea-level have been removed and elevation values are interpolated from the surroundings; see main text and analysis/script/03script.R).
+The third plot on the first row shows the site location within the study area relative to the isobases of the displacement curves in use.
+
+Subsequent rows for each site then show the simulation results.
+The first plot on these rows shows the probability density function from which site dates were drawn during simulation.
+The second plot shows the result of the simulation runs, where the intensity of the colour indicates the number of times the sea was simulated to be present at any given location.
+The third plot displays a violin plot of the measured distances between site and shore-line across all simulation runs.
+Measures for central tendency and dispersion associated with these are then given in the table
+
+```{r gunnarsrod5, include=FALSE}
+
+```
+
+Gunnarsrød 4 situated right by the railway (see also Gunnarsrød 5 below).
+The date is seen as possibly the result of old wood effect, a notion that is in line with simulation results here, as this date would the site located beneath the sea-level.
 

### 2021-11-02 analysis/paper/paper.Rmd
Should have rearranged radiocarbon dates correctly in radiocarbon.csv

@@
 # Data
-All data that are mentioned in the text, and that are required to run the analyses are freely available in online repositories.
+All data that are required to run the analyses and the derived data are freely available in an online repository. The dataset is slightly more extensive
 
@@
 This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hulls generated around the site features.
-A note on this is given for each site in the supplemental material.
 
@@
 The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shore-lines, each associated with their own range of temporal uncertainty.
-The first task was therefore to ascribe likely date ranges and a numerical expression of the associated uncertainty to the two phenomena.
+The first task was therefore to ascribe likely date ranges and associated uncertainty to the two phenomena.
 
@@
 To take account of this gradient, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase.
-This was done by first finding the difference between the two closest displacement curves.
-The elevation difference was found for each year along the entirety of the curves, with each value divided by the meter distance between the isobases.
+This was done by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves.
+Each value was then divided by the meter distance between the isobases.
 This was done for both ends of the confidence intervals.
 To find the adjusted displacement curve for a given site location the distance between the site polygon and closest isobase was multiplied with the values for difference per meter.
-If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east the values were instead added to the original curve.
+If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east, the values were instead added to the original curve.
 The result of this process is shown for an example site in Figure.
@@
 The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain.
-This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls through GRASS GIS.
-In addition to being available in the , it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.
+This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls from GRASS GIS in R through the rgrass7 package.
+In addition to code and original spatial data being available in the research compendia for this paper, it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.
 
-Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern , the simulation was performed.
+Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern, the simulation was performed.
 Each simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase.
@@
 The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing, and creating polygons of the resulting areas with missing values.
-The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the.
+The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon.
 The so-called topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM.
-This was done using the topoDist package.
+This was done using the topoDist package for R.
 In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.

### 2021-11-02 analysis/supplementary-materials/supplementary.Rmd
Should have rearranged radiocarbon dates correctly in radiocarbon.csv

@@
 Measures for central tendency and dispersion associated with these are then given in the table
+```{r anvik, include=FALSE}
+
+```
+Anvik is situated by the highway and a smaller road.
+Neither appear to significantly impact the simulation results.
+The dated fireplace is not seen as related to lithic material, which is believed to hail from the latest part of the Early Mesolithc (c.
+8500-8250 BCE).
 
@@
 Gunnarsrød 4 situated right by the railway (see also Gunnarsrød 5 below).
-The date is seen as possibly the result of old wood effect, a notion that is in line with simulation results here, as this date would the site located beneath the sea-level.
+The date was seen as possibly the result of old-wood effect, a notion that is in line with the simulation results here, as this date would have the site located beneath the sea-level.
 
@@
 The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated.
-While the interpolation appears to give reasonable results for the rest of the flatter marsh area where the railway runs today, the site itself would thus have been situated slightly higher, leading to the negative values on the distance measures.
+While the interpolation appears to give reasonable results for the rest of the flatter marsh area where the railway runs today, the site itself would thus have been situated slightly higher, likely leading to the negative values on the distance measures.
 As the simulation results compared to the site map in   indicate that the site would not have been situated any significant distance from the shoreline, the values have therefore been set to zero for the final aggregative analysis.

### 2021-11-04 analysis/paper/paper.Rmd
Analysis of individual sites

@@
 The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology.
-This follows not only from the dramatic changes to the landscape that this process represent, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
+This follows not only from the dramatic changes to the landscape that this process has represented throughout history, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
+This method has long history of use in the region, as is frequently applied within both the research and heritage management sector to date diverse archaeological phenomena such as rock art, harbour constructions and, as is the focus of this study, Stone Age sites,
 
@@
 Brøgger, in their respective theses of 1905 (see also @maccurdy1906).
-The method of shoreline dating has been fundamental to Stone Age research in Norway ever since, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
-Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites.
-Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age.
-However, this is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line.
-Consequently, adherence or deviation from this pattern has major implications for the socio-economic foundations of the societies in question.
+The method of shoreline dating has been fundamental to Stone Age research in Norway ever since.
+The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
+Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites.
+Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age, both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line.
+Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.
 
@@
 
-Despite the fundamental role of shoreline-dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods.
+Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods.
 The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach.
 The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age.
-This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement, as well as its precision.
+This can be used both to inform the certainty with which one can depend on dating with reference to shoreline displacement, as well as its precision.
 If there is chronological variation in the degree to which the sites were shore-bound, the investigation can also help identify societal shifts, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the area of analysis.
@@
 One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question.
-Taking inspiration from recent studies that have integrated multiple sources of spatio-temporal uncertainty, a Monte Carlo simulation-based approach was adopted for the analysis.
-
-The approach is only used to inform the question of the distance between sites and shorelines here, but this method can also be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
+Taking inspiration from recent studies that have integrated multiple sources of spatio-temporal uncertainty through Monte Carlo simulation, a similar approach is adopted here.
 
@@
 
+The approach is only used to inform the question of the distance between sites and shorelines here, but this method can also be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
+
 ```{r get-data, eval = FALSE}

### 2021-11-04 analysis/script/05script.R

@@ -88,7 +88,7 @@ sitearea <- rast(file.path(siterpath,
                            paste0(str_replace(sitename, " ", "_"), ".tif")))
 
 plot_results(sitename, output$sitel, output$datedat, sitearea, bmap,
-             sites_sa, isobases, output, date_groups, scale_dist = 1000,
+             sites_sa, isobases, output, date_groups, scale_dist = 1250,
              s_tdist = 3, s_xpos = 1250, s_ypos = 500,  s_bheight = 1.5)
 
 ggsave(file = here("analysis/figures/bakke.png"), width = 250,
@@ -641,10 +641,10 @@ ggsave(file = here("analysis/figures/langangenv1.png"), width = 250,
 sitename <- "Langangen Vestgård 3"
 date_groups <- c(2, 2, 1, 2, 2, 2)
 
-output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
-                           isobases, nsamp = 1000, loc_bbox = 400, siterpath)
-save(output,
-     file = here::here("analysis/data/derived_data/langangenv3.RData"))
+# output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
+#                            isobases, nsamp = 1000, loc_bbox = 400, siterpath)
+# save(output,
+#      file = here::here("analysis/data/derived_data/langangenv3.RData"))
 load(here("analysis/data/derived_data/langangenv3.RData"))
 
 sitearea <- rast(file.path(siterpath,
@@ -652,7 +652,7 @@ sitearea <- rast(file.path(siterpath,
 
 plot_results(sitename, output$sitel, output$datedat, sitearea, bmap,
              sites_sa, isobases, output, date_groups, scale_dist = 150,
-             s_tdist = 1, s_xpos = 110, s_ypos = 50,  s_bheight = 0.5)
+             s_tdist = 1.5, s_xpos = 150, s_ypos = 80,  s_bheight = 1.25)
 
 ggsave(file = here("analysis/figures/langangenv3.png"), width = 250,
        height = plot_height + (plot_height * length(unique(date_groups))),
@@ -676,18 +676,18 @@ plot_results(sitename, output$sitel, output$datedat, sitearea, bmap,
              sites_sa, isobases, output, date_groups, scale_dist = 150)
 
 ggsave(file = here("analysis/figures/langangenv5.png"), width = 250,
-       height = 228, units = "mm")
+       height = plot_height + (plot_height * length(unique(date_groups))),
+       units = "mm")
 
[…]

### 2021-11-08 analysis/paper/paper.Rmd
Ragnhildrød will take some time

@@
 The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology.
-This follows not only from the dramatic changes to the landscape that this process has represented throughout history, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
-This method has long history of use in the region, as is frequently applied within both the research and heritage management sector to date diverse archaeological phenomena such as rock art, harbour constructions and, as is the focus of this study, Stone Age sites,
+This follows not only from the dramatic changes to the landscape that this process has represented, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
+This method has long history of use in the region, as is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, harbour and other sea-side constructions and, as is the focus of this study, Stone Age sites.
 
@@
 The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age.
-This can be used both to inform the certainty with which one can depend on dating with reference to shoreline displacement, as well as its precision.
-If there is chronological variation in the degree to which the sites were shore-bound, the investigation can also help identify societal shifts, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the area of analysis.
 As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions.
@@
 Taking inspiration from recent studies that have integrated multiple sources of spatio-temporal uncertainty through Monte Carlo simulation, a similar approach is adopted here.
+The results of this study can be used both to inform the certainty with which one can depend on dating with reference to shoreline displacement, as well as its precision.
+If there is chronological variation in the degree to which the sites were shore-bound, the investigation can also help identify important societal changes, and any such variation can potentially also have implications for how archaeological investigations should be directed within the area of analysis.
 
 # Background
-Relative sea level is so named because the parameters of interest is the elevation of the sea surface relative to land.
+Relative sea-level is so named because the parameters of interest is the elevation of the sea surface relative to land.
 Variation in this relative distance follows from effects that can be divided into three main categories.
-The first of these is eustatic sea-level change.
-Eustatic sea-level is understood as the sea level if the water has been evenly distributed across the Earth's surface and without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself.
+The first of these is changes to the eustatic sea-level, which is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself.
 The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can either bind or release large amounts of water into the oceans, as well as by volumetric changes in the Earth's crust following from .
@@
 Following the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, despite corresponding eustatic sea-level rise.
-In other words, the relative sea level has been dropping throughout prehistory, a process that also continues to this day.
+In other words, the relative sea level has been dropping throughout prehistory---a process that also continues to this day.
 In other areas of Norway the situation is sometimes more complicated.
@@
 This can for example pertain to accessibility.
-If the site is situated on a ledge in a steep and jagged area of the present day landscape, it would make intuitive sense that the location was used when the ocean reached the elevation of the ledge and the site would have been accessible by means of watercraft.
-This line of reasoning is often also used in survey projects which typically do not produce material suitable for radiocarbon dating.
-The evaluation of the degree to which sites were shorebound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations.
+If the site is situated on a ledge in a steep and jagged area of the present day landscape, it would make intuitive sense that the location was used when the ocean reached the elevation of the site which would then have been accessible by means of watercraft.
+This line of reasoning is often also used in survey projects, which typically do not produce material suitable for radiocarbon dating.
+The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations.
 Although it appears that the arguments for such site locations is sensible and for a large part appear to be correct, few attempts at a comprehensive evaluation and quantification of this tendency has been done.
@@
 One of the only larger scale evaulations of this relationship in Norwegian archaeology has been done by Solheim and collegues (@solheim20, see also @solheim2020), who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
-They found a overlap of the confidence bands of the radiocarbon dates with the shoreline displacement curve for 86.5 % of the sites.
+They found a overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5 % of the sites.
 However, in several of the cases where there was a discrepancy, the
@@
 Most of the curves come provided with a confidence interval which is largely based on an evaluation by the geologists who developed the curve.
-The reason why these confidence intervals are not strictly mathematically derived is that the curves do not only contain uncertainy as related to radiometric dates, which are fairly well understood, but also hold potential error as related to the interpretation of sediment cores, placement of samples,  direction and adjustment to a common isobase based on the direction and tilt of the variation in rebound intensity, as well as, to name but a few.
-The error bands are consequently complex, and cannot necessarily be directly compared between curves as these have been developed at different times, using a wide range of different methods, and often using older and outdated calibration curves for both marine and terrestrial samples.
+The reason why these confidence intervals are not strictly mathematically derived is that the curves do not only contain uncertainy as related to radiometric dates, which are fairly well understood, but also hold potential error as related to the interpretation of sediment cores, placement of samples, as well as direction and adjustment to a common isobase based on the direction and gradient in rebound intensity, to name but a few.
+The error bands are consequently complex, and cannot necessarily be directly compared between curves as these have been developed at different times, using a wide range of different methods.
 However, any reinterpretation and standardisation of the curves is also hampered by a range of issues.
@@
 # Data
-All data that are required to run the analyses and the derived data are freely available in an online repository. The dataset is slightly more extensive
+All data that are required to run the analyses and the derived data are freely available in an online repository.
+The dataset is slightly more extensive, including excavated sites without radiocarbon dates, and radiocarbon dates to periods other than the Stone Age.
 
@@
 Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age, and nrow() sites have radiocarbon dates to the Stone Age that were evaluated by the original excavators to relate to the lithic inventory on the sites (see also below).
-Spatial data in the form of features and site limits as defined by the excavation was retrieved from local databases at the MCH.
+Spatial data in the form of features and site limits as defined by the excavation was retrieved from local databases at the Museum of Cultural History.
 Each radiocarbon date has been associated with site features or, where these weren't available, the spatial limit of the entire site.
@@
 .
-All dates not younger than 1700 BCE with 98 % certainty were then excluded from further analysis.
+All dates not younger than 1700 BCE with 99.7 % uncertainty range were then excluded from further analysis.
 Radiocarbon dates associated with each site were then inspected, and with reference to the evaluation done in the original reports, dates believed to belong to the same settlement phase were grouped.
@@
 
-Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern, the simulation was performed.
+Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulation was performed.
 Each simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase.
@@
 An elevation value was drawn uniformly from this range, using intervals of 5 cm.
-The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing, and creating polygons of the resulting areas with missing values.
+The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing.
+Polygons were then created from the resulting areas with missing values.
 The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon.

### 2021-11-15 analysis/paper/paper.Rmd
Set up sample_shoreline to keep info on site name and 14c correlation with artefact inventory. Need to look into details on how the topo vs. horisontal distances are measured

@@
 # Introduction
+
 The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology.
-This follows not only from the dramatic changes to the landscape that this process has represented, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
-This method has long history of use in the region, as is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, harbour and other sea-side constructions and, as is the focus of this study, Stone Age sites.
+This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
+This method has long history of use in the region, as is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, harbour constructions and, as is the focus of this study, Stone Age sites.
 
-The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was established by the geologist W.C.
+The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C.
 Brøgger and his son, the archaeologist A.W.
@@
 Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites.
-Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age, both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line.
+Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age.
+This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line.
 Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.
@@
 One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question.
-Taking inspiration from recent studies that have integrated multiple sources of spatio-temporal uncertainty through Monte Carlo simulation, a similar approach is adopted here.
-The results of this study can be used both to inform the certainty with which one can depend on dating with reference to shoreline displacement, as well as its precision.
-If there is chronological variation in the degree to which the sites were shore-bound, the investigation can also help identify important societal changes, and any such variation can potentially also have implications for how archaeological investigations should be directed within the area of analysis.
+Taking inspiration from recent studies that have integrated various and multiple sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g.
+@crema2010; @crema2012; @lewis2021], a similar approach is adopted here.
 
 # Background
-Relative sea-level is so named because the parameters of interest is the elevation of the sea surface relative to land.
+
+Relative sea-level is so named because the parameters of interest is the elevation of the surface of the sea relative to land.
 Variation in this relative distance follows from effects that can be divided into three main categories.
@@
 The second category is that of istostasy, which pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle.
-This is often the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust which causes its weight to be redistributed.
+This is often the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed.
 These effects thus causes the lithosphere to be either pushed into the mantle due to increased weight, or to rebound and to lift upwards due to lower weight.
@@
 As was exemplified in Solheims study, and which has been convincingly argued in several individual studies, many of the sites where there is a mismatch between the radiocarbon dates and the shoreline dates, the assemblages are clearly related to the shoreline date rather than the radiocarbon dates.
-Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates or whether they are the result of contaminated samples or by pedoturbation can be difficult to evaluate and will here be entirely dependent on the discretion of the original excavation report.
+Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates or whether they are the result of contaminated samples can be difficult to evaluate and will here be entirely dependent on the discretion of the original excavation report.
 While such evaluations can be difficult, it is nonetheless a necessity that will be held strictly separate from the formal equivalence tests in the analysis to follow.
@@
 # Data
+
 All data that are required to run the analyses and the derived data are freely available in an online repository.
@@
 
-horisontal and topographic distance between site and shore-line was measured.  
+horisontal and topographic distance between site and shore-line was measured.\
 This process was repeated 1000 times for each phase for each site.
@@
 
-<div id="refs"></div>
+::: {#refs}
+:::
 

### 2021-11-15 analysis/supplementary-materials/supplementary.Rmd
Set up sample_shoreline to keep info on site name and 14c correlation with artefact inventory. Need to look into details on how the topo vs. horisontal distances are measured

@@
 This document includes some notes and shows the simulation results for each individual site analysed in the study.
-The first plot on the first row for each site shows all calibrated radiocarbon dates from the Stone Age associated with the site if there are more than a single date (see analysis/data/raw_data/radiocarbon.csv for all dates, including those falling outside of the Stone Age).
+The first plot on the first row for each site shows all calibrated radiocarbon dates from the Stone Age associated with the site (see analysis/data/raw_data/radiocarbon.csv for all dates, including those falling outside of the Stone Age).
 The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase, and each sum indicates that the dates for a phase have been modelled using the Boundary function and then summed using OxCal.
 Each phase is assumed to be independent of other phases.
-A red outline indicates that the date(s) were not seen as related to the lithic inventory in the original report (see also main text).
+A red outline indicates that the date(s) were not seen as related to the artefact inventory in the original report (see also main text).
 The second plot on the first row displays the site location in the present day landscape on the edited DTM (i.e.
-highways and railways impacting the adjusted sea-level have been removed and elevation values are interpolated from the surroundings; see main text and analysis/script/03script.R).
-The third plot on the first row shows the site location within the study area relative to the isobases of the displacement curves in use.
+highways and railways impacting the adjusted sea-level have been removed, and elevation values are interpolated from the surroundings; see main text and analysis/script/03script.R).
+The third plot on the first row shows the site location within the study area relative to the isobases of the shoreline displacement curves in use.
 
@@
 The third plot displays a violin plot of the measured distances between site and shore-line across all simulation runs.
-Measures for central tendency and dispersion associated with these are then given in the table
+Measures for central tendency and dispersion associated with these are then given in the table.
+
 ```{r anvik, include=FALSE}
@@
 ```
+
 Anvik is situated by the highway and a smaller road.
@@
 
-```{r gunnarsrod5, include=FALSE}
+```{r gunnarsrod4, include=FALSE}
 
@@
 While the interpolation appears to give reasonable results for the rest of the flatter marsh area where the railway runs today, the site itself would thus have been situated slightly higher, likely leading to the negative values on the distance measures.
-As the simulation results compared to the site map in   indicate that the site would not have been situated any significant distance from the shoreline, the values have therefore been set to zero for the final aggregative analysis.
+As the simulation results compared to the site map in the report [@reitan2014d], in combination indicate that the site would not have been situated any significant distance from the shoreline, the values have been set to zero for the final aggregative analysis.
+A note can also be made on the date that is seen as unrelated to the artefact inventory.
+This was done on charcoal from a pit with ceramics which had a far younger date.
+The date was therefore interpreted as the result of contamination of some kind [@reitan2014d].

### 2021-11-15 analysis/script/04script.R

@@ -441,8 +441,11 @@ bboxpoly <- function(feature, xy_adjustment) {
 # displacement curve
 sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
 
-  results <- data.frame(matrix(ncol = 4, nrow = samps))
-  names(results) <- c("vertdist", "hordist", "topodist", "year")
+  results <- data.frame(matrix(ncol = 6, nrow = samps))
+  names(results) <- c("vertdist", "hordist", "topodist",
+                      "year", "rcarb_cor", "sitename")
+  results$rcarb_cor <- unique(posteriorprobs$rcarb_cor)
+  results$sitename <- unique(sitel$name)
   seapolygons <- list(length = samps)
   topopaths <- list(length = samps)
 
@@ -567,7 +570,7 @@ sea_overlaps <- function(sitearea, seapolygons){
 }
 
 # Function that combines all of the above
-apply_functions <- function(sitename, dtm, displacement_curves,
+apply_functions <- function(sitename, date_groups, dtm, displacement_curves,
                             isobases, nsamp = 1000, loc_bbox, siterpath,
                             rcarbcor_true = FALSE, sitelimit = TRUE){
 
@@ -619,7 +622,6 @@ apply_functions <- function(sitename, dtm, displacement_curves,
   # Retrieve the posterior density estimate for each date group
   for(i in 1:length(unique(date_groups))){
 
-
     # If there is only a single date from the site
     if(nrow(siter) == 1){
       posteriorprobs <- datedat
@@ -645,7 +647,6 @@ apply_functions <- function(sitename, dtm, displacement_curves,
 
   output$datedat <- datedat
   output$sitel <- sitel
-  output$date_groups <- date_groups
   output$sitecurve <- sitecurve
 
   # Store raster (replacing spaces and dashes in the file name)

### 2021-11-15 analysis/script/05script.R

@@ -51,6 +51,49 @@ plot_height <- 76
 # grouping of dates. Each site was therefore first simulated 50 times and
 # then rerun at 1000 when these issues were handled.
 
+######### Adal vestre 1 #########
+sitename <- "Adal vestre 1"
+date_groups <- group_dates(rcarb_sa, sitename)
+
+output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
+                          isobases, nsamp = 1000, loc_bbox = 400, siterpath)
+
+save(output,
+      file = here::here("analysis/data/derived_data/adalvestre1.RData"))
+load(here("analysis/data/derived_data/adalvestre1.RData"))
+
+sitearea <- rast(file.path(siterpath,
+                           paste0(str_replace(sitename, " ", "_"), ".tif")))
+
+plot_results(sitename, output$sitel, output$datedat, sitearea, bmap,
+             sites_sa, isobases, output, date_groups, scale_dist = 150,
+             s_tdist = 0.5, s_xpos = 150, s_ypos = 70, s_bheight = 0.25)
+
+ggsave(file = here("analysis/figures/adalvestre1.png"), width = 250,
+       height = plot_height + (plot_height * length(unique(date_groups))),
+       units = "mm")
+
+######### Adal vestre 2 #########
+sitename <- "Adal vestre 2"
+date_groups <- group_dates(rcarb_sa, sitename)
+
+output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
+                          isobases, nsamp = 1000, loc_bbox = 400, siterpath)
+save(output,
+     file = here::here("analysis/data/derived_data/adalvestre2.RData"))
+load(here("analysis/data/derived_data/adalvestre2.RData"))
+
+sitearea <- rast(file.path(siterpath,
+                           paste0(str_replace(sitename, " ", "_"), ".tif")))
+
+plot_results(sitename, output$sitel, output$datedat, sitearea, bmap,
+             sites_sa, isobases, output, date_groups, scale_dist = 150,
+             s_tdist = 1, s_xpos = 150, s_ypos = 70, s_bheight = 0.5)
+
+ggsave(file = here("analysis/figures/adalvestre2.png"), width = 250,
+       height = plot_height + (plot_height * length(unique(date_groups))),
+       units = "mm")
+
 ######### Anvik #########
 sitename <- "Anvik"
[…]

### 2021-11-15 analysis/paper/paper.Rmd
Struggeling with some strange crashes

@@
 .
-All dates not younger than 1700 BCE with 99.7 % uncertainty range were then excluded from further analysis.
-Radiocarbon dates associated with each site were then inspected, and with reference to the evaluation done in the original reports, dates believed to belong to the same settlement phase were grouped.
+All dates not younger than 1700 BCE with 99.7 % uncertainty were then excluded from further analysis.
+Radiocarbon dates associated with each site were then grouped if they overlaped with 99.7 % uncertainty, effectively taken to represent the same settlement phase.
 In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal.
@@
 Each simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase.
-This year then has a corresponding likely elevation range of the shoreline.
+This year then has a corresponding likely elevation range of the contemporaneous shoreline.
 An elevation value was drawn uniformly from this range, using intervals of 5 cm.

### 2021-11-15 analysis/script/04script.R

@@ -444,7 +444,7 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
   results <- data.frame(matrix(ncol = 6, nrow = samps))
   names(results) <- c("vertdist", "hordist", "topodist",
                       "year", "rcarb_cor", "sitename")
-  results$rcarb_cor <- unique(posteriorprobs$rcarb_cor)
+  results$rcarb_cor <- unique(na.omit(posteriorprobs$rcarb_cor))
   results$sitename <- unique(sitel$name)
   seapolygons <- list(length = samps)
   topopaths <- list(length = samps)
@@ -495,10 +495,6 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
       # Retrieve the two points
       distpts <- st_cast(distline, "POINT")
 
-      # Make the value negative to indicate that the site is located below
-      # sea level
-      results$hordist[i] <- st_length(distline) * -1
-
       # Find topographic distance
       coords <- rbind(st_coordinates(distpts[1]),
                       st_coordinates(distpts[2]))
@@ -509,8 +505,23 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
                            paths = TRUE)
       # Store distance (negative to indicate below sea-level)
       results$topodist[i] <- topodist[[1]]["loc", "sea"] * -1
-      # Store path geometry
-      topopaths[[i]] <- st_as_sf(topodist[[2]])
+
+      # Retrieve path
+      tpath <- st_as_sf(topodist[[2]])
+      topopaths[[i]] <- tpath
+
+      # Find end-points
+      endptns <- st_line_sample(tpath, sample = c(0, 1))
+
+      # Cast as line
+      hordistl <- st_cast(endptns, "LINESTRING")
+
+      # Find horisontal distance and make the value negative to indicate that
+      # the site is located below sea level. The reason for not doing this with
+      # distline defined above is that the topographic distance is measured to
+      # the adjacent cell.
+      results$hordist[i] <- st_length(hordistl) * -1
+
 
       # Else, that is not overlapping slightly or completely, i.e. the site is
       # located some distance from the sea, perform above steps bu without
@@ -522,9 +533,6 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
       # Retrieve the two points
[…]

### 2021-11-18 analysis/paper/paper.Rmd
rerunnig with new distance measures

@@
 This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
-This method has long history of use in the region, as is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, harbour constructions and, as is the focus of this study, Stone Age sites.
+This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, harbour constructions and, as is the focus of this study, Stone Age sites.
 
 The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C.
-Brøgger and his son, the archaeologist A.W.
-Brøgger, in their respective theses of 1905 (see also @maccurdy1906).
-The method of shoreline dating has been fundamental to Stone Age research in Norway ever since.
+Brøgger in 1905 (see @maccurdy1906).
+Shoreline dating has been fundamental to Stone Age research in Norway ever since.
 The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
@@
 
-Post-glacial isostatic rebound leading to shoreline fall
-
 Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods.
@@
 As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions.
-One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question.
-Taking inspiration from recent studies that have integrated various and multiple sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g.
+One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently spatial.
+Taking inspiration from recent studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g.
 @crema2010; @crema2012; @lewis2021], a similar approach is adopted here.
@@
 
-Relative sea-level is so named because the parameters of interest is the elevation of the surface of the sea relative to land.
-Variation in this relative distance follows from effects that can be divided into three main categories.
-The first of these is changes to the eustatic sea-level, which is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself.
-The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can either bind or release large amounts of water into the oceans, as well as by volumetric changes in the Earth's crust following from .
-The second category is that of istostasy, which pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle.
-This is often the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed.
-These effects thus causes the lithosphere to be either pushed into the mantle due to increased weight, or to rebound and to lift upwards due to lower weight.
-Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.
+Relative sea-level can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre [@shennan2015\].
+Variation in this relative distance follow from a range of effects \[e.g.
+@mörner1976; @milne2009].
+Of central importance here is eustasy and glacial istostasy.
+The eustatic sea level is understood as the sea level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015\].
+The eustatic sea level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that has volumetric impact on ocean basins.
+Istostasy, on the other hand, pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle.
+This is often the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed.
+These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight \[@milne2015].
 
-sea level is the distance from the solid surface of the Earth to the mean sea level, as averaged over time, and is also known as the geoid.
+Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet, the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise.
+In other words, the relative sea level has been dropping throughout prehistory.
+As this process is the result of glacial loading, the rate of uplift has been more severe towards the centre of the ice sheet.
+Thus, some areas on the outer coast have had a more stable sea-level, been subject to past marine transgression phases, or are still inundated today [e.g.
+@svendsen1987; @romundset2015\].
+In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating palimpsests of settlement events that are difficult to disentangle.
+Transgression, on the other hand, can lead to complete erosion of the sites, or bury them in marine sediments, leading for example to a hiatus in the archaeological record for certain sub-phases in the impacted areas \[@bjerck2008].
+Comparatively, given a continuous and relatively rapid shoreline regression from as high as c.
+210 m above present sea level in the inner Oslo fjord , any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgression phases.
+In sum, this makes the region especially useful for evaluating the assumption of a shore-bound settlement over time, and for identifying potential deviations from this pattern.
 
-Following the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, despite corresponding eustatic sea-level rise.
-In other words, the relative sea level has been dropping throughout prehistory---a process that also continues to this day.
-In other areas of Norway the situation is sometimes more complicated.
-On the outer coast, the relative sea-level change has transitioned between regression and transgression phases, and the archaeological remains can be destroyed.
-In other areas, the sea-level has been stable over longer periods of time, meaning coastal locations have often been reused multiple times over long time-spans, creating a palmpiset of settlement events that are impossible to disentangle.
-Comparatively, given the continuous and relatively rapid shoreline regression in south-eastern Norway, any one location in the region has only been shore-bound within a relatively limited time-span.
-As this general displacement trajectory is the same throughout the entirety of prehistory, this makes the region especially useful for testing of the assumption of a shore-bound settlement and to identify potential deviations from this pattern.
+such as Lista in southern Norway has In the most perhipheral coastal areas the relative sea-level change has transitioned between regression and transgression phases, leading to a situation where
 
-The most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves.
+Based on the premise that most Stone Age sites located lower than the marine limit were situated on or close to the shoreline, the most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves.
 If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this.
 In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions.
-Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was *not* shore bound [e.g.
+Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g.
 @jaksland2014].
 This can for example pertain to accessibility.
-If the site is situated on a ledge in a steep and jagged area of the present day landscape, it would make intuitive sense that the location was used when the ocean reached the elevation of the site which would then have been accessible by means of watercraft.
+If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if would have been located on an island.
+In these cases it would make intuitive sense that the location was used when the ocean reached the elevation of the site which would then have been accessible by means of watercraft.
 This line of reasoning is often also used in survey projects, which typically do not produce material suitable for radiocarbon dating.
 The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations.
-Although it appears that the arguments for such site locations is sensible and for a large part appear to be correct, few attempts at a comprehensive evaluation and quantification of this tendency has been done.
+Although it appears that the arguments for such site locations is sensible, few attempts at a comprehensive evaluation and quantification of this tendency has been done.
 
-One of the only larger scale evaulations of this relationship in Norwegian archaeology has been done by Solheim and collegues (@solheim20, see also @solheim2020), who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
+One of the only larger scale evaulations of this relationship in Norwegian archaeology has been done by Solheim and colleagues (@solheim20, see also @solheim2020), who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
 They found a overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5 % of the sites.
-However, in several of the cases where there was a discrepancy, the
+However, in several of the cases where there was a discrepancy, the.
+As was exemplified in Solheims study, and which has been convincingly argued in several individual studies, many of the sites where there is a mismatch between the radiocarbon dates and the shoreline dates, the assemblages are clearly related to the shoreline date rather than the radiocarbon dates.
+Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates or whether they are the result of contaminated samples can be difficult to evaluate and will here be entirely dependent on the discretion of the original excavation report.
+While such evaluations can be difficult, it is nonetheless a necessity that will be held strictly separate from the formal equivalence tests in the analysis to follow.
 
@@
 While this consitutes the formal dimension of the study, there must necessarily also be a qualitative and more subjectivly driven dimension as well.
-As was exemplified in Solheims study, and which has been convincingly argued in several individual studies, many of the sites where there is a mismatch between the radiocarbon dates and the shoreline dates, the assemblages are clearly related to the shoreline date rather than the radiocarbon dates.
-Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates or whether they are the result of contaminated samples can be difficult to evaluate and will here be entirely dependent on the discretion of the original excavation report.
-While such evaluations can be difficult, it is nonetheless a necessity that will be held strictly separate from the formal equivalence tests in the analysis to follow.
 
@@
 These number a total nrow() sites, of which nrow() have radiocarbon dates.
-Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age, and nrow() sites have radiocarbon dates to the Stone Age that were evaluated by the original excavators to relate to the lithic inventory on the sites (see also below).
-Spatial data in the form of features and site limits as defined by the excavation was retrieved from local databases at the Museum of Cultural History.
-Each radiocarbon date has been associated with site features or, where these weren't available, the spatial limit of the entire site.
+Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age.
+Spatial data in the form of features and site limits as defined by the excavating archaeologists was retrieved from local databases at the Museum of Cultural History.
+Each radiocarbon date has been associated with site features or where these weren't available, the spatial limit of the entire site.
 Due to somewhat variable practices between excavations, what spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports.
@@
 Each value was then divided by the meter distance between the isobases.
-This was done for both ends of the confidence intervals.
 To find the adjusted displacement curve for a given site location the distance between the site polygon and closest isobase was multiplied with the values for difference per meter.
@@
 .
-All dates not younger than 1700 BCE with 99.7 % uncertainty were then excluded from further analysis.
-Radiocarbon dates associated with each site were then grouped if they overlaped with 99.7 % uncertainty, effectively taken to represent the same settlement phase.
+All dates not younger than 1700 BCE with 99.7 % probability were then excluded from further analysis.
+Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase.
 In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal.
@@
 
-As the excavation of archaeological sites typically follow from residential and commercial development and the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances.
+As the excavation of archaeological sites typically follow from residential and commercial development, and the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances.
 In addition to employing 10 m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster.
-This involved manually defining the extent of problem areas such as railways, highways, quarries and the like, using polygon features.
+This involved manually defining the extent of problem areas such as railways, highways, quarries and the like.
 The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain.
@@
 
-Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulation was performed.
-Each simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase.
-This year then has a corresponding likely elevation range of the contemporaneous shoreline.
-An elevation value was drawn uniformly from this range, using intervals of 5 cm.
+Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulations were performed.
+A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site.
+This year then has a corresponding likely elevation range for the contemporaneous shoreline.
+An elevation value was then drawn uniformly from this range, using intervals of 5 cm.
 The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing.
 Polygons were then created from the resulting areas with missing values.
-The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon.
-The so-called topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM.
+The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea level from the lowest elevation of the site polygon.
+The topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM.
 This was done using the topoDist package for R.
-In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
-In the case that the sea-polygons completely contain the site, the horisontal and topographic distance measures were made negative.
-While its safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure and might also help identify discrepancies in displacement data or radiocarbon dates.
+The topographic distance was measured between site polygon and the point on the shoreline found to be horisontally closest.
+This was done using In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
+In the case that the sea-polygon completely contain the site, the horisontal and topographic distance measures were made negative.
+While its safe to assume that an archaeological site was not occupied when it was located beneath sea level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
+Negative values where therefore retained.
 
-horisontal and topographic distance between site and shore-line was measured.\
+\
 This process was repeated 1000 times for each phase for each site.
-The choice of 1000 iterations follows an evaluation of convergence on median distances using the example site Hovland 3.
+The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations on the example site Hovland 5, which has a uncertain date range and is located in area of quite complex topography.
 At 1000 simulation runs the analysis of each site ranged from hours, depending on the distance to the simulated shorelines and consequently the necessary size of the window of analysis.

### 2021-11-18 analysis/script/05script.R

@@ -44,6 +44,10 @@ sites_sa <- st_join(st_make_valid(sites_sa), isopolys,
 # Prespecified height of plot (to be multiplied by number of phases)
 plot_height <- 76
 
+# Specify number of simulation runs based on convergence below
+# (1000 seems adequate as only minute variation is observable beyond this point)
+nsamp = 1000
+
 # Identifying number of simulation runs to perform per site by evaluating
 # when mean distance values converges for a chosen test site with complex
 # topographic surroundings and uncertain date range.
@@ -83,15 +87,11 @@ simplt3 <- ggplot(simdat) +
   theme_classic()
 
 simplt1 + simplt2 + simplt3 +
-plot_annotation("Test-run on Hovland 5 to inform sample size")
+plot_annotation("Test-run on Hovland 5 to inform number of simulation runs")
 
 ggsave(file = here("analysis/figures/sample_size.png"), width = 300,
        height = 120,
        units = "mm")
-
-# Specify number of simulation runs based on convergence above
-# (1000 seems adequate as only minute variation is observable beyond this point)
-nsamp = 1000
 
 # Some manual inspection was necessary in case of inconsistencies in the DTM
 # (i.e. a highway running by the site) and for defining necessary size of the
@@ -587,7 +587,7 @@ ggsave(file = here("analysis/figures/hydal4.png"), width = 250,
 sitename <- "Krøgenes D1"
 date_groups <- group_dates(rcarb_sa, sitename)
 
-output <- apply_functions(sitename, dtm, displacement_curves,
+output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
                           isobases, nsamp = nsamp, loc_bbox = 400, siterpath)
 save(output,
      file = here::here("analysis/data/derived_data/krogenesd1.RData"))
@@ -596,7 +596,7 @@ sitearea <- rast(file.path(siterpath,
                            paste0(str_replace(sitename, " ", "_"), ".tif")))
 
 plot_results(sitename, output$sitel, output$datedat, sitearea, bmap,
-             sites_sa, isobases, output, scale_dist = 150)
+             sites_sa, isobases, output, date_groups, scale_dist = 150)
 
 ggsave(file = here("analysis/figures/krogenesd1.png"), width = 250,
        height = plot_height + (plot_height * length(unique(date_groups))),
[…]

### 2021-11-21 analysis/paper/paper.Rmd
Set up plotting. Chekc displacement curve for Løvås sites (cf. negative values)

@@
 Taking inspiration from recent studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g.
-@crema2010; @crema2012; @lewis2021], a similar approach is adopted here.
+@crema2010; @crema2012], a similar approach is adopted here.
 
@@
 Of central importance here is eustasy and glacial istostasy.
-The eustatic sea level is understood as the sea level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015\].
-The eustatic sea level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that has volumetric impact on ocean basins.
+The eustatic sea level is understood as the sea level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015].
+The eustatic sea level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that impact the volume of ocean basins.
 Istostasy, on the other hand, pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle.
 This is often the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed.
-These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight \[@milne2015].
+These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight [@milne2015].
 
@@
 Thus, some areas on the outer coast have had a more stable sea-level, been subject to past marine transgression phases, or are still inundated today [e.g.
-@svendsen1987; @romundset2015\].
+@svendsen1987; @romundset2015].
 In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating palimpsests of settlement events that are difficult to disentangle.
-Transgression, on the other hand, can lead to complete erosion of the sites, or bury them in marine sediments, leading for example to a hiatus in the archaeological record for certain sub-phases in the impacted areas \[@bjerck2008].
-Comparatively, given a continuous and relatively rapid shoreline regression from as high as c.
+Transgression, on the other hand, can lead to complete erosion of the sites, or bury them in marine sediments, leading for example to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008].
+Comparatively, given a continuous and still ongoing shoreline regression from as high as c.
 210 m above present sea level in the inner Oslo fjord , any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgression phases.
-In sum, this makes the region especially useful for evaluating the assumption of a shore-bound settlement over time, and for identifying potential deviations from this pattern.
+This makes the region especially useful for evaluating the assumption of a shore-bound settlement over time, and for identifying potential deviations from this pattern.
 
@@
 
-Based on the premise that most Stone Age sites located lower than the marine limit were situated on or close to the shoreline, the most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves.
-If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this.
+Previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically involved a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the site elevation [@breivik2018; @solheim2020; @åstveit2018\].
+This approach has a couple of limitations.
+Firstly, only a couple of studies have taken the variable uplift-rates into account when performing this comparison .
+Secondly, the question of interest in this case is effectively one of synchroneity between events, as opposed to difference \[@parnell2008].
+This framing consequently flips the direction of the common null-hypothesis of no significant difference too one of equivalence.
+While two overlapping confidence intervals means that one cannot conclude that they are different, this does not necessarily mean that they are the same.
+One implication of this method is consequently that the wider the uncertainty range for either radiocarbon- or shoreline date, the higher the liklihood that the confidence intervals overlap, and the higher the likelihood that we conclude in favour of our hypothesis.
+The final limitation of this method is that it only takes the vertical distance between site and shoreline into account.
+While this is the parameter of interest for shoreline dating, the practical implications of a vertical difference in elevation will be highly dependent on the local topography.
+
 In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions.
@@
 If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if would have been located on an island.
-In these cases it would make intuitive sense that the location was used when the ocean reached the elevation of the site which would then have been accessible by means of watercraft.
-This line of reasoning is often also used in survey projects, which typically do not produce material suitable for radiocarbon dating.
+In these cases it would make intuitive sense that the location was used when the ocean reached the elevation of the site, which would then have been accessible by means of watercraft.
 The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations.
 Although it appears that the arguments for such site locations is sensible, few attempts at a comprehensive evaluation and quantification of this tendency has been done.
+
+Based on the generally accepted premise that most Stone Age sites located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shorebound site location unless there is strong evidence to suggest otherwise.
+This is for example common in survey projects, which can be guided by a mental or digital reconstruction of the landscape.
+typically do not produce material suitable for radiocarbon dating.
+the most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves.
+If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this.
 
@@
 Spatial data in the form of features and site limits as defined by the excavating archaeologists was retrieved from local databases at the Museum of Cultural History.
-Each radiocarbon date has been associated with site features or where these weren't available, the spatial limit of the entire site.
+Each radiocarbon date has been associated with site features or, where these weren't available, the spatial limit of the entire site.
 Due to somewhat variable practices between excavations, what spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports.
@@
 Due to the dsfg there is a gradient in the rate of isostatic rebound, with increasing rebound rates towards the north-east.
-To take account of this gradient, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase.
+To take account of this, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase.
 This was done by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves.
 Each value was then divided by the meter distance between the isobases.
-To find the adjusted displacement curve for a given site location the distance between the site polygon and closest isobase was multiplied with the values for difference per meter.
+To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter.
 If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east, the values were instead added to the original curve.
@@
 The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain.
-This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls from GRASS GIS in R through the rgrass7 package.
-In addition to code and original spatial data being available in the research compendia for this paper, it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.
+This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls from GRASS GIS in R through the package rgrass7.
+In addition to code and original spatial data being available in the research compendium for this paper, it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.
 
@@
 Negative values where therefore retained.
+
+Some sources of likely varation and uncertainty have not been considered here.
+First of all the DTM has only been corrected for major modern disturbances.
+This means that erosion, although likely not that prevalent, have not been taken into account.
+Secondly, the DTM has a vertical error of.
+Thirdly, the tidal range has not been considered.
+This is around 0.5 m within the study area today, and there is no reason to think this was substantially different in the past.
+Fourthly, the exposure of the sites to wave action is also likely to have been of concern, and will presumably have implications for how close to the shoreline people settled.
 

### 2021-11-22 analysis/paper/paper.Rmd
Assembled preliminary results plot

@@
 
-Previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically involved a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the site elevation [@breivik2018; @solheim2020; @åstveit2018\].
+Previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically involved a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the site elevation [e.g.
+@breivik2018; @solheim2020; @åstveit2018\].
 This approach has a couple of limitations.
-Firstly, only a couple of studies have taken the variable uplift-rates into account when performing this comparison .
-Secondly, the question of interest in this case is effectively one of synchroneity between events, as opposed to difference \[@parnell2008].
-This framing consequently flips the direction of the common null-hypothesis of no significant difference too one of equivalence.
-While two overlapping confidence intervals means that one cannot conclude that they are different, this does not necessarily mean that they are the same.
-One implication of this method is consequently that the wider the uncertainty range for either radiocarbon- or shoreline date, the higher the liklihood that the confidence intervals overlap, and the higher the likelihood that we conclude in favour of our hypothesis.
-The final limitation of this method is that it only takes the vertical distance between site and shoreline into account.
-While this is the parameter of interest for shoreline dating, the practical implications of a vertical difference in elevation will be highly dependent on the local topography.
+First of all, only a couple of studies have taken the variable uplift-rates into account when performing this comparison.
+Secondly, following this method, the wider the uncertainty range associated with either radiocarbon- or shoreline date, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis.
+While two overlapping confidence intervals means that one cannot conclude that the dates are different, this does not necessarily mean that they are the same.
+The question thus necessitates a flip from a null-hypothesis of no signficant difference to one of equivalence, as the question of interest in this case is effectively one of synchroneity \[cf.
+@parnell2008].
+A final limitation of this method is that it only takes into account the vertical distance between site and shoreline.
+While this is the parameter of interest for shoreline dating, the practical implications of a vertical difference in elevation will be highly dependent on local topography and bathymetry.
 

### 2021-11-25 analysis/paper/paper.Rmd
Minor changes

@@
 This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
-This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, harbour constructions and, as is the focus of this study, Stone Age sites.
+This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites.
 
 The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C.
-Brøgger in 1905 (see @maccurdy1906).
+Brøgger in 1905 (see the review by @maccurdy1906 for an early mention in english).
 Shoreline dating has been fundamental to Stone Age research in Norway ever since.
@@
 
-Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods.
+Despite the fundamental role of shoreline dating for Norwegian Stone Age archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods.
 The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach.
@@
 In other words, the relative sea level has been dropping throughout prehistory.
-As this process is the result of glacial loading, the rate of uplift has been more severe towards the centre of the ice sheet.
-Thus, some areas on the outer coast have had a more stable sea-level, been subject to past marine transgression phases, or are still inundated today [e.g.
+As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet.
+Thus, some areas on the outer coast have had a more stable relative sea-level, been subject to past marine transgression phases, or remain inundated to this day [e.g.
 @svendsen1987; @romundset2015].
+The has direct archaeological impact.
 In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating palimpsests of settlement events that are difficult to disentangle.
-Transgression, on the other hand, can lead to complete erosion of the sites, or bury them in marine sediments, leading for example to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008].
+Transgression, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading for example to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008].
 Comparatively, given a continuous and still ongoing shoreline regression from as high as c.
-210 m above present sea level in the inner Oslo fjord , any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgression phases.
-This makes the region especially useful for evaluating the assumption of a shore-bound settlement over time, and for identifying potential deviations from this pattern.
+210 m above present sea level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions.
+This makes the region especially useful for evaluating the assumption of a shore-bound settlement over time.
 
 such as Lista in southern Norway has In the most perhipheral coastal areas the relative sea-level change has transitioned between regression and transgression phases, leading to a situation where
+
+Based on the generally accepted premise that most Stone Age sites located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
+This is for example common in survey projects, which can be guided by a both a digital and mental reconstruction of past sea level.
+Following an excavation, the most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves.
+If this corresponds then it is common to assume a shore bound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this.
+
+Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g.
+@jaksland2014].
+This can for example pertain to accessibility.
+If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if would have been located on an island, it would make intuitive sense that the site was used when the ocean reached close to its elevation, which would then have made the site accessible by means of watercraft.
+The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations.
+Although it appears that the arguments for such site locations is sensible, few attempts at a comprehensive evaluation and quantification of this tendency has been done.
 
@@
 This approach has a couple of limitations.
-First of all, only a couple of studies have taken the variable uplift-rates into account when performing this comparison.
+First of all, only a couple of studies have taken the variable uplift-rates into account when performing this comparison \[@fossum2020\].
 Secondly, following this method, the wider the uncertainty range associated with either radiocarbon- or shoreline date, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis.
@@
 @parnell2008].
-A final limitation of this method is that it only takes into account the vertical distance between site and shoreline.
-While this is the parameter of interest for shoreline dating, the practical implications of a vertical difference in elevation will be highly dependent on local topography and bathymetry.
+Another limitation of this often employed method is that it only takes into account the vertical distance between the sites and the shoreline.
+While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in elevation will be highly dependent on local topography and bathymetry.
 
 In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions.
-Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g.
-@jaksland2014].
-This can for example pertain to accessibility.
-If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if would have been located on an island.
-In these cases it would make intuitive sense that the location was used when the ocean reached the elevation of the site, which would then have been accessible by means of watercraft.
-The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations.
-Although it appears that the arguments for such site locations is sensible, few attempts at a comprehensive evaluation and quantification of this tendency has been done.
-
-Based on the generally accepted premise that most Stone Age sites located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shorebound site location unless there is strong evidence to suggest otherwise.
-This is for example common in survey projects, which can be guided by a mental or digital reconstruction of the landscape.
-typically do not produce material suitable for radiocarbon dating.
-the most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves.
-If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this.
 
@@
 If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east, the values were instead added to the original curve.
-The result of this process is shown for an example site in Figure.
+The result of this process is shown for an example site in .
 
@@
 
-Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulations were performed.
+Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulations were performed.
 A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site.
@@
 The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea level from the lowest elevation of the site polygon.
-The topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM.
-This was done using the topoDist package for R.
+The topographic distance between site and sea was also found by measuring the shortest distance while taking into account the slope of the terrain on the DTM.
+This was done using a topoDist package for R.
 The topographic distance was measured between site polygon and the point on the shoreline found to be horisontally closest.
-This was done using In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
+The shortest path was found using the Queen's case neighbourhood of eight cells.
+In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
 In the case that the sea-polygon completely contain the site, the horisontal and topographic distance measures were made negative.
@@
 
-Some sources of likely varation and uncertainty have not been considered here.
+Some sources of likely variation and uncertainty have not been considered here.
 First of all the DTM has only been corrected for major modern disturbances.
@@
 
-The approach is only used to inform the question of the distance between sites and shorelines here, but this method can also be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
+Here the approach was only used to inform the question of the distance between sites and shorelines, but this method can also be extended to a wide range of use-cases where one needs to visualise as well as quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore line, and the uncertainty inherent in this reconstruction.
 

### 2021-11-30 analysis/paper/paper.Rmd
Supplementary material. Created figures for text.

@@
 author:
-  - Myfirstname Mylastname:
-      email: fl@oneplace.edu
+  - Isak Roalkvam:
+      email: isak.roalkvam@iakh.uio.no
       institute: [UofO]
       correspondence: true
-  - Yourfirstname Yourlastname:
-      email: fl@another.edu
-      institute: [UofA]
-      correspondence: false
 institute:
-  - UofO: University of One Place
-  - UofA: University of Another Place
+  - UofO: University of Oslo
 date: "`r format(Sys.time(), '%d %B, %Y')`"
 output: 
-    bookdown::word_document2:
+    bookdown::pdf_document2:
+      toc: false
       fig_caption: yes
@@
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
-abstract: |
-  Text of abstract
-keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
 ---
-
-Keywords: `r rmarkdown::metadata$keywords`
-
-Highlights: `r rmarkdown::metadata$highlights`
 
@@
 
+```{r libraries, eval = TRUE, results = 'hide'}
+library(here)
+library(tidyverse)
+library(sf)
+library(ggridges)
+library(ggthemes)
+library(ggnewscale)
+library(patchwork)
+library(terra)
+library(raster)
+library(oxcAAR)
+library(vapour)
+library(topoDistance)
+library(IRanges)
+quickSetupOxcal()
+```
+
+```{r get-data-functions, include = FALSE}
+# Load all sites and radiocarbon data
+sites <- read.csv(here('analysis/data/raw_data/sites.csv'))
+rcarbon <- read.csv(here('analysis/data/raw_data/radiocarbon.csv'))
+
+# Background map
+bmap <- st_read(here('analysis/data/raw_data/naturalearth_countries.gpkg'))
+
+# Digital terrain model
+dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
+
+# Load processed site and radiocarbon data
+load(here("analysis/data/derived_data/01data.RData"))
+load(here("analysis/data/derived_data/02data.RData"))
+source(here("analysis/script/04script.R"), local = knitr::knit_global())
+```
+
 # Introduction
 
-The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology.
-This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
-This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites.
+The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology.
+This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
+This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g.
+@berg-hansen2009; @bjerck2005; @sognnes2003].
 
 The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C.
-Brøgger in 1905 (see the review by @maccurdy1906 for an early mention in english).
-Shoreline dating has been fundamental to Stone Age research in Norway ever since.
+Brøgger in 1905 (see the review by @maccurdy1906 for an early mention in English).
+Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g.
+@berg-hansen2009; @nummedal1923; @bjerck1990].
 The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
-Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites.
+Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites.
 Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age.
@@
 
-Relative sea-level can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre [@shennan2015\].
-Variation in this relative distance follow from a range of effects \[e.g.
+Relative sea-level (RSL) can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre [@shennan2015].
+Variation in this relative distance follow from a range of effects [e.g.
 @mörner1976; @milne2009].
@@
 The eustatic sea level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that impact the volume of ocean basins.
-Istostasy, on the other hand, pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle.
+Istostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle.
 This is often the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed.
@@
 
-Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet, the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise.
+Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g.
+@stroeven2016], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise .
 In other words, the relative sea level has been dropping throughout prehistory.
 As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet.
-Thus, some areas on the outer coast have had a more stable relative sea-level, been subject to past marine transgression phases, or remain inundated to this day [e.g.
+Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g.
 @svendsen1987; @romundset2015].
-The has direct archaeological impact.
-In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating palimpsests of settlement events that are difficult to disentangle.
-Transgression, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading for example to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008].
+These conditions are directly reflected in the archaeological record.
+In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating palimpsests of settlement events that are difficult to disentangle [@reitan2009].
+Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008].
 Comparatively, given a continuous and still ongoing shoreline regression from as high as c.
-210 m above present sea level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions.
-This makes the region especially useful for evaluating the assumption of a shore-bound settlement over time.
+220 m above present sea level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @sørensen1979].
+This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.
 
@@
 Based on the generally accepted premise that most Stone Age sites located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
-This is for example common in survey projects, which can be guided by a both a digital and mental reconstruction of past sea level.
+This is for example common in survey projects, which are often guided by a both a digital and mental reconstruction of past sea levels.
 Following an excavation, the most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves.
@@
 This can for example pertain to accessibility.
-If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if would have been located on an island, it would make intuitive sense that the site was used when the ocean reached close to its elevation, which would then have made the site accessible by means of watercraft.
+If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if it would have been located on an island when the sea level was higher, it would make intuitive sense that the site was used when the ocean reached close to its elevation, as this would have made the site accessible by means of watercraft.
 The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations.
@@
 
-Previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically involved a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the site elevation [e.g.
-@breivik2018; @solheim2020; @åstveit2018\].
+One of the larger evaluations of this relationship has been done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area---data that are also included in this study.
+They found a overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5 % of the sites.
+However, in the cases where there was a discrepancy, the original .
+As was exemplified in Solheims study, and which has been convincingly argued in several individual studies, many of the sites where there is a mismatch between the radiocarbon dates and the shoreline dates, the typological indicators in the assemblages are clearly unrelated to the radiocarbon dates.
+Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates, or whether they are entirely erroneous results of contamination can be difficult to evaluate, and will here be entirely dependent on the discretion of the original excavation report.
+
+While such evaluations can be difficult, it is nonetheless a necessity that will be held strictly separate from the formal equivalence tests in the analysis to follow.
+
+Previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
+@solheim2020; @åstveit2018].
 This approach has a couple of limitations.
-First of all, only a couple of studies have taken the variable uplift-rates into account when performing this comparison \[@fossum2020\].
-Secondly, following this method, the wider the uncertainty range associated with either radiocarbon- or shoreline date, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis.
-While two overlapping confidence intervals means that one cannot conclude that the dates are different, this does not necessarily mean that they are the same.
-The question thus necessitates a flip from a null-hypothesis of no signficant difference to one of equivalence, as the question of interest in this case is effectively one of synchroneity \[cf.
+First of all, the displacement curves are commonly applied directly to larger study areas, with only a couple of studies having taken the variable uplift-rates into account when performing this comparison [e.g.
+@møller1987; @fossum2020].
+Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis.
+This thus leads to an inferential framework that favours uncertainty, which is hardly desirable.
+In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same.
+The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g.
+@lakens2018], as the question of interest is effectively one of synchroneity between events [cf.
 @parnell2008].
-Another limitation of this often employed method is that it only takes into account the vertical distance between the sites and the shoreline.
-While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in elevation will be highly dependent on local topography and bathymetry.
+Another limitation of this often employed method is that it only takes into account the vertical distance between the sites and the sea level.
+While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry.
+RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horisontal displacement of the shoreline will be greater.
+Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of the relation between site and shore line, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.
+Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.
 
@@
 
-One of the only larger scale evaulations of this relationship in Norwegian archaeology has been done by Solheim and colleagues (@solheim20, see also @solheim2020), who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
-They found a overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5 % of the sites.
-However, in several of the cases where there was a discrepancy, the.
-As was exemplified in Solheims study, and which has been convincingly argued in several individual studies, many of the sites where there is a mismatch between the radiocarbon dates and the shoreline dates, the assemblages are clearly related to the shoreline date rather than the radiocarbon dates.
-Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates or whether they are the result of contaminated samples can be difficult to evaluate and will here be entirely dependent on the discretion of the original excavation report.
-While such evaluations can be difficult, it is nonetheless a necessity that will be held strictly separate from the formal equivalence tests in the analysis to follow.
-
-Given that the aim of this study is to test whether or not there is a match between dates provided by means of shoreline displaceent curves and radiocarbon dates, it is appropriate to evaulate this relationship by means of statistical equivalence tests.
+Given that the aim of this study is to test whether or not there is a match between dates provided by means of shoreline displacement curves and 14C-dates, it is appropriate to evaluate this relationship by means of statistical equivalence tests.
 This effectively shifts the common direction of the null-hypothesis, giving H0: The date given by means of radiocarbon dating and by shoreline displacement are different.
@@
 
-This study necessaitated the compilation of multiple relative sea-level curves developed for different parts of the region.
-Most of the curves come provided with a confidence interval which is largely based on an evaluation by the geologists who developed the curve.
-The reason why these confidence intervals are not strictly mathematically derived is that the curves do not only contain uncertainy as related to radiometric dates, which are fairly well understood, but also hold potential error as related to the interpretation of sediment cores, placement of samples, as well as direction and adjustment to a common isobase based on the direction and gradient in rebound intensity, to name but a few.
-The error bands are consequently complex, and cannot necessarily be directly compared between curves as these have been developed at different times, using a wide range of different methods.
+cannot necessarily be directly compared between curves as these have been developed at different times, using a wide range of different methods.
 However, any reinterpretation and standardisation of the curves is also hampered by a range of issues.
@@
 
-All data that are required to run the analyses and the derived data are freely available in an online repository.
-The dataset is slightly more extensive, including excavated sites without radiocarbon dates, and radiocarbon dates to periods other than the Stone Age.
+All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a digital research compendium following @marwick2018.
+The dataset is slightly more extensive than what is directly utilised in the study, and includes excavated sites without ^14^C-dates, as well as ^14^C-dates to periods other than the Stone Age.
 
-The data compiled for the analysis consists of all excavated Stone Age sites from the region between Horten county in the north-east, to Arendal in the south-west.
-These number a total nrow() sites, of which nrow() have radiocarbon dates.
-Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age.
-Spatial data in the form of features and site limits as defined by the excavating archaeologists was retrieved from local databases at the Museum of Cultural History.
-Each radiocarbon date has been associated with site features or, where these weren't available, the spatial limit of the entire site.
-Due to somewhat variable practices between excavations, what spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports.
-This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hulls generated around the site features.
+To get at the relationship between sites and contemporaneous shoreline, the analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
+While there is displacement data available for other areas of south eastern Norway [e.g.
+@hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Halden county in the north-east, to Arendal in the south-west.
+This area has newly compiled displacement curves for Skoppum in Halden [@romundset2021], Larvik [@sørensen2014; @sørensen2014b], Hanto in Tvedestrand [@romundset2018; @romundset2018b], and Bjørnebu in Arendal [@romundset2018b].
+
+The employed shoreline displacement curves are all based on the so-called isolation basin method, which represents an extremely precise method compared to other [@romundset2019].
+This involves extracting cores from a series of basins situated on bedrock in a constrained area of the landscape, and dating the transition from marine to lacustrine .
+Each curve is thus construed from a series of cored basins located at different elevations, each representing a sea-level index point.
+The reason why these confidence intervals are not strictly mathematically derived is that the curves do not only contain uncertainty as related to radiometric dates, which are well understood, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets, the adjustment to a common isobase based on the direction and gradient in rebound intensity, to name but a few [e.g.
+@romundset2010].
+The error bands and their trajectory are consequently quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
+For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications (see above).
+
+The archaeological data compiled for the analysis consists of all excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)).
+These number a total `r nrow(sites)` sites, of which `r length(unique(rcarbon$site_name))` sites are associated with the total of `r nrow(rcarbon)` radiocarbon dates.
+Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability.
+These sites form the basis for the analysis.
+Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History---the institution responsible for archaeological excavations in the region.
+In the compiled dataset, each radiocarbon date has been associated with the site features or excavation unit from where they originate, or, where these weren't available, the spatial limit of the entire site.
+Due to somewhat variable practices between excavations, what available spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports.
+This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.
+
+```{r map-iso,  fig.cap = "A) Distribution of the 70 analysed sites relative to the isobases of the displacement curves, B) Displacement curves.
+Note the higher  steepness of the curves that are situated towards the north-east.", fig.height = 13}
+bboxsites <- st_bbox(sites_sa)
+bboxsites[1] <- bboxsites[1] - 15000
+bboxsites[3] <- bboxsites[3] + 15000
+bboxsites[2] <- bboxsites[2] - 5000
+bboxsites[4] <- bboxsites[4] + 5000
+bboxsitespoly <- st_as_sf(st_as_sfc(bboxsites))
+
+bound_reproj <- st_transform(bboxsitespoly, st_crs(bmap))
+bmap2 <- bmap %>%
+    filter(st_intersects(., bound_reproj, sparse = FALSE))
+bmap_reproj <- st_transform(bmap2, st_crs(sites_sa))
+
+anc <- as.numeric(c(bboxsites$ymin, bboxsites$xmax))
+overview_map <- ggplot() +
+  geom_sf(data = bmap_reproj, fill = "grey", colour = NA) +
+  geom_sf(data = isobases, aes(colour = name)) +
+  geom_sf(data = st_centroid(sites_sa), fill = "red",
+          size = 2.25, shape = 21,
+          colour = "black") +
+  ggsn::scalebar(data = sites_sa, dist = 20, dist_unit = "km",
+                 transform = FALSE, st.size = 4, height = 0.02,
+                 border.size = 0.1, st.dist = 0.03,
+                 anchor = c(x = anc[2] - 15500, y = anc[1]) + 8000) +
+  coord_sf(xlim = c(bboxsites[1], bboxsites[3]),
+           ylim = c(bboxsites[2], bboxsites[4]),
+           expand = FALSE) +
+  scale_colour_manual(values = c("Arendal" = "black",
+                                   "Larvik" = "darkgreen",
+                                   "Tvedestrand" = "blue",
+                                   "Halden" = 'red')) +
+  theme_bw() + theme(axis.title=element_blank(),
+                     axis.text.y = element_blank(),
+                     axis.text.x = element_blank(),
+                     rect = element_rect(),
+                     axis.ticks = element_blank(),
+                     panel.grid.major = element_blank(),
+                     legend.position = "none")
+
+disp_curves <- ggplot(displacement_curves, aes(x = years, col = name)) +
+  geom_line(aes(y =  upperelev)) +
+  geom_line(aes(y = lowerelev)) +
+  ylab("Meters above present sea level") +
+  xlab("cal BCE/CE") +
+  scale_colour_manual(values = c("Arendal" = "black",
+                                   "Larvik" = "darkgreen",
+                                   "Tvedestrand" = "blue",
+                                   "Halden" = 'red')) +
+  theme_bw() +
+  theme(legend.title = element_blank(), legend.position = "bottom",
+        legend.direction = "horizontal")
+
+overview_map / disp_curves
+```
 
@@
 This was done by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves.
-Each value was then divided by the meter distance between the isobases.
+Each value was then divided by the distance between the isobases.
 To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter.
@@
 
-For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve using OxCal v.
-.
-All dates not younger than 1700 BCE with 99.7 % probability were then excluded from further analysis.
+For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v.
+[@bronkramsey2009] through the oxAAR package for R.
+All dates not older than 1700 BCE with 99.7 % probability were then excluded from further analysis.
 Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase.
@@
 
-As the excavation of archaeological sites typically follow from residential and commercial development, and the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances.
+As the excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances.
 In addition to employing 10 m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster.
@@
 The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain.
-This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls from GRASS GIS in R through the package rgrass7.
+This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls from GRASS GIS in R through the package rgrass7 [@bivand2021].
 In addition to code and original spatial data being available in the research compendium for this paper, it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.
@@
 The topographic distance between site and sea was also found by measuring the shortest distance while taking into account the slope of the terrain on the DTM.
-This was done using a topoDist package for R.
-The topographic distance was measured between site polygon and the point on the shoreline found to be horisontally closest.
-The shortest path was found using the Queen's case neighbourhood of eight cells.
+This was done using the topoDistance package for R.
+The topographic distance was measured between site polygon and the point on the shoreline found to be closest measured horisontally.
+The shortest topographic path was found using the Queen's case neighbourhood of eight cells.
 In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
-In the case that the sea-polygon completely contain the site, the horisontal and topographic distance measures were made negative.
-While its safe to assume that an archaeological site was not occupied when it was located beneath sea level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
-Negative values where therefore retained.
+In the case that the sea-polygon completely contain the site, the horisontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon.
+While it is safe to assume that an archaeological site was not occupied when it was located beneath sea level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
+Negative values where therefore retained unless otherwise stated.
 
@@
 
-\
 This process was repeated 1000 times for each phase for each site.
@@
 
-```{r get-data, eval = FALSE}
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
+```{r example-site, fig.show='hide'}
+
+# Assign closest isobases to site data.
+sites_sa <- st_join(st_make_valid(sites_sa), isopolys,
+                    join = st_intersects, largest = TRUE)
+
+# Specify site name
+sitename <- "Hegna vest 1"
+
+# Group dates that overlap with 99.7 % probability using the group_dates()
+# function. All functions in this chunk where package isnt specified are defined 
+# in 04script.R
+date_groups <- group_dates(rcarb_sa, sitename)
+
+# Retrieve site limit and site features
+siter <- dplyr::filter(rcarb_sa, site_name == sitename)
+sitel <- dplyr::filter(sites_sa, name == sitename)
+
+# Creating bounding box at 400 meters
+location_bbox <- bboxpoly(sitel, 400)
+
+# Clip DTM with the bbox
+sitearea <- terra::crop(dtm, location_bbox)
+
+# Model dates using the Boundary function
+datedat <- model_dates(sitedates = siter, date_groups = date_groups)
+
+# Find local displacement curve
+sitecurve <- interpolate_curve(years = xvals,
+                                 isobase1 = sitel$isobase1,
+                                 isobase2 = sitel$isobase2,
+                                 target = sitel,
+                                 dispdat = displacement_curves,
+                                 isodat = isobases,
+                                 direction_rel_curve1 = sitel$dir_rel_1)
+# Add site name
+sitecurve$name <- sitename
+
+# Plot of dates
+date_plot <- plot_dates(datedat, sitename, title = FALSE)
+site_map <- site_plot(sitearea, sitel, 150, date_groups,
+                      s_tdist = 0.5, s_xpos = 175,
+                       s_ypos = 85,  s_bheight = 0.3, s_tsize = 3)
+overview_map <- overview_plot(bmap, sitel, sites_sa, isobases,
+                               os_tdist = 0.03, os_xpos = 20000, os_ypos = 8000,
+                               os_bheight = 0.02, os_tsize = 3)
+
+dispdata <- filter(displacement_curves, name %in% c("Larvik", "Tvedestrand"))
+
+curve_plot <- ggplot() + 
+              geom_line(data = dispdata, 
+                        aes(x = years, y = upperelev, col = name)) +
+  geom_line(data = dispdata, 
+                        aes(x = years, y = lowerelev, col = name)) +
+  geom_line(data = sitecurve,
+            aes(x = years, y = upperelev, col = name)) +
+  geom_line(data = sitecurve,
+            aes(x = years, y = lowerelev, col = name)) +
+  scale_colour_manual(values = c("Larvik" = "darkgreen",
+                                  "Tvedestrand" = "blue",
+                                 "Hegna vest 1" = "hotpink2")) +
+  ylab("Meters above present sea level") +
+  xlab("Cal BCE/CE") +
+  theme_bw() +
+  theme(legend.title = element_blank(), legend.position = "bottom",
+        legend.direction = "horizontal")
+
+plt <- (site_map + date_plot) /
+(overview_map + curve_plot) +
+  plot_annotation(tag_levels = 'A')
+
+# It was easier to get correct dimensions by saving and then 
+ggsave(file = here("analysis/figures/example_map.png"), plot = plt, width = 250,
+       height = 76 + (76 * length(unique(date_groups))),
+       units = "mm")
+
 ```
 
-```{r demo-plot, fig.cap="A plot of random numbers"}
-plot(rnorm(10))
+```{r example-1, echo = FALSE, fig.cap = "Example site Hegna vest 1.
+A) Location of the site in the present day landscape.
+The red outline is the site limit.
+B) Radiocarbon dates associated with the site.
+Fill colour indicates what dates are assumed to belong to the same settlement phase.
+Multiple dates are modelled using the Boundary function in OxCal and then summed.
+The red outline indicates that the date does not match the typological indicators in the artefact assemblage of the site.
+C) The location of the site within the study area relative to isobases of the employed displacement curves.
+D) Displacement curve interpolated to the site location based on the distance between the site and the isobases of the two closest displacement curves.", out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/example_map.png"))
 ```
 
-```{r demo-inline-code}
-x <- round(pi, 2)
+```{r example-site-cont2, fig.show='hide'}
+set.seed(1)
+
+# Retrieve example date to be used. The posterior sum of date group 1
+posteriorprobs <- filter(datedat, group == 1 & class == "sum")
+
+# Set up to sample a single year
+samplingframe <- list(x = posteriorprobs[["dates"]],
+                      y = posteriorprobs[["probabilities"]])
+
+# Draw a single year
+sdate <- round(with(samplingframe, sample(x, size = 1, prob = y)))
+
+# Create plot showing the year drawn from the posterior sum
+datsamp <- ggplot(posteriorprobs) +
+  ggridges::geom_ridgeline(aes(x = dates, y = name,
+                               height = probabilities * 50),
+                           fill = "lightgrey") +
+  # Create segment to indicate which year was drawn
+  geom_segment(aes(x = sdate, xend = sdate, y = as.numeric(as.factor(name)),
+                   yend = as.numeric(as.factor(name)) +
+                     (posteriorprobs[round(dates) == sdate,
+                                     "probabilities"]) * 50),
+               colour = "red") +
+  labs(x = "cal BCE")  +
+  #scale_y_discrete(expand = c(0.01, 0)) +
+  theme_bw() + theme(
+    legend.position = "none",
+    panel.border = element_blank(),
+    axis.line = element_line("black"),
+    axis.title.y = element_blank(),
+    axis.ticks.y = element_blank()
+  ) +
+  scale_x_continuous(breaks = sort(c(-8200, -8000, sdate, -7600)))
+
+# Find upper and lower shoreline elevation at sdate
+uppers <- round(approx(sitecurve[,"years"], sitecurve[,"upperelev"],
+                       xout = sdate)[["y"]], 1)
+lowers <- round(approx(sitecurve[,"years"], sitecurve[,"lowerelev"],
+                       xout = sdate)[["y"]], 1)
+
+# Draw sealevel uniformly from elevation range, intervals of 5 cm
+sealevel <- sample(seq(lowers, uppers, 0.05), 1)
+
+# Plot the drawn elevation
+elevsamp <- ggplot(sitecurve, aes(x = years)) +
+  geom_line(aes(y =  upperelev), colour = "hotpink2") +
+  geom_line(aes(y = lowerelev), colour = "hotpink2") +
+  ylab("Meters above present sea level") +
+  xlab("cal BCE/CE") +
+  theme_bw() +
+  theme(legend.title = element_blank(), legend.position = "bottom",
+        legend.direction = "horizontal") +
+  geom_segment(aes(x = sdate, xend = sdate, y = uppers,
+                                           yend = -Inf), col = "black",
+               linetype = "solid") +
+  geom_segment(aes(x = -Inf, xend = sdate, y = uppers,
+                   yend = uppers), col = "black", size = 0.25) +
+  geom_segment(aes(x = -Inf, xend = sdate, y = lowers,
+                   yend = lowers), col = "black", size = 0.25) +
+  geom_segment(aes(x = -Inf, xend = sdate, y = lowers,
+                   yend = lowers), col = "black", size = 0.25) +
+  geom_segment(aes(x = -Inf, xend = sdate, y = sealevel,
+                   yend = sealevel), col = "red", size = 0.25,
+               linetype = "dashed") +
+  annotate("text", x = -8700, y = uppers + 2, label = as.character(uppers)) +
+  annotate("text", x = -8700, y = lowers - 2, label = as.character(lowers)) +
+  scale_x_continuous(breaks = sort(c(-9000, sdate, -6000, -3000, 0))) +
+  scale_y_continuous(breaks = sort(c(0, sealevel, 100)))
+
+# Adjust sea level on DTM
+rclmat <- matrix(c(sealevel, Inf, NA, 0, sealevel, 1),
+                 ncol = 3, byrow = TRUE)
+classarea <- terra::classify(sitearea, rclmat)
+seapoly <- st_combine(st_as_sf(terra::as.polygons(classarea)))
+
+# Retrieve lowest elevation on polygon
+siteelev <- extract(sitearea, vect(sitel), fun = min)[2]
+# Find difference to sea-level elevation
+vertdist <-  min(siteelev[1:nrow(siteelev),]) - sealevel
+
+# Find the nearest points on the site- and seapolygons,
+distline <- st_nearest_points(sitel, seapoly)
+
+# Retrieve the two points
+distpts <- st_cast(distline, "POINT")
+
+# Find topographic distance
+coords <- rbind(st_coordinates(distpts[1]),
+                st_coordinates(distpts[2]))
+rownames(coords) <- c("loc", "sea")
+
+# Find topographic distance (retrieve path)
+topodist <- topoDist(raster::raster(sitearea), coords,
+                     paths = TRUE)
+# Store distance
+tdist <- topodist[[1]]["loc", "sea"]
+
+# Retrieve path
+tpath <- st_as_sf(topodist[[2]])
+
+# Find end-points
+endptns <- st_line_sample(tpath, sample = c(0, 1))
+
+# Cast as line
+hordistl <- st_cast(endptns, "LINESTRING")
+
+# Find horisontal distance
+hordist <- st_length(hordistl)
+
+# Create label for map
+dist_label <- paste0("Horisontal distance = ", round(hordist, 1), "m", 
+                "\nTopographic distance = ", round(tdist, 1), "m",
+                "\nVertical distance = ", round(vertdist, 1), "m")
+
+# Anchor for scale bar
+anc <- as.numeric(c(st_bbox(location_bbox)$ymin, st_bbox(location_bbox)$xmax))
+
+# Create plot showing single simulation of sea level
+seasamp <- ggplot() +
+  geom_sf(data = seapoly, fill = "grey", colour = NA) +
+  geom_sf(data = sitel, fill = NA, colour = "red") +
+  #labs(title =  paste0("Sealevel at ", sealevel, " m above present")) +
+  coord_sf(xlim = c(st_bbox(location_bbox)[1], st_bbox(location_bbox)[3]),
+           ylim = c(st_bbox(location_bbox)[2], st_bbox(location_bbox)[4]),
+           expand = FALSE) +
+  ggsn::scalebar(data = sitel, dist = 150, dist_unit = "m",
+                 transform = FALSE, st.size = 3, height = 0.3,
+                 border.size = 0.1, st.dist = 0.5,
+                 anchor = c(x = anc[2] - 500, y = anc[1] + 800)) +
+  # Uncomment to annotate with distance measures
+  # annotate("text", x = st_bbox(location_bbox)[1] + 20,
+  #          y = st_bbox(location_bbox)[4] - 100,
+  #          label = dist_label, hjust = 0) +
+  theme_bw() + theme(axis.title=element_blank(),
+                     axis.text.y = element_blank(),
+                     axis.text.x = element_blank(),
+                     rect = element_rect(),
+                     axis.ticks = element_blank(),
+                     panel.grid.major = element_blank(),
+                     legend.position = "none")
+
+anot <- ggplot() +
+  theme_bw() +
+  annotate("text", x = 0,  y = 0, label = dist_label,
+           hjust = 0.5, size = 7)
+
+simplot <- (datsamp + elevsamp) / 
+  (seasamp + anot) + plot_annotation(tag_levels = 'A')
+
+ggsave(file = here("analysis/figures/example_sim_map.png"), plot = simplot, 
+       width = 300, height = 76 + (76 * 2),
+       units = "mm")
+
+```
+
+```{r example-2, echo = FALSE, fig.cap = "Example of a single simulation run on the site Hegna vest 1.
+A) The simulation starts by drawing a single year from the posterior density esimate.
+B) This then corresponds to an elevation range on the interpolated displacement curve.
+A single elevation is drawn uniformly from this range using 5cm intervals.
+C) The sea level is then adjusted on the DTM to this elevation and the various distance mesaures are found.
+D) The numerical result of the simulation run.", out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/example_sim_map.png"))
+```
+
+```{r example-site-cont3}
+# Load results from running 1000 simulations
+load(here("analysis/data/derived_data/hegnavest1.RData"))
+
+dat1 <- datsamp <- ggplot(posteriorprobs) +
+  ggridges::geom_ridgeline(aes(x = dates, y = name,
+                               height = probabilities * 50),
+                           fill = "lightgrey") +
+  labs(x = "")  +
+  #scale_y_discrete(expand = c(0.01, 0)) +
+  theme_bw() + theme(
+    legend.position = "none",
+    axis.title.y = element_blank(),
+    axis.ticks.y = element_blank()
+  )
+
+# Map of simulated sea-levels, date group 1
+simsea_plot1 <- shore_plot(output[[1]]$simsea, sitel, 150, date_groups,
+                      s_tdist = 0.5, s_xpos = 1300, s_ypos = 800,
+                      s_bheight = 0.3, s_tsize = 3)
+
+# Boxplot of distances from site to sea, date group 1
+simdist_plot1 <- distance_plot(output[[1]]$results)
+
+# Retrieve date group 2
+probsdat2 <- filter(datedat, group == 2 & class == "bposterior")
+
+dat2 <- ggplot(probsdat2) +
+  ggridges::geom_ridgeline(aes(x = dates, y = name,
+                                   height = probabilities * 50,
+                                   fill = as.factor(group),
+                                   alpha = class,
+                                   colour = as.factor(rcarb_cor))) +
+  labs(x = "cal BCE")  +
+  #scale_y_discrete(expand = c(0.01, 0)) +
+  theme_bw() + theme(
+    legend.position = "none",
+    axis.title.y = element_blank(),
+    axis.ticks.y = element_blank()
+  ) +
+      scale_colour_manual(breaks = c("f", "t", NA),
+                         values = c("red", "black", "yellow")) +
+      scale_alpha_manual(values = c("aprior" = 0.05,
+                                    "bposterior" = 0.4,
+                                    "sum" = 1)) +
+  scale_fill_manual(
+      values = as.character(ggthemes_data$colorblind[2, 2]))
+
+# Plot of date group 2
+
+# Map of simulated sea-levels, date group 2
+simsea_plot2 <- shore_plot(output[[2]]$simsea, sitel, 150, date_groups,
+                      s_tdist = 0.5, s_xpos = 1300, s_ypos = 800,
+                      s_bheight = 0.3, s_tsize = 3)
+
+# Boxplot of distances from site to sea
+simdist_plot2 <- distance_plot(output[[2]]$results)
+
+simresultplot <- (dat1 + simsea_plot1 + simdist_plot1) /
+(dat2 + simsea_plot2 + simdist_plot2)
+
+ggsave(file = here("analysis/figures/example_sim_result.png"),
+       plot = simresultplot,  width = 250, height = 76 + (76 * 2),
+       units = "mm")
+
+```
+
+```{r example-3, echo = FALSE, fig.cap = "The result of 1000 simulation runs for each of the two grops of dates on the site Hegna vest 1.
+The first column of plots shows the radiocarbon probability density estimate from where dates were drawn during simulation.
+The second column displays the result of simulating the raised sea-level 1000 times.
+The darker the colour, the more times the sea-level was simulated to that location.
+The third column shows violin plots of the different distance measures across all simulations.", out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/example_sim_result.png"))
+```
+
+```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites.
+Each data point is plotted with some transparency, meaning that the darker the colour, the more often those values occured.
+The first row shows the result of including all dates the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites.
+The second row shows the result of excluding these.
+The table under each plot lists some corresponding measures for central tendency and dispersion.", out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/results.png"))
 ```

### 2021-11-30 analysis/supplementary-materials/supplementary.Rmd
Supplementary material. Created figures for text.

@@
 output: pdf_document
+bibliography: references.bib
 ---
 
-This document includes some notes and shows the simulation results for each individual site analysed in the study.
-The first plot on the first row for each site shows all calibrated radiocarbon dates from the Stone Age associated with the site (see analysis/data/raw_data/radiocarbon.csv for all dates, including those falling outside of the Stone Age).
-The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase, and each sum indicates that the dates for a phase have been modelled using the Boundary function and then summed using OxCal.
+```{r,  echo = FALSE, results = 'hide'}
+library(here)
+
+```
+
+This document shows the simulation results for each individual site analysed in the study with some additional notes pertaining to edits to the DTM, treatment of ^14^C-dates and other relevant information.
+The sites are presented in alphabetical order.
+
+The first figure shows the result of a test run for the site Hovland 5 where the simulations where run 5000 times to identify when the means for the different distance measures converged.
+This was used to inform the use of 1000 simulation runs for the subsequent analysis of the other sites, as the means appear to have stabilised sufficiently by this point.
+Hovland 5 was chosen because the site only has a single ^14^C-date, giving a wide uncertainty range, and because it is located in an area of the landscape where the landscape results in quite large variability on the distance measures.
+
+```{r sample_size,  echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/sample_size.png"))
+
+```
+
+Below follows the presentation of the individual sites.
+The first plot on the first row for each site shows all calibrated radiocarbon dates from the Stone Age associated with the site (see analysis/data/raw_data/radiocarbon.csv for all dates, including those falling outside of the period).
+The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase, and each sum indicates that multiple dates for a phase have been modelled using the Boundary function and then summed using OxCal.
 Each phase is assumed to be independent of other phases.
-A red outline indicates that the date(s) were not seen as related to the artefact inventory in the original report (see also main text).
+A red outline indicates that the date(s) were not seen as related to the artefact inventory in the original report.
 The second plot on the first row displays the site location in the present day landscape on the edited DTM (i.e.
@@
 
-```{r anvik, include=FALSE}
+```{r adalvestre1,  echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/adalvestre1.png"))
 
@@
 
-Anvik is situated by the highway and a smaller road.
-Neither appear to significantly impact the simulation results.
+Adal vestre 1 is situated close to a field, but this does not appear to have resulted in any relevant changes to the terrain.
+
+```{r dalvestre2,  echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/adalvestre2.png"))
+
+```
+
+Adal vestre 2 as a similar location in the present day landscape to that of Adal vestre 1.
+
+```{r anvik,  echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/anvik.png"))
+
+```
+
+Anvik [@eymundsson2014a] is situated by the highway and a smaller road.
+Neither appear to impact the simulation results.
 The dated fireplace is not seen as related to lithic material, which is believed to hail from the latest part of the Early Mesolithc (c.
@@
 
-```{r gunnarsrod4, include=FALSE}
+```{r bakke, echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/bakke.png"))
+
+```
+
+The radiocarbon date is seen as unrelated to the site inventory, which is distinctly Early Mesolithic in character [@nyland2012b].
+
+```{r dordal, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/dordal.png"))
+
+```
+
+The ^14^C-dates from two fireplaces at Dørdal are not related to the lithic inventory, where there are no typological indicators pointing the Late Mesolithic.
+The report suggest that the results might reflect a contamination through a natural process of some sort [@solheim2017].
+
+```{r dybdalshei1, echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/dybdalshei1.png"))
+
+```
+
+Dybdalshei 1 was subject to a very limited excavation [@granum2018], and the lack of data might therefore lead one to question whether or not the site actually represent a settlement area.
+Furthermore, what little lithic material was retrieved is too generic to offer any typological support to the radiocarbon dates of the dated cooking pits.
+Nonetheless, the site and the ^14^C-dates are still treated the same way as the other sites here.
+This is based on the spatial relation between the lithic artefacts and the features, which offers some additional support to the fact that the cooking pits do in fact represent remnants from visits in the Mesolithic.
+
+```{r dybdalshei2, echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/dybdalshei2.png"))
+
+```
+
+The situation on Dybdalshei 2 is similar to that on Dybdalshei 1 [@granum2018], but with only a single date from a single feature and no lithics, the site is not included in the final results.
+
+```{r gunnarsrod4, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/gunnarsrod4.png"))
 
@@
 Gunnarsrød 4 situated right by the railway (see also Gunnarsrød 5 below).
-The date was seen as possibly the result of old-wood effect, a notion that is in line with the simulation results here, as this date would have the site located beneath the sea-level.
+The date was seen as possibly the result of old-wood effect, a notion that is in line with the simulation results here, as this date would have the site located beneath the sea-level [@reitan2014k].
 
-```{r gunnarsrod5, include=FALSE}
+```{r gunnarsrod5, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/gunnarsrod5.png"))
 
@@
 Gunnarsrød 5 is today situated at the exit of a railway tunnel.
-The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated.
-While the interpolation appears to give reasonable results for the rest of the flatter marsh area where the railway runs today, the site itself would thus have been situated slightly higher, likely leading to the negative values on the distance measures.
-As the simulation results compared to the site map in the report [@reitan2014d], in combination indicate that the site would not have been situated any significant distance from the shoreline, the values have been set to zero for the final aggregative analysis.
+The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated [@reitan2014d],.
+While the editing and interpolation of the DTM appears to give reasonable results for the rest of the flatter marsh area where the railway runs today, the site itself would thus have been situated slightly higher, likely leading to the negative values on the distance measures.
+In combination the simulation results and the site map in the report indicate that the site would not have been situated any significant distance from the shoreline.
+The values have been set to zero for the final aggregative analysis.
 A note can also be made on the date that is seen as unrelated to the artefact inventory.
 This was done on charcoal from a pit with ceramics which had a far younger date.
-The date was therefore interpreted as the result of contamination of some kind [@reitan2014d].
+The date was therefore interpreted as the result of contamination of some kind.
+
+```{r gunnarsrod6a, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/gunnarsrod6a.png"))
+
+```
+
+The ^14^C-date is not related to the lithic inventory on the site [@carrasco2014].
+
+```{r gunnarsrod6b, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/gunnarsrod6b.png"))
+
+```
+
+```{r gunnarsrod7, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/gunnarsrod7.png"))
+
+```
+
+```{r gunnarsrod10, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/gunnarsrod10.png"))
+
+```
+
+```{r hegnavest1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/hegnavest1.png"))
+```
+
+```{r hegnavest2, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/hegnavest2.png"))
+```
+
+```{r hegnavest3, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/hegnavest3.png"))
+```
+
+```{r hesthagc2, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/hesthagc2.png"))
+```
+
+```{r hesthagc4, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/hesthagc4.png"))
+```
+
+```{r hovland1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/hovland1.png"))
+```
+
+```{r hovland3, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/hovland3.png"))
+```
+
+```{r hovland4, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/hovland4.png"))
+```
+
+```{r hovland5, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/hovland5.png"))
+```
+
+```{r hydal4, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/hydal4.png"))
+```
+
+```{r krogenesd1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/krogenesd1.png"))
+```
+
+```{r krogenesd2, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/krogenesd2.png"))
+```
+
+```{r kvastada1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/kvastada1.png"))
+```
+
+```{r kvastada2, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/kvastada2.png"))
+```
+
+```{r kvastada4, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/kvastada4.png"))
+```
+
+```{r kvastada9, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/kvastada9.png"))
+```
+
+```{r langangenv1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/langangenv1.png"))
+```
+
+```{r langangenv3, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/langangenv3.png"))
+```
+
+```{r langangenv5, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/langangenv5.png"))
+```
+
+```{r langangenv6, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/langangenv6.png"))
+```
+
+```{r langangenv7, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/langangenv7.png"))
+```
+
+```{r langemyr, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/langemyr.png"))
+```
+
+```{r lovas1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/lovas1.png"))
+```
+
+```{r lovas2, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/lovas2.png"))
+```
+
+```{r lovas3, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/lovas3.png"))
+```
+
+```{r lunaveien, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/lunaveien.png"))
+```
+
+Lunaveien is a grave cairn site, and the report states that it is unlikely that any of the finds could relate to a settlement predating the construction of the cairn [@reitan2010].
+The results have therefore been excluded from the aggregative analysis.
+
+```{r nauena, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/nauena.png"))
+```
+
+A section of the highway located a few hundred meters to the south was pre-emptively edited in the DTM. This does not appear to be relevant for the simulated sea-level.
+
+```{r nauenc, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/nauenc.png"))
+```
+
+```{r nordby1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/nordby1.png"))
+```
+
+```{r nordby52, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/nordby52.png"))
+```
+
+```{r pauler1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/pauler1.png"))
+```
+
+```{r pauler2, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/pauler2.png"))
+```
+
+```{r pjonkerodr1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/pjonkerodr1.png"))
+```
+
+```{r prestemoen1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/prestemoen1.png"))
+```
+
+```{r ragnhildrod, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/ragnhildrod.png"))
+```
+
+```{r rodbol54, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/rodbol54.png"))
+```
+
+```{r rognlien, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/rognlien.png"))
+```
+
+```{r sageneb1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/sageneb1.png"))
+```
+
+```{r sageneb2, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/sageneb2.png"))
+```
+
+```{r solum3, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/solum3.png"))
+```
+
+```{r stokkep1, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/stokkep1.png"))
+```
+
+```{r stokkep5, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/stokkep5.png"))
+```
+
+```{r stokkep8, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/stokkep8.png"))
+```
+
+```{r torstvet, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/torstvet.png"))
+```
+
+```{r trondalen, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/trondalen.png"))
+```
+
+```{r tverdal, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/tverdal.png"))
+```
+
+```{r vallermyrene1a, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/vallermyrene1a.png"))
+```
+
+```{r vallermyrene1b, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/vallermyrene1b.png"))
+```
+
+```{r vallermyrene2, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/vallermyrene2.png"))
+```
+
+```{r vallermyrene4a, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/vallermyrene4a.png"))
+```
+
+```{r vallermyrene4b, echo = FALSE, out.width = "500px"}
+
+knitr::include_graphics(here("analysis/figures/vallermyrene4b.png"))
+```

### 2021-11-30 analysis/script/04script.R

@@ -473,11 +473,6 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
 
     seapolygons[[i]] <- seapoly
 
-    # Retrieve highest elevation on polygon
-    siteelev <- extract(sitearea, vect(sitel), fun = max)[2]
-    # Find difference to sea level elevation
-    results$vertdist[i] <-  max(siteelev[1:nrow(siteelev),]) - sealevel
-
     # If these are at the same coordinate (i.e. the polygons are overlapping),
     # return all values as 0
     if(st_overlaps(seapoly, sitel, sparse = FALSE)){
@@ -487,6 +482,11 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
 
       # Else...if the entire site polygon is in the sea
     } else if(st_contains(seapoly, sitel, sparse = FALSE)){
+
+      # Retrieve highest elevation on polygon
+      siteelev <- extract(sitearea, vect(sitel), fun = max)[2]
+      # Find difference to sea level elevation
+      results$vertdist[i] <-  max(siteelev[1:nrow(siteelev),]) - sealevel
 
       # Find the nearest points on site- and seapolygons, making the sea a
       # multiline feature as it contains the site
@@ -531,6 +531,12 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
       # located some distance from the sea, perform above steps but without
       # negative values
     } else {
+
+      # Retrieve lowest elevation on polygon
+      siteelev <- extract(sitearea, vect(sitel), fun = min)[2]
+      # Find difference to sea level elevation
+      results$vertdist[i] <-  min(siteelev[1:nrow(siteelev),]) - sealevel
+
       # Find the nearest points on the site- and seapolygons,
       distline <- st_nearest_points(sitel, seapoly)
 

### 2021-12-02 analysis/paper/paper.Rmd
Redid extraction of data from lists of lists of different length.

@@
 load(here("analysis/data/derived_data/02data.RData"))
+load(here("analysis/data/derived_data/06data.RData"))
 source(here("analysis/script/04script.R"), local = knitr::knit_global())
@@
 Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g.
-@berg-hansen2009; @nummedal1923; @bjerck1990].
+@berg-hansen2009; @bjerck1990; @mjærum2018].
 The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
@@
 Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age.
-This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line.
+This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline.
 Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.
 
-Despite the fundamental role of shoreline dating for Norwegian Stone Age archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods.
+Despite its fundamental role for Norwegian Stone Age archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods.
 The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach.
@@
 As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions.
-One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently spatial.
+One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shoreline is inherently spatial.
 Taking inspiration from recent studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g.
-@crema2010; @crema2012], a similar approach is adopted here.
+@crema2012; @crema2010; @lewis2021], a similar approach is adopted here.
 
@@
 Variation in this relative distance follow from a range of effects [e.g.
-@mörner1976; @milne2009].
+@milne2009; @mörner1976].
 Of central importance here is eustasy and glacial istostasy.
-The eustatic sea level is understood as the sea level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015].
-The eustatic sea level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that impact the volume of ocean basins.
+The eustatic sea-level is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015].
+The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that impact the volume of ocean basins.
 Istostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle.
@@
 Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g.
-@stroeven2016], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise .
-In other words, the relative sea level has been dropping throughout prehistory.
+@hughes2016; @stroeven2016], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g.
+@lambeck1998; @svendsen1987].
+In other words, the RSL has been dropping throughout prehistory.
 As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet.
 Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g.
-@svendsen1987; @romundset2015].
+@romundset2015; @svendsen1987].
 These conditions are directly reflected in the archaeological record.
@@
 Comparatively, given a continuous and still ongoing shoreline regression from as high as c.
-220 m above present sea level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @sørensen1979].
+220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @sørensen1979].
 This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.
 
-such as Lista in southern Norway has In the most perhipheral coastal areas the relative sea-level change has transitioned between regression and transgression phases, leading to a situation where
+```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from 18-10 ka (Hughes et al.
+2016).
+The studyarea defined later in the text is given a red outline.", out.width = "400px", fig.align="center"}
 
-Based on the generally accepted premise that most Stone Age sites located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
-This is for example common in survey projects, which are often guided by a both a digital and mental reconstruction of past sea levels.
-Following an excavation, the most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves.
-If this corresponds then it is common to assume a shore bound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this.
+knitr::include_graphics(here("analysis/figures/deglaciation.png"))
 
-Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g.
-@jaksland2014].
+```
+
+Based on the generally accepted premise that most Stone Age sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
+This is for example common in survey projects, which are often guided by a both a digital and mental reconstruction of past sea-levels.
+Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the possible time-span provided by the typological indicators are too wide to decisively verify this.
+It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g.
+@jaksland2014; @nummedal1923].
 This can for example pertain to accessibility.
-If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if it would have been located on an island when the sea level was higher, it would make intuitive sense that the site was used when the ocean reached close to its elevation, as this would have made the site accessible by means of watercraft.
-The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations.
-Although it appears that the arguments for such site locations is sensible, few attempts at a comprehensive evaluation and quantification of this tendency has been done.
+If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if it would have been located on an island when the sea-level was higher, it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft.
+Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are fewer, and are typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore-bound site location is largely based on a mosaic of smaller scale investigations.
 
-One of the larger evaluations of this relationship has been done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area---data that are also included in this study.
-They found a overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5 % of the sites.
-However, in the cases where there was a discrepancy, the original .
-As was exemplified in Solheims study, and which has been convincingly argued in several individual studies, many of the sites where there is a mismatch between the radiocarbon dates and the shoreline dates, the typological indicators in the assemblages are clearly unrelated to the radiocarbon dates.
-Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates, or whether they are entirely erroneous results of contamination can be difficult to evaluate, and will here be entirely dependent on the discretion of the original excavation report.
+One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 ^14^C-dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
+They found a overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
+However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates.
+This is based on typological and technological characteristics of the assemblages.
+Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether they are entirely erroneous results of contamination can be difficult to evaluate.
+Here this will be entirely dependent on, and follow the discretion of the original excavation reports.
 
-While such evaluations can be difficult, it is nonetheless a necessity that will be held strictly separate from the formal equivalence tests in the analysis to follow.
-
-Previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
+Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
 @solheim2020; @åstveit2018].
@@
 @parnell2008].
-Another limitation of this often employed method is that it only takes into account the vertical distance between the sites and the sea level.
+Another limitation of this often employed method is that it only takes into account the vertical distance between the sites and the sea-level.
 While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry.
@@
 Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.
-
-In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions.
-
-Given that the aim of this study is to test whether or not there is a match between dates provided by means of shoreline displacement curves and 14C-dates, it is appropriate to evaluate this relationship by means of statistical equivalence tests.
-This effectively shifts the common direction of the null-hypothesis, giving H0: The date given by means of radiocarbon dating and by shoreline displacement are different.
-Ha: The date given by means of radiocarbon dating and by shoreline displacement are the same.
-While this consitutes the formal dimension of the study, there must necessarily also be a qualitative and more subjectivly driven dimension as well.
-
-cannot necessarily be directly compared between curves as these have been developed at different times, using a wide range of different methods.
-However, any reinterpretation and standardisation of the curves is also hampered by a range of issues.
-The original data or necessary details about the sampling processes are not always available, and a reinterpretation based soley on radiometric dates, without taking into account the considerations and evaluations of the field geologists can be entirely misleading.
-In a rudimentary attempt to account for these differences when compiling the curves, the width of the confidence band for the sea level at a single date across the entire region is given as the largerst uncertainty of any one of the curves at that point in time.
-This averages to an uncertainty of +/- calendar years.
-This uncertainty is uniformly distributed within the confidence band.
 
@@
 To get at the relationship between sites and contemporaneous shoreline, the analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
-While there is displacement data available for other areas of south eastern Norway [e.g.
+While there is displacement data available for other areas of south-eastern Norway [e.g.
 @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Halden county in the north-east, to Arendal in the south-west.
-This area has newly compiled displacement curves for Skoppum in Halden [@romundset2021], Larvik [@sørensen2014; @sørensen2014b], Hanto in Tvedestrand [@romundset2018; @romundset2018b], and Bjørnebu in Arendal [@romundset2018b].
+This area has newly compiled displacement curves for Halden [@romundset2021], Larvik [@sørensen2014; @sørensen2014b], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
 
-The employed shoreline displacement curves are all based on the so-called isolation basin method, which represents an extremely precise method compared to other [@romundset2019].
-This involves extracting cores from a series of basins situated on bedrock in a constrained area of the landscape, and dating the transition from marine to lacustrine .
-Each curve is thus construed from a series of cored basins located at different elevations, each representing a sea-level index point.
-The reason why these confidence intervals are not strictly mathematically derived is that the curves do not only contain uncertainty as related to radiometric dates, which are well understood, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets, the adjustment to a common isobase based on the direction and gradient in rebound intensity, to name but a few [e.g.
-@romundset2010].
-The error bands and their trajectory are consequently quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
+The employed shoreline displacement curves are all based on the so-called isolation basin method [@romundset2011].
+This involves extracting cores from a series of basins located beneath the marine limit situated on bedrock in a constrained area of the landscape, and dating the transition from marine to lacustrine sediments in the cores.
+Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point.
+The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
+The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets, the adjustment to a common isobase based on the direction and gradient in rebound intensity, to name but a few [e.g.
+@romundset2011; @romundset2019].
 For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications (see above).
@@
 The archaeological data compiled for the analysis consists of all excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)).
-These number a total `r nrow(sites)` sites, of which `r length(unique(rcarbon$site_name))` sites are associated with the total of `r nrow(rcarbon)` radiocarbon dates.
-Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability.
+These number a total `r nrow(sites)` sites.
+Of these `r length(unique(rcarbon$site_name))` sites are associated with the total of `r nrow(rcarbon)` radiocarbon dates.
+In turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability.
 These sites form the basis for the analysis.
@@
 
-```{r map-iso,  fig.cap = "A) Distribution of the 70 analysed sites relative to the isobases of the displacement curves, B) Displacement curves.
-Note the higher  steepness of the curves that are situated towards the north-east.", fig.height = 13}
+[Table with sites]
+
+The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2019].
+After evaluating the options I decided to use 10m resolution DTM rather than the higher-resolution 1m version.
+In addition to resulting is less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by.
+The 10m DTM is a down-sampled version of the 1m resolution DTM, which is based on aerial laser scanning using a minimum of two and up to five points per m^2^ [@mapping2019].
+
+```{r map-iso,  fig.cap = "A) Distribution of the 68 analysed sites relative to the isobases of the displacement curves.
+The isobases have a direction of 327 $^{\\circ}$ (Romundset el al.
+2018), B) Displacement curves.
+Note the increasing steepness of the curves towards the north-east.", fig.height = 13}
 bboxsites <- st_bbox(sites_sa)
@@
   geom_line(aes(y = lowerelev)) +
-  ylab("Meters above present sea level") +
+  ylab("Meters above present sea-level") +
   xlab("cal BCE/CE") +
@@
 
-The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shore-lines, each associated with their own range of temporal uncertainty.
+The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shorelines, each associated with their own range of temporal uncertainty.
 The first task was therefore to ascribe likely date ranges and associated uncertainty to the two phenomena.
 
-Due to the dsfg there is a gradient in the rate of isostatic rebound, with increasing rebound rates towards the north-east.
-To take account of this, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase.
-This was done by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves.
+To take account of the gradient in the isostatic rebound, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase.
+This was done by for each site first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves.
 Each value was then divided by the distance between the isobases.
 To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter.
-If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east, the values were instead added to the original curve.
-The result of this process is shown for an example site in .
+These values were then subtracted from the dsiplacement curve located to the south-west of the site.
+The result of this process is shown for an example site in Figure \@ref(fig:example-1).
 
-For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v.
-[@bronkramsey2009] through the oxAAR package for R.
-All dates not older than 1700 BCE with 99.7 % probability were then excluded from further analysis.
+For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
 Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase.
 In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal.
-Multiple phases at a single site were treated as independent of each other (Figure x).
+Multiple phases at a single site were treated as independent of each other.
 
 As the excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances.
-In addition to employing 10 m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster.
+In addition to employing 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster.
 This involved manually defining the extent of problem areas such as railways, highways, quarries and the like.
 The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain.
-This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls from GRASS GIS in R through the package rgrass7 [@bivand2021].
+This was done using regularised spline interpolation with tension [e.g.
+@conolly2020], using the default settings of r.fillnulls from GRASS GIS in R through the package rgrass7 [@bivand2021].
 In addition to code and original spatial data being available in the research compendium for this paper, it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.
@@
 Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulations were performed.
-A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site.
-This year then has a corresponding likely elevation range for the contemporaneous shoreline.
-An elevation value was then drawn uniformly from this range, using intervals of 5 cm.
+A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \@ref(fig:example-2)).
+This year then has a corresponding likely elevation range for the contemporaneous shoreline from which an elevation value was drawn uniformly, using intervals of 5cm.
 The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing.
 Polygons were then created from the resulting areas with missing values.
-The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea level from the lowest elevation of the site polygon.
+The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon.
 The topographic distance between site and sea was also found by measuring the shortest distance while taking into account the slope of the terrain on the DTM.
-This was done using the topoDistance package for R.
-The topographic distance was measured between site polygon and the point on the shoreline found to be closest measured horisontally.
-The shortest topographic path was found using the Queen's case neighbourhood of eight cells.
+This was done using the topoDistance package for R [@wang2019].
+The topographic distance was measured between site polygon and the horisontally closest point on the shoreline.
+The shortest topographic path was found using the Queen's case neighbourhood of eight cells [@conolly2006].
 In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
 In the case that the sea-polygon completely contain the site, the horisontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon.
-While it is safe to assume that an archaeological site was not occupied when it was located beneath sea level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
-Negative values where therefore retained unless otherwise stated.
-
-Some sources of likely variation and uncertainty have not been considered here.
-First of all the DTM has only been corrected for major modern disturbances.
-This means that erosion, although likely not that prevalent, have not been taken into account.
-Secondly, the DTM has a vertical error of.
-Thirdly, the tidal range has not been considered.
-This is around 0.5 m within the study area today, and there is no reason to think this was substantially different in the past.
-Fourthly, the exposure of the sites to wave action is also likely to have been of concern, and will presumably have implications for how close to the shoreline people settled.
+While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
+Negative values where therefore retained.
 
 This process was repeated 1000 times for each phase for each site.
-The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations on the example site Hovland 5, which has a uncertain date range and is located in area of quite complex topography.
-At 1000 simulation runs the analysis of each site ranged from hours, depending on the distance to the simulated shorelines and consequently the necessary size of the window of analysis.
-
-The distances were then measured.
-
-The occupation of a site and shifting sea-levels.
-The synchroneity of these processes is tested by the comparison of singular events represented by the death of the dated organic materials.
-Additionally, defining what shore bound settlement represents can be problematic.
-In the literature this is often qualified with statements along the lines of "the sites were located on or close to the contemporenous shore-line".
-What constitutes close is of course highly variable, complicated further by the fact that vertical changes in sea level can results in widely different effects on the horisontal distance between a given point and the shore-line.
-Other elements of interest is the tidal range, and the exposure of the location in question, which can also have an impact on the distance of the site to the mean ocean surface.
-
-Here the approach was only used to inform the question of the distance between sites and shorelines, but this method can also be extended to a wide range of use-cases where one needs to visualise as well as quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore line, and the uncertainty inherent in this reconstruction.
+The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5.
+Hovland 5 was chosen for this evuluation as it has a fairly uncertain date range, and is located in area of quite complex topography.
+At 1000 simulation runs the analysis of each site ranged from hours to a couple of days, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.
 
@@
                                  "Hegna vest 1" = "hotpink2")) +
-  ylab("Meters above present sea level") +
+  ylab("Meters above present sea-level") +
   xlab("Cal BCE/CE") +
@@
 C) The location of the site within the study area relative to isobases of the employed displacement curves.
-D) Displacement curve interpolated to the site location based on the distance between the site and the isobases of the two closest displacement curves.", out.width = "500px"}
+D) Displacement curve interpolated to the site location based on the distance between the site and the isobases of the two closest displacement curves.", out.width = "500px", fig.align="center"}
 knitr::include_graphics(here("analysis/figures/example_map.png"))
@@
     legend.position = "none",
-    panel.border = element_blank(),
     axis.line = element_line("black"),
@@
 
-# Draw sealevel uniformly from elevation range, intervals of 5 cm
+# Draw sea-level uniformly from elevation range, intervals of 5 cm
 sealevel <- sample(seq(lowers, uppers, 0.05), 1)
@@
   geom_line(aes(y = lowerelev), colour = "hotpink2") +
-  ylab("Meters above present sea level") +
+  ylab("Meters above present sea-level") +
   xlab("cal BCE/CE") +
@@
                                            yend = -Inf), col = "black",
-               linetype = "solid") +
+                size = 0.25, linetype = "solid") +
   geom_segment(aes(x = -Inf, xend = sdate, y = uppers,
@@
                linetype = "dashed") +
-  annotate("text", x = -8700, y = uppers + 2, label = as.character(uppers)) +
-  annotate("text", x = -8700, y = lowers - 2, label = as.character(lowers)) +
-  scale_x_continuous(breaks = sort(c(-9000, sdate, -6000, -3000, 0))) +
+  annotate("text", x = -8700, y = uppers + 3.5, label = as.character(uppers)) +
+  annotate("text", x = -8700, y = lowers - 3.5, label = as.character(lowers)) +
+  scale_x_continuous(breaks = sort(c(sdate, -6000, -3000, 0))) +
   scale_y_continuous(breaks = sort(c(0, sealevel, 100)))
 
-# Adjust sea level on DTM
+# Adjust sea-level on DTM
 rclmat <- matrix(c(sealevel, Inf, NA, 0, sealevel, 1),
@@
 
-# Create plot showing single simulation of sea level
+# Create plot showing single simulation of sea-level
 seasamp <- ggplot() +
@@
   geom_sf(data = sitel, fill = NA, colour = "red") +
-  #labs(title =  paste0("Sealevel at ", sealevel, " m above present")) +
+  #labs(title =  paste0("Sea-level at ", sealevel, " m above present")) +
   coord_sf(xlim = c(st_bbox(location_bbox)[1], st_bbox(location_bbox)[3]),
@@
 anot <- ggplot() +
-  theme_bw() +
+  theme_void() +
   annotate("text", x = 0,  y = 0, label = dist_label,
@@
 A single elevation is drawn uniformly from this range using 5cm intervals.
-C) The sea level is then adjusted on the DTM to this elevation and the various distance mesaures are found.
-D) The numerical result of the simulation run.", out.width = "500px"}
+C) The sea-level is then adjusted on the DTM to this elevation and the various distance mesaures are found.
+D) The numerical result of the simulation run.", out.width = "500px", fig.align="center"}
 knitr::include_graphics(here("analysis/figures/example_sim_map.png"))
@@
 
+# Results
+
+The result of running the simulations across all sites are given in Figure \@ref(fig:results).
+Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, most of the sites have been situated close to the shoreline when they were in use.
+When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline, as could be expected.
+To what degree these do in fact represent contamination is discussed further below.
+However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
+It is interesting to note the distance for some of the earliest sites, which appears somewhat high.
+This, however, can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level.
+Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.
+From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horisontally and topograpchically close, they appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
+This corresponds well with the literature, where the introduction of the Neolithic proper is seen as occurring around this time [@prescott2020; @solheim2021].
+With the transition from a predominantly hunter-gatherer lifeway to early agro-pastrolism, settlements in coastal areas are to be increasingly withdrawn from the shoreline.
+Given this, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE.
+
 ```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites.
@@
 The second row shows the result of excluding these.
-The table under each plot lists some corresponding measures for central tendency and dispersion.", out.width = "500px"}
+The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center"}
 knitr::include_graphics(here("analysis/figures/results.png"))
+
 ```
+
+```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using only dates older than 2500 BCE that correspond to the site inventory.", out.width = "525px", fig.align="center"}
+knitr::include_graphics(here("analysis/figures/results2.png"))
+```
+
+# Discussion
+
+Some sources of likely variation and uncertainty have not been considered here.
+First of all the DTM has only been corrected for major modern disturbances.
+This means that erosion, although likely not that prevalent, has not been taken into account.
+Secondly, the DTM has a vertical error...
+Thirdly, the tidal range has not been considered...
+Fourthly, the exposure of the sites to wave action is also likely to have been of concern, and will presumably have implications for how close to the shoreline people settled.
+And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site
+
+# Conclusion
+
+Here the approach was used to inform the question of the distance between sites and shorelines, but this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore line, and the uncertainty inherent in this reconstruction.
 

### 2021-12-03 analysis/paper/paper.Rmd
First draft

@@
 
-```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from 18-10 ka (Hughes et al.
+```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from 19-10 ka (data from Hughes et al.
 2016).
@@
 While there is displacement data available for other areas of south-eastern Norway [e.g.
-@hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Halden county in the north-east, to Arendal in the south-west.
-This area has newly compiled displacement curves for Halden [@romundset2021], Larvik [@sørensen2014; @sørensen2014b], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
+@hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
+This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
 
@@
                                    "Tvedestrand" = "blue",
-                                   "Halden" = 'red')) +
+                                   "Horten" = 'red')) +
   theme_bw() + theme(axis.title=element_blank(),
@@
                                    "Tvedestrand" = "blue",
-                                   "Halden" = 'red')) +
+                                   "Horten" = 'red')) +
   theme_bw() +

### 2021-12-03 analysis/supplementary-materials/supplementary.Rmd
First draft

@@
 library(here)
+library(tidyverse)
 
+# Read in radiocarbon dates to be plotted alongside analysis for
+# each site.
+rcarbon <- read.csv(here('analysis/data/raw_data/radiocarbon.csv'))
 ```
@@
 ```{r adalvestre1,  echo = FALSE, out.width = "500px"}
+
+c14 <- filter(rcarbon, site_name  == "Adal vestre 1") %>% 
+  select(lab_code, c14_bp, error, material, context)
+
 knitr::include_graphics(here("analysis/figures/adalvestre1.png"))
-
 ```

### 2021-12-03 analysis/script/04script.R

@@ -786,7 +786,7 @@ overview_plot <- function(background_map, sitelimit, sites, isobases,
     scale_colour_manual(values = c("Arendal" = "black",
                                    "Larvik" = "darkgreen",
                                    "Tvedestrand" = "blue",
-                                   "Halden" = 'red')) +
+                                   "Horten" = 'red')) +
     theme_bw() + theme(axis.title=element_blank(),
                        axis.text.y = element_blank(),
                        axis.text.x = element_blank(),

### 2021-12-04 analysis/paper/paper.Rmd
Writing through supplementary

@@
 These conditions are directly reflected in the archaeological record.
-In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating palimpsests of settlement events that are difficult to disentangle [@reitan2009].
+In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [@reitan2009].
 Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008].
@@
 
-```{r example-1, echo = FALSE, fig.cap = "Example site Hegna vest 1.
+```{r example-1, echo = FALSE, fig.cap = "Example site Hegna vest 1 (Fossum 2017).
 A) Location of the site in the present day landscape.

### 2021-12-04 analysis/supplementary-materials/supplementary.Rmd
Writing through supplementary

--- 2021-11-15 analysis/supplementary-materials/supplementary.Rmd
@@
 output: pdf_document
+bibliography: references.bib
 ---
 
-This document includes some notes and shows the simulation results for each individual site analysed in the study.
-The first plot on the first row for each site shows all calibrated radiocarbon dates from the Stone Age associated with the site (see analysis/data/raw_data/radiocarbon.csv for all dates, including those falling outside of the Stone Age).
-The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase, and each sum indicates that the dates for a phase have been modelled using the Boundary function and then summed using OxCal.
+```{r,  echo = FALSE, results = 'hide'}
+library(here)
+library(tidyverse)
+library(knitr)
+library(kableExtra)
+
+# Read in radiocarbon dates to be plotted alongside analysis for
+# each site.
+rcarbon <- read.csv(here('analysis/data/raw_data/radiocarbon.csv'))
+```
+
+This document shows the simulation results for each individual site analysed in the study with some additional notes pertaining to edits to the DTM, treatment of ^14^C-dates and other relevant information.
+The sites are presented in alphabetical order.
+
+The first figure shows the result of a test run for the site Hovland 5 where the simulations where run 5000 times to identify when the means for the different distance measures converged.
+This was used to inform the use of 1000 simulation runs for the subsequent analysis of the other sites, as the means appear to have stabilised sufficiently by this point.
+Hovland 5 was chosen because the site only has a single ^14^C-date, giving a wide uncertainty range, and because it is located in an area of the landscape where the landscape results in quite large variability on the distance measures.
+
+```{r sample_size,  echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/sample_size.png"))
+
+```
+
+Below follows the presentation of the individual sites.
+The first plot on the first row for each site shows all calibrated radiocarbon dates from the Stone Age associated with the site (see analysis/data/raw_data/radiocarbon.csv for all dates, including those falling outside of the period).
+The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase, and each sum indicates that multiple dates for a phase have been modelled using the Boundary function and then summed using OxCal.
 Each phase is assumed to be independent of other phases.
-A red outline indicates that the date(s) were not seen as related to the artefact inventory in the original report (see also main text).
+A red outline indicates that the date(s) were not seen as related to the artefact inventory in the original report.
 The second plot on the first row displays the site location in the present day landscape on the edited DTM (i.e.
@@
 
-```{r anvik, include=FALSE}
+```{r adalvestre1,  echo = FALSE, out.width = "500px"}
+
+include_graphics(here("analysis/figures/adalvestre1.png"))
+
+filter(rcarbon, site_name  == "Adal vestre 1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Adal vestre 1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+Adal vestre 1 is situated close to a field, but this does not appear to have resulted in any relevant changes to the terrain.
+
+```{r dalvestre2,  echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/adalvestre2.png"))
+
+filter(rcarbon, site_name  == "Adal vestre 2") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Adal vestre 2", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
 
@@
 
-Anvik is situated by the highway and a smaller road.
-Neither appear to significantly impact the simulation results.
+Adal vestre 2 has a similar location in the present day landscape to that of Adal vestre 1 (above).
+
+```{r anvik,  echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/anvik.png"))
+
+filter(rcarbon, site_name  == "Anvik") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Anvik", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+
+```
+
+Anvik [@eymundsson2014a] is situated by the highway and a smaller road.
+Neither appear to impact the simulation results.
 The dated fireplace is not seen as related to lithic material, which is believed to hail from the latest part of the Early Mesolithc (c.
@@
 
-```{r gunnarsrod4, include=FALSE}
+```{r bakke, echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/bakke.png"))
+
+filter(rcarbon, site_name  == "Bakke") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Bakke", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+The radiocarbon date to the Stone Age is not related to the site inventory, which is distinctly Early Mesolithic in character [@nyland2012b].
+
+```{r dordal, echo = FALSE, out.width = "500px"}
+
+include_graphics(here("analysis/figures/dordal.png"))
+
+filter(rcarbon, site_name  == "Dørdal") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Dørdal", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
 
@@
 
-Gunnarsrød 4 situated right by the railway (see also Gunnarsrød 5 below).
-The date was seen as possibly the result of old-wood effect, a notion that is in line with the simulation results here, as this date would have the site located beneath the sea-level.
+The ^14^C-dates from two fireplaces at Dørdal are not related to the lithic inventory, where there are no typological indicators pointing the Late Mesolithic.
+The report suggest that the results might reflect a contamination through a natural process of some sort [@solheim2017].
 
-```{r gunnarsrod5, include=FALSE}
+```{r dybdalshei1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/dybdalshei1.png"))
+
+filter(rcarbon, site_name  == "Dybdalshei 1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Dybdalshei 1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
 
@@
 
-Gunnarsrød 5 is today situated at the exit of a railway tunnel.
-The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated.
-While the interpolation appears to give reasonable results for the rest of the flatter marsh area where the railway runs today, the site itself would thus have been situated slightly higher, likely leading to the negative values on the distance measures.
-As the simulation results compared to the site map in the report [@reitan2014d], in combination indicate that the site would not have been situated any significant distance from the shoreline, the values have been set to zero for the final aggregative analysis.
+Dybdalshei 1 was subject to a very limited excavation [@granum2018], and the lack of data might therefore lead one to question whether or not the site actually represent a settlement area.
+Furthermore, what little lithic material was retrieved is too generic to offer any typological support to the radiocarbon dates of the dated cooking pits.
+Nonetheless, the site and the ^14^C-dates are still treated the same way as the other sites.
+This is based on the spatial relation between the lithic artefacts and the features, which offers some additional support to the fact that the cooking pits do in fact represent remnants from visits in the Mesolithic.
+
+```{r dybdalshei2, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/dybdalshei2.png"))
+
+filter(rcarbon, site_name  == "Dybdalshei 2") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Dybdalshei 2", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+
+```
+
+The situation on Dybdalshei 2 is similar to that on Dybdalshei 1 [above, @granum2018], but with only a single date from a single feature and no lithics, the site is not included in the final results.
+
+```{r gunnarsrod4, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/gunnarsrod4.png"))
+
+filter(rcarbon, site_name  == "Gunnarsrød 4") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Gunnarsrød 4", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+
+```
+
+Gunnarsrød 4 [@reitan2014k], and Gunnarsrød 5 (below), were situated close to where the railway runs today, and therefore required editing of the DTM.
+Based on maps from the report, this appears to have given adequate, although not perfect results [@reitan2014k, fig.14.7.1].
+The date from the bottom of the cultural layer was seen as possibly impacted by old-wood effect, a notion that is in line with the simulation results, as this date would have the site located beneath the sea-level [@reitan2014k].
+
+```{r gunnarsrod5, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/gunnarsrod5.png"))
+
+filter(rcarbon, site_name  == "Gunnarsrød 5") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Gunnarsrød 5", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+
+```
+
+The location of Gunnarsrød 5 is today situated at the exit of a railway tunnel.
+The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated [@reitan2014d].
+While the editing and interpolation of the DTM appears to give reasonable results for the rest of the flatter marsh area where the railway runs today (see also Gunnarsrød 4, above), the site itself would thus have been situated slightly higher.
+This has likely led to the negative values on the distance measures.
+In combination, the simulation results and the site map in the report indicate that the site would not have been situated any significant distance from the shoreline.
+The values have therefore been set to zero for the final aggregative analysis.
 A note can also be made on the date that is seen as unrelated to the artefact inventory.
-This was done on charcoal from a pit with ceramics which had a far younger date.
-The date was therefore interpreted as the result of contamination of some kind [@reitan2014d].
+This is from charcoal from a pit with ceramics which had a far younger date.
+The date was therefore interpreted as the result of contamination of some kind.
+
+```{r gunnarsrod6a, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/gunnarsrod6a.png"))
+
+filter(rcarbon, site_name  == "Gunnarsrød 6a") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Gunnarsrød 6a", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+
+```
+
+The ^14^C-date to the Stone Age from Gunnarsrød 6a is not related to the lithic inventory on the site, which is clearly Mesolithic [@carrasco2014].
+
+```{r gunnarsrod6b, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/gunnarsrod6b.png"))
+
+filter(rcarbon, site_name  == "Gunnarsrød 6b") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Gunnarsrød 6b", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+
+```
+
+As with Gunnarsrød 6a above, the ^14^C-date to the Stone Age from Gunnarsrød 6b are not related to the Mesolithic inventory on the site [@carrasco2014].
+
+```{r gunnarsrod7, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/gunnarsrod7.png"))
+
+filter(rcarbon, site_name  == "Gunnarsrød 7") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Gunnarsrød 7", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+
+```
+
+Gunnarsrød 7 is situated on a slope, stretching from 55 masl up to 59 [@fossum2014b]. The lithic inventory is of a Middle Mesolithic character and does not match the ^14^C-dates.
+
+```{r gunnarsrod10, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/gunnarsrod10.png"))
+
+filter(rcarbon, site_name  == "Gunnarsrød 10") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Gunnarsrød 10", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+
+```
+
+Based on the location of Gunnarsrød 10 on a steep ledge, and the site inventory, the ^14^C-date is not believed to be related to the main occupation of the site [@reitan2014l].
+
+```{r hegnavest1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/hegnavest1.png"))
+
+filter(rcarbon, site_name  == "Hegna vest 1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Hegna vest 1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+Example site from the main text.
+Hegna vest 1 has been visited multiple times throughout prehistory.
+The lithic inventory is distinctly Middle Mesolithic and Neolithic in character, with some finds that might be even younger.
+This is also reflected in the pottery finds, which are deemed to be from the Bronze Age--Early Iron Age.
+The ^14^C-date to the Late Mesolithic is therefore considered unrelated to the main occupation of the site, as it cannot be related to any elements of the site inventory [@fossum2017a].
+The highway runs right by Hegna vest 1 and has been removed.
+This seem successful in that it doesn't appear to impact the simulated sea-levels.
+
+```{r hegnavest2, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/hegnavest2.png"))
+
+filter(rcarbon, site_name  == "Hegna vest 2") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Hegna vest 2", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+As with Hegna vest 1 (above), Hegna vest 2 has been used throughout prehistory.
+This includes the Middle Mesolithic, the Neolithc, the Bronze Age and the Early Iron Age.
+The ^14^C-dates are believed to be related to the site activities [@fossum2017].
+Hegna vest 2 is located where the highway runs today.
+This was removed, which appears to have been successful in relation to the simulation results.
+
+```{r hegnavest3, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/hegnavest3.png"))
+
+filter(rcarbon, site_name  == "Hegna vest 3") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Hegna vest 3", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+Along with Hegna vest 2 (above), Hegna vest 3 was situated where the highway runs today.
+The highway has been removed, and it would seem successfully so, as related to the sea-level adjustments.
+The ^14^C-date to the Middle Mesolithic fits well with the lithic inventory [@eigeland2017].
+
+```{r hesthagc2, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/hesthagc2.png"))
+
+filter(rcarbon, site_name  == "Hesthag C2") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Hesthag C2", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+Hesthag C2 has seen repeated use, as evidenced both by ^14^C-dates and the artefact inventory [@viken2018e].
+While only a few finds can be typologically dated to the Late Mesolithic, these were deemed substantial enough to trust the two ^14^C-dates to the period [@viken2018e, 271].
+No editing of the DTM was necessary.
+
+```{r hesthagc4, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/hesthagc4.png"))
+
+filter(rcarbon, site_name  == "Hesthag C4") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Hesthag C4", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+Hesthag C4
+
+```{r hovland1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/hovland1.png"))
+
+filter(rcarbon, site_name  == "Hovland 1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Hovland 1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r hovland3, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/hovland3.png"))
+
+filter(rcarbon, site_name  == "Hovland 3") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Hovland 3", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r hovland4, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/hovland4.png"))
+
+filter(rcarbon, site_name  == "Hovland 4") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Hovland 4", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r hovland5, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/hovland5.png"))
+
+filter(rcarbon, site_name  == "Hovland 5") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Hovland 5", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r hydal4, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/hydal4.png"))
+
+filter(rcarbon, site_name  == "Hydal 4") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Hydal 4", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r krogenesd1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/krogenesd1.png"))
+
+filter(rcarbon, site_name  == "Krøgenes D1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Krøgenes D1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r krogenesd2, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/krogenesd2.png"))
+
+filter(rcarbon, site_name  == "Krøgenes D2") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Krøgenes D2", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r kvastada1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/kvastada1.png"))
+
+filter(rcarbon, site_name  == "Kvastad A1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Kvastad A1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r kvastada2, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/kvastada2.png"))
+
+filter(rcarbon, site_name  == "Kvastad A2") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Kvastad A2", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r kvastada4, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/kvastada4.png"))
+
+filter(rcarbon, site_name  == "Kvastad A4") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Kvastad A4", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r kvastada9, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/kvastada9.png"))
+
+filter(rcarbon, site_name  == "Kvastad A9") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Kvastad A9", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r langangenv1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/langangenv1.png"))
+
+filter(rcarbon, site_name  == "Langangen Vestgård 1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Langangen Vestgård 1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r langangenv3, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/langangenv3.png"))
+
+filter(rcarbon, site_name  == "Langangen Vestgård 3") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Langangen Vestgård 3", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r langangenv5, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/langangenv5.png"))
+
+filter(rcarbon, site_name  == "Langangen Vestgård 5") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Langangen Vestgård 5", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r langangenv6, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/langangenv6.png"))
+
+filter(rcarbon, site_name  == "Langangen Vestgård 6") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Langangen Vestgård 6", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r langangenv7, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/langangenv7.png"))
+
+filter(rcarbon, site_name  == "Langangen Vestgård 7") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Langangen Vestgård 7", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r langemyr, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/langemyr.png"))
+
+filter(rcarbon, site_name  == "Langemyr") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Langemyr", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r lovas1, echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/lovas1.png"))
+
+filter(rcarbon, site_name  == "Løvås 1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Løvås 1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r lovas2, echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/lovas2.png"))
+
+filter(rcarbon, site_name  == "Løvås 2") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Løvås 2", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r lovas3, echo = FALSE, out.width = "500px"}
+knitr::include_graphics(here("analysis/figures/lovas3.png"))
+
+filter(rcarbon, site_name  == "Løvås 3") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Løvås 3", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r lunaveien, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/lunaveien.png"))
+
+filter(rcarbon, site_name  == "Lunaveien") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Lunaveien", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+Lunaveien is a grave cairn site, and the report states that it is unlikely that any of the finds could relate to a settlement predating the construction of the cairn [@reitan2010].
+The results have therefore been excluded from the aggregative analysis.
+
+```{r nauena, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/nauena.png"))
+
+filter(rcarbon, site_name  == "Nauen A") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Nauen A", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+A section of the highway located a few hundred meters to the south was pre-emptively edited in the DTM. This does not appear to be relevant for the simulated sea-level.
+
+```{r nauenc, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/nauenc.png"))
+
+filter(rcarbon, site_name  == "Nauen C") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Nauen C", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r nordby1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/nordby1.png"))
+
+filter(rcarbon, site_name  == "Nordby 1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Nordby 1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r nordby52, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/nordby52.png"))
+
+filter(rcarbon, site_name  == "Nordby 52") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Nordby 52", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r pauler1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/pauler1.png"))
+
+filter(rcarbon, site_name  == "Pauler 1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Pauler 1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r pauler2, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/pauler2.png"))
+
+filter(rcarbon, site_name  == "Pauler 2") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Pauler 2", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r pjonkerodr1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/pjonkerodr1.png"))
+
+filter(rcarbon, site_name  == "Pjonkerød R1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Pjonkerød R1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r prestemoen1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/prestemoen1.png"))
+
+filter(rcarbon, site_name  == "Prestemoen 1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Prestemoen 1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r ragnhildrod, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/ragnhildrod.png"))
+
+filter(rcarbon, site_name  == "Ragnhildrød") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Ragnhildrød", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r rodbol54, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/rodbol54.png"))
+
+filter(rcarbon, site_name  == "Rødbøl 54") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Rødbøl 54", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r rognlien, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/rognlien.png"))
+
+filter(rcarbon, site_name  == "Rognlien") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Rognlien", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r sageneb1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/sageneb1.png"))
+
+filter(rcarbon, site_name  == "Sagene B1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Sagene B1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r sageneb2, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/sageneb2.png"))
+
+filter(rcarbon, site_name  == "Sagene B2") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Sagene B2", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r solum3, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/solum3.png"))
+
+filter(rcarbon, site_name  == "Solum 3") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Solum 3", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r stokkep1, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/stokkep1.png"))
+
+filter(rcarbon, site_name  == "Stokke/Polland 1") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Stokke/Polland 1", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r stokkep5, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/stokkep5.png"))
+
+filter(rcarbon, site_name  == "Stokke/Polland 5") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Stokke/Polland 5", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r stokkep8, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/stokkep8.png"))
+
+filter(rcarbon, site_name  == "Stokke/Polland 8") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Stokke/Polland 8", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r torstvet, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/torstvet.png"))
+
+filter(rcarbon, site_name  == "Torstvet") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Torstvet", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r trondalen, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/trondalen.png"))
+
+filter(rcarbon, site_name  == "Trondalen") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Trondalen", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r tverdal, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/tverdal.png"))
+
+filter(rcarbon, site_name  == "Tverdal") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Tverdal", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r vallermyrene1a, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/vallermyrene1a.png"))
+
+filter(rcarbon, site_name  == "Vallermyrene 1a") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Vallermyrene 1a", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r vallermyrene1b, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/vallermyrene1b.png"))
+
+filter(rcarbon, site_name  == "Vallermyrene 1b") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Vallermyrene 1b", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r vallermyrene2, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/vallermyrene2.png"))
+
+filter(rcarbon, site_name  == "Vallermyrene 2") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Vallermyrene 2", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r vallermyrene4a, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/vallermyrene4a.png"))
+
+filter(rcarbon, site_name  == "Vallermyrene 4a") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Vallermyrene 4a", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r vallermyrene4b, echo = FALSE, out.width = "500px"}
+include_graphics(here("analysis/figures/vallermyrene4b.png"))
+
+filter(rcarbon, site_name  == "Vallermyrene 4b") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Vallermyrene 4b", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```

### 2021-12-07 analysis/supplementary-materials/supplementary.Rmd
Working through supplementary material

@@
 
-As with Gunnarsrød 6a above, the ^14^C-date to the Stone Age from Gunnarsrød 6b are not related to the Mesolithic inventory on the site [@carrasco2014].
+As with Gunnarsrød 6a above, the ^14^C-date to the Stone Age from Gunnarsrød 6b is not related to the Mesolithic inventory on the site [@carrasco2014].
 
@@
 
-Hesthag C4
+Typology and radiocarbon date match, and there are no modern disturbances in the vicinity of the site [@viken2018b].
+The second ^14^C-date is a reference date of charcoal from a context believed not be anthropogenic.
 
@@
 
+It was noted in the report that it is uncertain how close to the shoreline Hovland 1 would have been situated, based on the ^14^C-dates [@olsen2013b].
+The earliest dates match the lithic inventory.
+The later date is the only indication of activity from the Early Bronze Age/Early Iron Age.
+The site is not located far from the highway, which was been edited for the analysis of Hovland 1 and 4, as well as Torstvet (see below).
+
 ```{r hovland3, echo = FALSE, out.width = "500px"}
@@
 
+Hovland 3 is securely dated both typologically and radiometrically [@solheim2013]. It is, however, located just by the highway. This appears to have been successfully removed.
+
 ```{r hovland4, echo = FALSE, out.width = "500px"}
@@
 
+Hovland 4 is well-dated both typologically and with ^14^C-dates to the Mesolithic. It has a similar location to that of Hovland 3 (above), and editing of the highway seems to have been successful.
+
 ```{r hovland5, echo = FALSE, out.width = "500px"}
@@
 
+Hovland 5 is situated more withdrawn from the highway than the other Hovland sites (above).
+It only has a single ^14^C-date, but this does match the typological indicators of the assemblage [@mansrud2013].
+
 ```{r hydal4, echo = FALSE, out.width = "500px"}
@@
 
+The lithic inventory from Hydal 4 is distinctly Middle Mesolithic in character, and therefore does not match the ^14^C-dates to the Late Mesolithic [@koxvold2017e].
+The simulation of the sea-level does not appear impacted by any modern disturbances.
+
 ```{r krogenesd1, echo = FALSE, out.width = "500px"}
@@
 
+All ^14^C-dates from Krøgenes D1 are believed to be related to use of the site, which has been reused multiple times over a long time-span and which is evidenced in the artefact inventory as well [@reitan2018].
+To what degree the different dates are related to different or the same settlement phases is difficult to ascertain, but is here, as with for the rest of the sites, simply based on grouping dates overlapping with 99.7% probability.
+The site is situated around 100 meter from a road to the south, but this does not appear to impact simulation results, and is therefore left unedited in the DTM.
+
 ```{r krogenesd2, echo = FALSE, out.width = "500px"}
@@
 
+The oldest date is not seen as relevant for the main occupation on the site, in part also due to the large error [@mansrud2018b].
+The site location is similar to that of Krøgenes D1 (above), and did not require any editing.
+
 ```{r kvastada1, echo = FALSE, out.width = "500px"}
@@
 
+^14^C-date and lithic inventory at Kvastad A1 correspond with a date to the end of the Early Mesolithic [@stokke2018].
+The highway to north of the site is clearly visible on the DTM, but does not impact the simulation results.
+
 ```{r kvastada2, echo = FALSE, out.width = "500px"}
@@
 
+The two find areas are predominantly Mesolithic (to the south-west) and Nesolithic (to the north-east) but were treated in combination here as one Mesolithic date is from the Neolithic area [@stokke2018b].
+There is also a late Neolithic element in the lithic inventory that is not reflected in the ^14^C-dates.
+As with Kvastad A1 (above) the highway runs north of the site, clearly visible on the DTM, but does not impact the simulation results.
+
 ```{r kvastada4, echo = FALSE, out.width = "500px"}
@@
 
+While there are a few elements of the lithic inventory from Kvastad A4 that could be related to the Stone Age date, it is not related to main visit to the site, which are Early and Middle Mesolithic in character [@darmark2018c].
+The highway runs over the site today, but this is not relevant for the shore-line reconstruction done here.
+
 ```{r kvastada9, echo = FALSE, out.width = "500px"}
@@
 
+The Stone Age date from Kvastad A9 is not related to the Early Mesolithic assemblage [@darmark2018b]. The highway running by the site does not seem relevant to the simulation results.
+
 ```{r langangenv1, echo = FALSE, out.width = "500px"}
@@
 
+Langangen Vestgård 1 is a securely dated site with both ^14^C-dates and typology [@melvold2014]. No modern disturbances appear relevant for the sea-level adjustment.
+
 ```{r langangenv3, echo = FALSE, out.width = "500px"}
@@
 
+Typology of the site inventory from Langangen Vestgård 3 match the ^14^C-date [@eggen2014a].
+Typological indicators are too coarse to offer any insight on the division into two phases done here for the ^14^C-dates.
+No editing of the DTM was necessary.
+
 ```{r langangenv5, echo = FALSE, out.width = "500px"}
@@
 
+Langangen Vestgård 5. Typology match the ^14^C-dates and a division of two phases, one Late Mesolithic and one Early Neolithic [@reitan2014b]. No editing of the DTM necessary.
+
 ```{r langangenv6, echo = FALSE, out.width = "500px"}
@@
 
+Pottery and the lithic inventory of Langanven Vestgård 6 match ^14^C-dates to the Early Neolithic [@reitan2014c].
+Nothing in the artefact inventory could be related to the Late Neolithic date.
+The DTM did not require editing.
+
 ```{r langangenv7, echo = FALSE, out.width = "500px"}
@@
 
+Langangen Vestgård 7 is represented by a cooking pit found on one of two terraces scattered with lithics that was not known or originally part of the excavation project  [@reitan2014j].
+It was therefore only investigated to a limited degree.
+However, the material does typologically match the ^14^C-date.
+
 ```{r langemyr, echo = FALSE, out.width = "500px"}
@@
 
+Langemyr is typologically and radiometrically dated to the Middle Mesolithic [@koxvold2018a].
+Construction area just south the site was removed.
+This is the same highway-project and raster editing that was done for some of the Hovland sites above.
+
 ```{r lovas1, echo = FALSE, out.width = "500px"}
@@
 
+Løvås 1.
+No editing of the raster appeared to be necessary.
+Interestingly only the Løvås sites were simulated to be situated beneath the sea-level.
+The typologically secure inventory combined with the many overlapping dates from various features does indicate that the issue is with the displacement curve and not the dating of the site.
+Similar situation to that on Løvås 2 and 3 (below).
+
 ```{r lovas2, echo = FALSE, out.width = "500px"}
@@
 
+Løvås 2. Similar situation to that on Løvås 1 and 3 (above and below).
+
 ```{r lovas3, echo = FALSE, out.width = "500px"}
@@
 
+Similar situation to that on Løvås 1 and 2 (above).
+
 ```{r lunaveien, echo = FALSE, out.width = "500px"}
@@
 
-Lunaveien is a grave cairn site, and the report states that it is unlikely that any of the finds could relate to a settlement predating the construction of the cairn [@reitan2010].
+Lunaveien is a grave cairn site, and the report states that it is unlikely that any of the finds relate to a settlement pre-dating the construction of the cairn [@reitan2010].
 The results have therefore been excluded from the aggregative analysis.
@@
 
-A section of the highway located a few hundred meters to the south was pre-emptively edited in the DTM. This does not appear to be relevant for the simulated sea-level.
+Typology of the lithic inventory on Nauen A points to a Late Mesolithic date [@persson2008].
+The later date is seen as related to the establishment of a field for grazing or agriculture, which is also supported by dates that fall outside the analysis here.
+A section of the highway located a few hundred meters to the south was pre-emptively edited in the DTM.
+This does not appear to be relevant for the simulated sea-level.
+
+Nedre Holtan.
 
@@
 
+The typological indicators of the lithic inventory from Nauen C match the earliest dates [@persson2008].
+As with Nauen A, above, the DTM was edited pre-emptively and does not appear to have been necessary.
+
 ```{r nordby1, echo = FALSE, out.width = "500px"}
@@
 
+Building structures from the Late Neolithic, Bronze Age and Early Iron Age [@gjerpe2008b].
+Lithics and pottery finds support this.
+Site located on what is today a highway which was edited also for Nordby 52 (below), but is not relevant for the sea-level reconstruction for Nordby 1.
+
 ```{r nordby52, echo = FALSE, out.width = "500px"}
@@
 
+Nordby 52 is a rock-shelter tat has been visited throughout prehistory [@gjerpe2008].
+Clear typological date to both of the Stone Age periods defined by the ^14^C-dates treated here.
+The site is located by where the highway runs today (see also Nordby 1 above).
+Edit appears to have been successful.
+
 ```{r pauler1, echo = FALSE, out.width = "500px"}
@@
 
+Pauler 1 has a distinctly Early Mesolithic artefact inventory that is not related to the ^14^C-dates [@schallerahrberg2012].
+No modern disturbances appear to imapact the reconstruction of the sea-level corresponding to the ^14^C-dates.
+
 ```{r pauler2, echo = FALSE, out.width = "500px"}
@@
 
+As with Pauler 1 (above), Pauler 2 has a distinctly Early Mesolithic artefact inventory that is not related to the ^14^C-dates [@nyland2012].
+No modern disturbances appear to imapact the reconstruction of the sea-level corresponding to the ^14^C-dates.
+
 ```{r pjonkerodr1, echo = FALSE, out.width = "500px"}
@@
 
+Pjonkerød R1.
+Situated by a quarry today which has been fairly successfully edited.
+However, as the few negative values across the results are likely to result from the interpolation process and the fact that this is a quite large area, and not issues with the displacement curves or ^14^C-dates, the negative values were forced to zero for the aggregative analsysis.
+
 ```{r prestemoen1, echo = FALSE, out.width = "500px"}

### 2021-12-07 analysis/script/02script.R

@@ -153,4 +153,3 @@ isopolys$dir_rel_1 <- c("sw", "sw", "sw", "sw", "nw")
 # Save required data
 save(displacement_curves, isobases, isopolys, xvals,
      file = here("analysis/data/derived_data/02data.RData"))
-

### 2021-12-07 analysis/script/06script.R

@@ -6,6 +6,11 @@ library(grid)
 library(gridExtra)
 library(gtable)
 
+
+# VAlues to be set to 0:
+# Gunnarsrød
+# Pjonkerød R1
+
 # List all files except those starting with a number.
 # That is, all data files associated with sites
 datfiles <- grep("^[0-9]", list.files(here("analysis/data/derived_data")),

### 2021-12-08 analysis/paper/paper.Rmd
Working through supplementary material
Minor edits

@@
 This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g.
-@berg-hansen2009; @bjerck2005; @sognnes2003].
+@berg-hansen2009; @bjerck2005; @sognnes2003; @tallavaara2020].
 
@@
 
-```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from 19-10 ka (data from Hughes et al.
+```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from 19--10 ka (data from Hughes et al.
 2016).
@@
 ```
+
+The shore-bound settlement is believed to follow both from a resource perspective and from one of movement and communication, which could be more efficient on waterways.
+The is thus also extended to the hinterland and the inland, where sites are to be located along rivers and by inland lakes.
+This picture is to take a dramatic turn around 2400 BCE, where the introduction of Neolithic package associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline.
+That is not to say that aquatic resources were no longer exploited, but settlements and tool-production areas were no longer
 
@@
 First of all, the displacement curves are commonly applied directly to larger study areas, with only a couple of studies having taken the variable uplift-rates into account when performing this comparison [e.g.
-@møller1987; @fossum2020].
+@fossum2020; @møller1987; @persson2008].
 Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis.
@@
 While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry.
-RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horisontal displacement of the shoreline will be greater.
+RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater.
 Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of the relation between site and shore line, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.
@@
                                    "Tvedestrand" = "blue",
-                                   "Horten" = 'red')) +
+                                   "Horten" = "darkorange")) +
   theme_bw() + theme(axis.title=element_blank(),
@@
                                    "Tvedestrand" = "blue",
-                                   "Horten" = 'red')) +
+                                   "Horten" = "darkorange")) +
   theme_bw() +
@@
 Polygons were then created from the resulting areas with missing values.
-The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon.
+The horizontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon.
 The topographic distance between site and sea was also found by measuring the shortest distance while taking into account the slope of the terrain on the DTM.
 This was done using the topoDistance package for R [@wang2019].
-The topographic distance was measured between site polygon and the horisontally closest point on the shoreline.
+The topographic distance was measured between site polygon and the horizontally closest point on the shoreline.
 The shortest topographic path was found using the Queen's case neighbourhood of eight cells [@conolly2006].
 In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
-In the case that the sea-polygon completely contain the site, the horisontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon.
+In the case that the sea-polygon completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon.
 While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
@@
                                   "Tvedestrand" = "blue",
-                                 "Hegna vest 1" = "hotpink2")) +
+                                 "Hegna vest 1" = "red")) +
   ylab("Meters above present sea-level") +
-  xlab("Cal BCE/CE") +
+  xlab("cal BCE/CE") +
   theme_bw() +
@@
 
-# Find horisontal distance
+# Find horizontal distance
 hordist <- st_length(hordistl)
@@
 # Create label for map
-dist_label <- paste0("Horisontal distance = ", round(hordist, 1), "m", 
+dist_label <- paste0("horizontal distance = ", round(hordist, 1), "m", 
                 "\nTopographic distance = ", round(tdist, 1), "m",
@@
 Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.
-From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horisontally and topograpchically close, they appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
+From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, they appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
 This corresponds well with the literature, where the introduction of the Neolithic proper is seen as occurring around this time [@prescott2020; @solheim2021].

### 2021-12-08 analysis/script/04script.R

@@ -786,7 +786,7 @@ overview_plot <- function(background_map, sitelimit, sites, isobases,
     scale_colour_manual(values = c("Arendal" = "black",
                                    "Larvik" = "darkgreen",
                                    "Tvedestrand" = "blue",
-                                   "Horten" = 'red')) +
+                                   "Horten" = "darkorange")) +
     theme_bw() + theme(axis.title=element_blank(),
                        axis.text.y = element_blank(),
                        axis.text.x = element_blank(),

### 2021-12-21 analysis/paper/paper.Rmd
Minor edits

@@
 
-The shore-bound settlement is believed to follow both from a resource perspective and from one of movement and communication, which could be more efficient on waterways.
-The is thus also extended to the hinterland and the inland, where sites are to be located along rivers and by inland lakes.
-This picture is to take a dramatic turn around 2400 BCE, where the introduction of Neolithic package associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline.
-That is not to say that aquatic resources were no longer exploited, but settlements and tool-production areas were no longer
+The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from a resource perspective and from one of movement and communication, which would have been efficient on waterways.
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes.
+This general picture is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, where the introduction of the Neolithic proper inolved a.
+associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@solheim2018; @prescott2012].
+That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods.
 
-Based on the generally accepted premise that most Stone Age sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
+Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
 This is for example common in survey projects, which are often guided by a both a digital and mental reconstruction of past sea-levels.
-Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the possible time-span provided by the typological indicators are too wide to decisively verify this.
+Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the typologically informed date-span is too wide to decisively verify this.
 It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g.
@@
 RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater.
-Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of the relation between site and shore line, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.
+Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.
 Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.
@@
 All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a digital research compendium following @marwick2018.
-The dataset is slightly more extensive than what is directly utilised in the study, and includes excavated sites without ^14^C-dates, as well as ^14^C-dates to periods other than the Stone Age.
+The dataset is slightly more extensive than what is directly utilised in the study, and includes excavated sites without ^14^C-dates within the study area, as well as ^14^C-dates to periods other than the Stone Age.
 
-To get at the relationship between sites and contemporaneous shoreline, the analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
+To get at the relationship between sites and the contemporaneous shoreline, the analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
 While there is displacement data available for other areas of south-eastern Norway [e.g.
@@
 The employed shoreline displacement curves are all based on the so-called isolation basin method [@romundset2011].
-This involves extracting cores from a series of basins located beneath the marine limit situated on bedrock in a constrained area of the landscape, and dating the transition from marine to lacustrine sediments in the cores.
+This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments in the cores.
 Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point.
+Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible.
+Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located a significant distance from a common isobase perpendicular to this gradient.
 The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
-The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets, the adjustment to a common isobase based on the direction and gradient in rebound intensity, to name but a few [e.g.
+The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets, the adjustment to a common isobase, to name but a few [e.g.
 @romundset2011; @romundset2019].
@@
 These number a total `r nrow(sites)` sites.
-Of these `r length(unique(rcarbon$site_name))` sites are associated with the total of `r nrow(rcarbon)` radiocarbon dates.
+Of these, `r length(unique(rcarbon$site_name))` sites are associated with the total of `r nrow(rcarbon)` radiocarbon dates.
 In turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability.
-These sites form the basis for the analysis.
+These sites and ^14^C-dates form the basis for the analysis.
 Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History---the institution responsible for archaeological excavations in the region.
@@
 Due to somewhat variable practices between excavations, what available spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports.
-This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.
+This means that the limits are variably given as that defined during initial survey, area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.
 
@@
 The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2019].
-After evaluating the options I decided to use 10m resolution DTM rather than the higher-resolution 1m version.
-In addition to resulting is less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by.
+After evaluating the options I opted for the 10m resolution DTM rather than the higher-resolution 1m version.
+In addition to resulting in considerably less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by.
 The 10m DTM is a down-sampled version of the 1m resolution DTM, which is based on aerial laser scanning using a minimum of two and up to five points per m^2^ [@mapping2019].
@@
 To take account of the gradient in the isostatic rebound, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase.
-This was done by for each site first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves.
+This was done for each site by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves.
 Each value was then divided by the distance between the isobases.
@@
 Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase.
-In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal.
+In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed.
 Multiple phases at a single site were treated as independent of each other.
@@
 
-Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulations were performed.
+Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of relative sea-level change at that location, and a DTM edited to remove substantial modern disturbances, the simulations were performed.
 A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \@ref(fig:example-2)).
@@
 The topographic distance was measured between site polygon and the horizontally closest point on the shoreline.
+This means that the distance is not necessarily measured as the closest topographic distance to the shoreline, but rather as the shortest topographic path to the horizontally closest point on the shoreline.
+This significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact, given the distances considered.
 The shortest topographic path was found using the Queen's case neighbourhood of eight cells [@conolly2006].
@@
 While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
-Negative values where therefore retained.
+Negative values where therefore retained, with the exception of for the sites Gunnarsrød 5, Pjonkerød R1 and , where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
 
@@
 With the transition from a predominantly hunter-gatherer lifeway to early agro-pastrolism, settlements in coastal areas are to be increasingly withdrawn from the shoreline.
-Given this, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE.
+
+The negative values around originate from the sites Løvås 1 and 2.
+These are well-dated sites situated in a relatively undistributed area of the landscape.
+Thus, while there is a danger of circularity of having archaeological sites inform a reconstruction of the shoreline displacement, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper bound, as they would not have been in use when located beneath sea-level.
+Furthermore, as the Løvås sites are located at elevations from where the local displacement curve is interpolated between isololation basins at different elevation, it would seem that the sites represent a case where the archaeological material indicates discrepancies in the geological reconstructions of shoreline displacement in the area.
+
+Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE.
+Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations.
+Following the general and aggregative result of this, one recommendation could be that given no other information, the method of shoreline dating should be applied with an uncertainty of 0-15m offset to the Stone Age site elevation relative to the displacement curves, following the median vertical distance and associated interquartile range.
+This would naturally be best applied when combined with a more comprehensive consideration of the site location, which could lead one to reduce this uncertainty, given for example surroundings where it would be reasonable to assume that a certain sea-levels would be unlikely in light of the surroundings of the site.
 
@@
 Fourthly, the exposure of the sites to wave action is also likely to have been of concern, and will presumably have implications for how close to the shoreline people settled.
-And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site
+And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site.
+These dimensions will have implications for the shore-line dating for certain subsets of the sites and could be future avenues through which one could further improve the method.
+These dimensions will also be imprtant for more substantial considerations of the behavioural relation between sites and the shoreline.
 
@@
 
-Here the approach was used to inform the question of the distance between sites and shorelines, but this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore line, and the uncertainty inherent in this reconstruction.
+The most immediate contribution of this paper is a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline, which has largely been verified.
+Furthermore, the typical application of shoreline dating as an approximate method where date ranges are often for example rounded off to 200-year intervals, means that most previous research based on the method is given support by the findings reported here.
+However, the findings also suggest that simply using the method as a terminus ante quem for the occupation of the sites is, at least in my view, overly conservative and would stand to unnecessarily loose important chronological information.
+Attempting to move these considerations further through an explicit quantification has here resulted in a concrete recommendation of applying an offset of 0-15m when shoreline dating sites where no other information is available.
+
+This quantification was achieved by means of Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty.
+Here the approach was simply used to inform the question of the distance between sites and shorelines.
+However, this method can be extended to a wide range of use-cases where one needs to visualise and both quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
 

### 2021-12-21 analysis/supplementary-materials/supplementary.Rmd
Minor edits

@@
 
-Nordby 52 is a rock-shelter tat has been visited throughout prehistory [@gjerpe2008].
+Nordby 52 is a rock-shelter that has been visited throughout prehistory [@gjerpe2008].
 Clear typological date to both of the Stone Age periods defined by the ^14^C-dates treated here.
@@
 
-Pjonkerød R1.
-Situated by a quarry today which has been fairly successfully edited.
-However, as the few negative values across the results are likely to result from the interpolation process and the fact that this is a quite large area, and not issues with the displacement curves or ^14^C-dates, the negative values were forced to zero for the aggregative analsysis.
+The lithic inventory from Pjonkerød R1 is consistent with the ^14^C-date [@carrasco2015].
+The site is situated by a gravel pit today, which has been fairly successfully edited.
+However, as the few negative values in the results are likely to result from the interpolation process and the fact that this is a quite large area, as opposed to issues with the displacement curves or ^14^C-dates, the negative values were forced to zero for the aggregative analysis.
 
@@
 
+Prestemoen 1 has a lithic inventory consistent with the ^14^C-dates [@persson2014].
+The site is situated by a gravel pit that is clearly visible on the DTM to the north-east of the site.
+This was not edited as it has not impacted the simulated distance between site and sea.
+The landscape surrounding the site would, however, have been substantially different in the Mesolithic [@persson2014, fig.
+10.20].
+
 ```{r ragnhildrod, echo = FALSE, out.width = "500px"}

### 2021-12-26 analysis/paper/paper.Rmd
Added geometries to Brunstad sites and played around with colours for plots

@@
 )
+
+knitr::opts_knit$set(
+  eval.after = "fig.cap"
+)
 ```
@@
 
-# Load processed site and radiocarbon data
+# Load processed site and radiocarbon data, and load functions
 load(here("analysis/data/derived_data/01data.RData"))
@@
 These conditions are directly reflected in the archaeological record.
-In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [@reitan2009].
+In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [@reitan2009].
 Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008].
@@
 2016).
-The studyarea defined later in the text is given a red outline.", out.width = "400px", fig.align="center"}
+The study area defined later in the text is marked with a red outline.", out.width = "400px", fig.align="center"}
 
@@
 If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if it would have been located on an island when the sea-level was higher, it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft.
-Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are fewer, and are typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore-bound site location is largely based on a mosaic of smaller scale investigations.
+Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few, and are typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore-bound site location is largely based on a mosaic of smaller scale investigations.
 
 One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 ^14^C-dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
-They found a overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
+They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
 However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates.
@@
 @parnell2008].
-Another limitation of this often employed method is that it only takes into account the vertical distance between the sites and the sea-level.
+Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level.
 While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry.
@@
 
-```{r map-iso,  fig.cap = "A) Distribution of the 68 analysed sites relative to the isobases of the displacement curves.
-The isobases have a direction of 327 $^{\\circ}$ (Romundset el al.
+```{r map-iso,  fig.cap = paste("A) Distribution of the ", nrow(sites_sa), " analysed sites relative to the isobases of the displacement curves.
+The isobases have a direction of 327$^{\\circ}$ (Romundset el al.
 2018), B) Displacement curves.
-Note the increasing steepness of the curves towards the north-east.", fig.height = 13}
+Note the increasing steepness of the curves towards the north-east."), fig.height = 13}
 bboxsites <- st_bbox(sites_sa)
@@
 overview_map <- ggplot() +
-  geom_sf(data = bmap_reproj, fill = "grey", colour = NA) +
+  geom_sf(data = bmap_reproj, fill = "#fbecdc", colour = "black") +
   geom_sf(data = isobases, aes(colour = name)) +
@@
                      axis.text.x = element_blank(),
+                     panel.background = element_rect(fill = "#dbe3f3",
+                                                    colour = "black"),
                      rect = element_rect(),
@@
 Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase.
-In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed.
+In the case where there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed.
 Multiple phases at a single site were treated as independent of each other.
@@
 While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
-Negative values where therefore retained, with the exception of for the sites Gunnarsrød 5, Pjonkerød R1 and , where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
+Negative values were therefore retained, with the exception of for the sites Gunnarsrød 5, Pjonkerød R1 and , where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
 
@@
 The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5.
-Hovland 5 was chosen for this evuluation as it has a fairly uncertain date range, and is located in area of quite complex topography.
+Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography.
 At 1000 simulation runs the analysis of each site ranged from hours to a couple of days, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.
@@
 elevsamp <- ggplot(sitecurve, aes(x = years)) +
-  geom_line(aes(y =  upperelev), colour = "hotpink2") +
-  geom_line(aes(y = lowerelev), colour = "hotpink2") +
+  geom_line(aes(y =  upperelev), colour = "red") +
+  geom_line(aes(y = lowerelev), colour = "red") +
   ylab("Meters above present sea-level") +
@@
 # Create label for map
-dist_label <- paste0("horizontal distance = ", round(hordist, 1), "m", 
+dist_label <- paste0("Horizontal distance = ", round(hordist, 1), "m", 
                 "\nTopographic distance = ", round(tdist, 1), "m",
@@
 
-```{r example-3, echo = FALSE, fig.cap = "The result of 1000 simulation runs for each of the two grops of dates on the site Hegna vest 1.
+```{r example-3, echo = FALSE, fig.cap = "The result of 1000 simulation runs for each of the two groups of dates on the site Hegna vest 1.
 The first column of plots shows the radiocarbon probability density estimate from where dates were drawn during simulation.
@@
 
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-
 ::: {#refs}
@@
 
-\newpage
+<!-- \newpage -->
 
-### Colophon
+<!-- ### Colophon -->
 
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies: -->
 
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
-```
+<!-- ```{r colophon, cache = FALSE} -->
+<!-- # which R packages and versions? -->
+<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->
+<!-- ``` -->
 
-The current Git commit details are:
+<!-- The current Git commit details are: -->
 
-```{r}
-# what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
-```
+<!-- ```{r} -->
+<!-- # what commit is this file at?  -->
+<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->
+<!-- ``` -->

### 2021-12-26 analysis/supplementary-materials/supplementary.Rmd
Added geometries to Brunstad sites and played around with colours for plots

@@
 
-Hovland 5 is situated more withdrawn from the highway than the other Hovland sites (above).
-It only has a single ^14^C-date, but this does match the typological indicators of the assemblage [@mansrud2013].
+Hovland 5 is situated more withdrawn from the highway than the other Hovland sites (above), and the DTM did therefore not require any edits here.
+The site only has a single ^14^C-date to the Stone Age, but this does match the typological indicators of the assemblage [@mansrud2013].
 
@@
 
-The oldest date is not seen as relevant for the main occupation on the site, in part also due to the large error [@mansrud2018b].
+The oldest date is not seen as relevant for the main occupation on Krøgenes D2, in part also due to the large error [@mansrud2018b].
 The site location is similar to that of Krøgenes D1 (above), and did not require any editing.
@@
 
+Lithic inventory match the ^14^C-date from Rødbøl 54 [@mansrud2008].
+
 ```{r rognlien, echo = FALSE, out.width = "500px"}

### 2021-12-27 analysis/paper/paper.Rmd
Playing around with colours of plots

@@
 
-```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from 19--10 ka (data from Hughes et al.
+```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from c.
+17--8 kyr BCE (data from Hughes et al.
 2016).
@@
 This means that the distance is not necessarily measured as the closest topographic distance to the shoreline, but rather as the shortest topographic path to the horizontally closest point on the shoreline.
-This significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact, given the distances considered.
-The shortest topographic path was found using the Queen's case neighbourhood of eight cells [@conolly2006].
+Not finding the topographically closest point significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact on the results, given the distances considered.
+The shortest topographic path was found using the Queen's pattern neighbourhood of eight cells [@conolly2006].
 In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
@@
 Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography.
-At 1000 simulation runs the analysis of each site ranged from hours to a couple of days, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.
+At 1000 simulation runs the analysis of each site ranged from taking a few hours to a couple of days, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.
 
@@
 seasamp <- ggplot() +
-  geom_sf(data = seapoly, fill = "grey", colour = NA) +
+  geom_sf(data = seapoly, fill = "#dbe3f3", colour = NA) +
   geom_sf(data = sitel, fill = NA, colour = "red") +
@@
 # Map of simulated sea-levels, date group 1
-simsea_plot1 <- shore_plot(output[[1]]$simsea, sitel, 150, date_groups,
+simsea_plot1 <- shore_plot(sitearea, output[[1]]$simsea, sitel, 150, date_groups,
                       s_tdist = 0.5, s_xpos = 1300, s_ypos = 800,
@@
 # Map of simulated sea-levels, date group 2
-simsea_plot2 <- shore_plot(output[[2]]$simsea, sitel, 150, date_groups,
+simsea_plot2 <- shore_plot(sitearea, output[[2]]$simsea, sitel, 150, date_groups,
                       s_tdist = 0.5, s_xpos = 1300, s_ypos = 800,
@@
 The second column displays the result of simulating the raised sea-level 1000 times.
-The darker the colour, the more times the sea-level was simulated to that location.
+The more opaque the colour, the more times the sea-level was simulated to that location.
 The third column shows violin plots of the different distance measures across all simulations.", out.width = "500px"}
@@
 This corresponds well with the literature, where the introduction of the Neolithic proper is seen as occurring around this time [@prescott2020; @solheim2021].
-With the transition from a predominantly hunter-gatherer lifeway to early agro-pastrolism, settlements in coastal areas are to be increasingly withdrawn from the shoreline.
+With the transition from a predominantly hunter-gatherer lifeway to early agro-pastoralism, settlements in coastal areas are to be increasingly withdrawn from the shoreline.
 
-The negative values around originate from the sites Løvås 1 and 2.
-These are well-dated sites situated in a relatively undistributed area of the landscape.
-Thus, while there is a danger of circularity of having archaeological sites inform a reconstruction of the shoreline displacement, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper bound, as they would not have been in use when located beneath sea-level.
+The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3.
+These are recently excavated, well-dated sites situated in a relatively undistributed area of the landscape.
+Thus, while there is a danger of circularity of having archaeological sites inform a reconstruction of the shoreline displacement, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit as they would not have been in use when located beneath sea-level.
 Furthermore, as the Løvås sites are located at elevations from where the local displacement curve is interpolated between isololation basins at different elevation, it would seem that the sites represent a case where the archaeological material indicates discrepancies in the geological reconstructions of shoreline displacement in the area.

### 2021-12-27 analysis/script/05script.R

@@ -101,11 +101,11 @@ ggsave(file = here("analysis/figures/sample_size.png"), width = 300,
 sitename <- "Adal vestre 1"
 date_groups <- group_dates(rcarb_sa, sitename)
 
-output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
-                          isobases, nsamp = nsamp, loc_bbox = 400, siterpath)
-
-save(output,
-      file = here::here("analysis/data/derived_data/adalvestre1.RData"))
+# output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
+#                           isobases, nsamp = nsamp, loc_bbox = 400, siterpath)
+#
+# save(output,
+#       file = here::here("analysis/data/derived_data/adalvestre1.RData"))
 load(here("analysis/data/derived_data/adalvestre1.RData"))
 
 sitearea <- rast(file.path(siterpath,
@@ -123,11 +123,11 @@ ggsave(file = here("analysis/figures/adalvestre1.png"), width = 250,
 sitename <- "Adal vestre 2"
 date_groups <- group_dates(rcarb_sa, sitename)
 
-output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
-                          isobases, nsamp = nsamp, loc_bbox = 400, siterpath)
-save(output,
-     file = here::here("analysis/data/derived_data/adalvestre2.RData"))
-# load(here("analysis/data/derived_data/adalvestre2.RData"))
+# output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
+#                           isobases, nsamp = nsamp, loc_bbox = 400, siterpath)
+# save(output,
+#      file = here::here("analysis/data/derived_data/adalvestre2.RData"))
+load(here("analysis/data/derived_data/adalvestre2.RData"))
 
 sitearea <- rast(file.path(siterpath,
                            paste0(str_replace(sitename, " ", "_"), ".tif")))
@@ -144,11 +144,11 @@ ggsave(file = here("analysis/figures/adalvestre2.png"), width = 250,
 sitename <- "Alveberget 8"
 date_groups <- group_dates(rcarb_sa, sitename)
 
-output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
-                          isobases, nsamp = nsamp, loc_bbox = 400, siterpath)
-save(output,
-     file = here::here("analysis/data/derived_data/alveberget8.RData"))
-# load(here("analysis/data/derived_data/alveberget8.RData"))
[…]

### 2022-01-03 analysis/paper/paper.Rmd
Re-running analyses and writing

@@
 Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
-@solheim2020; @åstveit2018].
+@åstveit2018; @solheim2020].
 This approach has a couple of limitations.
@@
 ```{r map-iso,  fig.cap = paste("A) Distribution of the ", nrow(sites_sa), " analysed sites relative to the isobases of the displacement curves.
-The isobases have a direction of 327$^{\\circ}$ (Romundset el al.
+The isobases have a direction of 327$^{\\circ}$ (Romundset et al.
 2018), B) Displacement curves.
@@
 overview_map <- ggplot() +
-  geom_sf(data = bmap_reproj, fill = "#fbecdc", colour = "black") +
+  geom_sf(data = bmap_reproj, fill = "grey", colour = NA) +
   geom_sf(data = isobases, aes(colour = name)) +
@@
                      axis.text.x = element_blank(),
-                     panel.background = element_rect(fill = "#dbe3f3", 
-                                                    colour = "black"),
                      rect = element_rect(),
@@
 To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter.
-These values were then subtracted from the dsiplacement curve located to the south-west of the site.
+These values were then subtracted from the displacement curve located to the south-west of the site, and in the few cases where the site was located south-west of all isobases, the values were instead added.
 The result of this process is shown for an example site in Figure \@ref(fig:example-1).
@@
 
-# Plot the drawn elevation
+# Plot the drawn elevation on the displacement curve
 elevsamp <- ggplot(sitecurve, aes(x = years)) +
@@
 
-# Create label for map
-dist_label <- paste0("Horizontal distance = ", round(hordist, 1), "m", 
-                "\nTopographic distance = ", round(tdist, 1), "m",
-                "\nVertical distance = ", round(vertdist, 1), "m")
+# Create hillshade
+# For some reason the terra version returned pathchy hillshade (probably due 
+# to the defaults - I didn't bother experimenting too much as this is simply for
+# the purposes of visualisation).
+slope <- raster::terrain(raster(sitearea), 'slope')
+aspect <- raster::terrain(raster(sitearea), 'aspect')
+hill <- raster::hillShade(slope, aspect, 40, 270)
+
+# Make the raster amenable for plotting with ggplot (as.data.frame with terra
+# returned some strange errors)
+raster_df <- raster::as.data.frame(raster(sitearea), xy = TRUE)
+names(raster_df) <- c("x", "y", "value")
 
@@
 seasamp <- ggplot() +
+  geom_raster(data = raster::as.data.frame(hill, xy = TRUE),
+                aes(x = x, y = y, fill = layer)) +
+  scale_fill_gradient(low = "black", high = "grey40", na.value = NA) +
+  new_scale_fill() +
+  geom_raster(data = raster_df, aes(x = x, y = y, fill = value),
+                alpha = 0.3) +
+  scale_fill_gradient(low = "grey", high = "white", na.value = NA) +
+  new_scale_fill() +
+  geom_raster(data = raster_df, aes(x = x, y = y, fill = value)) +
+  scale_fill_gradient(low = NA, high = NA, na.value = "white") +
   geom_sf(data = seapoly, fill = "#dbe3f3", colour = NA) +
@@
                  anchor = c(x = anc[2] - 500, y = anc[1] + 800)) +
-  # Uncomment to annotate with distance measures
-  # annotate("text", x = st_bbox(location_bbox)[1] + 20,
-  #          y = st_bbox(location_bbox)[4] - 100,
-  #          label = dist_label, hjust = 0) +
   theme_bw() + theme(axis.title=element_blank(),
@@
                      legend.position = "none")
+
+# Create label of distance measures
+dist_label <- paste0("Horizontal distance = ", round(hordist, 1), "m", 
+                "\nTopographic distance = ", round(tdist, 1), "m",
+                "\nVertical distance = ", round(vertdist, 1), "m")
 
@@
 Here the approach was simply used to inform the question of the distance between sites and shorelines.
-However, this method can be extended to a wide range of use-cases where one needs to visualise and both quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
+However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
 

### 2022-01-03 analysis/script/04script.R

@@ -735,7 +735,7 @@ site_plot <- function(locationraster, sitelimit, dist, date_groups,
     new_scale_fill() +
     geom_raster(data = raster_df, aes(x = x, y = y, fill = value),
                 alpha = 0.3) +
-    scale_fill_gradient(low = "#fffcfc", high = "#f2dcd0", na.value = NA) +
+    scale_fill_gradient(low = "grey", high = "white", na.value = NA) +
     # scale_fill_gradient(low = "darkgrey", high = "grey", na.value = NA) +
     new_scale_fill() +
     geom_raster(data = raster_df, aes(x = x, y = y, fill = value)) +
@@ -773,7 +773,7 @@ overview_plot <- function(background_map, sitelimit, sites, isobases,
   anc <- as.numeric(c(bboxsites$ymin, bboxsites$xmax))
 
   ggplot() +
-    geom_sf(data = bmap_reproj, fill = "#fbecdc", colour = "black") +
+    geom_sf(data = bmap_reproj, fill = "grey", colour = NA) +
     geom_sf(data = isobases, aes(colour = name)) +
     geom_sf(data = st_centroid(sitelimit), size = 1.5, shape = 21,
             colour = "black", fill = "red") +
@@ -792,8 +792,6 @@ overview_plot <- function(background_map, sitelimit, sites, isobases,
                        axis.text.y = element_blank(),
                        axis.text.x = element_blank(),
                        rect = element_rect(),
-                       panel.background = element_rect(fill = "#dbe3f3",
-                                                       colour = "black"),
                        axis.ticks = element_blank(),
                        panel.grid.major = element_blank(),
                        legend.position = "none")
@@ -832,14 +830,6 @@ shore_plot <- function(locationraster, overlapgrid, sitelimit, dist, date_groups
   raster_df <- raster::as.data.frame(raster(locationraster), xy = TRUE)
   names(raster_df) <- c("x", "y", "value")
 
-  # scols <- c("#f2fbff", "#86a6d4")
-  scols <- c("grey98", "grey40")
-  scols <- c("#59bfff30", "#59bffff0")
-  scols <- c("#bfe6ff", "#59bfff")
-  scols <- c("#c7e3ff", "#59bfff")
-  #scols <- c("#accbff40","#accbf0")
-
-
   ggplot() +
     geom_raster(data = raster::as.data.frame(hill, xy = TRUE),
[…]

### 2022-01-06 analysis/paper/paper.Rmd
Re-running analyses

@@
 The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes.
-This general picture is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, where the introduction of the Neolithic proper inolved a.
+At the transition to the Early Neolithic there is an initial uptake of agriculture This general picture is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, where the introduction of the Neolithic proper inolved a.
 associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@solheim2018; @prescott2012].

### 2022-01-08 analysis/paper/paper.Rmd
Fixed unpacking of results

--- 2022-01-03 analysis/paper/paper.Rmd
@@
 This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g.
-@berg-hansen2009; @bjerck2005; @sognnes2003; @tallavaara2020].
+@berg-hansen2009; @bjerck2005; @sognnes2003; @solheim2018; @tallavaara2020].
 
@@
 
-The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from a resource perspective and from one of movement and communication, which would have been efficient on waterways.
+The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways.
 The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes.
-This general picture is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, where the introduction of the Neolithic proper inolved a.
+This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper.
 associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@solheim2018; @prescott2012].
 That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods.
+Earlier, at the transition to the Early Neolithic (c.
+4000 BCE) pottery is introduced to the sites, and there are some possible indications of an initial, small-scale uptake of agriculture at some sites in the Oslo fjord region.
+However, this initial uptake appears to be relatively small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a de-Neolithisation in the Middle Neolithic.
+[@nielsen2021] has recently argued that while there, the initial uptake of in the Early Neolithic is combined with a more complex settlement pattern, and argues that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data.
+The empirical expectation for this development would thus be a more varied relation between sites and the shore-line.
 
@@
 This is based on typological and technological characteristics of the assemblages.
-Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether they are entirely erroneous results of contamination can be difficult to evaluate.
+Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate.
 Here this will be entirely dependent on, and follow the discretion of the original excavation reports.

### 2022-01-10 analysis/paper/paper.Rmd
References for supplementary material

@@
 source(here("analysis/script/04script.R"), local = knitr::knit_global())
+
+# Exclude Lunaveien and Dybdalshei 2 (See supplementary material)
+sites_sa <- filter(sites_sa, !(name %in% c("Dybdalshei 2", "Lunaveien")))
+
 ```
@@
 This means that the limits are variably given as that defined during initial survey, area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.
+
+Five of the sites have been associated with farming activities, either in directly or in the form building structures.
+The first is Nordby 1, which has .
+The Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are related to farming.
+Both sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities.
+Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region.
 

### 2022-01-10 analysis/supplementary-materials/supplementary.Rmd
Fixed unpacking of results
References for supplementary material

@@
 date: "10/27/2021"
+header-includes:
+  - \usepackage{placeins}
 output: pdf_document
@@
 
-```{r,  echo = FALSE, results = 'hide'}
+```{r,  include=FALSE, message=FALSE, warning=FALSE}
 library(here)
@@
 library(kableExtra)
-
+knitr::opts_chunk$set(echo = FALSE, cache=FALSE)
+
+# Set up so that I can call floatbarrier using regfloat=TRUE in the chunk 
+# headings. See: 
+# https://gist.github.com/burchill/8873d2ade156b27e92a238a774ce2758
+knitr::knit_hooks$set(plot = function (x, options) {
+  float_correct <- function(f, y, opts)  {
+    if (is.null(opts$regfloat) || opts$regfloat==FALSE)
+      paste0(f(y, opts), "\n\n\\FloatBarrier\n")
+    else
+      f(y, opts)
+  }
+  if (!is.null(options$out.width) || !is.null(options$out.height) ||
+      !is.null(options$out.extra) || options$fig.align != "default" ||
+      !is.null(options$fig.subcap)) {
+    if (is.null(options$fig.scap))
+      options$fig.scap = NA
+    return(float_correct(knitr:::hook_plot_tex, x, options))
+  }
+  return(float_correct(knitr:::hook_plot_md_base, x, options))
+})
+```
+
+```{r}
 # Read in radiocarbon dates to be plotted alongside analysis for
@@
 The sites are presented in alphabetical order.
-
-The first figure shows the result of a test run for the site Hovland 5 where the simulations where run 5000 times to identify when the means for the different distance measures converged.
+Tables holding data on excavated sites and radiocarbon dates, and spatial data for site limits and site features are available in the online repoistory for the paper at https://osf.io/7F9SU
+
+Sites with ^14^C-dates to the Stone Age, but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975], Nedre Holtan, Rugtvedt [@odgaard1994], Auve [@østmo2008] and Torsrød [@østmo1975].
+These sites are therefore not included in the analysis.
+
+The first figure shows the result of a test run for the site Hovland 5, where the simulations were run 5000 times to identify when the means for the different distance measures converged.
 This was used to inform the use of 1000 simulation runs for the subsequent analysis of the other sites, as the means appear to have stabilised sufficiently by this point.
@@
 The third plot displays a violin plot of the measured distances between site and shore-line across all simulation runs.
-Measures for central tendency and dispersion associated with these are then given in the table.
-
-```{r adalvestre1,  echo = FALSE, out.width = "500px"}
+
+```{r adalvestre1,  echo = FALSE, out.width = "500px", regfloat=TRUE}
 
@@
 
-```{r dalvestre2,  echo = FALSE, out.width = "500px"}
+```{r dalvestre2,  echo = FALSE, out.width = "500px", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/adalvestre2.png"))
@@
 
-```{r anvik,  echo = FALSE, out.width = "500px"}
+```{r anvik,  echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/anvik.png"))
@@
 
-```{r bakke, echo = FALSE, out.width = "500px"}
+```{r bakke, echo = FALSE, out.width = "500px", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/bakke.png"))
@@
 
-```{r dordal, echo = FALSE, out.width = "500px"}
+```{r dordal, echo = FALSE, out.width = "500px", regfloat=TRUE}
 
@@
 
-```{r dybdalshei1, echo = FALSE, out.width = "500px"}
+```{r dybdalshei1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/dybdalshei1.png"))
@@
 Nonetheless, the site and the ^14^C-dates are still treated the same way as the other sites.
-This is based on the spatial relation between the lithic artefacts and the features, which offers some additional support to the fact that the cooking pits do in fact represent remnants from visits in the Mesolithic.
-
-```{r dybdalshei2, echo = FALSE, out.width = "500px"}
+This is based on the spatial relation between the lithic artefacts and the features, which offers some additional support to the notion that the cooking pits represent remnants from visits in the Mesolithic.
+
+```{r dybdalshei2, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/dybdalshei2.png"))
@@
 
-```{r gunnarsrod4, echo = FALSE, out.width = "500px"}
+```{r gunnarsrod4, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/gunnarsrod4.png"))
@@
 
-Gunnarsrød 4 [@reitan2014k], and Gunnarsrød 5 (below), were situated close to where the railway runs today, and therefore required editing of the DTM.
+The rock-shelter site Gunnarsrød 4 [@reitan2014k], and Gunnarsrød 5 (below), were situated close to where the railway runs today, and therefore required editing of the DTM.
 Based on maps from the report, this appears to have given adequate, although not perfect results [@reitan2014k, fig.14.7.1].
@@
 
-```{r gunnarsrod5, echo = FALSE, out.width = "500px"}
+```{r gunnarsrod5, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/gunnarsrod5.png"))
@@
 This has likely led to the negative values on the distance measures.
-In combination, the simulation results and the site map in the report indicate that the site would not have been situated any significant distance from the shoreline.
+In combination, the simulation results and the site map in the report [@reitan2014d, fig.9.1] indicate that the site would not have been situated any significant distance from the shoreline.
 The values have therefore been set to zero for the final aggregative analysis.
@@
 
-```{r gunnarsrod6a, echo = FALSE, out.width = "500px"}
+```{r gunnarsrod6a, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/gunnarsrod6a.png"))
@@
 
-The ^14^C-date to the Stone Age from Gunnarsrød 6a is not related to the lithic inventory on the site, which is clearly Mesolithic [@carrasco2014].
-
-```{r gunnarsrod6b, echo = FALSE, out.width = "500px"}
+The ^14^C-date to the Neolithic from Gunnarsrød 6a is not related to the lithic inventory on the site, which is clearly Mesolithic [@carrasco2014].
+
+```{r gunnarsrod6b, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/gunnarsrod6b.png"))
@@
 
-As with Gunnarsrød 6a above, the ^14^C-date to the Stone Age from Gunnarsrød 6b is not related to the Mesolithic inventory on the site [@carrasco2014].
-
-```{r gunnarsrod7, echo = FALSE, out.width = "500px"}
+As with Gunnarsrød 6a above, the ^14^C-date to the Neolithic from Gunnarsrød 6b is not related to the Mesolithic inventory on the site [@carrasco2014].
+
+```{r gunnarsrod7, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/gunnarsrod7.png"))
@@
 
-Gunnarsrød 7 is situated on a slope, stretching from 55 masl up to 59 [@fossum2014b]. The lithic inventory is of a Middle Mesolithic character and does not match the ^14^C-dates.
-
-```{r gunnarsrod10, echo = FALSE, out.width = "500px"}
+ The lithic inventory from Gunnarsrød 7 is of a Middle Mesolithic character and does not match the ^14^C-dates [@fossum2014b].
+
+```{r gunnarsrod10, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/gunnarsrod10.png"))
@@
 
-Based on the location of Gunnarsrød 10 on a steep ledge, and the site inventory, the ^14^C-date is not believed to be related to the main occupation of the site [@reitan2014l].
-
-```{r hegnavest1, echo = FALSE, out.width = "500px"}
+Based on the location of Gunnarsrød 10 on a steep ledge indicating a shore-bound location, and the site inventory, the ^14^C-date is not believed to be related to the main occupation of the site [@reitan2014l].
+
+```{r hegnavest1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/hegnavest1.png"))
@@
 
-```{r hegnavest2, echo = FALSE, out.width = "500px"}
+```{r hegnavest2, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/hegnavest2.png"))
@@
 
-```{r hegnavest3, echo = FALSE, out.width = "500px"}
+```{r hegnavest3, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/hegnavest3.png"))
@@
 
-```{r hesthagc2, echo = FALSE, out.width = "500px"}
+```{r hesthagc2, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/hesthagc2.png"))
@@
 
-```{r hesthagc4, echo = FALSE, out.width = "500px"}
+```{r hesthagc4, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/hesthagc4.png"))
@@
 
-Typology and radiocarbon date match, and there are no modern disturbances in the vicinity of the site [@viken2018b].
+Typology and radiocarbon date match, and there are no modern disturbances in the vicinity of Hesthag C4 [@viken2018b].
 The second ^14^C-date is a reference date of charcoal from a context believed not be anthropogenic.
 
-```{r hovland1, echo = FALSE, out.width = "500px"}
+```{r hovland1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/hovland1.png"))
@@
 The later date is the only indication of activity from the Early Bronze Age/Early Iron Age.
-The site is not located far from the highway, which was been edited for the analysis of Hovland 1 and 4, as well as Torstvet (see below).
-
-```{r hovland3, echo = FALSE, out.width = "500px"}
+The site is not located far from the highway, which was also edited for the analysis of Hovland 1 and 4, as well as Torstvet (see below).
+
+```{r hovland3, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/hovland3.png"))
@@
 
-```{r hovland4, echo = FALSE, out.width = "500px"}
+```{r hovland4, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/hovland4.png"))
@@
 
-Hovland 4 is well-dated both typologically and with ^14^C-dates to the Mesolithic. It has a similar location to that of Hovland 3 (above), and editing of the highway seems to have been successful.
-
-```{r hovland5, echo = FALSE, out.width = "500px"}
+Hovland 4 is well-dated both typologically and with ^14^C-dates to the Mesolithic [@mansrud2013a].
+It has a similar location to that of Hovland 3 (above), and editing of the highway seems to have been successful.
+
+```{r hovland5, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/hovland5.png"))
@@
 
-```{r hydal4, echo = FALSE, out.width = "500px"}
+```{r hydal4, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/hydal4.png"))
@@
 
-```{r krogenesd1, echo = FALSE, out.width = "500px"}
+```{r krogenesd1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/krogenesd1.png"))
@@
 
-All ^14^C-dates from Krøgenes D1 are believed to be related to use of the site, which has been reused multiple times over a long time-span and which is evidenced in the artefact inventory as well [@reitan2018].
+All ^14^C-dates from Krøgenes D1 are believed to be related to use of the site, which has been reused multiple times over a long time-span [@reitan2018].
+This is evidenced by the artefact inventory as well.
 To what degree the different dates are related to different or the same settlement phases is difficult to ascertain, but is here, as with for the rest of the sites, simply based on grouping dates overlapping with 99.7% probability.
@@
 
-```{r krogenesd2, echo = FALSE, out.width = "500px"}
+```{r krogenesd2, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/krogenesd2.png"))
@@
 
-```{r kvastada1, echo = FALSE, out.width = "500px"}
+```{r kvastada1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/kvastada1.png"))
@@
 
-```{r kvastada2, echo = FALSE, out.width = "500px"}
+```{r kvastada2, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/kvastada2.png"))
@@
 
-```{r kvastada4, echo = FALSE, out.width = "500px"}
+```{r kvastada4, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/kvastada4.png"))
@@
 
-While there are a few elements of the lithic inventory from Kvastad A4 that could be related to the Stone Age date, it is not related to main visit to the site, which are Early and Middle Mesolithic in character [@darmark2018c].
+While there are a few elements of the lithic inventory from Kvastad A4 that could be related to the late Middle Mesolithic ^14^C-date, it is not related to the main visits to the site, which are Early Mesolithic in character [@darmark2018c].
 The highway runs over the site today, but this is not relevant for the shore-line reconstruction done here.
 
-```{r kvastada9, echo = FALSE, out.width = "500px"}
+```{r kvastada9, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/kvastada9.png"))
@@
 
-```{r langangenv1, echo = FALSE, out.width = "500px"}
+```{r langangenv1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/langangenv1.png"))
@@
 
-```{r langangenv3, echo = FALSE, out.width = "500px"}
+```{r langangenv3, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/langangenv3.png"))
@@
 
-```{r langangenv5, echo = FALSE, out.width = "500px"}
+```{r langangenv5, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/langangenv5.png"))
@@
 
-```{r langangenv6, echo = FALSE, out.width = "500px"}
+```{r langangenv6, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/langangenv6.png"))
@@
 
-```{r langangenv7, echo = FALSE, out.width = "500px"}
+```{r langangenv7, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/langangenv7.png"))
@@
 
-```{r langemyr, echo = FALSE, out.width = "500px"}
+```{r langemyr, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/langemyr.png"))
@@
 
-```{r lovas1, echo = FALSE, out.width = "500px"}
+```{r lovas1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/lovas1.png"))
@@
 
-```{r lovas2, echo = FALSE, out.width = "500px"}
+```{r lovas2, echo = FALSE, out.width = "500px", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/lovas2.png"))
@@
 
-```{r lovas3, echo = FALSE, out.width = "500px"}
+```{r lovas3, echo = FALSE, out.width = "500px", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/lovas3.png"))
@@
 
-```{r lunaveien, echo = FALSE, out.width = "500px"}
+```{r lunaveien, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/lunaveien.png"))
@@
 
-```{r nauena, echo = FALSE, out.width = "500px"}
+```{r nauena, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/nauena.png"))
@@
 
-Nedre Holtan.
-
-```{r nauenc, echo = FALSE, out.width = "500px"}
+```{r nauenc, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/nauenc.png"))
@@
 
-```{r nordby1, echo = FALSE, out.width = "500px"}
+```{r nordby1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/nordby1.png"))
@@
 
-Building structures from the Late Neolithic, Bronze Age and Early Iron Age [@gjerpe2008b].
-Lithics and pottery finds support this.
+Site dated to the Late Neolithic, Bronze Age and Early Iron Age [@gjerpe2008b] with long-house 4 dated to the LN.
+Lithics and pottery finds support the dates.
 Site located on what is today a highway which was edited also for Nordby 52 (below), but is not relevant for the sea-level reconstruction for Nordby 1.
 
-```{r nordby52, echo = FALSE, out.width = "500px"}
+```{r nordby52, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/nordby52.png"))
@@
 
-```{r pauler1, echo = FALSE, out.width = "500px"}
+```{r pauler1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/pauler1.png"))
@@
 
-```{r pauler2, echo = FALSE, out.width = "500px"}
+```{r pauler2, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/pauler2.png"))
@@
 
-```{r pjonkerodr1, echo = FALSE, out.width = "500px"}
+```{r pjonkerodr1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/pjonkerodr1.png"))
@@
 
-```{r prestemoen1, echo = FALSE, out.width = "500px"}
+```{r prestemoen1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/prestemoen1.png"))
@@
 
-```{r ragnhildrod, echo = FALSE, out.width = "500px"}
+```{r ragnhildrod, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/ragnhildrod.png"))
@@
 
-```{r rodbol54, echo = FALSE, out.width = "500px"}
+```{r rodbol54, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/rodbol54.png"))
@@
 
-```{r rognlien, echo = FALSE, out.width = "500px"}
+```{r rognlien, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/rognlien.png"))
@@
 
-```{r sageneb1, echo = FALSE, out.width = "500px"}
+```{r sageneb1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/sageneb1.png"))
@@
 
-```{r sageneb2, echo = FALSE, out.width = "500px"}
+```{r sageneb2, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/sageneb2.png"))
@@
 
-```{r solum3, echo = FALSE, out.width = "500px"}
+```{r solum3, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/solum3.png"))
@@
 
-```{r stokkep1, echo = FALSE, out.width = "500px"}
+```{r stokkep1, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/stokkep1.png"))
@@
 
-```{r stokkep5, echo = FALSE, out.width = "500px"}
+```{r stokkep5, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/stokkep5.png"))
@@
 
-```{r stokkep8, echo = FALSE, out.width = "500px"}
+```{r stokkep8, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/stokkep8.png"))
@@
 
-```{r torstvet, echo = FALSE, out.width = "500px"}
+```{r torstvet, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/torstvet.png"))
@@
 
-```{r trondalen, echo = FALSE, out.width = "500px"}
+```{r trondalen, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/trondalen.png"))
@@
 
-```{r tverdal, echo = FALSE, out.width = "500px"}
+```{r tverdal, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/tverdal.png"))
@@
 
-```{r vallermyrene1a, echo = FALSE, out.width = "500px"}
+```{r vallermyrene1a, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/vallermyrene1a.png"))
@@
 
-```{r vallermyrene1b, echo = FALSE, out.width = "500px"}
+```{r vallermyrene1b, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/vallermyrene1b.png"))
@@
 
-```{r vallermyrene2, echo = FALSE, out.width = "500px"}
+```{r vallermyrene2, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/vallermyrene2.png"))
@@
 
-```{r vallermyrene4a, echo = FALSE, out.width = "500px"}
+```{r vallermyrene4a, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/vallermyrene4a.png"))
@@
 
-```{r vallermyrene4b, echo = FALSE, out.width = "500px"}
+```{r vallermyrene4b, echo = FALSE, out.width = "500px", regfloat=TRUE}
 include_graphics(here("analysis/figures/vallermyrene4b.png"))
@@
 ```
+
+Bratsberg was originally two sites [@wenn2012]. [@nielsen2021] has interpreted this as an indicator of early farming.

### 2022-01-10 analysis/script/06script.R

@@ -8,14 +8,14 @@ library(gtable)
 library(cowplot)
 
 # List all files except those starting with a number.
-# That is, all data files associated with sites
+# That is, all data files resulting from analysis in 05script.R
 datfiles <- grep("^[0-9]", list.files(here("analysis/data/derived_data")),
      invert = TRUE, value = TRUE)
 
 # Create empty list to hold results
 results_list <- list()
 
-# loop over, load results and assign them to the list
+# Loop over, load results and assign them to the list
 for(i in 1:length(datfiles)){
   # Load results
   load(file.path(here("analysis/data/derived_data", datfiles[i])))
@@ -23,9 +23,8 @@ for(i in 1:length(datfiles)){
                              c("datedat", "sitel", "sitecurve"))]
 }
 
-# results_list is a list of lists of variable length...
-# Anyway, these need to be unpacked. Might be a smoother purrr solution to
-# do all of this.
+# results_list is a list of lists of variable length that need to be unpacked.
+# (Might be a smoother purrr solution to do all of this?)
 results <- list()
 for(i in 1:length(results_list)){
   dat <- results_list[[i]]
@@ -41,7 +40,7 @@ distances <- results %>% bind_rows()
 distances <-  distances$results
 
 # Negative values for Gunnarsrød 5 and Pjonkerød R1 are to be set to zero
-# (see evaluation in supplementary material)
+# (see site evaluations in the supplementary material)
 distances <- distances %>%
   mutate(vertdist = ifelse(sitename %in% c("Gunnarsrød 5", "Pjonkerød R1")
                            & vertdist < 0,  0, vertdist),
@@ -49,8 +48,6 @@ distances <- distances %>%
                           & hordist < 0,  0, hordist),
          topodist = ifelse(sitename %in% c("Gunnarsrød 5", "Pjonkerød R1")
                            & topodist < 0,  0, topodist))
-
-max(distances[distances$sitename == "Gunnarsrød 5",]$vertdist, na.rm = TRUE)
 
 sum_all <- distances %>%
   summarise_at(c("hordist", "topodist", "vertdist"),
@@ -194,12 +191,6 @@ grd <- plot_grid(grd1, grd2, ncol = 1, scale = 0.9)
 
 ggsave(file = here("analysis/figures/results.png"), grd,
        bg = "white")
-
-# ggsave(file = here("analysis/figures/results_cor.png"), grd2,
[…]

### 2022-01-14 analysis/paper/paper.Rmd
Trying to identify recommended offset for shoreline dating

@@
 library(IRanges)
+library(stringr)
 quickSetupOxcal()
@@
 This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g.
-@berg-hansen2009; @bjerck2005; @sognnes2003; @solheim2018; @tallavaara2020].
+@berg-hansen2009; @bjerck2005; @sognnes2003; @tallavaara2020].
 
@@
 Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g.
-@berg-hansen2009; @bjerck1990; @mjærum2018].
+@berg-hansen2009; @bjerck1990; @mjærum2018; @solheim2018].
 The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
@@
 These number a total `r nrow(sites)` sites.
-Of these, `r length(unique(rcarbon$site_name))` sites are associated with the total of `r nrow(rcarbon)` radiocarbon dates.
-In turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability.
+Of these, `r length(unique(rcarbon$site_name))` sites are associated with a total of `r nrow(rcarbon)` radiocarbon dates.
+Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability.
 These sites and ^14^C-dates form the basis for the analysis.
@@
 
-Five of the sites have been associated with farming activities, either in directly or in the form building structures.
+Three of the sites have been associated with farming activities, either in directly or in the form building structures.
 The first is Nordby 1, which has .
-The Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are related to farming.
-Both sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities.
+Furthermore, the Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are both related to farming.
+Both of these sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities.
+These agricultural phases are thus treated seperately below.
 Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region.
@@
 
+```{r distances}
+distances2 <- distances %>%
+  filter(year <= -2500 & rcarb_cor == "t" & type == "forager") %>% 
+  mutate(hordist = ifelse(str_detect(sitename, "Løvås"), 0, hordist),
+         topodist = ifelse(str_detect(sitename, "Løvås"), 0, topodist),
+         vertdist = ifelse(str_detect(sitename, "Løvås"), 0, vertdist))
+
+distances3 <- distances %>%
+  filter(year <= -4000 & rcarb_cor == "t" & type == "forager") %>% 
+  mutate(hordist = ifelse(str_detect(sitename, "Løvås"), 0, hordist),
+         topodist = ifelse(str_detect(sitename, "Løvås"), 0, topodist),
+         vertdist = ifelse(str_detect(sitename, "Løvås"), 0, vertdist))
+```
 # Results
@@
 Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations.
+The best point estimate for the distance between site and shoreline for the pre-Late Neolithic is, following from the skewed distribution, given by the median at `r round(median(distances2$vertdist))`m.
+However, 95% of the distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
+That is, for 95% of the cases, the shoreline was situated on or down to m below the site location.
+This is constrained further when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m.
+When applying the method of shoreline dating to Stone Age sites in the region, it could thus be recommended, given no other information, to allow for a potential deviation between site and shoreline from `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))` in vertical distance.
+This can be reduced to `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m for Mesolithic sites.
+The sites treated here have been given a shoreline date using these values in Figure.
+However,
+
 Following the general and aggregative result of this, one recommendation could be that given no other information, the method of shoreline dating should be applied with an uncertainty of 0-15m offset to the Stone Age site elevation relative to the displacement curves, following the median vertical distance and associated interquartile range.
@@
 
+Figure shows the result of shoreline dating the sites using the recommended off-set with radiocarbon dates plotted against this.
+The steeper the curve the more narrow the date range resulting from the procedure.
+This means that older siter, and sites situated towards the north-east will have a more narrow date range.
+
 ```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites.
-Each data point is plotted with some transparency, meaning that the darker the colour, the more often those values occured.
+Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occured.
+Results associated with agricultural acitivites are in grey.
 The first row shows the result of including all dates the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites.
@@
 
-```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using only dates older than 2500 BCE that correspond to the site inventory.", out.width = "525px", fig.align="center"}
+```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using only dates older than 2500 BCE that correspond to the site inventory.
+Negative values have been set to zero.", out.width = "525px", fig.align="center"}
 knitr::include_graphics(here("analysis/figures/results2.png"))
@@
 
+```{r results3, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using only dates older than 4000 BCE that correspond to the site inventory.
+Negative values have been set to zero.", out.width = "525px", fig.align="center"}
+knitr::include_graphics(here("analysis/figures/results3.png"))
+```
+
+```{r shoredate, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites allowing for a sea-level potentially 4m above to 16m below the lowest site elevation.
+These are plotted against the oldest radiocarbon-informed dates for each site.", out.width = "525px", fig.align="center"}
+knitr::include_graphics(here("analysis/figures/shoredate.png"))
+```
+
 # Discussion
+
+While allowing for a deviation up to 16m might be recommended when no other information is available, this can likely be considerably reduced by evaluating the surroundings of the site.
 
@@
 Thirdly, the tidal range has not been considered...
-Fourthly, the exposure of the sites to wave action is also likely to have been of concern, and will presumably have implications for how close to the shoreline people settled.
+Fourthly, the exposure of the sites to wave action is also likely to have been of concern [@roalkvam2020], and will presumably have implications for how close to the shoreline people settled [@blankholm2020].
 And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site.

### 2022-01-14 analysis/supplementary-materials/supplementary.Rmd
Trying to identify recommended offset for shoreline dating

@@
 The sites are presented in alphabetical order.
-Tables holding data on excavated sites and radiocarbon dates, and spatial data for site limits and site features are available in the online repoistory for the paper at https://osf.io/7F9SU
-
-Sites with ^14^C-dates to the Stone Age, but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975], Nedre Holtan, Rugtvedt [@odgaard1994], Auve [@østmo2008] and Torsrød [@østmo1975].
+excavated sites and radiocarbon dates, and spatial data for site limits and site features are available in the online repoistory for the paper at https://osf.io/7F9SU
+
+Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan, Rugtvedt [@odgaard1994], Auve [@østmo2008] and Torsrød [@østmo1975].
 These sites are therefore not included in the analysis.
+Furthermore, @nielsen2021 has recently interpreted features from the otherwise younger sites Bratsberg and Larønningen as possibly related to farming activity in the Early Neolithic.
+However, given the limited number of features, the lack of any artefacts directly related to these, and the therefore somewhat speculative nature of this suggestion, the sites are kept out of the analysis here.
 
@@
 This was not edited as it has not impacted the simulated distance between site and sea.
-The landscape surrounding the site would, however, have been substantially different in the Mesolithic [@persson2014, fig.
+The landscape surrounding the site would, however, have been very different in the Mesolithic [@persson2014, fig.
 10.20].
@@
 
+The earliest ^14^C-date does not match the Middle Mesolithic inventory from Ragnhildrød [@mjaerum2012].
+The post-hole dated to the Late Neolithic/Bronze Age could be related to a single find of a pressure flaked object, but was not given much weight in the report [@mjaerum2012, 76].
+
 ```{r rodbol54, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
-Lithic inventory match the ^14^C-date from Rødbøl 54 [@mansrud2008].
+Lithic inventory match the ^14^C-date from Rødbøl 54 [@mansrud2008]. The site is situated where the highway runs today, which required editing.
 
@@
 
+Rognlien was originally excavated in [@ingstad1970].
+The site limit has been manually defined based on the description given in the report and the spatial geometry of the site as defined in the database Askeladden.
+It could be slightly off, but should match the elevation given in the publication of the site [@ingstad1970].The ^14^C-dates from the original excavation was not seen as contemporaneous with the lithic material on the site.
+Burnt elk bone from the original excavation was therefore recently subjected to radiocarbon dating and is believed to match the occupation of the site [pers.comm.
+Steinar Solheim, also provided in @stokke2017, 24].
+
 ```{r sageneb1, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+^14^C-dates from Sagene B1 do not match the Early Mesolithic inventory of the site [@viken2018].
+The wide discrepancy of the dates from the the post-holes also led @viken2018 to suggest that these features might actually be non-anthropogenic and rather be related to trees that have grown on the site.
+
 ```{r sageneb2, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+The ^14^C-dates from Sagene B2 do not match the clearly Early Mesolithic inventory of the site [@darmark2018a].
+
 ```{r solum3, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+Solum 3 is a Late Neolithic and Early Bronze Age site, with ^14^C-dates matching the lithic inventory [@fossum2014d].
+
 ```{r stokkep1, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+The artefact inventory from Stokke/Polland 1, containing both lithics and ceramics, indicates visits in the Late Mesolithic as well as the Early Neolithic [@koxvold2017i].
+Date groups were therefore manually separated, even though they overlap at 99.7% probability.
+The site is located right by the highway to the east, which was edited for the other Stokke/Polland sites, but this does not appear relevant for the reconstruction of the sea-level at Stokke/Polland 1, which would have reached the site from the west.
+
 ```{r stokkep5, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+The ^14^C-dates matches the lithic inventory from Stokke/Polland 5 [@mansrud2017].
+The site is located where the highway runs today.
+This appears to have been successfully edited.
+The excavated area is more extensive (field A--C in the report) that what is used as the site limit here, but the relation between the different areas is not clear.
+Field A, used as the site limit here, holds all the dated features.
+
 ```{r stokkep8, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+Stokke/Polland 8 have corresponding lithic inventory and ^14^C-date [@fossum2017c].
+The site is situated just east of where the highway runs today.
+This appears to have been successfully edited [cf.
+@fossum2017c, fig.28.3].
+
 ```{r torstvet, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+Typology of lithics match the ^14^C-dates from Torstvet [@mansrud2013b].
+The site is located where the highway runs today.
+This appears to have been successfully edited, although there is a possibility that this has led to the sea-level being simulated sightly further away from the site than what would otherwise have been the case.
+However, as with he Hovland-sites above, this appears to rather follow from the general topography of the landscape which makes the reconstruction more uncertain than in areas with a higher relief.
+
 ```{r trondalen, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+^14^C-dates match the typological indicators in the artefact assemblage at Trondalen [@mansrud2018c]. No disturbances required editing in the DTM.
+
 ```{r tverdal, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 ```
-
-Bratsberg was originally two sites [@wenn2012]. [@nielsen2021] has interpreted this as an indicator of early farming.

### 2022-01-14 analysis/script/06script.R

@@ -1,6 +1,7 @@
 library(here)
 library(dplyr)
 library(purrr)
+library(stringr)
 library(ggplot2)
 library(grid)
 library(gridExtra)
@@ -23,6 +24,9 @@ for(i in 1:length(datfiles)){
                              c("datedat", "sitel", "sitecurve"))]
 }
 
+# Sites and phases to label as agricultural
+agr <- c("Nordby 1", "Kvastad A2", "Nauen A")
+
 # results_list is a list of lists of variable length that need to be unpacked.
 # (Might be a smoother purrr solution to do all of this?)
 results <- list()
@@ -30,14 +34,23 @@ for(i in 1:length(results_list)){
   dat <- results_list[[i]]
   tmp <- list()
   for(j in 1:length(dat)){
-    tmp[[j]] <- dat[[j]]["results"]
+    pdat <- dat[[j]]["results"][[1]]
+    if(is.element(unique(pdat$sitename), agr) & any(pdat$year > -4000)){
+        pdat$type <- "agricultural"
+      } else{
+        pdat$type <- "forager"
+      }
+    tmp[[j]] <- pdat
   }
   results[[i]] <- tmp
 }
 
 # Collapse list of lists into a single data frame
 distances <- results %>% bind_rows()
-distances <-  distances$results
+# distances <-  distances$results
+
+# Lunaveien and Dybdalshei 2 are excluded
+# (see site evaluations in the supplementary material)
 
 # Negative values for Gunnarsrød 5 and Pjonkerød R1 are to be set to zero
 # (see site evaluations in the supplementary material)
@@ -49,6 +62,9 @@ distances <- distances %>%
          topodist = ifelse(sitename %in% c("Gunnarsrød 5", "Pjonkerød R1")
                            & topodist < 0,  0, topodist))
 
+save(distances,
+     file = here::here("analysis/data/derived_data/06data.RData"))
+
 sum_all <- distances %>%
   summarise_at(c("hordist", "topodist", "vertdist"),
                c(mean, median, sd, IQR)) %>%
@@ -59,15 +75,20 @@ sum_cor <- distances %>% filter(rcarb_cor == "t") %>%
                c(mean, median, sd, IQR)) %>%
    round()
 
+colsh = c("forager" = "#00ba38", "agricultural" = "darkgrey")
+colst = c("forager" = "#fad510", "agricultural" = "darkgrey")
+colsv = c("forager" = "steelblue", "agricultural" = "darkgrey")
+
[…]

### 2022-01-14 analysis/script/07script.R

@@ -1,0 +1,89 @@
+library(ggplot2)
+library(dplyr)
+library(here)
+library(terra)
+library(sf)
+
+# Load required functions and prepared data
+source(here("analysis/script/04script.R"))
+load(here("analysis/data/derived_data/01data.RData"))
+load(here("analysis/data/derived_data/02data.RData"))
+
+# Read in raster
+dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
+
+# List all files except those starting with a number.
+# That is, all data files resulting from analysis in 05script.R
+datfiles <- grep("^[0-9]", list.files(here("analysis/data/derived_data")),
+                 invert = TRUE, value = TRUE)
+
+# Create empty list to hold dat data
+posterior_list <- list()
+
+# Loop over, load and assign them to the list
+for(i in 1:length(datfiles)){
+  # Load date data
+  load(file.path(here("analysis/data/derived_data", datfiles[i])))
+  posterior_list[[i]] <- output[!(names(output) %in%
+                                  c("", "sitel", "sitecurve"))]$datedat %>%
+    filter(rcarb_cor == "t") %>%
+    filter(group == min(group)) %>%
+    # Make sure site name is included on all rows (sums have NA)
+    mutate(site_name = unique(na.omit(site_name)))
+}
+
+# Unpack list of lists
+rdates <- posterior_list %>% bind_rows()
+rdates <- rdates %>%  group_by(site_name) %>%
+  # If there is a sum, exclude the rest
+  filter(class == 'sum' | all(class != 'sum')) %>%
+  ungroup() %>%
+  filter(class != "aprior")
+
+
+# Assign closest isobases to site data and exclude sites not analysed
+sites_sa <- st_join(st_make_valid(sites_sa), isopolys,
+                    join = st_intersects, largest = TRUE) %>%
+  filter(!(name %in% c("Dybdalshei 2", "Lunaveien", "Frebergsvik C")))
+
+shorelinedates <- list()
+for(i in 1:nrow(sites_sa)){
+  print(sites_sa$name[i])
+  sdates <- shoreline_date(sitename = sites_sa$name[i],
+                 dtm = dtm,
+                 displacement_curves = displacement_curves,
+                 features = rcarb_sa,
+                 sites = sites_sa,
+                 isobases = isobases,
[…]

### 2022-01-19 analysis/paper/paper.Rmd
Exponential distance to shoreline

@@
   filter(year <= -2500 & rcarb_cor == "t" & type == "forager") %>% 
-  mutate(hordist = ifelse(str_detect(sitename, "Løvås"), 0, hordist),
-         topodist = ifelse(str_detect(sitename, "Løvås"), 0, topodist),
-         vertdist = ifelse(str_detect(sitename, "Løvås"), 0, vertdist))
+  mutate(hordist = ifelse(hordist < 0, 0, hordist),
+         topodist = ifelse(topodist < 0, 0, topodist),
+         vertdist = ifelse(vertdist < 0, 0, vertdist))
 
@@
   filter(year <= -4000 & rcarb_cor == "t" & type == "forager") %>% 
-  mutate(hordist = ifelse(str_detect(sitename, "Løvås"), 0, hordist),
-         topodist = ifelse(str_detect(sitename, "Løvås"), 0, topodist),
-         vertdist = ifelse(str_detect(sitename, "Løvås"), 0, vertdist))
+  mutate(hordist = ifelse(hordist < 0, 0, hordist),
+         topodist = ifelse(topodist < 0, 0, topodist),
+         vertdist = ifelse(vertdist < 0, 0, vertdist))
 ```

### 2022-01-25 analysis/paper/paper.Rmd
Minor edits
Applying exponential decay to the shoreline dating

@@
 However,
+An exponential function has been fit to the distribution.
+While it is clear that this does not perfectly match the data, it would seem reasonable from a theoretical perspective to assume that a process of exponential decay explains this relationship.
+The accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
 
@@
 
-```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using only dates older than 2500 BCE that correspond to the site inventory.
-Negative values have been set to zero.", out.width = "525px", fig.align="center"}
+```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using dates corresponding to the site inventory.
+Negative values have been set to zero.
+The first row only includes dates older than 2500 BCE and the second row only dates older than 4000 BCE.", out.width = "525px", fig.align="center"}
 knitr::include_graphics(here("analysis/figures/results2.png"))
@@
 
-```{r results3, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using only dates older than 4000 BCE that correspond to the site inventory.
-Negative values have been set to zero.", out.width = "525px", fig.align="center"}
-knitr::include_graphics(here("analysis/figures/results3.png"))
-```
+```{r shoredate_sites, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobases and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
+The shoreline dates in gray are plotted against the oldest radiocarbon-informed dates for each site in black.
+The red line indicates the 95% probability range for the shoreline date.", out.width = "525px", fig.align="center"}
 
-```{r shoredate, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites allowing for a sea-level potentially 4m above to 16m below the lowest site elevation.
-These are plotted against the oldest radiocarbon-informed dates for each site.", out.width = "525px", fig.align="center"}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))

### 2022-01-25 analysis/script/04script.R

@@ -1010,7 +1010,7 @@ plot_results <- function(sitename, sitelimit, datedata, sitearea,
 }
 
 
-# Function to shoreline date site, adding specified offsets
+# Function to shoreline date site, adding specified offsets uniformly
 shoreline_date <- function(sitename, dtm = dtm,
                            displacement_curves = displacement_curves,
                            features = rcarb_sa, sites = sites_sa,
@@ -1073,4 +1073,87 @@ shoreline_date <- function(sitename, dtm = dtm,
 
   shoredates <- c("latest_date" = latest, "earliest_date" = earliest)
   return(shoredates)
+}
+
+# Shoreline date, using exponential offset
+shoreline_date_exp <- function(sitename, dtm = dtm,
+                           displacement_curves = displacement_curves,
+                           features = rcarb_sa, sites = sites_sa,
+                           isobases = isobases, sitelimit = TRUE,
+                           expratio){
+  # Retrieve site features
+  siter <- filter(features, site_name == sitename)
+
+  # If the site limit is to be used
+  if(sitelimit == TRUE){
+    sitel <- filter(sites, name == sitename)
+    siteu <- st_union(sitel)
+    sitel <- st_as_sf(cbind(siteu, st_drop_geometry(sitel[1,])))
+  } else {
+    # If not create a convex hull around the dated features
+    sitel <- st_convex_hull(siter)
+    # And assign the values from the limit to the convex hull
+    sitel <- left_join(sitel, st_drop_geometry(filter(sites_sa,
+                name == sitename)), by = c("site_name" = "name", "ask_id"))
+  }
+
+  sitecurve <- interpolate_curve(years = xvals,
+                                 isobase1 = sitel$isobase1,
+                                 isobase2 = sitel$isobase2,
+                                 target = sitel,
+                                 dispdat = displacement_curves,
+                                 isodat = isobases,
+                                 direction_rel_curve1 = sitel$dir_rel_1)
+
+  siteelev <- extract(dtm, vect(sitel), fun = mean)[2]
+
+  probs <- c()
+  prob <- 1
+  i <- 1
+  while(prob > 0.00001) {
+    prob <- (1 * (1 - (expratio/10)))^i
+    probs[i] <- as.numeric(prob)
+    i <- i + 1
+  }
+
+
+  offsets <- seq(1, length(probs), 1)/10
[…]

### 2022-01-25 analysis/script/06script.R

@@ -7,6 +7,7 @@ library(grid)
 library(gridExtra)
 library(gtable)
 library(cowplot)
+library(MASS)
 
 # List all files except those starting with a number.
 # That is, all data files resulting from analysis in 05script.R
@@ -61,9 +62,6 @@ distances <- distances %>%
                           & hordist < 0,  0, hordist),
          topodist = ifelse(sitename %in% c("Gunnarsrød 5", "Pjonkerød R1")
                            & topodist < 0,  0, topodist))
-
-save(distances,
-     file = here::here("analysis/data/derived_data/06data.RData"))
 
 sum_all <- distances %>%
   summarise_at(c("hordist", "topodist", "vertdist"),
@@ -206,11 +204,11 @@ ggsave(file = here("analysis/figures/results.png"), grd,
        bg = "white")
 
 # Measures for vertical distance to inform recommendations for shoreline dating.
-# Using only corresponding dates, excluding dates to agricultural activity
+# Using only corresponding dates, including dates to agricultural activity
 
 # Dates older than 2500
 distances2 <- distances %>%
-  filter(year <= -2500 & rcarb_cor == "t" & type == "forager") %>%
+  filter(year <= -2500 & rcarb_cor == "t") %>%
   # mutate(hordist = ifelse(str_detect(sitename, "Løvås"), 0, hordist),
   #      topodist = ifelse(str_detect(sitename, "Løvås"), 0, topodist),
   #      vertdist = ifelse(str_detect(sitename, "Løvås"), 0, vertdist)) %>%
@@ -218,30 +216,35 @@ distances2 <- distances %>%
          topodist = ifelse(topodist < 0, 0, topodist),
          vertdist = ifelse(vertdist < 0, 0, vertdist))
 
-library(MASS)
-fitdistr(distances2$vertdist, "exponential")
-fitdistr(distances2$vertdist,"logistic")
+# Dates older than 4000 (which also excludes agricultural sites)
+distances3 <- distances %>%
+  filter(year <= -4000 & rcarb_cor == "t" & type == "forager") %>%
+  mutate(hordist = ifelse(hordist < 0, 0, hordist),
+         topodist = ifelse(topodist < 0, 0, topodist),
+         vertdist = ifelse(vertdist < 0, 0, vertdist))
 
-fitexp <- fitdistr(distances2$vertdist, "exponential")
[…]

### 2022-01-25 analysis/script/07script.R

@@ -8,6 +8,7 @@ library(sf)
 source(here("analysis/script/04script.R"))
 load(here("analysis/data/derived_data/01data.RData"))
 load(here("analysis/data/derived_data/02data.RData"))
+load(here("analysis/data/derived_data/06data.RData"))
 
 # Read in raster
 dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
@@ -45,6 +46,85 @@ rdates <- rdates %>%  group_by(site_name) %>%
 sites_sa <- st_join(st_make_valid(sites_sa), isopolys,
                     join = st_intersects, largest = TRUE) %>%
   filter(!(name %in% c("Dybdalshei 2", "Lunaveien", "Frebergsvik C")))
+
+shorelinedates <- list()
+for(i in 1:nrow(sites_sa)){
+  print(sites_sa$name[i])
+  shorelinedates[[i]] <- shoreline_date_exp(sitename = sites_sa$name[i],
+                                            dtm = dtm,
+                                            displacement_curves = displacement_curves,
+                                            features = rcarb_sa,
+                                            sites = sites_sa,
+                                            isobases = isobases,
+                                            sitelimit = TRUE,
+                                            expratio =  expfit$estimate)
+}
+
+sdates <- bind_rows(shorelinedates)
+
+# Radiocarbon dates corresponding to site inventory and older than 2500 BCE
+corsites <- rdates %>%
+  # Excluded, see supplmenetary
+  filter(!(site_name %in% c("Dybdalshei 2", "Lunaveien"))) %>%
+  group_by(site_name) %>%
+  # Exclude Late Neolithic sites
+  filter(n() != sum(is.na(name)) & min(dates) < -2500)
+
+# Retrieve shoreline date data for the same sites
+corshore <- sdates %>%  filter(site_name %in% unique(corsites$site_name))
+
+# Find 95 % probability range for shoreline dates and median shoreline date
+# for ordering in the plot
+hdr <- corshore %>%  group_by(site_name) %>%
+  filter(cumsum(probability) < 0.95) %>%
+  summarise(
+    comb_min = min(combined, na.rm = TRUE),
+    comb_max = max(combined, na.rm = TRUE),
+    comb_median = median(combined, na.rm = TRUE))
+
+# Call to plot
+shrplt <- ggplot(data = hdr,
[…]

### 2022-01-26 analysis/paper/paper.Rmd
Starting to include example sites from Brunlanes

@@
 
+# Shoreline dating sites from Brunlanes
+
+The pre-boreal sites from the Brunlanes-project are among the earliest in Norway.
+They have a clear and distinct Early Mesolithic artefact inventory and are situated in an extremely steep area of the landscape where any use of the sites after the sea retreated would be difficult to envision.
+In the publication of the sites, Jaksland provides a thorough discussion of shoreline dating in general and as used for the dating of the Brunlanes sites specifically.
+The result of comparing his results against the ones achieved using the above-outlined approach are given in Figure, and are largely in agreement.
+The sites are located more or less on the isobase of the Larvik curve and so no adjustment was needed.
+The slight discrepancies can result from the fact that a slightly updated version of the displacement curve is applied here, and the sites are dated according to their mean rather than their lowest elevation.
+Here Jaksland's dates are given a 200 year uncertainty range, provided earliest possible date The sites are located in such a constrained area of the landscape that, as argued by Jaksland, the difference in site elevation would arguably be enough to constrain the uncertainty range so that they don't overlap between sites -- information that is not integrated here.
+The steepness of the curve does nonetheless provide very precise dates, the accuracy of which are of course ultimately dependent on the veracity of the geological reconstruction.
+
 # Discussion

### 2022-01-26 analysis/paper/paper.Rmd
Figuring out how to plot Brunlanes dates

--- 2022-01-25 analysis/paper/paper.Rmd
@@
 This can be reduced to `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m for Mesolithic sites.
-The sites treated here have been given a shoreline date using these values in Figure.
-However,
+
 An exponential function has been fit to the distribution.
@@
 The accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
+The sites treated here have been given a shoreline date using these values in Figure Figure \@ref(fig:shoredate).
 
@@
 
-```{r shoredate_sites, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobases and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
-The shoreline dates in gray are plotted against the oldest radiocarbon-informed dates for each site in black.
+```{r shoredate, eval = TRUE, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
+The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.
 The red line indicates the 95% probability range for the shoreline date.", out.width = "525px", fig.align="center"}
@@
 ```
+
+# Shoreline dating the pioneer sites from Brunlanes
+
+The pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway.
+They have a distinct Early Mesolithic artefact inventory and are situated in an extremely steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location.
+Not least following from the fact that no pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of these earliest sites are of fundamental importance for understanding the pioneer settlement and intial colonisation of the Scandinavian peninsula.
+In the original publication of the sites, Jaksland provides a thorough discussion of shoreline dating in general and as used for the dating of the Brunlanes sites specifically.
+The result of comparing his results against the ones achieved using the above-outlined approach are given in Figure.
+These are largely in agreement.
+The sites are located more or less on the isobase of the displacement curve for Larvik and so neglible adjustment was needed.
+The slight discrepancies can result from the fact that a slightly updated version of the displacement curve is applied here, as well as Jaksland applying the lowest find occurence for the shoreline dating, which is slightly off from the lowest elevation of the site limits employed here.
+Jaksland's dates are a given a 200 and 50 year uncertainty range starting from what he gave as the earliest possible date.
+The sites are located in such a constrained area of the landscape that the difference in site elevation would arguably be enough to constrain the uncertainty range so that they don't overlap between sites -- information that is not integrated here.
+The steepness of the curve does nonetheless provide very precise dates, the accuracy of which are of course ultimately dependent on the veracity of the geological reconstruction.
 

### 2022-01-26 analysis/script/08script.R

@@ -19,31 +19,32 @@ sitespol <- read_sf(
 dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
 
 # Retrieve the Pauler sites
-pauler <- sitespol %>%
-  filter(str_detect(site_name, 'Pauler')) %>%
+brunlanes <- sitespol %>%
+  filter(str_detect(site_name, 'Pauler|Bakke|Sky 1')) %>%
   st_zm() %>%
   rename("name" = "site_name",
          "limit" = "name")
 
-pauler <- st_join(st_make_valid(pauler), isopolys,
+pauler <- st_join(st_make_valid(brunlanes), isopolys,
                     join = st_intersects, largest = TRUE)
 
 sdatesp <- list()
-for(i in 1:nrow(pauler)){
+for(i in 1:nrow(brunlanes)){
   print(pauler$name[i])
-  sdatesp[[i]] <- shoreline_date_exp(sitename = pauler$name[i],
-                                            dtm = dtm,
-                                            displacement_curves = displacement_curves,
-                                            sites = pauler,
-                                            isobases = isobases,
-                                            expratio =  expfit$estimate)
+  sdatesp[[i]] <- shoreline_date_exp(sitename = brunlanes$name[i],
+                                     dtm = dtm,
+                                     displacement_curves = displacement_curves,
+                                     sites = pauler,
+                                     isobases = isobases,
+                                     expratio =  expfit2$estimate,
+                                     siteelev = "min")
 }
 
-pdates <- bind_rows(sdatesp)
+bdates <- bind_rows(sdatesp)
 
 # Find 95 % probability range for shoreline dates and median shoreline date
 # for ordering in the plot
-hdr <- pdates %>%  group_by(site_name) %>%
+hdr <- bdates %>%  group_by(site_name) %>%
   filter(cumsum(probability) < 0.95) %>%
   summarise(
     comb_min = min(combined, na.rm = TRUE),
@@ -52,22 +53,24 @@ hdr <- pdates %>%  group_by(site_name) %>%
 
 # Lasse Jaksland's dates of the Brunlanes sites. He provided a maximum age,
 # and argued that the relative uncertainty would be around 50 years.
-jaksland <- data.frame(site_name = c(sort(unique(pdates$site_name))),
-                       earliest_date = c(-9200, -9150, -9000,
-                                         -8950, -8975, -8850, -8800)) %>%
+jaksland <- data.frame(site_name = sort(unique(bdates$site_name)),
[…]

### 2022-01-27 analysis/paper/paper.Rmd
Minor edits

--- 2022-01-25 analysis/paper/paper.Rmd
@@
 Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g.
-@hughes2016; @stroeven2016], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g.
+@hughes2016; @stroeven2016; see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g.
 @lambeck1998; @svendsen1987].
@@
 Comparatively, given a continuous and still ongoing shoreline regression from as high as c.
-220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @sørensen1979].
+220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@glorstad2020; @hafsten1957; @hafsten1983; @sørensen1979].
 This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.
@@
 The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes.
-This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper.
-associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@solheim2018; @prescott2012].
+This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2018; @prescott2020].
+associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@bakka1971; @prescott2012].
 That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods.
-Earlier, at the transition to the Early Neolithic (c.
-4000 BCE) pottery is introduced to the sites, and there are some possible indications of an initial, small-scale uptake of agriculture at some sites in the Oslo fjord region.
-However, this initial uptake appears to be relatively small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a de-Neolithisation in the Middle Neolithic.
+At an earlier stage, at the transition to the Early Neolithic (c.
+3900 BCE), pottery is introduced to the sites, and there are some possible indications of an initial, small-scale uptake of agriculture at some sites in the Oslo fjord region.
+However, this initial uptake appears to be relatively small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a complete de-Neolithisation in the Middle Neolithic.
 [@nielsen2021] has recently argued that while there, the initial uptake of in the Early Neolithic is combined with a more complex settlement pattern, and argues that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data.
-The empirical expectation for this development would thus be a more varied relation between sites and the shore-line.
+The empirical expectation for this development would thus be a more varied relation between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.
 
@@
 Both of these sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities.
-These agricultural phases are thus treated seperately below.
+Following from the expected deviance from the settlement patterns that characterise forager sites, these agricultural phases are thus treated seperately below.
 Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region.
+Due to the uncertain nature of this assertion these are omitted here.
 
@@
 
-```{r shoredate_sites, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobases and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
-The shoreline dates in gray are plotted against the oldest radiocarbon-informed dates for each site in black.
+```{r shoredate1, eval = FALSE, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
+The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.
 The red line indicates the 95% probability range for the shoreline date.", out.width = "525px", fig.align="center"}
@@
 ```
+
+```{r shoredate, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
+The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.
+The red line indicates the $95%$ probability range for the shoreline date.", out.width = "525px", fig.align="center"}
+knitr::include_graphics(here("analysis/figures/shoredate.png"))
+```
+
+# Shoreline dating the pioneer sites from Brunlanes
+
+The pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a].
+They have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location.
+Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of these earliest sites are of fundamental importance for understanding the pioneer settlement and intial colonisation of the Scandinavian peninsula [e.g.
+@glorstad2016; @bang-andersen2012].
+In the original publication of the sites, Jaksland [@jaksland2014] provides a thorough discussion of shoreline dating in general and as used for the dating of the Brunlanes sites specifically.
+The result of comparing his results against the ones achieved using the above-outlined approach are given in Figure.
+These are largely in agreement.
+The sites are located more or less on the isobase of the displacement curve for Larvik, and neglible adjustments were therefore needed.
+The small discrepancies between the results can instead follow from the fact that a slightly updated version of the displacement curve is applied here.
+Jaksland's dates are given a 200 and 50 year uncertainty range starting from what he gives as the earliest possible date.
+The argument for the uncertainty range of only 50 years is based on the location of the sites in such a constrained and steep area of the landscape that the difference in elevation between the sites establish their relative date and thus constrain the uncertainty range so that they don't overlap between sites located at different elevations.
+This information is not integrated here, but could justify reducing the upper bound of the uncertainty range, especially for Pauler 1 and Pauler 2.
+While the accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the steepness of the curve does nonetheless provide very precise dates, and the agreement with Jakslands thorough consideration speaks to general applicability of the approach to shoreline dating outlined in this study.
+This example also highlights the fact that a consideration of the surrouding terrain can help in reducing the date-span provided by the method, if certain sea-levels can be excluded as unlikely for when a site was in use.
+This makes less of an impact in this setting, where the 95% probability range is already limited, but would considerably improve the precision of the method in cases where the displacement curve is flatter (cf.
+Figure \@ref(fig:shoredate)).
 

### 2022-01-27 analysis/script/08script.R

@@ -63,13 +63,14 @@ ggplot(data = hdr,
                  aes(x = comb_median, y = reorder(site_name, -comb_min))) +
   geom_segment(data = hdr, aes(x = comb_min, xend = comb_max,
                                yend = site_name), col = "red", size = 1) +
-  ggridges::geom_ridgeline(data = pdates,
+  ggridges::geom_ridgeline(data = bdates,
                            aes(x = combined, y = site_name,
                                height = probability*50),
                            colour = "grey", fill = "grey") +
-  geom_linerange(data = jaksland, aes(xmin = earliest_date, xmax = latest_date,
-                                    ymin = site_name, ymax = site_name),
-                                    position = position_dodge2(width = 0.5),
+  geom_segment(data = hdr, aes(x = comb_min, xend = comb_max,
+                               yend = site_name), col = "red", size = 1) +
+  geom_segment(data = jaksland, aes(x = earliest_date, xend = latest_date,
+                                    y = site_name, yend = site_name),
                col = "blue", size = 1) +
   labs(y = "", x = "") +
   theme_bw()

### 2022-01-28 analysis/paper/paper.Rmd
Minor edits

@@
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways.
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes.
-This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2018; @prescott2020].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [but see @mjærum2018; @gundersen2013].
+This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
 associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@bakka1971; @prescott2012].
@@
 [@nielsen2021] has recently argued that while there, the initial uptake of in the Early Neolithic is combined with a more complex settlement pattern, and argues that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data.
-The empirical expectation for this development would thus be a more varied relation between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.
+The empirical expectation for this development would thus be a more varied relation between sites and the shore-line from around 3900 BCE followed by a more decisive shift around 2400 BCE.
 
@@
 
-```{r shoredate1, eval = FALSE, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
+```{r shoredate, echo = FALSE, fig.cap = "The result of shoreline dating  sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
 The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.
-The red line indicates the 95% probability range for the shoreline date.", out.width = "525px", fig.align="center"}
-
-knitr::include_graphics(here("analysis/figures/shoredate.png"))
-```
-
-```{r shoredate, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
-The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.
-The red line indicates the $95%$ probability range for the shoreline date.", out.width = "525px", fig.align="center"}
+The red line indicates the 95\\% probability range for the shoreline date.", out.width = "525px", fig.align="center"}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 They have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location.
-Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of these earliest sites are of fundamental importance for understanding the pioneer settlement and intial colonisation of the Scandinavian peninsula [e.g.
+Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of these earliest sites are of fundamental importance for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
 @glorstad2016; @bang-andersen2012].
-In the original publication of the sites, Jaksland [@jaksland2014] provides a thorough discussion of shoreline dating in general and as used for the dating of the Brunlanes sites specifically.
+In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically.
 The result of comparing his results against the ones achieved using the above-outlined approach are given in Figure.
@@
 The sites are located more or less on the isobase of the displacement curve for Larvik, and neglible adjustments were therefore needed.
-The small discrepancies between the results can instead follow from the fact that a slightly updated version of the displacement curve is applied here.
-Jaksland's dates are given a 200 and 50 year uncertainty range starting from what he gives as the earliest possible date.
-The argument for the uncertainty range of only 50 years is based on the location of the sites in such a constrained and steep area of the landscape that the difference in elevation between the sites establish their relative date and thus constrain the uncertainty range so that they don't overlap between sites located at different elevations.
-This information is not integrated here, but could justify reducing the upper bound of the uncertainty range, especially for Pauler 1 and Pauler 2.
-While the accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the steepness of the curve does nonetheless provide very precise dates, and the agreement with Jakslands thorough consideration speaks to general applicability of the approach to shoreline dating outlined in this study.
-This example also highlights the fact that a consideration of the surrouding terrain can help in reducing the date-span provided by the method, if certain sea-levels can be excluded as unlikely for when a site was in use.
-This makes less of an impact in this setting, where the 95% probability range is already limited, but would considerably improve the precision of the method in cases where the displacement curve is flatter (cf.
+The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here.
+Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date.
+The 200 year uncertainty range is given if the sites were considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other.
+Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty range so that they don't overlap between sites located at different elevations.
+This information is not integrated in the approach outlined here, but could justify reducing the upper bound of the uncertainty ranges, especially for Pauler 1 and Pauler 2.
+While the accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the steepness of the curve does nonetheless provide very precise dates, and the agreement with Jakslands thorough consideration does speak further to the general applicability of the approach to shoreline dating outlined in this study.
+This example also highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
+This makes less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where the displacement curve is flatter (cf.
 Figure \@ref(fig:shoredate)).

### 2022-01-28 analysis/supplementary-materials/supplementary.Rmd
Minor edits

@@
 
-Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan, Rugtvedt [@odgaard1994], Auve [@østmo2008] and Torsrød [@østmo1975].
+Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan, Rugtvedt [@odgaard1994], Tangen [@glorstad2005], Auve??
+[@østmo2008] and Torsrød [@østmo1975].
 These sites are therefore not included in the analysis.

### 2022-01-29 analysis/paper/paper.Rmd
included a couple of references

@@
 Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g.
-@hughes2016; @stroeven2016; see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g.
+@hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g.
 @lambeck1998; @svendsen1987].
@@
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways.
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [but see @mjærum2018; @gundersen2013].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [but see @mjærum2018; @gundersen2013; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 [@nielsen2021] has recently argued that while there, the initial uptake of in the Early Neolithic is combined with a more complex settlement pattern, and argues that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data.
-The empirical expectation for this development would thus be a more varied relation between sites and the shore-line from around 3900 BCE followed by a more decisive shift around 2400 BCE.
+The empirical expectation for the above outlined development would thus be a more varied relation between sites and the shore-line from around 3900 BCE followed by a more decisive shift around 2400 BCE.
 
@@
 This approach has a couple of limitations.
-First of all, the displacement curves are commonly applied directly to larger study areas, with only a couple of studies having taken the variable uplift-rates into account when performing this comparison [e.g.
+First of all, the displacement curves are commonly applied directly to larger study areas, with only a few studies having taken the variable uplift-rates into account when performing this comparison [e.g.
 @fossum2020; @møller1987; @persson2008].
@@
 @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
-This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
+This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b, @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
 
@@
 The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
-The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets, the adjustment to a common isobase, to name but a few [e.g.
-@romundset2011; @romundset2019].
-For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications (see above).
+The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
+@romundset2019; @romundset2011].
+For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.
 
@@
 Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region.
-Due to the uncertain nature of this assertion these are omitted here.
+Due to the uncertain and somewhat speculative nature of thi suggestion, these are omitted here.
 
@@
 This process was repeated 1000 times for each phase for each site.
-The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5.
+The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5, available in the supplementary material.
 Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography.
-At 1000 simulation runs the analysis of each site ranged from taking a few hours to a couple of days, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.
+At 1000 simulation runs the analysis of each site took up to a few hours, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.
 
@@
 Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, most of the sites have been situated close to the shoreline when they were in use.
-When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline, as could be expected.
-To what degree these do in fact represent contamination is discussed further below.
-However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
+When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline.
+However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, as indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
 It is interesting to note the distance for some of the earliest sites, which appears somewhat high.

### 2022-02-01 analysis/paper/paper.Rmd
Set up an initial pass at Brunlanes

@@
 quickSetupOxcal()
+
+# Set up so that I can call floatbarrier using regfloat=TRUE in the chunk 
+# headings. See: 
+# https://gist.github.com/burchill/8873d2ade156b27e92a238a774ce2758
+knitr::knit_hooks$set(plot = function (x, options) {
+  float_correct <- function(f, y, opts)  {
+    if (is.null(opts$regfloat) || opts$regfloat==FALSE)
+      paste0(f(y, opts), "\n\n\\FloatBarrier\n")
+    else
+      f(y, opts)
+  }
+  if (!is.null(options$out.width) || !is.null(options$out.height) ||
+      !is.null(options$out.extra) || options$fig.align != "default" ||
+      !is.null(options$fig.subcap)) {
+    if (is.null(options$fig.scap))
+      options$fig.scap = NA
+    return(float_correct(knitr:::hook_plot_tex, x, options))
+  }
+  return(float_correct(knitr:::hook_plot_md_base, x, options))
+})
 ```
@@
 Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g.
-@berg-hansen2009; @bjerck1990; @mjærum2018; @solheim2018].
+@berg-hansen2009; @bjerck1990; @breivik2014; @solheim2018].
 The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
@@
 2016).
-The study area defined later in the text is marked with a red outline.", out.width = "400px", fig.align="center"}
+The study area defined later in the text is marked with a red outline.", out.width = "400px", fig.align="center", regfloat=TRUE}
 
@@
 If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if it would have been located on an island when the sea-level was higher, it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft.
-Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few, and are typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore-bound site location is largely based on a mosaic of smaller scale investigations.
+Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few, and are typically conducted on a site-by-site or project-by-project basis, meaning the basis for assuming shore-bound site location is largely a mosaic of smaller scale investigations.
 
@@
 # Data
-
-All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a digital research compendium following @marwick2018.
-The dataset is slightly more extensive than what is directly utilised in the study, and includes excavated sites without ^14^C-dates within the study area, as well as ^14^C-dates to periods other than the Stone Age.
-
-To get at the relationship between sites and the contemporaneous shoreline, the analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
+To get at the relationship between sites and the contemporaneous shoreline, this analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
 While there is displacement data available for other areas of south-eastern Norway [e.g.
@@
 
+All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a digital research compendium following @marwick2018.
+The dataset is slightly more extensive than what is directly utilised in the study, and includes excavated sites without ^14^C-dates within the study area, as well as ^14^C-dates to periods other than the Stone Age.
+
 ```{r map-iso,  fig.cap = paste("A) Distribution of the ", nrow(sites_sa), " analysed sites relative to the isobases of the displacement curves.
@@
 2018), B) Displacement curves.
-Note the increasing steepness of the curves towards the north-east."), fig.height = 13}
+Note the increasing steepness of the curves towards the north-east."), fig.height = 13, regfloat=TRUE}
 bboxsites <- st_bbox(sites_sa)
@@
 C) The location of the site within the study area relative to isobases of the employed displacement curves.
-D) Displacement curve interpolated to the site location based on the distance between the site and the isobases of the two closest displacement curves.", out.width = "500px", fig.align="center"}
+D) Displacement curve interpolated to the site location based on the distance between the site and the isobases of the two closest displacement curves.", out.width = "500px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/example_map.png"))
@@
     legend.position = "none",
-    axis.line = element_line("black"),
     axis.title.y = element_blank(),
@@
 C) The sea-level is then adjusted on the DTM to this elevation and the various distance mesaures are found.
-D) The numerical result of the simulation run.", out.width = "500px", fig.align="center"}
+D) The numerical result of the simulation run.", out.width = "500px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/example_sim_map.png"))
@@
 The more opaque the colour, the more times the sea-level was simulated to that location.
-The third column shows violin plots of the different distance measures across all simulations.", out.width = "500px"}
+The third column shows violin plots of the different distance measures across all simulations.", out.width = "500px", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/example_sim_result.png"))
@@
 Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations.
-The best point estimate for the distance between site and shoreline for the pre-Late Neolithic is, following from the skewed distribution, given by the median at `r round(median(distances2$vertdist))`m.
+The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is, following from the skewed distribution, given by the median at `r round(median(distances2$vertdist))`m.
 However, 95% of the distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
-That is, for 95% of the cases, the shoreline was situated on or down to m below the site location.
-This is constrained further when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m.
-When applying the method of shoreline dating to Stone Age sites in the region, it could thus be recommended, given no other information, to allow for a potential deviation between site and shoreline from `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))` in vertical distance.
-This can be reduced to `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m for Mesolithic sites.
-
-An exponential function has been fit to the distribution.
+That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
+This is constrained slightly when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m.
+An exponential function has been fit to the distributions for vertical distance.
 While it is clear that this does not perfectly match the data, it would seem reasonable from a theoretical perspective to assume that a process of exponential decay explains this relationship.
-The accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
-The sites treated here have been given a shoreline date using these values in Figure Figure \@ref(fig:shoredate).
+Furthermore, the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
+Using the resulting exponential decay function, the sites treated here have been given a shoreline date in Figure \@ref(fig:shoredate).
 
@@
 The second row shows the result of excluding these.
-The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center"}
+The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/results.png"))
@@
 Negative values have been set to zero.
-The first row only includes dates older than 2500 BCE and the second row only dates older than 4000 BCE.", out.width = "525px", fig.align="center"}
+The first row only includes dates older than 2500 BCE and the second row only dates older than 4000 BCE.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/results2.png"))
@@
 The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.
-The red line indicates the 95\\% probability range for the shoreline date.", out.width = "525px", fig.align="center"}
+The red line indicates the 95\\% probability range for the shoreline date.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 
-The pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a].
-They have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location.
-Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of these earliest sites are of fundamental importance for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
-@glorstad2016; @bang-andersen2012].
+Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
+@bang-andersen2012; @breivik2014; @glorstad2016].
+The shoreline dated pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a].
+These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location.
 In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically.
-The result of comparing his results against the ones achieved using the above-outlined approach are given in Figure.
+A comparison of his results and the ones achieved using the above-outlined approach are given in Figure.
 These are largely in agreement.
-The sites are located more or less on the isobase of the displacement curve for Larvik, and neglible adjustments were therefore needed.
-The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here.
+The sites are located more or less on the isobase of the displacement curve for Larvik, and negligible adjustments were therefore needed.
+The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf.
+@sorenseninpress].
 Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date.
-The 200 year uncertainty range is given if the sites were considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other.
+The 200 year uncertainty range is given if the sites are considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other.
 Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty range so that they don't overlap between sites located at different elevations.
@@
 This example also highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
-This makes less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where the displacement curve is flatter (cf.
+One approach to this could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure, followed for example by a visual evaluation of the topography or by measuring the steepness of the slope to the shoreline.
+Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline.
+Such approaches would make less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where the displacement curve is flatter (cf.
 Figure \@ref(fig:shoredate)).
+
+```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating the Brunlanes sites.
+A) The result of applying the approach to shoreline dating outlined here, plotted against the dates provided by Jaksland (2014) given in black.
+B) Map of the Pauler sites and Sky 1 with simulated sea-level using the probability density associated with the shoreline date for Pauler 1.", out.width = "525px", fig.align="center", regfloat=TRUE}
+knitr::include_graphics(here("analysis/figures/brunlanes.png"))
+```
 

### 2022-02-01 analysis/script/04script.R

@@ -439,7 +439,8 @@ bboxpoly <- function(feature, xy_adjustment) {
 # Define function to simulate shoreline and measure distances given a single
 # radiocarbon probability density function, site limit, and adjusted
 # displacement curve
-sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
+sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs,
+                             paths = FALSE){
 
   results <- data.frame(matrix(ncol = 6, nrow = samps))
   names(results) <- c("vertdist", "hordist", "topodist",
@@ -447,7 +448,11 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
   results$rcarb_cor <- unique(na.omit(posteriorprobs$rcarb_cor))
   results$sitename <- unique(sitel$name)
   seapolygons <- list(length = samps)
-  # topopaths <- list(length = samps)
+
+  # If topographic paths are to be retrieved
+  if(paths == TRUE){
+    topopaths <- list()
+  }
 
   for(i in 1:samps){
     # Set up sampling frame from sum of posterior densities
@@ -503,6 +508,12 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
       # Find topographic distance (retrieve path)
       topodist <- topoDist(raster::raster(sitearea), coords,
                            paths = TRUE)
+
+      if(paths == TRUE){
+        topopaths[i] <- topodist[[2]]
+      }
+
+
       # Store distance (negative to indicate below sea-level)
       results$topodist[i] <- topodist[[1]]["loc", "sea"] * -1
 
@@ -551,6 +562,11 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
       # Find topographic distance (retrieve path)
       topodist <- topoDist(raster::raster(sitearea), coords,
                            paths = TRUE)
+
+      if(paths == TRUE){
+        topopaths[i] <- topodist[[2]]
+      }
+
       # Store distance
       results$topodist[i] <- topodist[[1]]["loc", "sea"]
 
@@ -574,11 +590,18 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs){
   }
   }
 
-  return(list(
-    results = results,
-    seapol = st_as_sf(st_sfc(do.call(rbind, seapolygons)),
[…]

### 2022-02-01 analysis/script/07script.R

@@ -1,5 +1,5 @@
 library(ggplot2)
-library(dplyr)
+library(tidyverse)
 library(here)
 library(terra)
 library(sf)
@@ -103,101 +103,14 @@ ggsave(file = here("analysis/figures/shoredate.png"), shrplt,
        width = 200, height = 200, units = "mm")
 
 
+# Example site for development
 
-
-
-
-
-
-  # Call to plot
-  ggplot(aes(x = combined, y = site_name)) +
-  ggridges::geom_ridgeline(aes(x = dates, height = probability), colour = "grey", fill = "grey") +
-  ggridges::geom_ridgeline(aes(x = dates, height = probabilities), colour = "red", fill = "white") +
-  labs(x = "cal BCE/CE", y = "") +
-  scale_x_continuous(breaks = seq(-9000, 1000, by = 2000)) +
-  theme_bw() +
-  theme(legend.position = "None")
-
-datedata %>%  ggplot(aes(x = median_date, y = reorder(site_name, -median_date))) +
-  ggridges::geom_ridgeline(aes(x = dates, height = probability), colour = "black", fill = "white")
-
-
-
-
-
-
-shorelinedates <- list()
-for(i in 1:nrow(sites_sa)){
-  print(sites_sa$name[i])
-  sdates <- shoreline_date(sitename = sites_sa$name[i],
-                 dtm = dtm,
-                 displacement_curves = displacement_curves,
-                 features = rcarb_sa,
-                 sites = sites_sa,
-                 isobases = isobases,
-                 sitelimit = TRUE,
-                 xvals = xvals,
-                 offset = c(16, 4))
-  shorelinedates[[i]] <- c("site_name" = sites_sa$name[i], sdates)
-}
-
-sdates <- shorelinedates %>% bind_rows() %>%
-  mutate(date_range = as.numeric(latest_date) - as.numeric(earliest_date)) %>%
-  rowwise() %>%
-  mutate(median_date = median(c(as.numeric(earliest_date), as.numeric(latest_date))))
-
-datedata <- full_join(rdates, sdates, by = "site_name")
-
-shrplt <- datedata %>%
-  # Excluded, see supplmenetary
-  filter(!(site_name %in% c("Dybdalshei 2", "Lunaveien"))) %>%
-  group_by(site_name) %>%
-  # Exclude Late Neolithic sites
-  filter(n() != sum(is.na(name)) & min(dates) < -2500) %>%
-  # Call to plot
-ggplot(aes(x = median_date, y = reorder(site_name, -median_date))) +
[…]

### 2022-02-01 analysis/paper/paper.Rmd
Wokring on Brunlanes sites

@@
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways.
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [but see @mjærum2018; @gundersen2013; @schulke2020].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brø but see @mjærum2018; @gundersen2013; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 At an earlier stage, at the transition to the Early Neolithic (c.
-3900 BCE), pottery is introduced to the sites, and there are some possible indications of an initial, small-scale uptake of agriculture at some sites in the Oslo fjord region.
-However, this initial uptake appears to be relatively small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a complete de-Neolithisation in the Middle Neolithic.
-[@nielsen2021] has recently argued that while there, the initial uptake of in the Early Neolithic is combined with a more complex settlement pattern, and argues that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data.
-The empirical expectation for the above outlined development would thus be a more varied relation between sites and the shore-line from around 3900 BCE followed by a more decisive shift around 2400 BCE.
+3900 BCE), pottery is introduced to the sites, and there are some possible indications of an initial uptake of agriculture at some sites in the Oslo fjord region.
+However, this appears to be small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955].
+[@nielsen2021b] has recently argued that while there, the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data.
+Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a more varied association between sites and the shore-line from around 3900 BCE, followed by a more decisive shift around 2400 BCE.
 
@@
 One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 ^14^C-dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
-They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
+They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
 However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates.
@@
 
-Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
+Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
 @åstveit2018; @solheim2020].
@@
 Three of the sites have been associated with farming activities, either in directly or in the form building structures.
-The first is Nordby 1, which has .
-Furthermore, the Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are both related to farming.
-Both of these sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities.
-Following from the expected deviance from the settlement patterns that characterise forager sites, these agricultural phases are thus treated seperately below.
-Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region.
-Due to the uncertain and somewhat speculative nature of thi suggestion, these are omitted here.
+The first is Nordby 1, at which the ^14^C-dates are associated with a Late Neolithic long-house.
+The Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are both related to farming.
+Both of these sites also have radiocarbon dates and lithic inventory associated with Mesolithic forager activities.
+Following from the expected deviance from the settlement patterns that are to characterise forager sites, these agricultural phases are thus treated separately below.
+Finally, @nielsen2021b has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratsberg and Larønningen could be related to early agricultural activity in the Oslo fjord region.
+Due to the uncertain and somewhat speculative nature of this suggestion, these are omitted here.
 
@@
 The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2019].
-After evaluating the options I opted for the 10m resolution DTM rather than the higher-resolution 1m version.
+It was here opted for the 10m resolution DTM rather than the higher-resolution 1m version.
 In addition to resulting in considerably less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by.
-The 10m DTM is a down-sampled version of the 1m resolution DTM, which is based on aerial laser scanning using a minimum of two and up to five points per m^2^ [@mapping2019].
+The 10m DTM is a down-sampled version of the 1m resolution DTM, which is based on aerial laser scanning using a minimum of two and up to five points per m^2^ and has a vertical error of  [@mapping2019].
 
@@
 In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically.
-A comparison of his results and the ones achieved using the above-outlined approach are given in Figure.
+A comparison of his results and the ones achieved using the above-outlined approach are given in Figure Figure \@ref(fig:brunlanes)A.
 These are largely in agreement.
@@
 Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date.
-The 200 year uncertainty range is given if the sites are considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other.
-Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty range so that they don't overlap between sites located at different elevations.
-This information is not integrated in the approach outlined here, but could justify reducing the upper bound of the uncertainty ranges, especially for Pauler 1 and Pauler 2.
-While the accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the steepness of the curve does nonetheless provide very precise dates, and the agreement with Jakslands thorough consideration does speak further to the general applicability of the approach to shoreline dating outlined in this study.
+The 200 year uncertainty range is given if the sites were to be considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other.
+Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap.
+This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges, especially for Pauler 1 and Pauler 2.
+While the accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the steepness of the curve does nonetheless provide very precise dates.
 This example also highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
-One approach to this could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure, followed for example by a visual evaluation of the topography or by measuring the steepness of the slope to the shoreline.
+One approach to this could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by measuring the steepness of the slope to the shoreline.
 Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline.
@@
 
-```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating the Brunlanes sites.
-A) The result of applying the approach to shoreline dating outlined here, plotted against the dates provided by Jaksland (2014) given in black.
-B) Map of the Pauler sites and Sky 1 with simulated sea-level using the probability density associated with the shoreline date for Pauler 1.", out.width = "525px", fig.align="center", regfloat=TRUE}
+```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating of the Brunlanes sites using elevations provided by Jaksland (2014:tab.4).
+A) The result of applying the approach to shoreline dating outlined here.
+Dates provided by Jaksland (2014) are plotted in black starting from the earliest possible date and where the box indicates a 50 year uncertainty range, and the line 200 years.
+B) Map of the Pauler sites and Sky 1.
+The sea-level has been simulated using the probability density associated with the shoreline date for Pauler 1 given as the yellow point (see also map in Jaksland 2014:fig.12a).", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/brunlanes.png"))

### 2022-02-02 analysis/paper/paper.Rmd
Unsure how to combine the probability density functions

@@
 Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.
-From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, they appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
-This corresponds well with the literature, where the introduction of the Neolithic proper is seen as occurring around this time [@prescott2020; @solheim2021].
-With the transition from a predominantly hunter-gatherer lifeway to early agro-pastoralism, settlements in coastal areas are to be increasingly withdrawn from the shoreline.
+From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
+This corresponds well with the literature and corresponds with the Neolithisation process.
 
@@
 
-Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE.
+Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE.
 Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations.
-The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is, following from the skewed distribution, given by the median at `r round(median(distances2$vertdist))`m.
+The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, following from the skewed distribution.
 However, 95% of the distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
@@
 
-```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating of the Brunlanes sites using elevations provided by Jaksland (2014:tab.4).
-A) The result of applying the approach to shoreline dating outlined here.
-Dates provided by Jaksland (2014) are plotted in black starting from the earliest possible date and where the box indicates a 50 year uncertainty range, and the line 200 years.
+```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating of the Brunlanes sites using site altitudes provided by Jaksland (2014:tab.4).
+A) The result of applying the approach to shoreline dating outlined above.
+Dates provided by Jaksland (2014) are plotted in black.
+The box indicates a 50 year uncertainty range which combined with the line indicates 200 years.
+Follwing Jaksland's approach these start from the earliest possible date.
 B) Map of the Pauler sites and Sky 1.
-The sea-level has been simulated using the probability density associated with the shoreline date for Pauler 1 given as the yellow point (see also map in Jaksland 2014:fig.12a).", out.width = "525px", fig.align="center", regfloat=TRUE}
+The sea-level has been simulated using the probability density associated with the shoreline date for Pauler 1 (see also map in Jaksland 2014:fig.12a).
+Pauler 1 is the yellow point.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/brunlanes.png"))

### 2022-02-02 analysis/script/04script.R

@@ -1122,9 +1122,9 @@ shoreline_date_exp <- function(sitename, dtm = dtm,
     siteelev <- specified_elev
   } else{
     if(siteelev == "mean") {
-      siteelev <- extract(dtm, vect(sitel), fun = mean)[2]
+      siteelev <- terra::extract(dtm, vect(sitel), fun = mean)[2]
     } else if(siteelev == "min"){
-      siteelev <- extract(dtm, vect(sitel), fun = min)[2]
+      siteelev <- terra::extract(dtm, vect(sitel), fun = min)[2]
     }
   }
 
@@ -1171,10 +1171,22 @@ shoreline_date_exp <- function(sitename, dtm = dtm,
 
     dates[i, 1:2] <- cbind(earliest, latest)
   }
+
+  # dates <- dates %>%
+  #   mutate(combined = (earliest_date + latest_date)/2,
+  #          probability = probability/sum(probability))
+
+  dates <- dates %>% group_by(r = row_number()) %>%
+    mutate(date_range = list(earliest_date:latest_date)) %>%
+    ungroup %>% dplyr::select(-r) %>%
+    # Unnest the list column to get the desired "long" data frame
+    unnest(date_range) %>%
+    mutate(probability = probability/sum(probability)) %>%
+    dplyr::select(-earliest_date, -latest_date)
+
+  # Make infinite values NA
+  is.na(dates) <- do.call(cbind,lapply(dates, is.infinite))
   dates$site_name <- sitename
 
-  dates <- dates %>%
-    mutate(combined = (earliest_date + latest_date)/2,
-           probability = probability/sum(probability))
   return(dates)
 }

### 2022-02-02 analysis/script/07script.R

@@ -192,53 +192,95 @@ ggplot(aes(y = probability)) +
                            colour = "red", fill = NA, alpha = 0.3) +
   labs(y = "Density", x = "BCE")
 
+is.na(dates) <- do.call(cbind,lapply(dates, is.infinite))
+yrs <- seq(-10500, 1950, 1)
+agegrid <- data.frame(year = yrs,
+                      earliest_prob = numeric(length(yrs)),
+                      latest_prob = numeric(length(yrs)))
 
-dat <- pivot_longer(dates,
-                    cols = c(earliest_date, latest_date), names_to = "year") %>%
-  # filter(!duplicated(value)) %>%
+erange <- seq(min(dates$earliest_date, na.rm = TRUE), max(dates$earliest_date, na.rm = TRUE), 1)
+efirst <- dates[match(unique(dates$earliest_date), dates$earliest_date),]
+
+which(erange == -8200)
+which(agegrid$year == -8200)
+
+for (i in 1:length(erange)) {
+  if(erange[i] %in% efirst$earliest_date) {
+    yr <- efirst$earliest_date[which(efirst$earliest_date == erange[i])]
+    agegrid[which(agegrid$year == yr),
+            "earliest_prob"] <- efirst[which(efirst$earliest_date == yr),"probability"]
+    j <- 1
+  } else {
+    yr <- efirst$earliest_date[which(efirst$earliest_date == erange[i - j])]
+    agegrid[which(agegrid$year == yr + j),
+            "earliest_prob"] <- efirst[which(efirst$earliest_date == yr),"probability"]
+    j <- j + 1
+  }
+}
+
+lrange <- seq(min(dates$latest_date), max(dates$latest_date), 1)
+lfirst <- dates[match(unique(dates$latest_date), dates$latest_date),]
+
+for (i in 1:length(lrange)) {
+  if(lrange[i] %in% lfirst$latest_date) {
+    yr <- lfirst$latest_date[which(lfirst$latest_date == lrange[i])]
+    agegrid[which(agegrid$year == yr),
+            "latest_prob"] <- lfirst[which(lfirst$latest_date == yr),"probability"]
+    j <- 1
+  } else {
+    yr <- lfirst$latest_date[which(lfirst$latest_date == lrange[i - j])]
+    agegrid[which(agegrid$year == yr + j),
+            "latest_prob"] <- lfirst[which(lfirst$latest_date == yr),"probability"]
[…]

### 2022-02-02 analysis/paper/paper.Rmd
Exponential function has been successfully integrated with the displacement curve (as far as I can tell)

@@
 @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
-This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b, @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
+This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
 
@@
 Not finding the topographically closest point significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact on the results, given the distances considered.
-The shortest topographic path was found using the Queen's pattern neighbourhood of eight cells [@conolly2006].
+The shortest topographic path was found using the Moore neighbourhood of eight cells [e.g.
+@conolly2006, 253; @herzog2013].
 In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
@@
 The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, following from the skewed distribution.
-However, 95% of the distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
+95% of the distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
 That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
@@
 Dates provided by Jaksland (2014) are plotted in black.
-The box indicates a 50 year uncertainty range which combined with the line indicates 200 years.
+The box indicates a 50 year uncertainty range which in combination with the black line extends 200 years.
 Follwing Jaksland's approach these start from the earliest possible date.

### 2022-02-03 analysis/paper/paper.Rmd
Shoreline date all sites without radiocarbon dates

@@
       email: isak.roalkvam@iakh.uio.no
-      institute: [UofO]
+      institute: [UiO]
       correspondence: true
 institute:
-  - UofO: University of Oslo
+  - UofO: University of Oslo, Institute of Archaeology, Conservation and History
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 @berg-hansen2009; @bjerck1990; @breivik2014; @solheim2018].
-The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available.
-Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites.
+The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are available.
+However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites.
 Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age.
-This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline.
+This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline.
 Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.
 
-Despite its fundamental role for Norwegian Stone Age archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods.
+Despite its important role for Norwegian Stone Age archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods.
 The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach.
@@
 
-```{r shoredate, echo = FALSE, fig.cap = "The result of shoreline dating  sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
+```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory.
+The sites are dated using the mean elevation of the site polygon, accounting for distance to closest isobase and by employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
 The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.
@@
 Follwing Jaksland's approach these start from the earliest possible date.
-B) Map of the Pauler sites and Sky 1.
+B) Map showing the centroids of the Pauler sites and Sky 1.
 The sea-level has been simulated using the probability density associated with the shoreline date for Pauler 1 (see also map in Jaksland 2014:fig.12a).
-Pauler 1 is the yellow point.", out.width = "525px", fig.align="center", regfloat=TRUE}
+Pauler 1 is the red point.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/brunlanes.png"))
@@
 
-While allowing for a deviation up to 16m might be recommended when no other information is available, this can likely be considerably reduced by evaluating the surroundings of the site.
+When no other information is available it can be recommended to apply the method for shoreline dating outlined above, which involves both taking the distance between sites and the isobase of displacement curves into consideration, and implementing to account for the distance between the sites and the shoreline.
+The precision of this method is, as shown above, both geographically and temporally contingent where older sites towards the north-east will get a more precise date than younger sites located towards the south-west.
+This approach will be useful either when shoreline dating a large number of sites where no other information is available [e.g.
+@solheim2018; @fossum2020; @nielsen2021b; @breivik2014], which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.
+
+Other information is of course typically available.
 

### 2022-02-03 analysis/supplementary-materials/supplementary.Rmd
Adding previous shoreline dates

@@
 
+```{r brunstad24, echo = FALSE, out.width = "500px", regfloat=TRUE}
+knitr::include_graphics(here("analysis/figures/brunstad24.png"))
+
+filter(rcarbon, site_name  == "Brunstad 24") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Brunstad 24", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
+```{r brunstad25, echo = FALSE, out.width = "500px", regfloat=TRUE}
+knitr::include_graphics(here("analysis/figures/brunstad25.png"))
+
+filter(rcarbon, site_name  == "Brunstad 25") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Brunstad 25", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
+
 ```{r dordal, echo = FALSE, out.width = "500px", regfloat=TRUE}

### 2022-02-03 analysis/script/08script.R

@@ -281,6 +281,13 @@ for(i in 1:nrow(sites_sl)){
 bdates <- bind_rows(sitdates)
 
 
+psdates <- read.csv((here("analysis/data/raw_data/previous_shoreline_dates.csv"))) %>%
+  mutate(start = start * -1,
+         end = end * -1) %>%
+  rename(site_name = name)
+
+bdates <- bdates %>%filter(site_name %in% psdates$site_name)
+
 # Find 95 % probability range for shoreline dates and median shoreline date
 # for ordering in the plot
 hdr <- bdates %>%  group_by(site_name) %>%
@@ -290,8 +297,18 @@ hdr <- bdates %>%  group_by(site_name) %>%
     year_max = max(year, na.rm = TRUE),
     year_median = median(year, na.rm = TRUE))
 
+hdr <- hdr[order(hdr$year_median, decreasing = TRUE),]
+hdr$site_name <- factor(hdr$site_name, levels = hdr$site_name)
+
+psdates <- psdates[order(match(psdates$site_name, hdr$site_name)),]
+psdates$site_name <- factor(psdates$site_name, levels = psdates$site_name)
+
+psdates1 <- psdates %>%  filter(start != end)
+psdates2 <- psdates %>%  filter(start == end)
+
+
 # Call to plot
-ggplot(data = hdr, aes(x = year_median, y = reorder(site_name, -year_median))) +
+ggplot(data = hdr, aes(x = year_median, y = site_name)) +
   geom_segment(data = hdr, aes(x = year_min, xend = year_max,
                                yend = site_name), col = "red") +
   ggridges::geom_ridgeline(data = bdates,
@@ -300,9 +317,11 @@ ggplot(data = hdr, aes(x = year_median, y = reorder(site_name, -year_median))) +
                            colour = "grey", fill = "grey") +
   geom_segment(data = hdr, aes(x = year_min, xend = year_max,
                                yend = site_name), col = "red") +
-  # geom_segment(data = jaksland, aes(x = earliest_date, xend = latest_date,
-  #                                   y = site_name, yend = site_name),
-  #              col = "black", size = 5) +
+  geom_linerange(data = psdates1, aes(xmin = start, xmax = end,
+                                      y = site_name), size = 1,
+                 position = position_dodge(width = 0.3, preserve = 'single'),
+                 inherit.aes = FALSE) +
+  geom_point(data = psdates2, aes(x = start), col = "black") +
   labs(y = "", x = "BCE") +
   theme_bw()
 

### 2022-02-04 analysis/paper/paper.Rmd
Re-dating previously shoreline dated sites

@@
 
+# Shoreline dating
+
+```{r redate, echo = FALSE, fig.cap = "Shoreline dating sites from the study area without radiocarbon dates or with radiocarbon dates that do not corresponding to the site inventories.
+These are comapred to the shoreline dates originally proposed by the excavation reports in black.", out.width = "525px", fig.align="center", regfloat=TRUE}
+knitr::include_graphics(here("analysis/figures/redate.png"))
+```
+
 # Discussion
@@
 When no other information is available it can be recommended to apply the method for shoreline dating outlined above, which involves both taking the distance between sites and the isobase of displacement curves into consideration, and implementing to account for the distance between the sites and the shoreline.
-The precision of this method is, as shown above, both geographically and temporally contingent where older sites towards the north-east will get a more precise date than younger sites located towards the south-west.
+The precision of this method is, as shown above, both geographically and temporally contingent, where older sites situated towards the north-east will get a more precise date than younger sites located towards the south-west.
 This approach will be useful either when shoreline dating a large number of sites where no other information is available [e.g.
 @solheim2018; @fossum2020; @nielsen2021b; @breivik2014], which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.
+
+The outlined approach is thus derived from the available radiocarbon record, and appears to be able to accommodate the available data.
+To further test the method will require independently derived radiocarbon dates.
 

### 2022-02-04 analysis/script/04script.R

@@ -1079,15 +1079,14 @@ shoreline_date_exp <- function(sitename, elev = dtm,
   for(i in 1:length(offsets)){
     negative_offset <- as.numeric(siteelev - offsets[i])
     if(!(negative_offset > 0)) {
+      # Make sure the sea level can not be lower than the present
       negative_offset <- 0.01
     }
     positive_offset <- as.numeric(siteelev + offsets[i])
 
-    # Find lower date, subtracting offset (defaults to 0)
     lowerd1 <- round(approx(sitecurve[,"lowerelev"],
                             xvals, xout = negative_offset)[['y']])
 
-    # Find upper date, subtracting offset (defaults to 0)
     upperd1 <- round(approx(sitecurve[,"upperelev"],
                             xvals, xout =  negative_offset)[['y']])
 

### 2022-02-04 analysis/script/08script.R

@@ -5,6 +5,7 @@ library(ggnewscale)
 library(dplyr)
 library(magrittr)
 library(stringr)
+library(tidyr)
 library(here)
 library(raster)
 library(terra)
@@ -256,13 +257,21 @@ ggsave(file = here("analysis/figures/brunlanes2.png"), width = 1325*2,
        units = "px")
 
 
-
+psdates <- read.csv((here("analysis/data/raw_data/previous_shoreline_dates.csv"))) %>%
+  mutate(start = start * -1,
+         end = end * -1) %>%
+  rename(site_name = name)
 
 sites_sl <- site_limits %>% filter(radiocarbon == "f" | radiocarbon == "t" & rcarb_cor == "f")
 
 sites_sl <- st_join(st_make_valid(sites_sl), isopolys,
                     join = st_intersects, largest = TRUE) %>%
-  filter(!is.na(dir_rel_1))
+  filter(!is.na(dir_rel_1)) %>%
+  filter(name %in% psdates$site_name)
+
+sites_sl <- inner_join(sites_sl,
+                       dplyr::select(psdates, site_name, min_elev, max_elev),
+                       by = c("name" = "site_name"))
 
 sitdates <- list()
 for(i in 1:nrow(sites_sl)){
@@ -273,20 +282,10 @@ for(i in 1:nrow(sites_sl)){
                                      sites = sites_sl,
                                      iso = isobases,
                                      expratio = expfit$estimate,
-                                     siteelev = "mean")
-
-
+                                     specified_elev = sites_sl$min_elev[i])
 }
 
 bdates <- bind_rows(sitdates)
-
-
-psdates <- read.csv((here("analysis/data/raw_data/previous_shoreline_dates.csv"))) %>%
-  mutate(start = start * -1,
-         end = end * -1) %>%
-  rename(site_name = name)
-
-bdates <- bdates %>%filter(site_name %in% psdates$site_name)
 
 # Find 95 % probability range for shoreline dates and median shoreline date
 # for ordering in the plot
@@ -308,20 +307,24 @@ psdates2 <- psdates %>%  filter(start == end)
 
 
 # Call to plot
-ggplot(data = hdr, aes(x = year_median, y = site_name)) +
+redateplt <- ggplot(data = hdr, aes(x = year_median, y = site_name)) +
   geom_segment(data = hdr, aes(x = year_min, xend = year_max,
                                yend = site_name), col = "red") +
-  ggridges::geom_ridgeline(data = bdates,
[…]

### 2022-02-07 analysis/paper/paper.Rmd
Solving the integration of the exponential decay function somewhat differently

@@
 ```
-# Results
+# Simulation results
 
@@
 The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, following from the skewed distribution.
-95% of the distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
+95% of the vertical distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
 That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
@@
 Results associated with agricultural acitivites are in grey.
-The first row shows the result of including all dates the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites.
+The first row shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites.
 The second row shows the result of excluding these.
@@
 Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap.
-This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges, especially for Pauler 1 and Pauler 2.
-While the accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the steepness of the curve does nonetheless provide very precise dates.
+This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges.
+The steepness of the curve does nonetheless provide very precise dates, although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction.
 This example also highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
@@
 Figure \@ref(fig:shoredate)).
+To further illustrate this point, excavated Stone Age sites in the region which have previously been dated with reference to relative sea-level change are subjected to the outlined approach below.
 
@@
 
-```{r redate, echo = FALSE, fig.cap = "Shoreline dating sites from the study area without radiocarbon dates or with radiocarbon dates that do not corresponding to the site inventories.
-These are comapred to the shoreline dates originally proposed by the excavation reports in black.", out.width = "525px", fig.align="center", regfloat=TRUE}
+```{r redate, echo = FALSE, fig.cap = "95% probability ranges of shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
+These are comapred to the dates originally proposed by the excavation reports in black.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/redate.png"))

### 2022-02-07 analysis/script/07script.R

@@ -142,6 +142,19 @@ siteelev <- terra::extract(dtm, vect(sitel), fun = min)[2]
 # a = original amount before decay (1 here)
 # x = time (distance here)
 
+# 1/200 x sannsynlighet
+inc <- seq(0, 70, 0.1)
+
+expdat <- data.frame(
+  offset = inc,
+  px = pexp(inc, rate = expfit$estimate))
+
+plot(y = expdat$px, x = inc)
+
+expdat <- expdat %>%
+  mutate(prob = px - lag(px, default =  first(px))) %>%
+  tail(-1)
+
 probs <- c()
 prob <- 1
 i <- 1
@@ -151,18 +164,15 @@ while(prob > 0.00001) {
   i <- i + 1
 }
 
-
-offsets <- seq(1, length(probs), 1)/10
 dates <- data.frame(matrix(ncol = 3, nrow = length(probs)))
 names(dates) <- c("earliest_date", "latest_date", "probability")
 dates$probability <- probs
 
-for(i in 1:length(offsets)){
-  negative_offset <- as.numeric(siteelev - offsets[i])
+for(i in 1:nrow(expdat)){
+  negative_offset <- as.numeric(siteelev - expdat$offset[i])
   if(!(negative_offset > 0)) {
     negative_offset <- 0.01
   }
-  positive_offset <- as.numeric(siteelev + offsets[i])
 
   # Find lower date, subtracting offset (defaults to 0)
   lowerd1 <- round(approx(sitecurve[,"lowerelev"],
@@ -172,17 +182,15 @@ for(i in 1:length(offsets)){
   upperd1 <- round(approx(sitecurve[,"upperelev"],
                           xvals, xout =  negative_offset)[['y']])
 
-  lowerd2 <- round(approx(sitecurve[,"lowerelev"],
-                          xvals, xout = positive_offset)[['y']])
-
-  upperd2 <- round(approx(sitecurve[,"upperelev"],
-                          xvals, xout =  positive_offset)[['y']])
-
   # Find youngest and oldest date
   earliest <- min(c(lowerd1, upperd1), na.rm = TRUE)
   latest <- max(c(lowerd1, upperd1), na.rm = TRUE)
 
-  yrs <- data.frame(year = seq(earliest, latest, 1), prob = probs[i])
+  # Assign probability to each year in range
+  yrs <- data.frame(year = seq(earliest, latest, 1)) %>%
[…]

### 2022-02-07 analysis/paper/paper.Rmd
Using IDW for the interpolation of the displacement curves. Re-running analyses

@@
 institute:
-  - UofO: University of Oslo, Institute of Archaeology, Conservation and History
+  - UiO: University of Oslo, Institute of Archaeology, Conservation and History
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 ```{r get-data-functions, include = FALSE}
+# For reproducibility
+set.seed(123)
+
 # Load all sites and radiocarbon data
@@
 The first task was therefore to ascribe likely date ranges and associated uncertainty to the two phenomena.
-
-To take account of the gradient in the isostatic rebound, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase.
-This was done for each site by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves.
-Each value was then divided by the distance between the isobases.
-To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter.
-These values were then subtracted from the displacement curve located to the south-west of the site, and in the few cases where the site was located south-west of all isobases, the values were instead added.
+To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was interpolated to each site location from the displacement curves using inverse distance weighting based on the distance between the site and the isobases of the displacement curves [e.g.
+@conolly2020; @conolly2006,94--97].
+This was done for each year along the entirety of the curves using the squared inverse of the distances.
 The result of this process is shown for an example site in Figure \@ref(fig:example-1).
-
 For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
@@
 ```{r example-site, fig.show='hide'}
-
-# Assign closest isobases to site data.
-sites_sa <- st_join(st_make_valid(sites_sa), isopolys,
-                    join = st_intersects, largest = TRUE)
-
 # Specify site name
@@
 sitecurve <- interpolate_curve(years = xvals,
-                                 isobase1 = sitel$isobase1,
-                                 isobase2 = sitel$isobase2,
                                  target = sitel,
                                  dispdat = displacement_curves,
-                                 isodat = isobases,
-                                 direction_rel_curve1 = sitel$dir_rel_1)
+                                 isodat = isobases)
 # Add site name
@@
 
-dispdata <- filter(displacement_curves, name %in% c("Larvik", "Tvedestrand"))
-
 curve_plot <- ggplot() + 
-              geom_line(data = dispdata, 
+              geom_line(data = displacement_curves, 
                         aes(x = years, y = upperelev, col = name)) +
-  geom_line(data = dispdata, 
+  geom_line(data =  displacement_curves, 
                         aes(x = years, y = lowerelev, col = name)) +
@@
             aes(x = years, y = lowerelev, col = name)) +
-  scale_colour_manual(values = c("Larvik" = "darkgreen",
+  scale_colour_manual(values = c("Arendal" = "black",
+                                   "Larvik" = "darkgreen",
                                    "Tvedestrand" = "blue",
+                                   "Horten" = "darkorange",
                                  "Hegna vest 1" = "red")) +
@@
 C) The location of the site within the study area relative to isobases of the employed displacement curves.
-D) Displacement curve interpolated to the site location based on the distance between the site and the isobases of the two closest displacement curves.", out.width = "500px", fig.align="center", regfloat=TRUE}
+D) Displacement curve interpolated to the site location.", out.width = "500px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/example_map.png"))
@@
 ```{r example-site-cont2, fig.show='hide'}
-set.seed(1)
-
 # Retrieve example date to be used. The posterior sum of date group 1
@@
 
-```{r redate, echo = FALSE, fig.cap = "95% probability ranges of shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
+```{r redate, echo = FALSE, fig.cap = "95\\% probability ranges of shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
 These are comapred to the dates originally proposed by the excavation reports in black.", out.width = "525px", fig.align="center", regfloat=TRUE}

### 2022-02-07 analysis/supplementary-materials/supplementary.Rmd
Using IDW for the interpolation of the displacement curves. Re-running analyses

@@
 
+```{r auve,  echo = FALSE, out.width = "500px", regfloat=TRUE}
+include_graphics(here("analysis/figures/anvik.png"))
+
+filter(rcarbon, site_name  == "Auve") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Auve", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+
+```
+
+The site limit for Auve was drawn manually based on the description in the publication of the site and the site geometry available in the database Askeladden.
+The area to the has reportedly been affected by land-slides, but the extent of this is difficult to ascertain based on maps of the area.
+As the simulation results appear reasonable, nothing was done to try to correct for this.
+
 ```{r bakke, echo = FALSE, out.width = "500px", regfloat=TRUE}

### 2022-02-07 analysis/script/02script.R

@@ -54,12 +54,17 @@ arendal$name <- "Arendal"
 horten <-  curve_interval(xvals, skoppum_raw)
 horten$name <- "Horten"
 
-# Transform to BCE/CE using Joe Roe's era package and make BCE negative
-# to corresponds to oxcAAR later on
+# # Transform to BCE/CE using Joe Roe's era package and make BCE negative
+# # to corresponds to oxcAAR later on. [This gave some issue with rounding
+# errors]
+# displacement_curves <- rbind(larvik, tvedestrand, arendal, horten)
+# displacement_curves$years <- yr(displacement_curves$years, "cal BP")
+# displacement_curves$years <- yr_transform(displacement_curves$years, "BCE") * -1
+# xvals <- as.integer(yr_transform(yr(xvals, "cal BP"), "BCE") * -1)
+
 displacement_curves <- rbind(larvik, tvedestrand, arendal, horten)
-displacement_curves$years <- yr(displacement_curves$years, "cal BP")
-displacement_curves$years <- yr_transform(displacement_curves$years, "BCE") * -1
-xvals <- as.integer(yr_transform(yr(xvals, "cal BP"), "BCE") * -1)
+displacement_curves$years <- (displacement_curves$years - 1950) * -1
+xvals <- (xvals - 1950) * -1
 
 # Specify arbitrarily long distance of the line to represent the isobase at
 # each centre-point.

### 2022-02-08 analysis/paper/paper.Rmd
Re-running analyses after implementing IDW

@@
 @hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g.
-@lambeck1998; @svendsen1987].
+@svendsen1987].
 In other words, the RSL has been dropping throughout prehistory.
@@
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways.
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brø but see @mjærum2018; @gundersen2013; @schulke2020].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @mjærum2018; @gundersen2013; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
-associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@bakka1971; @prescott2012].
+The shift to agro-pastoralism and the introduction of the farm led the site locations to be more withdrawn from the shoreline [e.g.
+@bakka1971; @prescott2012; @ostmo2008, 223].
 That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods.
@@
 3900 BCE), pottery is introduced to the sites, and there are some possible indications of an initial uptake of agriculture at some sites in the Oslo fjord region.
-However, this appears to be small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955].
-[@nielsen2021b] has recently argued that while there, the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data.
+However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955].
+@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data [cf.
+@solheim2012, 74] .
 Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a more varied association between sites and the shore-line from around 3900 BCE, followed by a more decisive shift around 2400 BCE.
@@
 While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
-Negative values were therefore retained, with the exception of for the sites Gunnarsrød 5, Pjonkerød R1 and , where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
+Negative values were therefore retained, with the exception of for the sites Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
 

### 2022-02-08 analysis/script/04script.R

@@ -346,7 +346,6 @@ group_dates <- function(data, sitename){
   return(date_groups)
   }
 
-
 # Interpolate displacement curve for
 interpolate_curve <- function(years, target, dispdat, isodat){
 
@@ -356,7 +355,13 @@ interpolate_curve <- function(years, target, dispdat, isodat){
   values <- data.frame(matrix(ncol = 3, nrow = length(years)))
   names(values) <- c("years", "lowerelev", "upperelev")
 
-  for(i in 1:length(years)){
+  # In the case that a site is on the isobase of a
+  # displacement curve (e.g. Alveberget 8), simply return that displacement
+  # curve
+  if(any(as.numeric(dists) == 0)){
+    values <- dispdat[dispdat$name == names(dists)[which(as.numeric(dists) == 0)],]
+
+  } else { for(i in 1:length(years)){
     for(j in 1:ncol(dists)){
       le <- dispdat[which(dispdat$name == names(dists)[j] &
                             dispdat$years == years[i]),
@@ -372,8 +377,8 @@ interpolate_curve <- function(years, target, dispdat, isodat){
     distdat <- as.data.frame(t(dists))
     names(distdat) <- c("distance", "lower", "upper")
 
-    # No sites are older than the lowest limit of any displacement curve
-    # so in case of NA values, simply assign NA
+      # No sites are older than the lowest limit of any displacement curve
+      # so in case of NA values, simply assign NA
     if(any(is.na(distdat))){
       lowerval <- upperval <- NA
     } else {
@@ -388,8 +393,8 @@ interpolate_curve <- function(years, target, dispdat, isodat){
     }
     values[i,] <- c(years[i], lowerval, upperval)
   }
-
-  return(values)
+  }
+    return(values)
 }
 
 # Function to identify and load the correct raster if it is split up into

### 2022-02-09 analysis/paper/paper.Rmd
Re-running analyses. Also adjusted the editing of the highway by the Hegna sites slightly

@@
 @bakka1971; @prescott2012; @ostmo2008, 223].
-That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods.
+That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a].
 At an earlier stage, at the transition to the Early Neolithic (c.
-3900 BCE), pottery is introduced to the sites, and there are some possible indications of an initial uptake of agriculture at some sites in the Oslo fjord region.
+3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region.
 However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955].
-@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data [cf.
-@solheim2012, 74] .
-Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a more varied association between sites and the shore-line from around 3900 BCE, followed by a more decisive shift around 2400 BCE.
+@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also @solheim2012, 74; @amundsen2006].
+Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a more varied association between sites and the shore-line from around 3900 BCE, followed by a more decisive shift around 2400 BCE with the introduction of a comprehensive Neolithic package.
 
@@
 
-One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 ^14^C-dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
+One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 ^14^C-dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
 They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
@@
 
-Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
+Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
 @åstveit2018; @solheim2020].
@@
 @conolly2020; @conolly2006,94--97].
-This was done for each year along the entirety of the curves using the squared inverse of the distances.
+This was done for each year along the entirety of the curves using the standard squared inverse of the distances.
 The result of this process is shown for an example site in Figure \@ref(fig:example-1).
@@
 From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
-This corresponds well with the literature and corresponds with the Neolithisation process.
+This corresponds well with the literature and the suggested impact of the Neolithisation process.
 

### 2022-02-10 analysis/paper/paper.Rmd
Re-running analyses. Alo revisiting shoreline dating. Requires an increment of 0.001m?

@@
 
-One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 ^14^C-dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
+One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 ^14^C-dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
 They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
@@
 
-# Model dates using the Boundary function
+# Model dates using the Boundary function in OxCal
 datedat <- model_dates(sitedates = siter, date_groups = date_groups)

### 2022-02-10 analysis/script/05script.R
Re-running analyses using IDW

@@ -1043,10 +1043,10 @@ ggsave(file = here("analysis/figures/nauenc.png"), width = 250,
 sitename <- "Nordby 1"
 date_groups <- group_dates(rcarb_sa, sitename)
 
-output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
-                          isobases, nsamp = nsamp, loc_bbox = 400, siterpath)
-save(output,
-     file = here::here("analysis/data/derived_data/nordby1.RData"))
+# output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
+#                           isobases, nsamp = nsamp, loc_bbox = 400, siterpath)
+# save(output,
+#      file = here::here("analysis/data/derived_data/nordby1.RData"))
 load(here("analysis/data/derived_data/nordby1.RData"))
 
 sitearea <- rast(file.path(siterpath,
@@ -1085,10 +1085,10 @@ ggsave(file = here("analysis/figures/nordby52.png"), width = 250,
 sitename <- "Pauler 1"
 date_groups <- group_dates(rcarb_sa, sitename)
 
-# output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
-#                           isobases, nsamp = nsamp, loc_bbox = 2500, siterpath)
-# save(output,
-#      file = here::here("analysis/data/derived_data/pauler1.RData"))
+output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
+                          isobases, nsamp = nsamp, loc_bbox = 2500, siterpath)
+save(output,
+     file = here::here("analysis/data/derived_data/pauler1.RData"))
 load(here("analysis/data/derived_data/pauler1.RData"))
 
 

### 2022-02-11 analysis/paper/paper.Rmd
Re-running analyses using IDW
Re-running analyses with IDW

@@
 This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g.
-@berg-hansen2009; @bjerck2005; @sognnes2003; @tallavaara2020].
+@akerlund1996; @bjerck2005; @nordqvist1995; @sognnes2003; @tallavaara2020].
 
-The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C.
-Brøgger in 1905 (see the review by @maccurdy1906 for an early mention in English).
-Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g.
-@berg-hansen2009; @bjerck1990; @breivik2014; @solheim2018].
-The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are available.
+The close association between Scandinavian Stone Age settlements and prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century.
+Shoreline dating has been fundamental to Stone Age research in Norway ever since [see for example @berg-hansen2009; @bergsvik2020; @bjerck2008; @damlien2021].
+The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.
 However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites.
@@
 The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach.
-The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age.
+The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age, and in turn have this inform a refined method for shoreline dating.
 As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions.
 One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shoreline is inherently spatial.
-Taking inspiration from recent studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g.
-@crema2012; @crema2010; @lewis2021], a similar approach is adopted here.
+Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [@bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here.
 
@@
 @hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g.
-@svendsen1987].
+@svendsen1987; @mörner1979].
 In other words, the RSL has been dropping throughout prehistory.
@@
 In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [@reitan2009].
-Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008].
+Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008; @glorstad2020].
 Comparatively, given a continuous and still ongoing shoreline regression from as high as c.
-220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@glorstad2020; @hafsten1957; @hafsten1983; @sørensen1979].
+220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @sørensen1979].
 This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.
@@
 However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955].
-@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also @solheim2012, 74; @amundsen2006].
-Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a more varied association between sites and the shore-line from around 3900 BCE, followed by a more decisive shift around 2400 BCE with the introduction of a comprehensive Neolithic package.
+@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also @amundsen2006; @solheim2012, 74].
+Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a more varied association between sites and the shore-line from around 3900 BCE, followed by a more decisive shift around 2400 BCE with a comprehensive Neolithisation.
+
+[@breivik2014; @solheim2018]
 
@@
 This is based on typological and technological characteristics of the assemblages.
-Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate.
-Here this will be entirely dependent on, and follow the discretion of the original excavation reports.
+Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate [@schulke2020], but this distinction is not critical here.
+What is of interest is the relation between settlments and tool-production areas and the shoreline, and not has argued that some of these deviating dates could represent later visits that due to the    The evaluation of the radiocarbon dates will here be entirely dependent on, and follow the discretion of the original excavation reports.
 
@@
 This means that erosion, although likely not that prevalent, has not been taken into account.
-Secondly, the DTM has a vertical error...
+Secondly, the DTM has a vertical error [@lewis2021] ...
 Thirdly, the tidal range has not been considered...

### 2022-02-12 analysis/paper/paper.Rmd
Re-running analyses after IDW

@@
 
-The close association between Scandinavian Stone Age settlements and prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century.
+The close association between Scandinavian Stone Age settlements and shifting prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and being applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century.
 Shoreline dating has been fundamental to Stone Age research in Norway ever since [see for example @berg-hansen2009; @bergsvik2020; @bjerck2008; @damlien2021].

### 2022-02-12 analysis/script/07script.R

@@ -133,7 +133,9 @@ siteelev <- terra::extract(dtm, vect(sitel), fun = min)[2]
 # a = original amount before decay (1 here)
 # x = time (distance here)
 
-inc <- seq(0, 70, 0.001)
+inc <- seq(0, 70, 0.1)
+
+plot(expdat$offset, expdat$probs)
 
 expdat <- data.frame(
   offset = inc,
@@ -146,17 +148,17 @@ dategrid <- data.frame(
   probability = 0)
 
 for(i in 1:nrow(expdat)){
-  negative_offset <- as.numeric(siteelev - expdat$offset[i])
-  if(!(negative_offset > 0)) {
-    negative_offset <- 0.01
+  adjusted_elev <- as.numeric(siteelev - expdat$offset[i])
+  if(!(adjusted_elev > 0)) {
+    adjusted_elev <- 0.01
   }
   # Find lower date, subtracting offset (defaults to 0)
   lowerd <- round(approx(sitecurve[,"lowerelev"],
-                          xvals, xout = negative_offset)[['y']])
+                          xvals, xout = adjusted_elev)[['y']])
 
   # Find upper date, subtracting offset (defaults to 0)
   upperd <- round(approx(sitecurve[,"upperelev"],
-                          xvals, xout =  negative_offset)[['y']])
+                          xvals, xout = adjusted_elev)[['y']])
 
   # Find youngest and oldest date
   earliest <- min(c(lowerd, upperd))

### 2022-02-14 analysis/paper/paper.Rmd
New method for shoreline dating

--- 2022-02-11 analysis/paper/paper.Rmd
@@
 
-The close association between Scandinavian Stone Age settlements and prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century.
-Shoreline dating has been fundamental to Stone Age research in Norway ever since [see for example @berg-hansen2009; @bergsvik2020; @bjerck2008; @damlien2021].
+The close association between Scandinavian Stone Age settlements and shifting prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century.
+Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g.
+@mikkelsen1975; @berg-hansen2009; @bergsvik2020; @bjerck2008; @damlien2021].
 The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.
@@
 @milne2009; @mörner1976].
-Of central importance here is eustasy and glacial istostasy.
+Of central importance here is eustasy and istostasy.
 The eustatic sea-level is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015].
@@
 @conolly2020; @conolly2006,94--97].
-This was done for each year along the entirety of the curves using the standard squared inverse of the distances.
+This was done for each year along the entirety of the curves weighting the interpolation by the squared inverse of the distances.
 The result of this process is shown for an example site in Figure \@ref(fig:example-1).
@@
 
-The result of running the simulations across all sites are given in Figure \@ref(fig:results).
-Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, most of the sites have been situated close to the shoreline when they were in use.
+Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, the simulations show that most of the sites have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)).
 When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline.
 However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, as indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
-It is interesting to note the distance for some of the earliest sites, which appears somewhat high.
-This, however, can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level.
+The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level.
 Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.

### 2022-02-14 analysis/script/07script.R

@@ -42,38 +42,27 @@ rdates <- rdates %>%  group_by(site_name) %>%
   filter(class != "aprior")
 
 
-# Assign closest isobases to site data and exclude sites not analysed
-sites_sa <- st_join(st_make_valid(sites_sa), isopolys,
-                    join = st_intersects, largest = TRUE) %>%
-  filter(!(name %in% c("Dybdalshei 2", "Lunaveien", "Frebergsvik C")))
+# Exclude sites not analysed
+sites_sa <- sites_sa %>%
+  filter(!(name %in% c("Dybdalshei 2", "Lunaveien", "Frebergsvik C",
+                       "Bratsberg", "Lar\xf8nningen")))
 
 shorelinedates <- list()
 for(i in 1:nrow(sites_sa)){
   print(sites_sa$name[i])
-  shorelinedates[[i]] <- shoreline_date_exp_avg(sitename = sites_sa$name[i],
-                                            dtm = dtm,
-                                            displacement_curves = displacement_curves,
-                                            features = rcarb_sa,
-                                            sites = sites_sa,
-                                            isobases = isobases,
-                                            sitelimit = TRUE,
-                                            expratio =  expfit$estimate)
+  shorelinedates[[i]] <- shoreline_date(sitename = sites_sa$name[i],
+                                        elev = dtm,
+                                        disp_curves = displacement_curves,
+                                        sites = sites_sa,
+                                        iso = isobases,
+                                        expratio = expfit$estimate,
+                                        siteelev = "min",
+                                        reso = 0.001,
+                                        specified_elev = NA)
 }
-
-shorelinedates <- list()
-for(i in 1:nrow(sites_sa)){
-  print(sites_sa$name[i])
-  shorelinedates[[i]] <- shoreline_date_exp(sitename = sites_sa$name[i],
-                                            elev = dtm,
-                                            disp_curves = displacement_curves,
-                                            sites = sites_sa,
-                                            iso = isobases,
[…]

### 2022-02-14 analysis/paper/paper.Rmd
Re-applying shoreline dating to the Brunlanes sites

@@
 ```{r get-data-functions, include = FALSE}
-# For reproducibility
-set.seed(123)
-
 # Load all sites and radiocarbon data
@@
   ylab("Meters above present sea-level") +
-  xlab("cal BCE/CE") +
+  xlab("BCE/CE") +
   scale_colour_manual(values = c("Arendal" = "black",
@@
   ylab("Meters above present sea-level") +
-  xlab("cal BCE/CE") +
+  xlab("BCE/CE") +
   theme_bw() +
@@
 ```{r example-site-cont2, fig.show='hide'}
+# For reproducibility
+set.seed(1234)
+
 # Retrieve example date to be used. The posterior sum of date group 1
@@
                colour = "red") +
-  labs(x = "cal BCE")  +
+  labs(x = "BCE")  +
   #scale_y_discrete(expand = c(0.01, 0)) +
@@
                                    colour = as.factor(rcarb_cor))) +
-  labs(x = "cal BCE")  +
+  labs(x = "BCE")  +
   #scale_y_discrete(expand = c(0.01, 0)) +
@@
 ```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory.
-The sites are dated using the mean elevation of the site polygon, accounting for distance to closest isobase and by employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
+The sites are dated using the lowest elevation of the site polygon and by employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
 The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.

### 2022-02-14 analysis/supplementary-materials/supplementary.Rmd
Re-running analyses after implementing IDW
Re-applying shoreline dating to the Brunlanes sites

@@
 
-Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan, Rugtvedt [@odgaard1994], Tangen [@glorstad2005], Auve??
-[@østmo2008] and Torsrød [@østmo1975].
+Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan, Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@østmo1975].
 These sites are therefore not included in the analysis.
@@
 ```{r auve,  echo = FALSE, out.width = "500px", regfloat=TRUE}
-include_graphics(here("analysis/figures/anvik.png"))
+include_graphics(here("analysis/figures/auve.png"))
 
@@
 The area to the has reportedly been affected by land-slides, but the extent of this is difficult to ascertain based on maps of the area.
-As the simulation results appear reasonable, nothing was done to try to correct for this.
+As the simulation results appear reasonable [see also map in @ostmo2008, fig.142], nothing was done to try to correct for this.
 
@@
 
-```{r ragnhildrod, echo = FALSE, out.width = "500px", regfloat=TRUE}
-include_graphics(here("analysis/figures/ragnhildrod.png"))
-
-filter(rcarbon, site_name  == "Ragnhildrød") %>% 
-  select(lab_code, c14_bp, error, material, context) %>% 
-   kable(caption = "Ragnhildrød", col.names = c("ID",
-                    "\\textsuperscript{14}C BP",
-                    "Error",
-                    "Material",
-                    "Context"),
-        booktabs = TRUE,  linesep = c(""), escape = FALSE)
-```
+# ```{r ragnhildrod, echo = FALSE, out.width = "500px", regfloat=TRUE}
+# include_graphics(here("analysis/figures/ragnhildrod.png"))
+# 
+# filter(rcarbon, site_name  == "Ragnhildrød") %>% 
+#   select(lab_code, c14_bp, error, material, context) %>% 
+#    kable(caption = "Ragnhildrød", col.names = c("ID",
+#                     "\\textsuperscript{14}C BP",
+#                     "Error",
+#                     "Material",
+#                     "Context"),
+#         booktabs = TRUE,  linesep = c(""), escape = FALSE)
+# ```
 

### 2022-02-17 analysis/paper/paper.Rmd
Setting up re-dating of shoreline dated sites
All analyses have been re-run using IDW for shoreline displacement

@@
 Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g.
-@mikkelsen1975; @berg-hansen2009; @bergsvik2020; @bjerck2008; @damlien2021].
+@berg-hansen2009; @bjerck2008; @breivik2014; @nummedal1923; @mikkelsen1975; @solheim2018].
 The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.
@@
 One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shoreline is inherently spatial.
-Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [@bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here.
+Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g.
+@bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here.
 
@@
 Variation in this relative distance follow from a range of effects [e.g.
-@milne2009; @mörner1976].
+@milne2009].
 Of central importance here is eustasy and istostasy.
 The eustatic sea-level is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015].
-The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that impact the volume of ocean basins.
+The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans [@mörner1976].
 Istostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle.
@@
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways.
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @mjærum2018; @gundersen2013; @schulke2020].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87, but see @mjærum2018; @gundersen2013; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a more varied association between sites and the shore-line from around 3900 BCE, followed by a more decisive shift around 2400 BCE with a comprehensive Neolithisation.
-
-[@breivik2014; @solheim2018]
 
@@
 # Data
-To get at the relationship between sites and the contemporaneous shoreline, this analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
+To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
 While there is displacement data available for other areas of south-eastern Norway [e.g.
@@
 Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible.
-Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located a significant distance from a common isobase perpendicular to this gradient.
+Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from a common isobase perpendicular to this gradient.
 The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
@@
 Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, the simulations show that most of the sites have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)).
-When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline.
-However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, as indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
+Some of the sites are situated considerable distances from the shoreline when all dates are included, also those that were believed to be erroneous in the original reports, but if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, as indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
 The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level.
-Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.
+Corresponding with the literature, another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.
 From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
-This corresponds well with the literature and the suggested impact of the Neolithisation process.
 
@@
 Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations.
-The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, following from the skewed distribution.
+The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m.
 95% of the vertical distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
@@
 This is constrained slightly when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m.
-An exponential function has been fit to the distributions for vertical distance.
+While the median distances for horizontal and topographic is only m, it also evident that the range of these values is substantially wider than that for vertical distance, illustrating the point that minor variations in the latter can have substantial consequences for these distance measures, depending on the topography.
+
+An exponential function has been fit to the distributions for vertical distance, estimated by means of maximum likelihood.
 While it is clear that this does not perfectly match the data, it would seem reasonable from a theoretical perspective to assume that a process of exponential decay explains this relationship.
 Furthermore, the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
-Using the resulting exponential decay function, the sites treated here have been given a shoreline date in Figure \@ref(fig:shoredate).
+Having derived a exponential decay function for describing the vertical distance between site and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account.
+
+Where x is, asdf.
+This formula is used to shoreline date the same sites from where this relationship was derived in Figure \@ref(fig:shoredate).
+The Løvås sites and Gunnarsrød 5 deviate expectedly from the shoreline date.
+Using the method for evaluating degree of contemporaneity from @parnell2008, given in figure , indicates that the shoreline date is conteporenous date with the radiocarbon dates phase for .
+
+Using the resulting exponential decay function, the sites treated here have been given a shoreline date
 

### 2022-02-17 analysis/supplementary-materials/supplementary.Rmd
All analyses have been re-run using IDW for shoreline displacement

@@
 
-Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan, Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@østmo1975].
+Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan [@sjursjeike], Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@østmo1975].
 These sites are therefore not included in the analysis.

### 2022-02-17 analysis/script/05script.R

@@ -1276,10 +1276,10 @@ ggsave(file = here("analysis/figures/sageneb2.png"), width = 250,
 sitename <- "Solum 3"
 date_groups <- group_dates(rcarb_sa, sitename)
 
-output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
-                          isobases, nsamp = nsamp, loc_bbox = 3500, siterpath)
-save(output,
-     file = here::here("analysis/data/derived_data/solum3.RData"))
+# output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
+#                           isobases, nsamp = nsamp, loc_bbox = 3500, siterpath)
+# save(output,
+#      file = here::here("analysis/data/derived_data/solum3.RData"))
 load(here("analysis/data/derived_data/solum3.RData"))
 
 sitearea <- rast(file.path(siterpath,

### 2022-02-18 analysis/paper/paper.Rmd
Minor edits

@@
 The close association between Scandinavian Stone Age settlements and shifting prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century.
-Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g.
+Shoreline dating has been fundamental to Stone Age archaeology in Norway ever since [e.g.
 @berg-hansen2009; @bjerck2008; @breivik2014; @nummedal1923; @mikkelsen1975; @solheim2018].
@@
 This is based on typological and technological characteristics of the assemblages.
-Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate [@schulke2020], but this distinction is not critical here.
-What is of interest is the relation between settlments and tool-production areas and the shoreline, and not has argued that some of these deviating dates could represent later visits that due to the    The evaluation of the radiocarbon dates will here be entirely dependent on, and follow the discretion of the original excavation reports.
+Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate [e.g.
+@persson2008; @schulke2020].
+This distinction is not deemed critical here, as what is of interest is settlements and tool-production areas, as indicated by artefact inventories or multiple site features, and not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates.
+The evaluation of the radiocarbon dates will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.
 
 Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
-@åstveit2018; @solheim2020].
+@åstveit2018; @solheim2020, see also @akerlund1995].
 This approach has a couple of limitations.
@@
 
-The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shorelines, each associated with their own range of temporal uncertainty.
-The first task was therefore to ascribe likely date ranges and associated uncertainty to the two phenomena.
+Shoreline dating could be characterised as a chorological dating method.
+That is, it is a method where the target phenomenon is dated based on its geographical location -- in this case its altitude relative to the present day sea-level.
+The method is based on the spatial relationship between two phenomena, sites and shoreline displacement, each associated with their own range of temporal uncertainty.
+The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions.
 To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was interpolated to each site location from the displacement curves using inverse distance weighting based on the distance between the site and the isobases of the displacement curves [e.g.
-@conolly2020; @conolly2006,94--97].
-This was done for each year along the entirety of the curves weighting the interpolation by the squared inverse of the distances.
+@conolly2020; @conolly2006, 94--97].
+This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances.
 The result of this process is shown for an example site in Figure \@ref(fig:example-1).
@@
 # For reproducibility
-set.seed(1234)
+set.seed(1)
 
@@
 
-Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, the simulations show that most of the sites have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)).
-Some of the sites are situated considerable distances from the shoreline when all dates are included, also those that were believed to be erroneous in the original reports, but if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, as indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
-The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level.
+Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, the simulations show that the sites tend to have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)).
+Some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included, but if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
+The distances for some of the earliest sites appears somewhat high, but this can likely be explained as mainly the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the sea-level.
 Corresponding with the literature, another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.
 From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
+There are also a couple of sites located some distance from the shoreline just after 4000 BCE.
+While the sample size is limited, this would thus be in line with a development that sees in settlements in the immediate inland
 
@@
 These are recently excavated, well-dated sites situated in a relatively undistributed area of the landscape.
-Thus, while there is a danger of circularity of having archaeological sites inform a reconstruction of the shoreline displacement, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit as they would not have been in use when located beneath sea-level.
+Thus, while there would be a danger of circularity of having archaeological sites inform a reconstruction of the shoreline displacement, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level as they would not have been in use when located under water.
 Furthermore, as the Løvås sites are located at elevations from where the local displacement curve is interpolated between isololation basins at different elevation, it would seem that the sites represent a case where the archaeological material indicates discrepancies in the geological reconstructions of shoreline displacement in the area.
@@
 ```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory.
-The sites are dated using the lowest elevation of the site polygon and by employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline.
+The sites are dated using the lowest elevation of the site polygon and by employing the exponential decay function identified in Figure 7 for characterising the distance between sites and shoreline.
 The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.
@@
 
+In Figure \@ref(fig:redate)), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are subjected to the method for shoreline dating outlined above.
+These are compared to the dates originally proposed in the excavation reports for the sites.
+This is an unfair exercise for a few reasons.
+First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implicitly or explicitly associated uncertainty.
+Secondly, seeing as these reports are from various dates in time, some are based on outdated data on RSL-change.
+Finally, they are often only meant to indicate a lower bound for when the sites could have been in use.
+I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, if these dates have been interpreted and informed research in an approximate manner.
+The results from applying the method devised here again highlights the spatial and temporal contingency of the method.
+In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.
+Furthermore, a few of the date ranges extend well beyond the Stone Age and could be severely and securely constrained with only cursory reference to typology.
+
 ```{r redate, echo = FALSE, fig.cap = "95\\% probability ranges of shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
@@
 
-When no other information is available it can be recommended to apply the method for shoreline dating outlined above, which involves both taking the distance between sites and the isobase of displacement curves into consideration, and implementing to account for the distance between the sites and the shoreline.
+When no other information is available it can be recommended to apply the method for shoreline dating outlined above, which involves both taking the distance between sites and the isobase of displacement curves into consideration, and implementing formula x to account for the distance between the sites and the shoreline.
 The precision of this method is, as shown above, both geographically and temporally contingent, where older sites situated towards the north-east will get a more precise date than younger sites located towards the south-west.
 This approach will be useful either when shoreline dating a large number of sites where no other information is available [e.g.
-@solheim2018; @fossum2020; @nielsen2021b; @breivik2014], which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.
+@breivik2014; @solheim2018; @fossum2020; @nielsen2021b], which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.
 

### 2022-02-18 analysis/script/04script.R

@@ -1084,16 +1084,28 @@ shoreline_date <- function(sitename, elev = dtm,
                            iso = isobases,
                            reso = 0.1,
                            expratio, siteelev = "mean",
-                           specified_elev = NA){
+                           specified_elev = NA,
+                           sitelimit = TRUE,
+                           features = NA){
 
-  sitel <- filter(sites, name == sitename)
-  siteu <- st_union(sitel)
-  sitel <- st_as_sf(cbind(siteu, st_drop_geometry(sitel[1,])))
+  # If site features and not site limit is to be used
+  siter <- filter(features, site_name == sitename)
 
-  sitecurve <- interpolate_curve(years = xvals,
-                                 target = sitel,
-                                 dispdat = displacement_curves,
-                                 isodat = isobases)
+  # If the site limit is to be used
+  if(sitelimit == TRUE){
+    sitel <- filter(sites, name == sitename)
+    siteu <- st_union(sitel)
+    sitel <- st_as_sf(cbind(siteu, st_drop_geometry(sitel[1,])))
+  } else {
+    # If not create a convex hull around the dated features
+    sitel <- st_convex_hull(siter)
+    # And assign the values from the limit to the convex hull
+    sitel <- left_join(sitel, st_drop_geometry(filter(sites_sa,
+                  name == sitename)), by = c("site_name" = "name", "ask_id"))
+    names(sitel)[which(names(sitel) == "site_name")] <- "name"
+    sitel <- st_as_sf(sitel)
+  }
+
 
   if(!(is.na(specified_elev))){
     siteelev <- specified_elev

### 2022-02-21 analysis/paper/paper.Rmd
Minor edits

@@
 @hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g.
-@svendsen1987; @mörner1979].
+@mörner1979; @svendsen1987].
 In other words, the RSL has been dropping throughout prehistory.
@@
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways.
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87, but see @mjærum2018; @gundersen2013; @schulke2020].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also @amundsen2006; @solheim2012, 74].
-Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a more varied association between sites and the shore-line from around 3900 BCE, followed by a more decisive shift around 2400 BCE with a comprehensive Neolithisation.
+Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.
 
@@
 
-Shoreline dating could be characterised as a chorological dating method.
-That is, it is a method where the target phenomenon is dated based on its geographical location -- in this case its altitude relative to the present day sea-level.
-The method is based on the spatial relationship between two phenomena, sites and shoreline displacement, each associated with their own range of temporal uncertainty.
+Shoreline dating could be termed a chorological dating method.
+That is, it is a method where the target phenomenon is given a likely date of occurence with reference to its geographical location -- in this case its altitude relative to the present day sea-level.
+The method is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty.
 The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions.
-To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was interpolated to each site location from the displacement curves using inverse distance weighting based on the distance between the site and the isobases of the displacement curves [e.g.
+To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location from the displacement curves using inverse distance weighting based on the distance between the site and the isobases of the displacement curves [e.g.
 @conolly2020; @conolly2006, 94--97].
@@
 For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
-Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase.
+Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same settlement phase.
 In the case where there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed.
@@
 Some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included, but if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
-The distances for some of the earliest sites appears somewhat high, but this can likely be explained as mainly the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the sea-level.
-Corresponding with the literature, another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.
+The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the sea-level.
+Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age, corresponding with the literature.
 From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
 There are also a couple of sites located some distance from the shoreline just after 4000 BCE.
-While the sample size is limited, this would thus be in line with a development that sees in settlements in the immediate inland
+While the sample size is limited, this would thus be in line with a development that sees an in settlements in the immediate inland
 
@@
 These are recently excavated, well-dated sites situated in a relatively undistributed area of the landscape.
-Thus, while there would be a danger of circularity of having archaeological sites inform a reconstruction of the shoreline displacement, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level as they would not have been in use when located under water.
-Furthermore, as the Løvås sites are located at elevations from where the local displacement curve is interpolated between isololation basins at different elevation, it would seem that the sites represent a case where the archaeological material indicates discrepancies in the geological reconstructions of shoreline displacement in the area.
+Thus, while there would be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level, as they would not have been in use when located under water.
+Furthermore, as the Løvås sites are located at elevations from where the local displacement curve is interpolated between isololation basins at different elevation, it would seem that the sites represent a case where the archaeological material indicates discrepancies in the geological reconstruction of shoreline displacement in the area.
 
@@
 Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations.
-The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m.
-95% of the vertical distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
+The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, while 95% of the vertical distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
 That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
 This is constrained slightly when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m.
-While the median distances for horizontal and topographic is only m, it also evident that the range of these values is substantially wider than that for vertical distance, illustrating the point that minor variations in the latter can have substantial consequences for these distance measures, depending on the topography.
+While the median distances for horizontal and topographic is only and respectively, it also evident that the range of these values is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the topography.
 
-An exponential function has been fit to the distributions for vertical distance, estimated by means of maximum likelihood.
-While it is clear that this does not perfectly match the data, it would seem reasonable from a theoretical perspective to assume that a process of exponential decay explains this relationship.
-Furthermore, the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
-Having derived a exponential decay function for describing the vertical distance between site and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account.
+An exponential function has been fit to the distributions for vertical distance using maximum likelihood.
+While it is clear that this does not perfectly match the data, this can in part be related to methodological factors, where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
+Furthermore, it would make intuitive sense that a process of exponential decay explains this relationship.
+Thus, having derived a exponential decay function for describing the vertical distance between sites and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account.
 
@@
 The precision of this method is, as shown above, both geographically and temporally contingent, where older sites situated towards the north-east will get a more precise date than younger sites located towards the south-west.
-This approach will be useful either when shoreline dating a large number of sites where no other information is available [e.g.
+This approach will be useful either when shoreline dating a large number of sites where no other information than site elevation is taken into account [e.g.
 @breivik2014; @solheim2018; @fossum2020; @nielsen2021b], which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.

### 2022-02-21 analysis/supplementary-materials/supplementary.Rmd
Minor edits

@@
 
+```{r alveberget8,  echo = FALSE, out.width = "500px", regfloat=TRUE}
+include_graphics(here("analysis/figures/alveberget8.png"))
+
+filter(rcarbon, site_name  == "Alveberget 8") %>% 
+  select(lab_code, c14_bp, error, material, context) %>% 
+   kable(caption = "Alveberget 8", col.names = c("ID",
+                    "\\textsuperscript{14}C BP",
+                    "Error",
+                    "Material",
+                    "Context"),
+        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+
+```
+
+The artefact inventory at Alveberget 8 matches the radiocarbon dates [@mansrud2021]. The site is located in an undisturbed area of the DTM.
+
 ```{r anvik,  echo = FALSE, out.width = "500px", regfloat=TRUE}

### 2022-02-22 analysis/paper/paper.Rmd
Added age difference between shoreline and radiocarbon dates

@@
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
-The shift to agro-pastoralism and the introduction of the farm led the site locations to be more withdrawn from the shoreline [e.g.
+The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm, led the site locations to be more withdrawn from the shoreline [e.g.
 @bakka1971; @prescott2012; @ostmo2008, 223].
@@
 However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955].
-@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also @amundsen2006; @solheim2012, 74].
+@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also e.g.
+@amundsen2006; @solheim2012, 74].
 Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.
@@
 Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
-This is for example common in survey projects, which are often guided by a both a digital and mental reconstruction of past sea-levels.
+This is for example common in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g.
+@berg-hansen2009; ] .
 Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the typologically informed date-span is too wide to decisively verify this.
@@
 The 10m DTM is a down-sampled version of the 1m resolution DTM, which is based on aerial laser scanning using a minimum of two and up to five points per m^2^ and has a vertical error of  [@mapping2019].
-
 All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a digital research compendium following @marwick2018.
-The dataset is slightly more extensive than what is directly utilised in the study, and includes excavated sites without ^14^C-dates within the study area, as well as ^14^C-dates to periods other than the Stone Age.
 
@@
 
-As the excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances.
+As the excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure, the area immediately surrounding the sites has sometimes been severely impacted by modern disturbances.
 In addition to employing 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster.
@@
 ```{r example-3, echo = FALSE, fig.cap = "The result of 1000 simulation runs for each of the two groups of dates on the site Hegna vest 1.
-The first column of plots shows the radiocarbon probability density estimate from where dates were drawn during simulation.
+The first column of plots shows the radiocarbon probability density function from where dates were drawn during simulation.
 The second column displays the result of simulating the raised sea-level 1000 times.
@@
 Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations.
-The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, while 95% of the vertical distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
+The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, while 95% of the values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
 That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
@@
 An exponential function has been fit to the distributions for vertical distance using maximum likelihood.
-While it is clear that this does not perfectly match the data, this can in part be related to methodological factors, where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
-Furthermore, it would make intuitive sense that a process of exponential decay explains this relationship.
-Thus, having derived a exponential decay function for describing the vertical distance between sites and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account.
+While it makes theoretical sense thata process of exponential decay explains this relationship, it  is clear that this does not perfectly match the data.
+However, this can in part be related to methodological factors, where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
+Having derived a exponential decay function for describing the vertical distance between sites and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account.
 
@@
 The Løvås sites and Gunnarsrød 5 deviate expectedly from the shoreline date.
-Using the method for evaluating degree of contemporaneity from @parnell2008, given in figure , indicates that the shoreline date is conteporenous date with the radiocarbon dates phase for .
+Nordby 52 is not equally expected to deviate from the shoreline date, but this could follow from the fact that a highway runs by the site and had to be edited.
+Using the method for evaluating degree of contemporaneity from @parnell2008, given in figure , indicates that the shoreline date is conteporenous date with the radiocarbon dates phase for.
 
@@
 Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occured.
-Results associated with agricultural acitivites are in grey.
+Results associated with agricultural acitivites are plotted in grey.
 The first row shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites.
@@
 ```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory.
-The sites are dated using the lowest elevation of the site polygon and by employing the exponential decay function identified in Figure 7 for characterising the distance between sites and shoreline.
-The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.
-The red line indicates the 95\\% probability range for the shoreline date.", out.width = "525px", fig.align="center", regfloat=TRUE}
+The sites are dated using the mean elevation of the site polygon and by employing the exponential decay function identified in Figure 7 for characterising the distance between sites and shoreline.
+A) The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.
+The red line indicates the 95\\% probability range for the shoreline date.
+B) Using the method of Parnell et al.
+(2008)", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))

### 2022-02-22 analysis/script/04script.R

@@ -1088,16 +1088,16 @@ shoreline_date <- function(sitename, elev = dtm,
                            sitelimit = TRUE,
                            features = NA){
 
-  # If site features and not site limit is to be used
-  siter <- filter(features, site_name == sitename)
-
   # If the site limit is to be used
   if(sitelimit == TRUE){
     sitel <- filter(sites, name == sitename)
     siteu <- st_union(sitel)
     sitel <- st_as_sf(cbind(siteu, st_drop_geometry(sitel[1,])))
   } else {
-    # If not create a convex hull around the dated features
+    # If site features and not site limit is to be used
+    siter <- filter(features, site_name == sitename)
+
+    # Create a convex hull around the dated features
     sitel <- st_convex_hull(siter)
     # And assign the values from the limit to the convex hull
     sitel <- left_join(sitel, st_drop_geometry(filter(sites_sa,
@@ -1106,6 +1106,10 @@ shoreline_date <- function(sitename, elev = dtm,
     sitel <- st_as_sf(sitel)
   }
 
+  sitecurve <- interpolate_curve(years = xvals,
+                                 target = sitel,
+                                 dispdat = disp_curves,
+                                 isodat = iso)
 
   if(!(is.na(specified_elev))){
     siteelev <- specified_elev

### 2022-02-24 analysis/paper/paper.Rmd
Minor edits

@@
   fig.path = "../figures/",
+  fig.align = 'center',
   dpi = 300
@@
 # Group dates that overlap with 99.7 % probability using the group_dates()
-# function. All functions in this chunk where package isnt specified are defined 
-# in 04script.R
+# function. (All functions in this chunk where package isn't specified are 
+# defined in 04script.R).
 date_groups <- group_dates(rcarb_sa, sitename)
@@
 The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3.
-These are recently excavated, well-dated sites situated in a relatively undistributed area of the landscape.
+These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape.
 Thus, while there would be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level, as they would not have been in use when located under water.
-Furthermore, as the Løvås sites are located at elevations from where the local displacement curve is interpolated between isololation basins at different elevation, it would seem that the sites represent a case where the archaeological material indicates discrepancies in the geological reconstruction of shoreline displacement in the area.
+It could therefore seem that the Løvås sites represent a case where the archaeological material indicates discrepancies in the geological reconstruction of shoreline displacement in the area.
 
@@
 
-# Shoreline dating the pioneer sites from Brunlanes
+# Shoreline dating
 
@@
 A comparison of his results and the ones achieved using the above-outlined approach are given in Figure Figure \@ref(fig:brunlanes)A.
-These are largely in agreement.
-The sites are located more or less on the isobase of the displacement curve for Larvik, and negligible adjustments were therefore needed.
 The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf.
@@
 The steepness of the curve does nonetheless provide very precise dates, although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction.
-This example also highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
-One approach to this could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by measuring the steepness of the slope to the shoreline.
-Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline.
+This example highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
+One approach to this could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the steepness of the slope to the shoreline.
+Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline for when the site was in use.
 Such approaches would make less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where the displacement curve is flatter (cf.
 Figure \@ref(fig:shoredate)).
-To further illustrate this point, excavated Stone Age sites in the region which have previously been dated with reference to relative sea-level change are subjected to the outlined approach below.
 
@@
 
-# Shoreline dating
-
-In Figure \@ref(fig:redate)), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are subjected to the method for shoreline dating outlined above.
+In Figure \@ref(fig:redate)), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are also subjected to the method for shoreline dating outlined above.
 These are compared to the dates originally proposed in the excavation reports for the sites.
-This is an unfair exercise for a few reasons.
-First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implicitly or explicitly associated uncertainty.
-Secondly, seeing as these reports are from various dates in time, some are based on outdated data on RSL-change.
+To avoid issues with the DTM, the sites are dates based on the mean elevation of those provided for each site in the reports.
+This is an illustrative, but unfair exercise for a few reasons.
+First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit, but undefined uncertainty range.
+Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change.
 Finally, they are often only meant to indicate a lower bound for when the sites could have been in use.
-I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, if these dates have been interpreted and informed research in an approximate manner.
+I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner.
 The results from applying the method devised here again highlights the spatial and temporal contingency of the method.
@@
 Furthermore, a few of the date ranges extend well beyond the Stone Age and could be severely and securely constrained with only cursory reference to typology.
+This thus points to another way that the dating method could be improved.
+In the above example from Brunlanes it was shown how this could conceivably be done by evaluating the geographical location of the sites.
+Another possibility is of course to draw on other temporal data, such as typological indicators or radiocarbon dates with which a shoreline date could be combined, for example within a Bayesian framework.
 
@@
 
-# Discussion
+# Concluding remarks
 
-When no other information is available it can be recommended to apply the method for shoreline dating outlined above, which involves both taking the distance between sites and the isobase of displacement curves into consideration, and implementing formula x to account for the distance between the sites and the shoreline.
-The precision of this method is, as shown above, both geographically and temporally contingent, where older sites situated towards the north-east will get a more precise date than younger sites located towards the south-west.
+The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline.
+This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c.
+3900 BCE after which a couple of sites become situated some distance from the shoreline, and the more decisive break at the transition to the Late Neolithic at c.
+2400 BCE -- in clear agreement with the literature.
+Based on these findings, a refined method for the shoreline dating of Stone Age sites has been proposed.
+This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula x to account for the distance between the sites and the shoreline.
+When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratios identified in Figure to characterise this relationship.
+However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the site and other temporal data such as typological indicators in the artefact inventories.
+The precision of the method is, as shown above, both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more accurate date than younger sites located towards the south-west.
+The impact of such additional information will therefore also vary.
+
+Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further test and adjust the findings reported here, but could also lay the foundations for refining the method.
+Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located.
+Identifying such subsets of sites will thus potentially further refine the method, allowing it to be adjusted depending on the site being dated.
+Furthermore, other dimensions related to the topographic location of the sites could be similarly explored.
+This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@blankholm2020].
+If such locational patterns can be discerned, this will also be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site is available.
+
+It is also worth highlithing some sources of likely variation and uncertainty have not been considered here.
+First of all the DTM has only been corrected for major modern disturbances.
+This means that erosion, although likely not that prevalent, has not been taken into account.
+Secondly, the DTM has a vertical error [cf.
+@lewis2021].
+
+Thirdly, the tidal range has not been considered...
+And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site.
+These dimensions will have implications for the shore-line dating for certain subsets of the sites and could be future avenues through which one could further improve the method.
+These dimensions will also be imprtant for more substantial considerations of the behavioural relation between sites and the shoreline.
+
 This approach will be useful either when shoreline dating a large number of sites where no other information than site elevation is taken into account [e.g.
@@
 
-Some sources of likely variation and uncertainty have not been considered here.
-First of all the DTM has only been corrected for major modern disturbances.
-This means that erosion, although likely not that prevalent, has not been taken into account.
-Secondly, the DTM has a vertical error [@lewis2021] ...
-Thirdly, the tidal range has not been considered...
-Fourthly, the exposure of the sites to wave action is also likely to have been of concern [@roalkvam2020], and will presumably have implications for how close to the shoreline people settled [@blankholm2020].
-And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site.
-These dimensions will have implications for the shore-line dating for certain subsets of the sites and could be future avenues through which one could further improve the method.
-These dimensions will also be imprtant for more substantial considerations of the behavioural relation between sites and the shoreline.
+This quantification was achieved by means of Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty.
+Here the approach was simply used to inform the question of the distance between sites and shorelines.
+However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
 
-# Conclusion
-
-The most immediate contribution of this paper is a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline, which has largely been verified.
  Furthermore, the typical application of shoreline dating as an approximate method where date ranges are often for example rounded off to 200-year intervals, means that most previous research based on the method is given support by the findings reported here.
@@
 Attempting to move these considerations further through an explicit quantification has here resulted in a concrete recommendation of applying an offset of 0-15m when shoreline dating sites where no other information is available.
-
-This quantification was achieved by means of Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty.
-Here the approach was simply used to inform the question of the distance between sites and shorelines.
-However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
 

### 2022-02-24 analysis/script/07script.R

@@ -73,7 +73,7 @@ for(i in 1:nrow(sites_sa)){
                                         sites = sites_sa,
                                         iso = isobases,
                                         expratio = expfit$estimate,
-                                        siteelev = "min",
+                                        siteelev = "mean",
                                         reso = 0.001,
                                         specified_elev = NA)
   }

### 2022-02-28 analysis/paper/paper.Rmd
Adding references

@@
 This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g.
-@akerlund1996; @bjerck2005; @nordqvist1995; @sognnes2003; @tallavaara2020].
+@akerlund1996; @bjerck2005; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @rosentau2013].
 
@@
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways.
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @gundersen2013; @mjærum2018; @schulke2020].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; see also discussion in @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region.
-However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955].
+However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227].
 @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also e.g.
-@amundsen2006; @solheim2012, 74].
+@amundsen2006; @solheim2012, 74; @ostmo1988].
 Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.
@@
 This is for example common in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g.
-@berg-hansen2009; ] .
+@berg-hansen2009; @eskeland2017].
 Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the typologically informed date-span is too wide to decisively verify this.
@@
 This can for example pertain to accessibility.
-If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if it would have been located on an island when the sea-level was higher, it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft.
-Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few, and are typically conducted on a site-by-site or project-by-project basis, meaning the basis for assuming shore-bound site location is largely a mosaic of smaller scale investigations.
+If the site is situated on a ledge in a steep and jagged area of the present day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft.
+Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few [see also @ilves2011], and are typically conducted on a site-by-site or project-by-project basis, meaning the assumed shore-bound site location is largely based on a mosaic of smaller scale investigations.
 
@@
 
-The employed shoreline displacement curves are all based on the so-called isolation basin method [@romundset2011].
+The employed shoreline displacement curves are all based on the so-called isolation basin method [e.g.
+@kjemperud1986; @romundset2011].
 This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments in the cores.
-Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point.
+Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point (SLIP).
 Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible.
 Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from a common isobase perpendicular to this gradient.
-The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
+The SLIPs from the isolaton basin method indicate the highest astronomical tide at the time of isolation.
+This is adjusted to mean sea-level based on the present day tidal range, which is assumed to have been the same throughout the Holocene [@sorensen2014, 44].
+Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
 The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
@@
 This process was repeated 1000 times for each phase for each site.
-The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5, available in the supplementary material.
+The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5, available in the supplementary material [cf.
+@crema2010, 1125].
 Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography.
@@
 An exponential function has been fit to the distributions for vertical distance using maximum likelihood.
-While it makes theoretical sense thata process of exponential decay explains this relationship, it  is clear that this does not perfectly match the data.
+While it makes theoretical sense that a process of exponential decay explains this relationship, it  is clear that this does not perfectly match the data.
 However, this can in part be related to methodological factors, where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
-Having derived a exponential decay function for describing the vertical distance between sites and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account.
+Having derived an exponential decay function for describing the vertical distance between sites and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account:
 
@@
 B) Using the method of Parnell et al.
-(2008)", out.width = "525px", fig.align="center", regfloat=TRUE}
+(2008), the synchroneity of the dates is evaluated ", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 
-In Figure \@ref(fig:redate)), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are also subjected to the method for shoreline dating outlined above.
+In Figure \@ref(fig:redate), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are also subjected to the method for shoreline dating outlined above.
 These are compared to the dates originally proposed in the excavation reports for the sites.
@@
 In the above example from Brunlanes it was shown how this could conceivably be done by evaluating the geographical location of the sites.
-Another possibility is of course to draw on other temporal data, such as typological indicators or radiocarbon dates with which a shoreline date could be combined, for example within a Bayesian framework.
+Another possibility is of course to draw on other temporal data, such as typological indicators or radiocarbon dates with which a shoreline date could be combined, also within a Bayesian framework.
 
@@
 
-The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline.
-This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c.
-3900 BCE after which a couple of sites become situated some distance from the shoreline, and the more decisive break at the transition to the Late Neolithic at c.
-2400 BCE -- in clear agreement with the literature.
-Based on these findings, a refined method for the shoreline dating of Stone Age sites has been proposed.
-This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula x to account for the distance between the sites and the shoreline.
-When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratios identified in Figure to characterise this relationship.
-However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the site and other temporal data such as typological indicators in the artefact inventories.
-The precision of the method is, as shown above, both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more accurate date than younger sites located towards the south-west.
-The impact of such additional information will therefore also vary.
-
-Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further test and adjust the findings reported here, but could also lay the foundations for refining the method.
-Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located.
-Identifying such subsets of sites will thus potentially further refine the method, allowing it to be adjusted depending on the site being dated.
-Furthermore, other dimensions related to the topographic location of the sites could be similarly explored.
-This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@blankholm2020].
-If such locational patterns can be discerned, this will also be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site is available.
-
-It is also worth highlithing some sources of likely variation and uncertainty have not been considered here.
+It is worth mentioning some sources of likely variation and uncertainty that have not been considered here.
 First of all the DTM has only been corrected for major modern disturbances.
@@
 
+The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline.
+This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c.
+3900 BCE, after which a couple of sites become situated some distance from the shoreline, and the more decisive break at the transition to the Late Neolithic at c.
+2400 BCE -- in clear agreement with the literature.
+Furthermore, based on the quantative nature of these findings, a refined method for the shoreline dating of Stone Age sites has been proposed.
+This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula x to account for the distance between the sites and the shoreline.
+When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratios identified in Figure \@ref(fig:redate) to characterise this relationship.
+However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data such as typological indicators in the artefact inventories.
+The precision of the method is, as shown above, both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more accurate date than younger sites located towards the south-west.
+The impact of such additional information will therefore also vary.
+
+Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further test and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted.
+Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located.
+Furthermore, other dimensions related to the topographic location of the sites could be similarly explored.
+This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [e.g.
+@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@blankholm2020].
+If such locational patterns can be discerned, this will also be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site is available.
+
+Finally, the analysis done here was Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty.
+Here the approach was simply used to inform the question of the distance between sites and shorelines.
+However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
+
 This approach will be useful either when shoreline dating a large number of sites where no other information than site elevation is taken into account [e.g.
@@
 Other information is of course typically available.
-
-This quantification was achieved by means of Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty.
-Here the approach was simply used to inform the question of the distance between sites and shorelines.
-However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
 

### 2022-02-28 analysis/script/08script.R

@@ -259,7 +259,7 @@ ggsave(file = here("analysis/figures/brunlanes2.png"),
 psdates <- read.csv((here("analysis/data/raw_data/previous_shoreline_dates.csv"))) %>%
   mutate(start = start * -1,
          end = end * -1) %>%
-  rename(site_name = name)
+  rename("site_name" = "name")
 
 sites_sl <- site_limits %>% filter(radiocarbon == "f" | radiocarbon == "t" & rcarb_cor == "f")
 
@@ -272,16 +272,17 @@ sites_sl <- inner_join(sites_sl,
 
 sitdates <- list()
 for(i in 1:nrow(sites_sl)){
-  print(sites_sl$name[i])
+  print(c(i, sites_sl$name[i]))
+
   sitdates[[i]] <- shoreline_date(sitename = sites_sl$name[i],
                                  elev = dtm,
                                  disp_curves = displacement_curves,
                                  sites = sites_sl,
                                  iso = isobases,
                                  expratio = expfit$estimate,
-                                 siteelev = "mean",
                                  reso = 0.1,
-                                 specified_elev = sites_sl$min_elev[i])
+                                 specified_elev = (sites_sl$min_elev[i]+
+                                                 sites_sl$max_elev[i])/2)
 }
 
 bdates <- bind_rows(sitdates)
@@ -327,4 +328,4 @@ redateplt <- ggplot(data = hdr, aes(x = year_median, y = site_name)) +
   theme_bw()
 
 ggsave(file = here("analysis/figures/redate.png"), redateplt,
-       width = 200, height = 230, units = "mm")
+       width = 180, height = 230, units = "mm")

### 2022-03-01 analysis/paper/paper.Rmd
Slight adjustment to figures. Added a couple references

@@
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways.
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; see also discussion in @gundersen2013; @mjærum2018; @schulke2020].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see discussion in @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
-@åstveit2018; @solheim2020, see also @akerlund1995].
+@åstveit2018; @solheim2020; see also @akerlund1995].
 This approach has a couple of limitations.
@@
 This is constrained slightly when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m.
-While the median distances for horizontal and topographic is only and respectively, it also evident that the range of these values is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the topography.
+While the median for horizontal and topographic distances is only `r round(median(distances2$hordist))`m and `r round(median(distances2$vertdist))`m respectively, it is also evident that their range is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the topography.
 
 An exponential function has been fit to the distributions for vertical distance using maximum likelihood.
-While it makes theoretical sense that a process of exponential decay explains this relationship, it  is clear that this does not perfectly match the data.
-However, this can in part be related to methodological factors, where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
-Having derived an exponential decay function for describing the vertical distance between sites and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account:
+While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data.
+However, this can at least in part be related to methodological factors where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
+If one accepts this, having derived an exponential decay function for describing the vertical distance between sites and shoreline can be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:
 
-Where x is, asdf.
-This formula is used to shoreline date the same sites from where this relationship was derived in Figure \@ref(fig:shoredate).
-The Løvås sites and Gunnarsrød 5 deviate expectedly from the shoreline date.
-Nordby 52 is not equally expected to deviate from the shoreline date, but this could follow from the fact that a highway runs by the site and had to be edited.
-Using the method for evaluating degree of contemporaneity from @parnell2008, given in figure , indicates that the shoreline date is conteporenous date with the radiocarbon dates phase for.
-
-Using the resulting exponential decay function, the sites treated here have been given a shoreline date
-
-Following the general and aggregative result of this, one recommendation could be that given no other information, the method of shoreline dating should be applied with an uncertainty of 0-15m offset to the Stone Age site elevation relative to the displacement curves, following the median vertical distance and associated interquartile range.
-This would naturally be best applied when combined with a more comprehensive consideration of the site location, which could lead one to reduce this uncertainty, given for example surroundings where it would be reasonable to assume that a certain sea-levels would be unlikely in light of the surroundings of the site.
-
-Figure shows the result of shoreline dating the sites using the recommended off-set with radiocarbon dates plotted against this.
-The steeper the curve the more narrow the date range resulting from the procedure.
-This means that older siter, and sites situated towards the north-east will have a more narrow date range.
+Where x is [...] This formula is used to shoreline date the same sites from where this relationship was derived (Figure \@ref(fig:shoredate).
+The result of comparing the synchroneity of the dates using the method of @parnell2008, is given in figure \@ref(fig:shoredate)B.
+Here, 100,000 age samples drawn from each ^14^C-date was subtracted from 100 000 age samples from the corresponding shoreline date.
+The result of this indicates that the shoreline date correspond with the radiocarbon dates for 44 out of 49 sites `r round(44/49, 2)*100`%.
+The deviation from the shoreline date at the Løvås sites and Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langangen Vestgård 1 and Nordby 52.
+For these two sites the zero age difference can be seen to fall just outside the 95% interval, and in the simulation of the distance to the shoreline the shoreline was simulated to be 0m from the site across virtually all of the 1000 simulation runs for both sites (see supplementary material).
 
@@
 @lewis2021].
+Thirdly, while mean sea-level is used for dating the sites, a consideration of the tidal range
 
-Thirdly, the tidal range has not been considered...
   And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site.
@@
 
-Finally, the analysis done here was Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty.
+Finally, this analysis was based on Monte Carlo-simulation for integrating multiple sources of spatio-temporal uncertainty.
 Here the approach was simply used to inform the question of the distance between sites and shorelines.
@@
 
- Furthermore, the typical application of shoreline dating as an approximate method where date ranges are often for example rounded off to 200-year intervals, means that most previous research based on the method is given support by the findings reported here.
-However, the findings also suggest that simply using the method as a terminus ante quem for the occupation of the sites is, at least in my view, overly conservative and would stand to unnecessarily loose important chronological information.
-Attempting to move these considerations further through an explicit quantification has here resulted in a concrete recommendation of applying an offset of 0-15m when shoreline dating sites where no other information is available.
+ges are often for example rounded off to 200-year intervals, means that most previous research based on the method is given support by the findings reported here.
+However, the findings also suggest that simply using the method as a terminus ante quem for the occupation of the sites is overly conservative and would stand to loose important chronological information unnecessarily.
+Attempting to move these considerations further through an explicit quantification has here resulted in a concrete recommendation of applying an offset when shoreline dating sites where no other information is available.
 

### 2022-03-02 analysis/paper/paper.Rmd
Writing and slight updated to a couple figures

@@
 This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g.
-@akerlund1996; @bjerck2005; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @rosentau2013].
+@akerlund1996; @bjerck2005; @ilves2011; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009].
 
@@
 17--8 kyr BCE (data from Hughes et al.
-2016).
+2016, although see Romundset et al.
+2019).
 The study area defined later in the text is marked with a red outline.", out.width = "400px", fig.align="center", regfloat=TRUE}
@@
 
-The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways.
+The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [discussed by @bergsvik2009].
 The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see discussion in @gundersen2013; @mjærum2018; @schulke2020].
@@
 There are also a couple of sites located some distance from the shoreline just after 4000 BCE.
-While the sample size is limited, this would thus be in line with a development that sees an in settlements in the immediate inland
+While the sample size is limited, this would thus be in line with a development that sees an increase in settlements located in the immediate inland around this time.
 
@@
 This is constrained slightly when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m.
-While the median for horizontal and topographic distances is only `r round(median(distances2$hordist))`m and `r round(median(distances2$vertdist))`m respectively, it is also evident that their range is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the topography.
+While the median for horizontal and topographic distances is only `r round(median(distances2$hordist))`m and `r round(median(distances2$vertdist))`m respectively, it is also evident that their range is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
 
-An exponential function has been fit to the distributions for vertical distance using maximum likelihood.
+An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation.
 While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data.
-However, this can at least in part be related to methodological factors where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
+However, this can at least in part be related to methodological factors, where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
 If one accepts this, having derived an exponential decay function for describing the vertical distance between sites and shoreline can be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:
 
-Where x is [...] This formula is used to shoreline date the same sites from where this relationship was derived (Figure \@ref(fig:shoredate).
-The result of comparing the synchroneity of the dates using the method of @parnell2008, is given in figure \@ref(fig:shoredate)B.
-Here, 100,000 age samples drawn from each ^14^C-date was subtracted from 100 000 age samples from the corresponding shoreline date.
-The result of this indicates that the shoreline date correspond with the radiocarbon dates for 44 out of 49 sites `r round(44/49, 2)*100`%.
-The deviation from the shoreline date at the Løvås sites and Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langangen Vestgård 1 and Nordby 52.
-For these two sites the zero age difference can be seen to fall just outside the 95% interval, and in the simulation of the distance to the shoreline the shoreline was simulated to be 0m from the site across virtually all of the 1000 simulation runs for both sites (see supplementary material).
+Where x is [...] In Figure \@ref(fig:shoredate) this formula is used to shoreline date the same sites from where this relationship was derived.
+Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons.
+The result of comparing the synchroneity of the dates using the method presented by @parnell2008 is given in Figure \@ref(fig:shoredate)B.
+Here, 100,000 age samples drawn from each ^14^C-date was subtracted from 100,000 age samples drawn from the corresponding shoreline date.
+The resulting range between the 2.5th and 97.5th percentile were then checked to see if they cross zero, in which case the dates are considered synchronous.
+The result of this procedure indicates that the shoreline date correspond with the radiocarbon dates for 44 out of 49 sites (`r round(44/49, 2)*100`%).
+The deviation from the shoreline date at the two Løvås sites and Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langangen Vestgård 1 and Nordby 52.
+For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B), and in the simulation of the distance to the shoreline the shoreline was simulated to be 0m from the site across virtually all of the 1000 simulation runs for both sites (see supplementary material).
+If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of the two Løsvås sites and Gunnarsrød 5 are instead due to extraneous factors, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.
 
@@
 The sites are dated using the mean elevation of the site polygon and by employing the exponential decay function identified in Figure 7 for characterising the distance between sites and shoreline.
-A) The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black.
-The red line indicates the 95\\% probability range for the shoreline date.
-B) Using the method of Parnell et al.
-(2008), the synchroneity of the dates is evaluated ", out.width = "525px", fig.align="center", regfloat=TRUE}
+A) The 95\\% probability range for the shoreline dates are plotted in gray and underlined in red.
+These are plotted against the oldest radiocarbon-informed phase for each site in black.
+B) Using the method presented by Parnell et al.
+(2008), the synchroneity of the dates is evaluated.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 
+In Figure \@ref(fig:redate), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are subjected to the method for shoreline dating outlined above.
+These are compared to the dates originally proposed in the excavation reports for the sites.
+To avoid issues with recent disturbances in the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
+This is an illustrative, but unfair exercise for a few reasons.
+First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit but undefined uncertainty range.
+Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change.
+Finally, they are often only meant to indicate a lower bound for when the sites could have been in use.
+I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner.
+The results from applying the method devised here highlights the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% intervals for the dates.
+In some cases it provides a very precise date range and in others it offers little more than a *terminus post quem*.
+This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise date ranges.
+However, a few of the date ranges extend well beyond major chronological divisions, even the entire Stone Age, and could be severely and securely constrained with only cursory reference to typology.
+This thus points to the possibility of drawing on other temporal data such as typological indicators or radiocarbon dates, for example within a Bayesian framework, to improve the precision of the dating of the sites.
+
+```{r redate, echo = FALSE, fig.cap = "Shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
+The 95\\% probability ranges in red are comapred to the dates originally proposed by the excavation reports in black.", out.width = "525px", fig.align="center", regfloat=TRUE}
+knitr::include_graphics(here("analysis/figures/redate.png"))
+```
+
 Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
@@
 The shoreline dated pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a].
-These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location.
+These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision use of the sites after the sea retreated any significant distance from their location.
 In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically.
-A comparison of his results and the ones achieved using the above-outlined approach are given in Figure Figure \@ref(fig:brunlanes)A.
+A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \@ref(fig:brunlanes)A.
 The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf.
@@
 This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges.
-The steepness of the curve does nonetheless provide very precise dates, although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction.
-This example highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
-One approach to this could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the steepness of the slope to the shoreline.
+Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the fast rate of RSL-change in this period does nonetheless result in very precise dates.
+Above it was shown how additional temporal data could be combined with the method to improve its accuracy and precision.
+This example, on the other hand, highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
+One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels based on the shoreline date, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the steepness of the slope to the shoreline.
 Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline for when the site was in use.
-Such approaches would make less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where the displacement curve is flatter (cf.
-Figure \@ref(fig:shoredate)).
+Such approaches would make less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change as been less severe (cf.
+Figure \@ref(fig:redate)).
 
@@
 
-In Figure \@ref(fig:redate), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are also subjected to the method for shoreline dating outlined above.
-These are compared to the dates originally proposed in the excavation reports for the sites.
-To avoid issues with the DTM, the sites are dates based on the mean elevation of those provided for each site in the reports.
-This is an illustrative, but unfair exercise for a few reasons.
-First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit, but undefined uncertainty range.
-Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change.
-Finally, they are often only meant to indicate a lower bound for when the sites could have been in use.
-I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner.
-The results from applying the method devised here again highlights the spatial and temporal contingency of the method.
-In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.
-Furthermore, a few of the date ranges extend well beyond the Stone Age and could be severely and securely constrained with only cursory reference to typology.
-This thus points to another way that the dating method could be improved.
-In the above example from Brunlanes it was shown how this could conceivably be done by evaluating the geographical location of the sites.
-Another possibility is of course to draw on other temporal data, such as typological indicators or radiocarbon dates with which a shoreline date could be combined, also within a Bayesian framework.
-
-```{r redate, echo = FALSE, fig.cap = "95\\% probability ranges of shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
-These are comapred to the dates originally proposed by the excavation reports in black.", out.width = "525px", fig.align="center", regfloat=TRUE}
-knitr::include_graphics(here("analysis/figures/redate.png"))
-```
-
 # Concluding remarks
 
-It is worth mentioning some sources of likely variation and uncertainty that have not been considered here.
+It is worth highlighting some sources of likely variation and uncertainty that have not been considered here.
 First of all the DTM has only been corrected for major modern disturbances.
@@
 @lewis2021].
-Thirdly, while mean sea-level is used for dating the sites, a consideration of the tidal range
+Thirdly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves away from the isobases of the displacement curves.
+Fourthly, the elevation range over which the sites themselves extend was not considered.
+And finally, while mean sea-level is used for dating the sites, a consideration of the tidal range
 
@@
 This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c.
-3900 BCE, after which a couple of sites become situated some distance from the shoreline, and the more decisive break at the transition to the Late Neolithic at c.
+3900 BCE, after which a couple of sites become situated some distance from the shoreline.
+This is followed by a more decisive break at the transition to the Late Neolithic at c.
 2400 BCE -- in clear agreement with the literature.
-Furthermore, based on the quantative nature of these findings, a refined method for the shoreline dating of Stone Age sites has been proposed.
+Furthermore, based on the quantitative nature of these findings, a refined method for the shoreline dating of Stone Age sites has been proposed.
 This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula x to account for the distance between the sites and the shoreline.
 When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratios identified in Figure \@ref(fig:redate) to characterise this relationship.
-However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data such as typological indicators in the artefact inventories.
-The precision of the method is, as shown above, both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more accurate date than younger sites located towards the south-west.
+However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data such as radiocarbon dates and typological indicators in the artefact inventories.
+The precision of the method is, as shown above, both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date than younger sites located towards the south-west.
 The impact of such additional information will therefore also vary.
@@
 @roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@blankholm2020].
-If such locational patterns can be discerned, this will also be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site is available.
+If such locational patterns can be discerned, this will also be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site beyond its location is available.
 
@@
 @breivik2014; @solheim2018; @fossum2020; @nielsen2021b], which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.
-
-The outlined approach is thus derived from the available radiocarbon record, and appears to be able to accommodate the available data.
-To further test the method will require independently derived radiocarbon dates.
-
-Other information is of course typically available.
-
-ges are often for example rounded off to 200-year intervals, means that most previous research based on the method is given support by the findings reported here.
-However, the findings also suggest that simply using the method as a terminus ante quem for the occupation of the sites is overly conservative and would stand to loose important chronological information unnecessarily.
-Attempting to move these considerations further through an explicit quantification has here resulted in a concrete recommendation of applying an offset when shoreline dating sites where no other information is available.
 

### 2022-03-03 analysis/paper/paper.Rmd
More writing

@@
 Shoreline dating has been fundamental to Stone Age archaeology in Norway ever since [e.g.
-@berg-hansen2009; @bjerck2008; @breivik2014; @nummedal1923; @mikkelsen1975; @solheim2018].
+@berg-hansen2009; @bjerck2008; @bjerck1990; @breivik2014; @nummedal1923; @mikkelsen1975; @solheim2018].
 The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.
@@
 Comparatively, given a continuous and still ongoing shoreline regression from as high as c.
-220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @sørensen1979].
+220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979].
 This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.
@@
 17--8 kyr BCE (data from Hughes et al.
-2016, although see Romundset et al.
+2016, but see Romundset et al.
 2019).
@@
 
-The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [discussed by @bergsvik2009].
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see discussion in @gundersen2013; @mjærum2018; @schulke2020].
+The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [ @berg-hansen2009; @bergsvik2009].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
-The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm, led the site locations to be more withdrawn from the shoreline [e.g.
+The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm is to have led site locations to be more withdrawn from the shoreline [e.g.
 @bakka1971; @prescott2012; @ostmo2008, 223].
@@
 However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227].
-@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also e.g.
+@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g.
 @amundsen2006; @solheim2012, 74; @ostmo1988].
@@
 @persson2008; @schulke2020].
-This distinction is not deemed critical here, as what is of interest is settlements and tool-production areas, as indicated by artefact inventories or multiple site features, and not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates.
-The evaluation of the radiocarbon dates will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.
+However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas evidenced by artefact inventories or multiple site features, and not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates.
+The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.
 
@@
 Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from a common isobase perpendicular to this gradient.
-The SLIPs from the isolaton basin method indicate the highest astronomical tide at the time of isolation.
-This is adjusted to mean sea-level based on the present day tidal range, which is assumed to have been the same throughout the Holocene [@sorensen2014, 44].
+The SLIPs from the isolaton basin method indicate the isolation from the highest astronomical tide.
+For the displacement curve, this is then adjusted to mean sea-level based on the present day tidal range, which is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
 Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
@@
 Three of the sites have been associated with farming activities, either in directly or in the form building structures.
-The first is Nordby 1, at which the ^14^C-dates are associated with a Late Neolithic long-house.
-The Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are both related to farming.
+The first is Nordby 1, at which the ^14^C-dates are associated with a Late Neolithic long-house [@gjerpe2008b].
+The Middle Neolithic phase at Kvastad A2 [@stokke2018b] and Late Neolithic phase at Nauen A [@persson2008] are both related to farming.
 Both of these sites also have radiocarbon dates and lithic inventory associated with Mesolithic forager activities.
 Following from the expected deviance from the settlement patterns that are to characterise forager sites, these agricultural phases are thus treated separately below.
-Finally, @nielsen2021b has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratsberg and Larønningen could be related to early agricultural activity in the Oslo fjord region.
+Finally, @nielsen2021b has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratsberg [@wenn2012] and Larønningen [@roberg2012] could be related to early agricultural activity in the Oslo fjord region.
 Due to the uncertain and somewhat speculative nature of this suggestion, these are omitted here.
@@
 This was done using regularised spline interpolation with tension [e.g.
-@conolly2020], using the default settings of r.fillnulls from GRASS GIS in R through the package rgrass7 [@bivand2021].
+@conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021].
 In addition to code and original spatial data being available in the research compendium for this paper, it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.
@@
               geom_line(data = displacement_curves, 
-                        aes(x = years, y = upperelev, col = name)) +
+                      aes(x = years, y = upperelev, col = name), alpha = 0.4) +
   geom_line(data =  displacement_curves, 
-                        aes(x = years, y = lowerelev, col = name)) +
+                      aes(x = years, y = lowerelev, col = name), alpha = 0.4) +
   geom_line(data = sitecurve,
@@
 That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
-This is constrained slightly when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m.
-While the median for horizontal and topographic distances is only `r round(median(distances2$hordist))`m and `r round(median(distances2$vertdist))`m respectively, it is also evident that their range is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
+While these values remain the same, the mean and standard deviation is constrained slightly when only considering the Mesolithic dates.
+While the median for horizontal and topographic distances is only `r round(median(distances2$hordist))`m and `r round(median(distances2$topodist))`m respectively, it is also evident that their range is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
 
-An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation.
+An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation (Figure \@ref(fig:results2)).
 While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data.
@@
 Where x is [...] In Figure \@ref(fig:shoredate) this formula is used to shoreline date the same sites from where this relationship was derived.
-Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons.
+Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation.
 The result of comparing the synchroneity of the dates using the method presented by @parnell2008 is given in Figure \@ref(fig:shoredate)B.
@@
 The deviation from the shoreline date at the two Løvås sites and Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langangen Vestgård 1 and Nordby 52.
-For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B), and in the simulation of the distance to the shoreline the shoreline was simulated to be 0m from the site across virtually all of the 1000 simulation runs for both sites (see supplementary material).
-If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of the two Løsvås sites and Gunnarsrød 5 are instead due to extraneous factors, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.
+For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B), and in the simulation of the distance to the shoreline this was simulated to be 0m across virtually all of the 1000 simulation runs for both sites (see supplementary material).
+If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of the two Løsvås sites and Gunnarsrød 5 are instead due to extraneous factors and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.
 
@@
 These are plotted against the oldest radiocarbon-informed phase for each site in black.
-B) Using the method presented by Parnell et al.
-(2008), the synchroneity of the dates is evaluated.", out.width = "525px", fig.align="center", regfloat=TRUE}
+B) Using the method described by Parnell et al.
+(2008), the synchroneity of the dates is evaluated.
+When the 95\\% interval for age difference crosses zero the dates are considered synchronous.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 In some cases it provides a very precise date range and in others it offers little more than a *terminus post quem*.
-This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise date ranges.
+This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise date ranges (Figure \@ref(fig:map-iso)B).
 However, a few of the date ranges extend well beyond major chronological divisions, even the entire Stone Age, and could be severely and securely constrained with only cursory reference to typology.
@@
 This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges.
-Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the fast rate of RSL-change in this period does nonetheless result in very precise dates.
+Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does nonetheless result in very precise dates.
 Above it was shown how additional temporal data could be combined with the method to improve its accuracy and precision.
 This example, on the other hand, highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
-One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels based on the shoreline date, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the steepness of the slope to the shoreline.
+One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the steepness of the slope to the shoreline.
 Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline for when the site was in use.
@@
 Thirdly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves away from the isobases of the displacement curves.
-Fourthly, the elevation range over which the sites themselves extend was not considered.
-And finally, while mean sea-level is used for dating the sites, a consideration of the tidal range
-
-  And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site.
-These dimensions will have implications for the shore-line dating for certain subsets of the sites and could be future avenues through which one could further improve the method.
-These dimensions will also be imprtant for more substantial considerations of the behavioural relation between sites and the shoreline.
+Finally, the definition of the site limits nor the elevation range over which these extend was not considered.
 
@@
 
-Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further test and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted.
+Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further test and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted and refined.
 Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located.
@@
 @roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@blankholm2020].
-If such locational patterns can be discerned, this will also be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site beyond its location is available.
+This is also related to the fact that while mean sea-level is used for dating the sites, a consideration of the tidal range could possibly have implications for the site location relative to the shoreline, depending on the topography.
+If such locational patterns can be discerned, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site beyond its location is available.
 
-Finally, this analysis was based on Monte Carlo-simulation for integrating multiple sources of spatio-temporal uncertainty.
-Here the approach was simply used to inform the question of the distance between sites and shorelines.
+Finally, this analysis was based on a simulation approach for integrating multiple sources of spatio-temporal uncertainty.
+Here this was simply used to inform the question of the distance between sites and shorelines.
 However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

### 2022-03-03 analysis/script/08script.R

@@ -47,9 +47,6 @@ for(i in 1:nrow(brunlanes)){
                                         siteelev = "min",
                                         reso = 0.01,
                                  specified_elev = brunlanes$elevation[i])
-
-
-
 }
 
 bdates <- bind_rows(sdatesp)

### 2022-03-04 analysis/paper/paper.Rmd
using HDR instead of quantiles for shoreline dates

@@
 The result of comparing the synchroneity of the dates using the method presented by @parnell2008 is given in Figure \@ref(fig:shoredate)B.
-Here, 100,000 age samples drawn from each ^14^C-date was subtracted from 100,000 age samples drawn from the corresponding shoreline date.
-The resulting range between the 2.5th and 97.5th percentile were then checked to see if they cross zero, in which case the dates are considered synchronous.
-The result of this procedure indicates that the shoreline date correspond with the radiocarbon dates for 44 out of 49 sites (`r round(44/49, 2)*100`%).
-The deviation from the shoreline date at the two Løvås sites and Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langangen Vestgård 1 and Nordby 52.
+Here, 100,000 age samples drawn from each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date.
+The resulting range of the 95% highest density region (HDR, @ ) was then checked to see if it crosses zero, in which case the dates are considered synchronous.
+The result of this procedure indicates that the shoreline date correspond with the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%).
+The deviation from the shoreline date at Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langemyr and Kvastad A2.
 For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B), and in the simulation of the distance to the shoreline this was simulated to be 0m across virtually all of the 1000 simulation runs for both sites (see supplementary material).

### 2022-03-04 analysis/script/07script.R

@@ -3,6 +3,10 @@ library(patchwork)
 library(here)
 library(terra)
 library(sf)
+library(hdrcde)
+
+# For reproducibility
+set.seed(1)
 
 # Load required functions and prepared data
 source(here("analysis/script/04script.R"))
@@ -44,8 +48,7 @@ rdates <- rdates %>%  group_by(site_name) %>%
 
 # Exclude sites not analysed
 sites_sa <- sites_sa %>%
-  filter(!(name %in% c("Dybdalshei 2", "Lunaveien", "Frebergsvik C",
-                       "Bratsberg", "Lar\xf8nningen")))
+  filter(!(name %in% c("Dybdalshei 2", "Lunaveien", "Frebergsvik C")))
 
 # Specify for which sites to not shoreline date the site limit
 # (see supplementary material for more on this)
@@ -103,12 +106,35 @@ corshore <- sdates %>%  filter(site_name %in% unique(corsites$site_name))
 # Find 95 % probability range for shoreline dates and median shoreline date
 # for ordering in the plot
 hdr <- corshore %>%  group_by(site_name) %>%
-  filter(cumsum(replace_na(probability, 0)) < 0.95 &
-           probability != 0) %>%
-  summarise(
-    year_min = min(years, na.rm = TRUE),
-    year_max = max(years, na.rm = TRUE),
-    year_median = median(years, na.rm = TRUE))
+  mutate(hdr = list(hdr(den = list("x" = years,
+                              "y" = probability), prob = 95)$hdr),
+         year_median = hdr(den = list("x" = years,
+                                      "y" = probability), prob = 95)$mode)
+
+
+ggplot(da)
+
+ggplot(data = hdrs,
+       aes(x = year_median, y = reorder(site_name, -year_median))) +
+  ggridges::geom_density_ridges_gradient(data = corshore,
+                               aes(x = years, y = site_name,
+                                   fill = stat(quantile)),
+                               quantile_lines = TRUE,
+                               quantile_fun = HDInterval::hdi,
+                               vline_linetype = 2) +
+  scale_fill_manual(values = c("transparent", "lightblue", "transparent"),
+                    guide = "none")
+
+
+  geom_segment(data = hdrs, aes(x = bind_rows(hdr), xend = bind_rows(hdr),
+                               yend = site_name), col = "red", size = 0.6)
+
+  mutate(year_min = min(hdr(den = list("x" = years,
+                                       "y" = probability), prob = 95)$hdr),
+         year_max = max(hdr(den = list("x" = years,
[…]

### 2022-03-07 analysis/paper/paper.Rmd
Set up plotting with HDR. Also miss a few sites for sites_sl due to a previous error in the subsetting code which caused a few sites to be excluded

@@
 This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g.
-@akerlund1996; @bjerck2005; @ilves2011; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009].
+@akerlund1996; @bjerck2005; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @wikell2009].
 
@@
 These conditions are directly reflected in the archaeological record.
-In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [@reitan2009].
+In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [e.g.
+@hagen1963; @reitan2009].
 Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008; @glorstad2020].
@@
 
-The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [ @berg-hansen2009; @bergsvik2009].
+The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [discussed in @berg-hansen2009; @bergsvik2009].
 The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @gundersen2013; @mjærum2018; @schulke2020].
@@
 It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g.
-@jaksland2014; @nummedal1923].
+@jaksland2014; @johansen1963; @nummedal1923].
 This can for example pertain to accessibility.
@@
 
-One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 ^14^C-dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
+One of the more comprehensive evaluations of the relationship between radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
 They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
@@
 The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.
+
+The method of shoreline dating has variably been said to contribute a method for relative dating for sites located at different elevations, for offering a lower limit for the site could have been in use, or been used to date a larger number of sites assign uniform uncertainty ranges of 200 or 500 years.
 
@@
 
-Shoreline dating could be termed a chorological dating method.
-That is, it is a method where the target phenomenon is given a likely date of occurence with reference to its geographical location -- in this case its altitude relative to the present day sea-level.
-The method is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty.
+The method of shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty.
 The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions.
@@
 Here, 100,000 age samples drawn from each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date.
-The resulting range of the 95% highest density region (HDR, @ ) was then checked to see if it crosses zero, in which case the dates are considered synchronous.
+The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered synchronous.
 The result of this procedure indicates that the shoreline date correspond with the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%).
@@
 ```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory.
-The sites are dated using the mean elevation of the site polygon and by employing the exponential decay function identified in Figure 7 for characterising the distance between sites and shoreline.
-A) The 95\\% probability range for the shoreline dates are plotted in gray and underlined in red.
-These are plotted against the oldest radiocarbon-informed phase for each site in black.
+A) The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black.
+These are plotted against the oldest radiocarbon-informed phase for each site in red.
 B) Using the method described by Parnell et al.
 (2008), the synchroneity of the dates is evaluated.
-When the 95\\% interval for age difference crosses zero the dates are considered synchronous.", out.width = "525px", fig.align="center", regfloat=TRUE}
+When the range of the 95\\% HDR for age difference crosses zero the dates are considered synchronous.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 
-In Figure \@ref(fig:redate), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are subjected to the method for shoreline dating outlined above.
+To further explore the implementation for shoreline dating outline above, excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the method outlined above (Figure \@ref(fig:redate)).
 These are compared to the dates originally proposed in the excavation reports for the sites.
@@
 This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise date ranges (Figure \@ref(fig:map-iso)B).
-However, a few of the date ranges extend well beyond major chronological divisions, even the entire Stone Age, and could be severely and securely constrained with only cursory reference to typology.
+However, some of the date ranges extend well beyond major chronological divisions, even the entire Stone Age, and could be severely and securely constrained with only cursory reference to typology.
 This thus points to the possibility of drawing on other temporal data such as typological indicators or radiocarbon dates, for example within a Bayesian framework, to improve the precision of the dating of the sites.
@@
 ```{r redate, echo = FALSE, fig.cap = "Shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
-The 95\\% probability ranges in red are comapred to the dates originally proposed by the excavation reports in black.", out.width = "525px", fig.align="center", regfloat=TRUE}
+The range of the 95\\% HDR in black are comapred to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/redate.png"))
@@
 Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline for when the site was in use.
-Such approaches would make less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change as been less severe (cf.
+Such approaches would make less of an impact in this setting, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change as been less severe (cf.
 Figure \@ref(fig:redate)).
@@
 A) The result of applying the approach to shoreline dating outlined above.
-Dates provided by Jaksland (2014) are plotted in black.
-The box indicates a 50 year uncertainty range which in combination with the black line extends 200 years.
+Dates provided by Jaksland (2014) are plotted in red.
+The box indicates a 50 year uncertainty range which in combination with the red line extends 200 years.
 Follwing Jaksland's approach these start from the earliest possible date.

### 2022-03-08 analysis/paper/paper.Rmd
Slight adustment to figures and more writing

@@
 
-The close association between Scandinavian Stone Age settlements and shifting prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century.
+The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century.
 Shoreline dating has been fundamental to Stone Age archaeology in Norway ever since [e.g.
-@berg-hansen2009; @bjerck2008; @bjerck1990; @breivik2014; @nummedal1923; @mikkelsen1975; @solheim2018].
+@berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @nummedal1923; @mikkelsen1975; @solheim2018].
 The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.
@@
 
+The method of shoreline dating has ranged from being met with scepticism as related to the degree sites would have been consistently shore-bound, to being deemed a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer little more than a earliest possible date for when a site could have been in use.
+The most common application in Norway has arguably been to use the method to provide a very approximate date for the occupation of the sites, typically in combination with other dating methods [see for example chapters in @; ; and below].
+Recently the method has also been used idenpendently to date a larger number sites to get a general impression of site frequency over time, typically by aggregating point estimates of shoreline dates in 200 or 500 year bins [e.g.
+@breivik2014; @solheim2018; @fossum2020; @nielsen2021b].
+In his discussion of the method, @nordqvist1999 -- in part drawing on @johansen1963 -- argues that there can be little doubt about the general applicability of the method, but what is less clear is the level of reliability and chronological resolution that it can offer.
+
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [discussed in @berg-hansen2009; @bergsvik2009].
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @gundersen2013; @mjærum2018; @schulke2020].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; although see @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227].
-@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g.
+@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic of the Oslo fjord area is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g.
 @amundsen2006; @solheim2012, 74; @ostmo1988].
@@
 They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
-However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates.
+However, where there was a discrepancy the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates.
 This is based on typological and technological characteristics of the assemblages.
 Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate [e.g.
-@persson2008; @schulke2020].
+@persson2008; @schulke2020; see also ].
 However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas evidenced by artefact inventories or multiple site features, and not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates.
 The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.
-
-The method of shoreline dating has variably been said to contribute a method for relative dating for sites located at different elevations, for offering a lower limit for the site could have been in use, or been used to date a larger number of sites assign uniform uncertainty ranges of 200 or 500 years.
 
@@
 
-The employed shoreline displacement curves are all based on the so-called isolation basin method [e.g.
+The employed shoreline displacement curves are all based on the so-called isolation basin method [see e.g.
 @kjemperud1986; @romundset2011].
-This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments in the cores.
+This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments.
 Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point (SLIP).
 Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible.
-Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from a common isobase perpendicular to this gradient.
-The SLIPs from the isolaton basin method indicate the isolation from the highest astronomical tide.
-For the displacement curve, this is then adjusted to mean sea-level based on the present day tidal range, which is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
+Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from a common isobase perpendicular to this gradient (Figure \@ref(fig:map-iso)).
+The resulting SLIPs indicate the isolation of the basin from the highest astronomical tide.
+This is adjusted to mean sea-level for the compilation of the displacement curve based on the present day tidal range, which is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
 Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
-The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
+The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well-defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
 @romundset2019; @romundset2011].
@@
 
-The archaeological data compiled for the analysis consists of all excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)).
-These number a total `r nrow(sites)` sites.
+The archaeological data compiled for the analysis consists of excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)).
+These number `r nrow(sites)` sites.
 Of these, `r length(unique(rcarbon$site_name))` sites are associated with a total of `r nrow(rcarbon)` radiocarbon dates.
-Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability.
+Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE), with 95 % probability.
 These sites and ^14^C-dates form the basis for the analysis.
@@
 These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape.
-Thus, while there would be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level, as they would not have been in use when located under water.
-It could therefore seem that the Løvås sites represent a case where the archaeological material indicates discrepancies in the geological reconstruction of shoreline displacement in the area.
+While there would be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level, as they would not have been in use when located under water.
+It could therefore seem that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geological reconstruction of shoreline displacement in the area.
 
@@
 Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations.
-The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, while 95% of the values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
-That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
-While these values remain the same, the mean and standard deviation is constrained slightly when only considering the Mesolithic dates.
-While the median for horizontal and topographic distances is only `r round(median(distances2$hordist))`m and `r round(median(distances2$topodist))`m respectively, it is also evident that their range is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
+The resulting best point estimate for the vertical distance between sites and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, while 95% of the values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
+That is, for 95% of the cases, the shoreline was simulated to be situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
+While these values remain the same, the mean and standard deviation is constrained slightly when only considering the Mesolithic dates in the seceond row of Figure \@ref(fig:results2).
+Furthermore, while the median for horizontal and topographic distances is only `r round(median(distances2$topodist))`m across all plots in Figure \@ref(fig:results2), it is also evident that their range is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
 
@@
 While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data.
-However, this can at least in part be related to methodological factors, where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
+However, this can at least in part be related to methodological factors where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
 If one accepts this, having derived an exponential decay function for describing the vertical distance between sites and shoreline can be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:
@@
 Where x is [...] In Figure \@ref(fig:shoredate) this formula is used to shoreline date the same sites from where this relationship was derived.
-Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation.
-The result of comparing the synchroneity of the dates using the method presented by @parnell2008 is given in Figure \@ref(fig:shoredate)B.
+Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation over the site limits.
+The result of comparing the radiocarbon and shoreline dates using the method presented by @parnell2008 for assessing synchroneity is given in Figure \@ref(fig:shoredate)B.
 Here, 100,000 age samples drawn from each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date.
-The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered synchronous.
+The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement.
 The result of this procedure indicates that the shoreline date correspond with the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%).
@@
 For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B), and in the simulation of the distance to the shoreline this was simulated to be 0m across virtually all of the 1000 simulation runs for both sites (see supplementary material).
-If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of the two Løsvås sites and Gunnarsrød 5 are instead due to extraneous factors and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.
+If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of Gunnarsrød 5 is instead due to extraneous factors related the DTM and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.
+Another implication of the failed dating of these two sites could be that sites from the absolute earliest part of the Mesolithic should be dated using a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate, to accommodate for the apparent deviation from the shoreline for sites older than c.
+6500 BCE (cf.
+Figure \@ref(fig:results)).
 
@@
 Negative values have been set to zero.
-The first row only includes dates older than 2500 BCE and the second row only dates older than 4000 BCE.", out.width = "525px", fig.align="center", regfloat=TRUE}
+The first row only includes simulated results older than 2500 BCE and the second row only results older than 4000 BCE.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/results2.png"))
@@
 
-```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory.
+```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory using the method proposed here.
 A) The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black.
@@
 
-```{r redate, echo = FALSE, fig.cap = "Shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
-The range of the 95\\% HDR in black are comapred to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
+```{r redate, echo = FALSE, fig.cap = "Re-dating of previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
+The range of the 95\\% HDR in black are compared to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/redate.png"))
@@
 
-Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further test and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted and refined.
+Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further evaluate and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted and refined.
 Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located.
@@
 
-This approach will be useful either when shoreline dating a large number of sites where no other information than site elevation is taken into account [e.g.
-@breivik2014; @solheim2018; @fossum2020; @nielsen2021b], which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.
+This approach will be useful either when shoreline dating a large number of sites where no other information than site elevation is taken into account , which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.
 

### 2022-03-08 analysis/supplementary-materials/supplementary.Rmd
Slight adustment to figures and more writing

@@
 The sites are presented in alphabetical order.
-excavated sites and radiocarbon dates, and spatial data for site limits and site features are available in the online repoistory for the paper at https://osf.io/7F9SU
-
-Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan [@sjursjeike], Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@østmo1975].
+All underlying and derived data are available in the online repository for the paper at https://osf.io/7F9SU
+
+Stone Age sites for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975], Nedre Holtan [@sjursjeike], Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@østmo1975].
 These sites are therefore not included in the analysis.

### 2022-03-08 analysis/script/08script.R

@@ -29,7 +29,7 @@ dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
 brunlanes <- sitespol %>%
   filter(str_detect(site_name, 'Pauler|Bakke|Sky 1')) %>%
   st_zm() %>%
-  rename("name" = "site_name",
+  dplyr::rename("name" = "site_name",
          "limit" = "name")
 
 # Use elevations provided by Jaksland
@@ -269,9 +269,9 @@ ggsave(file = here("analysis/figures/brunlanes.png"),
 psdates <- read.csv((here("analysis/data/raw_data/previous_shoreline_dates.csv"))) %>%
   mutate(start = start * -1,
          end = end * -1) %>%
-  rename("site_name" = "name")
+  dplyr::rename("site_name" = "name")
 
-sites_sl <- site_limits %>% filter(radiocarbon == "f" | radiocarbon == "t" && rcarb_cor == "f")
+sites_sl <- site_limits %>% filter(radiocarbon == "f" | radiocarbon == "t" & rcarb_cor == "f")
 
 sites_sl <- sites_sl %>%
   filter(name %in% psdates$site_name)

### 2022-03-09 analysis/paper/paper.Rmd
Slight adjustments to figures. Tried adjusting lambda as well

@@
 
-The method of shoreline dating has ranged from being met with scepticism as related to the degree sites would have been consistently shore-bound, to being deemed a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer little more than a earliest possible date for when a site could have been in use.
-The most common application in Norway has arguably been to use the method to provide a very approximate date for the occupation of the sites, typically in combination with other dating methods [see for example chapters in @; ; and below].
-Recently the method has also been used idenpendently to date a larger number sites to get a general impression of site frequency over time, typically by aggregating point estimates of shoreline dates in 200 or 500 year bins [e.g.
-@breivik2014; @solheim2018; @fossum2020; @nielsen2021b].
-In his discussion of the method, @nordqvist1999 -- in part drawing on @johansen1963 -- argues that there can be little doubt about the general applicability of the method, but what is less clear is the level of reliability and chronological resolution that it can offer.
+The method of shoreline dating has ranged from being met with scepticism as related to the degree sites would have been consistently shore-bound, to being characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer little more than a earliest possible date for when a site could have been in use.
+The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, typically in combination with other dating methods [see for example chapters in @; ; and below].
+Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time, typically by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [e.g.
+@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b].
+In his discussion of the method, @nordqvist1999, in part drawing on @johansen1963, argues that there can be little doubt about the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer.
 
@@
 However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227].
-@nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic of the Oslo fjord area is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g.
+@nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g.
 @amundsen2006; @solheim2012, 74; @ostmo1988].
@@
 Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
-@åstveit2018; @solheim2020; see also @akerlund1995].
+@akerlund1995; @åstveit2018; @solheim2020; see also @bjerck2008a; @kleppe1985].
 This approach has a couple of limitations.
-First of all, the displacement curves are commonly applied directly to larger study areas, with only a few studies having taken the variable uplift-rates into account when performing this comparison [e.g.
-@fossum2020; @møller1987; @persson2008].
+First of all, the displacement curves are sometimes applied directly to larger study areas, with only some studies having taken the variable uplift-rates into account when performing this comparison [e.g.
+@åstveit2018; @fossum2020; @møller1987; @persson2008].
 Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis.
@@
 Of these, `r length(unique(rcarbon$site_name))` sites are associated with a total of `r nrow(rcarbon)` radiocarbon dates.
-Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE), with 95 % probability.
+Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE), with 95% probability.
 These sites and ^14^C-dates form the basis for the analysis.
@@
 
-Three of the sites have been associated with farming activities, either in directly or in the form building structures.
+Three of the sites have been associated with farming activities, either directly or in the form building structures.
 The first is Nordby 1, at which the ^14^C-dates are associated with a Late Neolithic long-house [@gjerpe2008b].
-The Middle Neolithic phase at Kvastad A2 [@stokke2018b] and Late Neolithic phase at Nauen A [@persson2008] are both related to farming.
+The Middle Neolithic phase at Kvastad A2 [@stokke2018b] and Late Neolithic phase at Nauen A [@persson2008] are both directly related to farming.
 Both of these sites also have radiocarbon dates and lithic inventory associated with Mesolithic forager activities.
@@
 
-Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, the simulations show that the sites tend to have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)).
+Overall, as is indicated by the measures for central tendency and the almost solid line along the 0m mark on the y-axes, the simulations show that the sites tend to have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)).
 Some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included, but if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
@@
 The deviation from the shoreline date at Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langemyr and Kvastad A2.
-For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B), and in the simulation of the distance to the shoreline this was simulated to be 0m across virtually all of the 1000 simulation runs for both sites (see supplementary material).
+For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B).
 If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of Gunnarsrød 5 is instead due to extraneous factors related the DTM and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.
-Another implication of the failed dating of these two sites could be that sites from the absolute earliest part of the Mesolithic should be dated using a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate, to accommodate for the apparent deviation from the shoreline for sites older than c.
-6500 BCE (cf.
+Another implication of the failed dating of these two sites could be that for the earliest part of the Mesolithic, a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used.
+to accommodate the apparent deviation from the shoreline for sites older than c.
+7000 BCE (cf.
 Figure \@ref(fig:results)).
+The
 
@@
 
-```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using dates corresponding to the site inventory.
+```{r results2, echo = FALSE, fig.cap = "Histograms showing the simualted distance from the shoreline using dates corresponding to the site inventory.
 Negative values have been set to zero.
@@
 
-To further explore the implementation for shoreline dating outline above, excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the method outlined above (Figure \@ref(fig:redate)).
-These are compared to the dates originally proposed in the excavation reports for the sites.
+To further explore the implementation for shoreline dating outline above, excavated shoreline dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the method outlined above (Figure \@ref(fig:redate)).
+The resulting dates are compared to those originally proposed in the excavation reports for the sites.
 To avoid issues with recent disturbances in the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
@@
 However, some of the date ranges extend well beyond major chronological divisions, even the entire Stone Age, and could be severely and securely constrained with only cursory reference to typology.
-This thus points to the possibility of drawing on other temporal data such as typological indicators or radiocarbon dates, for example within a Bayesian framework, to improve the precision of the dating of the sites.
+This thus points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to improve the precision of the dating of the sites.
 
-```{r redate, echo = FALSE, fig.cap = "Re-dating of previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
-The range of the 95\\% HDR in black are compared to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
+```{r redate, echo = FALSE, fig.cap = "Re-dating previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
+The range of the 95\\% HDR in grey are compared to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/redate.png"))
@@
 This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [e.g.
-@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@blankholm2020].
-This is also related to the fact that while mean sea-level is used for dating the sites, a consideration of the tidal range could possibly have implications for the site location relative to the shoreline, depending on the topography.
+@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [helskog1978; @blankholm2020].
+This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range could possibly also have implications for the site location relative to the shoreline, depending on the topography [@helskog1978].
 If such locational patterns can be discerned, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site beyond its location is available.

### 2022-03-09 analysis/script/08script.R

@@ -271,7 +271,7 @@ psdates <- read.csv((here("analysis/data/raw_data/previous_shoreline_dates.csv")
          end = end * -1) %>%
   dplyr::rename("site_name" = "name")
 
-sites_sl <- site_limits %>% filter(radiocarbon == "f" | radiocarbon == "t" & rcarb_cor == "f")
+sites_sl <- site_limits %>% filter(radiocarbon == "f" | radiocarbon == "t" & rcarb_cor == "f" | name == "Viulsrød 2")
 
 sites_sl <- sites_sl %>%
   filter(name %in% psdates$site_name)
@@ -290,7 +290,7 @@ for(i in 1:nrow(sites_sl)){
                                  sites = sites_sl,
                                  iso = isobases,
                                  expratio = expfit$estimate,
-                                 reso = 0.1,
+                                 reso = 0.01,
                                  specified_elev = (sites_sl$min_elev[i]+
                                                  sites_sl$max_elev[i])/2)
 }
@@ -303,7 +303,7 @@ hdrdat <- bdates %>%  group_by(site_name) %>%
   mutate(year_min = min(hdrcde::hdr(den = list("x" = years, "y" = probability),
                                     prob = 95)$hdr),
          year_max = max(hdrcde::hdr(den = list("x" = years, "y" = probability), prob = 95)$hdr),
-         year_median = hdrcde::hdr(den = list("x" = years, "y" = probability), prob = 95)$mode)
+         year_median = median(hdrcde::hdr(den = list("x" = years, "y" = probability), prob = 95)$mode))
 
 
   r <- hdr(x, prob = 95)
@@ -319,10 +319,10 @@ r <- data.frame("ymin" = min(r$hdr), "median" = r$mode,
     year_median = median(years, na.rm = TRUE))
 
 hdrdat <- hdrdat[order(hdrdat$year_median, decreasing = TRUE),]
-hdrdat$site_name <- factor(hdrdat$site_name, levels = hdrdat$site_name)
+hdrdat$site_name <- factor(hdrdat$site_name, levels = unique(hdrdat$site_name))
 
 psdates <- psdates[order(match(psdates$site_name, hdrdat$site_name)),]
-psdates$site_name <- factor(psdates$site_name, levels = psdates$site_name)
+psdates$site_name <- factor(psdates$site_name, levels = unique(psdates$site_name))
 
 psdates1 <- psdates %>%  filter(start != end)
[…]

### 2022-03-10 analysis/paper/paper.Rmd
Slight adjustment to figures, more writing

@@
 ```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from c.
-17--8 kyr BCE (data from Hughes et al.
-2016, but see Romundset et al.
-2019).
-The study area defined later in the text is marked with a red outline.", out.width = "400px", fig.align="center", regfloat=TRUE}
+17--8 kyr BCE.
+The study area defined later in the text is marked with a red outline (deglaciation data from Hughes et al.
+2016, although see Romundset et al.
+2019 for the study area).", out.width = "400px", fig.align="center", regfloat=TRUE}
 
@@
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [discussed in @berg-hansen2009; @bergsvik2009].
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; although see @gundersen2013; @mjærum2018; @schulke2020].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 If the site is situated on a ledge in a steep and jagged area of the present day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft.
-Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few [see also @ilves2011], and are typically conducted on a site-by-site or project-by-project basis, meaning the assumed shore-bound site location is largely based on a mosaic of smaller scale investigations.
+Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @ilves2011].
 
@@
 They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
-However, where there was a discrepancy the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates.
+However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates.
 This is based on typological and technological characteristics of the assemblages.
-Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate [e.g.
-@persson2008; @schulke2020; see also ].
-However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas evidenced by artefact inventories or multiple site features, and not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates.
+Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results can be difficult to evaluate [e.g.
+@persson2008; @schulke2020].
+However, this distinction is not deemed critical here, as what is of interest is the relationship between the prehistoric shoreline and settlements and tool-production areas evidenced by artefact inventories or multiple site features.
+Not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates.
 The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.
@@
 If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of Gunnarsrød 5 is instead due to extraneous factors related the DTM and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.
-Another implication of the failed dating of these two sites could be that for the earliest part of the Mesolithic, a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used.
-to accommodate the apparent deviation from the shoreline for sites older than c.
-7000 BCE (cf.
+However, a different implication of the failed dating of these two sites could be that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for the earliest part of the Mesolithic (cf.
 Figure \@ref(fig:results)).
-The
 
@@
 
-To further explore the implementation for shoreline dating outline above, excavated shoreline dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the method outlined above (Figure \@ref(fig:redate)).
+To further explore the implementation for shoreline dating outlined above, excavated and shoreline-dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the method (Figure \@ref(fig:redate)).
 The resulting dates are compared to those originally proposed in the excavation reports for the sites.
 To avoid issues with recent disturbances in the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
-This is an illustrative, but unfair exercise for a few reasons.
-First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit but undefined uncertainty range.
+The results highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% intervals for the dates.
+In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.
+This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (Figure \@ref(fig:map-iso)B).
+
+The comparison with previously reported dates is an illustrative, but unfair exercise for a few reasons.
+First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit, but undefined uncertainty range.
 Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change.
 Finally, they are often only meant to indicate a lower bound for when the sites could have been in use.
-I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner.
-The results from applying the method devised here highlights the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% intervals for the dates.
-In some cases it provides a very precise date range and in others it offers little more than a *terminus post quem*.
-This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise date ranges (Figure \@ref(fig:map-iso)B).
-However, some of the date ranges extend well beyond major chronological divisions, even the entire Stone Age, and could be severely and securely constrained with only cursory reference to typology.
-This thus points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to improve the precision of the dating of the sites.
+Nonetheless, I do think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although not always, e.g.
+@roalkvam2022].
+That being said, the results do indicate that shoreline dating has at times been applied with an exaggerated degree of precision, and while the implications of a more stable RSL-change for the precision of shoreline dating is known, it appears to be somewhat under-appreciated.
+Furthermore, some of the date ranges resulting from the method outlined here extend well beyond major chronological divisions, even into the Iron Age, and could be severely and securely constrained with only cursory reference to typology.
+This is arguably an exercise  This also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to improve the precision of the dates.
 
@@
 
-Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
+Not least following from the fact that few Preboreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
 @bang-andersen2012; @breivik2014; @glorstad2016].
@@
 Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline for when the site was in use.
-Such approaches would make less of an impact in this setting, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change as been less severe (cf.
+Such approaches would make less of an impact in this setting, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf.
 Figure \@ref(fig:redate)).

### 2022-03-10 analysis/script/07script.R

@@ -88,7 +88,8 @@ feature_sites <- c("Langangen Vestgård 7", "Vallermyrene 2")
 load(here("analysis/data/derived_data/07data.RData"))
 
 sdates <- bind_rows(shorelinedates) %>% group_by(site_name) %>%
-  filter(cumsum(replace_na(probability, 0)) & probability != 0)
+  filter(cumsum(replace_na(probability, 0)) < 0.99  &
+           probability != 0)
 
 
 # Radiocarbon dates corresponding to site inventory and older than 2500 BCE
@@ -233,16 +234,65 @@ for(i in 1:nrow(esites)){
                                       specified_elev = NA)
 }
 
-shdates <- bind_rows(shoredates) %>% group_by(site_name) %>%
+shdates <- bind_rows(shoredates) %>% group_by(site_name)
   filter(cumsum(replace_na(probability, 0))  &
            probability != 0)
 
-ggplot()+
+cors <- filter(corsites, site_name %in% unique(shdates$site_name))
+
+ggplot() +
 ggridges::geom_ridgeline(data = shdates,
                          aes(x = years, y = site_name,
                              height = probability * 200),
-                         colour = "grey", fill = "grey")
+                         colour = "grey", fill = "grey") +
+  ggridges::geom_ridgeline(data = cors,
+                           aes(x = dates, y = site_name,
+                               height = probabilities * 50),
+                           colour = "red", fill = NA)
 
+agedif2 <- setNames(data.frame(matrix(ncol = 2, nrow = 0)), c("age_diff", "site_name"))
+for(i in 1:length(unique(shdates$site_name))){
+  samplingframe <- list(x = shdates[shdates$site_name ==
+                                             unique(shdates$site_name)[i],
+                                           "years"][[1]],
+                        y = shdates[shdates$site_name ==
+                                             unique(shdates$site_name)[i],
+                                           "probability"][[1]])
+  sdate <- with(samplingframe, sample(x, size = 100000, prob = y,
+                                      replace = TRUE))
+
+  samplingframe <- list(x = corsites[corsites$site_name ==
+                                       unique(shdates$site_name)[i],
+                                     "dates"]$dates,
+                        y = corsites[corsites$site_name ==
+                                       unique(shdates$site_name)[i],
+                                     "probabilities"]$probabilities)
+
[…]

### 2022-03-10 analysis/script/08script.R

@@ -43,7 +43,7 @@ for(i in 1:nrow(brunlanes)){
                                         disp_curves = displacement_curves,
                                         sites = brunlanes,
                                         iso = isobases,
-                                        expratio = expfit2$estimate,
+                                        expratio = expfit3$estimate,
                                         siteelev = "min",
                                         reso = 0.001,
                                  specified_elev = brunlanes$elevation[i])
@@ -75,6 +75,10 @@ for(i in 1:length(unique(bdates$site_name))){
   hdrdat <- rbind(hdrdat, groupdat)
 }
 
+bdates <- bdates %>%  group_by(site_name) %>%
+  filter(cumsum(replace_na(probability, 0)) < 0.99999  &
+           probability != 0)
+
 # Lasse Jaksland's dates of the Brunlanes sites. He provided a maximum age,
 # and argued that the relative uncertainty would be around 50 years.
 jaksland <- data.frame(site_name = sort(unique(bdates$site_name)),
@@ -93,19 +97,19 @@ dateplot <- ggplot(data = hdrdat, aes(x = year_median, y = reorder(site_name, -y
                            colour = "grey", fill = "grey") +
   geom_linerange(data = hdrdat, aes(xmin = start, xmax = end,
                                       y = site_name), size = 0.5, col = "black",
-                 position = position_dodge(width = 0.1, preserve = 'single'),
+                 position = position_dodge(width = 0.12, preserve = 'single'),
                  inherit.aes = FALSE) +
   geom_linerange(data = jaksland, aes(xmin = earliest_date, xmax = latest_date1,
-                                      y = site_name), size = 1.75, col = "red",
-                 position = position_dodge(width = 0.4, preserve = 'single'),
+                                      y = site_name), size = 2, col = "red",
+                 position = position_dodge(width = 0.5, preserve = 'single'),
                  inherit.aes = FALSE) +
   geom_linerange(data = jaksland, aes(xmin = earliest_date, xmax = latest_date2,
-                                      y = site_name), size = 0.5, col = "red",
-                 position = position_dodge(width = 0.4, preserve = 'single'),
[…]

### 2022-03-10 analysis/paper/paper.Rmd
Using saa style guide. Trying also to include all dates in the comparison

@@
 title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
+header-includes:
+   - \usepackage{lineno}
+   - \usepackage{afterpage}
+   - \usepackage{fancyhdr}
+   - \linenumbers
 author:
@@
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+csl: "../templates/saa.csl" # Insert path for the bib-style
 ---
+
+<!--- Page numbers in margin to avoid conflict with figure captions --->
+\fancypagestyle{plain}{%
+  \fancyhf{}
+  \fancyfoot[R]{\leavevmode\rlap{\thepage~}}
+}
+\pagestyle{plain}
 
@@
 
-The method of shoreline dating has ranged from being met with scepticism as related to the degree sites would have been consistently shore-bound, to being characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer little more than a earliest possible date for when a site could have been in use.
+The method of shoreline dating has been met with scepticism as related to the degree sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [@breivik2018a, 174].
 The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, typically in combination with other dating methods [see for example chapters in @; ; and below].
@@
 @breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b].
-In his discussion of the method, @nordqvist1999, in part drawing on @johansen1963, argues that there can be little doubt about the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer.
+In his discussion of the method, @nordqvist1999, in part drawing on @johansen1963, argues that there can be little doubt concerning its general applicability -- what is less clear is the level of reliability and chronological resolution that the method can offer.
 
@@
 @crema2010, 1125].
-Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography.
-At 1000 simulation runs the analysis of each site took up to a few hours, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.
+Hovland 5 was chosen for this evaluation as it has a fairly uncertain date, and is located in area of quite complex topography.
 
@@
 Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation over the site limits.
-The result of comparing the radiocarbon and shoreline dates using the method presented by @parnell2008 for assessing synchroneity is given in Figure \@ref(fig:shoredate)B.
-Here, 100,000 age samples drawn from each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date.
+The result of comparing the radiocarbon and shoreline dates using the method presented by @parnell2008 for assessing the synchroneity between dates is given in Figure \@ref(fig:shoredate)B.
+Here, 100,000 age samples drawn from the probability density function of each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date.
 The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement.
-The result of this procedure indicates that the shoreline date correspond with the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%).
+The result of this procedure indicates that the shoreline date correspond to the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%).
 The deviation from the shoreline date at Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langemyr and Kvastad A2.
@@
 If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of Gunnarsrød 5 is instead due to extraneous factors related the DTM and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.
-However, a different implication of the failed dating of these two sites could be that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for the earliest part of the Mesolithic (cf.
+However, another possible implication of the failed dating of these two sites could be that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites from the earliest part of the Mesolithic (cf.
 Figure \@ref(fig:results)).
@@
 To further explore the implementation for shoreline dating outlined above, excavated and shoreline-dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the method (Figure \@ref(fig:redate)).
-The resulting dates are compared to those originally proposed in the excavation reports for the sites.
-To avoid issues with recent disturbances in the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
+The resulting dates are compared to those originally proposed in the excavation reports for the sites (the numerical results are available in the supplementary material).
+To avoid issues with recent disturbances on the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
 The results highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% intervals for the dates.
 In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.
-This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (Figure \@ref(fig:map-iso)B).
+This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf.
+Figure \@ref(fig:map-iso)B).
 
@@
 Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change.
-Finally, they are often only meant to indicate a lower bound for when the sites could have been in use.
-Nonetheless, I do think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although not always, e.g.
+Finally, they are sometimes only meant to indicate a lower bound for when the sites could have been in use.
+Nonetheless, the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g.
 @roalkvam2022].
-That being said, the results do indicate that shoreline dating has at times been applied with an exaggerated degree of precision, and while the implications of a more stable RSL-change for the precision of shoreline dating is known, it appears to be somewhat under-appreciated.
-Furthermore, some of the date ranges resulting from the method outlined here extend well beyond major chronological divisions, even into the Iron Age, and could be severely and securely constrained with only cursory reference to typology.
-This is arguably an exercise  This also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to improve the precision of the dates.
+That being said, the results do also indicate that shoreline dating has at times been applied with an exaggerated degree of precision, and while the implications of a more stable RSL-change for the duration and re-use of site locations, and consequently the precision of shoreline dating is well known, it appears to be somewhat under-appreciated.
+Some of the date ranges resulting from the method outlined here clearly extend well beyond major chronological divisions, even into the Iron Age, and could be severely and securely constrained with only cursory reference to typology.
+However, while this is obvious in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be made explicit.
+This final point also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to further improve the precision of the dates that can be achieved with shoreline dating.
 
@@
 @bang-andersen2012; @breivik2014; @glorstad2016].
-The shoreline dated pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a].
+The shoreline-dated Preboreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a].
 These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision use of the sites after the sea retreated any significant distance from their location.
@@
 A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \@ref(fig:brunlanes)A.
-The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf.
+The sites have been dated using what @jaksland2014 gives as the lowest elevation of finds at each site, and by employing a exponential decay ratio of , to allow for more deviance in the distance between site and shoreline.
+This corresponds to the decay ratio for sites older than 7000 BCE in Figure \@ref(fig:results2).
+
+The small discrepancies between the achieved results mainly follow from the fact that a slightly updated version of the local displacement curve is applied here [cf.
 @sorenseninpress].
@@
 Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap.
-This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges.
+This information is not integrated in the approach outlined here, but could justify further reducing the uncertainty ranges.
 Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does nonetheless result in very precise dates.
@@
 This example, on the other hand, highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
-One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the steepness of the slope to the shoreline.
-Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline for when the site was in use.
+One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline.
+Based on this, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use.
 Such approaches would make less of an impact in this setting, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf.
@@
 @lewis2021].
-Thirdly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves away from the isobases of the displacement curves.
-Finally, the definition of the site limits nor the elevation range over which these extend was not considered.
+Thirdly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves futher away from the isobases of the displacement curves.
+Finally, the question of how site limits are defined nor the elevation range over which these extend was given much consideration.
+While each of these factors will have variable but likely not too large of an impact on the results, they clearly represent dimensions which would all benefit from further exploration.
 
@@
 2400 BCE -- in clear agreement with the literature.
-Furthermore, based on the quantitative nature of these findings, a refined method for the shoreline dating of Stone Age sites has been proposed.
-This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula x to account for the distance between the sites and the shoreline.
-When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratios identified in Figure \@ref(fig:redate) to characterise this relationship.
+Furthermore, based on the quantitative nature of these findings, a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed.
+This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula X to account for the distance between the sites and the shoreline.
+When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio identified in Figure \@ref(fig:redate) to characterise this relationship.
 However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data such as radiocarbon dates and typological indicators in the artefact inventories.
@@
 This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [e.g.
-@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [helskog1978; @blankholm2020].
+@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@helskog1978; @blankholm2020].
 This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range could possibly also have implications for the site location relative to the shoreline, depending on the topography [@helskog1978].
-If such locational patterns can be discerned, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site beyond its location is available.
+Dimensions such as these was given a cursory treatment here, with the estimation of the horizontal and topographic distance to the shoreline, but these dimensions could also be investigated further.
+If patterns related to such locational patterns can be discerned, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the site beyond its location is available.
 
 Finally, this analysis was based on a simulation approach for integrating multiple sources of spatio-temporal uncertainty.
-Here this was simply used to inform the question of the distance between sites and shorelines.
-However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
-
-This approach will be useful either when shoreline dating a large number of sites where no other information than site elevation is taken into account , which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.
+Here this was simply used to inform the question of the distance between sites and the shore-line.
+However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
 

### 2022-03-11 analysis/paper/paper.Rmd
Accommodating multi-phased comparison

@@
 ```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory using the method proposed here.
-A) The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black.
-These are plotted against the oldest radiocarbon-informed phase for each site in red.
-B) Using the method described by Parnell et al.
-(2008), the synchroneity of the dates is evaluated.
-When the range of the 95\\% HDR for age difference crosses zero the dates are considered synchronous.", out.width = "525px", fig.align="center", regfloat=TRUE}
+The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black.
+These are plotted against the modelled radiocarbon dates for each site, which are given colour from oldest to youngest phase.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
+```
+
+```{r shoredatediff, echo = FALSE, fig.cap = "The agreement between the shoreline dates and radiocarbon dates given in Figure ref(fig:shoredate) are evaluated using the method described by Parnell et al.
+(2008).
+When the range of the 95\\% HDRs for age difference crosses zero the dates are considered synchronous.
+The colour coding follows the divison of phases given in Figure ref(fig:shoredate).", out.width = "525px", fig.align="center", regfloat=TRUE}
+knitr::include_graphics(here("analysis/figures/shoredate2.png"))
 ```
@@
 Some of the date ranges resulting from the method outlined here clearly extend well beyond major chronological divisions, even into the Iron Age, and could be severely and securely constrained with only cursory reference to typology.
-However, while this is obvious in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be made explicit.
+However, while this is obvious in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed.
 This final point also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to further improve the precision of the dates that can be achieved with shoreline dating.
@@
 
-Not least following from the fact that few Preboreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
+Not least following from the fact that relatively few Preboreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway [@åstveit2018; @damlien2018; @kleppe2018], the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
 @bang-andersen2012; @breivik2014; @glorstad2016].

### 2022-03-11 analysis/paper/paper.Rmd
Trying to deal with multiple phases at each site

@@
 The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black.
-These are plotted against the modelled radiocarbon dates for each site, which are given colour from oldest to youngest phase.", out.width = "525px", fig.align="center", regfloat=TRUE}
+These are plotted against the modelled radiocarbon dates which are given colour from oldest to youngest phase for each site.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 
-```{r shoredatediff, echo = FALSE, fig.cap = "The agreement between the shoreline dates and radiocarbon dates given in Figure ref(fig:shoredate) are evaluated using the method described by Parnell et al.
+```{r shoredatediff, echo = FALSE, fig.cap = "Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure ref(fig:shoredate), based on the method described by Parnell et al.
 (2008).
-When the range of the 95\\% HDRs for age difference crosses zero the dates are considered synchronous.
+When the range of the 95\\% HDRs for age difference crosses zero and are coloured black, the shoreline- and radiocarbon dates are considered synchronous.
 The colour coding follows the divison of phases given in Figure ref(fig:shoredate).", out.width = "525px", fig.align="center", regfloat=TRUE}
@@
 ```{r redate, echo = FALSE, fig.cap = "Re-dating previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
-The range of the 95\\% HDR in grey are compared to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
+The range of the 95\\% HDRs in grey are compared to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/redate.png"))
@@
 A) The result of applying the approach to shoreline dating outlined above.
+The shoreline date in grey is underlined with the 95\\% HDR in black.
 Dates provided by Jaksland (2014) are plotted in red.
-The box indicates a 50 year uncertainty range which in combination with the red line extends 200 years.
-Follwing Jaksland's approach these start from the earliest possible date.
+Follwing his approach, the box indicates a 50 year uncertainty range which in combination with the red line extends 200 years.
 B) Map showing the centroids of the Pauler sites and Sky 1.
@@
 
-It is worth highlighting some sources of likely variation and uncertainty that have not been considered here.
-First of all the DTM has only been corrected for major modern disturbances.
+It is worth highlighting some limitations and sources of likely variation and uncertainty that have not been considered here.
+First of all the sample size is quite strained, and the future addition of only a few sites might skew the picture significantly.
+Secondly, the DTM has only been corrected for major modern disturbances.
 This means that erosion, although likely not that prevalent, has not been taken into account.
-Secondly, the DTM has a vertical error [cf.
+Thirdly, the DTM has a vertical error of, which could also benefit from being integrated in the analysis [cf.
 @lewis2021].
-Thirdly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves futher away from the isobases of the displacement curves.
-Finally, the question of how site limits are defined nor the elevation range over which these extend was given much consideration.
-While each of these factors will have variable but likely not too large of an impact on the results, they clearly represent dimensions which would all benefit from further exploration.
+Fourthly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves further away from the isobases of the displacement curves.
+Fifthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration.
+Finally, the division of phases at each sites was here simply done by treating  While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration.
 
@@
 
-Finally, this analysis was based on a simulation approach for integrating multiple sources of spatio-temporal uncertainty.
+Finally, this analysis employed a simulation approach to integrate multiple sources of spatio-temporal uncertainty.
 Here this was simply used to inform the question of the distance between sites and the shore-line.
-However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
+However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
 

### 2022-03-11 analysis/script/07script.R

@@ -5,8 +5,8 @@ library(terra)
 library(sf)
 library(hdrcde)
 
-# For reproducibility
-set.seed(123)
+# For reproducibility and the meaning of life
+set.seed(42)
 
 # Load required functions and prepared data
 source(here("analysis/script/04script.R"))
@@ -156,11 +156,11 @@ shrplt <- ggplot(data = hdrdat,
   ggridges::geom_ridgeline(data = corgroupsn2,
                            aes(x = dates, y = site_name,
                                height = probabilities * 50),
-                           col = "darkorange", fill = NA) +
+                           col = "blue", fill = NA) +
   ggridges::geom_ridgeline(data = corgroupsn3,
                            aes(x = dates, y = site_name,
                                height = probabilities * 50),
-                           col = "blue", fill = NA) +
+                           col = "darkorange", fill = NA) +
   ggridges::geom_ridgeline(data = corgroupsn4,
                            aes(x = dates, y = site_name,
                                height = probabilities * 50),
@@ -218,8 +218,8 @@ agedfplt2 <-  agedfplt + stat_summary(data = agedif2, aes(y = site_name,
                                       x = age_diff, colour = cross),
                                       fun.data = hdr_func, geom = "errorbar",
                                       size = 0.5) +
-  scale_colour_manual(values = c("darkorange", "black")) +
-  theme(panel.border = element_rect(colour = "darkorange"),
+  scale_colour_manual(values = c("blue", "black")) +
+  theme(panel.border = element_rect(colour = "blue"),
         axis.title.y = element_blank(),
         axis.text.y = element_blank(),
         axis.ticks.y = element_blank(),
@@ -229,8 +229,8 @@ agedfplt3 <-  agedfplt + stat_summary(data = agedif3, aes(y = site_name,
                                       x = age_diff, colour = cross),
                                       fun.data = hdr_func, geom = "errorbar",
                                       size = 0.5) +
-  scale_colour_manual(values = c("blue","black")) +
-  theme(panel.border= element_rect(colour = "blue"),
+  scale_colour_manual(values = c("darkorange","black")) +
+  theme(panel.border= element_rect(colour = "darkorange"),
         axis.title.y = element_blank(),
         axis.text.y = element_blank(),
         axis.ticks.y = element_blank(),
[…]

### 2022-03-12 analysis/paper/paper.Rmd
Working out how to solve age-difference plot

@@
 The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.
-However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites.
+However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise method by which one can hope to date the sites.
 Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age.
@@
 @amundsen2006; @solheim2012, 74; @ostmo1988].
-Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.
+Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and finally a decisive shift around 2400 BCE.
 
@@
 That is, for 95% of the cases, the shoreline was simulated to be situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
-While these values remain the same, the mean and standard deviation is constrained slightly when only considering the Mesolithic dates in the seceond row of Figure \@ref(fig:results2).
-Furthermore, while the median for horizontal and topographic distances is only `r round(median(distances2$topodist))`m across all plots in Figure \@ref(fig:results2), it is also evident that their range is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
+While these values remain the same when only the Mesolithic dates are included in the second row of Figure \@ref(fig:results2), the mean and standard deviation is slightly constrained.
+Furthermore, while the median for horizontal and topographic distance is only `r round(median(distances2$topodist))`m across all plots in Figure \@ref(fig:results2), the variation in statistics for dispersion is also highly more variable, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
 
@@
 While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data.
-However, this can at least in part be related to methodological factors where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
+However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
 If one accepts this, having derived an exponential decay function for describing the vertical distance between sites and shoreline can be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:
@@
 Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation over the site limits.
-The result of comparing the radiocarbon and shoreline dates using the method presented by @parnell2008 for assessing the synchroneity between dates is given in Figure \@ref(fig:shoredate)B.
-Here, 100,000 age samples drawn from the probability density function of each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date.
+The synchroneity between radiocarbon and shoreline dates was then evluated using the method presented by [@parnell2008, Figure \@ref(fig:shoredatediff)].
+Here, 100,000 age samples drawn from the probability density function of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding ^14^C-date.
 The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement.
+The overall success of this procedure is
+If one only includes sites older than 2500 BCE with 95% probability, this improves to, while only including dates older than 4000 BCE with 95% probability increases this further to a success rate of.
+Thus, if there in the dating of sites are reasons to
+
 The result of this procedure indicates that the shoreline date correspond to the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%).
@@
 The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black.
-These are plotted against the modelled radiocarbon dates which are given colour from oldest to youngest phase for each site.", out.width = "525px", fig.align="center", regfloat=TRUE}
+These are plotted against the modelled radiocarbon dates, which are given colour from oldest to youngest phase for each site.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 
-```{r shoredatediff, echo = FALSE, fig.cap = "Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure ref(fig:shoredate), based on the method described by Parnell et al.
-(2008).
-When the range of the 95\\% HDRs for age difference crosses zero and are coloured black, the shoreline- and radiocarbon dates are considered synchronous.
+```{r shoredatediff, echo = FALSE, fig.cap = "Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure ref(fig:shoredate).
+When the range of the 95\\% HDRs for age difference cross zero and are coloured black, the shoreline- and radiocarbon dates are considered synchronous.
 The colour coding follows the divison of phases given in Figure ref(fig:shoredate).", out.width = "525px", fig.align="center", regfloat=TRUE}

### 2022-03-17 analysis/paper/paper.Rmd
Minor edits in text and figures. shoredate2.png still needs some work but will do for now

@@
 
-The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century.
-Shoreline dating has been fundamental to Stone Age archaeology in Norway ever since [e.g.
+The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was established by the end of the 19th century [@degeer1896], and first applied as a dating method at the turn of the century [@brogger1905; @hollender1901].
+Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g.
 @berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @nummedal1923; @mikkelsen1975; @solheim2018].
@@
 
-```{r shoredatediff, echo = FALSE, fig.cap = "Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure ref(fig:shoredate).
-When the range of the 95\\% HDRs for age difference cross zero and are coloured black, the shoreline- and radiocarbon dates are considered synchronous.
-The colour coding follows the divison of phases given in Figure ref(fig:shoredate).", out.width = "525px", fig.align="center", regfloat=TRUE}
+```{r shoredatediff, echo = FALSE, fig.cap = "Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure 9.
+When the range of the 95\\% HDR for age difference crosses zero, indicated by a solid line, the shoreline- and radiocarbon dates are considered synchronous.
+The division and colour coding at the top of the plots reflects the divison of site phases given in Figure 9.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate2.png"))
@@
 Dates provided by Jaksland (2014) are plotted in red.
-Follwing his approach, the box indicates a 50 year uncertainty range which in combination with the red line extends 200 years.
+The box indicates a 50 year uncertainty range which in combination with the red line extends 200 years, following his approach.
 B) Map showing the centroids of the Pauler sites and Sky 1.

### 2022-03-17 analysis/script/07script.R

@@ -5,8 +5,8 @@ library(terra)
 library(sf)
 library(hdrcde)
 
-# For reproducibility and the meaning of life
-set.seed(42)
+# For reproducibility
+set.seed(321)
 
 # Load required functions and prepared data
 source(here("analysis/script/04script.R"))
@@ -114,7 +114,7 @@ corgroupsn <- rdates %>%
   filter(n() != sum(is.na(name)))
 
 # Struggled a bit with aes() for geom_ridgeline, so instead
-# of aes(col = group2), all three remaing groups are seperated here
+# of aes(col = group2), all three remaining groups are separated here
 corgroupsn2 <- corgroupsn %>%  filter(group2 == 2)
 corgroupsn3 <- corgroupsn %>%  filter(group2 == 3)
 corgroupsn4 <- corgroupsn %>%  filter(group2 == 4)
@@ -181,6 +181,8 @@ shrplt <- ggplot(data = hdrdat,
   scale_x_continuous(breaks = c(seq(-10000, -4000, 2000), -2500, 0, 2000) ) +
   theme_bw()
 
+#values = c("black", "#21908CFF", "#A6611A"),
+
 shorelinedates <- bind_rows(shorelinedates) %>%
   filter(site_name %in% unique(corsites$site_name))
 
@@ -189,6 +191,26 @@ agedif2 <- age_diff(shorelinedates, corgroupsn2)
 agedif3 <- age_diff(shorelinedates, corgroupsn3)
 agedif4 <- age_diff(shorelinedates, corgroupsn4)
 
+phasedt <- agedif %>% group_by(site_name) %>%
+  group_by(phase) %>%  summarise(n = n() / 100000) %>%
+  pivot_wider(names_from = phase, values_from = n)
+
+phasedt2 <- agedif2 %>% group_by(site_name) %>%
+  group_by(phase) %>%  summarise(n = n() / 100000) %>%
+  pivot_wider(names_from = phase, values_from = n)
+
+phasedt3 <- agedif3 %>% group_by(site_name) %>%
+  group_by(phase) %>%  summarise(n = n() / 100000) %>%
+  pivot_wider(names_from = phase, values_from = n) %>%
+  mutate(ln = NA)
+
+phasedt4 <- agedif4 %>% group_by(site_name) %>%
+  group_by(phase) %>%  summarise(n = n() / 100000) %>%
+  pivot_wider(names_from = phase, values_from = n) %>%
[…]

### 2022-03-18 analysis/paper/paper.Rmd
Set up agediff figure

@@
 load(here("analysis/data/derived_data/06data.RData"))
+load(here("analysis/data/derived_data/07agediff.RData"))
 source(here("analysis/script/04script.R"), local = knitr::knit_global())
@@
 @breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b].
-In his discussion of the method, @nordqvist1999, in part drawing on @johansen1963, argues that there can be little doubt concerning its general applicability -- what is less clear is the level of reliability and chronological resolution that the method can offer.
+In his review of the method, @nordqvist1999, in part drawing on @johansen1963, argues that there can be little doubt concerning its general applicability -- what is less clear is the level of reliability and chronological resolution that the method can offer.
 
@@
 @amundsen2006; @solheim2012, 74; @ostmo1988].
-Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and finally a decisive shift around 2400 BCE.
+Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and finally a decisive shift at around 2400 BCE.
 
@@
 
-One of the more comprehensive evaluations of the relationship between radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
+One of the more extensive evaluations of the relationship between archaeological radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
 They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
@@
 @persson2008; @schulke2020].
-However, this distinction is not deemed critical here, as what is of interest is the relationship between the prehistoric shoreline and settlements and tool-production areas evidenced by artefact inventories or multiple site features.
+However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas as evidenced by artefact inventories or multiple site features.
 Not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates.
@@
 Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.
-Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.
 
@@
 
-The employed shoreline displacement curves are all based on the so-called isolation basin method [see e.g.
+The employed shoreline displacement curves are all based on the so-called isolation basin method [e.g.
 @kjemperud1986; @romundset2011].
@@
 The resulting SLIPs indicate the isolation of the basin from the highest astronomical tide.
-This is adjusted to mean sea-level for the compilation of the displacement curve based on the present day tidal range, which is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
+This is adjusted to mean sea-level in the compilation of the displacement curve based on the present day tidal range, which is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
 Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
-The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well-defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
+The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
 @romundset2019; @romundset2011].
@@
 Three of the sites have been associated with farming activities, either directly or in the form building structures.
-The first is Nordby 1, at which the ^14^C-dates are associated with a Late Neolithic long-house [@gjerpe2008b].
+The first is Nordby 1 at which the ^14^C-dates are associated with a Late Neolithic long-house [@gjerpe2008b].
 The Middle Neolithic phase at Kvastad A2 [@stokke2018b] and Late Neolithic phase at Nauen A [@persson2008] are both directly related to farming.
 Both of these sites also have radiocarbon dates and lithic inventory associated with Mesolithic forager activities.
-Following from the expected deviance from the settlement patterns that are to characterise forager sites, these agricultural phases are thus treated separately below.
+These agricultural phases are treated separately below, following from their expected deviance from the settlement patterns that are to characterise forager sites.
 Finally, @nielsen2021b has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratsberg [@wenn2012] and Larønningen [@roberg2012] could be related to early agricultural activity in the Oslo fjord region.
 Due to the uncertain and somewhat speculative nature of this suggestion, these are omitted here.
-
-[Table with sites]
 
@@
 The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions.
-To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location from the displacement curves using inverse distance weighting based on the distance between the site and the isobases of the displacement curves [e.g.
+To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves using inverse distance weighting [e.g.
 @conolly2020; @conolly2006, 94--97].
@@
 @conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021].
-In addition to code and original spatial data being available in the research compendium for this paper, it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.
+In addition to code and original spatial data being available in the digital research compendium for this paper, the analysis of each individual site is also available in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.
 
@@
 A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \@ref(fig:example-2)).
-This year then has a corresponding likely elevation range for the contemporaneous shoreline from which an elevation value was drawn uniformly, using intervals of 5cm.
+This year then has a corresponding likely elevation range for the contemporaneous shoreline, from which an elevation value was drawn uniformly, using intervals of 5cm.
 The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing.
@@
 The horizontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon.
-The topographic distance between site and sea was also found by measuring the shortest distance while taking into account the slope of the terrain on the DTM.
+The topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM.
 This was done using the topoDistance package for R [@wang2019].
-The topographic distance was measured between site polygon and the horizontally closest point on the shoreline.
+The topographic distance was measured between the site polygon and the horizontally closest point on the shoreline.
 This means that the distance is not necessarily measured as the closest topographic distance to the shoreline, but rather as the shortest topographic path to the horizontally closest point on the shoreline.
-Not finding the topographically closest point significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact on the results, given the distances considered.
+Not finding the topographically closest point significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact on the results given the distances considered.
 The shortest topographic path was found using the Moore neighbourhood of eight cells [e.g.
@@
 While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
-Negative values were therefore retained, with the exception of for the sites Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
+Negative values were therefore retained with the exception of for the sites Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
 
@@
 # Group dates that overlap with 99.7 % probability using the group_dates()
-# function. (All functions in this chunk where package isn't specified are 
-# defined in 04script.R).
+# function. (bespoke functions are defined in 04script.R).
 date_groups <- group_dates(rcarb_sa, sitename)
@@
 
-curve_plot <- ggplot() + 
+curve_plot <- ggplot2::ggplot() + 
               geom_line(data = displacement_curves, 
@@
 
+```{r agediffdat}
+# Preparing some data for the text
+
+# Subtract one for Gunnarsrød 5 (see text)
+synchdata$meso[synchdata$cross == "false"] <- 
+  synchdata$meso[synchdata$cross == "false"] - 1
+
+cormeso <- synchdata %>% dplyr::filter(cross == "true") %>% 
+  dplyr::select(meso)
+corpreln <-  synchdata %>% dplyr::filter(cross == "true") %>% 
+  dplyr::select(pre_ln)
+corln <-  synchdata %>% filter(cross == "true") %>% dplyr::select(ln)
+phasesums <- summarise_if(synchdata, is.numeric, sum, na.rm = TRUE)
+
+```
 An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation (Figure \@ref(fig:results2)).
@@
 The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement.
-The overall success of this procedure is
-If one only includes sites older than 2500 BCE with 95% probability, this improves to, while only including dates older than 4000 BCE with 95% probability increases this further to a success rate of.
+The deviation from the shoreline date for the earliest phase at Gunnarsrød 5 is to be expected based on the simulation results (see above).
+When excluding this, the shoreline date correspond to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100, 1)`%).
+If one only includes dates modelled to be older than 2500 BCE with 95% probability, this improves to `r cormeso + corpreln` out of `r phasesums["meso"] + phasesums["pre_ln"]` cases (`r round((cormeso + corpreln)  / (phasesums["meso"] + phasesums["pre_ln"])*100, 1)`%), while only including dates older than 4000 BCE with 95% probability increases this further to a success rate of `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100, 1)`%).
 Thus, if there in the dating of sites are reasons to
@@
 The result of this procedure indicates that the shoreline date correspond to the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%).
-The deviation from the shoreline date at Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langemyr and Kvastad A2.
+, but not at Langemyr and Kvastad A2.
 For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B).
@@
 The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black.
-These are plotted against the modelled radiocarbon dates, which are given colour from oldest to youngest phase for each site.", out.width = "525px", fig.align="center", regfloat=TRUE}
+These are plotted against the modelled radiocarbon dates, which are given colour from oldest to youngest phase for each site, defined by non-overlapping dates at 99.7\\% probability.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 The comparison with previously reported dates is an illustrative, but unfair exercise for a few reasons.
-First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit, but undefined uncertainty range.
+First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an undefined, but implied or explicit uncertainty range.
 Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change.
@@
 
-It is worth highlighting some limitations and sources of likely variation and uncertainty that have not been considered here.
-First of all the sample size is quite strained, and the future addition of only a few sites might skew the picture significantly.
+Some limitations and sources of likely variation and uncertainty that have not been considered here are worth .
+First of all the sample size is quite strained, and it is worth bearing in mind that the future addition more sites might alter the picture considerably.
 Secondly, the DTM has only been corrected for major modern disturbances.

### 2022-03-18 analysis/script/04script.R

@@ -1293,13 +1293,13 @@ age_diff <- function(shorelinedata, radiocarbondata, nsamp = 100000){
                                           "year_min"][[1]] < -4000, "meso",
                     ifelse(prob95[prob95$site_name ==
                                     unique(shorelinedata$site_name)[i],
-                                  "year_min"][[1]] < -2500, "pre-ln", "ln")),
+                                  "year_min"][[1]] < -2500, "pre_ln", "ln")),
                     cross = ifelse(between(0, min(hdrcde::hdr(age_diff,
                                                               prob = 95)$hdr),
                                            max(hdrcde::hdr(age_diff,
                                        prob = 95)$hdr)), "true", "false")))
   }
 
-  agedif$phase <- factor(agedif$phase, levels=c("meso", "pre-ln", "ln"))
+  agedif$phase <- factor(agedif$phase, levels=c("meso", "pre_ln", "ln"))
   return(agedif)
 }

### 2022-03-18 analysis/script/07script.R

@@ -181,8 +181,11 @@ shrplt <- ggplot(data = hdrdat,
   scale_x_continuous(breaks = c(seq(-10000, -4000, 2000), -2500, 0, 2000) ) +
   theme_bw()
 
-#values = c("black", "#21908CFF", "#A6611A"),
+# Save plot
+ggsave(file = here("analysis/figures/shoredate.png"), shrplt,
+       width = 200, height = 250, units = "mm")
 
+# Next sections finds synchroneity between radiocarbon and shoreline dates
 shorelinedates <- bind_rows(shorelinedates) %>%
   filter(site_name %in% unique(corsites$site_name))
 
@@ -192,25 +195,33 @@ agedif3 <- age_diff(shorelinedates, corgroupsn3)
 agedif4 <- age_diff(shorelinedates, corgroupsn4)
 
 phasedt <- agedif %>% group_by(site_name) %>%
-  group_by(phase) %>%  summarise(n = n() / 100000) %>%
+  group_by(phase, cross) %>%  summarise(n = n() / 100000) %>%
   pivot_wider(names_from = phase, values_from = n)
 
 phasedt2 <- agedif2 %>% group_by(site_name) %>%
-  group_by(phase) %>%  summarise(n = n() / 100000) %>%
+  group_by(phase, cross) %>%  summarise(n = n() / 100000) %>%
   pivot_wider(names_from = phase, values_from = n)
 
 phasedt3 <- agedif3 %>% group_by(site_name) %>%
-  group_by(phase) %>%  summarise(n = n() / 100000) %>%
-  pivot_wider(names_from = phase, values_from = n) %>%
-  mutate(ln = NA)
+  group_by(phase, cross) %>%  summarise(n = n() / 100000) %>%
+  pivot_wider(names_from = phase, values_from = n)
 
 phasedt4 <- agedif4 %>% group_by(site_name) %>%
-  group_by(phase) %>%  summarise(n = n() / 100000) %>%
+  group_by(phase, cross) %>%  summarise(n = n() / 100000) %>%
   pivot_wider(names_from = phase, values_from = n) %>%
-  mutate(meso = NA, ln = NA) # Hacky solution given the single observation
+  mutate(meso = NA, ln = NA)
 
-phasedata <- rbind(phasedt, phasedt2, phasedt3, phasedt4) %>%
-  summarise_all(sum, na.rm = TRUE)
[…]

### 2022-03-21 analysis/paper/paper.Rmd
Minor edits

@@
 load(here("analysis/data/derived_data/07agediff.RData"))
-source(here("analysis/script/04script.R"), local = knitr::knit_global())
+source(here("analysis/script/04functions.R"), local = knitr::knit_global())
 
@@
 
-The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was established by the end of the 19th century [@degeer1896], and first applied as a dating method at the turn of the century [@brogger1905; @hollender1901].
+The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was established by the end of the 19th century [@degeer1896], and was first applied as a dating method at the turn of the century [@brogger1905; @hollender1901].
 Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g.
-@berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @nummedal1923; @mikkelsen1975; @solheim2018].
+@berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @nummedal1923; @mikkelsen1975; @shetelig1922; @solheim2018].
 The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.
@@
 @hagen1963; @reitan2009].
-Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008; @glorstad2020].
+Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, under water today, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008; @glorstad2020].
 Comparatively, given a continuous and still ongoing shoreline regression from as high as c.
 220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979].
-This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.
+This makes the region especially useful for evaluating the assumption of a shore-bound settlement pattern over a long and continuous time-span.
 
@@
 
-The method of shoreline dating has been met with scepticism as related to the degree sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [@breivik2018a, 174].
-The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, typically in combination with other dating methods [see for example chapters in @; ; and below].
-Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time, typically by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [e.g.
-@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b].
-In his review of the method, @nordqvist1999, in part drawing on @johansen1963, argues that there can be little doubt concerning its general applicability -- what is less clear is the level of reliability and chronological resolution that the method can offer.
+The method of shoreline dating has been met with scepticism as related to the degree sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [@breivik2018a, 174, see @nordqvist1999].
+The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @; ; and below].
+Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time, by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b].
+In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963].
 
-The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [discussed in @berg-hansen2009; @bergsvik2009].
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @gundersen2013; @mjærum2018; @schulke2020].
+The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@bjerck1990; also discussed in @berg-hansen2009; @bergsvik2009].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; although see @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227].
-@nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g.
+@nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g.
 @amundsen2006; @solheim2012, 74; @ostmo1988].
@@
 
-Three of the sites have been associated with farming activities, either directly or in the form building structures.
+Three of the sites have been associated with agriculture, either directly or in the form building structures.
 The first is Nordby 1 at which the ^14^C-dates are associated with a Late Neolithic long-house [@gjerpe2008b].
-The Middle Neolithic phase at Kvastad A2 [@stokke2018b] and Late Neolithic phase at Nauen A [@persson2008] are both directly related to farming.
+The Middle Neolithic phase at Kvastad A2 [@stokke2018b] and Late Neolithic phase at Nauen A [@persson2008] are both directly related to farming acitivities.
 Both of these sites also have radiocarbon dates and lithic inventory associated with Mesolithic forager activities.
-These agricultural phases are treated separately below, following from their expected deviance from the settlement patterns that are to characterise forager sites.
+Following from the expected deviance from the settlement patterns that are to characterise forager sites, these agricultural phases are treated separately below.
 Finally, @nielsen2021b has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratsberg [@wenn2012] and Larønningen [@roberg2012] could be related to early agricultural activity in the Oslo fjord region.
@@
 
-The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2019].
+The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2018, https://hoydedata.no].
 It was here opted for the 10m resolution DTM rather than the higher-resolution 1m version.
 In addition to resulting in considerably less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by.
-The 10m DTM is a down-sampled version of the 1m resolution DTM, which is based on aerial laser scanning using a minimum of two and up to five points per m^2^ and has a vertical error of  [@mapping2019].
+The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m [@mapping2018].
 All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a digital research compendium following @marwick2018.
@@
 
-overview_map / disp_curves
+overview_map / disp_curves + plot_annotation(tag_levels = "A")
 ```
@@
 @crema2010, 1125].
-Hovland 5 was chosen for this evaluation as it has a fairly uncertain date, and is located in area of quite complex topography.
+Hovland 5 was chosen for this evaluation as it has a fairly un-precise date and is located in area of quite complex topography.
 
@@
 Overall, as is indicated by the measures for central tendency and the almost solid line along the 0m mark on the y-axes, the simulations show that the sites tend to have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)).
-Some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included, but if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
+Some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included (Figure \@ref(fig:results)A), but if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, Figure \@ref(fig:results)B gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
 The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the sea-level.
@@
 
-Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE.
+Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results for from Figure \@ref(fig:results)B is given again in Figure \@ref(fig:results2)A, excluding all simulation results younger than 2500 BCE.
 Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations.
@@
 That is, for 95% of the cases, the shoreline was simulated to be situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
-While these values remain the same when only the Mesolithic dates are included in the second row of Figure \@ref(fig:results2), the mean and standard deviation is slightly constrained.
-Furthermore, while the median for horizontal and topographic distance is only `r round(median(distances2$topodist))`m across all plots in Figure \@ref(fig:results2), the variation in statistics for dispersion is also highly more variable, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
+While these values remain the same when only the Mesolithic dates are included (Figure \@ref(fig:results2)B), the mean and standard deviation is slightly constrained.
+Furthermore, while the median for horizontal and topographic distance is only `r round(median(distances2$topodist))`m across all plots in Figure \@ref(fig:results2), the variation in the statistics for dispersion is greater, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
 
@@
 
+site_elevation, shoreline_displacement (upper and lower bound?), relationship between site and shoreline. For each offset, subtract
+
 Where x is [...] In Figure \@ref(fig:shoredate) this formula is used to shoreline date the same sites from where this relationship was derived.
-Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation over the site limits.
-The synchroneity between radiocarbon and shoreline dates was then evluated using the method presented by [@parnell2008, Figure \@ref(fig:shoredatediff)].
-Here, 100,000 age samples drawn from the probability density function of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding ^14^C-date.
+The Late Neolithic sites were also included for illustrative purposes.
+Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits.
+The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by @parnell2008 [Figure \@ref(fig:shoredatediff)].
+Here, 100,000 age samples drawn from the probability density function of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding modelled ^14^C-dates.
 The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement.
-The deviation from the shoreline date for the earliest phase at Gunnarsrød 5 is to be expected based on the simulation results (see above).
-When excluding this, the shoreline date correspond to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100, 1)`%).
-If one only includes dates modelled to be older than 2500 BCE with 95% probability, this improves to `r cormeso + corpreln` out of `r phasesums["meso"] + phasesums["pre_ln"]` cases (`r round((cormeso + corpreln)  / (phasesums["meso"] + phasesums["pre_ln"])*100, 1)`%), while only including dates older than 4000 BCE with 95% probability increases this further to a success rate of `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100, 1)`%).
-Thus, if there in the dating of sites are reasons to
-
-The result of this procedure indicates that the shoreline date correspond to the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%).
-, but not at Langemyr and Kvastad A2.
-For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B).
-If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of Gunnarsrød 5 is instead due to extraneous factors related the DTM and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.
-However, another possible implication of the failed dating of these two sites could be that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites from the earliest part of the Mesolithic (cf.
+When excluding the earliest phase at Gunnarsrød 5, the deviation of which is to be expected based on the simulation results (see above), the shoreline date correspond to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%).
+If one only includes dates modelled to be older than 2500 BCE with 95% probability, i.e.
+older than the Late Neolithic, this improves to `r cormeso + corpreln` out of `r phasesums["meso"] + phasesums["pre_ln"]` cases (`r round((cormeso + corpreln)  / (phasesums["meso"] + phasesums["pre_ln"])*100)`%).
+When only including dates older than 4000 BCE with 95% probability, i.e.
+only Mesolithic, this is further increased to a success rate of `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100)`%).
+The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the likely implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (cf.
 Figure \@ref(fig:results)).
@@
 
-```{r results2, echo = FALSE, fig.cap = "Histograms showing the simualted distance from the shoreline using dates corresponding to the site inventory.
+```{r results2, echo = FALSE, fig.cap = "Histograms showing the simulated distance from the shoreline using dates corresponding to the site inventory.
 Negative values have been set to zero.
-The first row only includes simulated results older than 2500 BCE and the second row only results older than 4000 BCE.", out.width = "525px", fig.align="center", regfloat=TRUE}
+A) Simulated results older than 2500 BCE, and B) simulated results older than 4000 BCE.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/results2.png"))
@@
 ```{r shoredatediff, echo = FALSE, fig.cap = "Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure 9.
-When the range of the 95\\% HDR for age difference crosses zero, indicated by a solid line, the shoreline- and radiocarbon dates are considered synchronous.
+When the range of the 95\\% HDR for age difference crosses zero, the shoreline and radiocarbon dates are considered synchronous.
+Line segments with vertical bars indicate that this does not cross zero and that the dates are not in agreement.
 The division and colour coding at the top of the plots reflects the divison of site phases given in Figure 9.", out.width = "525px", fig.align="center", regfloat=TRUE}
@@
 
-# Shoreline dating
+# Re-dating previously shoreline dated sites
 
 To further explore the implementation for shoreline dating outlined above, excavated and shoreline-dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the method (Figure \@ref(fig:redate)).
-The resulting dates are compared to those originally proposed in the excavation reports for the sites (the numerical results are available in the supplementary material).
+The resulting dates are compared to those originally proposed in the excavation reports for the sites (numerical results are available in the supplementary material).
 To avoid issues with recent disturbances on the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
@@
 Finally, they are sometimes only meant to indicate a lower bound for when the sites could have been in use.
-Nonetheless, the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g.
+Overall, the results indicate that shoreline dating has generally been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g.
 @roalkvam2022].
-That being said, the results do also indicate that shoreline dating has at times been applied with an exaggerated degree of precision, and while the implications of a more stable RSL-change for the duration and re-use of site locations, and consequently the precision of shoreline dating is well known, it appears to be somewhat under-appreciated.
+That being said, the results do also indicate that shoreline dating has at times been applied with an exaggerated degree of precision.
+While the implications of a more stable RSL-change for the duration of use and re-use of site locations, and consequently the precision of shoreline dating is well known, it appears to be somewhat under-appreciated.
 Some of the date ranges resulting from the method outlined here clearly extend well beyond major chronological divisions, even into the Iron Age, and could be severely and securely constrained with only cursory reference to typology.
@@
 
-Not least following from the fact that relatively few Preboreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway [@åstveit2018; @damlien2018; @kleppe2018], the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
-@bang-andersen2012; @breivik2014; @glorstad2016].
+Not least following from the fact that relatively few Preboreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway [@åstveit2018; @damlien2018b; @kleppe2018], the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
+@bang-andersen2012; @breivik2014; @glorstad2016; @fuglestvedt2012].
 The shoreline-dated Preboreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a].
@@
 This means that erosion, although likely not that prevalent, has not been taken into account.
-Thirdly, the DTM has a vertical error of, which could also benefit from being integrated in the analysis [cf.
+Thirdly, the DTM has a vertical error which could also benefit from being integrated in the analysis [cf.
 @lewis2021].
@@
 This is followed by a more decisive break at the transition to the Late Neolithic at c.
-2400 BCE -- in clear agreement with the literature.
+2400 BCE.
+This development is in clear agreement with the literature.
 Furthermore, based on the quantitative nature of these findings, a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed.
@@
 This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range could possibly also have implications for the site location relative to the shoreline, depending on the topography [@helskog1978].
-Dimensions such as these was given a cursory treatment here, with the estimation of the horizontal and topographic distance to the shoreline, but these dimensions could also be investigated further.
+With the estimation of the horizontal and topographic distance to the shoreline dimensions such as these was given a very cursory treatment here that could certainly be investigated further.
 If patterns related to such locational patterns can be discerned, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the site beyond its location is available.

### 2022-03-21 analysis/supplementary-materials/supplementary.Rmd
Minor edits

@@
                     "Context"),
-        booktabs = TRUE,  linesep = c(""), escape = FALSE)
+        booktabs = TRUE,  linesep = c(""), escape = FALSE) %>% 
+  kable_styling(latex_options="scale_down")
 ```

### 2022-03-21 analysis/script/04functions.R

@@ -400,7 +400,8 @@ interpolate_curve <- function(years, target, dispdat, isodat){
   # displacement curve (e.g. Alveberget 8), simply return that displacement
   # curve
   if(any(as.numeric(dists) == 0)){
-    values <- dispdat[dispdat$name == names(dists)[which(as.numeric(dists) == 0)],]
+    values <-
+      dispdat[dispdat$name == names(dists)[which(as.numeric(dists) == 0)],]
 
   } else { for(i in 1:length(years)){
     for(j in 1:ncol(dists)){
@@ -868,8 +869,9 @@ overview_plot <- function(background_map, sitelimit, sites, isobases,
 }
 
 # Define function to plot site relative to simulated sea-levels
-shore_plot <- function(locationraster, overlapgrid, sitelimit, dist, date_groups,
-                       s_tdist, s_xpos, s_ypos, s_bheight, s_tsize) {
+shore_plot <- function(locationraster, overlapgrid, sitelimit, dist,
+                       date_groups, s_tdist, s_xpos, s_ypos, s_bheight,
+                       s_tsize) {
 
   bboxgrid <- st_bbox(overlapgrid)
   anc <- as.numeric(c(bboxgrid$ymin, bboxgrid$xmax))
@@ -975,7 +977,7 @@ plot_dates <- function(datedata, sitename, multigroup = TRUE, groupn = NA,
        colour = as.factor(rcarb_cor),
        alpha = class))+
       scale_alpha_manual(values = c("aprior" = 0.05,
-                                    "bposterior" = 0.4,
+                                    "bposterior" = 0.3, #0.4
                                     "sum" = 1)) +
       scale_colour_manual(breaks = c("f", "t", NA),
                           values = c("red", "black", "yellow")) +
@@ -994,7 +996,7 @@ plot_dates <- function(datedata, sitename, multigroup = TRUE, groupn = NA,
       scale_colour_manual(breaks = c("f", "t", NA),
                          values = c("red", "black", "yellow")) +
       scale_alpha_manual(values = c("aprior" = 0.05,
-                                    "bposterior" = 0.4,
+                                    "bposterior" = 0.3, #0.4
                                     "sum" = 1)) +
       theme_bw() +
       labs(y = "", x = "") +
@@ -1215,13 +1217,14 @@ shoreline_date1 <- function(sitename, elev = dtm,
     if(!(negative_offset > 0)) {
       negative_offset <- 0.01
     }
[…]

### 2022-03-21 analysis/script/06simulation_results.R

@@ -77,12 +77,12 @@ colsh = c("forager" = "#00ba38", "agricultural" = "darkgrey")
 colst = c("forager" = "#fad510", "agricultural" = "darkgrey")
 colsv = c("forager" = "steelblue", "agricultural" = "darkgrey")
 
-# Horisontal distance, all dates to the Stone Age
+# Horizontal distance, all dates to the Stone Age
 h1 <- ggplot(distances, aes(x = year, y = hordist,
                             colour = type)) +
   geom_point(shape = 16, alpha = 0.01) +
   scale_colour_manual(values = colsh) +
-  labs(y = "Horisontal distance (m)", x = "BCE",
+  labs(y = "Horizontal distance (m)", x = "BCE",
        title = "All dates") +
   scale_x_continuous(breaks = seq(-8000, -1000, by = 2000)) +
   theme_bw() +
@@ -99,13 +99,13 @@ th <- sum(th1$heights)
 # Arrange plot and table
 hor1 <- grid.arrange(h1, th1, heights = unit.c(unit(1, "null"), th))
 
-# Horisontal distance, dates corresponding to inventory
+# Horizontal distance, dates corresponding to inventory
 h2 <- distances %>%  filter(rcarb_cor == "t") %>%
   ggplot(aes(x = year, y = hordist,
                         colour = type)) +
   geom_point(shape = 16, alpha = 0.01) +
   scale_colour_manual(values = colsh) +
-  labs(y = "Horisontal distance (m)", x = "BCE",
+  labs(y = "Horizontal distance (m)", x = "BCE",
        title = "Corresponding dates") +
   scale_x_continuous(breaks = seq(-8000, -1000, by = 2000)) +
   theme_bw() +
@@ -198,7 +198,7 @@ vert2 <- grid.arrange(v2, vt2, heights = unit.c(unit(1, "null"), th))
 grd1 <- grid.arrange(hor1, topo1, vert1, nrow = 1)
 grd2 <- grid.arrange(hor2, topo2, vert2, nrow = 1)
 
-grd <- plot_grid(grd1, grd2, ncol = 1, scale = 0.9)
+grd <- plot_grid(grd1, grd2, labels = c("A", "B"), ncol = 1, scale = 0.9)
 
 ggsave(file = here("analysis/figures/results.png"), grd,
        bg = "white")
@@ -270,10 +270,10 @@ histhor <- distances2 %>%
                             colour = "black",
[…]

### 2022-03-22 analysis/paper/paper.Rmd
Minor edits

@@
 csl: "../templates/saa.csl" # Insert path for the bib-style
+keywords: |
+  Stone Age; Settlement patterns; Scandinavia; Relative sea-level change; Shoreline dating; Monte Carlo simulation
 ---
@@
 @hagen1963; @reitan2009].
-Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, under water today, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008; @glorstad2020].
+Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, still submerged today [@bjerck2008; @glorstad2020].
+This can lead to a hiatus in the archaeological record for certain sub-phases in the impacted areas.
 Comparatively, given a continuous and still ongoing shoreline regression from as high as c.
@@
 
-The method of shoreline dating has been met with scepticism as related to the degree sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [@breivik2018a, 174, see @nordqvist1999].
-The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @; ; and below].
-Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time, by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b].
+The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [see review by @nordqvist1999].
+The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @damlien2021; @jaksland2014; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below].
+Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time.
+This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b].
 In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963].
@@
 @bakka1971; @prescott2012; @ostmo2008, 223].
-That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a].
+That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2012].
 At an earlier stage, at the transition to the Early Neolithic (c.
@@
 @amundsen2006; @solheim2012, 74; @ostmo1988].
-Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and finally a decisive shift at around 2400 BCE.
+Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the Early Neolithic around 3900 BCE, and finally a decisive shift at the transition to the Late Neolithic c.
+2400 BCE.
 
 Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
-This is for example common in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g.
+This is for example commonly reflected in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g.
 @berg-hansen2009; @eskeland2017].
-Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the typologically informed date-span is too wide to decisively verify this.
+Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date-span is too wide to decisively verify this.
 It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g.
@@
 Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
-@akerlund1995; @åstveit2018; @solheim2020; see also @bjerck2008a; @kleppe1985].
+@akerlund1995; @åstveit2018; @solheim2020; see also @bjerck2008a; @kleppe1985; @ramstad2009].
 This approach has a couple of limitations.
@@
 RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater.
-Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.
+Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and help move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.
 
@@
 While there is displacement data available for other areas of south-eastern Norway [e.g.
-@hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
+@hafsten1957; @sørensen1979; @sørensen1999; @creel2022], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
 This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
@@
 Both of these sites also have radiocarbon dates and lithic inventory associated with Mesolithic forager activities.
-Following from the expected deviance from the settlement patterns that are to characterise forager sites, these agricultural phases are treated separately below.
+Following from the expected deviance from the settlement patterns that are to characterise forager sites, these agricultural phases are highlighted in the analysis below.
 Finally, @nielsen2021b has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratsberg [@wenn2012] and Larønningen [@roberg2012] could be related to early agricultural activity in the Oslo fjord region.
@@
 In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
-In the case that the sea-polygon completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon.
+In the case that the sea-polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon.
 While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
@@
 @crema2010, 1125].
-Hovland 5 was chosen for this evaluation as it has a fairly un-precise date and is located in area of quite complex topography.
+Hovland 5 was chosen for this evaluation as it has a fairly imprecise date and is located in area of quite complex topography.
 
@@
 Line segments with vertical bars indicate that this does not cross zero and that the dates are not in agreement.
-The division and colour coding at the top of the plots reflects the divison of site phases given in Figure 9.", out.width = "525px", fig.align="center", regfloat=TRUE}
+The division and colour coding at the top of the plots reflect the divison of site phases given in Figure 9.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate2.png"))
@@
 To avoid issues with recent disturbances on the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
-The results highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% intervals for the dates.
+The results highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% HDRs for the dates.
 In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.
@@
 
-Some limitations and sources of likely variation and uncertainty that have not been considered here are worth .
-First of all the sample size is quite strained, and it is worth bearing in mind that the future addition more sites might alter the picture considerably.
+Some limitations and sources of likely variation and uncertainty that have not been considered here should be mentioned.
+First of all the sample size is quite strained, and the future addition more sites might alter the picture considerably.
 Secondly, the DTM has only been corrected for major modern disturbances.
@@
 Fourthly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves further away from the isobases of the displacement curves.
+This is also related to the fact that the RSL data can be treated in different ways than what was done for the compilation of the displacement curves employed here [cf.
+@creel2022].
 Fifthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration.
-Finally, the division of phases at each sites was here simply done by treating  While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration.
+Finally, the division of phases at each site was here simply done by treating  While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration.
 
@@
 :::
-
-<!-- \newpage -->
-
-<!-- ### Colophon -->
-
-<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies: -->
-
-<!-- ```{r colophon, cache = FALSE} -->
-<!-- # which R packages and versions? -->
-<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->
-<!-- ``` -->
-
-<!-- The current Git commit details are: -->
-
-<!-- ```{r} -->
-<!-- # what commit is this file at?  -->
-<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->
-<!-- ``` -->

### 2022-03-25 analysis/paper/paper.Rmd
Updated figures, minor text edits

@@
 2016, although see Romundset et al.
-2019 for the study area).", out.width = "400px", fig.align="center", regfloat=TRUE}
+2019 in relation to the study area).", out.width = "525px", fig.align = "center", regfloat = TRUE}
 
@@
 
-The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@bjerck1990; also discussed in @berg-hansen2009; @bergsvik2009].
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; although see @gundersen2013; @mjærum2018; @schulke2020].
+The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; also discussed in @berg-hansen2009; @bergsvik2009].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; although see @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 @amundsen2006; @solheim2012, 74; @ostmo1988].
-Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the Early Neolithic around 3900 BCE, and finally a decisive shift at the transition to the Late Neolithic c.
+Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c.
 2400 BCE.
@@
 Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
-This is for example commonly reflected in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g.
+This is for example reflected in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g.
 @berg-hansen2009; @eskeland2017].
@@
 This is based on typological and technological characteristics of the assemblages.
-Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results can be difficult to evaluate [e.g.
+Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these dates are entirely erroneous can be difficult to evaluate [e.g.
 @persson2008; @schulke2020].
@@
 # Data
-To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
+To get at the relationship between sites and the contemporaneous shoreline this analysis was dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
 While there is displacement data available for other areas of south-eastern Norway [e.g.
@@
 
-The employed shoreline displacement curves are all based on the so-called isolation basin method [e.g.
-@kjemperud1986; @romundset2011].
-This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments.
-Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point (SLIP).
+The employed shoreline displacement data is based on the so-called isolation basin method [e.g.
+@kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments.
+Each basin thus represent a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to a continuous time series for RSL-change, projected to a common isobase [see @creel2022, 5].
 Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible.
-Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from a common isobase perpendicular to this gradient (Figure \@ref(fig:map-iso)).
-The resulting SLIPs indicate the isolation of the basin from the highest astronomical tide.
-This is adjusted to mean sea-level in the compilation of the displacement curve based on the present day tidal range, which is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
+Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from the common isobase perpendicular to this gradient (Figure \@ref(fig:map-iso)).
+The SLIPs indicate the isolation of the basin from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves based on the present day tidal range.
+This is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
 Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
 The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
-@romundset2019; @romundset2011].
+@romundset2019; @romundset2011, for an alternative approach see @creel2022].
 For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.
@@
 Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same settlement phase.
-In the case where there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed.
+In the case where there are multiple dates believed to belong to a single phase, these were modelled using the Boundary function in OxCal and then summed.
 Multiple phases at a single site were treated as independent of each other.
@@
 @conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021].
-In addition to code and original spatial data being available in the digital research compendium for this paper, the analysis of each individual site is also available in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.
+In addition to code and original spatial data being available in the digital research compendium for this paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.
 
@@
 A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \@ref(fig:example-2)).
-This year then has a corresponding likely elevation range for the contemporaneous shoreline, from which an elevation value was drawn uniformly, using intervals of 5cm.
+This year then has a corresponding likely elevation range for the contemporaneous shoreline from which an elevation value was drawn uniformly, using intervals of 5cm.
 The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing.
@@
 However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
-If one accepts this, having derived an exponential decay function for describing the vertical distance between sites and shoreline can be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:
+If one accepts this, the exponential decay function used for describing the vertical distance between sites and shoreline can be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:
 
-site_elevation, shoreline_displacement (upper and lower bound?), relationship between site and shoreline. For each offset, subtract
+\begin{equation} 
+  p\left(t\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
+  (\#eq:shoredate)
+\end{equation}
 
-Where x is [...] In Figure \@ref(fig:shoredate) this formula is used to shoreline date the same sites from where this relationship was derived.
-The Late Neolithic sites were also included for illustrative purposes.
+Where x is [...] site_elevation, shoreline_displacement (upper and lower bound?), relationship between site and shoreline. For each offset, subtract
+
+\@ref(eq:shoredate) is used to shoreline date the same sites from where this relationship was derived, using the decay ratio of `r round(expfit$estimate, 3)` (Figure \@ref(fig:shoredate)) which was found when including all pre-Late Neolithic results (Figure \@ref(fig:results2)A).
+For illustrative purposes the Late Neolithic sites were also included.
 Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits.
@@
 When excluding the earliest phase at Gunnarsrød 5, the deviation of which is to be expected based on the simulation results (see above), the shoreline date correspond to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%).
-If one only includes dates modelled to be older than 2500 BCE with 95% probability, i.e.
-older than the Late Neolithic, this improves to `r cormeso + corpreln` out of `r phasesums["meso"] + phasesums["pre_ln"]` cases (`r round((cormeso + corpreln)  / (phasesums["meso"] + phasesums["pre_ln"])*100)`%).
+Only including dates modelled to be older than 2500 BCE with 95% probability, i.e.
+older than the Late Neolithic, improves this to `r cormeso + corpreln` out of `r phasesums["meso"] + phasesums["pre_ln"]` cases (`r round((cormeso + corpreln)  / (phasesums["meso"] + phasesums["pre_ln"])*100)`%).
 When only including dates older than 4000 BCE with 95% probability, i.e.
-only Mesolithic, this is further increased to a success rate of `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100)`%).
-The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the likely implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (cf.
-Figure \@ref(fig:results)).
+only Mesolithic, the success rate is further increased to `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100)`%).
+The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the possible implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (see also Figure \@ref(fig:results)).
 
@@
 Results associated with agricultural acitivites are plotted in grey.
-The first row shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites.
-The second row shows the result of excluding these.
+The first row A) shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites.
+The second row B) shows the result of excluding these.
 The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center", regfloat=TRUE}
@@
 
-```{r shoredatediff, echo = FALSE, fig.cap = "Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure 9.
+```{r shoredatediff, echo = FALSE, fig.cap = "Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure 8.
 When the range of the 95\\% HDR for age difference crosses zero, the shoreline and radiocarbon dates are considered synchronous.
 Line segments with vertical bars indicate that this does not cross zero and that the dates are not in agreement.
-The division and colour coding at the top of the plots reflect the divison of site phases given in Figure 9.", out.width = "525px", fig.align="center", regfloat=TRUE}
+The division and colour coding at the top of the plots reflect the divison of site phases given in Figure 8.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate2.png"))
@@
 
-To further explore the implementation for shoreline dating outlined above, excavated and shoreline-dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the method (Figure \@ref(fig:redate)).
+To further explore the implementation for shoreline dating outlined above, excavated and shoreline dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the approach (Figure \@ref(fig:redate)).
 The resulting dates are compared to those originally proposed in the excavation reports for the sites (numerical results are available in the supplementary material).
 To avoid issues with recent disturbances on the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
-The results highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% HDRs for the dates.
-In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.
-This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf.
-Figure \@ref(fig:map-iso)B).
 
@@
 Finally, they are sometimes only meant to indicate a lower bound for when the sites could have been in use.
-Overall, the results indicate that shoreline dating has generally been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g.
+Overall, the results suggest that shoreline dating has generally been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g.
 @roalkvam2022].
 That being said, the results do also indicate that shoreline dating has at times been applied with an exaggerated degree of precision.
-While the implications of a more stable RSL-change for the duration of use and re-use of site locations, and consequently the precision of shoreline dating is well known, it appears to be somewhat under-appreciated.
-Some of the date ranges resulting from the method outlined here clearly extend well beyond major chronological divisions, even into the Iron Age, and could be severely and securely constrained with only cursory reference to typology.
-However, while this is obvious in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed.
-This final point also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to further improve the precision of the dates that can be achieved with shoreline dating.
+While the implications of a more stable RSL-change for the duration of use and re-use of site locations, and consequently the precision of shoreline dating, is well known, it appears somewhat under-appreciated.
+The results also highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% HDRs for the dates.
+In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.
+This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf.
+Figure \@ref(fig:map-iso)B).
+Further, as some of the date ranges extend well beyond major chronological divisions, even into the Iron Age, they could be severely and securely constrained with only cursory reference to typology.
+While this would be trivial in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed.
+This also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to further improve the precision of the dates that can be achieved with shoreline dating.
 
@@
 Not least following from the fact that relatively few Preboreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway [@åstveit2018; @damlien2018b; @kleppe2018], the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
-@bang-andersen2012; @breivik2014; @glorstad2016; @fuglestvedt2012].
-The shoreline-dated Preboreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a].
+@bang-andersen2012; @berg-hansen2018; @breivik2014; @glorstad2016; @fuglestvedt2012].
+The shoreline dated Preboreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a].
 These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision use of the sites after the sea retreated any significant distance from their location.
@@
 A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \@ref(fig:brunlanes)A.
-The sites have been dated using what @jaksland2014 gives as the lowest elevation of finds at each site, and by employing a exponential decay ratio of , to allow for more deviance in the distance between site and shoreline.
+The sites have been dated using what @jaksland2014 gives as the lowest elevation of finds at each site, and by employing a exponential decay ratio of `r `round(expfit3$estimate, 3)`, to allow for more deviance in the distance between site and shoreline.
 This corresponds to the decay ratio for sites older than 7000 BCE in Figure \@ref(fig:results2).
@@
 Dates provided by Jaksland (2014) are plotted in red.
-The box indicates a 50 year uncertainty range which in combination with the red line extends 200 years, following his approach.
+The box indicates a 50 year uncertainty range which in combination with the red line extends 200 years.
 B) Map showing the centroids of the Pauler sites and Sky 1.
@@
 
-Some limitations and sources of likely variation and uncertainty that have not been considered here should be mentioned.
-First of all the sample size is quite strained, and the future addition more sites might alter the picture considerably.
+The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline.
+This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c.
+4000 BCE, after which a couple of sites become situated some distance from the sea, followed by a more decisive break at the transition to the Late Neolithic at c.
+2500 BCE.
+This development is in agreement with the literature.
+Furthermore, based on the quantitative nature of these findings, an initial formulation of a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed.
+In addition to taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves implementing equation \@ref(eq:shoredate) to account for the distance between the sites and the shoreline.
+When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio of `r round(expfit$estimate, 3)` (Figure \@ref(fig:redate)A) to characterise this relationship.
+Furthermore, while this remains to be formalised and explored further, it was also showed how the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data.
+As the precision of the method is both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date than younger sites located towards the south-west, the impact of such additional information will also vary.
+
+Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further evaluate and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted.
+Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located.
+Furthermore, other dimensions related to the topographic location of the sites could be similarly explored.
+This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [e.g.
+@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@helskog1978; @blankholm2020].
+This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range could possibly also have implications for the site location relative to the shoreline, depending on the topography [@helskog1978].
+The potential of dimensions such as these was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline.
+If patterns related to such locational patterns can be discerned and unpicked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the site beyond its location is available.
+
+Some limitations and sources of likely variation and uncertainty that have not been considered should also be mentioned.
+First of all the sample size is quite strained and the future addition more sites might alter the picture considerably.
 Secondly, the DTM has only been corrected for major modern disturbances.
@@
 Fourthly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves further away from the isobases of the displacement curves.
-This is also related to the fact that the RSL data can be treated in different ways than what was done for the compilation of the displacement curves employed here [cf.
+This is also related to the fact that the RSL data can be handled in different ways than with the isobase method that has been used for the compilation of the employed displacement curves [cf.
 @creel2022].
 Fifthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration.
-Finally, the division of phases at each site was here simply done by treating  While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration.
-
-The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline.
-This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c.
-3900 BCE, after which a couple of sites become situated some distance from the shoreline.
-This is followed by a more decisive break at the transition to the Late Neolithic at c.
-2400 BCE.
-This development is in clear agreement with the literature.
-Furthermore, based on the quantitative nature of these findings, a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed.
-This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula X to account for the distance between the sites and the shoreline.
-When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio identified in Figure \@ref(fig:redate) to characterise this relationship.
-However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data such as radiocarbon dates and typological indicators in the artefact inventories.
-The precision of the method is, as shown above, both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date than younger sites located towards the south-west.
-The impact of such additional information will therefore also vary.
-
-Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further evaluate and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted and refined.
-Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located.
-Furthermore, other dimensions related to the topographic location of the sites could be similarly explored.
-This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [e.g.
-@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@helskog1978; @blankholm2020].
-This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range could possibly also have implications for the site location relative to the shoreline, depending on the topography [@helskog1978].
-With the estimation of the horizontal and topographic distance to the shoreline dimensions such as these was given a very cursory treatment here that could certainly be investigated further.
-If patterns related to such locational patterns can be discerned, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the site beyond its location is available.
+Finally, the division of phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7% as unrelated.
+This could also be done differently.
+While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration.
 
@@
 Here this was simply used to inform the question of the distance between sites and the shore-line.
-However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
+However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent in this reconstruction.
 

### 2022-03-28 analysis/paper/paper.Rmd
Updates to text and figure

@@
 Istostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle.
-This is often the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed.
+This can be the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed.
 These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight [@milne2015].
@@
 @hagen1963; @reitan2009].
-Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, still submerged today [@bjerck2008; @glorstad2020].
+Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, leave them still submerged today [@bjerck2008; @glorstad2020].
 This can lead to a hiatus in the archaeological record for certain sub-phases in the impacted areas.
@@
 The study area defined later in the text is marked with a red outline (deglaciation data from Hughes et al.
-2016, although see Romundset et al.
+2016, but see also Romundset et al.
 2019 in relation to the study area).", out.width = "525px", fig.align = "center", regfloat = TRUE}
 
+# This map was pre-compiled for convenience. The deglaciation data from DATED-1 
+# Hughes et al. 2015 is available as supplemental material to their paper:
+# https://doi.org/10.1111/bor.12142
 knitr::include_graphics(here("analysis/figures/deglaciation.png"))
@@
 The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [see review by @nordqvist1999].
-The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @damlien2021; @jaksland2014; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below].
+The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @jaksland2014; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below].
 Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time.
@@
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; also discussed in @berg-hansen2009; @bergsvik2009].
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; although see @gundersen2013; @mjærum2018; @schulke2020].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 # Data
-To get at the relationship between sites and the contemporaneous shoreline this analysis was dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
+To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
 While there is displacement data available for other areas of south-eastern Norway [e.g.
-@hafsten1957; @sørensen1979; @sørensen1999; @creel2022], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
+@creel2022; @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
 This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
@@
 The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
-@romundset2019; @romundset2011, for an alternative approach see @creel2022].
+@romundset2011; @romundset2019; for an alternative approach see @creel2022].
 For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.
@@
 @conolly2020; @conolly2006, 94--97].
-This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances.
+This was done for each year along the entirety of the curves, weighting the interpolation by the standard squared inverse of the distances.
 The result of this process is shown for an example site in Figure \@ref(fig:example-1).
@@
 ```{r example-1, echo = FALSE, fig.cap = "Example site Hegna vest 1 (Fossum 2017).
-A) Location of the site in the present day landscape.
+A) Location of the site on the edited 10m resolution DTM.
 The red outline is the site limit.
@@
 Some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included (Figure \@ref(fig:results)A), but if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, Figure \@ref(fig:results)B gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
-The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the sea-level.
+The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the simulated sea-level.
 Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age, corresponding with the literature.
@@
 However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
-If one accepts this, the exponential decay function used for describing the vertical distance between sites and shoreline can be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:
+If one accepts this, the probability density function for exponential decay can be used to characterise the vertical distance between sites and the shoreline:
 
+<!-- d = 1(1-\lambda)x -->
 \begin{equation} 
-  p\left(t\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
+  f(x) = \lambda {e}^{- \lambda x}
+  (\#eq:decay)
+\end{equation}
+
+Where $\lambda$ is the decay ratio and $x$ is the distance to the shoreline.
+This can then be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:
+
+<!-- p\left(t\right) = \binom{x}{x} x^x\left(x-x\right)^{x-x} -->
+\begin{equation} 
+  placeholder
   (\#eq:shoredate)
@@
 
-Where x is [...] site_elevation, shoreline_displacement (upper and lower bound?), relationship between site and shoreline. For each offset, subtract
+Where x is [...] the site elevation, $i$ is the trajectory for shoreline displacement, here given as a uniform probability range between the upper and lower bound of the interpolated displacement curve.
 
-\@ref(eq:shoredate) is used to shoreline date the same sites from where this relationship was derived, using the decay ratio of `r round(expfit$estimate, 3)` (Figure \@ref(fig:shoredate)) which was found when including all pre-Late Neolithic results (Figure \@ref(fig:results2)A).
-For illustrative purposes the Late Neolithic sites were also included.
+Equation \@ref(eq:shoredate) is used to shoreline date the same sites from where this relationship was derived (Figure \@ref(fig:shoredate)), employing $\lambda$ = `r round(expfit$estimate, 3)` for the decay ratio in \@ref(eq:decay), which was found when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
+The Late Neolithic sites were also included for illustrative purposes.
 Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits.
@@
 
-```{r results2, echo = FALSE, fig.cap = "Histograms showing the simulated distance from the shoreline using dates corresponding to the site inventory.
+```{r results2, echo = FALSE, fig.cap = "Histograms showing the simulated distance from the shoreline using dates corresponding to the site inventories.
 Negative values have been set to zero.
@@
 
-To further explore the implementation for shoreline dating outlined above, excavated and shoreline dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the approach (Figure \@ref(fig:redate)).
-The resulting dates are compared to those originally proposed in the excavation reports for the sites (numerical results are available in the supplementary material).
+To further explore the implementation for shoreline dating given above, excavated and shoreline dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the outlined approach (Figure \@ref(fig:redate)).
+The resulting dates are compared to those originally proposed in the excavation reports for the sites (the numerical results are available in the supplementary material).
 To avoid issues with recent disturbances on the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
@@
 A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \@ref(fig:brunlanes)A.
-The sites have been dated using what @jaksland2014 gives as the lowest elevation of finds at each site, and by employing a exponential decay ratio of `r `round(expfit3$estimate, 3)`, to allow for more deviance in the distance between site and shoreline.
+The sites have been dated using what @jaksland2014 gives as the lowest elevation of finds at each site, and by employing a exponential decay ratio of `r round(expfit3$estimate, 3)`, to allow for more deviance in the distance between site and shoreline.
 This corresponds to the decay ratio for sites older than 7000 BCE in Figure \@ref(fig:results2).
@@
 This information is not integrated in the approach outlined here, but could justify further reducing the uncertainty ranges.
-Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does nonetheless result in very precise dates.
-Above it was shown how additional temporal data could be combined with the method to improve its accuracy and precision.
+Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does result in very precise dates.
+Above it was suggested that additional temporal data could be combined with the method to improve its accuracy and precision.
 This example, on the other hand, highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
 One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline.
-Based on this, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use.
+If this is developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use.
 Such approaches would make less of an impact in this setting, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf.
@@
 Furthermore, based on the quantitative nature of these findings, an initial formulation of a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed.
-In addition to taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves implementing equation \@ref(eq:shoredate) to account for the distance between the sites and the shoreline.
+Apart from taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves implementing equation \@ref(eq:shoredate) to account for the distance between the sites and the shoreline.
 When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio of `r round(expfit$estimate, 3)` (Figure \@ref(fig:redate)A) to characterise this relationship.
@@
 Furthermore, other dimensions related to the topographic location of the sites could be similarly explored.
-This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [e.g.
-@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@helskog1978; @blankholm2020].
+This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@helskog1978; @blankholm2020].
 This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range could possibly also have implications for the site location relative to the shoreline, depending on the topography [@helskog1978].
@@
 Fifthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration.
-Finally, the division of phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7% as unrelated.
-This could also be done differently.
+Finally, the radiocarbon dates and division of phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7% as representing unrelated events.
+This could also be handled differently [@bronkramsey2015].
 While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration.

### 2022-03-28 analysis/supplementary-materials/supplementary.Rmd
Minor edits
Updates to text and figure

@@
 These sites are therefore not included in the analysis.
-Furthermore, @nielsen2021 has recently interpreted features from the otherwise younger sites Bratsberg and Larønningen as possibly related to farming activity in the Early Neolithic.
+Furthermore, @nielsen2021b has recently interpreted features from the otherwise younger sites Bratsberg and Larønningen as possibly related to farming activity in the Early Neolithic.
 However, given the limited number of features, the lack of any artefacts directly related to these, and the therefore somewhat speculative nature of this suggestion, the sites are kept out of the analysis here.
@@
 The site limit for Auve was drawn manually based on the description in the publication of the site and the site geometry available in the database Askeladden.
-The area to the has reportedly been affected by land-slides, but the extent of this is difficult to ascertain based on maps of the area.
-As the simulation results appear reasonable [see also map in @ostmo2008, fig.142], nothing was done to try to correct for this.
+The site is located in and around an intersection of the road, parts of the area has been levelled and farmed, and an area to north-west of the site has reportedly been affected by land-slides [@ostmo2008, 16--29] , but the extent of this is difficult to ascertain based on maps of the area.
+As the simulation results appear reasonable [see also map in @ostmo2008, fig.142], nothing was done to try to correct for these factors.
 
@@
 
-The radiocarbon date to the Stone Age is not related to the site inventory, which is distinctly Early Mesolithic in character [@nyland2012b].
+The radiocarbon date to the Stone Age is not related to the site inventory at Bakke, which is distinctly Early Mesolithic in character [@nyland2012b].
 

### 2022-03-28 analysis/script/07backwards_dating.R

@@ -9,7 +9,7 @@ library(hdrcde)
 set.seed(321)
 
 # Load required functions and prepared data
-source(here("analysis/script/04script.R"))
+source(here("analysis/script/04functions.R"))
 load(here("analysis/data/derived_data/01data.RData"))
 load(here("analysis/data/derived_data/02data.RData"))
 load(here("analysis/data/derived_data/06data.RData"))

### 2022-03-28 analysis/script/08redate.R

@@ -267,12 +267,14 @@ ggsave(file = here("analysis/figures/brunlanes.png"),
        width = 200, height = 150, units = "mm")
 
 
-psdates <- read.csv((here("analysis/data/raw_data/previous_shoreline_dates.csv"))) %>%
+psdates <- read.csv((here(
+  "analysis/data/raw_data/previous_shoreline_dates.csv"))) %>%
   mutate(start = start * -1,
          end = end * -1) %>%
   dplyr::rename("site_name" = "name")
 
-sites_sl <- site_limits %>% filter(radiocarbon == "f" | radiocarbon == "t" & rcarb_cor == "f" | name == "Viulsrød 2")
+sites_sl <- site_limits %>% filter(radiocarbon == "f" | radiocarbon == "t" &
+                                     rcarb_cor == "f" | name == "Viulsrød 2")
 
 sites_sl <- sites_sl %>%
   filter(name %in% psdates$site_name)

### 2022-03-30 analysis/paper/paper.Rmd
Minor edits

@@
 The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach.
-The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age, and in turn have this inform a refined method for shoreline dating.
+The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age, and in turn have this inform an improved method for shoreline dating.
 As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions.
@@
 The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [see review by @nordqvist1999].
-The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @jaksland2014; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below].
+The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @jaksland2014; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below].
 Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time.
@@
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; also discussed in @berg-hansen2009; @bergsvik2009].
-The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020].
+The same logic has also been extended to the hinter- and inland regions, where sites are to be predominantly located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
@@
 While there is displacement data available for other areas of south-eastern Norway [e.g.
-@creel2022; @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
+@hafsten1957; @sørensen1979; @sørensen1999; and recent compilation by @creel2022], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
 This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
@@
 @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments.
-Each basin thus represent a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to a continuous time series for RSL-change, projected to a common isobase [see @creel2022, 5].
+Each basin thus represent a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to devise a continuous time series for RSL-change, projected to a common isobase [see @creel2022, 5].
 Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible.
 Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from the common isobase perpendicular to this gradient (Figure \@ref(fig:map-iso)).
-The SLIPs indicate the isolation of the basin from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves based on the present day tidal range.
+The SLIPs indicate the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves based on the present day tidal range.
 This is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
@@
 For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
-Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same settlement phase.
+Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same settlement phases.
 In the case where there are multiple dates believed to belong to a single phase, these were modelled using the Boundary function in OxCal and then summed.
@@
 Where $\lambda$ is the decay ratio and $x$ is the distance to the shoreline.
-This can then be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:
+This can then be combined with the elevation of the site and the shoreline displacement to provide a method for shoreline dating that takes this distance into account:
 
@@
 Finally, this analysis employed a simulation approach to integrate multiple sources of spatio-temporal uncertainty.
-Here this was simply used to inform the question of the distance between sites and the shore-line.
+Here this was simply used to inform the question of the distance between sites and the shoreline.
 However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent in this reconstruction.

### 2022-03-30 analysis/script/07backwards_dating.R

@@ -45,9 +45,6 @@ rdates <- rdates %>%  group_by(site_name, group) %>%
   ungroup() %>%
   filter(class != "aprior")
 
-
-
-
 # Exclude sites not analysed
 sites_sa <- sites_sa %>%
   filter(!(name %in% c("Dybdalshei 2", "Lunaveien", "Frebergsvik C")))

### 2022-04-09 analysis/paper/paper.Rmd
Minor edits

@@
 Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g.
-@berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @nummedal1923; @mikkelsen1975; @shetelig1922; @solheim2018].
+@berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018].
 The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.
@@
 These conditions are directly reflected in the archaeological record.
-In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [e.g.
+In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement phases that are difficult to disentangle [e.g.
 @hagen1963; @reitan2009].
@@
 Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time.
-This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b].
+This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b, see also @jorgensen2020; @tallavaara2020].
 In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963].
@@
 For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
-Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same settlement phases.
-In the case where there are multiple dates believed to belong to a single phase, these were modelled using the Boundary function in OxCal and then summed.
+Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same settlement phase.
+In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed.
 Multiple phases at a single site were treated as independent of each other.
@@
 The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the simulated sea-level.
-Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age, corresponding with the literature.
+Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.
 From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
@@
 Where $\lambda$ is the decay ratio and $x$ is the distance to the shoreline.
-This can then be combined with the elevation of the site and the shoreline displacement to provide a method for shoreline dating that takes this distance into account:
+This can then be combined with the elevation of the site and the shoreline displacement data to provide a method for shoreline dating that takes this distance into account:
 
@@
 
-Equation \@ref(eq:shoredate) is used to shoreline date the same sites from where this relationship was derived (Figure \@ref(fig:shoredate)), employing $\lambda$ = `r round(expfit$estimate, 3)` for the decay ratio in \@ref(eq:decay), which was found when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
-The Late Neolithic sites were also included for illustrative purposes.
+Eq.
+\@ref(eq:shoredate) is used to shoreline date the same sites from where this relationship was derived (Figure \@ref(fig:shoredate)), employing $\lambda$ = `r round(expfit$estimate, 3)` for the decay ratio in \@ref(eq:decay).
+This is the ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
+For illustrative purposes the Late Neolithic sites were also subjected to the method for shoreline dating.
 Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits.
@@
 The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement.
-When excluding the earliest phase at Gunnarsrød 5, the deviation of which is to be expected based on the simulation results (see above), the shoreline date correspond to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%).
+When excluding the earliest occupation phase at Gunnarsrød 5, the deviation of which is to be expected based on the simulation results (see above), the shoreline date correspond to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%).
 Only including dates modelled to be older than 2500 BCE with 95% probability, i.e.
@@
 only Mesolithic, the success rate is further increased to `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100)`%).
-The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the possible implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (see also Figure \@ref(fig:results)).
+The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the likely implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (see also Figure \@ref(fig:results)).
 
@@
 The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black.
-These are plotted against the modelled radiocarbon dates, which are given colour from oldest to youngest phase for each site, defined by non-overlapping dates at 99.7\\% probability.", out.width = "525px", fig.align="center", regfloat=TRUE}
+These are plotted against the modelled radiocarbon dates, which are given colour from oldest to youngest occupation phase for each site, defined by non-overlapping dates at 99.7\\% probability.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 
-To further explore the implementation for shoreline dating given above, excavated and shoreline dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the outlined approach (Figure \@ref(fig:redate)).
+To further explore the implementation for shoreline dating presented above, excavated and shoreline dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the outlined approach (Figure \@ref(fig:redate)).
 The resulting dates are compared to those originally proposed in the excavation reports for the sites (the numerical results are available in the supplementary material).
@@
 That being said, the results do also indicate that shoreline dating has at times been applied with an exaggerated degree of precision.
-While the implications of a more stable RSL-change for the duration of use and re-use of site locations, and consequently the precision of shoreline dating, is well known, it appears somewhat under-appreciated.
+While the implications of a more stable RSL-change for the duration of use and re-use of site locations are well known, this also appears to be somewhat under-appreciated for the purposes of shoreline dating.
 The results also highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% HDRs for the dates.
@@
 Furthermore, based on the quantitative nature of these findings, an initial formulation of a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed.
-Apart from taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves implementing equation \@ref(eq:shoredate) to account for the distance between the sites and the shoreline.
+Apart from taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves implementing Eq.
+\@ref(eq:shoredate) to account for the distance between the sites and the shoreline.
 When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio of `r round(expfit$estimate, 3)` (Figure \@ref(fig:redate)A) to characterise this relationship.
@@
 The potential of dimensions such as these was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline.
-If patterns related to such locational patterns can be discerned and unpicked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the site beyond its location is available.
+If patterns related to such locational patterns can be discerned and unpicked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information beyond their location is available.
 
@@
 Secondly, the DTM has only been corrected for major modern disturbances.
-This means that erosion, although likely not that prevalent, has not been taken into account.
+This means that other forms of erosion, although likely not that prevalent, has not been taken into account.
 Thirdly, the DTM has a vertical error which could also benefit from being integrated in the analysis [cf.
@@
 @creel2022].
-Fifthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration.
-Finally, the radiocarbon dates and division of phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7% as representing unrelated events.
-This could also be handled differently [@bronkramsey2015].
+Fifthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration [cf.
+@mjærum2022].
+Finally, the radiocarbon dates and division of settlement phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7% as representing unrelated occupations.
+This could also be handled differently [@bronkramsey2009; @bronkramsey2015].
 While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration.

### 2022-04-10 analysis/paper/paper.Rmd
Adding the Pauler data to all the other pre-reported shoreline dates and elevations

@@
 Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time.
-This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b, see also @jorgensen2020; @tallavaara2020].
+This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b; see also @jorgensen2020; @tallavaara2020].
 In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963].
@@
 
-As the excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure, the area immediately surrounding the sites has sometimes been severely impacted by modern disturbances.
+The excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure.
+As the data collection for the DTM was started by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has therefore sometimes been severely impacted by disturbances after the excavation of the site.
 In addition to employing 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster.

### 2022-04-10 analysis/supplementary-materials/supplementary.Rmd
Adding the Pauler data to all the other pre-reported shoreline dates and elevations

@@
 
+Typological indicators in the lithic inventory from Brunstad 24 matches the ^14^C-dates [@danielsen2018].
+The footprint of a large building structure covering a large part of the site from the south required editing of the DTM.
+Apart from that the landscape appears relatively undisturbed.
+8 fragments from a pot could potentially correspond with the ^14^C-dates to the Late Bronze Age and Early Iron Age, but given their uncetaint nature and the limited number of sherds this was disregarded here.
+
 ```{r brunstad25, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 ```
+
+As with Brunstad 24 (above), the artefact inventory from Brunstad 25 matches the radiocarbon dates [@reitan2018], and the building structure covering the southern part of the site had impacted the DTM and had to be removed.
 

### 2022-04-11 analysis/paper/paper.Rmd
Minor edits

@@
 For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
-Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same settlement phase.
+Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same event, here termed settlement phases.
 In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed.
@@
 ```{r redate, echo = FALSE, fig.cap = "Re-dating previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
-The range of the 95\\% HDRs in grey are compared to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
+The 95\\% HDRs in grey are compared to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/redate.png"))

### 2022-04-13 analysis/paper/paper.Rmd
Minor edits

@@
 Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time.
-This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b; see also @jorgensen2020; @tallavaara2020].
+This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020].
 In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963].
@@
 This is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
+The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021; 30cm at the standard port Helgeroa in Larvik].
 Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
@@
 For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
-Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same event, here termed settlement phases.
+Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same event, here termed settlement or site phases.
 In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed.
@@
 That is, for 95% of the cases, the shoreline was simulated to be situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
-While these values remain the same when only the Mesolithic dates are included (Figure \@ref(fig:results2)B), the mean and standard deviation is slightly constrained.
+While these values remain the same when only the Mesolithic dates are included (Figure \@ref(fig:results2)B), the mean and standard deviation are slightly constrained.
 Furthermore, while the median for horizontal and topographic distance is only `r round(median(distances2$topodist))`m across all plots in Figure \@ref(fig:results2), the variation in the statistics for dispersion is greater, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
@@
 Eq.
-\@ref(eq:shoredate) is used to shoreline date the same sites from where this relationship was derived (Figure \@ref(fig:shoredate)), employing $\lambda$ = `r round(expfit$estimate, 3)` for the decay ratio in \@ref(eq:decay).
+\@ref(eq:shoredate) is used to shoreline date the same sites from where this relationship was derived (Figure \@ref(fig:shoredate)), employing $\lambda$ = `r round(expfit$estimate, 3)` for the decay ratio in Eq.
+\@ref(eq:decay).
 This is the ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
-For illustrative purposes the Late Neolithic sites were also subjected to the method for shoreline dating.
+For illustrative purposes the Late Neolithic sites were also subjected to the proposed method for shoreline dating.
 Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits.
-The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by @parnell2008 [Figure \@ref(fig:shoredatediff)].
+The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by @parnell2008.
 Here, 100,000 age samples drawn from the probability density function of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding modelled ^14^C-dates.
-The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement.
-When excluding the earliest occupation phase at Gunnarsrød 5, the deviation of which is to be expected based on the simulation results (see above), the shoreline date correspond to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%).
+The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement (Figure \@ref(fig:shoredatediff)).
+When excluding the earliest occupation phase at Gunnarsrød 5, the deviation of which is to be expected based on issues with the DTM (see above), the shoreline date correspond to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%).
 Only including dates modelled to be older than 2500 BCE with 95% probability, i.e.
@@
 When only including dates older than 4000 BCE with 95% probability, i.e.
-only Mesolithic, the success rate is further increased to `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100)`%).
+only Mesolithic site phases, the success rate is further increased to `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100)`%).
 The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the likely implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (see also Figure \@ref(fig:results)).
@@
 
-To further explore the implementation for shoreline dating presented above, excavated and shoreline dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the outlined approach (Figure \@ref(fig:redate)).
+To further explore the implementation for shoreline dating presented above, excavated and shoreline dated Stone Age sites within the study area where ^14^C-dates are not available or these are not believed to date the main occupation of the sites have been subjected to the outlined approach (Figure \@ref(fig:redate)).
 The resulting dates are compared to those originally proposed in the excavation reports for the sites (the numerical results are available in the supplementary material).
@@
 
-```{r redate, echo = FALSE, fig.cap = "Re-dating previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
+```{r redate, echo = FALSE, fig.cap = "Re-dating excavated and previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
 The 95\\% HDRs in grey are compared to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
@@
 2500 BCE.
-This development is in agreement with the literature.
+This development is in clear agreement with the literature.
 Furthermore, based on the quantitative nature of these findings, an initial formulation of a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed.
@@
 @mjærum2022].
-Finally, the radiocarbon dates and division of settlement phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7% as representing unrelated occupations.
-This could also be handled differently [@bronkramsey2009; @bronkramsey2015].
+Finally, the radiocarbon dates and division of settlement phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7% as representing unrelated occupation events.
+This could also be handled differently [e.g.
+@bronkramsey2009; @bronkramsey2015].
 While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration.

### 2022-04-13 analysis/supplementary-materials/supplementary.Rmd
Minor edits

@@
 Apart from that the landscape appears relatively undisturbed.
-8 fragments from a pot could potentially correspond with the ^14^C-dates to the Late Bronze Age and Early Iron Age, but given their uncetaint nature and the limited number of sherds this was disregarded here.
+8 fragments from a pot could potentially correspond with the ^14^C-dates to the Late Bronze Age and Early Iron Age, but given their uncetain nature and the limited number of sherds, this was disregarded here.
 
@@
 Nonetheless, the site and the ^14^C-dates are still treated the same way as the other sites.
-This is based on the spatial relation between the lithic artefacts and the features, which offers some additional support to the notion that the cooking pits represent remnants from visits in the Mesolithic.
+This is based on the agreement between the ^14^C-dates ans the spatial relation between the lithic artefacts and the features, which offers some additional support to the notion that the cooking pits represent remnants from visits in the Mesolithic.
 
@@
 
-The situation on Dybdalshei 2 is similar to that on Dybdalshei 1 [above, @granum2018], but with only a single date from a single feature and no lithics, the site is not included in the final results.
+The situation on Dybdalshei 2 is similar to that on Dybdalshei 1 [above, @granum2018], but with only a single date from a single feature and no lithics, the site is not included in subsequent analysis.
 
@@
 The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated [@reitan2014d].
-While the editing and interpolation of the DTM appears to give reasonable results for the rest of the flatter marsh area where the railway runs today (see also Gunnarsrød 4, above), the site itself would thus have been situated slightly higher.
+While the editing and interpolation of the DTM appears to give reasonable results for the rest of the flatter marsh area where the railway runs today (see also Gunnarsrød 4, above), the site itself would have been situated slightly higher.
 This has likely led to the negative values on the distance measures.
 In combination, the simulation results and the site map in the report [@reitan2014d, fig.9.1] indicate that the site would not have been situated any significant distance from the shoreline.
-The values have therefore been set to zero for the final aggregative analysis.
+The negative values have therefore been set to zero for the final aggregative analysis.
 A note can also be made on the date that is seen as unrelated to the artefact inventory.
@@
 The site is not located far from the highway, which was also edited for the analysis of Hovland 1 and 4, as well as Torstvet (see below).
+It could be noted that all of the Hovland sites and Torstvet (sites from the E18 Brunlanes project) are located a relatively short vertical distance from the sea, but the terrain makes this result in quite high horizontal and topographic distances.
 
@@
 
-Hovland 3 is securely dated both typologically and radiometrically [@solheim2013]. It is, however, located just by the highway. This appears to have been successfully removed.
+Hovland 3 is securely dated both typologically and radiometrically [@solheim2013]. The site located right by the highway, but this appears to have been successfully removed.
 
@@
 
-Hovland 5 is situated more withdrawn from the highway than the other Hovland sites (above), and the DTM did therefore not require any edits here.
+Hovland 5 is situated more withdrawn from the highway than the other Hovland sites (above), and the DTM did therefore not require any editing.
 The site only has a single ^14^C-date to the Stone Age, but this does match the typological indicators of the assemblage [@mansrud2013].
@@
 To what degree the different dates are related to different or the same settlement phases is difficult to ascertain, but is here, as with for the rest of the sites, simply based on grouping dates overlapping with 99.7% probability.
-The site is situated around 100 meter from a road to the south, but this does not appear to impact simulation results, and is therefore left unedited in the DTM.
+The site is situated around 100 meter from a road to the south, but this does not appear to impact simulation results, which is therefore left unedited in the DTM.
 
@@
 
-The oldest date is not seen as relevant for the main occupation on Krøgenes D2, in part also due to the large error [@mansrud2018b].
-The site location is similar to that of Krøgenes D1 (above), and did not require any editing.
+Dates match the artefact inventory of Krøgenes D2, with the exception of the oldest date which is not seen as relevant for the main occupation of the site, in part also due to the large error [@mansrud2018b].
+The site location is similar to that of Krøgenes D1 (above), and did not require any editing of the DTM.
 
@@
 
-The two find areas are predominantly Mesolithic (to the south-west) and Nesolithic (to the north-east) but were treated in combination here as one Mesolithic date is from the Neolithic area [@stokke2018b].
+The two find areas at Kvastad A2 are predominantly Mesolithic (to the south-west) and Neolithic (to the north-east) but were treated in combination here as one Mesolithic date was from the Neolithic area [@stokke2018b].
 There is also a late Neolithic element in the lithic inventory that is not reflected in the ^14^C-dates.
@@
 Typology of the site inventory from Langangen Vestgård 3 match the ^14^C-date [@eggen2014a].
-Typological indicators are too coarse to offer any insight on the division into two phases done here for the ^14^C-dates.
+Typological indicators are too coarse to offer any insight on the division into two phases done here for the ^14^C-dates (see main text).
 No editing of the DTM was necessary.
@@
 
-Langangen Vestgård 5. Typology match the ^14^C-dates and a division of two phases, one Late Mesolithic and one Early Neolithic [@reitan2014b]. No editing of the DTM necessary.
+Langangen Vestgård 5. Typology match the ^14^C-dates and a division into two settlement phases, one Late Mesolithic and one Early Neolithic [@reitan2014b]. No editing of the DTM deemed necessary.
 
@@
 
-Langangen Vestgård 7 is represented by a cooking pit found on one of two terraces scattered with lithics that was not known or originally part of the excavation project  [@reitan2014j].
-It was therefore only investigated to a limited degree.
+Langangen Vestgård 7 is represented by a cooking pit found on one of two terraces scattered with lithics that was not known and not originally part of the excavation project [@reitan2014j].
+The site was therefore only investigated to a limited degree.
 However, the material does typologically match the ^14^C-date.
@@
 Langemyr is typologically and radiometrically dated to the Middle Mesolithic [@koxvold2018a].
-Construction area just south the site was removed.
+A construction area just south the site was removed.
 This is the same highway-project and raster editing that was done for some of the Hovland sites above.
@@
 
-Lunaveien is a grave cairn site, and the report states that it is unlikely that any of the finds relate to a settlement pre-dating the construction of the cairn [@reitan2010].
+Lunaveien is a grave cairn site, and the report states that it is unlikely that any of the finds relate to a settlement [@reitan2010].
 The results have therefore been excluded from the aggregative analysis.
@@
 Typology of the lithic inventory on Nauen A points to a Late Mesolithic date [@persson2008].
-The later date is seen as related to the establishment of a field for grazing or agriculture, which is also supported by dates that fall outside the analysis here.
+The later date is seen as related to the establishment of a field for grazing or agriculture, which is also supported by younger dates that fall outside the analysis here.
 A section of the highway located a few hundred meters to the south was pre-emptively edited in the DTM.
@@
 
-Site dated to the Late Neolithic, Bronze Age and Early Iron Age [@gjerpe2008b] with long-house 4 dated to the LN.
-Lithics and pottery finds support the dates.
-Site located on what is today a highway which was edited also for Nordby 52 (below), but is not relevant for the sea-level reconstruction for Nordby 1.
+Nordby 1 is dated to the Late Neolithic, Bronze Age and Early Iron Age [@gjerpe2008b] with long-house 4 dated to the Late Neolithic.
+Lithics and pottery finds support the ^14^C-dates.
+The site located on what is today a highway which was edited also for Nordby 52 (below), but is not relevant for the sea-level reconstruction for Nordby 1.
 
@@
 Nordby 52 is a rock-shelter that has been visited throughout prehistory [@gjerpe2008].
-Clear typological date to both of the Stone Age periods defined by the ^14^C-dates treated here.
+The site has a clear typological date to both of the Stone Age periods defined by the ^14^C-dates treated here.
 The site is located by where the highway runs today (see also Nordby 1 above).
@@
 Pauler 1 has a distinctly Early Mesolithic artefact inventory that is not related to the ^14^C-dates [@schallerahrberg2012].
-No modern disturbances appear to imapact the reconstruction of the sea-level corresponding to the ^14^C-dates.
 
@@
 As with Pauler 1 (above), Pauler 2 has a distinctly Early Mesolithic artefact inventory that is not related to the ^14^C-dates [@nyland2012].
-No modern disturbances appear to imapact the reconstruction of the sea-level corresponding to the ^14^C-dates.
 
@@
 The lithic inventory from Pjonkerød R1 is consistent with the ^14^C-date [@carrasco2015].
-The site is situated by a gravel pit today, which has been fairly successfully edited.
+The site is situated by a large gravel pit today, which has been fairly successfully edited.
 However, as the few negative values in the results are likely to result from the interpolation process and the fact that this is a quite large area, as opposed to issues with the displacement curves or ^14^C-dates, the negative values were forced to zero for the aggregative analysis.
@@
 
-Lithic inventory match the ^14^C-date from Rødbøl 54 [@mansrud2008]. The site is situated where the highway runs today, which required editing.
+Lithic inventory match the ^14^C-date from Rødbøl 54 [@mansrud2008]. The site is situated where the highway runs today, which required editing and which is deemed to have been successful.
 
@@
 The site limit has been manually defined based on the description given in the report and the spatial geometry of the site as defined in the database Askeladden.
-It could be slightly off, but should match the elevation given in the publication of the site [@ingstad1970].The ^14^C-dates from the original excavation was not seen as contemporaneous with the lithic material on the site.
+The manually defined limit could be slightly off, but should match the elevation of the site given in the publication [@ingstad1970].The ^14^C-dates from the original excavation was not seen as contemporaneous with the lithic material on the site.
 Burnt elk bone from the original excavation was therefore recently subjected to radiocarbon dating and is believed to match the occupation of the site [pers.comm.
 Steinar Solheim, also provided in @stokke2017, 24].
+Given the 0m distance that the older dates give here, however, there does not appear to be grounds on which to reject the older dates based on sea-level change.
 
@@
 ^14^C-dates from Sagene B1 do not match the Early Mesolithic inventory of the site [@viken2018].
-The wide discrepancy of the dates from the the post-holes also led @viken2018 to suggest that these features might actually be non-anthropogenic and rather be related to trees that have grown on the site.
+The wide discrepancy of the dates from the the post-holes also led @viken2018 to suggest that these features might actually be non-anthropogenic and that they could instead be related to trees that have grown on the site.
 
@@
 
+Tverdal is an unpublished site. Preliminary analysis indicates that radiocarbon dates and typological indicators match. The area surrounding the site appears relatively undisturbed by modern activity.
+
 ```{r vallermyrene1a, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+Vallermyrene 1 was divided into 1a and 1b based on elevation difference, typology and 14C-dates [@reitan2014f].
+Vallermyene 1a is the oldest of the two.
+Although situated not that far from the railway, the area immediatly surrounding the site has not been disturbed in a way that appears to impact the reconstruction of the past sea-level.
+
 ```{r vallermyrene1b, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+Vallermyrene 1b is the younger than Vallermyrene 1a (above), as indicated both by the radiocarbon date, typology and elevation [@reitan2014f]. The DTM did not require editing.
+
 ```{r vallermyrene2, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+Vallermyrene 2 was a large site situated in a field [@reitan2014o].
+The site has ^14^C-dates and artefacts dating occupations at the sites to multiple phases throughout prehistory, including the Bronze Age and phases up until the Viking Age.
+The Stone Age activity on the site was evidenced by 359 lithics and a fireplace dated to the Early Neolithic.
+Given it's uncertain extent, the site limit used here is defined by the fireplace.
+
 ```{r vallermyrene4a, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 
+Vallermyrene 4 was, as with Vallermyrene 1 (above), divided into Vallermyrene 4a and 4b based on radiocarbon dates and differences in elevation [@eigeland2014].
+Typologically the sites are identical, and match the ^14^C-dates.
+Vallermyrene 4a is the older of the two and was less disturbed by modern acitivity before the excavation.
+The railway runs just east and north-east of the site and had to be removed.
+In addition, what appears to be a gravel pit was located to the south-west.
+Both railway and gravel pit appear to have been adequately edited in the employed DTM.
+
 ```{r vallermyrene4b, echo = FALSE, out.width = "500px", regfloat=TRUE}
@@
 ```
+
+Vallermyrene 4b is slightly younger than Vallermyrene 4a (above).
+Typology and radiocarbon dates match [@eigeland2014].
+The DTM was edited to remove the railway track running by the site and a gravel pit.

### 2022-04-20 analysis/paper/paper.Rmd
Adding a figure to visualise the similarity with 14C calibration

@@
 This is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
-The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021; 30cm at the standard port Helgeroa in Larvik].
+The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021, 30cm at the standard port Helgeroa in Larvik].
 Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.

### 2022-04-20 analysis/supplementary-materials/supplementary.Rmd
Adding a figure to visualise the similarity with 14C calibration

@@
 
-# ```{r ragnhildrod, echo = FALSE, out.width = "500px", regfloat=TRUE}
-# include_graphics(here("analysis/figures/ragnhildrod.png"))
-# 
-# filter(rcarbon, site_name  == "Ragnhildrød") %>% 
-#   select(lab_code, c14_bp, error, material, context) %>% 
-#    kable(caption = "Ragnhildrød", col.names = c("ID",
-#                     "\\textsuperscript{14}C BP",
-#                     "Error",
-#                     "Material",
-#                     "Context"),
-#         booktabs = TRUE,  linesep = c(""), escape = FALSE)
-# ```
+```{r ragnhildrod, echo = FALSE, out.width = "500px", regfloat=TRUE, warning = FALSE}
+include_graphics(here("analysis/figures/ragnhildrod.png"))
+
+filter(rcarbon, site_name  == "Ragnhildrød") %>% 
+   select(lab_code, c14_bp, error, material, context) %>% 
+    kable(caption = "Ragnhildrød", col.names = c("ID",
+                     "\\textsuperscript{14}C BP",
+                     "Error",
+                     "Material",
+                     "Context"),
+         booktabs = TRUE,  linesep = c(""), escape = FALSE)
+```
 

### 2022-04-20 analysis/script/05site_analysis.R

@@ -1171,10 +1171,10 @@ ggsave(file = here("analysis/figures/prestemoen1.png"), width = 250,
 sitename <- "Ragnhildrød"
 date_groups <- group_dates(rcarb_sa, sitename)
 
-output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
-                          isobases, nsamp = nsamp, loc_bbox = 4000, siterpath)
-save(output,
-     file = here::here("analysis/data/derived_data/ragnhildrod.RData"))
+# output <- apply_functions(sitename, date_groups, dtm, displacement_curves,
+#                           isobases, nsamp = nsamp, loc_bbox = 4000, siterpath)
+# save(output,
+#      file = here::here("analysis/data/derived_data/ragnhildrod.RData"))
 load(here("analysis/data/derived_data/ragnhildrod.RData"))
 
 sitearea <- rast(file.path(siterpath,

### 2022-04-20 analysis/paper/paper.Rmd
Formulating the dating procedure

@@
 
-The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [see review by @nordqvist1999].
+The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than an earliest possible date for when a site could have been in use [see review by @nordqvist1999].
 The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @jaksland2014; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below].
@@
 ```
+
+```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites.
+Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occured.
+Results associated with agricultural acitivites are plotted in grey.
+The first row A) shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites.
+The second row B) shows the result of excluding these.
+The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center", regfloat=TRUE}
+knitr::include_graphics(here("analysis/figures/results.png"))
+
+```
+
+```{r results2, echo = FALSE, fig.cap = "Histograms showing the simulated distance from the shoreline using dates corresponding to the site inventories.
+Negative values have been set to zero.
+A) Simulated results older than 2500 BCE, and B) simulated results older than 4000 BCE.", out.width = "525px", fig.align="center", regfloat=TRUE}
+knitr::include_graphics(here("analysis/figures/results2.png"))
+```
+
+```{r exampledat, echo = FALSE, fig.cap = "Shoreline dating Hegna vest 1.
+The dashed line indicates the elevation of the site.
+The decay ratio follows Figure 7A.
+The shoreline date in grey is underlined with the 95\\% HDR in black.", out.width = "500px", fig.align="center", regfloat=TRUE}
+knitr::include_graphics(here("analysis/figures/sdate.png"))
+
+```
+
+# Shoreline dating
+
 An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation (Figure \@ref(fig:results2)).
@@
 However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
-If one accepts this, the probability density function for exponential decay can be used to characterise the vertical distance between sites and the shoreline:
+If one accepts this, the probability density function for exponential decay can be used to characterise the vertical distance between sites and the shoreline and leveraged for devising a method for shoreline dating that takes this into account.
+
+The procedure is thus aimed at determining the likely age of the occupation of a site based on its altitude above present day sea-level, with reference to shoreline displacement and the likely elevation of the site above sea-level when it was in use.
+For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly.
+This leads the problem to become analogous to that of the calibration of a radiocarbon date.
+Drawing on the standard Bayesian formulation for the calibration procedure for 14C-dates [e.g.
+@bronkramsey2009], the probability density associated with the calendar age can then be given as
+
+\begin{equation} 
+  p(x_{i} | \thetha_{i}) ~ p(\theta_{i} | x_{i})p(x_{i})
+  (\#eq:bayes)
+\end{equation}
+
+where for any site i the unknown calendar date is denoted $\thetha$ and the elevation of the contemporaneous sea-level $x$.
+The elevation of the sea-level is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $d$.
+Based on the simulation results above, the distance from the elevation of the site to the contemporaneous shoreline is defined by the probability density function for exponential decay
 
@@
 \begin{equation}
-  f(x) = \lambda {e}^{- \lambda x}
+  p(d) = \lambda {e}^{- \lambda x}
   (\#eq:decay)
@@
 
-Where $\lambda$ is the decay ratio and $x$ is the distance to the shoreline.
-This can then be combined with the elevation of the site and the shoreline displacement data to provide a method for shoreline dating that takes this distance into account:
+Where $\lambda$ is the decay ratio.
+This can then be coupled with the trajectory of relative sea-level change, denoted RSL, to find the probability density for the calendar date.
+Here the relative sea-level change is defined by a uniform probability density function over the range between the lower and upper displacement curves interpolated to the site location
+
+\begin{equation}
+p(RSL) = U[RSL_{lower}, RSL_{upper}]	
+  (\#eq:rsl)
+\end{equation}
 
@@
 \end{equation}
-
-Where x is [...] the site elevation, $i$ is the trajectory for shoreline displacement, here given as a uniform probability range between the upper and lower bound of the interpolated displacement curve.
 
@@
 The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the likely implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (see also Figure \@ref(fig:results)).
-
-```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites.
-Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occured.
-Results associated with agricultural acitivites are plotted in grey.
-The first row A) shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites.
-The second row B) shows the result of excluding these.
-The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center", regfloat=TRUE}
-knitr::include_graphics(here("analysis/figures/results.png"))
-
-```
-
-```{r results2, echo = FALSE, fig.cap = "Histograms showing the simulated distance from the shoreline using dates corresponding to the site inventories.
-Negative values have been set to zero.
-A) Simulated results older than 2500 BCE, and B) simulated results older than 4000 BCE.", out.width = "525px", fig.align="center", regfloat=TRUE}
-knitr::include_graphics(here("analysis/figures/results2.png"))
-```
 

### 2022-04-20 analysis/paper/paper.tex
Formulating the dating procedure

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\title{A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast}
\author{Isak Roalkvam\textsuperscript{1,*}}
\date{20 April, 2022}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast},
  pdfauthor={Isak Roalkvam1,},
  pdfkeywords={Stone Age; Settlement patterns; Scandinavia; Relative sea-level change; Shoreline dating; Monte Carlo simulation},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{lineno}
\usepackage{afterpage}
\usepackage{fancyhdr}
\linenumbers
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\begin{document}
\maketitle

\textsuperscript{1} University of Oslo, Institute of Archaeology, Conservation and History

\textsuperscript{*} Correspondence: \href{mailto:isak.roalkvam@iakh.uio.no}{Isak Roalkvam \textless{}\href{mailto:isak.roalkvam@iakh.uio.no}{\nolinkurl{isak.roalkvam@iakh.uio.no}}\textgreater{}}

\fancypagestyle{plain}{%
  \fancyhf{}
  \fancyfoot[R]{\leavevmode\rlap{\thepage~}}
}
\pagestyle{plain}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites (e.g. Åkerlund 1996; Bjerck 2005; Nordqvist 1995; Schmitt et al. 2009; Sognnes 2003; Tallavaara and Pesonen 2020; Wikell et al. 2009).

The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was established by the end of the 19th century (De Geer 1896), and was first applied as a dating method at the turn of the century (Brøgger 1905; Hollender 1901). Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since (e.g. Berg-Hansen 2009; Bjerck 1990, 2008a; Breivik 2014; Johansen 1963; Mikkelsen 1975; Mjærum 2022; Nummedal 1923; Shetelig 1922; Solheim and Persson 2018). The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

Despite its important role for Norwegian Stone Age archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age, and in turn have this inform an improved method for shoreline dating. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shoreline is inherently spatial. Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation (e.g. Bevan et al. 2013; Crema et al. 2010; Crema 2012, 2015; Yubero-Gómez et al. 2016), a similar approach is adopted here.

\hypertarget{background}{%
\section{Background}\label{background}}

Relative sea-level (RSL) can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre (Shennan 2015). Variation in this relative distance follow from a range of effects (e.g. Milne et al. 2009). Of central importance here is eustasy and istostasy. The eustatic sea-level is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself (Shennan 2015). The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans (Mörner 1976). Istostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle. This can be the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed. These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight (Milne 2015).

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet (e.g. Hughes et al. 2016; Stroeven et al. 2016, see Figure \ref{fig:deglaciation}), the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise (e.g. Mörner 1979; Svendsen and Mangerud 1987). In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression (e.g. Romundset et al. 2015; Svendsen and Mangerud 1987). These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement phases that are difficult to disentangle (e.g. Hagen 1963; Reitan and Berg-Hansen 2009). Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, leave them still submerged today (Bjerck 2008a; Glørstad et al. 2020). This can lead to a hiatus in the archaeological record for certain sub-phases in the impacted areas. Comparatively, given a continuous and still ongoing shoreline regression from as high as c.~220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions (Hafsten 1957, 1983; Romundset et al. 2018; Sørensen 1979). This makes the region especially useful for evaluating the assumption of a shore-bound settlement pattern over a long and continuous time-span.

\begin{figure}

{\centering \includegraphics[width=525px]{/home/isak/phd/assessing.sealevel.dating/analysis/figures/deglaciation}

}

\caption{Deglaciation at 1000 year intervals from c. 17--8 kyr BCE. The study area defined later in the text is marked with a red outline (deglaciation data from Hughes et al. 2016, but see also Romundset et al. 2019 in relation to the study area).}\label{fig:deglaciation}
\end{figure}

The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than an earliest possible date for when a site could have been in use (see review by Nordqvist 1999). The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods (see for example chapters in Jaksland 2014; Melvold and Persson 2014; Reitan and Persson 2014; Reitan and Sundström 2018; Solheim 2017 and below). Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time. This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins (Breivik 2014; Breivik and Bjerck 2018; Fossum 2020; Mjærum 2022; Nielsen 2021; Solheim and Persson 2018; see also Jørgensen et al. 2020; Tallavaara and Pesonen 2020). In his review, Nordqvist (1999) argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer (see also Johansen 1963).

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways (Bjerck 1990; Brøgger 1905:166; also discussed in Berg-Hansen 2009; Bergsvik 2009). The same logic has also been extended to the hinter- and inland regions, where sites are to be predominantly located along rivers and lakes (Brøgger 1905:166; Glørstad 2010:57--87; but see also Gundersen 2013; Mjærum 2018; Schülke 2020). This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper (Prescott 2020; Solheim 2021). The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm is to have led site locations to be more withdrawn from the shoreline (e.g. Bakka and Kaland 1971; Østmo 2008:223; \textbf{prescott2012?}). That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods (Glørstad 2012). At an earlier stage, at the transition to the Early Neolithic (c.~3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic (Hinsch 1955; Nielsen et al. 2019; Østmo 1988:225--227). Nielsen (2021) has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data (see also e.g. Amundsen et al. 2006; Østmo 1988; Solheim 2012:74). Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c.~2400 BCE.

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example reflected in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels (e.g. Berg-Hansen 2009; Eskeland 2017). Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date-span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound (e.g. Jaksland 2014; Johansen 1963; Nummedal 1923). This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few (see also Ilves and Darmark 2011).

One of the more extensive evaluations of the relationship between archaeological radiocarbon dates and RSL-change was done by Solheim and colleagues (Breivik et al. 2018; Solheim 2020), who compared 102 radiocarbon dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5\% of the sites. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating \textsuperscript{14}C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these dates are entirely erroneous can be difficult to evaluate (e.g. Persson 2008; Schülke 2020). However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas as evidenced by artefact inventories or multiple site features. Not remnants of stays as ephemeral to only be discernible by individual features or dubious \textsuperscript{14}C-dates. The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.

Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of Breivik et al. (2018), involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site (e.g. Åkerlund et al. 1995; Åstveit 2018; Solheim 2020; see also Bjerck 2008b; Kleppe 1985; Ramstad 2009). This approach has a couple of limitations. First of all, the displacement curves are sometimes applied directly to larger study areas, with only some studies having taken the variable uplift-rates into account when performing this comparison (e.g. Åstveit 2018; Fossum 2020; Møller 1987; Persson 2008). Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence (e.g. Lakens et al. 2018), as the question of interest is effectively one of synchroneity between events (cf. Parnell et al. 2008). Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and help move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.

\hypertarget{data}{%
\section{Data}\label{data}}

To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway (e.g. Hafsten 1957; Sørensen 1979, 1999; and recent compilation by Creel et al. 2022), considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten (\textbf{romundset2021?}), Larvik (Sørensen et al. in prep; Sørensen, Henningsmoen, et al. 2014; Sørensen, Høeg, et al. 2014), Tvedestrand (Romundset 2018; Romundset et al. 2018), and Arendal (Romundset 2018).

The employed shoreline displacement data is based on the so-called isolation basin method (e.g. Kjemperud 1986; Romundset et al. 2011), which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments. Each basin thus represent a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to devise a continuous time series for RSL-change, projected to a common isobase (see Creel et al. 2022:5). Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from the common isobase perpendicular to this gradient (Figure \ref{fig:map-iso}). The SLIPs indicate the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves based on the present day tidal range. This is assumed to have been the same throughout the Holocene (Sørensen, Henningsmoen, et al. 2014:44). The highest astronomical tide in the study area reaches around 30cm above mean sea-level (Norwegian Mapping Authority 2021:30cm at the standard port Helgeroa in Larvik). Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few (e.g. Romundset et al. 2011, 2019; for an alternative approach see Creel et al. 2022). For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

The archaeological data compiled for the analysis consists of excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \ref{fig:map-iso}). These number 155 sites. Of these, 91 sites are associated with a total of 547 radiocarbon dates. Of these, in turn, 67 sites are related to the 259 radiocarbon dates that fall within the Stone Age (9500--1700 BCE), with 95\% probability. These sites and \textsuperscript{14}C-dates form the basis for the analysis. Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History---the institution responsible for archaeological excavations in the region. In the compiled dataset, each radiocarbon date has been associated with the site features or excavation unit from where they originate, or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what available spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as that defined during initial survey, area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.

Three of the sites have been associated with agriculture, either directly or in the form building structures. The first is Nordby 1 at which the \textsuperscript{14}C-dates are associated with a Late Neolithic long-house (Gjerpe and Bukkemoen 2008). The Middle Neolithic phase at Kvastad A2 (Stokke and Reitan 2018) and Late Neolithic phase at Nauen A (Persson 2008) are both directly related to farming acitivities. Both of these sites also have radiocarbon dates and lithic inventory associated with Mesolithic forager activities. Following from the expected deviance from the settlement patterns that are to characterise forager sites, these agricultural phases are highlighted in the analysis below. Finally, Nielsen (2021) has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratsberg (Wenn 2012) and Larønningen (Røberg 2012) could be related to early agricultural activity in the Oslo fjord region. Due to the uncertain and somewhat speculative nature of this suggestion, these are omitted here.

The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority (Norwegian Mapping Authority 2018, \url{https://hoydedata.no}). It was here opted for the 10m resolution DTM rather than the higher-resolution 1m version. In addition to resulting in considerably less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by. The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m (Norwegian Mapping Authority 2018). All data and R programming code (R Core Team 2021) required to run the analyses, as well as the derived data are freely available in an online repository at \url{https://osf.io/7f9su/}, organised as a digital research compendium following Marwick et al. (2018).

\begin{figure}

{\centering \includegraphics{../figures/map-iso-1}

}

\caption{A) Distribution of the  67  analysed sites relative to the isobases of the displacement curves. The isobases have a direction of 327$^{\circ}$ (Romundset et al. 2018), B) Displacement curves. Note the increasing steepness of the curves towards the north-east.}\label{fig:map-iso}
\end{figure}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

The method of shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves using inverse distance weighting (e.g. Conolly 2020; Conolly and Lake 2006:94--97). This was done for each year along the entirety of the curves, weighting the interpolation by the standard squared inverse of the distances. The result of this process is shown for an example site in Figure \ref{fig:example-1}. For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve (Reimer et al. 2020) using OxCal v4.4.4 (Bronk Ramsey 2009) through the oxcAAR package for R (Hinz et al. 2021). Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7\% probability, meaning these were effectively taken to represent the same event, here termed settlement or site phases. In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

The excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure. As the data collection for the DTM was started by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has therefore sometimes been severely impacted by disturbances after the excavation of the site. In addition to employing 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension (e.g. Conolly 2020), using the default settings of r.fillnulls from GRASS GIS (GRASS Development Team 2017) in R through the package rgrass7 (Bivand 2021). In addition to code and original spatial data being available in the digital research compendium for this paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.

Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of relative sea-level change at that location, and a DTM edited to remove substantial modern disturbances, the simulations were performed. A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \ref{fig:example-2}). This year then has a corresponding likely elevation range for the contemporaneous shoreline from which an elevation value was drawn uniformly, using intervals of 5cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing. Polygons were then created from the resulting areas with missing values. The horizontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon. The topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM. This was done using the topoDistance package for R (Wang 2019). The topographic distance was measured between the site polygon and the horizontally closest point on the shoreline. This means that the distance is not necessarily measured as the closest topographic distance to the shoreline, but rather as the shortest topographic path to the horizontally closest point on the shoreline. Not finding the topographically closest point significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact on the results given the distances considered. The shortest topographic path was found using the Moore neighbourhood of eight cells (e.g. Conolly and Lake 2006:253; Herzog 2013). In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon. While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values were therefore retained with the exception of for the sites Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the \textsuperscript{14}C-dates or displacement curves (see supplementary material for more details).

This process was repeated 1000 times for each phase for each site. The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5, available in the supplementary material (cf. Crema et al. 2010:1125). Hovland 5 was chosen for this evaluation as it has a fairly imprecise date and is located in area of quite complex topography.

\begin{figure}

{\centering \includegraphics[width=500px]{/home/isak/phd/assessing.sealevel.dating/analysis/figures/example_map}

}

\caption{Example site Hegna vest 1 (Fossum 2017). A) Location of the site on the edited 10m resolution DTM. The red outline is the site limit. B) Radiocarbon dates associated with the site. Fill colour indicates what dates are assumed to belong to the same settlement phase. Multiple dates are modelled using the Boundary function in OxCal and then summed. The red outline indicates that the date does not match the typological indicators in the artefact assemblage of the site. C) The location of the site within the study area relative to isobases of the employed displacement curves. D) Displacement curve interpolated to the site location.}\label{fig:example-1}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=500px]{/home/isak/phd/assessing.sealevel.dating/analysis/figures/example_sim_map}

}

\caption{Example of a single simulation run on the site Hegna vest 1. A) The simulation starts by drawing a single year from the posterior density esimate.  B) This then corresponds to an elevation range on the interpolated displacement curve. A single elevation is drawn uniformly from this range using 5cm intervals. C) The sea-level is then adjusted on the DTM to this elevation and the various distance mesaures are found. D) The numerical result of the simulation run.}\label{fig:example-2}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=500px]{/home/isak/phd/assessing.sealevel.dating/analysis/figures/example_sim_result}

}

\caption{The result of 1000 simulation runs for each of the two groups of dates on the site Hegna vest 1. The first column of plots shows the radiocarbon probability density function from where dates were drawn during simulation. The second column displays the result of simulating the raised sea-level 1000 times. The more opaque the colour, the more times the sea-level was simulated to that location. The third column shows violin plots of the different distance measures across all simulations.}\label{fig:example-3}
\end{figure}

\hypertarget{simulation-results}{%
\section{Simulation results}\label{simulation-results}}

Overall, as is indicated by the measures for central tendency and the almost solid line along the 0m mark on the y-axes, the simulations show that the sites tend to have been situated close to the shoreline when they were in use (Figure \ref{fig:results}). Some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included (Figure \ref{fig:results}A), but if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, Figure \ref{fig:results}B gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \ref{fig:map-iso}B), which leads the uncertainty of the \textsuperscript{14}C-dates to give a wider possible elevation range for the simulated sea-level. Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance. There are also a couple of sites located some distance from the shoreline just after 4000 BCE. While the sample size is limited, this would thus be in line with a development that sees an increase in settlements located in the immediate inland around this time.

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape. While there would be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level, as they would not have been in use when located under water. It could therefore seem that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geological reconstruction of shoreline displacement in the area.

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results for from Figure \ref{fig:results}B is given again in Figure \ref{fig:results2}A, excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The resulting best point estimate for the vertical distance between sites and shoreline for the pre-Late Neolithic is given by the median at 4m, while 95\% of the values fall within the range 0--18m. That is, for 95\% of the cases, the shoreline was simulated to be situated on or down to 18m below the site location. While these values remain the same when only the Mesolithic dates are included (Figure \ref{fig:results2}B), the mean and standard deviation are slightly constrained. Furthermore, while the median for horizontal and topographic distance is only 10m across all plots in Figure \ref{fig:results2}, the variation in the statistics for dispersion is greater, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.

\begin{figure}

{\centering \includegraphics[width=500px]{/home/isak/phd/assessing.sealevel.dating/analysis/figures/results}

}

\caption{The result of running the analysis across all sites. Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occured. Results associated with agricultural acitivites are plotted in grey. The first row A) shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites. The second row B) shows the result of excluding these. The table under each plot lists some corresponding statistics for central tendency and dispersion.}\label{fig:results}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=525px]{/home/isak/phd/assessing.sealevel.dating/analysis/figures/results2}

}

\caption{Histograms showing the simulated distance from the shoreline using dates corresponding to the site inventories. Negative values have been set to zero. A) Simulated results older than 2500 BCE, and B) simulated results older than 4000 BCE.}\label{fig:results2}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=500px]{/home/isak/phd/assessing.sealevel.dating/analysis/figures/sdate}

}

\caption{Shoreline dating Hegna vest 1. The dashed line indicates the elevation of the site. The decay ratio follows Figure 7A. The shoreline date in grey is underlined with the 95\% HDR in black.}\label{fig:exampledat}
\end{figure}

\hypertarget{shoreline-dating}{%
\section{Shoreline dating}\label{shoreline-dating}}

An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation (Figure \ref{fig:results2}). While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data. However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero. If one accepts this, the probability density function for exponential decay can be used to characterise the vertical distance between sites and the shoreline and leveraged for devising a method for shoreline dating that takes this into account.

The procedure is thus aimed at determining the likely age of the occupation of a site based on its altitude above present day sea-level, with reference to shoreline displacement and the likely elevation of the site above sea-level when it was in use. For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly. This leads the problem to become analogous to that of the calibration of a radiocarbon date. Drawing on the standard Bayesian formulation for the calibration procedure for 14C-dates (e.g. Bronk Ramsey 2009), the probability density associated with the calendar age can then be given as

\begin{equation} 
  p(x_{i} | \thetha_{i}) ~ p(\theta_{i} | x_{i})p(x_{i})
  \label{eq:bayes}
\end{equation}

where for any site i the unknown calendar date is denoted \(\thetha\) and the elevation of the contemporaneous sea-level \(x\). The elevation of the sea-level is dependent on the present day elevation of the site \(\alpha\) and the distance between site and the shoreline \(d\). Based on the simulation results above, the distance from the elevation of the site to the contemporaneous shoreline is defined by the probability density function for exponential decay

\begin{equation}
  p(d) = \lambda {e}^{- \lambda x}
  \label{eq:decay}
\end{equation}

Where \(\lambda\) is the decay ratio. This can then be coupled with the trajectory of relative sea-level change, denoted RSL, to find the probability density for the calendar date. Here the relative sea-level change is defined by a uniform probability density function over the range between the lower and upper displacement curves interpolated to the site location

\begin{equation}
p(RSL) = U[RSL_{lower}, RSL_{upper}]    
  \label{eq:rsl}
\end{equation}

\begin{equation} 
  placeholder
  \label{eq:shoredate}
\end{equation}

Eq. \eqref{eq:shoredate} is used to shoreline date the same sites from where this relationship was derived (Figure \ref{fig:shoredate}), employing \(\lambda\) = 0.173 for the decay ratio in Eq. \eqref{eq:decay}. This is the ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \ref{fig:results2}A). For illustrative purposes the Late Neolithic sites were also subjected to the proposed method for shoreline dating. Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits. The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by Parnell et al. (2008). Here, 100,000 age samples drawn from the probability density function of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding modelled \textsuperscript{14}C-dates. The resulting range of the 95\% highest density region (HDR, Hyndman 1996) was then checked to see if it crosses zero, in which case the dates are considered to be in agreement (Figure \ref{fig:shoredatediff}). When excluding the earliest occupation phase at Gunnarsrød 5, the deviation of which is to be expected based on issues with the DTM (see above), the shoreline date correspond to the radiocarbon dates in 58 out of 68 cases (84\%). Only including dates modelled to be older than 2500 BCE with 95\% probability, i.e.~older than the Late Neolithic, improves this to 56 out of 61 cases (92\%). When only including dates older than 4000 BCE with 95\% probability, i.e.~only Mesolithic site phases, the success rate is further increased to 46/49 (94\%). The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the likely implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (see also Figure \ref{fig:results}).

\begin{figure}

{\centering \includegraphics[width=525px]{/home/isak/phd/assessing.sealevel.dating/analysis/figures/shoredate}

}

\caption{The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory using the method proposed here. The shoreline dates are plotted in grey and underlined with the 95\% HDR in black. These are plotted against the modelled radiocarbon dates, which are given colour from oldest to youngest occupation phase for each site, defined by non-overlapping dates at 99.7\% probability.}\label{fig:shoredate}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=525px]{/home/isak/phd/assessing.sealevel.dating/analysis/figures/shoredate2}

}

\caption{Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure 8. When the range of the 95\% HDR for age difference crosses zero, the shoreline and radiocarbon dates are considered synchronous. Line segments with vertical bars indicate that this does not cross zero and that the dates are not in agreement.  The division and colour coding at the top of the plots reflect the divison of site phases given in Figure 8.}\label{fig:shoredatediff}
\end{figure}

\hypertarget{re-dating-previously-shoreline-dated-sites}{%
\section{Re-dating previously shoreline dated sites}\label{re-dating-previously-shoreline-dated-sites}}

To further explore the implementation for shoreline dating presented above, excavated and shoreline dated Stone Age sites within the study area where \textsuperscript{14}C-dates are not available or these are not believed to date the main occupation of the sites have been subjected to the outlined approach (Figure \ref{fig:redate}). The resulting dates are compared to those originally proposed in the excavation reports for the sites (the numerical results are available in the supplementary material). To avoid issues with recent disturbances on the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.

The comparison with previously reported dates is an illustrative, but unfair exercise for a few reasons. First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an undefined, but implied or explicit uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Finally, they are sometimes only meant to indicate a lower bound for when the sites could have been in use. Overall, the results suggest that shoreline dating has generally been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner (although see e.g. Roalkvam 2022). That being said, the results do also indicate that shoreline dating has at times been applied with an exaggerated degree of precision. While the implications of a more stable RSL-change for the duration of use and re-use of site locations are well known, this also appears to be somewhat under-appreciated for the purposes of shoreline dating. The results also highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95\% HDRs for the dates. In some cases the method provides a very precise date range and in others it offers little more than a \emph{terminus post quem}. This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf.~Figure \ref{fig:map-iso}B). Further, as some of the date ranges extend well beyond major chronological divisions, even into the Iron Age, they could be severely and securely constrained with only cursory reference to typology. While this would be trivial in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed. This also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to further improve the precision of the dates that can be achieved with shoreline dating.

\begin{figure}

{\centering \includegraphics[width=525px]{/home/isak/phd/assessing.sealevel.dating/analysis/figures/redate}

}

\caption{Re-dating excavated and previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories. The 95\% HDRs in grey are compared to the dates originally proposed by the excavation reports in red.}\label{fig:redate}
\end{figure}

Not least following from the fact that relatively few Preboreal \textsuperscript{14}C-dates associated with anthropogenic activity have been achieved in Norway (Åstveit 2018; Damlien and Solheim 2018; Kleppe 2018), the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula (e.g. Bang-Andersen 2012; Berg-Hansen 2018; Breivik 2014; Fuglestvedt 2012; Glørstad 2016). The shoreline dated Preboreal sites from the Brunlanes-project are among the earliest known sites in Norway (Jaksland 2012a, 2012b; Jaksland and Persson 2014). These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, Jaksland (2014) provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \ref{fig:brunlanes}A. The sites have been dated using what Jaksland (2014) gives as the lowest elevation of finds at each site, and by employing a exponential decay ratio of 0.13, to allow for more deviance in the distance between site and shoreline. This corresponds to the decay ratio for sites older than 7000 BCE in Figure \ref{fig:results2}.

The small discrepancies between the achieved results mainly follow from the fact that a slightly updated version of the local displacement curve is applied here (cf. Sørensen et al. in prep). Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were to be considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap. This information is not integrated in the approach outlined here, but could justify further reducing the uncertainty ranges. Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does result in very precise dates. Above it was suggested that additional temporal data could be combined with the method to improve its accuracy and precision. This example, on the other hand, highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use. One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \ref{fig:brunlanes}B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline. If this is developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use. Such approaches would make less of an impact in this setting, where the 95\% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf.~Figure \ref{fig:redate}).

\begin{figure}

{\centering \includegraphics[width=525px]{/home/isak/phd/assessing.sealevel.dating/analysis/figures/brunlanes}

}

\caption{Shoreline dating of the Brunlanes sites using site altitudes provided by Jaksland (2014:tab.4). A) The result of applying the approach to shoreline dating outlined above. The shoreline date in grey is underlined with the 95\% HDR in black. Dates provided by Jaksland (2014) are plotted in red. The box indicates a 50 year uncertainty range which in combination with the red line extends 200 years. B) Map showing the centroids of the Pauler sites and Sky 1. The sea-level has been simulated using the probability density associated with the shoreline date for Pauler 1 (see also map in Jaksland 2014:fig.12a). Pauler 1 is the red point.}\label{fig:brunlanes}
\end{figure}

\hypertarget{concluding-remarks}{%
\section{Concluding remarks}\label{concluding-remarks}}

The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline. This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c.~4000 BCE, after which a couple of sites become situated some distance from the sea, followed by a more decisive break at the transition to the Late Neolithic at c.~2500 BCE. This development is in clear agreement with the literature. Furthermore, based on the quantitative nature of these findings, an initial formulation of a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed. Apart from taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves implementing Eq. \eqref{eq:shoredate} to account for the distance between the sites and the shoreline. When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio of 0.173 (Figure \ref{fig:redate}A) to characterise this relationship. Furthermore, while this remains to be formalised and explored further, it was also showed how the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data. As the precision of the method is both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date than younger sites located towards the south-west, the impact of such additional information will also vary.

Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further evaluate and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted. Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located. Furthermore, other dimensions related to the topographic location of the sites could be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern (Roalkvam 2020), and which presumably has implications for how close to the shoreline people settled (Blankholm 2020; Helskog 1978). This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range could possibly also have implications for the site location relative to the shoreline, depending on the topography (Helskog 1978). The potential of dimensions such as these was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline. If patterns related to such locational patterns can be discerned and unpicked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information beyond their location is available.

Some limitations and sources of likely variation and uncertainty that have not been considered should also be mentioned. First of all the sample size is quite strained and the future addition more sites might alter the picture considerably. Secondly, the DTM has only been corrected for major modern disturbances. This means that other forms of erosion, although likely not that prevalent, has not been taken into account. Thirdly, the DTM has a vertical error which could also benefit from being integrated in the analysis (cf. Lewis 2021). Fourthly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves further away from the isobases of the displacement curves. This is also related to the fact that the RSL data can be handled in different ways than with the isobase method that has been used for the compilation of the employed displacement curves (cf. Creel et al. 2022). Fifthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration (cf. Mjærum 2022). Finally, the radiocarbon dates and division of settlement phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7\% as representing unrelated occupation events. This could also be handled differently (e.g. Bronk Ramsey 2009, 2015). While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration.

Finally, this analysis employed a simulation approach to integrate multiple sources of spatio-temporal uncertainty. Here this was simply used to inform the question of the distance between sites and the shoreline. However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent in this reconstruction.

\newpage

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-akerlund1996}{}}%
\CSLBlock{Åkerlund, Agneta}
\CSLLeftMargin{ 1996}
\CSLRightInline{\emph{{Human responses to shore displacement: Living by the sea in Eastern Middle Sweden during the Stone Age}}. Riksantikvarieämbetet, Stockholm.}

\leavevmode\vadjust pre{\hypertarget{ref-akerlund1995}{}}%
\CSLBlock{Åkerlund, Agneta, Jan Risberg, Urve Miller, and Per Gustafsson}
\CSLLeftMargin{ 1995}
\CSLRightInline{{On the applicability of the \(^{14}\)C method to interdisciplinary studies on shore displacement and settlement location}. \emph{PACT} 49:53--84.}

\leavevmode\vadjust pre{\hypertarget{ref-amundsen2006}{}}%
\CSLBlock{Amundsen, Øystein, Stig Knutsen, Axel Mjærum, and Gaute Reitan}
\CSLLeftMargin{ 2006}
\CSLRightInline{{Nøkleby i Ski -- en tidligneolittisk jordbruksboplass?} \emph{Primitive tider} 9:85--96.}

\leavevmode\vadjust pre{\hypertarget{ref-uxe5stveit2018}{}}%
\CSLBlock{Åstveit, Leif Inge}
\CSLLeftMargin{ 2018}
\CSLRightInline{{The Early Mesolithic of Western Norway}. In \emph{{Early Economy and Settlement in Northern Europe. Pioneering, Resource Use, Coping with Change}}, edited by Hans Peter Blankholm, pp. 231--274. Equinox, Sheffield.}

\leavevmode\vadjust pre{\hypertarget{ref-bakka1971}{}}%
\CSLBlock{Bakka, Egil, and Peter Emil Kaland}
\CSLLeftMargin{ 1971}
\CSLRightInline{Early farming in Hordaland, western Norway. Problems and approaches in archaeology and pollen analysis. \emph{Norwegian Archaeological Review} 4:1--17. DOI:\href{https://doi.org/10.1080/00293652.1971.9965136}{10.1080/00293652.1971.9965136}.}

\leavevmode\vadjust pre{\hypertarget{ref-bang-andersen2012}{}}%
\CSLBlock{Bang-Andersen, Sveinung}
\CSLLeftMargin{ 2012}
\CSLRightInline{Colonizing Contrasting Landscapes. The Pioneer Coast Settlement and Inland Utilization in Southern Norway 10,000-9500 Years Before Present. \emph{Oxford Journal of Archaeology} 31:103--120. DOI:\href{https://doi.org/10.1111/j.1468-0092.2012.00381.x}{10.1111/j.1468-0092.2012.00381.x}.}

\leavevmode\vadjust pre{\hypertarget{ref-berg-hansen2009}{}}%
\CSLBlock{Berg-Hansen, Inger Marie}
\CSLLeftMargin{ 2009}
\CSLRightInline{\emph{{Steinalderregistrering. Metodologi og forskningshistorie i Norge 1900-2000 med en feltstudie fra Lista i Vest-Agder}}. Museum of Cultural History, University of Oslo, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-berg-hansen2018}{}}%
\CSLLeftMargin{ 2018 }
\CSLRightInline{{Continuity and Change in Late- and Post-glacial Social Networks: Knowledge Transmission and Blade Production Methods in Ahrensburgian and Early Mesolithic North West Europe}. In \emph{{The Early Settlement of Northern Europe. Transmission of Knowledge and Culture}}, edited by Kjel Knutsson, Helena Knutsson, Jan Apel, and Håkon Glørstad, pp. 63--98. Equinox, Sheffield.}

\leavevmode\vadjust pre{\hypertarget{ref-bergsvik2009}{}}%
\CSLBlock{Bergsvik, Knut Andreas}
\CSLLeftMargin{ 2009}
\CSLRightInline{{Caught in the middle: functional and ideological aspects of Mesolithic shores in Norway}. In \emph{{Mesolithic Horizons: Papers presented at the Seventh International Conference on the Mesolithic in Europe, Belfast 2005}}, edited by Sinéad B. McCartan, Rick Schulting, Graeme Warren, and Peter Woodman, pp. 602--609. Oxbow Books, Oxford.}

\leavevmode\vadjust pre{\hypertarget{ref-bevan2013a}{}}%
\CSLBlock{Bevan, Andrew, Enrico R. Crema, Xiuzhen Li, and Alessio Palmisano}
\CSLLeftMargin{ 2013}
\CSLRightInline{{Intensities, Interactions, and Uncertainties: Some New Approaches to Archaeological Distributions}. In \emph{{Computational Approaches to Archaeological Spaces}}, edited by Andrew Bevan and Mark Lake, pp. 27--52. Left Coast Press, Walnut Creek.}

\leavevmode\vadjust pre{\hypertarget{ref-bivand2021}{}}%
\CSLBlock{Bivand, Roger}
\CSLLeftMargin{ 2021}
\CSLRightInline{\emph{{rgrass7: Interface Between GRASS 7 Geographical Information System and R. R package version 0.2-6.}}}

\leavevmode\vadjust pre{\hypertarget{ref-bjerck1990}{}}%
\CSLBlock{Bjerck, Hein Bjartmann}
\CSLLeftMargin{ 1990}
\CSLRightInline{{Mesolithic site types and settlement patterns at Vega, Northern Norway}. \emph{Acta Archaeologica} 60:1--32.}

\leavevmode\vadjust pre{\hypertarget{ref-bjerck2005}{}}%
\CSLLeftMargin{ 2005 }
\CSLRightInline{Strandlinjedatering. In \emph{Norsk arkeologisk leksikon}, edited by Einar Østmo and Lotte Hedeager, pp. 363--364. Pax, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-bjerck2008}{}}%
\CSLLeftMargin{ 2008a }
\CSLRightInline{{Norwegian Mesolithic Trends: A Review}. In \emph{{Mesolithic Europe}}, edited by Geoff Bailey and Penny Spikins, pp. 60--106. Cambridge University Press, Cambridge.}

\leavevmode\vadjust pre{\hypertarget{ref-bjerck2008a}{}}%
\CSLLeftMargin{ 2008b }
\CSLRightInline{Innledende betraktninger. In \emph{{NTNU Vitenskapsmuseets arkeologiske undersøkelser Ormen Lange Nyhamna}}, edited by Hein Bjartmann Bjerck, Leif Inge Åstveit, Trond Meling, Jostein Gundersen, Guro Jørgensen, and Staale Normann, pp. 548--551. Tapir Akademisk Forlag, Trondheim.}

\leavevmode\vadjust pre{\hypertarget{ref-blankholm2020}{}}%
\CSLBlock{Blankholm, Hans Peter}
\CSLLeftMargin{ 2020}
\CSLRightInline{{In the wake of the wake. An investigation of the impact of the Storegga tsunami on the human settlement of inner Varangerfjord, northern Norway}. \emph{Quaternary International} 549:65--73. DOI:\url{https://doi.org/10.1016/j.quaint.2018.05.050}.}

\leavevmode\vadjust pre{\hypertarget{ref-breivik2014}{}}%
\CSLBlock{Breivik, Heidi Mjelva}
\CSLLeftMargin{ 2014}
\CSLRightInline{Palaeo-oceanographic development and human adaptive strategies in the Pleistocene--Holocene transition: A study from the Norwegian coast. \emph{The Holocene} 24:1478--1490. DOI:\href{https://doi.org/10.1177/0959683614544061}{10.1177/0959683614544061}.}

\leavevmode\vadjust pre{\hypertarget{ref-breivik2018}{}}%
\CSLBlock{Breivik, Heidi Mjelva, Guro Fossum, and Steinar Solheim}
\CSLLeftMargin{ 2018}
\CSLRightInline{Exploring human responses to climatic fluctuations and environmental diversity: Two stories from Mesolithic Norway. \emph{Quaternary International} 465. Impacts of gradual and abrupt environmental changes on Late glacial to Middle Holocene cultural changes in Europe:258--275. DOI:\href{https://doi.org/10.1016/j.quaint.2016.12.019}{10.1016/j.quaint.2016.12.019}.}

\leavevmode\vadjust pre{\hypertarget{ref-breivik2018a}{}}%
\CSLBlock{Breivik, Heidi, and Hein Bjartmann Bjerck}
\CSLLeftMargin{ 2018}
\CSLRightInline{{Early Mesolithic Central Norway: A Review of Research History, Settlements, and Tool Tradition}. In \emph{{Early Economy and Settlement in Northern Europe. Pioneering, Resource Use, Coping with Change}}, edited by Hans Peter Blankholm, pp. 169--206. Equinox, Sheffield.}

\leavevmode\vadjust pre{\hypertarget{ref-brogger1905}{}}%
\CSLBlock{Brøgger, Waldemar Christofer}
\CSLLeftMargin{ 1905}
\CSLRightInline{\emph{{Strandliniens Beliggenhed under Stenalderen i Det Sydøstlige Norge}}. Norges geologiske undersøkelse, Kristiania.}

\leavevmode\vadjust pre{\hypertarget{ref-bronkramsey2009}{}}%
\CSLBlock{Bronk Ramsey, Christopher}
\CSLLeftMargin{ 2009}
\CSLRightInline{Bayesian Analysis of Radiocarbon Dates. \emph{Radiocarbon} 51(1):337--360. DOI:\href{https://doi.org/10.1017/S0033822200033865}{10.1017/S0033822200033865}.}

\leavevmode\vadjust pre{\hypertarget{ref-bronkramsey2015}{}}%
\CSLLeftMargin{ 2015 }
\CSLRightInline{{Bayesian Approaches to the Building of Archaeological Chronologies}. In \emph{{Mathematics and Archaeology}}, edited by Juan A. Barcelo and Igor Bogdanovic, pp. 272--292. CRC Press, Boca Raton.}

\leavevmode\vadjust pre{\hypertarget{ref-conolly2020}{}}%
\CSLBlock{Conolly, James}
\CSLLeftMargin{ 2020}
\CSLRightInline{Spatial interpolation. In \emph{{Archaeological Spatial Analysis: A Methodological Guide}}, edited by Mark Gillings, Piraye Hacıgüzeller, and Gary Lock, pp. 118--134. Routledge, London \& New York.}

\leavevmode\vadjust pre{\hypertarget{ref-conolly2006}{}}%
\CSLBlock{Conolly, James, and Mark Lake}
\CSLLeftMargin{ 2006}
\CSLRightInline{\emph{{Geographical Information Systems in Archaeology}}. Cambridge University Press, Cambridge.}

\leavevmode\vadjust pre{\hypertarget{ref-creel2022}{}}%
\CSLBlock{Creel, Roger C., Jacqueline Austermann, Nicole S. Khan, William J. D'Andrea, Nicholas Balascio, Blake Dyer, Erica Ashe, and William Menke}
\CSLLeftMargin{ 2022}
\CSLRightInline{Postglacial relative sea level change in Norway. \emph{Quaternary Science Reviews} 282:107422. DOI:\href{https://doi.org/10.1016/j.quascirev.2022.107422}{10.1016/j.quascirev.2022.107422}.}

\leavevmode\vadjust pre{\hypertarget{ref-crema2012}{}}%
\CSLBlock{Crema, Enrico R.}
\CSLLeftMargin{ 2012}
\CSLRightInline{Modelling Temporal Uncertainty in Archaeological Analysis. \emph{Journal of Archaeological Method and Theory} 19(3):440--461. DOI:\href{https://doi.org/10.1007/s10816-011-9122-3}{10.1007/s10816-011-9122-3}.}

\leavevmode\vadjust pre{\hypertarget{ref-crema2015}{}}%
\CSLLeftMargin{ 2015 }
\CSLRightInline{{Time and Probabilistic Reasoning in Settlement Analysis}. In \emph{{Mathematics and Archaeology}}, edited by Juan A. Barcelo and Igor Bogdanovic, pp. 314--334. CRC Press, Boca Raton.}

\leavevmode\vadjust pre{\hypertarget{ref-crema2010}{}}%
\CSLBlock{Crema, Enrico R., Andrew Bevan, and Mark W. Lake}
\CSLLeftMargin{ 2010}
\CSLRightInline{A probabilistic framework for assessing spatio-temporal point patterns in the archaeological record. \emph{Journal of Archaeological Science} 37(5):1118--1130. DOI:\href{https://doi.org/10.1016/j.jas.2009.12.012}{10.1016/j.jas.2009.12.012}.}

\leavevmode\vadjust pre{\hypertarget{ref-damlien2018b}{}}%
\CSLBlock{Damlien, Hege, and Steinar Solheim}
\CSLLeftMargin{ 2018}
\CSLRightInline{{The Pioneer Settlement of Eastern Norway}. In \emph{{Early Economy and Settlement in Northern Europe. Pioneering, Resource Use, Coping with Change}}, edited by Hans Peter Blankholm, pp. 335--367. Equinox, Sheffield.}

\leavevmode\vadjust pre{\hypertarget{ref-degeer1896}{}}%
\CSLBlock{De Geer, Gerard}
\CSLLeftMargin{ 1896}
\CSLRightInline{\emph{{Om Skandinaviens geografiska utveckling efter Istiden}}. P. A. Norstedt \& Söner, Stockholm.}

\leavevmode\vadjust pre{\hypertarget{ref-eskeland2017}{}}%
\CSLBlock{Eskeland, Knut Fossdal}
\CSLLeftMargin{ 2017}
\CSLRightInline{\emph{{Rapport, arkeologisk registrering. E18 Langangen Rugtvedt, 16/06999, Porsgrunn og Bamble kommune}}. Skien.}

\leavevmode\vadjust pre{\hypertarget{ref-fossum2020}{}}%
\CSLBlock{Fossum, Guro}
\CSLLeftMargin{ 2020}
\CSLRightInline{{Specialists facing climate change. The 8200 cal BP event and its impact on the coastal settlement in the inner Oslo fjord, southeast Norway}. In \emph{{Coastal Landscapes of the Mesolithic: Human Engagement with the Coast from the Atlantic to the Baltic Sea}}, edited by Almut Schülke, pp. 179--201. Routledge, London \& New York.}

\leavevmode\vadjust pre{\hypertarget{ref-fuglestvedt2012}{}}%
\CSLBlock{Fuglestvedt, Ingrid}
\CSLLeftMargin{ 2012}
\CSLRightInline{{The Pioneer Condition on the Scandinavian Peninsula: the Last Frontier of a {`}Palaeolithic Way{'} in Europe}. \emph{Norwegian Archaeological Review} 45(1):1--29. DOI:\href{https://doi.org/10.1080/00293652.2012.669998}{10.1080/00293652.2012.669998}.}

\leavevmode\vadjust pre{\hypertarget{ref-gjerpe2008b}{}}%
\CSLBlock{Gjerpe, Lars Erik, and Grethe Bjørkan Bukkemoen}
\CSLLeftMargin{ 2008}
\CSLRightInline{{Nordby 1 -- Toskipede hus fra neolitikum-bronsealder og boplasspor fra jernalder}. In \emph{{E18-prosjektet Vestfold. Bind 3. Hus, boplass- og dyrkningsspor}}, edited by Lars Erik Gjerpe, pp. 7--38. University of Oslo, Museum of Cultural History, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-glorstad2010}{}}%
\CSLBlock{Glørstad, Håkon}
\CSLLeftMargin{ 2010}
\CSLRightInline{\emph{{The Structure and History of the Late Mesolithic Societies in the Oslo Fjord Area 6300-3800 BC}}. Bricoleur Press, Lindome.}

\leavevmode\vadjust pre{\hypertarget{ref-glorstad2012}{}}%
\CSLLeftMargin{ 2012 }
\CSLRightInline{{Historical ideal types and the transition to the Late Neolithic in South Norway}. In \emph{{Becoming European. The transformation of third millenium Northern and Western Europe}}, edited by Christopher Prescott and Håkon Glørstad, pp. 82--99. Oxbow Books, Oxford \& Oakville.}

\leavevmode\vadjust pre{\hypertarget{ref-glorstad2016}{}}%
\CSLLeftMargin{ 2016 }
\CSLRightInline{Deglaciation, sea-level change and the Holocene colonization of Norway. \emph{Geological Society, London, Special Publications} 411:9--25. DOI:\href{https://doi.org/10.1144/SP411.7}{10.1144/SP411.7}.}

\leavevmode\vadjust pre{\hypertarget{ref-glorstad2020}{}}%
\CSLBlock{Glørstad, Håkon, Jostein Gundersen, Frode Kvalø, Pål Nymoen, David Simpson, and Birgitte Skar}
\CSLLeftMargin{ 2020}
\CSLRightInline{{Submerged Stone Age from a Norwegian Perspective}. In \emph{{The Archaeology of Europe's Drowned Landscapes}}, edited by Geoff Bailey, Nena Galanidou, Hans Peeters, Hauke Jöns, and Moritz Mennenga, pp. 125-\/-140. Springer, Cham.}

\leavevmode\vadjust pre{\hypertarget{ref-grass2017}{}}%
\CSLBlock{GRASS Development Team}
\CSLLeftMargin{ 2017}
\CSLRightInline{\emph{{Geographic Resources Analysis Support System (GRASS) Software, Version 7.2}}. Open Source Geospatial Foundation.}

\leavevmode\vadjust pre{\hypertarget{ref-gundersen2013}{}}%
\CSLBlock{Gundersen, Jostein}
\CSLLeftMargin{ 2013}
\CSLRightInline{{Verken fjord eller fjell -- steinalderen i det kystnære innlandet. Gamle og nye funn fra Notodden i Telemark}. \emph{Viking} 76:35--62.}

\leavevmode\vadjust pre{\hypertarget{ref-hafsten1957}{}}%
\CSLBlock{Hafsten, Ulf}
\CSLLeftMargin{ 1957}
\CSLRightInline{{De senkvartære strandlinje-forskyvningene i Oslotrakten belyst ved pollenanalytiske undersøkelser}. \emph{Norwegian Journal of Geography} 16(1-8):74--99. DOI:\href{https://doi.org/10.1080/00291955708622137}{10.1080/00291955708622137}.}

\leavevmode\vadjust pre{\hypertarget{ref-hafsten1983}{}}%
\CSLLeftMargin{ 1983 }
\CSLRightInline{{Shore-level changes in South Norway during the last 13,000 years, traced by biostratigraphical methods and radiometric datings}. \emph{Norwegian Journal of Geography} 37(2):63--79. DOI:\href{https://doi.org/10.1080/00291958308552089}{10.1080/00291958308552089}.}

\leavevmode\vadjust pre{\hypertarget{ref-hagen1963}{}}%
\CSLBlock{Hagen, Anders}
\CSLLeftMargin{ 1963}
\CSLRightInline{{Problemkompleks Fosna. Opphav -- kontakt med kontinentale grupper -- forholdet til Komsa}. In \emph{{Boplatsproblem vid Kattegat och Skagerack}}, pp. 53--59. Göteborg och Bohusläns fornminnesförening \& Institutionen för nordisk fornkunskap, Gothenburg University, Gothenburg.}

\leavevmode\vadjust pre{\hypertarget{ref-helskog1978}{}}%
\CSLBlock{Helskog, Knut}
\CSLLeftMargin{ 1978}
\CSLRightInline{Late Holocene sea-level changes seen from prehistoric settlements. \emph{Norwegian Journal of Geography} 32:111--119. DOI:\href{https://doi.org/10.1080/00291957808552032}{10.1080/00291957808552032}.}

\leavevmode\vadjust pre{\hypertarget{ref-herzog2013}{}}%
\CSLBlock{Herzog, Irmela}
\CSLLeftMargin{ 2013}
\CSLRightInline{{The Potential and Limits of Optimal Path Analysis}. In \emph{{Computational Approaches to Archaeological Spaces}}, edited by Andrew Bevan and Mark Lake, pp. 179--211. Left Coast Press, Walnut Creek.}

\leavevmode\vadjust pre{\hypertarget{ref-hinsch1955}{}}%
\CSLBlock{Hinsch, Erik}
\CSLLeftMargin{ 1955}
\CSLRightInline{{Traktbegerkultur -- Megalitkultur. En studie av Øst-Norges eldste neolitiske gruppe}. \emph{Universitetets Oldsaksamling Årbok} 1951--1953:10--177.}

\leavevmode\vadjust pre{\hypertarget{ref-hinz2021}{}}%
\CSLBlock{Hinz, Martin, Clemens Schmid, Daniel Knitter, and Carolin Tietze}
\CSLLeftMargin{ 2021}
\CSLRightInline{\emph{oxcAAR: Interface to 'OxCal' radiocarbon calibration. R package version 1.1.0.}}

\leavevmode\vadjust pre{\hypertarget{ref-hollender1901}{}}%
\CSLBlock{Hollender, Artur}
\CSLLeftMargin{ 1901}
\CSLRightInline{Om sveriges nivåförändringar efter menniskans invandring. \emph{Geolgiska Föreningen i Stockholm Förhandlingar} 23(4):1118--1130. DOI:\href{https://doi.org/10.1080/00293652.1975.9965220}{10.1080/00293652.1975.9965220}.}

\leavevmode\vadjust pre{\hypertarget{ref-hughes2016}{}}%
\CSLBlock{Hughes, Anna L. C., Richard Gyllencreutz, Øystein S. Lohne, Jan Mangerud, and John Inge Svendsen}
\CSLLeftMargin{ 2016}
\CSLRightInline{The last Eurasian ice sheets {\textendash} a chronological database and time-slice reconstruction, DATED-1. \emph{Boreas} 45(1):1--45. DOI:\url{https://doi.org/10.1111/bor.12142}.}

\leavevmode\vadjust pre{\hypertarget{ref-hyndman1996}{}}%
\CSLBlock{Hyndman, Rob J}
\CSLLeftMargin{ 1996}
\CSLRightInline{Computing and Graphing Highest Density Regions. \emph{The American Statistician} 50(2):120--126.}

\leavevmode\vadjust pre{\hypertarget{ref-ilves2011}{}}%
\CSLBlock{Ilves, Kristin, and Kim Darmark}
\CSLLeftMargin{ 2011}
\CSLRightInline{Some Critical and Methodological Aspects of Shoreline Determination: Examples from the Baltic Sea Region. \emph{Journal of Archaeological Method and Theory} 18:147--165. DOI:\href{https://doi.org/10.1007/s10816-010-9084-x}{10.1007/s10816-010-9084-x}.}

\leavevmode\vadjust pre{\hypertarget{ref-jaksland2012}{}}%
\CSLBlock{Jaksland, Lasse (editor)}
\CSLLeftMargin{ 2012a}
\CSLRightInline{\emph{{E18 Brunlanesprosjektet. Bind II. Undersøkte lokaliteter fra tidligmesolitikum}}. University of Oslo, Museum of Cultural History, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-jaksland2012a}{}}%
\CSLBlock{ (editor)}
\CSLLeftMargin{ 2012b}
\CSLRightInline{\emph{{E18 Brunlanesprosjektet. Bind III. Undersøkte lokaliteter fra tidligmesolitikum og senere}}. University of Oslo, Museum of Cultural History, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-jaksland2014}{}}%
\CSLLeftMargin{ 2014 }
\CSLRightInline{Kulturhistorisk sammenstilling. In \emph{E18 brunlanesprosjektet. Bind i. Forutsetninger og kulturhistorisk sammenstilling}, edited by Lasse Jaksland and Per Persson, pp. 11--46. University of Oslo, Museum of Cultural History, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-jaksland2014a}{}}%
\CSLBlock{Jaksland, Lasse, and Per Persson (editors)}
\CSLLeftMargin{ 2014}
\CSLRightInline{\emph{{E18 Brunlanesprosjektet. Bind I. Forutsetninger og kulturhistorisk sammenstilling}}. University of Oslo, Museum of Cultural History, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-johansen1963}{}}%
\CSLBlock{Johansen, Erling}
\CSLLeftMargin{ 1963}
\CSLRightInline{{Kyst(fangst)boplassenes strandbundenhet og strandlinjekronologien}. In \emph{{Boplatsproblem vid Kattegat och Skagerack}}, pp. 90--92. Göteborg och Bohusläns fornminnesförening \& Institutionen för nordisk fornkunskap, Gothenburg University, Gothenburg.}

\leavevmode\vadjust pre{\hypertarget{ref-jorgensen2020}{}}%
\CSLBlock{Jørgensen, Erlend Kirkeng, Petro Pesonen, and Miikka Tallavaara}
\CSLLeftMargin{ 2020}
\CSLRightInline{{Climatic changes cause synchronous population dynamics and adaptive strategies among coastal hunter-gatherers in Holocene northern Europe}. \emph{Quaternary Research}:1--16. DOI:\href{https://doi.org/10.1017/qua.2019.86}{10.1017/qua.2019.86}.}

\leavevmode\vadjust pre{\hypertarget{ref-kjemperud1986}{}}%
\CSLBlock{Kjemperud, Alfred}
\CSLLeftMargin{ 1986}
\CSLRightInline{{Late Weichselian and Holocene shoreline displacement in the Trondheimsfjord area, central Norway}. \emph{Boreas} 15(1):61--82. DOI:\href{https://doi.org/10.1111/j.1502-3885.1986.tb00744.x}{10.1111/j.1502-3885.1986.tb00744.x}.}

\leavevmode\vadjust pre{\hypertarget{ref-kleppe1985}{}}%
\CSLBlock{Kleppe, Else Johansen}
\CSLLeftMargin{ 1985}
\CSLRightInline{\emph{{Archaeological Data on Shore Displacements in Norway}}. Norges geografiske oppmåling, Hønefoss.}

\leavevmode\vadjust pre{\hypertarget{ref-kleppe2018}{}}%
\CSLBlock{Kleppe, Jan Ingolf}
\CSLLeftMargin{ 2018}
\CSLRightInline{{The Pioneer Colonization of Northern Norway}. In \emph{{Early Economy and Settlement in Northern Europe. Pioneering, Resource Use, Coping with Change}}, edited by Hans Peter Blankholm, pp. 13--57. Equinox, Sheffield.}

\leavevmode\vadjust pre{\hypertarget{ref-lakens2018}{}}%
\CSLBlock{Lakens, Daniël, Anne M. Scheel, and Peder M. Isager}
\CSLLeftMargin{ 2018}
\CSLRightInline{Equivalence Testing for Psychological Research: A Tutorial. \emph{Advances in Methods and Practices in Psychological Science} 1(2):259--269. DOI:\href{https://doi.org/10.1177/2515245918770963}{10.1177/2515245918770963}.}

\leavevmode\vadjust pre{\hypertarget{ref-lewis2021}{}}%
\CSLBlock{Lewis, Joseph}
\CSLLeftMargin{ 2021}
\CSLRightInline{Probabilistic Modelling for Incorporating Uncertainty in Least Cost Path Results: a Postdictive Roman Road Case Study. \emph{Journal of Archaeological Method and Theory} 28(3):911--924. DOI:\href{https://doi.org/10.1007/s10816-021-09522-w}{10.1007/s10816-021-09522-w}.}

\leavevmode\vadjust pre{\hypertarget{ref-marwick2018}{}}%
\CSLBlock{Marwick, Ben, Carl Boettiger, and Lincoln Mullen}
\CSLLeftMargin{ 2018}
\CSLRightInline{{Packaging Data Analytical Work Reproducibly Using R (and Friends)}. \emph{The American Statistician} 72(1):80--88. DOI:\href{https://doi.org/10.1080/00031305.2017.1375986}{10.1080/00031305.2017.1375986}.}

\leavevmode\vadjust pre{\hypertarget{ref-melvold2014b}{}}%
\CSLBlock{Melvold, Stine, and Per Persson (editors)}
\CSLLeftMargin{ 2014}
\CSLRightInline{\emph{{Vestfoldbaneprosjektet. Arkeologiske undersøkelser i forbindelse med ny jernbane mellom Larvik og Porsgrunn. Bind 1. Tidlig- Og mellommesolittiske lokaliteter i Vestfold og Telemark}}. Portal forlag, Kristiansand.}

\leavevmode\vadjust pre{\hypertarget{ref-mikkelsen1975}{}}%
\CSLBlock{Mikkelsen, Egil}
\CSLLeftMargin{ 1975}
\CSLRightInline{{Mesolithic in South-Eastern Norway}. \emph{Norwegian Archaeological Review} 8(1):1118--1130. DOI:\href{https://doi.org/10.1080/11035890109445866}{10.1080/11035890109445866}.}

\leavevmode\vadjust pre{\hypertarget{ref-milne2015}{}}%
\CSLBlock{Milne, Glenn A}
\CSLLeftMargin{ 2015}
\CSLRightInline{Glacial isostatic adjustment. In \emph{Handbook of sea-level research}, edited by Ian Shennan, Antony J Long, and Benjamin P Horton, pp. 421--437. Wiley, Chichester.}

\leavevmode\vadjust pre{\hypertarget{ref-milne2009}{}}%
\CSLBlock{Milne, Glenn A., W. Roland Gehrels, Chris W. Hughes, and Mark E. Tamisiea}
\CSLLeftMargin{ 2009}
\CSLRightInline{Identifying the causes of sea-level change. \emph{Nature Geoscience} 2(7):471--478. DOI:\href{https://doi.org/10.1038/ngeo544}{10.1038/ngeo544}.}

\leavevmode\vadjust pre{\hypertarget{ref-mjuxe6rum2018}{}}%
\CSLBlock{Mjærum, Axel}
\CSLLeftMargin{ 2018}
\CSLRightInline{Hinterland discoveries: Middle Mesolithic woodland utilization and the case of the site Eidsberg, eastern Norway. \emph{Current Swedish Archaeology} 26(1):159--188. DOI:\href{https://doi.org/10.37718/CSA.2018.11}{10.37718/CSA.2018.11}.}

\leavevmode\vadjust pre{\hypertarget{ref-mjuxe6rum2022}{}}%
\CSLLeftMargin{ 2022 }
\CSLRightInline{{A Matter of Scale: Responses to Landscape Changes in the Oslo Fjord, Norway, in the Mesolithic}. \emph{Open Archaeology} 8(1):62--84. DOI:\href{https://doi.org/10.1515/opar-2022-0225}{10.1515/opar-2022-0225}.}

\leavevmode\vadjust pre{\hypertarget{ref-muxf8ller1987}{}}%
\CSLBlock{Møller, Jakob J}
\CSLLeftMargin{ 1987}
\CSLRightInline{Shoreline relation and prehistoric settlement in northern norway. \emph{Norwegian Journal of Geography} 41:45--60. DOI:\url{http://dx.doi.org/10.1080/00291958708552171}.}

\leavevmode\vadjust pre{\hypertarget{ref-muxf6rner1976}{}}%
\CSLBlock{Mörner, Nils-Axel}
\CSLLeftMargin{ 1976}
\CSLRightInline{{Eustasy and Geoid Changes}. \emph{The Journal of Geology} 84(2):123--151. DOI:\href{https://doi.org/10.1086/628184}{10.1086/628184}.}

\leavevmode\vadjust pre{\hypertarget{ref-muxf6rner1979}{}}%
\CSLLeftMargin{ 1979 }
\CSLRightInline{{The Fennoscandian Uplift and Late Cenozoic Geodynamics: Geological Evidence}. \emph{GeoJournal} 3(3):287--318. DOI:\href{https://doi.org/10.1007/BF00177634}{10.1007/BF00177634}.}

\leavevmode\vadjust pre{\hypertarget{ref-nielsen2021b}{}}%
\CSLBlock{Nielsen, Svein Vatsvåg}
\CSLLeftMargin{ 2021}
\CSLRightInline{Early farming in Southeastern Norway: New evidence and interpretations. \emph{Journal of Neolithic Archaeology} 23:83--113. DOI:\href{https://doi.org/10.12766/jna.2021.4}{10.12766/jna.2021.4}.}

\leavevmode\vadjust pre{\hypertarget{ref-nielsen2019}{}}%
\CSLBlock{Nielsen, Svein Vatsvåg, Per Persson, and Steinar Solheim}
\CSLLeftMargin{ 2019}
\CSLRightInline{De-Neolithisation in southern Norway inferred from statistical modelling of radiocarbon dates. \emph{Journal of Anthropological Archaeology} 53:82--91. DOI:\href{https://doi.org/10.1016/j.jaa.2018.11.004}{10.1016/j.jaa.2018.11.004}.}

\leavevmode\vadjust pre{\hypertarget{ref-nordqvist1995}{}}%
\CSLBlock{Nordqvist, Bengt}
\CSLLeftMargin{ 1995}
\CSLRightInline{{The Mesolithic settlement of the west coast of Sweden - with special emphasis on chronology and topography of coastal settlements}. In \emph{{Man and Sea in the Mesolithic. Coastal settlement above and below present sea level}}, edited by Anders Fischer, pp. 185--196. Oxbow Books, Oxford.}

\leavevmode\vadjust pre{\hypertarget{ref-nordqvist1999}{}}%
\CSLLeftMargin{ 1999 }
\CSLRightInline{{The Chronology of the Western Swedish Mesolithic and Late Paleolithic: Old Answers in Spite of New Methods}. In \emph{{The Mesolithic of Central Scandinavia}}, edited by Joel Boaz, pp. 235--253. University of Oslo, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-mapping2018}{}}%
\CSLBlock{Norwegian Mapping Authority}
\CSLLeftMargin{ 2018}
\CSLRightInline{\emph{Produktspesifikasjon. Nasjonal modell for høydedata fra laserskanning (FKB-laser). FKB-laser\_v30}.}

\leavevmode\vadjust pre{\hypertarget{ref-mapping2021}{}}%
\CSLLeftMargin{ 2021 }
\CSLRightInline{\emph{{Tidevannstabeller for den norske kyst med Svalbard samt Dover, England}}.}

\leavevmode\vadjust pre{\hypertarget{ref-nummedal1923}{}}%
\CSLBlock{Nummedal, Anders}
\CSLLeftMargin{ 1923}
\CSLRightInline{Om flintpladsene. \emph{Norwegian Journal of Geography} 7(2):89--141.}

\leavevmode\vadjust pre{\hypertarget{ref-ostmo1988}{}}%
\CSLBlock{Østmo, Einar}
\CSLLeftMargin{ 1988}
\CSLRightInline{\emph{{Etableringen av jordbrukskultur i Østfold i steinalderen}}. The University Collection of National Antiquities, University of Oslo, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-ostmo2008}{}}%
\CSLLeftMargin{ 2008 }
\CSLRightInline{\emph{{Auve. En fangstboplass fra yngre steinalder på Vesterøya i Sandefjord. I. Den arkeologiske del}}. Museum of Cultural History, University of Oslo, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-parnell2008}{}}%
\CSLBlock{Parnell, A. C., J. Haslett, J. R. M. Allen, C. E. Buck, and B. Huntley}
\CSLLeftMargin{ 2008}
\CSLRightInline{A flexible approach to assessing synchroneity of past events using Bayesian reconstructions of sedimentation history. \emph{Quaternary Science Reviews} 27(19-20):1872--1885. DOI:\href{https://doi.org/10.1016/j.quascirev.2008.07.009}{10.1016/j.quascirev.2008.07.009}.}

\leavevmode\vadjust pre{\hypertarget{ref-persson2008}{}}%
\CSLBlock{Persson, Per}
\CSLLeftMargin{ 2008}
\CSLRightInline{{Nauen 5.2 -- Stenåldersboplatser och fossil åkermark}. In \emph{{E18-prosjektet Vestfold. Bind 2. Steinalderboplasser, boplasspor, graver og dyrkningsspor}}, edited by Lars Erik Gjerpe, pp. 163--198. University of Oslo, Museum of Cultural History, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-prescott2020}{}}%
\CSLBlock{Prescott, Christopher}
\CSLLeftMargin{ 2020}
\CSLRightInline{{Interpreting Complex Diachronic {"}Neolithic{"}-Period Data in Norway}. In \emph{{Farmers at the Frontier -- A Pan European Perspective on Neolithisation}}, edited by Kurt J. Gron, Lasse Sørensen, and Peter Rowley-Conwy, pp. 381--400. Oxbow Books, Oxford.}

\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2021}{}}%
\CSLBlock{R Core Team}
\CSLLeftMargin{ 2021}
\CSLRightInline{\emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing, Vienna, Austria.}

\leavevmode\vadjust pre{\hypertarget{ref-ramstad2009}{}}%
\CSLBlock{Ramstad, Morten}
\CSLLeftMargin{ 2009}
\CSLRightInline{{Eldre steinalder på Melkøya, representativitet, strandlinjer og transgresjon}. In \emph{{Undersøkelsene på Melkøya. Melkøyaprosjektet -- Kulturhistoriske registreringer og utgravninger 2001 og 2002}}, edited by Anders Hesjedal, Morten Ramstad, and Anja R. Niemi, pp. 491--495. Tromsø museum, Universitetesmuseet, Tromsø.}

\leavevmode\vadjust pre{\hypertarget{ref-reimer2020}{}}%
\CSLBlock{Reimer, Paula J., William E. N. Austin, Edouard Bard, Alex Bayliss, Paul G. Blackwell, Christopher Bronk Ramsey, Martin Butzin, Hai Cheng, R. Lawrence Edwards, Michael Friedrich, Pieter M. Grootes, Thomas P. Guilderson, Irka Hajdas, Timothy J. Heaton, Alan G. Hogg, Konrad A. Hughen, Bernd Kromer, Sturt W. Manning, Raimund Muscheler, Jonathan G. Palmer, Charlotte Pearson, Johannes van der Plicht, Ron W. Reimer, David A. Richards, E. Marian Scott, John R. Southon, Christian S. M. Turney, Lukas Wacker, Florian Adolphi, Ulf Büntgen, Manuela Capano, Simon M. Fahrni, Alexandra Fogtmann-Schulz, Ronny Friedrich, Peter Köhler, Sabrina Kudsk, Fusa Miyake, Jesper Olsen, Frederick Reinig, Minoru Sakamoto, Adam Sookdeo, and Sahra Talamo}
\CSLLeftMargin{ 2020}
\CSLRightInline{The IntCal20 Northern Hemisphere Radiocarbon Age Calibration Curve (0{\textendash}55 cal kBP). \emph{Radiocarbon} 62(4):725--757. DOI:\href{https://doi.org/10.1017/RDC.2020.41}{10.1017/RDC.2020.41}.}

\leavevmode\vadjust pre{\hypertarget{ref-reitan2009}{}}%
\CSLBlock{Reitan, Gaute, and Inger Marie Berg-Hansen}
\CSLLeftMargin{ 2009}
\CSLRightInline{\emph{{Lundevågenprosjektet, delrapport 1. Sammenfattende rapport. Lunde, 6/1, 6/35 og Skjolnes 7/23, 7/27, Farsund kommune, Vest-Agder}}. Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-reitan2014l}{}}%
\CSLBlock{Reitan, Gaute, and Per Persson (editors)}
\CSLLeftMargin{ 2014}
\CSLRightInline{\emph{{Vestfoldbaneprosjektet. Arkeologiske undersøkelser i forbindelse med ny jernbane mellom Larvik og Porsgrunn. Bind 2. Seinmesolittiske, neolittiske og yngre lokaliteter i Vestfold og Telemark}}. Portal forlag, Kristiansand.}

\leavevmode\vadjust pre{\hypertarget{ref-reitan2018c}{}}%
\CSLBlock{Reitan, Gaute, and Lars Sundström (editors)}
\CSLLeftMargin{ 2018}
\CSLRightInline{\emph{{The Stone Age Coastal Settlement in Aust-Agder, Southeast Norway}}. Cappelen Damm Akademisk, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-roalkvam2020}{}}%
\CSLBlock{Roalkvam, Isak}
\CSLLeftMargin{ 2020}
\CSLRightInline{{Algorithmic classification and statistical modelling of coastal settlement patterns in Mesolithic south-eastern Norway}. \emph{Journal of Computer Applications in Archaeology} 3(1):288--307. DOI:\url{https://doi.org/10.5334/jcaa.60}.}

\leavevmode\vadjust pre{\hypertarget{ref-roalkvam2022}{}}%
\CSLLeftMargin{ 2022 }
\CSLRightInline{{Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway}. \emph{Journal of Archaeological Science: Reports} 42:103371. DOI:\href{https://doi.org/10.1016/j.jasrep.2022.103371}{10.1016/j.jasrep.2022.103371}.}

\leavevmode\vadjust pre{\hypertarget{ref-roberg2012}{}}%
\CSLBlock{Røberg, Frank Halvar N.}
\CSLLeftMargin{ 2012}
\CSLRightInline{\emph{{Bosetnings- og aktivitetsspor. Larønningen, 221/2138. Skien, Telemark}}. University of Oslo, Museum of Cultural History, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-romundset2018b}{}}%
\CSLBlock{Romundset, Anders}
\CSLLeftMargin{ 2018}
\CSLRightInline{{Postglacial shoreline displacement in the Tvedestrand-Arendal area}. In \emph{{The Stone Age Coastal Settlement in Aust-Agder, Southeast Norway}}, edited by Gaute Reitan and Lars Sundström, pp. 463--478. Cappelen Damm Akademisk, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-romundset2011}{}}%
\CSLBlock{Romundset, Anders, Stein Bondevik, and Ole Bennike}
\CSLLeftMargin{ 2011}
\CSLRightInline{Postglacial uplift and relative sea level changes in Finnmark, northern Norway. \emph{Quaternary Science Reviews} 30(19-20):2398--2421. DOI:\href{https://doi.org/10.1016/j.quascirev.2011.06.007}{10.1016/j.quascirev.2011.06.007}.}

\leavevmode\vadjust pre{\hypertarget{ref-romundset2015}{}}%
\CSLBlock{Romundset, Anders, Ola Fredin, and Fredrik Høgaas}
\CSLLeftMargin{ 2015}
\CSLRightInline{A Holocene sea{-}level curve and revised isobase map based on isolation basins from near the southern tip of Norway. \emph{Boreas} 44:383--400. DOI:\href{https://doi.org/10.1111/bor.12105}{10.1111/bor.12105}.}

\leavevmode\vadjust pre{\hypertarget{ref-romundset2018}{}}%
\CSLBlock{Romundset, Anders, Thomas R. Lakeman, and Fredrik Høgaas}
\CSLLeftMargin{ 2018}
\CSLRightInline{Quantifying variable rates of postglacial relative sea level fall from a cluster of 24 isolation basins in southern Norway. \emph{Quaternary Science Reviews} 197:175--192. DOI:\href{https://doi.org/10.1016/j.quascirev.2018.07.041}{10.1016/j.quascirev.2018.07.041}.}

\leavevmode\vadjust pre{\hypertarget{ref-romundset2019}{}}%
\CSLLeftMargin{ 2019 }
\CSLRightInline{Coastal lake records add constraints to the age and magnitude of the Younger Dryas ice-front oscillation along the Skagerrak coastline in southern Norway. \emph{Journal of Quaternary Science} 34(2):112--124. DOI:\url{https://doi.org/10.1002/jqs.3085}.}

\leavevmode\vadjust pre{\hypertarget{ref-schmitt2009}{}}%
\CSLBlock{Schmitt, Lou, Stephan Larsson, Jan Burdukiewicz, John Ziker, Krister Svedhage, Jeanette Zamon, and Steffen Holger}
\CSLLeftMargin{ 2009}
\CSLRightInline{Chronological Insights, Cultural Change, and Resource Exploitation on the West Coast of Sweden During the Late Palaeolithic/Early Mesolithic Transition. \emph{Oxford Journal of Archaeology} 28:1--27. DOI:\href{https://doi.org/10.1111/j.1468-0092.2008.00317.x}{10.1111/j.1468-0092.2008.00317.x}.}

\leavevmode\vadjust pre{\hypertarget{ref-schulke2020}{}}%
\CSLBlock{Schülke, Almut}
\CSLLeftMargin{ 2020}
\CSLRightInline{{First visit or revisit? Motivations of mobility and the use and reuse of sites in the changing coastal areas of Mesolithic southeastern Norway}. In \emph{{Coastal Landscapes of the Mesolithic: Human Engagement with the Coast from the Atlantic to the Baltic Sea}}, edited by Almut Schülke, pp. 359--393. Routledge, London \& New York.}

\leavevmode\vadjust pre{\hypertarget{ref-shennan2015}{}}%
\CSLBlock{Shennan, Ian}
\CSLLeftMargin{ 2015}
\CSLRightInline{Handbook of sea-level research: Framing research questions. In \emph{{Handbook of Sea-Level Research}}, edited by Ian Shennan, Antony J Long, and Benjamin P Horton, pp. 3--25. Wiley, Chichester.}

\leavevmode\vadjust pre{\hypertarget{ref-shetelig1922}{}}%
\CSLBlock{Shetelig, Haakon}
\CSLLeftMargin{ 1922}
\CSLRightInline{\emph{{Primitive Tider i Norge -- En oversigt over stenalderen}}. John Griegs Forlag, Bergen.}

\leavevmode\vadjust pre{\hypertarget{ref-sognnes2003}{}}%
\CSLBlock{Sognnes, Kalle}
\CSLLeftMargin{ 2003}
\CSLRightInline{On shoreline dating of rock art. \emph{Acta Archaeologica} 74:189--209.}

\leavevmode\vadjust pre{\hypertarget{ref-solheim2012}{}}%
\CSLBlock{Solheim, Steinar}
\CSLLeftMargin{ 2012}
\CSLRightInline{{Lokal praksis og fremmed opphav. Arbeidsdeling, sosiale relasjoner og differensiering i østnorsk tidligneolitikum}. Unpublished PhD thesis, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-solheim2017b}{}}%
\CSLBlock{ (editor)}
\CSLLeftMargin{ 2017}
\CSLRightInline{\emph{{E18 Rugtvedt-Dørdal. Arkeologiske undersøkelser av lokaliteter fra steinalder og jernalder i Bamble kommune, Telemark fylke}}. Portal forlag, Kristiansand.}

\leavevmode\vadjust pre{\hypertarget{ref-solheim2020}{}}%
\CSLLeftMargin{ 2020 }
\CSLRightInline{{Mesolithic coastal landscapes. Demography, settlement patterns and subsistence economy in southeastern Norway}. In \emph{{Coastal Landscapes of the Mesolithic: Human Engagement with the Coast from the Atlantic to the Baltic Sea}}, edited by Almut Schülke. Routledge, London \& New York.}

\leavevmode\vadjust pre{\hypertarget{ref-solheim2021}{}}%
\CSLLeftMargin{ 2021 }
\CSLRightInline{Timing the Emergence and Development of Arable Farming in Southeastern Norway by Using Summed Probability Distribution of Radiocarbon Dates and a Bayesian Age Model. \emph{Radiocarbon} 63(5):1--22. DOI:\href{https://doi.org/10.1017/RDC.2021.80}{10.1017/RDC.2021.80}.}

\leavevmode\vadjust pre{\hypertarget{ref-solheim2018}{}}%
\CSLBlock{Solheim, Steinar, and Per Persson}
\CSLLeftMargin{ 2018}
\CSLRightInline{Early and mid-Holocene coastal settlement and demography in southeastern Norway: Comparing distribution of radiocarbon dates and shoreline-dated sites, 8500--2000 cal. BCE. \emph{Journal of Archaeological Science: Reports} 19:334--343. DOI:\href{https://doi.org/10.1016/j.jasrep.2018.03.007}{10.1016/j.jasrep.2018.03.007}.}

\leavevmode\vadjust pre{\hypertarget{ref-suxf8rensen1979}{}}%
\CSLBlock{Sørensen, Rolf}
\CSLLeftMargin{ 1979}
\CSLRightInline{{Late Weichselian deglaciation in the Oslofjord area, south Norway}. \emph{Boreas} 8(2):241--246. DOI:\url{https://doi.org/10.1111/j.1502-3885.1979.tb00806.x}.}

\leavevmode\vadjust pre{\hypertarget{ref-suxf8rensen1999}{}}%
\CSLLeftMargin{ 1999 }
\CSLRightInline{{En \(^{14}\)C datert og dendrokronologisk kalibrert strandforksyvningskurve for søndre Østfold. Sørøst-Norge}. In \emph{{Museumslandskap. Artikkelsamling til Kerstin Griffin på 60-årsdagen. Bind A}}, edited by Lotte Selsing and Grete Lillehammer, pp. 227--242. AmS-rapport 12A. Museum of Archaeology, Stavanger.}

\leavevmode\vadjust pre{\hypertarget{ref-suxf8rensen2014}{}}%
\CSLBlock{Sørensen, Rolf, Kari E. Henningsmoen, Helge I. Høeg, and Veronika Gälman}
\CSLLeftMargin{ 2014}
\CSLRightInline{{Holocene landhevningsstudier i søndre Vestfold og sørøstre Telemark -- Revidert kurve}. In \emph{{Vestfoldbaneprosjektet. Arkeologiske undersøkelser i forbindelse med ny jernbane mellom Larvik og Porsgrunn. Bind 1}}, edited by Stine Melvold and Per Persson, pp. 36--47. Portal, Kristiansand.}

\leavevmode\vadjust pre{\hypertarget{ref-sorenseninpress}{}}%
\CSLBlock{Sørensen, Rolf, Kari E Henningsmoen, Helge I Høeg, and Veronika Gälman}
\CSLLeftMargin{ in prep}
\CSLRightInline{{Holocen vegetasjonshistorie og landhevning i søndre Vestfold og sørøstre Telemark}. In \emph{{The Stone Age in Telemark. Archaeological results and scientific analysis from Vestfoldbaneprosjektet and E18 Rugtvedt-Dørdal}}, edited by Per Persson and Steinar Solheim.}

\leavevmode\vadjust pre{\hypertarget{ref-suxf8rensen2014b}{}}%
\CSLBlock{Sørensen, Rolf, Helge I. Høeg, Kari E. Henningsmoen, Göran Skog, Solveig F. Labowsky, and Bjørg Stabell}
\CSLLeftMargin{ 2014}
\CSLRightInline{{Utviklingen av det senglasiale og tidlig preboreale landskapet og vegetasjonen omkring steinalderboplassene ved Pauler}. In \emph{{E18 Brunlanesprosjektet. Bind I. Forutsetninger og kulturhistorisk sammenstilling}}, edited by Lasse Jaksland and Per Persson, pp. 171--213. University of Oslo, Museum of Cultural History, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-stokke2018b}{}}%
\CSLBlock{Stokke, Jo-Simon Frøshaug, and Gaute Reitan}
\CSLLeftMargin{ 2018}
\CSLRightInline{Kvastad A2. Lokalitet med funn fra tidlig- og mellommesolitikum og dyrkningsspor fra mellom- og senneolitikum. In \emph{{The Stone Age Coastal Settlement in Aust-Agder, Southeast Norway}}, edited by Gaute Reitan and Lars Sundström, pp. 375--407. Cappelen Damm Akademisk, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-stroeven2016}{}}%
\CSLBlock{Stroeven, Arjen P., Clas Hättestrand, Johan Kleman, Jakob Heyman, Derek Fabel, Ola Fredin, Bradley W. Goodfellow, Jonathan M. Harbor, John D. Jansen, Lars Olsen, Marc W. Caffee, David Fink, Jan Lundqvist, Gunhild C. Rosqvist, Bo Strömberg, and Krister N. Jansson}
\CSLLeftMargin{ 2016}
\CSLRightInline{Deglaciation of Fennoscandia. \emph{Quaternary Science Reviews} 147:91--121. DOI:\href{https://doi.org/10.1016/j.quascirev.2015.09.016}{10.1016/j.quascirev.2015.09.016}.}

\leavevmode\vadjust pre{\hypertarget{ref-svendsen1987}{}}%
\CSLBlock{Svendsen, John Inge, and Jan Mangerud}
\CSLLeftMargin{ 1987}
\CSLRightInline{Late Weichselian and Holocene sea-level history for a cross-section of western Norway. \emph{Journal of Quaternary Science} 2(2):113--132. DOI:\href{https://doi.org/10.1002/jqs.3390020205}{10.1002/jqs.3390020205}.}

\leavevmode\vadjust pre{\hypertarget{ref-tallavaara2020}{}}%
\CSLBlock{Tallavaara, Miikka, and Petro Pesonen}
\CSLLeftMargin{ 2020}
\CSLRightInline{Human ecodynamics in the north-west coast of Finland 10,000--2000 years ago. \emph{Quaternary International} 549:26--35. DOI:\href{https://doi.org/10.1016/j.quaint.2018.06.032}{10.1016/j.quaint.2018.06.032}.}

\leavevmode\vadjust pre{\hypertarget{ref-wang2019}{}}%
\CSLBlock{Wang, Ian}
\CSLLeftMargin{ 2019}
\CSLRightInline{\emph{topoDistance: Calculating topographic paths and distances. R package version 1.0.1. Https://CRAN.r-project.org/package=topoDistance}.}

\leavevmode\vadjust pre{\hypertarget{ref-wenn2012}{}}%
\CSLBlock{Wenn, Camilla Cecilie}
\CSLLeftMargin{ 2012}
\CSLRightInline{\emph{{Bosetningsspor, produksjonsområde og dyrkningsspor fra Neolitikum til Folkevandringstid. Bratsberg, 63/69, 244. Skien kommune, Telemark}}. University of Oslo, Museum of Cultural History, Oslo.}

\leavevmode\vadjust pre{\hypertarget{ref-wikell2009}{}}%
\CSLBlock{Wikell, Roger, Fredrik Molin, and Mattias Pettersson}
\CSLLeftMargin{ 2009}
\CSLRightInline{{The Archipelago of Eastern Middle Sweden - Mesolithic Settlement in Comparison with \(^{14}\)C and Shoreline Dating}. In \emph{{Chronology and Evolution within the Mesolithic of North-West Europe}}, edited by Philippe Crombé, Mark van Strydonck, Joris Sergant, Mathieu Boudin, and Machteld Bats, pp. 417--434. Cambridge Scholar Publishing, Brussels.}

\leavevmode\vadjust pre{\hypertarget{ref-yubero-gomez2016}{}}%
\CSLBlock{Yubero-Gómez, Maria, Xavier Rubio-Campillo, and Javier López-Cachero}
\CSLLeftMargin{ 2016}
\CSLRightInline{{The study of spatiotemporal patterns integrating temporal uncertainty in late prehistoric settlements in northeastern Spain}. \emph{Archaeological and Anthropological Sciences} 8(3):477--490. DOI:\href{https://doi.org/10.1007/s12520-015-0231-x}{10.1007/s12520-015-0231-x}.}

\end{CSLReferences}

\end{document}

### 2022-04-20 analysis/script/09shorelinedate.R

@@ -84,17 +84,19 @@ groupdat$year_median <- dat$mode
 plt <- dategrid %>%
   filter(probability != 0) %>%
 ggplot() +
+  geom_ribbon(data= sitecurve,
+              aes(x = years, ymin = lowerelev, ymax = upperelev), fill = "red",
+              alpha = 0.2) +
   geom_line(data = sitecurve, aes(x = years, y = upperelev, col = "red")) +
   geom_line(data = sitecurve, aes(x = years, y = lowerelev, col = "red")) +
   ggridges::geom_ridgeline(aes(x = years, y = 1, height = probability * 18000),
-                              colour = "black", fill = "grey", alpha = 0.7) +
+                              colour = NA, fill = "darkgrey", alpha = 0.7) +
   geom_segment(data = groupdat, aes(x = start, xend = end,
                                   y = 0, yend = 0), col = "black") +
-  labs(y = "Meters above present sea-level", x = "BCE",
+  labs(y = "Meters above present sea-level", x = "Shoreline date (BCE)",
        title = paste("\U03BB =", as.numeric(round(expfit$estimate, 3)))) +
-  scale_x_continuous(limits = c(-9500, -4500),
-                     breaks = (seq(-9000, -5000, 1000)), expand = c(0, 0)) +
-  scale_y_continuous(limits = c(0, 80)) +
+  scale_x_continuous(breaks = (seq(-9000, -5000, 1000)), expand = c(0, 0)) +
+  coord_cartesian(ylim = c(0, 80), xlim = c(-9500, -4500)) +
   theme_bw() +
   theme(legend.position = "None")
 
@@ -109,11 +111,13 @@ x_extent <- ggplot_build(plt)$layout$panel_scales_x[[1]]$range$range
 expdatt$probs_scaled <- expdatt$probs / max(expdatt$probs) *
   diff(x_extent)/7 + x_extent[1]
 
-plt <- plt + geom_polygon(data = expdatt, aes(x =  probs_scaled,
+dplt <- plt + geom_polygon(data = expdatt, aes(x =  probs_scaled,
                                        y = as.numeric(siteelev) - offset),
-                   fill = "#046c9a", col = "black", alpha = 0.6)
-  # geom_hline(yintercept = as.numeric(siteelev), linetype = "dashed",
-  #            col = "black")
+                   fill = "#046c9a", alpha = 0.6) +
[…]

### 2022-04-21 analysis/paper/paper.Rmd
Working out some kinks in figures from backwards_dating.R and the main text

@@
 keywords: |
-  Stone Age; Settlement patterns; Scandinavia; Relative sea-level change; Shoreline dating; Monte Carlo simulation
+  Shoreline dating; Stone Age; Settlement patterns; Scandinavia; Relative sea-level change
 ---
@@
 However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas as evidenced by artefact inventories or multiple site features.
-Not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates.
+Not remnants of stays as ephemeral to only be discernible by isolated features or dubious ^14^C-dates.
 The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.
@@
 
-```{r results2, echo = FALSE, fig.cap = "Histograms showing the simulated distance from the shoreline using dates corresponding to the site inventories.
+```{r results2, echo = FALSE, fig.cap = "Histograms showing the simulated distance from the shoreline using radiocarbon dates corresponding to the site inventories.
 Negative values have been set to zero.
@@
 
-```{r exampledat, echo = FALSE, fig.cap = "Shoreline dating Hegna vest 1.
-The dashed line indicates the elevation of the site.
-The decay ratio follows Figure 7A.
-The shoreline date in grey is underlined with the 95\\% HDR in black.", out.width = "500px", fig.align="center", regfloat=TRUE}
-knitr::include_graphics(here("analysis/figures/sdate.png"))
-
-```
+An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation (Figure \@ref(fig:results2)).
+While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data.
+However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
+If one accepts this, the probability density function for exponential decay can be used to characterise the vertical distance between sites and the shoreline, and be used to inform a method for shoreline dating that takes this into account.
 
@@
 
-An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation (Figure \@ref(fig:results2)).
-While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data.
-However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
-If one accepts this, the probability density function for exponential decay can be used to characterise the vertical distance between sites and the shoreline and leveraged for devising a method for shoreline dating that takes this into account.
-
-The procedure is thus aimed at determining the likely age of the occupation of a site based on its altitude above present day sea-level, with reference to shoreline displacement and the likely elevation of the site above sea-level when it was in use.
+The procedure to be outlined is thus aimed at determining the likely age of the occupation of a site based on its altitude above present day sea-level, with reference to shoreline displacement and the likely elevation of the site above the sea-level when it was in use.
 For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly.
-This leads the problem to become analogous to that of the calibration of a radiocarbon date.
-Drawing on the standard Bayesian formulation for the calibration procedure for 14C-dates [e.g.
+This leads the problem to become analogous to that of the calibration of a radiocarbon date (see also Figure \@ref(fig:exampledat)).
+Drawing on the standard Bayesian formulation for the calibration procedure of ^14^C-dates [e.g.
 @bronkramsey2009], the probability density associated with the calendar age can then be given as
@@
 \begin{equation} 
-  p(x_{i} | \thetha_{i}) ~ p(\theta_{i} | x_{i})p(x_{i})
+  p(x_{i} \mid \theta_{i}) = p(\theta_{i} \mid x_{i} )p(x_{i} )
   (\#eq:bayes)
@@
 
-where for any site i the unknown calendar date is denoted $\thetha$ and the elevation of the contemporaneous sea-level $x$.
-The elevation of the sea-level is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $d$.
+where for any site $i$ the unknown calendar date is denoted $\theta$ and $x$ denotes the elevation of the contemporaneous sea-level relative to its present level.
+First, finding the elevation of the sea-level is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $d$.
 Based on the simulation results above, the distance from the elevation of the site to the contemporaneous shoreline is defined by the probability density function for exponential decay
@@
 Where $\lambda$ is the decay ratio.
-This can then be coupled with the trajectory of relative sea-level change, denoted RSL, to find the probability density for the calendar date.
-Here the relative sea-level change is defined by a uniform probability density function over the range between the lower and upper displacement curves interpolated to the site location
+This can then be coupled with the trajectory of relative sea-level change, denoted $RSL$, to find the likely elevation of the sea-level.
+Here the relative sea-level change is defined by a uniform probability density function over the range between the lower and upper bounds of the displacement curve interpolated to the site location
 
@@
 \end{equation}
+
+Finding the probability density for the calendar date then becomes a matter of coupling the probability of the distance between site and shoreline to the probability of the altitude of the shoreline, and transferring this probability to each calendar year.
 
@@
 
-Eq.
-\@ref(eq:shoredate) is used to shoreline date the same sites from where this relationship was derived (Figure \@ref(fig:shoredate)), employing $\lambda$ = `r round(expfit$estimate, 3)` for the decay ratio in Eq.
-\@ref(eq:decay).
-This is the ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
-For illustrative purposes the Late Neolithic sites were also subjected to the proposed method for shoreline dating.
-Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits.
+An example of the implementation of Eq.
+\@ref(eq:shoredate) is given in Figure \@ref(fig:shoredate), where $\lambda$ = `r round(expfit$estimate, 3)`.
+This is the decay ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
+In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date all of the sites from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes.
+Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were all dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits.
 The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by @parnell2008.
@@
 
+```{r exampledat, echo = FALSE, fig.cap = "Shoreline dating of Hegna vest 1.
+The dashed line marks the mean elevation of the site polygon.
+The exponential function decays following the $\\lambda$ from Figure 7A.
+The resulting shoreline date in grey is underlined with the 95\\% HDR in black.", out.width = "500px", fig.align="center", regfloat=TRUE}
+knitr::include_graphics(here("analysis/figures/sdate.png"))
+
+```
+
 ```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory using the method proposed here.
@@
 
-```{r shoredatediff, echo = FALSE, fig.cap = "Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure 8.
+```{r shoredatediff, echo = FALSE, fig.cap = "Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure 9.
 When the range of the 95\\% HDR for age difference crosses zero, the shoreline and radiocarbon dates are considered synchronous.
 Line segments with vertical bars indicate that this does not cross zero and that the dates are not in agreement.
-The division and colour coding at the top of the plots reflect the divison of site phases given in Figure 8.", out.width = "525px", fig.align="center", regfloat=TRUE}
+The division and colour coding at the top of the plots reflect the divison of site phases given in Figure 9.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate2.png"))
@@
 Finally, they are sometimes only meant to indicate a lower bound for when the sites could have been in use.
-Overall, the results suggest that shoreline dating has generally been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g.
+Overall, the results could, with some danger of circularity, suggest that shoreline dating has generally been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g.
 @roalkvam2022].
@@
 Figure \@ref(fig:map-iso)B).
-Further, as some of the date ranges extend well beyond major chronological divisions, even into the Iron Age, they could be severely and securely constrained with only cursory reference to typology.
+Furthermore, as some of the date ranges extend well beyond major chronological divisions, even into the Iron Age, they could be severely and securely constrained with only cursory reference to typology.
 While this would be trivial in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed.
@@
 Furthermore, while this remains to be formalised and explored further, it was also showed how the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data.
-As the precision of the method is both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date than younger sites located towards the south-west, the impact of such additional information will also vary.
+As the precision of the method is both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date, the impact of such additional information will also vary.
 
@@
 This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range could possibly also have implications for the site location relative to the shoreline, depending on the topography [@helskog1978].
-The potential of dimensions such as these was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline.
+The potential of exploring dimensions such as these was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline.
 If patterns related to such locational patterns can be discerned and unpicked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information beyond their location is available.
@@
 @bronkramsey2009; @bronkramsey2015].
-While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration.
+While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration and which indicate that some of the precision following from the outlined approach is likely to be spurious.
 

### 2022-04-21 analysis/script/07backwards_dating.R

@@ -53,43 +53,43 @@ sites_sa <- sites_sa %>%
 # (see supplementary material for more on this)
 feature_sites <- c("Langangen Vestgård 7", "Vallermyrene 2")
 
-# Uncomment to rerun
-shorelinedates <- list()
-for(i in 1:nrow(sites_sa)){
-  print(paste(i, sites_sa$name[i]))
-  if(sites_sa$name[i] %in% feature_sites){
-    shorelinedates[[i]] <- shoreline_date(sitename = sites_sa$name[i],
-                                          elev = dtm,
-                                          disp_curves = displacement_curves,
-                                          sites = sites_sa,
-                                          iso = isobases,
-                                          expratio = expfit$estimate,
-                                          siteelev = "mean",
-                                          reso = 0.001,
-                                          specified_elev = NA,
-                                          sitelimit = FALSE,
-                                          features = rcarb_sa)
-  } else {
-  shorelinedates[[i]] <- shoreline_date(sitename = sites_sa$name[i],
-                                        elev = dtm,
-                                        disp_curves = displacement_curves,
-                                        sites = sites_sa,
-                                        iso = isobases,
-                                        expratio = expfit$estimate,
-                                        siteelev = "mean",
-                                        reso = 0.001,
-                                        specified_elev = NA)
-  }
-}
-
+# # Uncomment to rerun
+# shorelinedates <- list()
+# for(i in 1:nrow(sites_sa)){
+#   print(paste(i, sites_sa$name[i]))
+#   if(sites_sa$name[i] %in% feature_sites){
+#     shorelinedates[[i]] <- shoreline_date(sitename = sites_sa$name[i],
+#                                           elev = dtm,
+#                                           disp_curves = displacement_curves,
+#                                           sites = sites_sa,
+#                                           iso = isobases,
+#                                           expratio = expfit$estimate,
+#                                           siteelev = "mean",
+#                                           reso = 0.001,
+#                                           specified_elev = NA,
+#                                           sitelimit = FALSE,
+#                                           features = rcarb_sa)
[…]

### 2022-04-23 analysis/paper/paper.Rmd
Fixing figures and trying to get the mathematical notation correct

@@
 Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from the common isobase perpendicular to this gradient (Figure \@ref(fig:map-iso)).
-The SLIPs indicate the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves based on the present day tidal range.
+The SLIPs indicate the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present day tidal range.
 This is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
-The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021, 30cm at the standard port Helgeroa in Larvik].
+The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021; 30cm at the standard port Helgeroa in Larvik].
 Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
@@
 The excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure.
-As the data collection for the DTM was started by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has therefore sometimes been severely impacted by disturbances after the excavation of the site.
+As the data collection for the employed DTM was begun by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has sometimes been severely impacted by disturbances after the excavation.
 In addition to employing 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster.
@@
 
-The procedure to be outlined is thus aimed at determining the likely age of the occupation of a site based on its altitude above present day sea-level, with reference to shoreline displacement and the likely elevation of the site above the sea-level when it was in use.
+The procedure for shoreline dating to be outlined is thus aimed at determining the likely age of the occupation of a site based on its altitude above present day sea-level, with reference to shoreline displacement and the likely elevation of the site above the sea-level when it was in use.
 For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly.
 This leads the problem to become analogous to that of the calibration of a radiocarbon date (see also Figure \@ref(fig:exampledat)).
-Drawing on the standard Bayesian formulation for the calibration procedure of ^14^C-dates [e.g.
+Drawing on the standard Bayesian formulation for the calibration procedure of individual ^14^C-dates [e.g.
 @bronkramsey2009], the probability density associated with the calendar age can then be given as
@@
 
-where for any site $i$ the unknown calendar date is denoted $\theta$ and $x$ denotes the elevation of the contemporaneous sea-level relative to its present level.
+where for any site $i$ the unknown calendar date is denoted $\theta$ and $x$ denotes the elevation of the contemporaneous sea-level, relative to its present level.
 First, finding the elevation of the sea-level is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $d$.
@@
 \begin{equation}
-  p(d) = \lambda {e}^{- \lambda x}
+  p(\alpha - d) = \lambda {e}^{- \lambda d}
   (\#eq:decay)
@@
 
-Where $\lambda$ is the decay ratio.
+where $\lambda$ is the decay ratio.
 This can then be coupled with the trajectory of relative sea-level change, denoted $RSL$, to find the likely elevation of the sea-level.
-Here the relative sea-level change is defined by a uniform probability density function over the range between the lower and upper bounds of the displacement curve interpolated to the site location
+Here the relative sea-level change is defined by a uniform probability density function over the range between the lower and upper bounds of the displacement curve, interpolated to the site location
 
@@
 The dashed line marks the mean elevation of the site polygon.
-The exponential function decays following the $\\lambda$ from Figure 7A.
+The exponential function decays with ratio $\\lambda$ from Figure 7A.
 The resulting shoreline date in grey is underlined with the 95\\% HDR in black.", out.width = "500px", fig.align="center", regfloat=TRUE}
@@
 ```{r shoredatediff, echo = FALSE, fig.cap = "Evaluation of the agreement between the shoreline dates and radiocarbon dates given in Figure 9.
-When the range of the 95\\% HDR for age difference crosses zero, the shoreline and radiocarbon dates are considered synchronous.
-Line segments with vertical bars indicate that this does not cross zero and that the dates are not in agreement.
+When the range of the 95\\% HDR for age difference crosses zero, the shoreline and radiocarbon dates are considered to be in agreement.
+Line segments with vertical bars indicate the HDR does not cross zero and that the dates do not correspond.
 The division and colour coding at the top of the plots reflect the divison of site phases given in Figure 9.", out.width = "525px", fig.align="center", regfloat=TRUE}
@@
 Some limitations and sources of likely variation and uncertainty that have not been considered should also be mentioned.
-First of all the sample size is quite strained and the future addition more sites might alter the picture considerably.
+First of all the sample size is quite strained and the future addition of more sites might alter the picture considerably.
 Secondly, the DTM has only been corrected for major modern disturbances.
@@
 @bronkramsey2009; @bronkramsey2015].
-While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration and which indicate that some of the precision following from the outlined approach is likely to be spurious.
+While each of these factors will have variable impact on the final results, they clearly represent dimensions which would all benefit from further consideration and which means that some of the precision following from the outlined approach is likely to be spurious.
 

### 2022-04-23 analysis/script/07backwards_dating.R

@@ -185,10 +185,11 @@ shrplt <- ggplot(data = hdrdat,
   theme_bw()
 
 # Save plot
-ggsave(file = here("analysis/figures/shoredate_2.png"), shrplt,
+ggsave(file = here("analysis/figures/shoredate.png"), shrplt,
        width = 200, height = 250, units = "mm")
 
 # Next sections finds synchroneity between radiocarbon and shoreline dates
+# Grab the probablities again without filtering for plotting
 shorelinedates <- bind_rows(shorelinedates) %>%
   filter(site_name %in% unique(corsites$site_name))
 
@@ -313,7 +314,7 @@ agedfplt4 <- agedfplt + stat_summary(data = agedif4, aes(y = site_name,
   scale_color_manual(name = "", labels = c("Mesolithic",
                                            "Early and Middle Neolithic",
                                            "Late Neolithic"),
-                     values = c("black", "#00BA38", "deeppink1"),
+                     values = c("#00BA38", "deeppink1", "black"),
                      guide = "none") +
   geom_hline(yintercept = Inf, col = "forestgreen", size = 2) +
   theme(axis.title.y = element_blank(),

### 2022-04-26 analysis/paper/paper.Rmd
Added some references

@@
 The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than an earliest possible date for when a site could have been in use [see review by @nordqvist1999].
-The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @jaksland2014; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below].
+The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @glorstad2002; @glorstad2003; @glorstad2004; @jaksland2001; @jaksland2012; @jaksland2012a; @jaksland2014a; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below].
 Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time.
@@
 
-The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; also discussed in @berg-hansen2009; @bergsvik2009].
+The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; @bjerck2017; also discussed by @berg-hansen2009; @bergsvik2009].
 The same logic has also been extended to the hinter- and inland regions, where sites are to be predominantly located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020].
-This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020].
+This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@prescott2020; cf.
+@solheim2021].
 The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm is to have led site locations to be more withdrawn from the shoreline [e.g.
-@bakka1971; @prescott2012; @ostmo2008, 223].
+@bakka1971; @prescott2020; @ostmo2008, 223].
 That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2012].

### 2022-04-27 analysis/paper/paper.Rmd
Revised equations for shoreline dating

@@
 For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly.
-This leads the problem to become analogous to that of the calibration of a radiocarbon date (see also Figure \@ref(fig:exampledat)).
-Drawing on the standard Bayesian formulation for the calibration procedure of individual ^14^C-dates [e.g.
-@bronkramsey2009], the probability density associated with the calendar age can then be given as
+This leads the problem to become analogous to that of the calibration of a radiocarbon date [@bronkramsey2009, see also Figure \@ref(fig:exampledat)].
 
-\begin{equation} 
-  p(x_{i} \mid \theta_{i}) = p(\theta_{i} \mid x_{i} )p(x_{i} )
-  (\#eq:bayes)
-\end{equation}
-
-where for any site $i$ the unknown calendar date is denoted $\theta$ and $x$ denotes the elevation of the contemporaneous sea-level, relative to its present level.
-First, finding the elevation of the sea-level is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $d$.
-Based on the simulation results above, the distance from the elevation of the site to the contemporaneous shoreline is defined by the probability density function for exponential decay
+First, finding the elevation of the sea-level at the time the site was in use is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $d$.
+Based on the simulation results above, the distance from the elevation of the site to the contemporaneous shoreline is defined by the probability density function for exponential decay:
 
@@
 \begin{equation}
-  p(\alpha - d) = \lambda {e}^{- \lambda d}
+  p(\alpha - d) = \lambda {e}^{- \lambda (\alpha - d)}
   (\#eq:decay)
@@
 where $\lambda$ is the decay ratio.
-This can then be coupled with the trajectory of relative sea-level change, denoted $RSL$, to find the likely elevation of the sea-level.
-Here the relative sea-level change is defined by a uniform probability density function over the range between the lower and upper bounds of the displacement curve, interpolated to the site location
+This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $t$.
+Here this is defined by a uniform probability density function over the range between the lower $t_{l}$ and upper $t_{u}$ bounds of the displacement curve that as been interpolated to the site location:
 
 \begin{equation}
-p(RSL) = U[RSL_{lower}, RSL_{upper}]	
+p(t | \alpha - d) = U[t_{l | \alpha - d}, t_{u | \alpha - d}]	
   (\#eq:rsl)
@@
 
-Finding the probability density for the calendar date then becomes a matter of coupling the probability of the distance between site and shoreline to the probability of the altitude of the shoreline, and transferring this probability to each calendar year.
+Finding the probability for calendar dates then becomes a matter of coupling the probability of the distance between site and shoreline to that of the displacement curve:
+
+\begin{equation}
+p(t | \alpha - d) = p(t | \alpha -d)p(\alpha - d)
+  (\#eq:comb)
+\end{equation}
+
+We can then get rid of parameter $d$ by summarising over the possible distances between site and the shoreline.
+The probability distribution for the calendar date of the use of the site, given its elevation, is then given by:
 
@@
 \begin{equation} 
-  placeholder
+  p(t | \alpha) = \sum_{d} p(t | \alpha -d)p(\alpha -d)
   (\#eq:shoredate)
@@
 This is the decay ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
+For the numerical implementation, $d$ is defined by increments of 0.001 from the site elevation $\alpha$.
 In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date all of the sites from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes.
@@
 ```{r exampledat, echo = FALSE, fig.cap = "Shoreline dating of Hegna vest 1.
-The dashed line marks the mean elevation of the site polygon.
+The dashed line marks the mean elevation of the site polygon which is used for $\\alpha$ in the dating of the site.
 The exponential function decays with ratio $\\lambda$ from Figure 7A.
@@
 The sites have been dated using what @jaksland2014 gives as the lowest elevation of finds at each site, and by employing a exponential decay ratio of `r round(expfit3$estimate, 3)`, to allow for more deviance in the distance between site and shoreline.
-This corresponds to the decay ratio for sites older than 7000 BCE in Figure \@ref(fig:results2).
+This corresponds to the decay ratio for results older than 7000 BCE in Figure \@ref(fig:results2).
 
@@
 
+Given that shoreline dating is contingent on regularities in human behaviour it should naturally be applied with care.
+Furthermore, formulating and visualising the method along the lines of how radiocarbon dates are treated, as was done here, does stand the chance of giving a veneer of radiometric accuracy that is not warranted.
+That being said, the best chance we have of not throwing away precious temporal data is to rigorously evaluate the method using independent data such as radiocarbon dates, by offering a precise formulation of how it could be applied, by specifying what sources of uncertainty are accounted for, and by making this process open through the dissemination of underlying data and programming code.
+
 Finally, this analysis employed a simulation approach to integrate multiple sources of spatio-temporal uncertainty.

### 2022-04-29 analysis/paper/paper.Rmd
Minor edits

@@
 csl: "../templates/saa.csl" # Insert path for the bib-style
+abstract: |
+  A central premise for Norwegian Stone Age archaeology is that most coastal sites were located on or close to the contemporary shoreline when they were in use.
+By reconstructing the trajectory of relative sea-level change, this offers a dating method termed shoreline dating, separate from both radiometric and typological methods which has been widely applied.
+However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which, due to the relative rarity of the method, has been under limited scrutiny compared to more traditional dating techniques in archaeology.
+This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast.
+This is done by combining independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement through Monte Carlo simulation.
+The findings largely confirm previous evaluations of this relationship, indicting that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied.
+Drawing on the quantative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed, and compared to previous applications of the technique.
+
 keywords: |
@@
 
+## Highlights {-}
+
+* Simulates the spatial relation between sites and the prehistoric shoreline
+* Quantification of horizontal, topographic and vertical distance
+* Confirms close association between sites and the contemporaneous shoreline
+* Proposes a formalised method for the shoreline dating of pre-Late Neolithic sites
+
+Keywords: `r rmarkdown::metadata$keywords`
 ```{r setup, echo = FALSE}
@@
 
-The method of shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty.
+Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty.
 The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions.
-To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves using inverse distance weighting [e.g.
+To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves, using inverse distance weighting [e.g.
 @conolly2020; @conolly2006, 94--97].
-This was done for each year along the entirety of the curves, weighting the interpolation by the standard squared inverse of the distances.
+This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances.
 The result of this process is shown for an example site in Figure \@ref(fig:example-1).
 For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
-Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same event, here termed settlement or site phases.
+Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same event, here termed settlement or site phase.
 In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed.
@@
 @conolly2006, 253; @herzog2013].
-In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero.
-In the case that the sea-polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon.
+In the case where the sea polygons intersects the site polygon, all distance measures were set to zero.
+In the case that the sea polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon.
 While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
@@
 This leads the problem to become analogous to that of the calibration of a radiocarbon date [@bronkramsey2009, see also Figure \@ref(fig:exampledat)].
-
-First, finding the elevation of the sea-level at the time the site was in use is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $d$.
+First, finding the elevation of the sea-level at the time the site was in use is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $D$.
 Based on the simulation results above, the distance from the elevation of the site to the contemporaneous shoreline is defined by the probability density function for exponential decay:
@@
 \begin{equation}
-  p(\alpha - d) = \lambda {e}^{- \lambda (\alpha - d)}
+  p(\alpha - D) = \lambda {e}^{- \lambda (\alpha - D)}
   (\#eq:decay)
@@
 where $\lambda$ is the decay ratio.
-This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $t$.
-Here this is defined by a uniform probability density function over the range between the lower $t_{l}$ and upper $t_{u}$ bounds of the displacement curve that as been interpolated to the site location:
+This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $T$.
+Here this is defined by a uniform probability density function over the range between the lower $T_{l}$ and upper $T_{u}$ bounds of the displacement curve that as been interpolated to the site location:
 
 \begin{equation}
-p(t | \alpha - d) = U[t_{l | \alpha - d}, t_{u | \alpha - d}]	
+p(T | \alpha - D) = U[T_{l | \alpha - D}, T_{u | \alpha - D}]	
   (\#eq:rsl)
@@
 
-Finding the probability for calendar dates then becomes a matter of coupling the probability of the distance between site and shoreline to that of the displacement curve:
+Finding the probability for the date of the site then becomes a matter of transferring the probability of the distance between site and shoreline and that of the displacement curve to calendar dates:
 
 \begin{equation}
-p(t | \alpha - d) = p(t | \alpha -d)p(\alpha - d)
+p(T | \alpha - D) = p(T | \alpha - D)p(\alpha - D)
   (\#eq:comb)
@@
 
-We can then get rid of parameter $d$ by summarising over the possible distances between site and the shoreline.
-The probability distribution for the calendar date of the use of the site, given its elevation, is then given by:
+We can then get rid of parameter $D$ by summarising over all possible distances between site and the shoreline. Given its elevation, the probability of the date for the occupation of the site is then:
 
@@
 \begin{equation} 
-  p(t | \alpha) = \sum_{d} p(t | \alpha -d)p(\alpha -d)
+  p(T | \alpha) = \sum_{D} p(T | \alpha - D)p(\alpha - D)
   (\#eq:shoredate)
@@
 
-An example of the implementation of Eq.
-\@ref(eq:shoredate) is given in Figure \@ref(fig:shoredate), where $\lambda$ = `r round(expfit$estimate, 3)`.
+An example implementation of this approach is given in Figure \@ref(fig:shoredate), where $\lambda$ = `r round(expfit$estimate, 3)`.
 This is the decay ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
-For the numerical implementation, $d$ is defined by increments of 0.001 from the site elevation $\alpha$.
+For the numerical implementation, $D$ is here defined as a sequence of increments of 0.001m from the site elevation $\alpha$.
 In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date all of the sites from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes.
@@
 ```{r exampledat, echo = FALSE, fig.cap = "Shoreline dating of Hegna vest 1.
-The dashed line marks the mean elevation of the site polygon which is used for $\\alpha$ in the dating of the site.
+The dashed line marks the mean elevation of the site polygon which is used to inform $\\alpha$ in the dating of the site.
 The exponential function decays with ratio $\\lambda$ from Figure 7A.
@@
 That being said, the results do also indicate that shoreline dating has at times been applied with an exaggerated degree of precision.
-While the implications of a more stable RSL-change for the duration of use and re-use of site locations are well known, this also appears to be somewhat under-appreciated for the purposes of shoreline dating.
+While the implications of a more stable RSL-change for shoreline dating are well known, this also appears to be somewhat under-appreciated in the practical implementation of the method.
 The results also highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% HDRs for the dates.
@@
 Furthermore, based on the quantitative nature of these findings, an initial formulation of a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed.
-Apart from taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves implementing Eq.
-\@ref(eq:shoredate) to account for the distance between the sites and the shoreline.
+Apart from taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves accounting for the distance between the sites and the shoreline.
 When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio of `r round(expfit$estimate, 3)` (Figure \@ref(fig:redate)A) to characterise this relationship.
@@
 First of all the sample size is quite strained and the future addition of more sites might alter the picture considerably.
-Secondly, the DTM has only been corrected for major modern disturbances.
+Secondly, the validity of the method was evaluated by applying it to the data from where the input parameters were derived.
+Fitting and evaluating a model using the exact same data will likely exaggerate its performance.
+Thirdly, the DTM has only been corrected for major modern disturbances.
 This means that other forms of erosion, although likely not that prevalent, has not been taken into account.
-Thirdly, the DTM has a vertical error which could also benefit from being integrated in the analysis [cf.
+Fourthly, the DTM has a vertical error which could also benefit from being integrated in the analysis [cf.
 @lewis2021].
-Fourthly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves further away from the isobases of the displacement curves.
+Fifthly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves further away from the isobases of the displacement curves.
 This is also related to the fact that the RSL data can be handled in different ways than with the isobase method that has been used for the compilation of the employed displacement curves [cf.
 @creel2022].
-Fifthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration [cf.
+Sixthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration [cf.
 @mjærum2022].
@@
 Furthermore, formulating and visualising the method along the lines of how radiocarbon dates are treated, as was done here, does stand the chance of giving a veneer of radiometric accuracy that is not warranted.
-That being said, the best chance we have of not throwing away precious temporal data is to rigorously evaluate the method using independent data such as radiocarbon dates, by offering a precise formulation of how it could be applied, by specifying what sources of uncertainty are accounted for, and by making this process open through the dissemination of underlying data and programming code.
+That being said, the best chance we have of not throwing away precious temporal data, or exaggerating our handle on it, is arguably to rigorously evaluate the method using independent data such as radiocarbon dates, by offering a precise formulation of how it could be applied, by specifying what sources of uncertainty are accounted for and by making this process open through the dissemination of underlying data and programming code.
 

### 2022-04-29 analysis/supplementary-materials/supplementary.Rmd
Minor edits

@@
   - \usepackage{placeins}
-output: pdf_document
+output: 
+  pdf_document:
+    toc: true
 bibliography: references.bib
@@
 
-This document shows the simulation results for each individual site analysed in the study with some additional notes pertaining to edits to the DTM, treatment of ^14^C-dates and other relevant information.
+## Introduction
+
+This document presents the sensitivity analysis performed for informing the number of simulation runs in the analysis of each site, the analysis of each individual site, and the results of re-dating previously shoreline dated sites.
+All underlying and derived data are available in the online repository for the paper at https://osf.io/7F9SU
+
+## Sensitivity analysis
+
+The first figure shows the result of a test run for the site Hovland 5, where the simulations were run 5000 times to identify when the means for the different distance measures converged.
+This was used to inform the use of 1000 simulation runs for the subsequent analysis of the other sites, as the means appear to have stabilised sufficiently by this point.
+Hovland 5 was chosen because the site only has a single ^14^C-date, giving a wide uncertainty range, and because it is located in an area of the landscape where the landscape results in quite large variability on the distance measures.
+
+```{r sample_size,  echo = FALSE, out.width = "500px", warning=FALSE}
+include_graphics(here("analysis/figures/sample_size.png"))
+
+```
+
+## Site analysis
+
+The presentation of simulation results for each individual site also has some additional notes pertaining to edits to the DTM, treatment of ^14^C-dates and other relevant information.
 The sites are presented in alphabetical order.
-All underlying and derived data are available in the online repository for the paper at https://osf.io/7F9SU
 
@@
 
-The first figure shows the result of a test run for the site Hovland 5, where the simulations were run 5000 times to identify when the means for the different distance measures converged.
-This was used to inform the use of 1000 simulation runs for the subsequent analysis of the other sites, as the means appear to have stabilised sufficiently by this point.
-Hovland 5 was chosen because the site only has a single ^14^C-date, giving a wide uncertainty range, and because it is located in an area of the landscape where the landscape results in quite large variability on the distance measures.
-
-```{r sample_size,  echo = FALSE, out.width = "500px"}
-include_graphics(here("analysis/figures/sample_size.png"))
-
-```
-
-Below follows the presentation of the individual sites.
 The first plot on the first row for each site shows all calibrated radiocarbon dates from the Stone Age associated with the site (see analysis/data/raw_data/radiocarbon.csv for all dates, including those falling outside of the period).
@@
 
-```{r adalvestre1,  echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r adalvestre1,  echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 
@@
 
-```{r dalvestre2,  echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r dalvestre2,  echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 knitr::include_graphics(here("analysis/figures/adalvestre2.png"))
@@
 
-```{r alveberget8,  echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r alveberget8,  echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/alveberget8.png"))
@@
 
-```{r anvik,  echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r anvik,  echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/anvik.png"))
@@
 
-```{r auve,  echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r auve,  echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/auve.png"))
@@
 
-```{r bakke, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r bakke, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 knitr::include_graphics(here("analysis/figures/bakke.png"))
@@
 
-```{r brunstad24, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r brunstad24, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 knitr::include_graphics(here("analysis/figures/brunstad24.png"))
@@
 
-```{r brunstad25, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r brunstad25, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 knitr::include_graphics(here("analysis/figures/brunstad25.png"))
@@
 
-```{r dordal, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r dordal, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 
@@
 
-```{r dybdalshei1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r dybdalshei1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/dybdalshei1.png"))
@@
 
-```{r dybdalshei2, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r dybdalshei2, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/dybdalshei2.png"))
@@
 
-```{r gunnarsrod4, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r gunnarsrod4, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/gunnarsrod4.png"))
@@
 
-```{r gunnarsrod5, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r gunnarsrod5, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/gunnarsrod5.png"))
@@
 
-```{r gunnarsrod6a, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r gunnarsrod6a, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/gunnarsrod6a.png"))
@@
 
-```{r gunnarsrod6b, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r gunnarsrod6b, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/gunnarsrod6b.png"))
@@
 
-```{r gunnarsrod7, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r gunnarsrod7, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/gunnarsrod7.png"))
@@
 
-```{r gunnarsrod10, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r gunnarsrod10, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/gunnarsrod10.png"))
@@
 
-```{r hegnavest1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r hegnavest1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/hegnavest1.png"))
@@
 
-```{r hegnavest2, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r hegnavest2, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/hegnavest2.png"))
@@
 
-```{r hegnavest3, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r hegnavest3, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/hegnavest3.png"))
@@
 
-```{r hesthagc2, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r hesthagc2, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/hesthagc2.png"))
@@
 
-```{r hesthagc4, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r hesthagc4, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/hesthagc4.png"))
@@
 
-```{r hovland1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r hovland1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/hovland1.png"))
@@
 
-```{r hovland3, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r hovland3, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/hovland3.png"))
@@
 
-```{r hovland4, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r hovland4, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/hovland4.png"))
@@
 
-```{r hovland5, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r hovland5, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/hovland5.png"))
@@
 
-```{r hydal4, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r hydal4, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/hydal4.png"))
@@
 
-```{r krogenesd1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r krogenesd1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/krogenesd1.png"))
@@
 
-```{r krogenesd2, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r krogenesd2, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/krogenesd2.png"))
@@
 
-```{r kvastada1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r kvastada1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/kvastada1.png"))
@@
 
-```{r kvastada2, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r kvastada2, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/kvastada2.png"))
@@
 
-```{r kvastada4, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r kvastada4, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/kvastada4.png"))
@@
 
-```{r kvastada9, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r kvastada9, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/kvastada9.png"))
@@
 
-```{r langangenv1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r langangenv1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/langangenv1.png"))
@@
 
-```{r langangenv3, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r langangenv3, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/langangenv3.png"))
@@
 
-```{r langangenv5, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r langangenv5, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/langangenv5.png"))
@@
 
-```{r langangenv6, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r langangenv6, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/langangenv6.png"))
@@
 
-```{r langangenv7, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r langangenv7, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/langangenv7.png"))
@@
 
-```{r langemyr, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r langemyr, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/langemyr.png"))
@@
 
-```{r lovas1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r lovas1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 knitr::include_graphics(here("analysis/figures/lovas1.png"))
@@
 
-```{r lovas2, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r lovas2, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 knitr::include_graphics(here("analysis/figures/lovas2.png"))
@@
 
-```{r lovas3, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r lovas3, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 knitr::include_graphics(here("analysis/figures/lovas3.png"))
@@
 
-```{r lunaveien, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r lunaveien, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/lunaveien.png"))
@@
 
-```{r nauena, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r nauena, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/nauena.png"))
@@
 
-```{r nauenc, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r nauenc, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/nauenc.png"))
@@
 
-```{r nordby1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r nordby1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/nordby1.png"))
@@
 
-```{r nordby52, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r nordby52, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/nordby52.png"))
@@
 
-```{r pauler1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r pauler1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/pauler1.png"))
@@
 
-```{r pauler2, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r pauler2, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/pauler2.png"))
@@
 
-```{r pjonkerodr1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r pjonkerodr1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/pjonkerodr1.png"))
@@
 
-```{r prestemoen1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r prestemoen1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/prestemoen1.png"))
@@
 
-```{r ragnhildrod, echo = FALSE, out.width = "500px", regfloat=TRUE, warning = FALSE}
+```{r ragnhildrod, echo = FALSE, out.width = "500px", regfloat=TRUE, warning = FALSE, warning=FALSE}
 include_graphics(here("analysis/figures/ragnhildrod.png"))
@@
 
-```{r rodbol54, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r rodbol54, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/rodbol54.png"))
@@
 
-```{r rognlien, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r rognlien, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/rognlien.png"))
@@
 
-```{r sageneb1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r sageneb1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/sageneb1.png"))
@@
 
-```{r sageneb2, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r sageneb2, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/sageneb2.png"))
@@
 
-```{r solum3, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r solum3, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/solum3.png"))
@@
 
-```{r stokkep1, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r stokkep1, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/stokkep1.png"))
@@
 
-```{r stokkep5, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r stokkep5, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/stokkep5.png"))
@@
 
-```{r stokkep8, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r stokkep8, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/stokkep8.png"))
@@
 
-```{r torstvet, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r torstvet, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/torstvet.png"))
@@
 
-```{r trondalen, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r trondalen, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/trondalen.png"))
@@
 
-```{r tverdal, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r tverdal, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/tverdal.png"))
@@
 
-```{r vallermyrene1a, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r vallermyrene1a, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/vallermyrene1a.png"))
@@
 
-```{r vallermyrene1b, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r vallermyrene1b, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/vallermyrene1b.png"))
@@
 
-```{r vallermyrene2, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r vallermyrene2, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/vallermyrene2.png"))
@@
 
-```{r vallermyrene4a, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r vallermyrene4a, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/vallermyrene4a.png"))
@@
 
-```{r vallermyrene4b, echo = FALSE, out.width = "500px", regfloat=TRUE}
+```{r vallermyrene4b, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
 include_graphics(here("analysis/figures/vallermyrene4b.png"))
@@
 The DTM was edited to remove the railway track running by the site and a gravel pit.
+
+## Numerical result of re-dating

### 2022-05-01 analysis/paper/paper.Rmd
Setting up Supplementary table

@@
 ---
-title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
+title: "Simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
 header-includes:
@@
   A central premise for Norwegian Stone Age archaeology is that most coastal sites were located on or close to the contemporary shoreline when they were in use.
-By reconstructing the trajectory of relative sea-level change, this offers a dating method termed shoreline dating, separate from both radiometric and typological methods which has been widely applied.
+By reconstructing the trajectory of relative sea-level change, this offers a dating method termed shoreline dating separate from both radiometric and typological methods, which is widely applied.
 However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which, due to the relative rarity of the method, has been under limited scrutiny compared to more traditional dating techniques in archaeology.
@@
 The findings largely confirm previous evaluations of this relationship, indicting that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied.
-Drawing on the quantative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed, and compared to previous applications of the technique.
+Drawing on the quantative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.
 
@@
 
-The procedure for shoreline dating to be outlined is thus aimed at determining the likely age of the occupation of a site based on its altitude above present day sea-level, with reference to shoreline displacement and the likely elevation of the site above the sea-level when it was in use.
+The procedure for shoreline dating to be outlined is aimed at determining the likely age of the occupation of a site based on its altitude above present day sea-level, with reference to shoreline displacement and the likely elevation of the site above the sea-level when it was in use.
 For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly.
@@
 
-We can then get rid of parameter $D$ by summarising over all possible distances between site and the shoreline. Given its elevation, the probability of the date for the occupation of the site is then:
+We can then get rid of parameter $D$ by summarising over all possible distances between site and the shoreline. Given its elevation, the probability of the date of the occupation of the site is then:
 
@@
 
-An example implementation of this approach is given in Figure \@ref(fig:shoredate), where $\lambda$ = `r round(expfit$estimate, 3)`.
+An example implementation of this approach is given in Figure \@ref(fig:exampledat), where $\lambda$ = `r round(expfit$estimate, 3)`.
 This is the decay ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
-For the numerical implementation, $D$ is here defined as a sequence of increments of 0.001m from the site elevation $\alpha$.
+For the numerical implementation, $D$ is here defined as a sequence of increments of 0.001m, starting from the site elevation $\alpha$.
 In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date all of the sites from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes.

### 2022-05-01 analysis/supplementary-materials/supplementary.Rmd
Setting up Supplementary table

@@
 output: 
-  pdf_document:
+  bookdown::pdf_document2:
     toc: true
+    fig_caption: yes
+    reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+    pandoc_args:
+    - --lua-filter=../templates/scholarly-metadata.lua
+    - --lua-filter=../templates/author-info-blocks.lua
+    - --lua-filter=../templates/pagebreak.lua
 bibliography: references.bib
+csl: "../templates/saa.csl" # Insert path for the bib-style
 ---
@@
 rcarbon <- read.csv(here('analysis/data/raw_data/radiocarbon.csv'))
-```
-
-## Introduction
+
+# Read in site data for references in the final table
+sitdat <- read.csv(here('analysis/data/raw_data/sites.csv'))
+
+# Read in reported shoreline dates
+prevdates <- read.csv((here(
+  "analysis/data/raw_data/previous_shoreline_dates.csv"))) %>%
+  mutate(start = start * -1,
+         end = end * -1) %>%
+  dplyr::rename("site_name" = "name")
+
+# Load results of re-dating previously shoreline dated sites
+load(here("analysis/data/derived_data/08data.RData"))
+```
+
+## Introduction {-}
 
@@
 
-## Sensitivity analysis
+## Sensitivity analysis {-}
 
@@
 
-## Site analysis
+## Site analyses {-}
 
@@
 
-The earliest ^14^C-date does not match the Middle Mesolithic inventory from Ragnhildrød [@mjaerum2012].
+The earliest ^14^C-date, from an uncertain context, is argued to be older and not match the typological indicators in the inventory from Ragnhildrød [@mjaerum2012, 50, 71].
 The post-hole dated to the Late Neolithic/Bronze Age could be related to a single find of a pressure flaked object, but was not given much weight in the report [@mjaerum2012, 76].
@@
 
-## Numerical result of re-dating
+## Numerical result of re-dating {-}
+
+```{r}
+# Find 95% HDR for shoreline dates and reorder the sitesby median date 
+# (Taken directly from 08redate.R)
+
+bdates3 <- rbind(bdates, bdates2)
+
+hdrdat <- data.frame()
+for(i in 1:length(unique(bdates3$site_name))){
+  dat <-hdrcde::hdr(den = list("x" = bdates3[bdates3$site_name ==
+                                                unique(bdates3$site_name)[i],
+                                              "years"][[1]],
+                               "y" = bdates3[bdates3$site_name ==
+                                                unique(bdates3$site_name)[i],
+                                              "probability"][[1]]), prob = 95)
+  segdat <- data.frame(t(dat$hdr))
+  groupdat <- data.frame(matrix(nrow = length(segdat[seq(1, nrow(segdat), 2),]),
+                                ncol = 5))
+  names(groupdat) <- c("site_name", "start", "end", "group", "year_median")
+
+  groupdat$site_name <-  unique(bdates3$site_name)[i]
+  groupdat$start <- segdat[seq(1, nrow(segdat), 2), "X95."]
+  groupdat$end <- segdat[seq(2, nrow(segdat), 2), "X95."]
+  groupdat$group <- seq(1:length(dat$hdr[c(TRUE, FALSE)]))
+  groupdat$year_median <- dat$mode
+
+  hdrdat <- rbind(hdrdat, groupdat)
+}
+
+hdrdat <- hdrdat[order(hdrdat$year_median, decreasing = TRUE),]
+hdrdat$site_name <- factor(hdrdat$site_name, levels = unique(hdrdat$site_name))
+
+prevdates <- prevdates[order(match(prevdates$site_name, hdrdat$site_name)),]
+prevdates$site_name <- factor(prevdates$site_name,
+                            levels = unique(prevdates$site_name))
+prevdates$site_name <- as.character(prevdates$site_name)
+
+tabdat <- left_join(prevdates, hdrdat, by = "site_name")
+
+redat <- left_join(tabdat, sitdat, by = c("site_name" = "name")) %>% 
+  arrange(site_name) %>% 
+  group_by(site_name) %>% 
+  mutate(start95 = plyr::round_any(min(start.y), 10),
+         end95 = plyr::round_any(max(end.y), 10)) %>% 
+  # Replace minus sign/not minus sign with BCE/CE
+  mutate(start.x = ifelse(grepl("^[-]", start.x), 
+            gsub("^[-]", "", start.x), 
+            gsub("$", " CE", start.x)),
+         end.x = ifelse(grepl("^[-]", end.x), 
+            gsub("^[-](.*)", "\\1 BCE", end.x), 
+            gsub("$", " CE", end.x)),
+         end95 = ifelse(grepl("^[-]", end95), 
+            gsub("^[-](.*)", "\\1 BCE", end95), 
+            gsub("$", " CE", end95))) %>%
+  mutate(start95 = ifelse(str_detect(end95, "BCE"),
+                          ifelse(grepl("^[-]", start95), 
+            gsub("^[-](.*)", "\\1", start95), 
+            gsub("$", "", start95)),
+            ifelse(grepl("^[-]", start95), 
+            gsub("^[-](.*)", "\\1 BCE", start95), 
+            gsub("$", "CE", start95)))) %>% 
+  mutate(elev = paste0(min_elev, "\u2013", max_elev),
+         repdate = paste0(start.x, "\u2013", end.x),
+         hdrdate = paste0(start95, "\u2013", end95)) %>% 
+  select(site_name, elev, repdate, hdrdate, reference)
+
+redat[match(unique(redat$site_name), redat$site_name),] %>% 
+   rowid_to_column("id") %>% 
+   kable(caption = "Shoreline dating previously shorline dated sites. The range of the 95\\% HDR is rounded off to the nearest 10 years to reduce spurious precision.",
+         col.names = c("", 
+                     "Site",
+                    "Elevation",
+                    "Reported date",
+                    "95% HDR",
+                    "Reference"),
+        booktabs = TRUE,  linesep = c(""), escape = TRUE, longtable = TRUE)
+
+```
+
+\newpage
+
+## References {-}
+
+::: {#refs}
+:::

### 2022-05-01 analysis/script/08redate.R

@@ -298,8 +298,8 @@ for(i in 1:nrow(sites_sl)){
 bdates2 <- bind_rows(sitdates) %>% group_by(site_name) %>%
   filter(probability != 0)
 
-save(bdates, simsea, bdates2,
-     file = here("analysis/data/derived_data/08data.RData"))
+# save(bdates, simsea, bdates2,
+#      file = here("analysis/data/derived_data/08data.RData"))
 
 load(here("analysis/data/derived_data/08data.RData"))
 
@@ -348,7 +348,8 @@ redateplt <- ggplot(data = hdrdat, aes(x = year_median, y = site_name)) +
   #                          colour = "grey", fill = "grey") +
   geom_linerange(data = hdrdat, aes(xmin = start, xmax = end,
                                     y = site_name), col = "grey", size = 2.5) +
-# position = position_dodge(width = 0.4, preserve = 'single'), inherit.aes = FALSE
+# position = position_dodge(width = 0.4, preserve = 'single'),
+      # inherit.aes = FALSE
   geom_segment(data = psdates1, aes(x = start, xend = end, yend = site_name),
                size = 1, col = "red", alpha = 1) +
   geom_point(data = psdates2, aes(x = start), col = "red",

### 2022-05-03 analysis/paper/paper.Rmd
Minor edits for midway assessment

@@
 ---
-title: "Simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
+title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
 header-includes:
@@
 abstract: |
-  A central premise for Norwegian Stone Age archaeology is that most coastal sites were located on or close to the contemporary shoreline when they were in use.
-By reconstructing the trajectory of relative sea-level change, this offers a dating method termed shoreline dating separate from both radiometric and typological methods, which is widely applied.
-However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which, due to the relative rarity of the method, has been under limited scrutiny compared to more traditional dating techniques in archaeology.
+  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use.
+By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating that is separate from both radiometric and typological methods and which is widely applied.
+However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which, due to the geographical contingency and relative rarity of the method, has been under limited scrutiny compared to more ubiquitous dating techniques in archaeology.
 This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast.
-This is done by combining independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement through Monte Carlo simulation.
+This is done by means of Monte Carlo simulation, which is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement.
 The findings largely confirm previous evaluations of this relationship, indicting that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied.
-Drawing on the quantative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.
+Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.
 
@@
 This approach has a couple of limitations.
-First of all, the displacement curves are sometimes applied directly to larger study areas, with only some studies having taken the variable uplift-rates into account when performing this comparison [e.g.
+First of all, the displacement curves are sometimes applied directly to larger study areas, analogous to what @borreggine2022 term a bathtub model, with only some studies having taken the variable uplift-rates into account when performing this comparison [e.g.
 @åstveit2018; @fossum2020; @møller1987; @persson2008].
@@
 For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly.
-This leads the problem to become analogous to that of the calibration of a radiocarbon date [@bronkramsey2009, see also Figure \@ref(fig:exampledat)].
+This leads the problem to become analogous to that of the calibration of a radiocarbon date [@bronkramsey2009; @vanderplicht1993, see also Figure \@ref(fig:exampledat)].
 First, finding the elevation of the sea-level at the time the site was in use is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $D$.
@@
 where $\lambda$ is the decay ratio.
-This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $T$.
+This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $T$ for the occupation of the site.
 Here this is defined by a uniform probability density function over the range between the lower $T_{l}$ and upper $T_{u}$ bounds of the displacement curve that as been interpolated to the site location:
@@
 \begin{equation}
-p(T | \alpha - D) = p(T | \alpha - D)p(\alpha - D)
+p(T | \alpha - D) = p(T | \alpha - D)p(\alpha - D) \mbox{ [*This notation is a bit questionable*] }
   (\#eq:comb)
@@
 
-We can then get rid of parameter $D$ by summarising over all possible distances between site and the shoreline. Given its elevation, the probability of the date of the occupation of the site is then:
+We can then get rid of parameter $D$ by summarising over all possible distances between site and the shoreline. Given its elevation, the probability for the date of the occupation of the site is then:
 
@@
 When the range of the 95\\% HDR for age difference crosses zero, the shoreline and radiocarbon dates are considered to be in agreement.
-Line segments with vertical bars indicate the HDR does not cross zero and that the dates do not correspond.
+Line segments with vertical bars indicate that the HDR does not cross zero and that the dates do not correspond.
 The division and colour coding at the top of the plots reflect the divison of site phases given in Figure 9.", out.width = "525px", fig.align="center", regfloat=TRUE}
@@
 Some limitations and sources of likely variation and uncertainty that have not been considered should also be mentioned.
-First of all the sample size is quite strained and the future addition of more sites might alter the picture considerably.
+First of all the sample size is strained and the future addition of more sites might alter the picture considerably.
 Secondly, the validity of the method was evaluated by applying it to the data from where the input parameters were derived.

### 2022-05-03 analysis/supplementary-materials/supplementary.Rmd
Minor edits for midway assessment

@@
 ---
-title: "Supplementary material"
+title: "Supplementary material to the paper 'A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast'"
 author: "Isak Roalkvam"

### 2022-05-04 analysis/paper/paper.Rmd
Minor edits

@@
 By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating that is separate from both radiometric and typological methods and which is widely applied.
-However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which, due to the geographical contingency and relative rarity of the method, has been under limited scrutiny compared to more ubiquitous dating techniques in archaeology.
+However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which has been under limited scrutiny compared to more ubiquitous dating techniques due to its geographical contingency and relative rarity.
 This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast.
@@
 load(here("analysis/data/derived_data/06data.RData"))
-load(here("analysis/data/derived_data/07agediff.RData"))
+load(here("analysis/data/derived_data/08agediff.RData"))
 source(here("analysis/script/04functions.R"), local = knitr::knit_global())
@@
 The isobases have a direction of 327$^{\\circ}$ (Romundset et al.
-2018), B) Displacement curves.
+2018, but see Sørensen et al.
+2014), B) Displacement curves.
 Note the increasing steepness of the curves towards the north-east."), fig.height = 13, regfloat=TRUE}
@@
 This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $T$ for the occupation of the site.
-Here this is defined by a uniform probability density function over the range between the lower $T_{l}$ and upper $T_{u}$ bounds of the displacement curve that as been interpolated to the site location:
+Here this is defined by a uniform probability density function over the range between the lower $T_{l}$ and upper $T_{u}$ bounds of the displacement curve that has been interpolated to the site location:
 
@@
 \begin{equation}
-p(T | \alpha - D) = p(T | \alpha - D)p(\alpha - D) \mbox{ [*This notation is a bit questionable*] }
+p(T | \alpha - D) = p(T | \alpha - D)p(\alpha - D) \mbox{ [This notation is somewhat questionable] }
   (\#eq:comb)
@@
 
-We can then get rid of parameter $D$ by summarising over all possible distances between site and the shoreline. Given its elevation, the probability for the date of the occupation of the site is then:
+We can then get rid of parameter $D$ by summarising over all possible distances between site and the shoreline. Given its elevation, the probability for the date of the occupation of a site is then:
 

### 2022-05-04 analysis/script/04functions.R

@@ -1123,7 +1123,7 @@ shoreline_date <- function(sitename, elev = dtm,
     }
   }
 
-  inc <- seq(0, 80, reso)
+  inc <- seq(0, as.numeric(siteelev), reso)
 
   expdat <- data.frame(
     offset = inc,
@@ -1139,7 +1139,8 @@ shoreline_date <- function(sitename, elev = dtm,
   for(i in 1:nrow(expdat)){
     adjusted_elev <- as.numeric(siteelev - expdat$offset[i])
     if(!(adjusted_elev > 0)) {
-      adjusted_elev <- 0.01
+      adjusted_elev <- reso # Just to make sure sea-level is never
+                            #lower than present (unnecessary due to inc?)
     }
     # Find lower date, subtracting offset (defaults to 0)
     lowerd <- round(approx(sitecurve[,"lowerelev"],

### 2022-05-04 analysis/script/07shorelinedate.R

@@ -1,0 +1,128 @@
+library(tidyverse)
+library(patchwork)
+library(here)
+library(terra)
+library(sf)
+library(hdrcde)
+
+# For reproducibility
+set.seed(321)
+
+# Load required functions and prepared data
+source(here("analysis/script/04functions.R"))
+load(here("analysis/data/derived_data/01data.RData"))
+load(here("analysis/data/derived_data/02data.RData"))
+load(here("analysis/data/derived_data/06data.RData"))
+
+# Read in raster
+dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
+
+# Example site (effectively stepping through the code from shoreline_date in
+# 04.functions.R)
+sitename <- "Hegna vest 1"
+sitel <- filter(sites_sa, name == sitename)
+
+sitecurve <- interpolate_curve(years = xvals,
+                               target = sitel,
+                               dispdat = displacement_curves,
+                               isodat = isobases)
+sitecurve$name <- sitename
+siteelev <- terra::extract(dtm, vect(sitel), fun = mean)[2]
+
+inc <- seq(0, as.numeric(siteelev), 0.001)
+
+expdat <- data.frame(
+  offset = inc,
+  # Cumulative exponential distribution function
+  px = pexp(inc, rate = expfit$estimate)) %>%
+  # Probability for each elevation increment is given by subtracting the
+  # previous cumulative probability from the current
+  mutate(probs = px - lag(px, default =  dplyr::first(px))) %>%
+  tail(-1) %>%
+  filter(px < 0.99999) # Probability cut-off
+
+dategrid <- data.frame(
+  years = seq(-10000, 2000, 1),
+  probability = 0)
+
+for(i in 1:nrow(expdat)){
+  adjusted_elev <- as.numeric(siteelev - expdat$offset[i])
+  if(!(adjusted_elev > 0)) { # Do not allow sea-level to go below present
+    adjusted_elev <- 0.01
+  }
+  # Find lower date
+  lowerd <- round(approx(sitecurve[,"lowerelev"],
+                         xvals, xout = adjusted_elev)[['y']])
+
+  # Find upper date
+  upperd <- round(approx(sitecurve[,"upperelev"],
+                         xvals, xout = adjusted_elev)[['y']])
+
+  # Find youngest and oldest date
[…]

### 2022-05-04 analysis/script/08backwards_dating.R

@@ -1,0 +1,419 @@
+library(tidyverse)
+library(patchwork)
+library(here)
+library(terra)
+library(sf)
+library(hdrcde)
+
+# For reproducibility
+set.seed(321)
+
+# Load required functions and prepared data
+source(here("analysis/script/04functions.R"))
+load(here("analysis/data/derived_data/01data.RData"))
+load(here("analysis/data/derived_data/02data.RData"))
+load(here("analysis/data/derived_data/06data.RData"))
+
+# Read in raster
+dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
+
+# List all files except those starting with a number.
+# That is, all data files resulting from analysis in 05script.R
+datfiles <- grep("^[0-9]", list.files(here("analysis/data/derived_data")),
+                 invert = TRUE, value = TRUE)
+
+# Create empty list to hold dat data
+posterior_list <- list()
+
+# Loop over, load and assign them to the list
+for(i in 1:length(datfiles)){
+  # Load date data
+  load(file.path(here("analysis/data/derived_data", datfiles[i])))
+  posterior_list[[i]] <- output[!(names(output) %in%
+                                  c("", "sitel", "sitecurve"))]$datedat %>%
+    filter(rcarb_cor == "t") %>%
+    # filter(group == min(group)) %>%
+    # Make sure site name is included on all rows (sums have NA)
+    mutate(site_name = unique(na.omit(site_name)))
+}
+
+# Unpack list of lists
+rdates <- posterior_list %>% bind_rows()
+rdates <- rdates %>%  group_by(site_name, group) %>%
+  # If there is a sum, exclude the rest
+  filter(class == 'sum' | all(class != 'sum')) %>%
+  ungroup() %>%
+  filter(class != "aprior")
+
+# Exclude sites not analysed
+# (see supplementary material for more on this)
+sites_sa <- sites_sa %>%
+  filter(!(name %in% c("Dybdalshei 2", "Lunaveien", "Frebergsvik C")))
+
+# Specify for which sites to not shoreline date the site limit
+# (see supplementary material)
+feature_sites <- c("Langangen Vestgård 7", "Vallermyrene 2")
+
+# # Uncomment to rerun
+# shorelinedates <- list()
+# for(i in 1:nrow(sites_sa)){
+#   print(paste(i, sites_sa$name[i]))
+#   if(sites_sa$name[i] %in% feature_sites){
+#     shorelinedates[[i]] <- shoreline_date(sitename = sites_sa$name[i],
[…]

### 2022-05-04 analysis/script/08redate.R

@@ -1,364 +1,0 @@
-library(ggplot2)
-library(ggrepel)
-library(patchwork)
-library(ggnewscale)
-library(dplyr)
-library(magrittr)
-library(stringr)
-library(tidyr)
-library(here)
-library(raster)
-library(terra)
-library(sf)
-library(topoDistance)
-library(cowplot)
-
-source(here("analysis/script/04functions.R"))
-load(here("analysis/data/derived_data/01data.RData"))
-load(here("analysis/data/derived_data/02data.RData"))
-load(here("analysis/data/derived_data/06data.RData"))
-
-# Read in all site limits
-sitespol <- read_sf(
-  here('analysis/data/raw_data/site_limits/site_limits.gpkg'))
-
-# Read in previous shoreline dates and reported elevations
-prevdates <- read.csv((here(
-  "analysis/data/raw_data/previous_shoreline_dates.csv"))) %>%
-  mutate(start = start * -1,
-         end = end * -1) %>%
-  dplyr::rename("site_name" = "name")
-
-# Read in raster
-dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
-
-# Retrieve the Pauler sites
-brunlanes <- sitespol %>%
-  filter(str_detect(site_name, 'Pauler|Bakke|Sky 1')) %>%
-  st_zm() %>%
-  dplyr::rename("name" = "site_name",
-         "limit" = "name")
-
-brun_rep <- prevdates %>% filter(str_detect(site_name, 'Pauler|Bakke|Sky 1'))
-
-brunlanes <- full_join(brunlanes, brun_rep, by = c("name" = "site_name"))
-
-sdatesp <- list()
-for(i in 1:nrow(brunlanes)){
-  print(brunlanes$name[i])
-  sdatesp[[i]] <- shoreline_date(sitename = brunlanes$name[i],
-                                        elev = dtm,
-                                        disp_curves = displacement_curves,
-                                        sites = brunlanes,
-                                        iso = isobases,
-                                        expratio = expfit3$estimate,
-                                        siteelev = "min",
-                                        reso = 0.001,
-                                 specified_elev = brunlanes$min_elev[i])
-}
-
-bdates <- as_tibble(bind_rows(sdatesp))
-
-# Find 95 % probability range for shoreline dates and median shoreline date
-# for ordering in the plot
-hdrdat <- data.frame()
-for(i in 1:length(unique(bdates$site_name))){
[…]

### 2022-05-04 analysis/script/eaa.R

@@ -1,1151 +1,0 @@
-library(tidyverse)
-library(sf)
-library(ggridges)
-library(tmap)
-library(tmaptools)
-library(cowplot)
-library(terra)
-library(oxcAAR)
-library(topoDistance)
-library(nord)
-quickSetupOxcal()
-
-# This script is an absolute mess and will not run. The main components are
-# available in the other scripts and defintion of relevant functions mainly in
-# 04script.R
-
-source(here::here("analysis/script/02script.R"))
-load(here::here("analysis/data/derived_data/01data.RData"))
-load(here::here("analysis/data/derived_data/02data.RData"))
-load(here::here("analysis/data/derived_data/sp1.RData"))
-
-dtm1 <- rast("/home/isak/phd/eaa_presentation/dtm1/dtm1_33_120_109.tif") # LV1
-dtm2 <- rast("/home/isak/phd/eaa_presentation/dtm1/dtm1_33_119_108.tif")
-dtm3 <- rast("/home/isak/phd/eaa_presentation/dtm1/dtm_33_115_116_105.tif")
-dtm4 <- rast("/home/isak/phd/eaa_presentation/dtm1/vm1/dtm1_33_119_109.tif")
-dtm5 <- rast("/home/isak/phd/eaa_presentation/dtm1/trondal/trondal.tif")
-dtm1[dtm1 <= 0] <- NA
-
-# Tverdal, Sandvigen 1, Langangen Vestgård 1, Hesthag C2,
-# Stokke/Polland 1 (partly highway), Stokke/Polland 5 (highway)
-# Vallermyrene 1a
-
-countries <- st_read(here::here('analysis/data/raw_data/naturalearth_countries.gpkg'))
-
-# assign isobases to sites
-sites_sa <- st_join(st_make_valid(sites_sa), isopolys, join = st_intersects, largest = TRUE)
-site_pts <- st_centroid(sites_sa)
-
-
-model_rcarb <- function(data){
-
-  # Format individual dates for R_Date in Oxcal
-  dates <- R_Date(data[["lab_code"]], data[["c14_bp"]], data[["error"]])
-
-  # Manually compile rest of OxCal code
-  oxcal_code <- paste0('
-  Plot()
-   {
-    Sequence()
-    {
-     Boundary("Start");
-     Phase()
-     {
-    ',
-                       dates,
-                       '
-     Sum("Sum");
-     };
-     Boundary("End");
-    };
-  };')
-
[…]

### 2022-05-07 analysis/paper/paper.Rmd
Minor edits. Adding references to supplementary

@@
 By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating that is separate from both radiometric and typological methods and which is widely applied.
-However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which has been under limited scrutiny compared to more ubiquitous dating techniques due to its geographical contingency and relative rarity.
+However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which have been under limited scrutiny compared to those related to more ubiquitous dating techniques in archaeology.
+This is due to the geographical contingency and thus relative rarity of the method.
 This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast.
@@
 One of the more extensive evaluations of the relationship between archaeological radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
-They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
+They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
 However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates.
@@
 While there is displacement data available for other areas of south-eastern Norway [e.g.
-@hafsten1957; @sørensen1979; @sørensen1999; and recent compilation by @creel2022], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
+@hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
 This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
@@
 @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments.
-Each basin thus represent a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to devise a continuous time series for RSL-change, projected to a common isobase [see @creel2022, 5].
-Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible.
+Each basin thus represent a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to devise a continuous time series for RSL-change adjusted to a common isobase.
+To minimise the impact of variable uplift rates, the cored basins are therefore located in a as constrained area of the landscape as possible.
 Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from the common isobase perpendicular to this gradient (Figure \@ref(fig:map-iso)).
@@
 This is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
-The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021, 30cm at the standard port Helgeroa in Larvik].
+The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021 30cm at the standard port Helgeroa in Larvik].
 Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
 The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
-@romundset2011; @romundset2019; for an alternative approach see @creel2022].
+@romundset2011; @romundset2019; for alternative approaches see e.g.
+@barnett2020; @creel2022].
 For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.
@@
 ```{r example-3, echo = FALSE, fig.cap = "The result of 1000 simulation runs for each of the two groups of dates on the site Hegna vest 1.
-The first column of plots shows the radiocarbon probability density function from where dates were drawn during simulation.
+The first column of plots shows the calibrated radiocarbon probability distribution from where dates were drawn during simulation.
 The second column displays the result of simulating the raised sea-level 1000 times.
@@
 From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
-There are also a couple of sites located some distance from the shoreline just after 4000 BCE.
-While the sample size is limited, this would thus be in line with a development that sees an increase in settlements located in the immediate inland around this time.
+While the sample size is limited, there are also a couple of sites located some distance from the shoreline just after 4000 BCE.
+That the findings appear to be off from the chronological framework by around a century must be seen in relation to chronological smearing from the uncertainty in the ^14^C-dates, and the findings are thus in clear agreement with the literature.
 
@@
 For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly.
-This leads the problem to become analogous to that of the calibration of a radiocarbon date [@bronkramsey2009; @vanderplicht1993, see also Figure \@ref(fig:exampledat)].
+This leads the problem to become analogous to that of the calibration of a radiocarbon date [@bronkramsey2009; @stuvier1989; @vanderplicht1993, see also Figure \@ref(fig:exampledat)].
 First, finding the elevation of the sea-level at the time the site was in use is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $D$.
@@
 This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $T$ for the occupation of the site.
-Here this is defined by a uniform probability density function over the range between the lower $T_{l}$ and upper $T_{u}$ bounds of the displacement curve that has been interpolated to the site location:
+This is defined by a uniform probability density function over the range between the lower $T_{l}$ and upper $T_{u}$ bounds of the displacement curve that has been interpolated to the site location:
 
@@
 
-Finding the probability for the date of the site then becomes a matter of transferring the probability of the distance between site and shoreline and that of the displacement curve to calendar dates:
+Finding the probability for the date of the site then becomes a matter of transferring the probability of the distance between site and shoreline to calendar dates using the displacement curve:
 
@@
 
-An example implementation of this approach is given in Figure \@ref(fig:exampledat), where $\lambda$ = `r round(expfit$estimate, 3)`.
+An example of an implementation of the outlined approach is given in Figure \@ref(fig:exampledat), where $\lambda$ = `r round(expfit$estimate, 3)`.
 This is the decay ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
-For the numerical implementation, $D$ is here defined as a sequence of increments of 0.001m, starting from the site elevation $\alpha$.
+For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, starting from the site elevation $\alpha$ and extending down to the present sea-level.
+As a note, this approach is analogous to Procedure 1 as described by @stuvier1989 for the calibration of ^14^C-dates.
+Given the monotonic nature of the displacement curves there is no issue with multiple intercepts here, which usually comes up in the calibration of a ^14^C-date duo to the wiggly character of the calibration curve.
 In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date all of the sites from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes.
@@
 The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by @parnell2008.
-Here, 100,000 age samples drawn from the probability density function of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding modelled ^14^C-dates.
+Here, 100,000 age samples drawn from the probability distribution of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding modelled ^14^C-dates.
 The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement (Figure \@ref(fig:shoredatediff)).
@@
 Above it was suggested that additional temporal data could be combined with the method to improve its accuracy and precision.
-This example, on the other hand, highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
+Drawing on @jaksland2014, this example instead highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
 One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline.
@@
 B) Map showing the centroids of the Pauler sites and Sky 1.
-The sea-level has been simulated using the probability density associated with the shoreline date for Pauler 1 (see also map in Jaksland 2014:fig.12a).
+The sea-level has been simulated using the probability distribution associated with the shoreline date for Pauler 1 (see also map in Jaksland 2014:fig.12a).
 Pauler 1 is the red point.", out.width = "525px", fig.align="center", regfloat=TRUE}
@@
 First of all the sample size is strained and the future addition of more sites might alter the picture considerably.
-Secondly, the validity of the method was evaluated by applying it to the data from where the input parameters were derived.
+Secondly, the validity of the outlined method was evaluated by applying it to the data from where the input parameters were derived.
 Fitting and evaluating a model using the exact same data will likely exaggerate its performance.
@@
 Fifthly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves further away from the isobases of the displacement curves.
-This is also related to the fact that the RSL data can be handled in different ways than with the isobase method that has been used for the compilation of the employed displacement curves [cf.
-@creel2022].
+This is also related to the fact that the RSL data can be handled in different ways than with the isobase method that has been used for the compilation of the employed displacement curves.
 Sixthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration [cf.
@@
 Furthermore, formulating and visualising the method along the lines of how radiocarbon dates are treated, as was done here, does stand the chance of giving a veneer of radiometric accuracy that is not warranted.
-That being said, the best chance we have of not throwing away precious temporal data, or exaggerating our handle on it, is arguably to rigorously evaluate the method using independent data such as radiocarbon dates, by offering a precise formulation of how it could be applied, by specifying what sources of uncertainty are accounted for and by making this process open through the dissemination of underlying data and programming code.
+That being said, the best chance we have of not throwing away precious temporal data, or exaggerating our handle on it, is arguably to rigorously evaluate the method using independent data such as radiocarbon dates, by offering a precise formulation of how it could be applied, by specifying what sources of uncertainty are accounted for and by making this process open through the open dissemination of underlying data and programming code.
 

### 2022-05-07 analysis/supplementary-materials/supplementary.Rmd
Minor edits
Minor edits. Adding references to supplementary

--- 2022-05-01 analysis/supplementary-materials/supplementary.Rmd
@@
 csl: "../templates/saa.csl" # Insert path for the bib-style
+nocite: | 
+  @*
 ---
@@
 # Load results of re-dating previously shoreline dated sites
-load(here("analysis/data/derived_data/08data.RData"))
+load(here("analysis/data/derived_data/09data.RData"))
 ```
@@
 
+<!-- @nyland2012b -->
+<!-- @danielsen2018 -->
+<!-- @koxvold2018 -->
+<!-- @solheim2017 -->
+<!-- @granum2018 -->
+<!-- @reitan2014l -->
+<!-- @reitan2014f -->
+<!-- @carrasco2014 -->
+<!-- @carrasco2014 -->
+<!-- @fossum2014b -->
+<!-- @fossum2014c -->
+<!-- @koxvold2017l -->
+<!-- @koxvold2017n -->
+<!-- @havstein2017a -->
+<!-- @havstein2017 -->
+<!-- @koxvold2017h -->
+<!-- @eigeland2017a -->
+<!-- @koxvold2013 -->
+<!-- @koxvold2017d -->
+<!-- @koxvold2017e -->
+<!-- @koxvold2016 -->
+<!-- @koxvold2017m -->
+<!-- @koxvold2017f -->
+<!-- @koxvold2017g -->
+<!-- @koxvold2020a -->
+<!-- @stokke2018a -->
+<!-- @reitan2018a -->
+<!-- @stokke2018a -->
+<!-- @nielsen2017 -->
+<!-- @nielsen2017 -->
+<!-- @darmark2018c -->
+<!-- @viken2018a -->
+<!-- @darmark2018b -->
+<!-- @eymundsson2014 -->
+<!-- @eggen2014b -->
+<!-- @reitan2014h -->
+<!-- @mansrud2018 -->
+<!-- @persson2008 -->
+<!-- @eigeland2014a -->
+<!-- @eigeland2014 -->
+<!-- @fossum2014a -->
+<!-- @olsen2013a -->
+<!-- @koxvold2013a -->
+<!-- @mansrud2013c -->
+<!-- @hårstad2021 -->
+<!-- @berge2012 -->
+<!-- @schallerahrberg2012 -->
+<!-- @nyland2012 -->
+<!-- @amundsen2012 -->
+<!-- @nyland2012a -->
+<!-- @amundsen2012a -->
+<!-- @jaksland2012 -->
+<!-- @jaksland2012a -->
+@carrasco2015
+@carrasco2015
+@persson2014a
+@hårstad2021b
+@koxvold2020
+@viken2018
+@darmark2018a
+@darmark2018
+@darmark2018
+@nielsen2017a
+@ekstrand2013
+
 \newpage

### 2022-05-10 analysis/paper/paper.Rmd
Minor edits and adding some references

@@
   A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use.
-By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating that is separate from both radiometric and typological methods and which is widely applied.
-However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which have been under limited scrutiny compared to those related to more ubiquitous dating techniques in archaeology.
-This is due to the geographical contingency and thus relative rarity of the method.
+By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied.
+However, while the potentially immense benefits of an additional source of temporal data in addition to radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiqutous dating techniques in archaeology.
 This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast.
@@
 This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g.
-@akerlund1996; @bjerck2005; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @wikell2009].
+@akerlund1996; @bjerck2005; @loken1977; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @wikell2009].
 
@@
 This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020].
-In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963].
+In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963; @johansen1997].
 

### 2022-05-10 analysis/supplementary-materials/supplementary.Rmd
Minor edits and adding some references

@@
 
-Stone Age sites for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975], Nedre Holtan [@sjursjeike], Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@østmo1975].
+Stone Age sites for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975], Nedre Holtan [@sjurseike1991; @glorstad1998], Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@ostmo1976].
 These sites are therefore not included in the analysis.
@@
 
-Rognlien was originally excavated in [@ingstad1970].
-The site limit has been manually defined based on the description given in the report and the spatial geometry of the site as defined in the database Askeladden.
-The manually defined limit could be slightly off, but should match the elevation of the site given in the publication [@ingstad1970].The ^14^C-dates from the original excavation was not seen as contemporaneous with the lithic material on the site.
+The first archaeological finds from Rognlien were done in the 1880s and, following some smaller investigations, was excavated by @ingstad1970 in 1957.
+The site limit has been manually defined based on the description given in the publication and the spatial geometry of the site as defined in the database Askeladden.
+The manually defined limit could be slightly off, but should match the elevation of the site given in the publication [@ingstad1970, 26].
+The ^14^C-dates from the original excavation was not seen as contemporaneous with the lithic material on the site.
 Burnt elk bone from the original excavation was therefore recently subjected to radiocarbon dating and is believed to match the occupation of the site [pers.comm.
@@
 <!-- @jaksland2012a -->
-@carrasco2015
-@carrasco2015
-@persson2014a
-@hårstad2021b
-@koxvold2020
-@viken2018
-@darmark2018a
-@darmark2018
-@darmark2018
-@nielsen2017a
-@ekstrand2013
+<!-- @carrasco2015 -->
+<!-- @carrasco2015 -->
+<!-- @persson2014a -->
+<!-- @hårstad2021b -->
+<!-- @koxvold2020 -->
+<!-- @viken2018 -->
+<!-- @darmark2018a -->
+<!-- @darmark2018 -->
+<!-- @darmark2018 -->
+<!-- @nielsen2017a -->
+<!-- @ekstrand2013 -->
+<!-- @ekstrand2013 -->
+<!-- @ekstrand2013 -->
+<!-- @amundsen2012b -->
+<!-- @fossum2014 -->
+<!-- @fossum2017c -->
+<!-- @mansrud2016 -->
+<!-- @koxvold2017j -->
+<!-- @fossum2017d -->
+<!-- @eggen2014 -->
+<!-- @melvold2014a -->
+<!-- @stokke2017 -->
+<!-- @stokke2017 -->
+<!-- @koxvold2017c -->
+<!-- @koxvold2017b -->
+<!-- @koxvold2017a -->
+<!-- @reitan2021 -->
+<!-- @reitan2021 -->
+<!-- @reitan2014i -->
 

### 2022-05-12 analysis/paper/paper.Rmd
Minor edits, sent off to Anders

@@
 By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied.
-However, while the potentially immense benefits of an additional source of temporal data in addition to radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiqutous dating techniques in archaeology.
+However, while the potentially immense benefits of an additional source of temporal data seperate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiqutous dating techniques in archaeology.
 This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast.
@@
 @romundset2011; @romundset2019; for alternative approaches see e.g.
-@barnett2020; @creel2022].
+@barnett2020; @cahill2016; @creel2022].
 For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.
@@
 This is the decay ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
-For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, starting from the site elevation $\alpha$ and extending down to the present sea-level.
+For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, starting from the site elevation $\alpha$ and extending down to the present sea-level, that is until $\alpha - D = 0$.
 As a note, this approach is analogous to Procedure 1 as described by @stuvier1989 for the calibration of ^14^C-dates.
-Given the monotonic nature of the displacement curves there is no issue with multiple intercepts here, which usually comes up in the calibration of a ^14^C-date duo to the wiggly character of the calibration curve.
+Given the monotonic nature of the displacement curves there is no issue with multiple intercepts here, which usually comes up in the calibration of a ^14^C-date due to the wiggly character of the calibration curve.
 In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date all of the sites from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes.

### 2022-05-19 analysis/paper/paper.Rmd
Minor edits

@@
 Comparatively, given a continuous and still ongoing shoreline regression from as high as c.
-220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979].
+220m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979].
 This makes the region especially useful for evaluating the assumption of a shore-bound settlement pattern over a long and continuous time-span.

### 2022-06-09 analysis/paper/paper.Rmd
Had to correct for a mix-up with the dates from Gunnarsrød 5

@@
 This is done by means of Monte Carlo simulation, which is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement.
-The findings largely confirm previous evaluations of this relationship, indicting that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied.
+The findings largely confirm previous evaluations of this relationship, indicating that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied.
 Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

### 2022-06-09 analysis/script/07shorelinedate.R

@@ -124,5 +124,5 @@ dplt <- plt + geom_polygon(data = expdatt, aes(x =  probs_scaled,
   geom_hline(yintercept = as.numeric(siteelev), linetype = "dashed",
              col = "#046c9a", alpha = 0.6)
 
-ggsave(file = here("analysis/figures/sdate2.png"), dplt,
+ggsave(file = here("analysis/figures/sdate.png"), dplt,
        width = 100, height = 100, units = "mm")

### 2022-06-10 analysis/paper/paper.Rmd
Minor edit

@@
 When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio of `r round(expfit$estimate, 3)` (Figure \@ref(fig:redate)A) to characterise this relationship.
-Furthermore, while this remains to be formalised and explored further, it was also showed how the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data.
+Furthermore, while this remains to be formalised and explored further, it was also showed how the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data.
 As the precision of the method is both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date, the impact of such additional information will also vary.

### 2022-06-22 analysis/paper/paper.Rmd
Re-ran redating with Melau and Austein included

@@
 This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g.
-@akerlund1996; @bjerck2005; @loken1977; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @wikell2009].
+@akerlund1996; @bjerck2005; @gjerde2021; @loken1977; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @wikell2009].
 
@@
 This is the decay ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
-For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, starting from the site elevation $\alpha$ and extending down to the present sea-level, that is until $\alpha - D = 0$.
+For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, starting from the site elevation $\alpha$ and extending down to the present sea-level, that is, until $\alpha - D = 0$.
 As a note, this approach is analogous to Procedure 1 as described by @stuvier1989 for the calibration of ^14^C-dates.
@@
 Furthermore, formulating and visualising the method along the lines of how radiocarbon dates are treated, as was done here, does stand the chance of giving a veneer of radiometric accuracy that is not warranted.
-That being said, the best chance we have of not throwing away precious temporal data, or exaggerating our handle on it, is arguably to rigorously evaluate the method using independent data such as radiocarbon dates, by offering a precise formulation of how it could be applied, by specifying what sources of uncertainty are accounted for and by making this process open through the open dissemination of underlying data and programming code.
+That being said, the best chance we have of not throwing away precious temporal data, or exaggerating our handle on it, is arguably to rigorously evaluate the method using independent data such as radiocarbon dates, by offering a precise formulation of how it could be applied, by specifying what sources of uncertainty are accounted for and by making this process transparent through the open dissemination of underlying data and programming code.
 

### 2022-06-29 analysis/paper/paper.Rmd
Minor edits

@@
 For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly.
-This leads the problem to become analogous to that of the calibration of a radiocarbon date [@bronkramsey2009; @stuvier1989; @vanderplicht1993, see also Figure \@ref(fig:exampledat)].
+This leads the problem to become analogous to that of the calibration of a radiocarbon date [see Figure \@ref(fig:exampledat), @bronkramsey2009; @stuvier1989; @vanderplicht1993].
 First, finding the elevation of the sea-level at the time the site was in use is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $D$.
@@
 For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, starting from the site elevation $\alpha$ and extending down to the present sea-level, that is, until $\alpha - D = 0$.
-As a note, this approach is analogous to Procedure 1 as described by @stuvier1989 for the calibration of ^14^C-dates.
-Given the monotonic nature of the displacement curves there is no issue with multiple intercepts here, which usually comes up in the calibration of a ^14^C-date due to the wiggly character of the calibration curve.
+
 In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date all of the sites from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes.
@@
 Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date.
-The 200 year uncertainty range is given if the sites were to be considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other.
+The 200 year uncertainty range is given if the sites were to be considered in isolation, while his argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other.
 Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap.
 This information is not integrated in the approach outlined here, but could justify further reducing the uncertainty ranges.
+
 Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does result in very precise dates.

### 2022-06-29 analysis/script/08backwards_dating.R

@@ -83,8 +83,8 @@ for(i in 1:nrow(sites_sa)){
   }
 }
 
-save(shorelinedates,
-     file = here("analysis/data/derived_data/08data.RData"))
+# save(shorelinedates,
+#      file = here("analysis/data/derived_data/08data.RData"))
 
 load(here("analysis/data/derived_data/08data.RData"))
 
@@ -189,8 +189,8 @@ shrplt <- ggplot(data = hdrdat,
 ggsave(file = here("analysis/figures/shoredate.png"), shrplt,
        width = 200, height = 250, units = "mm")
 
-# Next sections finds synchroneity between radiocarbon and shoreline dates
-# Grab the probablities again without filtering for plotting
+# Next section finds synchroneity between radiocarbon- and shoreline dates
+# Grab the probabilities again without filtering for plotting
 shorelinedates <- bind_rows(shorelinedates) %>%
   filter(site_name %in% unique(corsites$site_name))
 
@@ -225,8 +225,8 @@ synchdata <- synchdt  %>% group_by(cross) %>%
   ungroup() %>%
   mutate(row_total = rowSums(across(where(is.numeric))))
 
-# save(synchdata,
-#      file = here("analysis/data/derived_data/08agediff.RData"))
+save(synchdata,
+     file = here("analysis/data/derived_data/08agediff.RData"))
 
 hdr_func <- function(x) {
   r <- hdrcde::hdr(x, prob = 95)

### 2022-07-12 analysis/paper/paper.Rmd
Updated RStudio and packages

@@
 
-# Set up so that I can call floatbarrier using regfloat=TRUE in the chunk 
+# Set-up so that I can call floatbarrier using regfloat=TRUE in the chunk 
 # headings. See: 
@@
 For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly.
-This leads the problem to become analogous to that of the calibration of a radiocarbon date [see Figure \@ref(fig:exampledat), @bronkramsey2009; @stuvier1989; @vanderplicht1993].
+This leads the problem to become similar to that of the calibration of a radiocarbon date [see Figure \@ref(fig:exampledat), @bronkramsey2009; @stuvier1989; @vanderplicht1993].
 First, finding the elevation of the sea-level at the time the site was in use is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $D$.
@@
 This is the decay ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
-For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, starting from the site elevation $\alpha$ and extending down to the present sea-level, that is, until $\alpha - D = 0$.
+For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, starting from the site elevation $\alpha$.
+The exponential function is stepped through in it's cumulative form, where the probability from the previous 0.001m step is subtracted from the probability at the current step.
+This probability is then divided equally across the individual calendar years in the range between the lower and the upper limit of the displacement curve at the current 0.001m step.
+The histogram that is the resulting shoreline date is the sum of performing this procedure on all possible 0.001m values of $D$, which in practice is until $\alpha - D = 0$ or when 99.999% of the exponential function has been covered.
 

### 2022-07-12 analysis/script/05site_analysis.R

@@ -17,6 +17,11 @@ quickSetupOxcal()
 # source("http://bioconductor.org/biocLite.R")
 # biocLite("IRanges")
 
+#or:
+# if (!require("BiocManager", quietly = TRUE))
+#   install.packages("BiocManager")
+# BiocManager::install("IRanges")
+
 # For reproducibility
 set.seed(123)
 

### 2022-07-25 analysis/paper/paper.Rmd
Issues with example_simulation result

@@
 author: 
-  - Isak Roalkvam:
-      email: isak.roalkvam@iakh.uio.no
-      institute: [UiO]
-      correspondence: true
-institute:
-  - UiO: University of Oslo, Institute of Archaeology, Conservation and History
+- 'Isak Roalkvam'
+- 'University of Oslo, Institute of Archaeology, Conservation and History'
+  # - Isak Roalkvam:
+  #     email: isak.roalkvam@iakh.uio.no
+  #     institute: [UiO]
+  #     correspondence: true
+# institute:
+#   - UiO: University of Oslo, Institute of Archaeology, Conservation and History
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
       pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
+      # - --lua-filter=../templates/scholarly-metadata.lua
+      # - --lua-filter=../templates/author-info-blocks.lua
       - --lua-filter=../templates/pagebreak.lua
@@
 \begin{equation}
-p(T | \alpha - D) = p(T | \alpha - D)p(\alpha - D) \mbox{ [This notation is somewhat questionable] }
+p(T | \alpha - D) = p(T | \alpha - D)p(\alpha - D)
   (\#eq:comb)

### 2022-07-26 analysis/script/06simulation_results.R
Final adjustments before publishing pre-print

@@ -29,7 +29,7 @@ for(i in 1:length(datfiles)){
 agr <- c("Nordby 1", "Kvastad A2", "Nauen A")
 
 # results_list is a list of lists of variable length that need to be unpacked.
-# (Might be a smoother purrr solution to do all of this?)
+# (Probably a smoother purrr solution to do all of this?)
 results <- list()
 for(i in 1:length(results_list)){
   dat <- results_list[[i]]
@@ -73,9 +73,12 @@ sum_cor <- distances %>% filter(rcarb_cor == "t") %>%
                c(mean, median, sd, IQR)) %>%
    round()
 
-colsh = c("forager" = "#00ba38", "agricultural" = "darkgrey")
-colst = c("forager" = "#fad510", "agricultural" = "darkgrey")
-colsv = c("forager" = "steelblue", "agricultural" = "darkgrey")
+# Colour for horisontal distance
+colsh = c("forager" = "#00ba38", "agricultural" = "grey30")
+# Colour for topographic distance
+colst = c("forager" = "yellow3", "agricultural" = "grey30") # "#fad510"
+# Colour for vertical distance
+colsv = c("forager" = "#046c9a", "agricultural" = "grey30") # "steelblue"
 
 # Horizontal distance, all dates to the Stone Age
 h1 <- ggplot(distances, aes(x = year, y = hordist,
@@ -397,7 +400,7 @@ grd <- plot_grid(grd3, grd4, labels = c("A", "B"), ncol = 1, scale = 0.9)
 
 # Save plot
 ggsave(file = here("analysis/figures/results2.png"), grd,
-       bg = "white", width = 250, height = 225, units = "mm")
+       bg = "white", width = 250, height = 200, units = "mm")
 
 # Save results and exponential functions
 save(distances, expfit, expfit2, expfit3,

### 2022-08-02 analysis/paper/paper.Rmd
Final adjustments before publishing pre-print
Minor edits

@@
 By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied.
-However, while the potentially immense benefits of an additional source of temporal data seperate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiqutous dating techniques in archaeology.
+However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiquitous dating techniques in archaeology.
 This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast.
@@
 Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
-This is for example reflected in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g.
-@berg-hansen2009; @eskeland2017].
+This is for example reflected in surveys, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g.
+@berg-hansen2009; @eskeland2017; @nummedal1923].
 Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date-span is too wide to decisively verify this.
@@
 
-Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty.
-The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions.
+Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with temporal uncertainty.
+The first task was therefore to ascribe a likely date and associated uncertainty to these dimensions.
 To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves, using inverse distance weighting [e.g.

### 2022-08-02 analysis/paper/paper.Rmd
Minor edits

@@
 The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m [@mapping2018].
-All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a digital research compendium following @marwick2018.
+All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a research compendium following @marwick2018.
 
@@
 The result of this process is shown for an example site in Figure \@ref(fig:example-1).
-For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
+For the the sites all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
 Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same event, here termed settlement or site phase.
@@
 As the data collection for the employed DTM was begun by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has sometimes been severely impacted by disturbances after the excavation.
-In addition to employing 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster.
+In addition to employing the 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster.
 This involved manually defining the extent of problem areas such as railways, highways, quarries and the like.
@@
 @conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021].
-In addition to code and original spatial data being available in the digital research compendium for this paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.
+In addition to code and original spatial data being available in the research compendium for this paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.
 

### 2022-08-06 analysis/paper/paper.Rmd
Minor edits

@@
 Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g.
-@berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018].
+@berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @olsen1984; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018].
 The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.

### 2022-08-06 analysis/supplementary-materials/supplementary.Rmd
Minor edits

@@
 title: "Supplementary material to the paper 'A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast'"
-author: "Isak Roalkvam"
-date: "10/27/2021"
+author: 
+- 'Isak Roalkvam'
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 header-includes:
@@
     pandoc_args:
-    - --lua-filter=../templates/scholarly-metadata.lua
-    - --lua-filter=../templates/author-info-blocks.lua
+    # - --lua-filter=../templates/scholarly-metadata.lua
+    # - --lua-filter=../templates/author-info-blocks.lua
     - --lua-filter=../templates/pagebreak.lua
@@
 
-The earliest ^14^C-date, from an uncertain context, is argued to be older and not match the typological indicators in the inventory from Ragnhildrød [@mjaerum2012, 50, 71].
+The earliest ^14^C-date, from an uncertain context, is argued to be younger and not match the typological indicators in the inventory from Ragnhildrød [@mjaerum2012, 50, 71].
 The post-hole dated to the Late Neolithic/Bronze Age could be related to a single find of a pressure flaked object, but was not given much weight in the report [@mjaerum2012, 76].

### 2022-09-01 analysis/paper/paper.Rmd
Minor edits after read-through

@@
 By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied.
-However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiquitous dating techniques in archaeology.
+However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more established dating techniques in archaeology.
 This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast.
-This is done by means of Monte Carlo simulation, which is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement.
-The findings largely confirm previous evaluations of this relationship, indicating that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied.
+Monte Carlo simulation is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement.
+The findings largely confirm previous hypotheses that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied.
 Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.
@@
 The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology.
-This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
+This follows not only from the dramatic changes to the landscape that this process created throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.
 This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g.
@@
 
-The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was established by the end of the 19th century [@degeer1896], and was first applied as a dating method at the turn of the century [@brogger1905; @hollender1901].
+The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was proposed at the end of the 19th century [@degeer1896], and it was first applied as a dating method at the turn of the century [@brogger1905; @hollender1901].
 Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g.
@@
 The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.
-However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise method by which one can hope to date the sites.
+However, given the coarse and fuzzy resolution of established typological frameworks, the vast number of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise method by which one can hope to date the sites.
 Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age.
@@
 
-Despite its important role for Norwegian Stone Age archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods.
+Despite its important role for Fennoscandian archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods.
 The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach.
-The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age, and in turn have this inform an improved method for shoreline dating.
+The goal here is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age in a relatively well sampled portion of Scandinavia, and in turn have this quantification inform the development of a formalised method for shoreline dating.
 As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions.
@@
 Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g.
-@bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here.
+@bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here and adapted to the parameters of post-glacial sea-level change and the Stone Age settlement of southern Norway.
 
@@
 @milne2009].
-Of central importance here is eustasy and istostasy.
-The eustatic sea-level is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015].
+Of central importance here is eustasy and isostasy.
+Eustatic sea-level is understood to be the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015].
 The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans [@mörner1976].
-Istostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle.
-This can be the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed.
-These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight [@milne2015].
+Isostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle caused by mass loading and unloading, which occurs with glaciation and deglaciation.
+These effects causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight [@milne2015].
 
 Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g.
-@hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g.
+@hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has caused most areas of Norway to have been subjected to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g.
 @mörner1979; @svendsen1987].
 In other words, the RSL has been dropping throughout prehistory.
-As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet.
-Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g.
+As this process is the result of glacial loading, the rate of uplift is faster towards the centre of the ice sheet relative to the distal aspects.
+Thus, there is differential glacio-isostatic impact to a site’s location depending on it's relation to the ice sheet’s centre of mass, leading some areas on the outer coast to have had a more stable RSL or been subject to marine transgression [e.g.
 @romundset2015; @svendsen1987].
 These conditions are directly reflected in the archaeological record.
-In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement phases that are difficult to disentangle [e.g.
+In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time spans, creating a mix of settlement phases that are difficult to disentangle [e.g.
 @hagen1963; @reitan2009].
-Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, leave them still submerged today [@bjerck2008; @glorstad2020].
-This can lead to a hiatus in the archaeological record for certain sub-phases in the impacted areas.
+Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, submerge them [@bjerck2008; @glorstad2020].
+Transgression can therefore lead to a hiatus in the archaeological record for certain sub-phases in the impacted areas despite the fact that there were likely coastal settlements during the inferred hiatuses.
 Comparatively, given a continuous and still ongoing shoreline regression from as high as c.
-220m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979].
-This makes the region especially useful for evaluating the assumption of a shore-bound settlement pattern over a long and continuous time-span.
+220m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979].
+This makes the region especially useful for evaluating the assumption of a shore-bound settlement pattern over a long and continuous time span.
 
 ```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from c.
-17--8 kyr BCE.
+17--8 thousand years BCE.
 The study area defined later in the text is marked with a red outline (deglaciation data from Hughes et al.
@@
 Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time.
-This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020].
-In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963; @johansen1997].
+This is done by aggregating point estimates of shoreline dates in 100-, 200- or 500-year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020].
+In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963; @johansen1997; @mikkelsen1975a, 100].
 
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; @bjerck2017; also discussed by @berg-hansen2009; @bergsvik2009].
-The same logic has also been extended to the hinter- and inland regions, where sites are to be predominantly located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020].
+The same logic has also been extended to the hinterland and inland regions, where sites are belived to be predominantly located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@prescott2020; cf.
 @solheim2021].
-The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm is to have led site locations to be more withdrawn from the shoreline [e.g.
+The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the development of settled farmsteads is to have led site locations to be more withdrawn from the shoreline [e.g.
 @bakka1971; @prescott2020; @ostmo2008, 223].
@@
 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region.
-However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227].
+However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation and return to foraging in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227].
 @nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g.
@@
 Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
-This is for example reflected in surveys, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g.
+This is for example reflected in archaeological survey practices, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g.
 @berg-hansen2009; @eskeland2017; @nummedal1923].
-Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date-span is too wide to decisively verify this.
-It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g.
+Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date span is too wide to decisively verify this.
+It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would appear to have been sensible if the site was not shore bound based on an archaeological understanding [e.g.
 @jaksland2014; @johansen1963; @nummedal1923].
 This can for example pertain to accessibility.
-If the site is situated on a ledge in a steep and jagged area of the present day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft.
+If the site is situated on a ledge in a steep and jagged area of the present-day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft.
 Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @ilves2011].
@@
 Not remnants of stays as ephemeral to only be discernible by isolated features or dubious ^14^C-dates.
-The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.
+The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent upon, and follow the discretion of the original excavation reports.
 
@@
 This approach has a couple of limitations.
-First of all, the displacement curves are sometimes applied directly to larger study areas, analogous to what @borreggine2022 term a bathtub model, with only some studies having taken the variable uplift-rates into account when performing this comparison [e.g.
+First, the displacement curves are sometimes applied directly to larger study areas, analogous to what @borreggine2022 term a bathtub model, with only some studies having taken the variable uplift-rates into account when performing this comparison [e.g.
 @åstveit2018; @fossum2020; @møller1987; @persson2008].
-Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis.
+Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that the conclusion supports the hypothesis.
 This thus leads to an inferential framework that favours uncertainty, which is hardly desirable.
-In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same.
+In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are actually the same.
 The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g.
@@
 RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater.
-Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and help move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.
+Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation and help move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.
 
 # Data
-To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on a study area with good control of the trajectory of prehistoric shoreline displacement.
+To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on identifying a study area with good control of the trajectory of prehistoric shoreline displacement.
 While there is displacement data available for other areas of south-eastern Norway [e.g.
@@
 
-The employed shoreline displacement data is based on the so-called isolation basin method [e.g.
+The shoreline displacement data used in this study is based on the so-called isolation basin method [e.g.
 @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments.
-Each basin thus represent a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to devise a continuous time series for RSL-change adjusted to a common isobase.
-To minimise the impact of variable uplift rates, the cored basins are therefore located in a as constrained area of the landscape as possible.
-Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from the common isobase perpendicular to this gradient (Figure \@ref(fig:map-iso)).
-The SLIPs indicate the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present day tidal range.
-This is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
+Each basin thus represents a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to devise a continuous time series for RSL-change adjusted to a common isobase.
+To minimise the impact of variable uplift rates, the cored basins are therefore located in as constrained of an area of the landscape as possible.
+Following from the morphology of the retreating ice sheet, the uplift is more stark towards the north-east, which needs to be adjusted for in the case that any basins are located any significant distance from the common isobase that runs perpendicular to this uplift gradient (Figure \@ref(fig:map-iso)).
+The SLIPs indicate the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present-day tidal range.
+For simplicity, the tidal range is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
 The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021 30cm at the standard port Helgeroa in Larvik].
-Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters.
-The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
+Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and they are the integrated result of both expert knowledge and more objectively quantifiable parameters.
+The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but they also contain potential error related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
 @romundset2011; @romundset2019; for alternative approaches see e.g.
@@
 The archaeological data compiled for the analysis consists of excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)).
-These number `r nrow(sites)` sites.
-Of these, `r length(unique(rcarbon$site_name))` sites are associated with a total of `r nrow(rcarbon)` radiocarbon dates.
-Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE), with 95% probability.
+These number `r nrow(sites)` sites of which `r length(unique(rcarbon$site_name))` sites are associated with a total of `r nrow(rcarbon)` radiocarbon dates.
+Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon date ranges that intersect the Stone Age (9500--1700 BCE), with 95% probability.
 These sites and ^14^C-dates form the basis for the analysis.
-Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History---the institution responsible for archaeological excavations in the region.
+Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History of the University of Oslo---the institution responsible for archaeological excavations and data curation in the region.
 In the compiled dataset, each radiocarbon date has been associated with the site features or excavation unit from where they originate, or, where these weren't available, the spatial limit of the entire site.
@@
 The first is Nordby 1 at which the ^14^C-dates are associated with a Late Neolithic long-house [@gjerpe2008b].
-The Middle Neolithic phase at Kvastad A2 [@stokke2018b] and Late Neolithic phase at Nauen A [@persson2008] are both directly related to farming acitivities.
+The Middle Neolithic phase at Kvastad A2 [@stokke2018b] and Late Neolithic phase at Nauen A [@persson2008] are both directly related to farming activities.
 Both of these sites also have radiocarbon dates and lithic inventory associated with Mesolithic forager activities.
@@
 The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2018, https://hoydedata.no].
-It was here opted for the 10m resolution DTM rather than the higher-resolution 1m version.
-In addition to resulting in considerably less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by.
+The 10m resolution DTM was used rather than the higher-resolution 1m version, both because this resulted in considerably less processing time and because the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances.
 The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m [@mapping2018].
@@
 The result of this process is shown for an example site in Figure \@ref(fig:example-1).
-For the the sites all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
-Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same event, here termed settlement or site phase.
-In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed.
+For the sites all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
+Radiocarbon dates associated with each site were then grouped if their date ranges intersected at 99.7% probability, meaning these were effectively taken to be associated with the same event, here termed settlement or site phase.
+In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed using the Sum function.
 Multiple phases at a single site were treated as independent of each other.
 
-The excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure.
-As the data collection for the employed DTM was begun by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has sometimes been severely impacted by disturbances after the excavation.
+The excavation of archaeological sites in Norway typically occur in advance of residential and commercial infrastructure development.
+As the data collection for the utilised DTM was begun by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has sometimes been severely impacted by disturbances after the excavation.
 In addition to employing the 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster.
@@
 @conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021].
-In addition to code and original spatial data being available in the research compendium for this paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.
-
-Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of relative sea-level change at that location, and a DTM edited to remove substantial modern disturbances, the simulations were performed.
+In addition to code and original spatial data being available in the online repository for the paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.
+
+Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of RSL change at that location, and a DTM edited to remove substantial modern disturbances, the simulations were performed.
 A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \@ref(fig:example-2)).
@@
 @conolly2006, 253; @herzog2013].
-In the case where the sea polygons intersects the site polygon, all distance measures were set to zero.
+In the case where the sea polygons intersect the site polygon, all distance measures were set to zero.
 In the case that the sea polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon.
 While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
-Negative values were therefore retained with the exception of for the sites Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
+Negative values were therefore retained with the exception of the sites of Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
 
@@
 @crema2010, 1125].
-Hovland 5 was chosen for this evaluation as it has a fairly imprecise date and is located in area of quite complex topography.
+Hovland 5 was chosen for this evaluation as it has an imprecise age and is located in area of quite complex topography.
 
@@
 ```{r example-2, echo = FALSE, fig.cap = "Example of a single simulation run on the site Hegna vest 1.
-A) The simulation starts by drawing a single year from the posterior density esimate.
+A) The simulation starts by drawing a single year from the posterior density estimate.
 B) This then corresponds to an elevation range on the interpolated displacement curve.
 A single elevation is drawn uniformly from this range using 5cm intervals.
-C) The sea-level is then adjusted on the DTM to this elevation and the various distance mesaures are found.
+C) The sea-level is then adjusted on the DTM to this elevation and the various distance measures are found.
 D) The numerical result of the simulation run.", out.width = "500px", fig.align="center", regfloat=TRUE}
@@
 ```{r example-3, echo = FALSE, fig.cap = "The result of 1000 simulation runs for each of the two groups of dates on the site Hegna vest 1.
-The first column of plots shows the calibrated radiocarbon probability distribution from where dates were drawn during simulation.
-The second column displays the result of simulating the raised sea-level 1000 times.
-The more opaque the colour, the more times the sea-level was simulated to that location.
-The third column shows violin plots of the different distance measures across all simulations.", out.width = "500px", regfloat=TRUE}
+The leftmost column of plots shows the calibrated radiocarbon probability distribution from where dates were drawn during simulation.
+The centre column displays the result of simulating the raised sea-level 1000 times.
+The more opaque the colour appears, the more times the sea-level was simulated in that location.
+The rightmost column shows violin plots of the different distance measures across all simulations.", out.width = "500px", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/example_sim_result.png"))
@@
 Some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included (Figure \@ref(fig:results)A), but if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, Figure \@ref(fig:results)B gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
-The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the simulated sea-level.
+The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the rapid RSL fall in the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the simulated sea-level.
 Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.
-From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.
+After around 2500 BCE several sites are situated a considerable distance from the reconstructed shoreline, and while a couple remain horizontally and topographically close, most appear to be considerably elevated above the sea-level, as indicated on the plot for vertical distance.
 While the sample size is limited, there are also a couple of sites located some distance from the shoreline just after 4000 BCE.
@@
 These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape.
-While there would be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level, as they would not have been in use when located under water.
-It could therefore seem that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geological reconstruction of shoreline displacement in the area.
-
-Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results for from Figure \@ref(fig:results)B is given again in Figure \@ref(fig:results2)A, excluding all simulation results younger than 2500 BCE.
+While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent an upper constraining limit for the sea-level, as they would not have been in use when located under water.
+It therefore seems that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geologic reconstruction of shoreline displacement in the area.
+
+Accepting that shoreline dating appears to lose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results from Figure \@ref(fig:results)B are presented again in Figure \@ref(fig:results2)A, excluding all simulation results younger than 2500 BCE.
 Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations.
-The resulting best point estimate for the vertical distance between sites and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, while 95% of the values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
-That is, for 95% of the cases, the shoreline was simulated to be situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
+The resulting best point estimate for the vertical distance between sites and shoreline for the pre-Late Neolithic is given by the median distance of `r round(median(distances2$vertdist))`m, while 95% of the values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m.
+That is, for 95% of the cases, the shoreline was simulated to be situated on or lower than `r round(quantile(distances2$vertdist, 0.95))`m below the site location.
 While these values remain the same when only the Mesolithic dates are included (Figure \@ref(fig:results2)B), the mean and standard deviation are slightly constrained.
@@
 ```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites.
-Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occured.
-Results associated with agricultural acitivites are plotted in grey.
+Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occurred.
+Results associated with agricultural activites are plotted in grey.
 The first row A) shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites.
@@
 While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data.
-However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.
-If one accepts this, the probability density function for exponential decay can be used to characterise the vertical distance between sites and the shoreline, and be used to inform a method for shoreline dating that takes this into account.
+However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a zero distance.
+If one accepts this, the probability density function for exponential decay can be used to characterise the vertical distance between sites and the shoreline and be used to inform a method for shoreline dating that takes this into account.
 
@@
 This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $T$ for the occupation of the site.
-This is defined by a uniform probability density function over the range between the lower $T_{l}$ and upper $T_{u}$ bounds of the displacement curve that has been interpolated to the site location:
+This is defined by a uniform probability density function ($U$) over the range between the lower $T_{l}$ and upper $T_{u}$ bounds of the displacement curve that has been interpolated to the site location:
 
@@
 
-We can then get rid of parameter $D$ by summarising over all possible distances between site and the shoreline. Given its elevation, the probability for the date of the occupation of a site is then:
+We can then get rid of parameter $D$ by summaing all possible distances between site and the shoreline. Given its elevation, the probability for the date of the occupation of a site is then:
 
@@
 An example of an implementation of the outlined approach is given in Figure \@ref(fig:exampledat), where $\lambda$ = `r round(expfit$estimate, 3)`.
-This is the decay ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
+This is the decay ratio identified when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
 For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, starting from the site elevation $\alpha$.
@@
 This probability is then divided equally across the individual calendar years in the range between the lower and the upper limit of the displacement curve at the current 0.001m step.
-The histogram that is the resulting shoreline date is the sum of performing this procedure on all possible 0.001m values of $D$, which in practice is until $\alpha - D = 0$ or when 99.999% of the exponential function has been covered.
-
-In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date all of the sites from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes.
-Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were all dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits.
+The histogram that is the resulting shoreline date is the sum of performing this procedure on all possible 0.001m values of $D$, which, in practice, is until $\alpha - D = 0$ or when 99.999% of the exponential function has been stepped through.
+
+In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date the sites with radiocarbon dates from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes.
+Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits.
 The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by @parnell2008.
@@
 The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement (Figure \@ref(fig:shoredatediff)).
-When excluding the earliest occupation phase at Gunnarsrød 5, the deviation of which is to be expected based on issues with the DTM (see above), the shoreline date correspond to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%).
+When excluding the earliest occupation phase at Gunnarsrød 5, the deviation of which is to be expected based on issues with the DTM (see above), the shoreline date corresponds to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%).
 Only including dates modelled to be older than 2500 BCE with 95% probability, i.e.
@@
 Line segments with vertical bars indicate that the HDR does not cross zero and that the dates do not correspond.
-The division and colour coding at the top of the plots reflect the divison of site phases given in Figure 9.", out.width = "525px", fig.align="center", regfloat=TRUE}
+The division and colour coding at the top of the plots reflect the division of site phases given in Figure 9.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate2.png"))
@@
 To avoid issues with recent disturbances on the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
+As all of the included sites have been excavated after the turn of the millennium, and the wide adoption of GNSS technology, the reported elevations should be trustworthy.
 
 The comparison with previously reported dates is an illustrative, but unfair exercise for a few reasons.
-First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an undefined, but implied or explicit uncertainty range.
+First, the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an undefined, but implied or explicit uncertainty range.
 Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change.
 Finally, they are sometimes only meant to indicate a lower bound for when the sites could have been in use.
-Overall, the results could, with some danger of circularity, suggest that shoreline dating has generally been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g.
+Overall, the results could, with some danger of circularity, suggest that shoreline dating has generally been applied with a reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g.
 @roalkvam2022].
-That being said, the results do also indicate that shoreline dating has at times been applied with an exaggerated degree of precision.
+With these considerations in mind, the results also indicate that shoreline dating has at times been applied with an exaggerated degree of precision.
 While the implications of a more stable RSL-change for shoreline dating are well known, this also appears to be somewhat under-appreciated in the practical implementation of the method.
@@
 
-Not least following from the fact that relatively few Preboreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway [@åstveit2018; @damlien2018b; @kleppe2018], the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
+Not least following from the fact that relatively few ^14^C-dates older than c.
+8000 BCE associated with anthropogenic activity have been achieved in Norway [@åstveit2018; @damlien2018b; @kleppe2018], the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g.
 @bang-andersen2012; @berg-hansen2018; @breivik2014; @glorstad2016; @fuglestvedt2012].
 The shoreline dated Preboreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a].
-These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision use of the sites after the sea retreated any significant distance from their location.
+These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where use of the sites would have been difficult after the sea retreated any significant distance from their location due to accessibility.
 In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically.
@@
 @sorenseninpress].
-Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date.
+Jaksland's dates are given a flat 200- and 50-year uncertainty range starting from what he gives as the earliest possible date.
 The 200 year uncertainty range is given if the sites were to be considered in isolation, while his argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other.
-Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap.
-This information is not integrated in the approach outlined here, but could justify further reducing the uncertainty ranges.
+Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they do not overlap.
+This information is not integrated in the approach outlined here, but it could justify further reducing the uncertainty ranges.
 
@@
 Above it was suggested that additional temporal data could be combined with the method to improve its accuracy and precision.
-Drawing on @jaksland2014, this example instead highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use.
+Drawing on @jaksland2014, this example instead highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help to increase the precision of the method if this can be used to exclude certain RSLs as unlikely for when a site was in use.
 One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline.
-If this is developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use.
+If this method is developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use.
 Such approaches would make less of an impact in this setting, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf.
@@
 
-The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline.
-This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c.
-4000 BCE, after which a couple of sites become situated some distance from the sea, followed by a more decisive break at the transition to the Late Neolithic at c.
+The most significant finding of this paper is a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline.
+This is indicated by the close proximity of sites and the shoreline until the transition to the Neolithic at c.
+4000 BCE, after which a few sites are situated some distance from the sea, followed by a more decisive break at the transition to the Late Neolithic at c.
 2500 BCE.
@@
 Apart from taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves accounting for the distance between the sites and the shoreline.
-When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio of `r round(expfit$estimate, 3)` (Figure \@ref(fig:redate)A) to characterise this relationship.
+When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio of `r round(expfit$estimate, 3)` (Figure \@ref(fig:results2)A) to characterise this relationship.
 Furthermore, while this remains to be formalised and explored further, it was also showed how the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data.
@@
 Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further evaluate and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted.
-Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located.
+Given its behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located.
 Furthermore, other dimensions related to the topographic location of the sites could be similarly explored.
@@
 The potential of exploring dimensions such as these was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline.
-If patterns related to such locational patterns can be discerned and unpicked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information beyond their location is available.
+If patterns related to such locational patterns can be discerned and unpacked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information beyond their location is available.
 
 Some limitations and sources of likely variation and uncertainty that have not been considered should also be mentioned.
-First of all the sample size is strained and the future addition of more sites might alter the picture considerably.
+First, the sample size is limited and the future addition of more sites might alter the picture considerably.
 Secondly, the validity of the outlined method was evaluated by applying it to the data from where the input parameters were derived.
@@
 Thirdly, the DTM has only been corrected for major modern disturbances.
-This means that other forms of erosion, although likely not that prevalent, has not been taken into account.
-Fourthly, the DTM has a vertical error which could also benefit from being integrated in the analysis [cf.
-@lewis2021].
+This means that other forms of erosion, although likely not that prevalent, have not been considered.
+Fourthly, the DTM has a vertical error which could also benefit from being integrated in the analysis [@fisher1993; @lewis2021].
 Fifthly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves further away from the isobases of the displacement curves.
 This is also related to the fact that the RSL data can be handled in different ways than with the isobase method that has been used for the compilation of the employed displacement curves.
-Sixthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration [cf.
-@mjærum2022].
-Finally, the radiocarbon dates and division of settlement phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7% as representing unrelated occupation events.
+Sixthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration [@mjærum2022].
+Finally, the aggregation and division of settlement phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7% as representing unrelated occupation events.
 This could also be handled differently [e.g.
@@
 
-Given that shoreline dating is contingent on regularities in human behaviour it should naturally be applied with care.
-Furthermore, formulating and visualising the method along the lines of how radiocarbon dates are treated, as was done here, does stand the chance of giving a veneer of radiometric accuracy that is not warranted.
+Given that shoreline dating is contingent on regular patterns of human behaviour it should naturally be applied with care.
+Furthermore, formulating and visualising the method along the lines of how radiocarbon dates are treated, as was done here, does stand the chance of giving a veneer of radiometric accuracy to shoreline dating that is not warranted.
 That being said, the best chance we have of not throwing away precious temporal data, or exaggerating our handle on it, is arguably to rigorously evaluate the method using independent data such as radiocarbon dates, by offering a precise formulation of how it could be applied, by specifying what sources of uncertainty are accounted for and by making this process transparent through the open dissemination of underlying data and programming code.

### 2022-09-06 analysis/paper/paper.Rmd
Updated supplementary and figures 1+2

@@
 
+# Point features with city names
+cpts <- st_read(here('analysis/data/raw_data/central_points.gpkg'))
+
 # Digital terrain model
@@
 ```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from c.
-17--8 thousand years BCE.
+17--8 thousand years (kyr) BCE.
 The study area defined later in the text is marked with a red outline (deglaciation data from Hughes et al.
@@
 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region.
-However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation and return to foraging in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227].
+However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a return to foraging and complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227].
 @nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g.
@@
 Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date span is too wide to decisively verify this.
-It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would appear to have been sensible if the site was not shore bound based on an archaeological understanding [e.g.
+It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would appear to have been sensible if the site was not shore bound [e.g.
 @jaksland2014; @johansen1963; @nummedal1923].
@@
 If the site is situated on a ledge in a steep and jagged area of the present-day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft.
-Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @ilves2011].
+Although it appears that the arguments for such site locations can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @ilves2011].
 
@@
 While there is displacement data available for other areas of south-eastern Norway [e.g.
-@hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west.
-This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
+@hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)).
+This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
 
@@
 
-```{r map-iso,  fig.cap = paste("A) Distribution of the ", nrow(sites_sa), " analysed sites relative to the isobases of the displacement curves.
+```{r map-iso,  fig.cap = paste("A) Location of the study area and the distribution of the ", nrow(sites_sa), " analysed sites relative to the isobases of the displacement curves.
 The isobases have a direction of 327$^{\\circ}$ (Romundset et al.
-2018, but see Sørensen et al.
+2018, although see Sørensen et al.
 2014), B) Displacement curves.
-Note the increasing steepness of the curves towards the north-east."), fig.height = 13, regfloat=TRUE}
+Note the increasing steepness of the curves towards the north-east."), fig.height = 13, fig.width = 12, regfloat=TRUE}
+# Create a bounding box around site points to limit the overview map
+sitbbox <- st_bbox(sites_sa)
+sitbbox[1:2] <- sitbbox[1:2] - 300000
+sitbbox[3:4] <- sitbbox[3:4] + 300000
+boundingpoly <- st_as_sf(st_as_sfc(sitbbox))
+
+# Reproject the bounding box to match world map, and crop the world map
+# with the bounding box polygon
+bound_reproj <- st_transform(boundingpoly, st_crs(bmap))
+mapcountries <- bmap %>% 
+  filter(st_intersects(., bound_reproj, sparse = FALSE))
+count_reproj <- st_transform(mapcountries, st_crs(sites_sa))
+
 bboxsites <- st_bbox(sites_sa)
@@
 
+cpts <- filter(cpts, !(NAME %in% c("Tvedestrand", "Porsgrunn")))
+
+overview <-
+  ggplot() +
+  geom_sf(data = count_reproj, fill = "grey", colour = NA) +
+  geom_sf(data = bboxsitespoly,
+          fill = NA, colour = "red", size = 0.25) +
+  geom_sf(data = cpts) +
+   ggrepel::geom_text_repel(
+    data = cpts,
+    aes(label = NAME, geometry = geom),
+    stat = "sf_coordinates",
+    min.segment.length = 0,
+    size = 4) +
+  coord_sf(xlim = c(sitbbox[1], sitbbox[3]), ylim = c(sitbbox[2], sitbbox[4]),
+         expand = FALSE) +
+  labs(title = "A") +
+  theme_bw() + theme(axis.title=element_blank(),
+                     axis.text.y = element_blank(),
+                     axis.text.x = element_blank(),
+                     rect = element_rect(),
+                     axis.ticks = element_blank(),
+                     panel.grid.major = element_blank())
+
 bound_reproj <- st_transform(bboxsitespoly, st_crs(bmap))
@@
 anc <- as.numeric(c(bboxsites$ymin, bboxsites$xmax))
-overview_map <- ggplot() +
+site_map <- ggplot() +
   geom_sf(data = bmap_reproj, fill = "grey", colour = NA) +
@@
   geom_line(aes(y = lowerelev)) +
-  ylab("Meters above present sea-level") +
-  xlab("BCE/CE") +
+  labs(title = "B", y = "Meters above present sea-level", x = "BCE/CE") +
   scale_colour_manual(values = c("Arendal" = "black",
@@
 
-overview_map / disp_curves + plot_annotation(tag_levels = "A")
+(overview + site_map) /
+(disp_curves + theme(plot.margin = unit(c(0,135,0,135), "pt")))
 ```
@@
 The red outline indicates that the date does not match the typological indicators in the artefact assemblage of the site.
-C) The location of the site within the study area relative to isobases of the employed displacement curves.
+C) The location of the site within the study area relative to isobases of the displacement curves.
 D) Displacement curve interpolated to the site location.", out.width = "500px", fig.align="center", regfloat=TRUE}
@@
 ```{r example-site-cont3}
-# Load results from running 1000 simulations
+# Load simulation results
 load(here("analysis/data/derived_data/hegnavest1.RData"))
@@
   labs(x = "")  +
-  #scale_y_discrete(expand = c(0.01, 0)) +
   theme_bw() + theme(
@@
 While the sample size is limited, there are also a couple of sites located some distance from the shoreline just after 4000 BCE.
-That the findings appear to be off from the chronological framework by around a century must be seen in relation to chronological smearing from the uncertainty in the ^14^C-dates, and the findings are thus in clear agreement with the literature.
+The chronological smearing following from the uncertainty in the ^14^C-dates means that while the results cannot be used to directly inform discussions that deal with the century scale around these chronological transitions [e.g.
+@solheim2021; @prescott2020], the findings are nonetheless in clear agreement with the general chronological framework in the literature.
 
 The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3.
-These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape.
+These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape [@reitan2022b].
 While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent an upper constraining limit for the sea-level, as they would not have been in use when located under water.
@@
 
-In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date the sites with radiocarbon dates from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes.
+To evaluate the outlined procedure it is used to shoreline date the sites from where the method was derived to check if the resulting shoreline dates correspond to the radiocarbon dates of the sites (Figure \@ref(fig:shoredate)).
+The Late Neolithic sites are also included here for illustrative purposes, even though these have not informed the decay ratio in use.
 Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits.
@@
 The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black.
-These are plotted against the modelled radiocarbon dates, which are given colour from oldest to youngest occupation phase for each site, defined by non-overlapping dates at 99.7\\% probability.", out.width = "525px", fig.align="center", regfloat=TRUE}
+These are plotted against the modelled radiocarbon dates, which are given colour from oldest to youngest occupation phase for each site, defined by non-intersecting dates at 99.7\\% probability.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 The small discrepancies between the achieved results mainly follow from the fact that a slightly updated version of the local displacement curve is applied here [cf.
-@sorenseninpress].
+@sørensen2014; @sorensen2023].
 Jaksland's dates are given a flat 200- and 50-year uncertainty range starting from what he gives as the earliest possible date.

### 2022-09-06 analysis/supplementary-materials/supplementary.Rmd
Updated supplementary and figures 1+2

@@
 
-Løvås 1.
+The artefact inventory from Løvås 1 matches the radiocarbon dates [@reitan2022].
 No editing of the raster appeared to be necessary.
@@
 
-Løvås 2. Similar situation to that on Løvås 1 and 3 (above and below).
+The artefact inventory from Løvås 2 matches the radiocarbon dates [@reitan2022]. Similar setting to that of Løvås 1 and 3 (above and below).
 
@@
 
-Similar situation to that on Løvås 1 and 2 (above).
+The artefact inventory from Løvås 3 matches the radiocarbon dates [@reitan2022]. Similar setting to that of Løvås 1 and 2 (above).
 
@@
 <!-- @reitan2014h -->
+<!-- @reitan20140 -->
 <!-- @mansrud2018 -->
 <!-- @persson2008 -->
+<!-- @eigeland2014 -->
 <!-- @eigeland2014a -->
-<!-- @eigeland2014 -->
 <!-- @fossum2014a -->
@@
 <!-- @darmark2018 -->
+<!-- @darmark2018d -->
 <!-- @nielsen2017a -->
@@
 <!-- @reitan2014i -->
+<!-- @reitan2014f -->
 

### 2022-09-14 analysis/paper/paper.Rmd
Adding a couple more sites for re-dating and updating supplementary material references

@@
   A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use.
-By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied.
+By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed 'shoreline dating' which is widely applied.
 However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more established dating techniques in archaeology.
@@
 @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)).
-This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].
+This area has newly compiled displacement curves for Skoppum in Horten [@romundset2021], Gunnarsrød in Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Hanto in Tvedestrand [@romundset2018; @romundset2018b], and Bjørnebu in Arendal [@romundset2018b].
 
@@
 Note the increasing steepness of the curves towards the north-east."), fig.height = 13, fig.width = 12, regfloat=TRUE}
-# Create a bounding box around site points to limit the overview map
+# To make sure labelling ends up the same every knit.
+set.seed(1)
+
+# Create a bounding box around site points to limit the overview map.
 sitbbox <- st_bbox(sites_sa)
-sitbbox[1:2] <- sitbbox[1:2] - 300000
-sitbbox[3:4] <- sitbbox[3:4] + 300000
+sitbbox[1:2] <- sitbbox[1:2] - 100000
+sitbbox[3:4] <- sitbbox[3:4] + 100000
 boundingpoly <- st_as_sf(st_as_sfc(sitbbox))
@@
 # Reproject the bounding box to match world map, and crop the world map
-# with the bounding box polygon
+# with the bounding box polygon.
 bound_reproj <- st_transform(boundingpoly, st_crs(bmap))
@@
 bboxsitespoly <- st_as_sf(st_as_sfc(bboxsites))
-
-cpts <- filter(cpts, !(NAME %in% c("Tvedestrand", "Porsgrunn")))
 
@@
 
-disp_curves <- ggplot(displacement_curves, aes(x = years, col = name)) +
+# Rename to use the precise placenames
+rname_disp <- displacement_curves
+rname_disp$name <- ifelse(displacement_curves$name == "Larvik",
+                                   "Gunnarsrød", 
+                                   ifelse(
+                                     displacement_curves$name == "Tvedestrand",
+                                     "Hanto", 
+                                     ifelse(
+                                       displacement_curves$name == "Arendal",
+                                       "Bjørnebu", "Skoppum")))
+
+disp_curves <- ggplot(rname_disp, aes(x = years, col = name)) +
   geom_line(aes(y =  upperelev)) +
@@
   labs(title = "B", y = "Meters above present sea-level", x = "BCE/CE") +
-  scale_colour_manual(values = c("Arendal" = "black",
-                                   "Larvik" = "darkgreen",
-                                   "Tvedestrand" = "blue",
-                                   "Horten" = "darkorange")) +
+  scale_colour_manual(values = c("Skoppum" = "darkorange",
+                                 "Gunnarsrød" = "darkgreen",
+                                 "Hanto" = "blue",
+                                 "Bjørnebu" = "black")) +
   theme_bw() +
@@
 Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with temporal uncertainty.
-The first task was therefore to ascribe a likely date and associated uncertainty to these dimensions.
+The first task was therefore to ascribe a likely date and associated degree of uncertainty to these dimensions.
 To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves, using inverse distance weighting [e.g.
@@
 @conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021].
-In addition to code and original spatial data being available in the online repository for the paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.
+In addition to code and original spatial data being available in the online repository for the paper, the location and analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.
 
@@
 curve_plot <- ggplot2::ggplot() + 
-              geom_line(data = displacement_curves, 
+              geom_line(data = rname_disp, 
                       aes(x = years, y = upperelev, col = name), alpha = 0.4) +
-  geom_line(data =  displacement_curves, 
+  geom_line(data =  rname_disp, 
                       aes(x = years, y = lowerelev, col = name), alpha = 0.4) +
@@
             aes(x = years, y = lowerelev, col = name)) +
-  scale_colour_manual(values = c("Arendal" = "black",
-                                   "Larvik" = "darkgreen",
-                                   "Tvedestrand" = "blue",
-                                   "Horten" = "darkorange",
+  scale_colour_manual(values = c("Skoppum" = "darkorange",
+                                 "Gunnarsrød" = "darkgreen",
+                                 "Hanto" = "blue",
+                                 "Bjørnebu" = "black",
                                  "Hegna vest 1" = "red")) +
@@
 
-We can then get rid of parameter $D$ by summaing all possible distances between site and the shoreline. Given its elevation, the probability for the date of the occupation of a site is then:
+We can then get rid of parameter $D$ by summing all possible distances between site and the shoreline. Given its elevation, the probability for the date of the occupation of a site is then:
 
@@
 
-To evaluate the outlined procedure it is used to shoreline date the sites from where the method was derived to check if the resulting shoreline dates correspond to the radiocarbon dates of the sites (Figure \@ref(fig:shoredate)).
+To evaluate the outlined procedure it is used to shoreline date the sites from where the method was derived to check if the resulting shoreline dates correspond to the radiocarbon dates associated with the sites (Figure \@ref(fig:shoredate)).
 The Late Neolithic sites are also included here for illustrative purposes, even though these have not informed the decay ratio in use.
@@
 Here this was simply used to inform the question of the distance between sites and the shoreline.
-However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent in this reconstruction.
+However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent to this reconstruction.
 

### 2022-09-14 analysis/supplementary-materials/supplementary.Rmd
Adding a couple more sites for re-dating and updating supplementary material references

@@
 Interestingly only the Løvås sites were simulated to be situated beneath the sea-level.
-The typologically secure inventory combined with the many overlapping dates from various features does indicate that the issue is with the displacement curve and not the dating of the site.
+The typologically secure inventory combined with the many overlapping dates from various features does indicate that the issue is with the displacement curve and not the dating of the site (see also main text).
 Similar situation to that on Løvås 2 and 3 (below).
@@
 This appears to have been successfully edited.
-The excavated area is more extensive (field A--C in the report) that what is used as the site limit here, but the relation between the different areas is not clear.
+The excavated area is more extensive (field A--C in the report) than what is used as the site limit here, but the relation between the different areas is not clear.
 Field A, used as the site limit here, holds all the dated features.
@@
 This appears to have been successfully edited, although there is a possibility that this has led to the sea-level being simulated sightly further away from the site than what would otherwise have been the case.
-However, as with he Hovland-sites above, this appears to rather follow from the general topography of the landscape which makes the reconstruction more uncertain than in areas with a higher relief.
+However, as with he Hovland sites above, this appears to rather follow from the general topography of the landscape characterised by a low relief which makes the reconstruction more uncertain.
 
@@
 
-Vallermyrene 1 was divided into 1a and 1b based on elevation difference, typology and 14C-dates [@reitan2014f].
-Vallermyene 1a is the oldest of the two.
-Although situated not that far from the railway, the area immediatly surrounding the site has not been disturbed in a way that appears to impact the reconstruction of the past sea-level.
+Vallermyrene 1 was divided into 1a and 1b based on elevation difference, typology and ^14^C-dates [@reitan2014f].
+Vallermyrene 1a is the oldest of the two.
+Although situated not that far from the railway, the area immediately surrounding the site has not been disturbed in a way that appears to impact the reconstruction of the past sea-level.
 
@@
 Vallermyrene 2 was a large site situated in a field [@reitan2014o].
-The site has ^14^C-dates and artefacts dating occupations at the sites to multiple phases throughout prehistory, including the Bronze Age and phases up until the Viking Age.
+The site has ^14^C-dates and artefacts dating occupations at the site to multiple phases throughout prehistory, including the Bronze Age and phases up until the Viking Age.
 The Stone Age activity on the site was evidenced by 359 lithics and a fireplace dated to the Early Neolithic.
@@
 <!-- @fossum2017c -->
+<!-- @fossum2017b -->
 <!-- @mansrud2016 -->
@@
 <!-- @reitan2014f -->
+<!-- @reitan2014s -->
 

### 2022-09-21 analysis/paper/paper.Rmd
Comments from Anders

@@
 However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more established dating techniques in archaeology.
-This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast.
+This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located below the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast.
 Monte Carlo simulation is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement.
@@
 
-The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was proposed at the end of the 19th century [@degeer1896], and it was first applied as a dating method at the turn of the century [@brogger1905; @hollender1901].
+The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was proposed at the end of the 19th century [@degeer1896], and was first applied as a dating method at the turn of the century [@brogger1905; @hollender1901].
 Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g.
@@
 
-The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than an earliest possible date for when a site could have been in use [see review by @nordqvist1999].
+The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, it has been characterised as a relative dating method for sites located within a constrained geographical area, or it has been argued to offer no more than an earliest possible date for when a site could have been in use [see review by @nordqvist1999].
 The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @glorstad2002; @glorstad2003; @glorstad2004; @jaksland2001; @jaksland2012; @jaksland2012a; @jaksland2014a; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below].
@@
 This is done by aggregating point estimates of shoreline dates in 100-, 200- or 500-year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020].
-In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963; @johansen1997; @mikkelsen1975a, 100].
+In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method---what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963; @johansen1997; @mikkelsen1975a, 100].
 
 The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; @bjerck2017; also discussed by @berg-hansen2009; @bergsvik2009].
-The same logic has also been extended to the hinterland and inland regions, where sites are belived to be predominantly located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020].
+The same logic has also been extended to the hinterland and inland regions, where sites are believed to be predominantly located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020].
 This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@prescott2020; cf.
@@
 
-Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
-This is for example reflected in archaeological survey practices, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g.
-@berg-hansen2009; @eskeland2017; @nummedal1923].
+Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit (the highest elevation of the sea after the retreat of the ice) were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise.
+This is for example reflected in archaeological survey practices, which are often guided by both a digital and mental reconstruction of past sea-levels [see e.g.
+@berg-hansen2009; @eskeland2017; @nummedal1923; @simpson2009].
 Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date span is too wide to decisively verify this.
@@
 
-One of the more extensive evaluations of the relationship between archaeological radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
-They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites.
+One of the more extensive evaluations of the relationship between archaeological radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
+They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.3% of the dates [@solheim2020, 48] .
 However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates.
@@
 This thus leads to an inferential framework that favours uncertainty, which is hardly desirable.
-In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are actually the same.
+In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same.
 The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g.
@@
 
-The shoreline displacement data used in this study is based on the so-called isolation basin method [e.g.
+The shoreline displacement data used in this study are based on the so-called isolation basin method [e.g.
 @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments.
-Each basin thus represents a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to devise a continuous time series for RSL-change adjusted to a common isobase.
-To minimise the impact of variable uplift rates, the cored basins are therefore located in as constrained of an area of the landscape as possible.
+Each basin thus represents a high precision sea-level index point (SLIP) which are combined in a continuous time series for RSL-change adjusted to a common shoreline isobase.
+The isobases are here contours indicating equal shoreline displacement over the same time span [@svendsen1987, 116].
+To minimise the impact of variable uplift rates, the cored basins are located in as constrained of an area of the landscape as possible.
 Following from the morphology of the retreating ice sheet, the uplift is more stark towards the north-east, which needs to be adjusted for in the case that any basins are located any significant distance from the common isobase that runs perpendicular to this uplift gradient (Figure \@ref(fig:map-iso)).
-The SLIPs indicate the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present-day tidal range.
+Furthermore, as the uplift has been greater immediately following the retreat of the ice, such adjustments, and thus potential uncertainty, will be more critical further back in time.
+The resulting SLIPs are most commonly interpreted as representing the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present-day tidal range.
 For simplicity, the tidal range is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
-The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021 30cm at the standard port Helgeroa in Larvik].
-Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and they are the integrated result of both expert knowledge and more objectively quantifiable parameters.
-The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but they also contain potential error related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g.
-@romundset2011; @romundset2019; for alternative approaches see e.g.
-@barnett2020; @cahill2016; @creel2022].
+The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021; 30cm at the standard port Helgeroa in Larvik].
+Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs as they are the integrated result of both expert knowledge and more objectively quantifiable parameters.
+The reason for this is that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but they are for example also dependent on the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, as well as being based on expert knowledge of regional post-glacial geologic developments and local geomorphology, to name but a few factors [e.g.
+@romundset2011; @romundset2018; @svendsen1987; for an alternative approach see @creel2022].
 For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.
@@
 The chronological smearing following from the uncertainty in the ^14^C-dates means that while the results cannot be used to directly inform discussions that deal with the century scale around these chronological transitions [e.g.
-@solheim2021; @prescott2020], the findings are nonetheless in clear agreement with the general chronological framework in the literature.
+@prescott2020; @solheim2021], the findings are nonetheless in clear agreement with the general chronological framework in the literature.
 
@@
 Fourthly, the DTM has a vertical error which could also benefit from being integrated in the analysis [@fisher1993; @lewis2021].
-Fifthly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves further away from the isobases of the displacement curves.
-This is also related to the fact that the RSL data can be handled in different ways than with the isobase method that has been used for the compilation of the employed displacement curves.
+Fifthly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves further away from the isobases of the displacement curves---an uncertainty that is likely higher for RSL-change further back in time due to the shoreline gradient.
+This is also related to the fact that the geologic reconstructions hold uncertainty that is not represented in the displacement curves, relating for example to variation in the methods and quality of the data used for the compilation of the curves, as well as the expert interpretations underlying these.
 Sixthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration [@mjærum2022].

### 2022-09-21 analysis/supplementary-materials/supplementary.Rmd
Comments from Anders

@@
 This was used to inform the use of 1000 simulation runs for the subsequent analysis of the other sites, as the means appear to have stabilised sufficiently by this point.
-Hovland 5 was chosen because the site only has a single ^14^C-date, giving a wide uncertainty range, and because it is located in an area of the landscape where the landscape results in quite large variability on the distance measures.
+Hovland 5 was chosen because the site only has a single ^14^C-date, giving an uncertain age, and because it is located in an area of the landscape where the landscape results in quite large variability on the distance measures.
 
@@
 The second plot on the first row displays the site location in the present day landscape on the edited DTM (i.e.
-highways and railways impacting the adjusted sea-level have been removed, and elevation values are interpolated from the surroundings; see main text and analysis/script/03script.R).
+highways and railways impacting the adjusted sea-level have been removed, and elevation values are interpolated from the surroundings; see main text and analysis/script/03dtm_edit.R).
 The third plot on the first row shows the site location within the study area relative to the isobases of the shoreline displacement curves in use.
@@
 
-Adal vestre 1 is situated close to a field, but this does not appear to have resulted in any relevant changes to the terrain.
+The single-phased lithic inventory matches the radiocarbon dates to the Middle Mesolithic at Adal vestre 1 [@granados2022].
+The site is situated close to a field, but this does not appear to have resulted in any relevant changes to the terrain.
 
@@
 
-Adal vestre 2 has a similar location in the present day landscape to that of Adal vestre 1 (above).
+The artefact inventory from Adal vestre 2 is single-phased and matches the ^14^C-dates [@granados2022].
+The site has a similar location in the present day landscape to that of Adal vestre 1 (above) with no apparent indications of relevant impact to the DTM from modern activities.
 
@@
 
-The site limit for Auve was drawn manually based on the description in the publication of the site and the site geometry available in the database Askeladden.
-The site is located in and around an intersection of the road, parts of the area has been levelled and farmed, and an area to north-west of the site has reportedly been affected by land-slides [@ostmo2008, 16--29] , but the extent of this is difficult to ascertain based on maps of the area.
+The site limit for Auve was drawn manually based on the description in the publication of the site and the site geometry available in the national heritage database Askeladden.
+The site is located in and around an intersection of the road, parts of the area has been levelled and farmed, and an area to north-west of the site has reportedly been affected by land-slides [@ostmo2008, 16--29], but the extent of this is difficult to ascertain based on maps of the area.
 As the simulation results appear reasonable [see also map in @ostmo2008, fig.142], nothing was done to try to correct for these factors.

### 2022-09-23 analysis/paper/paper.Rmd
Final edits before submission

@@
 Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g.
-@berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @olsen1984; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018].
+@berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @olsen1984; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018; @solheim2020b].
 The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.
@@
 For simplicity, the tidal range is assumed to have been the same throughout the Holocene [@sørensen2014, 44].
-The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021; 30cm at the standard port Helgeroa in Larvik].
+The highest astronomical tide in the study area reaches around 30cm above mean sea-level [30cm at the standard port Helgeroa in Larvik, @mapping2021].
 Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs as they are the integrated result of both expert knowledge and more objectively quantifiable parameters.
@@
 If this method is developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use.
-Such approaches would make less of an impact in this setting, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf.
+Such approaches would make less of an impact for the Brunlanes sites, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf.
 Figure \@ref(fig:redate)).
@@
 
-# References
+# References {-}
 

### 2022-09-23 analysis/supplementary-materials/supplementary.Rmd
Final edits before submission

@@
 
-The site limit for Auve was drawn manually based on the description in the publication of the site and the site geometry available in the national heritage database Askeladden.
+The site limit for Auve was drawn manually based on the description in the publication of the site and the site geometry available in the national heritage database Askeladden [@directorate2018].
 The site is located in and around an intersection of the road, parts of the area has been levelled and farmed, and an area to north-west of the site has reportedly been affected by land-slides [@ostmo2008, 16--29], but the extent of this is difficult to ascertain based on maps of the area.
@@
 The first archaeological finds from Rognlien were done in the 1880s and, following some smaller investigations, was excavated by @ingstad1970 in 1957.
-The site limit has been manually defined based on the description given in the publication and the spatial geometry of the site as defined in the database Askeladden.
+The site limit has been manually defined based on the description given in the publication and the spatial geometry of the site as defined in the database Askeladden [@directorate2018].
 The manually defined limit could be slightly off, but should match the elevation of the site given in the publication [@ingstad1970, 26].

### 2022-09-23 analysis/paper/paper.Rmd
Setting up tiling of edited DTM to distribute with research compendium

@@
 
+```{r load_dtm, include = FALSE}
+# This chunk tried to the load the edited DTM. Due to the file sizes involved
+# this is not distributed with the repository but had to be assembled by running
+# 03dtm_edit.R which compiles the original DTM stored in 400 tiles, edits this
+# and stores the derived version in the derived data folder. This requires
+# a local installation of GRASS GIS (see the script for more details).
+
+# Digital terrain model
+dtm <- rast(here("analysis/data/derived_data/dtm10.tif"))
+
+```
+
 ```{r get-data-functions, include = FALSE}
@@
 cpts <- st_read(here('analysis/data/raw_data/central_points.gpkg'))
-
-# Digital terrain model
-dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
 

### 2022-09-23 analysis/script/03dtm_edit.R

@@ -7,21 +7,34 @@ library(rgrass7)
 # created manually based on an inspection of satellite data and the DTM.
 
 # This path has to be set to the local installation of GRASS
-grasspath <- "/usr/lib/grass78"
-# Path to raster data. Stored locally for now
-rpath <- "/home/isak/phd/eaa_presentation/dtm10/"
+# grasspath <- "/usr/lib/grass78"
 
-# Read in data
-dtm <- rast(file.path(rpath, "dtm10_32.tif"))
+try(!is.na(grasspath)) # Uncomment and det path to local GRASS installation
+                       #  above or this will throw and error.
+
+# Due to the file size (c. 1.6 GB), the DTM is stored in 400 tiles that are
+# first merged and then saved at this destination
+dtmpath <-  here::here("analysis/data/derived_data")
+
+# Get path to DTM tiles
+dtmtiles <- list.files(here::here("analysis/data/raw_data/tiled_dtm10"),
+                       pattern= ".*.tif$", full.names = TRUE)
+
+# Load the tiles and merge
+tiles <- sapply(dtmtiles, raster::raster)
+names(tiles)[1:2] <- c('x', 'y')
+dtm <- rast(do.call(raster::merge, tiles))
+
+# Read in vector data defining problem areas to be edited
 clip <- st_read(here::here("analysis/data/raw_data/clipping_poly.gpkg"))
 
 # Make the vector feature a raster
-rclip <- rasterize(vect(clip), dtm)
+rclip <- terra::rasterize(vect(clip), dtm)
 
 # Make the raster NA at the polygons
 dmask <- mask(dtm, rclip, inverse = TRUE)
 
-edrast <- file.path(rpath, "dtm10.tif")
+edrast <- file.path(dtmpath, "dtm10.tif")
 writeRaster(dmask, edrast,
             overwrite = TRUE)
 

### 2022-09-23 analysis/paper/paper.Rmd
Added the edited DTM as tiles

--- 2022-09-23 analysis/paper/paper.Rmd
@@
 
+```{r load_dtm, include = FALSE}
+# This chunk tries to the load the edited DTM. Due to the file sizes involved,
+# this is distributed as tiles which have to be merged by running 00dtm_prep.R
+
+# Digital terrain model
+dtm <- rast(here("analysis/data/derived_data/dtm10.tif"))
+
+```
+
 ```{r get-data-functions, include = FALSE}

### 2022-09-23 analysis/script/03dtm_edit.R

@@ -6,24 +6,25 @@ library(rgrass7)
 # new values. Polygon data with which to remove these features has been
 # created manually based on an inspection of satellite data and the DTM.
 
-# This path has to be set to the local installation of GRASS
+
+# --- Only the edited DTM is distributed with this research compendium ---
+#                         (see 00dtm_prep.R)
+# To run, this script requires a local installation of GRASS GIS as well as
+# the manual download of unedited DTM from hoydedata.no
+
+
+##  This path has to be set to the local installation of GRASS
 # grasspath <- "/usr/lib/grass78"
 
-try(!is.na(grasspath)) # Uncomment and det path to local GRASS installation
+try(!is.na(grasspath)) # Uncomment and set path to local GRASS installation
                        #  above or this will throw and error.
 
-# Due to the file size (c. 1.6 GB), the DTM is stored in 400 tiles that are
-# first merged and then saved at this destination
-dtmpath <-  here::here("analysis/data/derived_data")
+# Path to locally stored unedited version of the DTM from the
+# Norwegian Mapping Authority (freely available from hoydedata.no)
+# dtm <- rast("/home/isak/spatial_data/dtm10.tif")
 
-# Get path to DTM tiles
-dtmtiles <- list.files(here::here("analysis/data/raw_data/tiled_dtm10"),
-                       pattern= ".*.tif$", full.names = TRUE)
-
-# Load the tiles and merge
-tiles <- sapply(dtmtiles, raster::raster)
-names(tiles)[1:2] <- c('x', 'y')
-dtm <- rast(do.call(raster::merge, tiles))
+try(!is.na(dtm)) # Uncomment and read in locally stored unedited DTM
+                 # above or this will throw and error.
 
 # Read in vector data defining problem areas to be edited
 clip <- st_read(here::here("analysis/data/raw_data/clipping_poly.gpkg"))
@@ -34,6 +35,7 @@ rclip <- terra::rasterize(vect(clip), dtm)
 # Make the raster NA at the polygons
 dmask <- mask(dtm, rclip, inverse = TRUE)
 
+dtmpath <-  here::here("analysis/data/derived_data")
 edrast <- file.path(dtmpath, "dtm10.tif")
 writeRaster(dmask, edrast,
             overwrite = TRUE)

### 2022-09-23 analysis/script/07shorelinedate.R

@@ -14,8 +14,10 @@ load(here("analysis/data/derived_data/01data.RData"))
 load(here("analysis/data/derived_data/02data.RData"))
 load(here("analysis/data/derived_data/06data.RData"))
 
-# Read in raster
-dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
+# This tries to the load the edited DTM. Due to the file sizes involved,
+# this is distributed as tiles which have to be merged by first
+# running 00dtm_prep.R
+dtm <- rast(here("analysis/data/derived_data/dtm10.tif"))
 
 # Example site (effectively stepping through the code from shoreline_date in
 # 04.functions.R)

### 2022-09-23 README.Rmd
Update readme

@@
 
-<!-- Our pre-print is online here: -->
+<!-- The pre-print is online here: -->
 
@@
     generated by the analysis. It also has a rendered version,
-    `paper.docx`, suitable for reading (the code is replaced by figures
+    `paper.pdf`, suitable for reading (the code is replaced by figures
     and tables in this file)
@@
     supplementary-materials](/analysis/supplementary-materials):
-    Supplementary materials including notes and other documents
-    prepared and collected during the analysis.
+    Supplementary materials.
 
@@
 - run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
-[DESCRIPTION](/DESCRIPTION) file).
-- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console
+[DESCRIPTION](/DESCRIPTION) file). - open and run `analysis/script/00dtm_prep.R` to compile the necessary raster data.
+- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.pdf`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console
 
@@
 
-**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse
+**Raster data in *analysis/data/raw_data/tiled_dtm* :** [NLOD](https://data.norge.no/nlod/en/)
+
+**Remaining data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse
 

### 2022-10-20 analysis/paper/paper.Rmd
Working on comments from peer reviewers

@@
 
-An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation (Figure \@ref(fig:results2)).
-While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data.
-However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a zero distance.
+It is clear from the distributions that the simulation results have a severe right skew.
+Most sites were likely situated less than a meter from the shoreline, and from this there is a decline in density as one moves further along the x-axis.
+This trend is reasonably approximated by a exponential function that has been fit to the distributions for vertical distance using maximum likelihood estimation (Figure \@ref(fig:results2)).
+An expanded consideration of this relationship and comparison with some standard models is provided in the supplementary material.
+While exponential decay reasonably captures this relationship, it is also clear that it does not perfectly match the data.
+However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from the resolution of the spatial data, from defining intersecting sea- and site polygon as having a zero distance, from sampling error and, although they are relatively few, from forcing negative values to zero.
 If one accepts this, the probability density function for exponential decay can be used to characterise the vertical distance between sites and the shoreline and be used to inform a method for shoreline dating that takes this into account.

### 2022-10-20 analysis/supplementary-materials/supplementary.Rmd
Working on comments from peer reviewers

@@
 
-Vallermyrene 1b is the younger than Vallermyrene 1a (above), as indicated both by the radiocarbon date, typology and elevation [@reitan2014f]. The DTM did not require editing.
+Vallermyrene 1b is younger than Vallermyrene 1a (above), as indicated both by the radiocarbon date, typology and elevation [@reitan2014f]. The DTM did not require editing.
 
@@
 The DTM was edited to remove the railway track running by the site and a gravel pit.
+
+## Model comparison
+
+This section provides some further elaboration on the choice of an exponential model to describe the vertical distance between sites and shoreline.
+The normal distribution has been fit when including the negative values.
+After this, two exponential models were fit to the distrbutions.
+One when forcing negative values to zero and one where these are made NA, to evaluate the relative impact of these different choices.
+As the differences are negligeble, the choice was made to use the first solution.
+This is also the to which the power law distribution has been fit.
+To be able to fit a power law distribution, 0.01 have been added to all values, including for the normal and exponential functions, to avoid having values of zero.
+All models are fit using maximum likelihood estimation.
+
+Although it appears evident that the exponential model is the best fit simply from inspecting the, the models have also been compared using the Schwarz criterion, often also termed the Bayesian information criteria [BIC, see e.g.
+@timpson2021; @crema2019 for previous use in archaeology].
+This evaluates the ability of the models to describe the data while also taking into account the number of parameters in the models to avoid over-fitting.
 

### 2022-10-20 analysis/script/02displacement_prep.R

@@ -1,7 +1,6 @@
 library(tidyverse)
 library(here)
 library(sf)
-library(cowplot)
 
 # Load radiocarbon and site data from the first script
 load(here("analysis/data/derived_data/01data.RData"))
@@ -52,15 +51,7 @@ arendal$name <- "Arendal"
 horten <-  curve_interval(xvals, skoppum_raw)
 horten$name <- "Horten"
 
-# # Transform to BCE/CE using Joe Roe's era package and make BCE negative
-# # to corresponds to oxcAAR later on. [This gave some issues with rounding
-# errors and has been solved differently below]
-# displacement_curves <- rbind(larvik, tvedestrand, arendal, horten)
-# displacement_curves$years <- yr(displacement_curves$years, "cal BP")
-# displacement_curves$years <- yr_transform(displacement_curves$years,
-#                              "BCE") * -1
-# xvals <- as.integer(yr_transform(yr(xvals, "cal BP"), "BCE") * -1)
-
+# # Transform to BCE/CE
 displacement_curves <- rbind(larvik, tvedestrand, arendal, horten)
 displacement_curves$years <- (displacement_curves$years - 1950) * -1
 xvals <- (xvals - 1950) * -1

### 2022-10-20 analysis/script/05site_analysis.R

@@ -25,7 +25,7 @@ quickSetupOxcal()
 set.seed(123)
 
 # Load required functions and prepared data
-source(here("analysis/script/04functions.R"))
+source(here("analysis/script/04analysis_functions.R"))
 load(here("analysis/data/derived_data/01data.RData"))
 load(here("analysis/data/derived_data/02data.RData"))
 

### 2022-10-20 analysis/script/06simulation_results.R

@@ -48,10 +48,6 @@ for(i in 1:length(results_list)){
 
 # Collapse list of lists into a single data frame
 distances <- results %>% bind_rows()
-# distances <-  distances$results
-
-# Lunaveien and Dybdalshei 2 are excluded
-# (see site evaluations in the supplementary material)
 
 # Negative values for Gunnarsrød 5 and Pjonkerød R1 are to be set to zero
 # (see site evaluations in the supplementary material)
@@ -405,3 +401,4 @@ ggsave(file = here("analysis/figures/results2.png"), grd,
 # Save results and exponential functions
 save(distances, expfit, expfit2, expfit3,
      file = here("analysis/data/derived_data/06data.RData"))
+

### 2022-10-20 analysis/script/07shorelinedate.R

@@ -6,10 +6,10 @@ library(sf)
 library(hdrcde)
 
 # For reproducibility
-set.seed(321)
+set.seed(1)
 
 # Load required functions and prepared data
-source(here("analysis/script/04functions.R"))
+source(here("analysis/script/04analysis_functions.R"))
 load(here("analysis/data/derived_data/01data.RData"))
 load(here("analysis/data/derived_data/02data.RData"))
 load(here("analysis/data/derived_data/06data.RData"))

### 2022-10-20 analysis/script/08backwards_dating.R

@@ -9,7 +9,7 @@ library(hdrcde)
 set.seed(1)
 
 # Load required functions and prepared data
-source(here("analysis/script/04functions.R"))
+source(here("analysis/script/04analysis_functions.R"))
 load(here("analysis/data/derived_data/01data.RData"))
 load(here("analysis/data/derived_data/02data.RData"))
 load(here("analysis/data/derived_data/06data.RData"))

### 2022-10-30 analysis/paper/paper (copy).Rmd
Wokring on comments from reviewers

--- 2022-10-20 analysis/paper/paper.Rmd
@@
 # Digital terrain model
-dtm <- rast(here("analysis/data/derived_data/dtm10.tif"))
+# dtm <- rast(here("analysis/data/derived_data/dtm10.tif"))
+dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
 
@@
 load(here("analysis/data/derived_data/08agediff.RData"))
-source(here("analysis/script/04functions.R"), local = knitr::knit_global())
+source(here("analysis/script/04analysis_functions.R"), local = knitr::knit_global())
 
@@
 @nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g.
-@amundsen2006; @solheim2012, 74; @ostmo1988].
+@amundsen2006; @solheim2012, 74; @ostmo1988, see e.g.
+@bergsvik2002; @bergsvik2012 for related perspectives on Neolithic western Norway].
 Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c.

### 2022-10-30 analysis/paper/paper.Rmd
Wokring on comments from reviewers

--- 2022-09-23 analysis/paper/paper.Rmd
@@
 library(stringr)
+library(kableExtra)
 quickSetupOxcal()
@@
 Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g.
-@bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here and adapted to the parameters of post-glacial sea-level change and the Stone Age settlement of southern Norway.
+@bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here and adapted to post-glacial sea-level change and the Stone Age settlement of southern Norway.
 
@@
 @nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g.
-@amundsen2006; @solheim2012, 74; @ostmo1988].
+@amundsen2006; @solheim2012, 74; @ostmo1988; see e.g.
+@bergsvik2002; @bergsvik2012 for similarly nuanced considerations of the coastal settlement of Neolithic western Norway].
 Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c.
@@
 One of the more extensive evaluations of the relationship between archaeological radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area.
-They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.3% of the dates [@solheim2020, 48] .
+They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.3% of the dates [@solheim2020, 48].
 However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates.
@@
 While these values remain the same when only the Mesolithic dates are included (Figure \@ref(fig:results2)B), the mean and standard deviation are slightly constrained.
-Furthermore, while the median for horizontal and topographic distance is only `r round(median(distances2$topodist))`m across all plots in Figure \@ref(fig:results2), the variation in the statistics for dispersion is greater, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
+Furthermore, while the median for horizontal and topographic distance is only `r round(median(distances2$topodist))`m across all plots in Figure \@ref(fig:results2), the magnitude of the statistics for dispersion is greater than what it is for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.
 
@@
 Negative values have been set to zero.
-A) Simulated results older than 2500 BCE, and B) simulated results older than 4000 BCE.", out.width = "525px", fig.align="center", regfloat=TRUE}
+A) Simulated results older than 2500 BCE, and B) simulated results older than 4000 BCE.
+Note that the superimposed gamma distributions have been fit when adding a constant of 0.001 to the distance values, and have been cut off on the y-axis for visualisaton.", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/results2.png"))
@@
 
-An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation (Figure \@ref(fig:results2)).
-While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data.
-However, this can at least in part be related to methodological factors, where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a zero distance.
-If one accepts this, the probability density function for exponential decay can be used to characterise the vertical distance between sites and the shoreline and be used to inform a method for shoreline dating that takes this into account.
+It is clear from the distributions that the simulation results have a severe right skew.
+Most sites were likely situated less than a meter from the shoreline, and from this there is a sharp decline in density as one moves further along the x-axis.
+To characterise this relationship, a series of standard models have been fit to the simulation results older than 2500 BCE by means of maximum likelihood estimation (Table \@ref(tab:tab1), these are visualised in the supplementary material).
+The models have only been fit to positive values, with negative values first forced to zero before adding a constant of 0.001.
+The performances of the models were then compared by means of the Akaike information criterion (AIC) and the Bayesian (or Schwarz) information criterion (BIC).
+The AIC and BIC evaluate the degree to which the models fit to the data, while penalising for the number of model parameters to avoid over-fitting [e.g @burnham2002; for applications in archaeology see e.g.
+@eve2014; @timpson2021].
+As lower values point to a better model, it is evident from both the AIC and BIC results that the gamma is the best model among the candidates.
+This could also have benefited from a more sophisticated treatment of the zero-values, which is related to the fact that these are likely to be a mix of both exact zeros, the case when there is an actual intersection between site and sea, and, although probably to a far lesser extent, zeroes that result from the case when the distance between site and sea is below the detection limit due to the employed methods and the resolution of the spatial data [see e.g.
+@dunn2005; @helsel2005].
+In conclusion, however, the gamma appears to represent a reasonable approximation of the data.
+If one accepts this, the probability density function for the gamma distribution can be used to characterise the vertical distance between sites and the shoreline and be used to inform a method for shoreline dating that takes this into account.
+
+``` {r tab1}
+# Table with model parameters and BIC+AIC scores
+
+mcdf$parameters <- linebreak(mcdf$parameters)
+
+mcdf[,c(1,6, 3, 5)] %>% 
+kableExtra::kable(caption = "Comparison of models fit to the simulated vertical distances older than 2500 BCE, with negative results set to zero and a constant of 0.001 added to the values.
+The models are listed in the order of performance.",
+         col.names = c("Model",
+                    "Parameters",
+                    "AIC",
+                    "BIC"),
+        booktabs = TRUE, escape = FALSE,
+        longtable = TRUE, row.names = FALSE) %>% 
+        column_spec(2, width = "6cm") %>% 
+        kableExtra::kable_styling(latex_options = "striped")
+
+```
 
@@
 This leads the problem to become similar to that of the calibration of a radiocarbon date [see Figure \@ref(fig:exampledat), @bronkramsey2009; @stuvier1989; @vanderplicht1993].
-First, finding the elevation of the sea-level at the time the site was in use is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $D$.
-Based on the simulation results above, the distance from the elevation of the site to the contemporaneous shoreline is defined by the probability density function for exponential decay:
-
-<!-- d = 1(1-\lambda)x -->
+First, finding the elevation of the sea-level at the time the site was in use is dependent on the present day elevation of the site $E$ and the distance between site and the shoreline $D$.
+Based on the simulation results above, the distance from the elevation of the site to the contemporaneous shoreline is defined by the probability density function for the gamma distribution:
+
 \begin{equation}
-  p(\alpha - D) = \lambda {e}^{- \lambda (\alpha - D)}
-  (\#eq:decay)
+p(E - D) = \frac{1}{\sigma ^ {\alpha}\Gamma(\alpha)}(E - D) ^ {\alpha - 1}{e} ^ {-(E - D) / \sigma}
+  (\#eq:gamma)
 \end{equation}
 
-where $\lambda$ is the decay ratio.
+where the parameters $\alpha$ is the shape and $\sigma$ the rate for the distribution, and $\Gamma(\alpha)$ denotes the gamma function.
 This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $T$ for the occupation of the site.
-This is defined by a uniform probability density function ($U$) over the range between the lower $T_{l}$ and upper $T_{u}$ bounds of the displacement curve that has been interpolated to the site location:
+This is defined by a discrete uniform probability mass function ($Ud$) over the range between the lower $T_{l}$ and upper $T_{u}$ bounds of the displacement curve that has been interpolated to the site location:
 
 \begin{equation}
-p(T | \alpha - D) = U[T_{l | \alpha - D}, T_{u | \alpha - D}]	
+p(T | E - D) = Ud[T_{l | E - D}, T_{u | E - D}]	
   (\#eq:rsl)
@@
 \begin{equation}
-p(T | \alpha - D) = p(T | \alpha - D)p(\alpha - D)
+p(T | E - D) = p(T | E - D)p(E - D)
   (\#eq:comb)
@@
 
-<!-- p\left(t\right) = \binom{x}{x} x^x\left(x-x\right)^{x-x} -->
 \begin{equation} 
-  p(T | \alpha) = \sum_{D} p(T | \alpha - D)p(\alpha - D)
+  p(T | E) = \sum_{D} p(T | E - D)p(E - D)
   (\#eq:shoredate)
@@
 
-An example of an implementation of the outlined approach is given in Figure \@ref(fig:exampledat), where $\lambda$ = `r round(expfit$estimate, 3)`.
-This is the decay ratio identified when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
-For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, starting from the site elevation $\alpha$.
-The exponential function is stepped through in it's cumulative form, where the probability from the previous 0.001m step is subtracted from the probability at the current step.
+An example of an implementation of the outlined approach is given in Figure \@ref(fig:exampledat), where $\alpha$ = `r round(gammafit$estimate[1], 3)` and $\sigma$ = `r round(gammafit$estimate[2], 3)`.
+These are the parameters for the gamma distribution identified when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
+For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, which, following from the adjustment of the values for fitting the gamma distribution, starts from 0.001m.
+The gamma distribution is stepped through in it's cumulative form, where the probability from the previous 0.001m step is subtracted from the probability at the current step.
 This probability is then divided equally across the individual calendar years in the range between the lower and the upper limit of the displacement curve at the current 0.001m step.
-The histogram that is the resulting shoreline date is the sum of performing this procedure on all possible 0.001m values of $D$, which, in practice, is until $\alpha - D = 0$ or when 99.999% of the exponential function has been stepped through.
+The probability mass function that is the resulting shoreline date is the sum of performing this procedure on all possible 0.001m values of $D$, which, in practice, is down to and including $E - D = 0.001$ or when 99.999% of the gamma distribution has been stepped through.
 
 To evaluate the outlined procedure it is used to shoreline date the sites from where the method was derived to check if the resulting shoreline dates correspond to the radiocarbon dates associated with the sites (Figure \@ref(fig:shoredate)).
-The Late Neolithic sites are also included here for illustrative purposes, even though these have not informed the decay ratio in use.
+The Late Neolithic sites are also included here for illustrative purposes, even though these have not informed the gamma parameters in use.
 Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits.
 The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by @parnell2008.
-Here, 100,000 age samples drawn from the probability distribution of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding modelled ^14^C-dates.
+Here, 100,000 age samples drawn from the probability mass function of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding probability mass function of the modelled ^14^C-dates.
 The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement (Figure \@ref(fig:shoredatediff)).
@@
 only Mesolithic site phases, the success rate is further increased to `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100)`%).
-The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the likely implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (see also Figure \@ref(fig:results)).
-
-```{r exampledat, echo = FALSE, fig.cap = "Shoreline dating of Hegna vest 1.
-The dashed line marks the mean elevation of the site polygon which is used to inform $\\alpha$ in the dating of the site.
-The exponential function decays with ratio $\\lambda$ from Figure 7A.
-The resulting shoreline date in grey is underlined with the 95\\% HDR in black.", out.width = "500px", fig.align="center", regfloat=TRUE}
+
+```{r exampledat, echo = FALSE, fig.cap = paste0("Shoreline dating of Hegna vest 1.
+The dashed line marks the mean elevation of the site polygon which is used to inform $E$ in the dating of the site.
+The gamma function on the y-axis have the parameters $\\alpha$ = ", round(gammafit$estimate[1], 3), " and $\\sigma$ = ", round(gammafit$estimate[1], 3),  " (see Figure 7A).
+The resulting shoreline date in grey is underlined with the 95\\% HDR in black."), out.width = "500px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/sdate.png"))
@@
 Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change.
-Finally, they are sometimes only meant to indicate a lower bound for when the sites could have been in use.
+Thirdly, they are sometimes only meant to indicate a lower bound for when the sites could have been in use.
+And finally, they are often stated to be the combined result of considering shoreline displacement along with characteristics of local topography, as well as artefact typology--although precisely how these are weighted and used to inform the suggested date is not always clear.
 Overall, the results could, with some danger of circularity, suggest that shoreline dating has generally been applied with a reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g.
@@
 While the implications of a more stable RSL-change for shoreline dating are well known, this also appears to be somewhat under-appreciated in the practical implementation of the method.
-The results also highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% HDRs for the dates.
-In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.
+However, the results also highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% HDRs for the dates.
+In some cases the implementation suggested here provides a very precise date range and in others it offers little more than a *terminus post quem*.
 This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf.
@@
 While this would be trivial in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed.
-This also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to further improve the precision of the dates that can be achieved with shoreline dating.
+This also points to the possibility of drawing on other temporal data, for example within a Bayesian framework [e.g.
+@buck1996], to further improve the precision of the dates that can be achieved with shoreline dating.
 
 ```{r redate, echo = FALSE, fig.cap = "Re-dating excavated and previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
-The 95\\% HDRs in grey are compared to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
+The 95\\% HDRs in grey are compared to the dates originally proposed by the excavation reports in red.
+For clarity in the figure, only the 95\\% HDRs of the shoreline dates are displayed.
+However, the reader is asked to keep in mind that these are associated with a probability mass function that form a better foundation for any further analysis (see e.g.
+Bronk Ramsey).", out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/redate.png"))
@@
 A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \@ref(fig:brunlanes)A.
-The sites have been dated using what @jaksland2014 gives as the lowest elevation of finds at each site, and by employing a exponential decay ratio of `r round(expfit3$estimate, 3)`, to allow for more deviance in the distance between site and shoreline.
-This corresponds to the decay ratio for results older than 7000 BCE in Figure \@ref(fig:results2).
+The sites have been dated using what @jaksland2014 gives as the lowest elevation of finds at each site.
 
@@
 Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does result in very precise dates.
-Above it was suggested that additional temporal data could be combined with the method to improve its accuracy and precision.
+Above it was suggested that additional temporal data could be combined with the method to improve its precision.
 Drawing on @jaksland2014, this example instead highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help to increase the precision of the method if this can be used to exclude certain RSLs as unlikely for when a site was in use.
@@
 Apart from taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves accounting for the distance between the sites and the shoreline.
-When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio of `r round(expfit$estimate, 3)` (Figure \@ref(fig:results2)A) to characterise this relationship.
+When no other information is available, it can at present be recommended to use the empirically derived gamma distribution with a shape of `r round(gammafit$estimate[1], 3)` and scale of `r round(gammafit$estimate[2], 3)` (Figure \@ref(fig:results2)A) to characterise this relationship.
 Furthermore, while this remains to be formalised and explored further, it was also showed how the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data.

### 2022-10-30 analysis/supplementary-materials/supplementary.Rmd
Wokring on comments from reviewers

--- 2022-09-23 analysis/supplementary-materials/supplementary.Rmd
@@
 
+# Load simulation results and model selection
+load(here("analysis/data/derived_data/06data.RData"))
+
 # Load results of re-dating previously shoreline dated sites
@@
 
-## Introduction {-}
-
-This document presents the sensitivity analysis performed for informing the number of simulation runs in the analysis of each site, the analysis of each individual site, and the results of re-dating previously shoreline dated sites.
+## Contents
+
+This document presents the sensitivity analysis performed for informing the number of simulation runs in the analysis of each site, the analysis of each individual site, the evaluation of different for models for characterising the results, and the results of re-dating previously shoreline dated sites.
 All underlying and derived data are available in the online repository for the paper at https://osf.io/7F9SU
 
-## Sensitivity analysis {-}
+## Sensitivity analysis
 
@@
 
-## Site analyses {-}
+## Site analyses
 
@@
 
-## Numerical result of re-dating {-}
+## Model comparison and selection
+
+In the figure below a series of standard models have been fit to the simulation results older than 2500 BCE (presented with negative values set to zero in Figure 7A in the main text).
+The models were fit by means of maximum likelihood estimation.
+Before fitting all models, the values have been adjusted by adding a constant of 0.001 to avoid zero values for the logistic, log normal, power law and gamma distributions.
+The models have only been fit to the positive values.
+This was done both when removing the negative values and when forcing these to zero, before adjusting by 0.001.
+As the difference between these two solutions was negligible and as the assumption here is that the negative values in actuality reflect a distance of zero (see also main text), the result of employing the latter approach is reported on here.
+The performance of the models were then compared by means of the Akaike information criterion (AIC) and the Bayesian (or Schwarz) information criterion (BIC).
+The AIC and BIC evaluate the degree to which the models fit to the data, while penalising for the number of model parameters to avoid over-fitting [e.g @burnham2002; for applications in archaeology see e.g.
+@eve2014; @timpson2021].
+As lower values point to a better model, it is evident from the results that the gamma distribution is the best candidate.
+The approach taken here could have benefited from a more sophisticated treatment of the zero-values, which is related to the fact that these are likely to be a mix of both exact zeros, the case when there is an actual intersection between sea and site, and, although probably to a far lesser extent, zeroes that result from the case when the distance between site and sea is below the detection limit due to the employed methods and the resolution of the spatial data [see e.g.
+@dunn2005; @helsel2005].
+In conclusion, however, the gamma represents a reasonable first approximation.
+
+```{r model_selection, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
+include_graphics(here("analysis/figures/model_comparison.png"))
+```
+
+``` {r}
+# Table with model paramters and BIC+AIC scores
+
+mcdf$parameters <- linebreak(mcdf$parameters)
+
+mcdf[,c(1,6, 3, 5, 4)] %>% 
+kable("latex", caption = "Model comparison using AIC and BIC. The models are listed in the order of performance.",
+         col.names = c("Model",
+                    "Parameters",
+                    "AIC",
+                    "BIC",
+                    "Log-liklihood"),
+        booktabs = TRUE, escape = FALSE,
+        longtable = TRUE, row.names = FALSE) %>% 
+        column_spec(2, width = "6cm") %>% 
+        kable_styling(latex_options = "striped")
+        # kableExtra::kable_styling(latex_options= c("scale_down"))
+
+```
+
+## Numerical result of re-dating
 
@@
          hdrdate = paste0(start95, "\u2013", end95)) %>%
-  select(site_name, elev, repdate, hdrdate, reference)
+  dplyr::select(site_name, elev, repdate, hdrdate, reference)
 
@@
    rowid_to_column("id") %>% 
-   kable(caption = "Shoreline dating previously shorline dated sites. The range of the 95\\% HDR is rounded off to the nearest 10 years to reduce spurious precision.",
+   kable(caption = "The result of shoreline dating previously shorline dated sites using the method proposed here.
+The range of the 95\\% HDR is rounded to the nearest 10 years to reduce spurious precision.",
          col.names = c("",
                      "Site",
-                    "Elevation",
+                    "Reported elevation",
                     "Reported date",
@@
                     "Reference"),
-        booktabs = TRUE,  linesep = c(""), escape = TRUE, longtable = TRUE)
+        booktabs = TRUE,  linesep = c(""), escape = TRUE, longtable = TRUE) %>% 
+  kable_styling(font_size = 8, latex_options = "striped")
 

### 2022-10-30 analysis/script/04analysis_functions.R

@@ -627,6 +627,36 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs,
   #                       crs = st_crs(sitel))))
 }
 
+# Function to only simulate the sea without finding distances
+simulate_sea <- function(nsim, sitel, sitecurve, sitearea, posteriorprobs){
+
+  searasters <- list(length = nsim)
+
+  for(i in 1:nsim){
+    # Set up sampling frame from sum of posterior densities
+    samplingframe <- list(x = posteriorprobs[["dates"]],
+                          y = posteriorprobs[["probabilities"]])
+    # Draw single sample date
+    sdate <- with(samplingframe, sample(x, size = 1, prob = y))
+
+    # Find sealevel-range at sampled date
+    uppers <- approx(sitecurve[,"years"], sitecurve[,"upperelev"],
+                     xout = sdate)[["y"]]
+    lowers <- approx(sitecurve[,"years"], sitecurve[,"lowerelev"],
+                     xout = sdate)[["y"]]
+
+    sealevel <- sample(seq(lowers, uppers, 0.05), 1)
+
+    # Create polygon representing the elevation of the sea level
+    rclmat <- matrix(c(sealevel, Inf, NA, 0, sealevel, 1),
+                     ncol = 3, byrow = TRUE)
+    searasters[[i]] <- raster(terra::classify(sitearea, rclmat))
+  }
+  rs <- stack(searasters)
+  searaster <- calc(rs, sum, na.rm = TRUE)
+  return(searaster)
+}
+
 # Define function to count the number of times the sea is simulated as present
 # for all raster cells
 sea_overlaps <- function(sitearea, seapolygons){
@@ -1043,8 +1073,8 @@ shoreline_date <- function(sitename, elev = dtm,
                            disp_curves = displacement_curves,
                            sites = sites_sa,
                            iso = isobases,
-                           reso = 0.1,
-                           expratio, siteelev = "mean",
+                           reso = 0.1, exponential = TRUE,
+                           modelfit, siteelev = "mean",
                            specified_elev = NA,
                            sitelimit = TRUE,
                            features = NA){
@@ -1082,14 +1112,24 @@ shoreline_date <- function(sitename, elev = dtm,
     }
   }
 
+  if(exponential){
   inc <- seq(0, as.numeric(siteelev), reso)
[…]

### 2022-10-30 analysis/script/06simulation_results.R

@@ -8,6 +8,13 @@ library(gridExtra)
 library(gtable)
 library(cowplot)
 library(MASS)
+library(poweRlaw)
+library(AICcmodavg)
+
+# This script treats the result of the simulations. This involves visualisations
+# of the results as well as model fitting and comparison (which is treated in
+# the supplementary material)
+
 
 # List all files except those starting with a number.
 # That is, all data files resulting from analysis in 05script.R
@@ -48,6 +55,8 @@ for(i in 1:length(results_list)){
 
 # Collapse list of lists into a single data frame
 distances <- results %>% bind_rows()
+
+##### Visualisation of distance measures across time ########
 
 # Negative values for Gunnarsrød 5 and Pjonkerød R1 are to be set to zero
 # (see site evaluations in the supplementary material)
@@ -202,8 +211,157 @@ grd <- plot_grid(grd1, grd2, labels = c("A", "B"), ncol = 1, scale = 0.9)
 ggsave(file = here("analysis/figures/results.png"), grd,
        bg = "white", width = 250, height = 200, units = "mm")
 
+
+###### Model comparison using density of vertical distance measures ########
+
 # Measures for vertical distance to inform recommendations for shoreline dating.
 # Using only corresponding dates, including dates to agricultural activity
+
+# Vertical distances (older than 2500 BCE) , including negative values
+vertdistneg <- distances %>%
+  filter(year <= -2500 & rcarb_cor == "t")
+vertdistneg$vertdist <- vertdistneg$vertdist + 0.001
+
+# Vertical distances, forcing negative values to zero
+vertdist <- distances %>%
+  filter(year <= -2500 & rcarb_cor == "t") %>%
+  mutate(vertdist = ifelse(vertdist < 0, 0, vertdist))
+vertdist$vertdist <- vertdist$vertdist + 0.001
+
+# Vertical distances, excluding negative values
+vertdistna <- distances %>%
+  filter(year <= -2500 & rcarb_cor == "t") %>%
+  mutate(vertdist = ifelse(vertdist < 0, NA, vertdist)) %>%
+  filter(!is.na(vertdist))
+vertdistna$vertdist <- vertdistna$vertdist + 0.001
+
+# Fit exponential function to only positive values
+expfit <- fitdistr(vertdist$vertdist, "exponential")
+expdat <- data.frame(
[…]

### 2022-10-30 analysis/script/07shorelinedate.R

@@ -31,12 +31,16 @@ sitecurve <- interpolate_curve(years = xvals,
 sitecurve$name <- sitename
 siteelev <- terra::extract(dtm, vect(sitel), fun = mean)[2]
 
-inc <- seq(0, as.numeric(siteelev), 0.001)
+inc <- seq(0.001, as.numeric(siteelev), 0.001)
 
 expdat <- data.frame(
   offset = inc,
   # Cumulative exponential distribution function
-  px = pexp(inc, rate = expfit$estimate)) %>%
+  # px = pexp(inc, rate = expfit$estimate)) %>%
+  # px = plnorm(inc, meanlog = lognormfit$estimate[1],
+  #             sdlog = lognormfit$estimate[2]))  %>%
+  px = pgamma(inc, shape = gammafit$estimate[1],
+              rate = gammafit$estimate[2])) %>%
   # Probability for each elevation increment is given by subtracting the
   # previous cumulative probability
   mutate(probs = px - lag(px, default =  dplyr::first(px))) %>%
@@ -50,7 +54,7 @@ dategrid <- data.frame(
 for(i in 1:nrow(expdat)){
   adjusted_elev <- as.numeric(siteelev - expdat$offset[i])
   if(!(adjusted_elev > 0)) { # Do not allow sea-level to go below present
-    adjusted_elev <- 0.01
+    adjusted_elev <- 0.001
   }
   # Find lower date
   lowerd <- round(approx(sitecurve[,"lowerelev"],
@@ -109,18 +113,17 @@ ggplot() +
   theme_bw() +
   theme(legend.position = "None")
 
-# , breaks = c(0, 20, 40, as.numeric(siteelev), 80),
-# labels = c("0", "20", "40", expression(alpha), "80")
-
 #  For plotting purposes to close the geom_polygon on the y-axis
 expdatt <- rbind(c(0, 0, 0), expdat)
 
 # Code taken from oxcAAR to plot density to y-axis
 x_extent <- ggplot_build(plt)$layout$panel_scales_x[[1]]$range$range
-expdatt$probs_scaled <- expdatt$probs / max(expdatt$probs) *
-  diff(x_extent)/6 + x_extent[1]
+expdatt$probs_scaled <- expdatt$probs / max(expdatt$probs) * 1 *
+  diff(x_extent) + x_extent[1]
 
-dplt <- plt + geom_polygon(data = expdatt, aes(x =  probs_scaled,
+
+dplt <-
[…]

### 2022-10-30 analysis/script/08backwards_dating.R

@@ -18,11 +18,14 @@ load(here("analysis/data/derived_data/06data.RData"))
 # this is distributed as tiles which have to be merged by first
 # running 00dtm_prep.R
 dtm <- rast(here("analysis/data/derived_data/dtm10.tif"))
+dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
 
 # List all files except those starting with a number.
 # That is, all data files resulting from analysis in 05script.R
 datfiles <- grep("^[0-9]", list.files(here("analysis/data/derived_data")),
                  invert = TRUE, value = TRUE)
+# Exclude the temp directory
+datfiles <- datfiles[ !grepl("sitearea_temp", datfiles) ]
 
 # Create empty list to hold dat data
 posterior_list <- list()
@@ -56,8 +59,10 @@ sites_sa <- sites_sa %>%
 # (see supplementary material)
 feature_sites <- c("Langangen Vestgård 7", "Vallermyrene 2")
 
+
+# Load shoreline date results (uncomment below to rerun)
 load(here("analysis/data/derived_data/08data.RData"))
-# # Uncomment to rerun
+
 # shorelinedates <- list()
 # for(i in 1:nrow(sites_sa)){
 #   print(paste(i, sites_sa$name[i]))
@@ -67,7 +72,9 @@ load(here("analysis/data/derived_data/08data.RData"))
 #                                           disp_curves = displacement_curves,
 #                                           sites = sites_sa,
 #                                           iso = isobases,
-#                                           expratio = expfit$estimate,
+#                                           exponential = FALSE,
+#                                           modelfit = gammafit,
+#                                           # expratio = expfit$estimate,
 #                                           siteelev = "mean",
 #                                           reso = 0.001,
 #                                           specified_elev = NA,
@@ -79,19 +86,19 @@ load(here("analysis/data/derived_data/08data.RData"))
 #                                         disp_curves = displacement_curves,
 #                                         sites = sites_sa,
 #                                         iso = isobases,
-#                                         expratio = expfit$estimate,
+#                                         exponential = FALSE,
+#                                         modelfit = gammafit,
+#                                         # expratio = expfit$estimate,
 #                                         siteelev = "mean",
 #                                         reso = 0.001,
[…]

### 2022-10-30 analysis/script/mdl.R

@@ -1,6 +1,7 @@
 library(tidyverse)
 library(MASS)
 library(poweRlaw)
+library(AICcmodavg)
 
 # Load simulation results
 load(here::here("analysis/data/derived_data/06data.RData"))
@@ -8,17 +9,24 @@ load(here::here("analysis/data/derived_data/06data.RData"))
 # Vertical distances, including negative values
 vertdistneg <- distances %>%
   filter(year <= -2500 & rcarb_cor == "t")
+vertdistneg$vertdist <- vertdistneg$vertdist + 0.001
+
+vertdisto <- distances %>%
+  filter(year <= -2500 & rcarb_cor == "t") %>%
+  mutate(vertdist = ifelse(vertdist < 0, 0, vertdist))
 
 # Vertical distances, forcing negative values to zero
 vertdist <- distances %>%
   filter(year <= -2500 & rcarb_cor == "t") %>%
   mutate(vertdist = ifelse(vertdist < 0, 0, vertdist))
+vertdist$vertdist <- vertdist$vertdist + 0.001
 
 # Vertical distances, excluding negative values
 vertdistna <- distances %>%
   filter(year <= -2500 & rcarb_cor == "t") %>%
   mutate(vertdist = ifelse(vertdist < 0, NA, vertdist)) %>%
   filter(!is.na(vertdist))
+vertdistna$vertdist <- vertdistna$vertdist + 0.001
 
 # Fit normal distribution to all values
 normfit <- MASS::fitdistr(vertdistneg$vertdist, "normal")
@@ -28,41 +36,116 @@ normdat <- data.frame(
              mean =  normfit$estimate[1],
              sd =  normfit$estimate[1]))
 
-# Fit exponential functions to only positive values
+# Fit exponential function to only positive values
 expfit <- fitdistr(vertdist$vertdist, "exponential")
 expdat <- data.frame(
   x = vertdist$vertdist,
   px = dexp(as.numeric(vertdist$vertdist), rate = expfit$estimate))
+expdatcum <- data.frame(
+  x = vertdist$vertdist,
+  px = pexp(as.numeric(vertdist$vertdist), rate = expfit$estimate))
 
-# Fit exponential functions to only positive values
-expfitna <- fitdistr((vertdistna$vertdist+0.1), "exponential")
+# Fit exponential function to only positive values
+expfitna <- fitdistr(vertdistna$vertdist, "exponential")
[…]

### 2022-11-09 analysis/paper/paper.Rmd
Changes after peer review

@@
    - \linenumbers
+   - \usepackage[font={small}]{caption}
 author: 
@@
 date: "`r format(Sys.time(), '%d %B, %Y')`"
+# always_allow_html: yes # Uncomment if compiling to word
 output:
@@
   Shoreline dating; Stone Age; Settlement patterns; Scandinavia; Relative sea-level change
+nocite: | 
+  @telford2004
 ---
@@
 library(kableExtra)
+library(english)
 quickSetupOxcal()
@@
 
-# Load processed site and radiocarbon data, and load functions
+# Load preprocessed site and radiocarbon data, results, and load functions
 load(here("analysis/data/derived_data/01data.RData"))
@@
 load(here("analysis/data/derived_data/08agediff.RData"))
-source(here("analysis/script/04analysis_functions.R"), local = knitr::knit_global())
-
-# Exclude Lunaveien and Dybdalshei 2 (See supplementary material)
-sites_sa <- filter(sites_sa, !(name %in% c("Dybdalshei 2", "Lunaveien")))
+load(here("analysis/data/derived_data/08posterior_rcarbon.RData"))
+load(here("analysis/data/derived_data/09data.RData"))
+source(here("analysis/script/04analysis_functions.R"),
+       local = knitr::knit_global())
+
+# Exclude Lunaveien, Dybdalshei 2 and Frebergsvik C (see supplementary material)
+sites_sa <- filter(sites_sa, !(name %in% c("Dybdalshei 2", 
+                                           "Lunaveien", "Frebergsvik C")))
+
+# Do the same for radiocarbon dates intersecting the Stone Age.
+rcarb_sa <- filter(rcarb_sa, site_name %in% unique(sites_sa$name))
 
@@
 If the site is situated on a ledge in a steep and jagged area of the present-day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft.
-Although it appears that the arguments for such site locations can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @ilves2011].
+Although it appears that the arguments for such site locations can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @berg-hansen2022, 644; @ilves2011].
 
@@
 
-Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
-@akerlund1995; @åstveit2018; @solheim2020; see also @bjerck2008a; @kleppe1985; @ramstad2009].
+Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability mass functions plotted against local shoreline displacement curves based on the elevation of the site [e.g.
+@akerlund1995; @åstveit2018; @berg-hansen2022; @solheim2020; see also @bjerck2008a; @kleppe1985; @ramstad2009].
 This approach has a couple of limitations.
 First, the displacement curves are sometimes applied directly to larger study areas, analogous to what @borreggine2022 term a bathtub model, with only some studies having taken the variable uplift-rates into account when performing this comparison [e.g.
-@åstveit2018; @fossum2020; @møller1987; @persson2008].
+@åstveit2018; @fossum2020; @møller1987; @persson2008; @rosenvinge2022].
 Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that the conclusion supports the hypothesis.
@@
 The highest astronomical tide in the study area reaches around 30cm above mean sea-level [30cm at the standard port Helgeroa in Larvik, @mapping2021].
-Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs as they are the integrated result of both expert knowledge and more objectively quantifiable parameters.
-The reason for this is that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but they are for example also dependent on the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, as well as being based on expert knowledge of regional post-glacial geologic developments and local geomorphology, to name but a few factors [e.g.
+
+As the displacement curves and their trajectory are quite complex constructs and the integrated result of both expert knowledge and more objectively quantifiable parameters, the geologists that have undertaken the studies have not found reason to assign variable uncertainty within the confidence envelopes of the displacement curves [@sørensen2014, 44; @romundset2018, 187].
+The reason for this is that the trajectory of the curves is not only based on radiometric dates, the uncertainty of which are well defined, but are for example also dependent on the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, as well as being based on expert knowledge of regional post-glacial geologic developments and local geomorphology, to name but a few factors [e.g.
 @romundset2011; @romundset2018; @svendsen1987; for an alternative approach see @creel2022].
@@
 The archaeological data compiled for the analysis consists of excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)).
-These number `r nrow(sites)` sites of which `r length(unique(rcarbon$site_name))` sites are associated with a total of `r nrow(rcarbon)` radiocarbon dates.
+These number `r nrow(sites)` sites, of which `r length(unique(rcarbon$site_name))` are associated with the total of `r nrow(rcarbon)` radiocarbon dates.
 Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon date ranges that intersect the Stone Age (9500--1700 BCE), with 95% probability.
@@
 Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of RSL change at that location, and a DTM edited to remove substantial modern disturbances, the simulations were performed.
-A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \@ref(fig:example-2)).
+A single simulation run involved first drawing a single year weighted by the posterior probability distribution of a given occupation phase of a site (Figure \@ref(fig:example-2)).
 This year then has a corresponding likely elevation range for the contemporaneous shoreline from which an elevation value was drawn uniformly, using intervals of 5cm.
@@
 This was done using the topoDistance package for R [@wang2019].
-The topographic distance was measured between the site polygon and the horizontally closest point on the shoreline.
-This means that the distance is not necessarily measured as the closest topographic distance to the shoreline, but rather as the shortest topographic path to the horizontally closest point on the shoreline.
-Not finding the topographically closest point significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact on the results given the distances considered.
+The topographic distance was measured between the points on the site and sea polygons that were identified as being the closest when measured horizontally.
+Because it is measured as the shortest topographic path between the horizontally closest points, this means that the distance does not necessarily match the closest topographic distance if the entirety of the polygons had been considered.
+Not finding the topographically closest points significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact on the results, given the distances considered.
 The shortest topographic path was found using the Moore neighbourhood of eight cells [e.g.
 @conolly2006, 253; @herzog2013].
+
 In the case where the sea polygons intersect the site polygon, all distance measures were set to zero.
@@
 
-This process was repeated 1000 times for each phase for each site.
-The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5, available in the supplementary material [cf.
+This process was repeated 1000 times for each phase for each site (Figure \@ref(fig:example-3)).
+The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5 [cf.
 @crema2010, 1125].
-Hovland 5 was chosen for this evaluation as it has an imprecise age and is located in area of quite complex topography.
+This evaluation is presented in the supplementary material.
+Hovland 5 was chosen for this assessment as it has an imprecise age and is located in area of quite complex topography.
 
@@
 ```{r example-2, echo = FALSE, fig.cap = "Example of a single simulation run on the site Hegna vest 1.
-A) The simulation starts by drawing a single year from the posterior density estimate.
+A) The simulation starts by drawing a single year, weighted by the the posterior probability distribution.
 B) This then corresponds to an elevation range on the interpolated displacement curve.
@@
 # Map of simulated sea-levels, date group 1
-simsea_plot1 <- shore_plot(sitearea, output[[1]]$simsea, sitel, 150, date_groups,
-                      s_tdist = 0.5, s_xpos = 1300, s_ypos = 800,
-                      s_bheight = 0.3, s_tsize = 3)
+simsea_plot1 <- shore_plot(sitearea, output[[1]]$simsea, sitel,
+                           150, date_groups, s_tdist = 0.5, s_xpos = 1300,
+                           s_ypos = 800, s_bheight = 0.3, s_tsize = 3)
 
@@
 # Map of simulated sea-levels, date group 2
-simsea_plot2 <- shore_plot(sitearea, output[[2]]$simsea, sitel, 150, date_groups,
-                      s_tdist = 0.5, s_xpos = 1300, s_ypos = 800,
-                      s_bheight = 0.3, s_tsize = 3)
+simsea_plot2 <- shore_plot(sitearea, output[[2]]$simsea, sitel, 150,
+                           date_groups, s_tdist = 0.5, s_xpos = 1300,
+                           s_ypos = 800, s_bheight = 0.3, s_tsize = 3)
 
@@
 ```{r distances}
+# Various subsets for which some statistics are reported in the text.
+
+distcor <- filter(distances, rcarb_cor == "t")
+
+# Results older than 7500 BCE
+r1 <- distcor %>% filter(year < -7500)
+
+# Statistics for vertical distance
+rs1 <- r1 %>% 
+  summarise(median = median(vertdist),
+          max = max(vertdist),
+          min = min(vertdist))
+
+# Number of sites with results higher than 5m
+r1a5 <- r1 %>% group_by(sitename) %>% summarise(min = min(vertdist)) %>%
+  filter(min < 5) %>% nrow()
+
+# Results older than 2500 BCE
+r2 <- distcor %>% filter(year > -2500)
+
+# Statistics
+rs2 <- r2 %>% 
+  summarise(
+    mediandist = median(vertdist),
+    maxdist = max(vertdist)
+  )
+
+# Number of sites with results that include a vertical distance of zero
+r2a0 <- r2 %>% group_by(sitename) %>% summarise(min = min(vertdist)) %>%
+  filter(!(min > 0)) %>% nrow()
+
+# Results between 4000 BCE and 2500 BCE
+r3 <- distcor %>% filter(year < -2500 & year > -4000)
+
+# Statistics
+rs3 <- r3 %>% 
+  summarise(median = median(vertdist))
+
+# Number of sites with results higher than 25m
+r3a25 <- r3 %>% group_by(sitename) %>% summarise(min = min(vertdist)) %>% 
+  filter(min > 25) %>% nrow()
+
+# Data when setting negative results to zero, to be used in the text
 distances2 <- distances %>%
@@
          vertdist = ifelse(vertdist < 0, 0, vertdist))
+
 ```
@@
 Overall, as is indicated by the measures for central tendency and the almost solid line along the 0m mark on the y-axes, the simulations show that the sites tend to have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)).
-Some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included (Figure \@ref(fig:results)A), but if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, Figure \@ref(fig:results)B gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
-The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the rapid RSL fall in the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the simulated sea-level.
+As is also illustrated by the measures for dispersion, some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included (Figure \@ref(fig:results)A).
+However, if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, Figure \@ref(fig:results)B gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline.
+The distances for the earliest sites appears somewhat high, with the highest vertical distance of the results older than 7500 BCE being `r round(rs1$max, 1)`m.
+But this can likely be explained as the result of the rapid RSL fall in the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the simulated sea-level.
+This is also indicated by the fact that the median vertical distance for the same simulation results is `r round(rs1$median, 1)`m, and `r r1a5` of the `r length(unique(r1$sitename))` sites associated with these results have simulated vertical distances that extend below 5m.
+
 Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.
-After around 2500 BCE several sites are situated a considerable distance from the reconstructed shoreline, and while a couple remain horizontally and topographically close, most appear to be considerably elevated above the sea-level, as indicated on the plot for vertical distance.
+After around 2500 BCE several sites are situated a considerable distance from the reconstructed shoreline, and while a couple remain horizontally and topographically close, most appear to be considerably elevated above the sea-level.
+Of the results from after 2500 BCE, which are associated with `r length(unique(r2$sitename))` sites, only `r as.english(r2a0)` has simulation results for vertical distance that includes zero.
+The highest simulated vertical distance among these is `r round(rs2$max, 1)`m and the median is `r round(rs2$median, 1)`m.
 While the sample size is limited, there are also a couple of sites located some distance from the shoreline just after 4000 BCE.
+Of the `r length(unique(r3$sitename))` sites associated with the period between 4000 and 2500 BCE, `r r3a25` sites have all vertical distance results above 25m.
+However, the median vertical distance of the results from this period is only `r round(rs3$median, 1)`m, indicating that while some sites have a markedly withdrawn location, most are still situated close to the shoreline.
 The chronological smearing following from the uncertainty in the ^14^C-dates means that while the results cannot be used to directly inform discussions that deal with the century scale around these chronological transitions [e.g.
-@prescott2020; @solheim2021], the findings are nonetheless in clear agreement with the general chronological framework in the literature.
+@prescott2020; @solheim2021], the findings are nonetheless in clear agreement with the general chronological developments suggested in the literature.
 
 The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3.
-These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape [@reitan2022b].
+@berg-hansen2022[644] made a similar observation in their assessment of the correspondence between shoreline displacement and radiocarbon dates from these sites.
+These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022].
 While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent an upper constraining limit for the sea-level, as they would not have been in use when located under water.
@@
 
+# Number of site phases for all dates and only dates corresponding to site
+# inventories.
+nphases <- ardates %>%  group_by(site_name) %>% distinct(group) %>% nrow()
+nphasescor <- rdates %>% group_by(site_name) %>% distinct(group) %>% nrow()
+
+# Number of sites for simulation results older than 2500 BCE
+nsitesdist25 <- rdates %>% filter(dates < -2500 & probabilities > 0) %>%
+  group_by(site_name) %>% distinct(site_name) %>% nrow()
+
+# Number of sites for simulation results older than 4000 BCE
+nsitesdist4 <- rdates %>% filter(dates < -4000 & probabilities > 0) %>%
+  group_by(site_name) %>% distinct(site_name) %>% nrow()
+
+# Number of site phases for simulation results older than 2500 BCE
+nphasesdist25 <- rdates %>% filter(dates < -2500 & probabilities > 0) %>%
+  group_by(site_name) %>% distinct(group) %>% nrow()
+# Number of site phases for simulation results older than 4000 BCE
+nphasesdist4 <- rdates %>% filter(dates < -4000 & probabilities > 0) %>%
+  group_by(site_name) %>% distinct(group) %>% nrow()
+
+# Correspondence between shoreline and radiocarbon dates.
 # Subtract one for Gunnarsrød 5 (see text)
@@
 
-```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites.
+```{r results, echo = FALSE, fig.cap = paste0("The result of running the analysis across all sites.
 Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occurred.
-Results associated with agricultural activites are plotted in grey.
-The first row A) shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites.
-The second row B) shows the result of excluding these.
-The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center", regfloat=TRUE}
+Results associated with agricultural activities are plotted in grey.
+The first row A) shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites (", length(unique(distances$sitename)) ," sites and ", nphases ," site phases).
+The second row B) shows the result of excluding these (resulting in ", length(unique(distances[distances$rcarb_cor == "t",]$sitename)), " sites and ", nphasescor ," site phases).
+The table under each plot lists some corresponding statistics for central tendency and dispersion."), out.width = "500px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/results.png"))
@@
 
-```{r results2, echo = FALSE, fig.cap = "Histograms showing the simulated distance from the shoreline using radiocarbon dates corresponding to the site inventories.
+```{r results2, echo = FALSE, fig.cap = paste0("Histograms showing the simulated distance from the shoreline using radiocarbon dates corresponding to the site inventories.
 Negative values have been set to zero.
-A) Simulated results older than 2500 BCE, and B) simulated results older than 4000 BCE.
-Note that the superimposed gamma distributions have been fit when adding a constant of 0.001 to the distance values, and have been cut off on the y-axis for visualisaton.", out.width = "525px", fig.align="center", regfloat=TRUE}
+A) Simulated results older than 2500 BCE (", nsitesdist25 , " sites and ", nphasesdist25 ," site phases) and B) simulated results older than 4000 BCE (", nsitesdist4 , " sites and ", nphasesdist4 ," site phases).
+Note that the cut-off is done based on the calendar year associated with each distance value.
+Consequently, sites and site phases are only completely excluded if the entire posterior probability of the radiocarbon dates falls earlier than the cut-off.
+Furthermore, the superimposed gamma distributions have been fit when adding a constant of 0.001 to the distance values and have been cut off on the y-axis for visualisaton.
+The gamma distribution in A forms the basis for the analysis to follow, but a version has also been fit to the vertical distances in B to further illustrate the difference between the distributions."), out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/results2.png"))
@@
 
-It is clear from the distributions that the simulation results have a severe right skew.
-Most sites were likely situated less than a meter from the shoreline, and from this there is a sharp decline in density as one moves further along the x-axis.
-To characterise this relationship, a series of standard models have been fit to the simulation results older than 2500 BCE by means of maximum likelihood estimation (Table \@ref(tab:tab1), these are visualised in the supplementary material).
-The models have only been fit to positive values, with negative values first forced to zero before adding a constant of 0.001.
-The performances of the models were then compared by means of the Akaike information criterion (AIC) and the Bayesian (or Schwarz) information criterion (BIC).
-The AIC and BIC evaluate the degree to which the models fit to the data, while penalising for the number of model parameters to avoid over-fitting [e.g @burnham2002; for applications in archaeology see e.g.
+It is clear that the distributions in Figure \@ref(fig:results2) have a severe right skew.
+Most sites were likely situated less than a meter from the shoreline, and from this there is a sharp decline in density as one moves further along the x-axes.
+To characterise this relationship, a series of standard models for distributions with a right skew have been fit to the simulation results for vertical distance older than 2500 BCE (Figure \@ref(fig:results2)A) by means of maximum likelihood estimation (Table \@ref(tab:tab1)).
+As most of the models only accept positive values a constant of 0.001 was added to avoid values of zero.
+It was attempted to both remove negative values and force these to zero before adding the constant.
+As the difference between these two solutions was negligible, and as the assumption here is that negative values in actuality reflect a distance of zero, the latter approach was chosen (a plot displaying the negative values and the compared models is available in the supplementary material).
+
+The performance of the models was then compared by means of the Akaike information criterion (AIC) and the Bayesian (or Schwarz) information criterion (BIC).
+The AIC and BIC evaluate the degree to which the models fit to the data, while penalising for the number of model parameters to avoid over-fitting [e.g.
+@burnham2002; for applications in archaeology see e.g.
 @eve2014; @timpson2021].
-As lower values point to a better model, it is evident from both the AIC and BIC results that the gamma is the best model among the candidates.
-This could also have benefited from a more sophisticated treatment of the zero-values, which is related to the fact that these are likely to be a mix of both exact zeros, the case when there is an actual intersection between site and sea, and, although probably to a far lesser extent, zeroes that result from the case when the distance between site and sea is below the detection limit due to the employed methods and the resolution of the spatial data [see e.g.
+As lower values point to a better model, it is evident from both the AIC and BIC that the gamma is the best among the candidate models.
+It is worth noting that this could have benefited from a more sophisticated treatment of the zero-values.
+This is because these are likely to be a mix of both exact zeros, the case when there is an actual intersection between site and sea, and, although probably to a far lesser extent, zeroes that result from the case when the distance between site and sea is below the detection limit due to the employed methods and the resolution of the spatial data [e.g.
 @dunn2005; @helsel2005].
@@
 kableExtra::kable(caption = "Comparison of models fit to the simulated vertical distances older than 2500 BCE, with negative results set to zero and a constant of 0.001 added to the values.
-The models are listed in the order of performance.",
+The models are listed in the order of performance.
+A plot with all of the models is available in the supplementary material.",
          col.names = c("Model",
@@
 
-where the parameters $\alpha$ is the shape and $\sigma$ the rate for the distribution, and $\Gamma(\alpha)$ denotes the gamma function.
+wheres $\alpha$ is the shape and $\sigma$ the rate of the distribution, and $\Gamma(\alpha)$ denotes the gamma function.
 This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $T$ for the occupation of the site.
@@
 An example of an implementation of the outlined approach is given in Figure \@ref(fig:exampledat), where $\alpha$ = `r round(gammafit$estimate[1], 3)` and $\sigma$ = `r round(gammafit$estimate[2], 3)`.
-These are the parameters for the gamma distribution identified when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A).
+These are the parameters for the gamma distribution identified when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A) and are the parameters used in all applications of the proposed method that follow below.
 For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, which, following from the adjustment of the values for fitting the gamma distribution, starts from 0.001m.
@@
 ```{r exampledat, echo = FALSE, fig.cap = paste0("Shoreline dating of Hegna vest 1.
-The dashed line marks the mean elevation of the site polygon which is used to inform $E$ in the dating of the site.
-The gamma function on the y-axis have the parameters $\\alpha$ = ", round(gammafit$estimate[1], 3), " and $\\sigma$ = ", round(gammafit$estimate[1], 3),  " (see Figure 7A).
+The mean elevation of the site polygon is used to inform $E$ in the dating of the site.
+The gamma distribution in blue on the y-axis extends the full range of possible values for $E - D$ and has the parameters $\\alpha$ = ", round(gammafit$estimate[1], 3), " and $\\sigma$ = ", round(gammafit$estimate[1], 3),  " (see Figure 7A).
+The red line marks the shoreline displacement curve interpolated to the site location.
 The resulting shoreline date in grey is underlined with the 95\\% HDR in black."), out.width = "500px", fig.align="center", regfloat=TRUE}
@@
 
-```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory using the method proposed here.
+```{r shoredate, echo = FALSE, fig.cap = paste("The result of backwards shoreline dating the", length(unique(distances[distances$rcarb_cor == "t",]$sitename)),  "sites with radiocarbon dates corresponding to the artefact inventory using the method proposed here.
 The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black.
-These are plotted against the modelled radiocarbon dates, which are given colour from oldest to youngest occupation phase for each site, defined by non-intersecting dates at 99.7\\% probability.", out.width = "525px", fig.align="center", regfloat=TRUE}
+These are plotted against the modelled radiocarbon dates, which are given colour from oldest to youngest occupation phase for each site, defined by non-intersecting dates at 99.7\\% probability."), out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/shoredate.png"))
@@
 
-To further explore the implementation for shoreline dating presented above, excavated and shoreline dated Stone Age sites within the study area where ^14^C-dates are not available or these are not believed to date the main occupation of the sites have been subjected to the outlined approach (Figure \@ref(fig:redate)).
+To further explore the implementation for shoreline dating presented above, `r length(unique(bdates2$site_name))` excavated and shoreline dated Stone Age sites within the study area where ^14^C-dates are not available or these are not believed to date the main occupation of the sites have been subjected to the outlined approach (Figure \@ref(fig:redate)).
 The resulting dates are compared to those originally proposed in the excavation reports for the sites (the numerical results are available in the supplementary material).
-To avoid issues with recent disturbances on the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
+To avoid issues with recent disturbances in the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site.
 As all of the included sites have been excavated after the turn of the millennium, and the wide adoption of GNSS technology, the reported elevations should be trustworthy.
 
-The comparison with previously reported dates is an illustrative, but unfair exercise for a few reasons.
+This comparison is useful for illustrating both how the method has previously been employed, and for revealing nuances of the implementation that is proposed here.
+However, the comparison is also unfair to the previously proposed dates for a few reasons.
 First, the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an undefined, but implied or explicit uncertainty range.
@@
 Thirdly, they are sometimes only meant to indicate a lower bound for when the sites could have been in use.
-And finally, they are often stated to be the combined result of considering shoreline displacement along with characteristics of local topography, as well as artefact typology--although precisely how these are weighted and used to inform the suggested date is not always clear.
-Overall, the results could, with some danger of circularity, suggest that shoreline dating has generally been applied with a reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g.
+Additionally, the dates are often stated to be the result of also considering artefact typology and characteristics of local topography to inform the likely elevation of the sea when the site was in use---although precisely how these are weighted and used to inform the suggested date is often not as clear.
+
+With a couple exceptions, the previously hypothesised dates and the ones achieved here appear to roughly correspond when it comes to the start date for the occupation of the sites.
+The most clear difference mainly pertains to the fact that the previously proposed date ranges are, without exception, more constrained than the 95% HDRs resulting from the proposed method.
+Considering the right skew of the probability mass functions underlying the 95% HDRs and the general overlap for the start dates, these results could, with some danger of circularity, suggest that shoreline dating has generally been applied with a reasonable degree of success.
+This also follows from the fact that these dates have typically informed research in an approximate manner [although see e.g.
 @roalkvam2022].
+
 With these considerations in mind, the results also indicate that shoreline dating has at times been applied with an exaggerated degree of precision.
 While the implications of a more stable RSL-change for shoreline dating are well known, this also appears to be somewhat under-appreciated in the practical implementation of the method.
-However, the results also highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% HDRs for the dates.
-In some cases the implementation suggested here provides a very precise date range and in others it offers little more than a *terminus post quem*.
+The results indicate that the spatial and temporal contingency of the method is better captured by the implementation suggested here, as is illustrated by the variation in the range of the 95% HDRs for the dates.
+In some cases the proposed method provides a fairly precise date and in others it offers little more than a *terminus post quem*.
 This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf.
 Figure \@ref(fig:map-iso)B).
-Furthermore, as some of the date ranges extend well beyond major chronological divisions, even into the Iron Age, they could be severely and securely constrained with only cursory reference to typology.
+However, as some of the 95% HDRs extend well beyond major chronological divisions, even into the Iron Age, it is also clear that some of these could be severely and securely constrained with only cursory reference to typology.
 While this would be trivial in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed.
-This also points to the possibility of drawing on other temporal data, for example within a Bayesian framework [e.g.
-@buck1996], to further improve the precision of the dates that can be achieved with shoreline dating.
-
-```{r redate, echo = FALSE, fig.cap = "Re-dating excavated and previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
+This also points to the possibility of drawing on other temporal data to further improve the precision of the dates that can be achieved with shoreline dating.
+
+```{r redate, echo = FALSE, fig.cap = paste("Re-dating", length(unique(bdates2$site_name)), "excavated and previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories.
 The 95\\% HDRs in grey are compared to the dates originally proposed by the excavation reports in red.
 For clarity in the figure, only the 95\\% HDRs of the shoreline dates are displayed.
-However, the reader is asked to keep in mind that these are associated with a probability mass function that form a better foundation for any further analysis (see e.g.
-Bronk Ramsey).", out.width = "525px", fig.align="center", regfloat=TRUE}
+However, the reader is asked to keep in mind that these are associated with a probability mass function with a right skew that form a better foundation for any further analysis (see e.g.
+Telford et al.
+2004)."), out.width = "525px", fig.align="center", regfloat=TRUE}
 knitr::include_graphics(here("analysis/figures/redate.png"))
@@
 Drawing on @jaksland2014, this example instead highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help to increase the precision of the method if this can be used to exclude certain RSLs as unlikely for when a site was in use.
-One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline.
-If this method is developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use.
+One potential way to do this could be through the analysis of phosphate concentrations in soils, which has the potential to offer insights on the likely position of the shoreline when a site was in use [@ilves2011].
+This has been done in the Baltic sea region [e.g.
+@broadbent1979; @sundström2006; @ilves2011], but has yet to provide reliable results in Norway [e.g.
+@melvold2014a; @viken2018b].
+The identification of other physical traces of shore formation processes and the deposition of beach sediments in relation to archaeological material also holds similar potential [e.g.
+@bondevik2019].
+Finally, another approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline.
+If such methods are developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use.
 Such approaches would make less of an impact for the Brunlanes sites, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf.
@@
 When no other information is available, it can at present be recommended to use the empirically derived gamma distribution with a shape of `r round(gammafit$estimate[1], 3)` and scale of `r round(gammafit$estimate[2], 3)` (Figure \@ref(fig:results2)A) to characterise this relationship.
-Furthermore, while this remains to be formalised and explored further, it was also showed how the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data.
+Furthermore, while this remains to be formalised and explored further, it was also showed how the method could potentially be improved by including more information on both the topographic location of the sites and other temporal data.
+To the degree that making such a distinction is useful, this could be derived from assessments of both a qualitative and quantitative nature, with Bayesian inference forming a natural framework for integrating such considerations [e.g.
+@buck1996; @otarola-castillo2023].
 As the precision of the method is both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date, the impact of such additional information will also vary.
@@
 Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further evaluate and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted.
-Given its behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located.
-Furthermore, other dimensions related to the topographic location of the sites could be similarly explored.
-This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@helskog1978; @blankholm2020].
-This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range could possibly also have implications for the site location relative to the shoreline, depending on the topography [@helskog1978].
+For example, from Figure \@ref(fig:results2) it is clear that the Mesolithic sites have generally been located closer to the shoreline than the later sites.
+It was not attempted to explore this further here, given the constrained sample size and the good accuracy that was achieved with the parameters in use.
+However, the future addition of more data might give justification for using different models or parameter settings when dating sites from certain time intervals.
+Furthermore, following from its behavioural nature, it would also seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline the sites were located.
+This is illustrated here by the site phases associated with agricultural activity, marked in Figure \@ref(fig:results), which were all found to be located some distance from the sea.
+A wide range of different behavioural dimensions could potentially provide nuance to how the method should be applied.
+Other factors related to the topographic location of the sites could also be similarly explored.
+This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled.
+This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range and potential impact of storm surges could also have implications for the location of a site relative to the shoreline, depending on the topography [@helskog1978; @blankholm2020; @bondevik2019].
 The potential of exploring dimensions such as these was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline.
@@
 Sixthly, neither the question of how site limits are defined nor the elevation range over which these extend was given much consideration [@mjærum2022].
-Finally, the aggregation and division of settlement phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7% as representing unrelated occupation events.
+Finally, the aggregation and division of settlement phases at each site was here simply done by treating radiocarbon dates not overlapping at 99.7% as representing unrelated occupation events, which were then modelled by use of the Boundary and Sum functions in OxCal.
 This could also be handled differently [e.g.
@@
 
-Finally, this analysis employed a simulation approach to integrate multiple sources of spatio-temporal uncertainty.
-Here this was simply used to inform the question of the distance between sites and the shoreline.
+As the nature of the relationship between sites and sea is likely to vary temporally and geographically [e.g.
+@nyland2020], the proposed implementation and parametrisation of shoreline dating can not be expected to be directly applicable elsewhere.
+When this is combined with the fact that the rate of RSL-change also varies geographically and temporally [e.g.
+@svendsen1987], this means that the accuracy and precision of the method can also be expected to vary.
+However, the methodological framework used to evaluate the relationship between sites and sea is readily extendible to other regions of northern Scandinavia where reliable data on shoreline displacement is available, thus making such extensions feasible.
+Furthermore, the simulation approach used to integrate multiple sources of spatio-temporal uncertainty was here simply used to inform the question of the distance between sites and the shoreline.
 However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent to this reconstruction.

### 2022-11-09 analysis/supplementary-materials/supplementary.Rmd
Changes after peer review

@@
 
-In the figure below a series of standard models have been fit to the simulation results older than 2500 BCE (presented with negative values set to zero in Figure 7A in the main text).
-The models were fit by means of maximum likelihood estimation.
-Before fitting all models, the values have been adjusted by adding a constant of 0.001 to avoid zero values for the logistic, log normal, power law and gamma distributions.
-The models have only been fit to the positive values.
-This was done both when removing the negative values and when forcing these to zero, before adjusting by 0.001.
-As the difference between these two solutions was negligible and as the assumption here is that the negative values in actuality reflect a distance of zero (see also main text), the result of employing the latter approach is reported on here.
-The performance of the models were then compared by means of the Akaike information criterion (AIC) and the Bayesian (or Schwarz) information criterion (BIC).
-The AIC and BIC evaluate the degree to which the models fit to the data, while penalising for the number of model parameters to avoid over-fitting [e.g @burnham2002; for applications in archaeology see e.g.
-@eve2014; @timpson2021].
-As lower values point to a better model, it is evident from the results that the gamma distribution is the best candidate.
-The approach taken here could have benefited from a more sophisticated treatment of the zero-values, which is related to the fact that these are likely to be a mix of both exact zeros, the case when there is an actual intersection between sea and site, and, although probably to a far lesser extent, zeroes that result from the case when the distance between site and sea is below the detection limit due to the employed methods and the resolution of the spatial data [see e.g.
-@dunn2005; @helsel2005].
-In conclusion, however, the gamma represents a reasonable first approximation.
+Visualisation of the raw simulation results for vertical distances older than 2500 BCE, as well as the models fit for Table 1 in the main text.
+The table is also reproduced below.
+Note that the models have been fit when forcing negative values to zero and adding a constant of 0.001 to avoid zeroes.
+The data is presented with negative values set to zero in Figure 7A in the main text.
 
@@
 
-mcdf[,c(1,6, 3, 5, 4)] %>% 
-kable("latex", caption = "Model comparison using AIC and BIC. The models are listed in the order of performance.",
+mcdf[,c(1,6, 3, 5)] %>% 
+kable("latex", caption = "Same as Table 1. in the main text. Model comparison using AIC and BIC. The models are listed in the order of performance.",
          col.names = c("Model",
@@
                     "AIC",
-                    "BIC",
-                    "Log-liklihood"),
+                    "BIC"),
         booktabs = TRUE, escape = FALSE,
@@
 
+\pagebreak
 ## Numerical result of re-dating

### 2022-11-09 analysis/script/06simulation_results.R

@@ -18,8 +18,15 @@ library(AICcmodavg)
 
 # List all files except those starting with a number.
 # That is, all data files resulting from analysis in 05script.R
-datfiles <- grep("^[0-9]", list.files(here("analysis/data/derived_data")),
-     invert = TRUE, value = TRUE)
+datfiles <- grep("^[0-9]", list.files(here("analysis/data/derived_data"),
+                                      include.dirs = FALSE), invert = TRUE,
+                 value = TRUE)
+
+# Exclude directories
+datfiles <-setdiff(grep("^[0-9]", list.files(here("analysis/data/derived_data"),
+                                  include.dirs = FALSE), invert = TRUE,
+             value = TRUE), list.dirs(here("analysis/data/derived_data"),
+                                      recursive = FALSE, full.names = FALSE))
 
 # Create empty list to hold results
 results_list <- list()
@@ -27,6 +34,7 @@ results_list <- list()
 # Loop over, load results and assign them to the list
 for(i in 1:length(datfiles)){
   # Load results
+  print(paste(i, datfiles[i]))
   load(file.path(here("analysis/data/derived_data", datfiles[i])))
   results_list[[i]] <- output[!(names(output) %in%
                              c("datedat", "sitel", "sitecurve"))]
@@ -39,6 +47,7 @@ agr <- c("Nordby 1", "Kvastad A2", "Nauen A")
 # (Probably a smoother purrr solution to do all of this?)
 results <- list()
 for(i in 1:length(results_list)){
+  print(i)
   dat <- results_list[[i]]
   tmp <- list()
   for(j in 1:length(dat)){
@@ -55,6 +64,10 @@ for(i in 1:length(results_list)){
 
 # Collapse list of lists into a single data frame
 distances <- results %>% bind_rows()
+
+# Exclude Lunaveien and Dybdalshei 4 (see supplementary meterial)
+distances <- distances %>%
+  filter(!(sitename %in% c("Lunaveien", "Dybdalshei 2")))
 
 ##### Visualisation of distance measures across time ########
 
@@ -239,7 +252,7 @@ vertdistna$vertdist <- vertdistna$vertdist + 0.001
 expfit <- fitdistr(vertdist$vertdist, "exponential")
 expdat <- data.frame(
   x = vertdist$vertdist,
-  px = dexp(vertdist$vertdist, rate = expfit2$estimate))
[…]

### 2022-11-09 analysis/script/07shorelinedate.R

@@ -6,7 +6,7 @@ library(sf)
 library(hdrcde)
 
 # For reproducibility
-set.seed(1)
+set.seed(123)
 
 # Load required functions and prepared data
 source(here("analysis/script/04analysis_functions.R"))
@@ -18,6 +18,7 @@ load(here("analysis/data/derived_data/06data.RData"))
 # this is distributed as tiles which have to be merged by first
 # running 00dtm_prep.R
 dtm <- rast(here("analysis/data/derived_data/dtm10.tif"))
+dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
 
 # Example site (effectively stepping through the code from shoreline_date in
 # 04.functions.R)
@@ -29,60 +30,17 @@ sitecurve <- interpolate_curve(years = xvals,
                                dispdat = displacement_curves,
                                isodat = isobases)
 sitecurve$name <- sitename
+
+# retrieve for plotting
 siteelev <- terra::extract(dtm, vect(sitel), fun = mean)[2]
 
-inc <- seq(0.001, as.numeric(siteelev), 0.001)
 
-expdat <- data.frame(
-  offset = inc,
-  # Cumulative exponential distribution function
-  # px = pexp(inc, rate = expfit$estimate)) %>%
-  # px = plnorm(inc, meanlog = lognormfit$estimate[1],
-  #             sdlog = lognormfit$estimate[2]))  %>%
-  px = pgamma(inc, shape = gammafit$estimate[1],
-              rate = gammafit$estimate[2])) %>%
-  # Probability for each elevation increment is given by subtracting the
-  # previous cumulative probability
-  mutate(probs = px - lag(px, default =  dplyr::first(px))) %>%
-  tail(-1) %>%
-  filter(px < 0.99999) # Probability cut-off
-
-dategrid <- data.frame(
-  years = seq(-10000, 2000, 1),
-  probability = 0)
-
-for(i in 1:nrow(expdat)){
-  adjusted_elev <- as.numeric(siteelev - expdat$offset[i])
-  if(!(adjusted_elev > 0)) { # Do not allow sea-level to go below present
-    adjusted_elev <- 0.001
-  }
-  # Find lower date
-  lowerd <- round(approx(sitecurve[,"lowerelev"],
-                         xvals, xout = adjusted_elev)[['y']])
-
-  # Find upper date
-  upperd <- round(approx(sitecurve[,"upperelev"],
[…]

### 2022-11-09 analysis/script/08backwards_dating.R

@@ -29,7 +29,6 @@ datfiles <- datfiles[ !grepl("sitearea_temp", datfiles) ]
 
 # Create empty list to hold dat data
 posterior_list <- list()
-
 # Loop over, load and assign them to the list
 for(i in 1:length(datfiles)){
   # Load date data
@@ -48,52 +47,52 @@ rdates <- rdates %>%  group_by(site_name, group) %>%
   # If there is a sum, exclude the rest
   filter(class == 'sum' | all(class != 'sum')) %>%
   ungroup() %>%
-  filter(class != "aprior")
+  filter(class != "aprior") %>%
+  filter(!(site_name %in% c("Dybdalshei 2", "Lunaveien", "Frebergsvik C")))
 
 # Exclude sites not analysed
 # (see supplementary material for more on this)
 sites_sa <- sites_sa %>%
   filter(!(name %in% c("Dybdalshei 2", "Lunaveien", "Frebergsvik C")))
 
-# Specify for which sites to not shoreline date the site limit
-# (see supplementary material)
+# Specify for which sites to not shoreline date the site limit but rather
+# use features (see supplementary material)
 feature_sites <- c("Langangen Vestgård 7", "Vallermyrene 2")
-
 
 # Load shoreline date results (uncomment below to rerun)
 load(here("analysis/data/derived_data/08data.RData"))
 
-# shorelinedates <- list()
-# for(i in 1:nrow(sites_sa)){
-#   print(paste(i, sites_sa$name[i]))
-#   if(sites_sa$name[i] %in% feature_sites){
-#     shorelinedates[[i]] <- shoreline_date(sitename = sites_sa$name[i],
-#                                           elev = dtm,
-#                                           disp_curves = displacement_curves,
-#                                           sites = sites_sa,
-#                                           iso = isobases,
-#                                           exponential = FALSE,
-#                                           modelfit = gammafit,
-#                                           # expratio = expfit$estimate,
-#                                           siteelev = "mean",
-#                                           reso = 0.001,
-#                                           specified_elev = NA,
-#                                           sitelimit = FALSE,
-#                                           features = rcarb_sa)
-#   } else {
-#   shorelinedates[[i]] <- shoreline_date(sitename = sites_sa$name[i],
-#                                         elev = dtm,
-#                                         disp_curves = displacement_curves,
[…]

### 2022-11-09 analysis/paper/paper.Rmd
Further changes from peer review and final comments from supervisors

@@
 Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g.
-@berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @olsen1984; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018; @solheim2020b].
+@berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @olsen1984; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018; @solheim2020b; @mansrud2018].
 The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates.
@@
 @bergsvik2002; @bergsvik2012 for similarly nuanced considerations of the coastal settlement of Neolithic western Norway].
-Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c.
+Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shoreline with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c.
 2400 BCE.
@@
 Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date span is too wide to decisively verify this.
-It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would appear to have been sensible if the site was not shore bound [e.g.
+It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would appear to have been sensible if the site was not shore-bound [e.g.
 @jaksland2014; @johansen1963; @nummedal1923].
@@
 The shoreline displacement data used in this study are based on the so-called isolation basin method [e.g.
-@kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments.
+@kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations below the marine limit, and dating the transition from marine to lacustrine sediments.
 Each basin thus represents a high precision sea-level index point (SLIP) which are combined in a continuous time series for RSL-change adjusted to a common shoreline isobase.
@@
 As the displacement curves and their trajectory are quite complex constructs and the integrated result of both expert knowledge and more objectively quantifiable parameters, the geologists that have undertaken the studies have not found reason to assign variable uncertainty within the confidence envelopes of the displacement curves [@sørensen2014, 44; @romundset2018, 187].
-The reason for this is that the trajectory of the curves is not only based on radiometric dates, the uncertainty of which are well defined, but are for example also dependent on the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, as well as being based on expert knowledge of regional post-glacial geologic developments and local geomorphology, to name but a few factors [e.g.
+The reason for this is that the trajectory of the curves is not only based on radiometric dates, the uncertainty of which are well-defined, but are for example also dependent on the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, as well as being based on expert knowledge of regional post-glacial geologic developments and local geomorphology, to name but a few factors [e.g.
 @romundset2011; @romundset2018; @svendsen1987; for an alternative approach see @creel2022].
@@
 In the case that the sea polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon.
-While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
+While it is safe to assume that an archaeological site was not occupied when it was located below sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
 Negative values were therefore retained with the exception of the sites of Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
@@
 The highest simulated vertical distance among these is `r round(rs2$max, 1)`m and the median is `r round(rs2$median, 1)`m.
-While the sample size is limited, there are also a couple of sites located some distance from the shoreline just after 4000 BCE.
-Of the `r length(unique(r3$sitename))` sites associated with the period between 4000 and 2500 BCE, `r r3a25` sites have all vertical distance results above 25m.
+Furthermore, some deviation is evident from just after 4000 BCE we well.
+The sample size is limited, but of the `r length(unique(r3$sitename))` sites associated with the period between 4000 and 2500 BCE, `r as.english(r3a25)` sites have all vertical distance results above 25m.
 However, the median vertical distance of the results from this period is only `r round(rs3$median, 1)`m, indicating that while some sites have a markedly withdrawn location, most are still situated close to the shoreline.
@@
 However, the comparison is also unfair to the previously proposed dates for a few reasons.
-First, the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an undefined, but implied or explicit uncertainty range.
+First, the dates provided in the reports are typically stated to be a very rough estimate and are sometimes given as a point estimate with an undefined, but implied or explicit uncertainty range.
 Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change.
@@
 
-With a couple exceptions, the previously hypothesised dates and the ones achieved here appear to roughly correspond when it comes to the start date for the occupation of the sites.
-The most clear difference mainly pertains to the fact that the previously proposed date ranges are, without exception, more constrained than the 95% HDRs resulting from the proposed method.
+With a few exceptions, the previously hypothesised dates and the ones achieved here appear to roughly correspond when it comes to the start date for the occupation of the sites.
+The clearest difference mainly pertains to the fact that the previously proposed date ranges are, without exception, more constrained than the 95% HDRs resulting from the proposed method.
 Considering the right skew of the probability mass functions underlying the 95% HDRs and the general overlap for the start dates, these results could, with some danger of circularity, suggest that shoreline dating has generally been applied with a reasonable degree of success.
@@
 The results indicate that the spatial and temporal contingency of the method is better captured by the implementation suggested here, as is illustrated by the variation in the range of the 95% HDRs for the dates.
-In some cases the proposed method provides a fairly precise date and in others it offers little more than a *terminus post quem*.
+In some cases the proposed method provides a relatively precise date and in others it offers little more than a *terminus post quem*.
 This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf.
@@
 When no other information is available, it can at present be recommended to use the empirically derived gamma distribution with a shape of `r round(gammafit$estimate[1], 3)` and scale of `r round(gammafit$estimate[2], 3)` (Figure \@ref(fig:results2)A) to characterise this relationship.
-Furthermore, while this remains to be formalised and explored further, it was also showed how the method could potentially be improved by including more information on both the topographic location of the sites and other temporal data.
+Furthermore, while this remains to be formalised and explored further, it was also demonstrated how the method could potentially be improved by including more information on both the topographic location of the sites and other temporal data.
 To the degree that making such a distinction is useful, this could be derived from assessments of both a qualitative and quantitative nature, with Bayesian inference forming a natural framework for integrating such considerations [e.g.
@@
 For example, from Figure \@ref(fig:results2) it is clear that the Mesolithic sites have generally been located closer to the shoreline than the later sites.
-It was not attempted to explore this further here, given the constrained sample size and the good accuracy that was achieved with the parameters in use.
+It was not attempted to explore this further here, given the constrained sample size and the accuracy that was achieved with the parameters in use.
 However, the future addition of more data might give justification for using different models or parameter settings when dating sites from certain time intervals.
-Furthermore, following from its behavioural nature, it would also seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline the sites were located.
+Furthermore, following from its behavioural nature, it is also likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline the sites were located.
 This is illustrated here by the site phases associated with agricultural activity, marked in Figure \@ref(fig:results), which were all found to be located some distance from the sea.
 A wide range of different behavioural dimensions could potentially provide nuance to how the method should be applied.
+
 Other factors related to the topographic location of the sites could also be similarly explored.
 This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled.
-This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range and potential impact of storm surges could also have implications for the location of a site relative to the shoreline, depending on the topography [@helskog1978; @blankholm2020; @bondevik2019].
-The potential of exploring dimensions such as these was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline.
+This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range and potential impact of storm surges could also have implications for the location of a site relative to the shoreline, depending on the topography [@helskog1978; @bondevik2019].
+The potential of exploring such dimensions was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline.
 If patterns related to such locational patterns can be discerned and unpacked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information beyond their location is available.
+A mention should also be made here of the fact that catastrophic events such as tsunamis might also be of relevance [e.g.
+@blankholm2020; @nielsen2020; @nyland2021].
+Evidence for the impact of tsunamis in the Stone Age has not been identified in south-eastern Norway as of yet [see @romundset2015, 398], and might therefore not be of direct relevance to the coastal settlement in the region.
+However, the outburst flood resulting from the catastrophic drainage of the glacial lake Nedre Glomsjø around 8500--8000 BCE [@høgaas2019], located in Mid-Norway some 230 km north of present-day Oslo, could have had consequences for how the coast was utilised [@solheim2020b, 9].
 
@@
 When this is combined with the fact that the rate of RSL-change also varies geographically and temporally [e.g.
-@svendsen1987], this means that the accuracy and precision of the method can also be expected to vary.
+@svendsen1987], this means that the accuracy and precision of the method will also vary.
 However, the methodological framework used to evaluate the relationship between sites and sea is readily extendible to other regions of northern Scandinavia where reliable data on shoreline displacement is available, thus making such extensions feasible.
-Furthermore, the simulation approach used to integrate multiple sources of spatio-temporal uncertainty was here simply used to inform the question of the distance between sites and the shoreline.
+Furthermore, the simulation approach used to integrate multiple sources of spatio-temporal uncertainty was used here to inform the question of the distance between sites and the shoreline.
 However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent to this reconstruction.

### 2022-11-09 analysis/paper/paper.Rmd
Attempting to use Dockerfile

@@
 This information is not integrated in the approach outlined here, but it could justify further reducing the uncertainty ranges.
-
-Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does result in very precise dates.
+Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does nonetheless result in very precise dates.
+
 Above it was suggested that additional temporal data could be combined with the method to improve its precision.
@@
 If such methods are developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use.
-Such approaches would make less of an impact for the Brunlanes sites, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf.
+Such approaches would make less of an impact for the Brunlanes sites, where the 95% HDRs are already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf.
 Figure \@ref(fig:redate)).
@@
 Evidence for the impact of tsunamis in the Stone Age has not been identified in south-eastern Norway as of yet [see @romundset2015, 398], and might therefore not be of direct relevance to the coastal settlement in the region.
-However, the outburst flood resulting from the catastrophic drainage of the glacial lake Nedre Glomsjø around 8500--8000 BCE [@høgaas2019], located in Mid-Norway some 230 km north of present-day Oslo, could have had consequences for how the coast was utilised [@solheim2020b, 9].
+However, the outburst flood resulting from the catastrophic drainage of the glacial lake Nedre Glomsjø around 8500--8000 BCE [@høgaas2019], located in Mid-Norway some 230km north of present-day Oslo, could have had consequences for how the coast was utilised [@solheim2020b, 9].
 

### 2022-11-09 analysis/paper/paper.Rmd
Updated packages, checked that paper still compiles

@@
 # the purposes of visualisation).
-slope <- raster::terrain(raster(sitearea), 'slope')
-aspect <- raster::terrain(raster(sitearea), 'aspect')
+sraster <- raster(sitearea)
+crs(sraster) <- st_crs(sitearea)$proj4string
+
+slope <- raster::terrain(sraster, 'slope')
+aspect <- raster::terrain(sraster, 'aspect')
 hill <- raster::hillShade(slope, aspect, 40, 270)

### 2022-11-09 analysis/script/04analysis_functions.R

@@ -737,10 +737,13 @@ site_plot <- function(locationraster, sitelimit, dist, date_groups,
   bboxraster <- st_bbox(locationraster)
   anc <- as.numeric(c(bboxraster$ymin, bboxraster$xmax))
 
+  lraster <- raster(locationraster)
+  crs(lraster) <- st_crs(locationraster)$proj4string
+
   # Create hillshade
   # For some reason the terra version returned pathchy hillshade
-  slope <- raster::terrain(raster(locationraster), 'slope')
-  aspect <- raster::terrain(raster(locationraster), 'aspect')
+  slope <- raster::terrain(lraster, 'slope')
+  aspect <- raster::terrain(lraster, 'aspect')
   hill <- raster::hillShade(slope, aspect, 40, 270)
 
   # Now make sea-level NA
@@ -841,10 +844,13 @@ shore_plot <- function(locationraster, overlapgrid, sitelimit, dist,
   bboxraster <- st_bbox(locationraster)
   anc <- as.numeric(c(bboxraster$ymin, bboxraster$xmax))
 
+  lraster <- raster(locationraster)
+  crs(lraster) <- st_crs(locationraster)$proj4string
+
   # Create hillshade
   # For some reason the terra version returned pathchy hillshade
-  slope <- raster::terrain(raster(locationraster), 'slope')
-  aspect <- raster::terrain(raster(locationraster), 'aspect')
+  slope <- raster::terrain(lraster, 'slope')
+  aspect <- raster::terrain(lraster, 'aspect')
   hill <- raster::hillShade(slope, aspect, 40, 270)
 
   # Now make sea-level NA

### 2022-11-10 analysis/paper/paper.Rmd
Set up as.character(english::as.english()) to print numbers in written form a couple of places

@@
 library(oxcAAR)
-library(vapour)
 library(topoDistance)
@@
 Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age.
-After around 2500 BCE several sites are situated a considerable distance from the reconstructed shoreline, and while a couple remain horizontally and topographically close, most appear to be considerably elevated above the sea-level.
-Of the results from after 2500 BCE, which are associated with `r length(unique(r2$sitename))` sites, only `r as.english(r2a0)` has simulation results for vertical distance that includes zero.
+Of the results from after 2500 BCE, which are associated with `r length(unique(r2$sitename))` sites, only `r as.character(as.english(r2a0))` has simulation results for vertical distance that includes zero.
 The highest simulated vertical distance among these is `r round(rs2$max, 1)`m and the median is `r round(rs2$median, 1)`m.
-Furthermore, some deviation is evident from just after 4000 BCE we well.
-The sample size is limited, but of the `r length(unique(r3$sitename))` sites associated with the period between 4000 and 2500 BCE, `r as.english(r3a25)` sites have all vertical distance results above 25m.
+Furthermore, some deviation from the shoreline is evident from just after 4000 BCE as well.
+The sample size is limited, but of the `r length(unique(r3$sitename))` sites associated with the period between 4000 and 2500 BCE, `r as.character(as.english(r3a25))` sites have all vertical distance results above 25m.
 However, the median vertical distance of the results from this period is only `r round(rs3$median, 1)`m, indicating that while some sites have a markedly withdrawn location, most are still situated close to the shoreline.

### 2022-11-10 analysis/paper/paper.Rmd
Only using the DTM tile for Hegna Vest 1 in paper.rmd, instead of the entire DTM

@@
 ```{r load_dtm, include = FALSE}
-# This chunk tries to the load the edited DTM. Due to the file sizes involved,
-# this is distributed as tiles which have to be merged by running 00dtm_prep.R
-
-# Digital terrain model
-# dtm <- rast(here("analysis/data/derived_data/dtm10.tif"))
-dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
+# Load DTM tile for the example site Hegna Vest 1
+dtm <- rast(here("analysis/data/raw_data/tiled_dtm/output210d2773e79a4907927a671b18efd762-004-009.tif"))
 
@@
 These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022].
-While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent an upper constraining limit for the sea-level, as they would not have been in use when located under water.
+While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly
+represent an upper constraining limit for the sea-level, as they would not have been in use when located under water.
 It therefore seems that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geologic reconstruction of shoreline displacement in the area.

### 2022-11-10 analysis/script/00dtm_prep.R

@@ -1,16 +1,13 @@
 library(raster)
 
-# This script has to be run for it to be possible to compile paper.Rmd
-
 # As the raster data is very large, only the edited version is distributed with
 # the research compendium as a series of raster tiles. These have to be merged
 # for the rest of the scripts to run.
 
 # The unedited DTM is not available here, but the polygon data used to edit the
-# raster are and the script run to do the editing is available in 03dtm_edit.R,
+# raster are, and the script run to do the editing is available in 03dtm_edit.R,
 # making it possible to re-run the editing by retrieving raster data for the
 # study area from hoydedata.no
-
 
 # Get path to DTM tiles
 dtmtiles <- list.files(here::here("analysis/data/raw_data/tiled_dtm"),
@@ -22,8 +19,7 @@ names(tiles)[1:2] <- c('x', 'y')
 dtm <- do.call(raster::merge, tiles)
 
 # Save merged raster to the derived data folder
-dtmpath <-  here::here("analysis/data/derived_data")
+dtmpath <-  here::here("analysis/data/derived_data/dtm")
 edrast <- file.path(dtmpath, "dtm10.tif")
 writeRaster(dtm, edrast,
             overwrite = TRUE)
-

### 2022-11-10 analysis/script/05site_analysis.R

@@ -37,7 +37,7 @@ bmap <- st_read(here("analysis/data/raw_data/naturalearth_countries.gpkg"))
 # This tries to the load the edited DTM. Due to the file sizes involved,
 # this is distributed as tiles which have to be merged by first
 # running 00dtm_prep.R
-dtm <- rast(here("analysis/data/derived_data/dtm10.tif"))
+dtm <- rast(here("analysis/data/derived_data/dtm/dtm10.tif"))
 
 # Path to directory to hold smaller site area rasters
 siterpath <- here("analysis/data/derived_data/sitearea_temp")

### 2022-11-10 analysis/script/09redate.R

@@ -29,8 +29,10 @@ prevdates <- read.csv((here(
          end = end * -1) %>%
   dplyr::rename("site_name" = "name")
 
-# Read in raster
-dtm <- rast("/home/isak/phd/eaa_presentation/dtm10/dtm10.tif")
+# This tries to the load the edited DTM. Due to the file sizes involved,
+# this is distributed as tiles which have to be merged by first
+# running 00dtm_prep.R
+dtm <- rast(here("analysis/data/derived_data/dtm/dtm10.tif"))
 
 # Retrieve the Pauler sites
 brunlanes <- sitespol %>%

### 2022-11-10 README.Rmd
Update readme
Update README.Rmd

@@
 [![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/isakro/assessing.sealevel.dating/master?urlpath=rstudio)
+[![R-CMD-check](https://github.com/isakro/assessing.sealevel.dating/actions/workflows/R-CMD-check.yml/badge.svg)](https://github.com/isakro/assessing.sealevel.dating/actions/workflows/R-CMD-check.yml)
 
@@
 
+  - [:file\_folder: script](/analysis/script): R programming scripts for
+  analysing the data. The derived data is loaded in the R Markdown source 
+  document (below).
   - [:file\_folder: paper](/analysis/paper): R Markdown source document

### 2022-11-11 analysis/script/04analysis_functions.R
Updated how simulated sea-level is plotted. Attempting to run r script with Docker

@@ -420,7 +420,9 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs,
                       "year", "rcarb_cor", "sitename")
   results$rcarb_cor <- unique(na.omit(posteriorprobs$rcarb_cor))
   results$sitename <- unique(sitel$name)
-  seapolygons <- list(length = samps)
+
+  # seapolygons <- list(length = samps)
+  searasters <- list(length = samps)
 
   # If topographic paths are to be retrieved
   if(paths == TRUE){
@@ -449,7 +451,8 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs,
     classarea <- terra::classify(sitearea, rclmat)
     seapoly <- st_combine(st_as_sf(terra::as.polygons(classarea)))
 
-    seapolygons[[i]] <- seapoly
+    # seapolygons[[i]] <- seapoly
+    searasters[[i]] <- raster::raster(classarea)
 
     # If these are at the same coordinate (i.e. the polygons are overlapping),
     # return all values as 0
@@ -563,17 +566,23 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs,
   }
   }
 
+  rs <- stack(searasters)
+  searaster <- calc(rs, sum, na.rm = TRUE)
+
+  #seapolygons
   if(paths == TRUE){
     return(list(
       results = results,
-      seapol = st_as_sf(st_sfc(do.call(rbind, seapolygons)),
-                        crs = st_crs(sitel)),
+      searaster = searaster,
+      # seapol = st_as_sf(st_sfc(do.call(rbind, seapolygons)),
+      #                   crs = st_crs(sitel)),
       topopaths = topopaths))
   } else {
     return(list(
       results = results,
-      seapol = st_as_sf(st_sfc(do.call(rbind, seapolygons)),
-                        crs = st_crs(sitel))))
+      searaster = searaster))
+      # seapol = st_as_sf(st_sfc(do.call(rbind, searaster)),
+      #                   crs = st_crs(sitel))))
   }
   # # If there are no topopaths at all, do not try to return these
   # if(any(!(summary(topopaths)[,2]) %in% c("-none-", "sf"))){
@@ -703,11 +712,11 @@ apply_functions <- function(sitename, date_groups, dtm, displacement_curves,
[…]

### 2022-11-11 README.Rmd
Minor edits to main text and readme

@@
     and tables in this file)
-  - [:file\_folder: data](/analysis/data): Data used in the analysis.
+  - [:file\_folder: data](/analysis/data): Raw and derived data
   - [:file\_folder: figures](/analysis/figures): Plots and other

### 2022-11-11 analysis/paper/paper.Rmd
Updated how simulated sea-level is plotted. Attempting to run r script with Docker
Minor edits to main text and readme

@@
 The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m [@mapping2018].
-All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a research compendium following @marwick2018.
+All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in a version-controlled repository at <https://osf.io/7f9su/>, organised as a research compendium [see @Marwick2017; @marwick2018].
 
@@
 Furthermore, some deviation from the shoreline is evident from just after 4000 BCE as well.
-The sample size is limited, but of the `r length(unique(r3$sitename))` sites associated with the period between 4000 and 2500 BCE, `r as.character(as.english(r3a25))` sites have all vertical distance results above 25m.
+Of the `r length(unique(r3$sitename))` sites associated with the period between 4000 and 2500 BCE, `r as.character(as.english(r3a25))` sites have all vertical distance results above 25m.
 However, the median vertical distance of the results from this period is only `r round(rs3$median, 1)`m, indicating that while some sites have a markedly withdrawn location, most are still situated close to the shoreline.
@@
 These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022].
-While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly
-represent an upper constraining limit for the sea-level, as they would not have been in use when located under water.
+While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent an upper constraining limit for the sea-level, as they would not have been in use when located under water.
 It therefore seems that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geologic reconstruction of shoreline displacement in the area.
@@
 @blankholm2020; @nielsen2020; @nyland2021].
-Evidence for the impact of tsunamis in the Stone Age has not been identified in south-eastern Norway as of yet [see @romundset2015, 398], and might therefore not be of direct relevance to the coastal settlement in the region.
+Evidence for the impact of tsunamis in the Stone Age has not been identified in south-eastern Norway as of yet [see @romundset2015, 398; cf.
+@romundset2018; @sørensen2014], and might therefore not be of direct relevance to the coastal settlement in the region.
 However, the outburst flood resulting from the catastrophic drainage of the glacial lake Nedre Glomsjø around 8500--8000 BCE [@høgaas2019], located in Mid-Norway some 230km north of present-day Oslo, could have had consequences for how the coast was utilised [@solheim2020b, 9].

### 2022-11-14 analysis/paper/paper.Rmd
Changed how simulated sea-levels are visualised

@@
 
+# Site to directory holding simulated sea level raster
+searasterpath <- here("analysis/data/derived_data/sealevel_rasters")
+
 # Exclude Lunaveien, Dybdalshei 2 and Frebergsvik C (see supplementary material)
@@
 # This map was pre-compiled for convenience. The deglaciation data from DATED-1 
-# Hughes et al. 2015 is available as supplemental material to their paper:
+# (Hughes et al. 2015) is available as supplemental material to their paper:
 # https://doi.org/10.1111/bor.12142
@@
 The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m [@mapping2018].
-All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in a version-controlled repository at <https://osf.io/7f9su/>, organised as a research compendium [see @Marwick2017; @marwick2018].
+All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in a version-controlled repository at <https://osf.io/7f9su/>, organised as a research compendium following Marwick et al.
+[-@marwick2018; @Marwick2017].
 
@@
   geom_sf(data = seapoly, fill = "#dbe3f3", colour = NA) +
-  geom_sf(data = sitel, fill = NA, colour = "red") +
+  geom_sf(data = sitel, linewidth = 0.5, fill = NA, colour = "red") +
   #labs(title =  paste0("Sea-level at ", sealevel, " m above present")) +
@@
 # Map of simulated sea-levels, date group 1
-simsea_plot1 <- shore_plot(sitearea, output[[1]]$simsea, sitel,
+simsea_plot1 <- shore_plot(sitearea, sitel,
                            150, date_groups, s_tdist = 0.5, s_xpos = 1300,
-                           s_ypos = 800, s_bheight = 0.3, s_tsize = 3)
+                           s_ypos = 800, s_bheight = 0.3,
+                           s_tsize = 3, grouping = 1, searasterpath)
 
@@
 # Map of simulated sea-levels, date group 2
-simsea_plot2 <- shore_plot(sitearea, output[[2]]$simsea, sitel, 150,
+simsea_plot2 <- shore_plot(sitearea, sitel, 150,
                            date_groups, s_tdist = 0.5, s_xpos = 1300,
-                           s_ypos = 800, s_bheight = 0.3, s_tsize = 3)
+                           s_ypos = 800, s_bheight = 0.3,
+                           s_tsize = 3, grouping = 2, searasterpath)
 
@@
 These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022].
-While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent an upper constraining limit for the sea-level, as they would not have been in use when located under water.
+While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent
+an upper constraining limit for the sea-level, as they would not have been in use when located under water.
 It therefore seems that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geologic reconstruction of shoreline displacement in the area.
@@
 Note that the cut-off is done based on the calendar year associated with each distance value.
-Consequently, sites and site phases are only completely excluded if the entire posterior probability of the radiocarbon dates falls earlier than the cut-off.
+Consequently, sites and site phases are only completely excluded if the entire posterior probability of the radiocarbon dates falls later than the cut-off.
 Furthermore, the superimposed gamma distributions have been fit when adding a constant of 0.001 to the distance values and have been cut off on the y-axis for visualisaton.
@@
 
-wheres $\alpha$ is the shape and $\sigma$ the rate of the distribution, and $\Gamma(\alpha)$ denotes the gamma function.
+where $\alpha$ is the shape and $\sigma$ the rate of the distribution, and $\Gamma(\alpha)$ denotes the gamma function.
 This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $T$ for the occupation of the site.
@@
 The mean elevation of the site polygon is used to inform $E$ in the dating of the site.
-The gamma distribution in blue on the y-axis extends the full range of possible values for $E - D$ and has the parameters $\\alpha$ = ", round(gammafit$estimate[1], 3), " and $\\sigma$ = ", round(gammafit$estimate[1], 3),  " (see Figure 7A).
+The gamma distribution in blue on the y-axis extends the full range of possible values for $E - D$ and has the parameters $\\alpha$ = ", round(gammafit$estimate[1], 3), " and $\\sigma$ = ", round(gammafit$estimate[2], 3),  " (see Figure 7A).
 The red line marks the shoreline displacement curve interpolated to the site location.
@@
 Furthermore, formulating and visualising the method along the lines of how radiocarbon dates are treated, as was done here, does stand the chance of giving a veneer of radiometric accuracy to shoreline dating that is not warranted.
-That being said, the best chance we have of not throwing away precious temporal data, or exaggerating our handle on it, is arguably to rigorously evaluate the method using independent data such as radiocarbon dates, by offering a precise formulation of how it could be applied, by specifying what sources of uncertainty are accounted for and by making this process transparent through the open dissemination of underlying data and programming code.
+That being said, the best chance we have of not throwing away precious temporal data, or exaggerate our handle on it, is arguably to rigorously evaluate the method using independent data such as radiocarbon dates, by offering a precise formulation of how it could be applied, by specifying what sources of uncertainty are accounted for and by making this process transparent through the open dissemination of underlying data and programming code.
 

### 2022-11-14 analysis/script/04analysis_functions.R

@@ -413,7 +413,7 @@ bboxpoly <- function(feature, xy_adjustment) {
 # radiocarbon probability density function, site limit, and adjusted
 # displacement curve
 sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs,
-                             paths = FALSE){
+                             paths = FALSE, searasterpath, grouping = NULL){
 
   results <- data.frame(matrix(ncol = 6, nrow = samps))
   names(results) <- c("vertdist", "hordist", "topodist",
@@ -568,19 +568,26 @@ sample_shoreline <- function(samps, sitel, sitecurve, sitearea, posteriorprobs,
 
   rs <- stack(searasters)
   searaster <- calc(rs, sum, na.rm = TRUE)
+  crs(searaster) <- st_crs(sitel)$proj4string
+
+  writeRaster(searaster, file.path(searasterpath,
+                                  paste0(str_replace_all(str_replace_all(
+                                    unique(sitel$name)," ", "_"),
+                                                     "/", "_"), grouping,
+                                    ".tif")),
+              overwrite = TRUE)
+
 
   #seapolygons
   if(paths == TRUE){
     return(list(
       results = results,
-      searaster = searaster,
       # seapol = st_as_sf(st_sfc(do.call(rbind, seapolygons)),
       #                   crs = st_crs(sitel)),
       topopaths = topopaths))
   } else {
     return(list(
-      results = results,
-      searaster = searaster))
+      results = results))
       # seapol = st_as_sf(st_sfc(do.call(rbind, searaster)),
       #                   crs = st_crs(sitel))))
   }
@@ -647,7 +654,8 @@ sea_overlaps <- function(sitearea, seapolygons){
 # Function that combines all of the above
 apply_functions <- function(sitename, date_groups, dtm, displacement_curves,
                             isobases, nsamp = 1000, loc_bbox, siterpath,
-                            rcarbcor_true = FALSE, sitelimit = TRUE){
+                            searasterpath, rcarbcor_true = FALSE,
+                            sitelimit = TRUE){
 
   # Retrieve site features
   siter <- filter(rcarb_sa, site_name == sitename)
@@ -709,25 +717,24 @@ apply_functions <- function(sitename, date_groups, dtm, displacement_curves,
     }
 
     # Simulate sea-level and retrieve distances (takes time)
[…]

### 2022-11-14 analysis/paper/paper.Rmd
Added a more detailed background map for the study area

@@
 
+# Background map, more detailed for Norway
+bmap2 <- st_read(here('analysis/data/raw_data/naturalearth_norway.gpkg'))
+
 # Point features with city names
-cpts <- st_read(here('analysis/data/raw_data/central_points.gpkg'))
+cpts <- st_read(here('analysis/data/raw_data/naturalearth_cities.gpkg'))
 
@@
 
-bound_reproj <- st_transform(bboxsitespoly, st_crs(bmap))
-bmap2 <- bmap %>%
+bound_reproj <- st_transform(bboxsitespoly, st_crs(sites_sa))
+bmap2 <- bmap2 %>%
     filter(st_intersects(., bound_reproj, sparse = FALSE))
-bmap_reproj <- st_transform(bmap2, st_crs(sites_sa))
+bmap2_reproj <- st_transform(bmap2, st_crs(sites_sa))
 
@@
 site_map <- ggplot() +
-  geom_sf(data = bmap_reproj, fill = "grey", colour = NA) +
+  geom_sf(data = bmap2_reproj, fill = "grey", colour = NA) +
   geom_sf(data = isobases, aes(colour = name)) +
@@
                                        "Bjørnebu", "Skoppum")))
+rname_disp$name <- factor(as.factor(rname_disp$name),
+                          levels = c("Bjørnebu", "Hanto",
+                                     "Gunnarsrød", "Skoppum"))
 
@@
 
+dispdat <- rbind(rname_disp, sitecurve)
+
 # Plot of dates
@@
                                os_tdist = 0.03, os_xpos = 20000, os_ypos = 8000,
-                               os_bheight = 0.02, os_tsize = 3)
+                               os_bheight = 0.02, os_tsize = 3,
+                              detailed_map = bmap2)
 
 curve_plot <- ggplot2::ggplot() + 
-              geom_line(data = rname_disp, 
+              geom_line(data = dispdat, 
                       aes(x = years, y = upperelev, col = name), alpha = 0.4) +
-  geom_line(data =  rname_disp, 
+  geom_line(data =  dispdat, 
                       aes(x = years, y = lowerelev, col = name), alpha = 0.4) +
@@
 @berg-hansen2022[644] made a similar observation in their assessment of the correspondence between shoreline displacement and radiocarbon dates from these sites.
-These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022].
+The sites are recently excavated, well-dated and situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022].
 While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent
@@
 To characterise this relationship, a series of standard models for distributions with a right skew have been fit to the simulation results for vertical distance older than 2500 BCE (Figure \@ref(fig:results2)A) by means of maximum likelihood estimation (Table \@ref(tab:tab1)).
-As most of the models only accept positive values a constant of 0.001 was added to avoid values of zero.
+As most of the models only accept positive values, a constant of 0.001 was added to avoid values of zero.
 It was attempted to both remove negative values and force these to zero before adding the constant.
@@
 These are the parameters for the gamma distribution identified when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A) and are the parameters used in all applications of the proposed method that follow below.
-For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, which, following from the adjustment of the values for fitting the gamma distribution, starts from 0.001m.
+For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, which, following from the adjustment of the values for fitting the compared models, starts from 0.001m.
 The gamma distribution is stepped through in it's cumulative form, where the probability from the previous 0.001m step is subtracted from the probability at the current step.
@@
 The gamma distribution in blue on the y-axis extends the full range of possible values for $E - D$ and has the parameters $\\alpha$ = ", round(gammafit$estimate[1], 3), " and $\\sigma$ = ", round(gammafit$estimate[2], 3),  " (see Figure 7A).
-The red line marks the shoreline displacement curve interpolated to the site location.
+The red envelope marks the shoreline displacement curve interpolated to the site location.
 The resulting shoreline date in grey is underlined with the 95\\% HDR in black."), out.width = "500px", fig.align="center", regfloat=TRUE}
@@
 However, as some of the 95% HDRs extend well beyond major chronological divisions, even into the Iron Age, it is also clear that some of these could be severely and securely constrained with only cursory reference to typology.
-While this would be trivial in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed.
+While this would be trivial in some cases, the nature and uncertainty inherent to the method still means that this is arguably an exercise that should be explicitly performed.
 This also points to the possibility of drawing on other temporal data to further improve the precision of the dates that can be achieved with shoreline dating.

### 2022-11-14 analysis/script/04analysis_functions.R

@@ -800,7 +800,8 @@ site_plot <- function(locationraster, sitelimit, dist, date_groups,
 }
 
 overview_plot <- function(background_map, sitelimit, sites, isobases,
-                          os_tdist, os_xpos, os_ypos, os_bheight, os_tsize) {
+                          os_tdist, os_xpos, os_ypos, os_bheight, os_tsize,
+                          detailed_map = NA) {
   bboxsites <- st_bbox(sites)
   bboxsites[1] <- bboxsites[1] - 15000
   bboxsites[3] <- bboxsites[3] + 15000
@@ -808,11 +809,18 @@ overview_plot <- function(background_map, sitelimit, sites, isobases,
   bboxsites[4] <- bboxsites[4] + 5000
   bboxsitespoly <- st_as_sf(st_as_sfc(bboxsites))
 
-  bound_reproj <- st_transform(bboxsitespoly, st_crs(background_map))
-  bmap2 <- background_map %>%
-    filter(st_intersects(., bound_reproj, sparse = FALSE))
-  bmap_reproj <- st_transform(bmap2, st_crs(sites))
 
+  if(!(any(is.na(detailed_map)))){
+    bound_reproj <- st_transform(bboxsitespoly, st_crs(detailed_map))
+    bmap2 <- detailed_map %>%
+      filter(st_intersects(., bound_reproj, sparse = FALSE))
+  }else{
+    bound_reproj <- st_transform(bboxsitespoly, st_crs(background_map))
+    bmap2 <- background_map %>%
+      filter(st_intersects(., bound_reproj, sparse = FALSE))
+  }
+
+  bmap_reproj <- st_transform(bmap2, st_crs(sites))
   anc <- as.numeric(c(bboxsites$ymin, bboxsites$xmax))
 
   ggplot() +
@@ -1068,7 +1076,9 @@ plot_results <- function(sitename, sitelimit, datedata, sitearea,
     }
 }
 
-# Shoreline date, using exponential function for elevation offset
+# Shoreline date. Defaults to using exponential, as this was used originally.
+# However, the dates performed for the paper are done with exponential set to
+# FALSE as the gamma was later identified as the better fit.
 shoreline_date <- function(sitename, elev = dtm,
                            disp_curves = displacement_curves,
                            sites = sites_sa,

### 2022-11-15 analysis/paper/paper.Rmd
Revised manuscript

@@
   comment = "#>",
+  fig.hide = TRUE,
   fig.path = "../figures/",
@@
 As this process is the result of glacial loading, the rate of uplift is faster towards the centre of the ice sheet relative to the distal aspects.
-Thus, there is differential glacio-isostatic impact to a site’s location depending on it's relation to the ice sheet’s centre of mass, leading some areas on the outer coast to have had a more stable RSL or been subject to marine transgression [e.g.
+Thus, there is differential glacio-isostatic impact to a site’s location depending on its relation to the ice sheet’s centre of mass, leading some areas on the outer coast to have had a more stable RSL or been subject to marine transgression [e.g.
 @romundset2015; @svendsen1987].
@@
 
-```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from c.
+```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000-year intervals from c.
 17--8 thousand years (kyr) BCE.
@@
 The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, it has been characterised as a relative dating method for sites located within a constrained geographical area, or it has been argued to offer no more than an earliest possible date for when a site could have been in use [see review by @nordqvist1999].
-The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @glorstad2002; @glorstad2003; @glorstad2004; @jaksland2001; @jaksland2012; @jaksland2012a; @jaksland2014a; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below].
-Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time.
+The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @glorstad2002; @glorstad2003; @glorstad2004; @jaksland2001; @jaksland2012; @jaksland2012a; @jaksland2014a; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b; @solheim2013 and below].
+Recently the method has also been used independently to date a larger number of sites to get a general impression of site frequency over time.
 This is done by aggregating point estimates of shoreline dates in 100-, 200- or 500-year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020].
@@
 Not remnants of stays as ephemeral to only be discernible by isolated features or dubious ^14^C-dates.
-The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent upon, and follow the discretion of the original excavation reports.
+The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent upon and follow the discretion of the original excavation reports.
 
@@
 The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m [@mapping2018].
-All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in a version-controlled repository at <https://osf.io/7f9su/>, organised as a research compendium following Marwick et al.
-[-@marwick2018; @Marwick2017].
+All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in a version-controlled repository at <https://osf.io/7f9su/>, organised as a research compendium following Marwick [@marwick2018; -@Marwick2017].
 
@@
 For the sites all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021].
-Radiocarbon dates associated with each site were then grouped if their date ranges intersected at 99.7% probability, meaning these were effectively taken to be associated with the same event, here termed settlement or site phase.
+Radiocarbon dates associated with each site were then grouped if their date ranges intersected at 99.7% probability, meaning these were effectively taken to be associated with the same occupation event, here termed settlement or site phase.
 In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed using the Sum function.
@@
 While it is safe to assume that an archaeological site was not occupied when it was located below sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates.
-Negative values were therefore retained with the exception of the sites of Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
+Negative values were therefore retained except of for the sites Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).
 
@@
 This evaluation is presented in the supplementary material.
-Hovland 5 was chosen for this assessment as it has an imprecise age and is located in area of quite complex topography.
+Hovland 5 was chosen for this assessment as it has an imprecise age and is located in area of quite complex topography [@mansrud2013].
 
@@
 ```{r example-2, echo = FALSE, fig.cap = "Example of a single simulation run on the site Hegna vest 1.
-A) The simulation starts by drawing a single year, weighted by the the posterior probability distribution.
+A) The simulation starts by drawing a single year, weighted by the posterior probability distribution.
 B) This then corresponds to an elevation range on the interpolated displacement curve.
@@
 @berg-hansen2022[644] made a similar observation in their assessment of the correspondence between shoreline displacement and radiocarbon dates from these sites.
-The sites are recently excavated, well-dated and situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022].
-While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent
-an upper constraining limit for the sea-level, as they would not have been in use when located under water.
+The sites are recently excavated, well-dated and are situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022].
+While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent an upper constraining limit for the sea-level, as they would not have been in use when located under water.
 It therefore seems that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geologic reconstruction of shoreline displacement in the area.
@@
 Consequently, sites and site phases are only completely excluded if the entire posterior probability of the radiocarbon dates falls later than the cut-off.
-Furthermore, the superimposed gamma distributions have been fit when adding a constant of 0.001 to the distance values and have been cut off on the y-axis for visualisaton.
+Furthermore, the superimposed gamma distributions have been fit when adding a constant of 0.001 to the distance values and have been cut off on the y-axis for visualisation.
 The gamma distribution in A forms the basis for the analysis to follow, but a version has also been fit to the vertical distances in B to further illustrate the difference between the distributions."), out.width = "525px", fig.align="center", regfloat=TRUE}
@@
 For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, which, following from the adjustment of the values for fitting the compared models, starts from 0.001m.
-The gamma distribution is stepped through in it's cumulative form, where the probability from the previous 0.001m step is subtracted from the probability at the current step.
+The gamma distribution is stepped through in its cumulative form, where the probability from the previous 0.001m step is subtracted from the probability at the current step.
 This probability is then divided equally across the individual calendar years in the range between the lower and the upper limit of the displacement curve at the current 0.001m step.
@@
 Jaksland's dates are given a flat 200- and 50-year uncertainty range starting from what he gives as the earliest possible date.
-The 200 year uncertainty range is given if the sites were to be considered in isolation, while his argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other.
+The 200-year uncertainty range is given if the sites were to be considered in isolation, while his argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other.
 Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they do not overlap.
@@
 One potential way to do this could be through the analysis of phosphate concentrations in soils, which has the potential to offer insights on the likely position of the shoreline when a site was in use [@ilves2011].
-This has been done in the Baltic sea region [e.g.
+This has been done in the Baltic Sea region [e.g.
 @broadbent1979; @sundström2006; @ilves2011], but has yet to provide reliable results in Norway [e.g.
@@
 Dates provided by Jaksland (2014) are plotted in red.
-The box indicates a 50 year uncertainty range which in combination with the red line extends 200 years.
+The box indicates a 50-year uncertainty range which in combination with the red line extends 200 years.
 B) Map showing the centroids of the Pauler sites and Sky 1.
@@
 As the nature of the relationship between sites and sea is likely to vary temporally and geographically [e.g.
-@nyland2020], the proposed implementation and parametrisation of shoreline dating can not be expected to be directly applicable elsewhere.
+@nyland2020], the proposed implementation and parametrisation of shoreline dating cannot be expected to be directly applicable elsewhere.
 When this is combined with the fact that the rate of RSL-change also varies geographically and temporally [e.g.

### 2022-12-05 analysis/paper/paper.Rmd
Minor edits

@@
   @telford2004
+  @fossum2017a
 ---
@@
 The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m [@mapping2018].
-All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in a version-controlled repository at <https://osf.io/7f9su/>, organised as a research compendium following Marwick [@marwick2018; -@Marwick2017].
+All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in a version-controlled repository at <https://doi.org/10.17605/osf.io/7f9su>, organised as a research compendium following Marwick [@marwick2018; -@Marwick2017].
 
@@
 
-where $\alpha$ is the shape and $\sigma$ the rate of the distribution, and $\Gamma(\alpha)$ denotes the gamma function.
+where $\alpha$ is the shape and $\sigma$ the scale of the distribution, and $\Gamma(\alpha)$ denotes the gamma function.
 This can then be coupled with the trajectory of relative sea-level change to find the corresponding calendar date $T$ for the occupation of the site.
